### 2013-03-11 envelope.tex
initial import

\documentclass{article}
\usepackage{amsmath,amssymb}
\usepackage{fullpage}
\usepackage{graphicx}

\newcommand{\mudn}{\mu_\downarrow}
\newcommand{\muup}{\mu_\uparrow}

\begin{document}

\begin{center}
\includegraphics[width=\textwidth]{hcg-trees}
\end{center}
\vskip 2cm

We observe alleles that are polymorphic in humans (left branch),
and the same allele observed in one chimp (middle) and one gorilla (right).
Depicted are the four most parsimonious scenarios: 
in any case it requires one mutation;
and in other cases it requires at least two more rare(-ish) events:
incomplete lineage sorting (ILS) and/or another mutation.
The likelihoods are, I think:
\begin{align}
  p_A > p_B > p_C > p_D ,
\end{align}
(see below); but $D$ will create the bump on the right.

\begin{enumerate}

\item[A] The standard model, mutation sometime in the tree leading to modern humans.
Frequency distribution skewed towards low frequency.

\item[B] Two mutations, one leading to chimps and humans; the other on the gorilla lineage.  Requires the polymorphism in humans to be fairly old, so has a flat frequency distribution.

\item[C] One mutation, but long ago.  Requires the polymorphism to be old, again.

\item[D] Two mutations, one in chimp-gorilla and one in humans.  The red allele, being recently derived, has a skew towards low frequencies; which we miscall as high derived frequencies.

\end{enumerate}

\section{Asymmetric mutation}

Suppose possibly asymmetric mutation rates: $\muup$ from black to red,
and $\mudn$ from red to black.
Suppose that the probability of ILS in the whole human lineage is $\eta_C$,
and in the human-chimp ancestor is $\eta_G$.
Finally, suppose that the typical coalscence times in human, human-chip ancestor, and human-chimp-gorilla ancestor are, respectively, $T_H$, $T_C$, and $T_G$.
The the relative rates of $A$--$D$ are, roughly:
\begin{align}
A&: \mudn T_H & B&: \muup^2 T_C T_G \eta_C \\
C&: \muup T_G \eta_C \eta_G & D&: \muup^2 T_H T_G \eta_G  .
\end{align}

The human-chimp divergence is like 7 mya, 
and the human-chimp-gorilla only a couple more million years back;
and ancestral effective population sizes are thought to have been $\sim 10 \times$ larger than the human effective population size; so $\eta_G$ is not so small at all: 
like, 0.2?
On the other hand, we have lots of samples, so the chance of more than one lineage making it back to the human-chimp ancestor is pretty good, maybe of similar order?
If $\muup/\mudn \approx 10$, this suggests that the ratio of $A$ to $C$ is about 10:1.

In $C$ and $D$ the relevant numbers of generations for mutation should be around $10^5$; for $T_H$ in $D$ this is thanks to the large number of samples; 
and for $T_G$ in both this is thanks to the large effective population size of the ancestral population.  
Even if $\muup \approx 10^{-7}$, this still leaves $D$ about 1000 times less likely than $A$.

As for $B$, note that the mutation on gorilla has the full branch to occur on, i.e.~$\sim10^7$ years; so $\muup T_G$ is only a small factor.
Because there are fewer lineages to mutate, the mutation in human-chimp is less likely than the one occurring in $A$ by ad additional factor; and the incomplete lineage sorting is another factor. 
Incorporating $\muup/\mudn\approx10$, this suggest a $A:D$ ratio of only about 10:1.

%%%%%%%%
\section{Selection}

Selection (or biased gene conversion) will do several things:
effectively increase the mutation rate of the beneficial allele;
make ILS \emph{less} likely;
and distort the frequency spectrum.

\ldots effects on $A$--$D$?
\end{document}

### 2013-05-13 context-dependence.tex
tree inference

\documentclass{article}
\usepackage{amsmath,amssymb}
\newcommand{\E}{\mathbb{E}}
\renewcommand{\P}{\mathbb{P}}

\begin{document}

\section*{Introduction}

Interacting particle systems with neighborhood structure: transition probabilities generally not expressable (except TASEP).
Markov random fields with Glauber dynamics a case of this.

Example: Ising model.

Context-dependent substitution processes are an example of interest for phylogenetics and evolutionary inference.
Usually dealt with either by using larger and larger windows (but, only up to four bp),
or by doing MCMC over additional information (entire history of changes).
Also, Markov-along-the-genome method.

The difficulty comes from the fact that the rates of change at any one site depend in principle on arbitrarily far away sites,
so transition matrix is ginormous.

Goal: parameter inference, given before-after observations, or else observations descended from a common ancestor.

Here restrict to one-dimensional neighborhood structure,
to make it easier.

\section*{Notation}

\end{document}

### 2013-05-13 simple-tree-inference.R

@@ -1,0 +1,119 @@
+# Evolve from a common root
+source("codons.R")
+source("codon-inference-fns.R")
+source("sim-context-fns.R")
+
+# maximum size of pattern (for simulation)
+patlen <- 2
+mutpats <- list( 
+        fwd=combn( bases, 2, simplify=FALSE ), 
+        rev=lapply( combn( bases, 2, simplify=FALSE ), rev),
+        gc=list( c("GC","TC") ) ,
+        ga=list( c("GA","CA") ) ,
+        ta=list( c("AT","GT") ) 
+    ) 
+mutrates <- runif( length(mutpats) )*1e-8
+selpats <- c( "[GC]", "[AT]" )
+selcoef <- runif( length(selpats) )*1e-3
+# other params
+Ne <- c(1e4,1e4)
+seqlen <- 1e4
+tlen <- 3e7  # total length 6e7 gives lots of transitions (but still signal); 1e7 not so many
+branchlens <- c(1,1)
+
+# tlen <- .1e7: 2.9 sec ; 3e7: 188 sec
+
+initfreqs <- c(.3,.3,.2,.2)
+initseq <- rinitseq(seqlen,bases,basefreqs=initfreqs)
+system.time( 
+        simseqs <- lapply(1:2, function (k) simseq( seqlen, tlen*branchlens[k], patlen=patlen, mutpats=mutpats, selpats=selpats, mutrates=mutrates, selcoef=selcoef, Ne=Ne[k], initseq=initseq ) ) 
+    )
+
+thisone <- formatC( floor(runif(1)*1e6) , digits=6,flag='0')
+now <- format(Sys.time(), "%Y-%m-%d-%H-%M")
+save( thisone, now, patlen, mutpats, selpats, selcoef, Ne, seqlen, tlen, branchlens, initfreqs, simseqs, file=paste(now,thisone,"selsims.RData",sep='') )
+
+
+####
+# Inference.
+lwin <- 0
+rwin <- 2
+win <- 1
+winlen <- lwin+win+rwin
+
+genmatrix <- makegenmatrix( mutpats, selpats, patlen=winlen)
+genmatrix@x <- update(genmatrix,mutrates*tlen,selcoef,Ne)
+projmatrix <- collapsepatmatrix( ipatterns=rownames(genmatrix), lwin=lwin, rwin=rwin )
+subtransmatrix <- computetransmatrix( genmatrix, projmatrix, names=TRUE )
+
+counts <- list(
[…]

### 2013-05-14 context-dependence.tex
writeup

@@
 \usepackage{amsmath,amssymb}
+\usepackage{amsthm}
 \newcommand{\E}{\mathbb{E}}
 \renewcommand{\P}{\mathbb{P}}
+\newcommand{\calS}{\mathcal{S}}  % set of states
+\newcommand{\calT}{\mathcal{T}}  % set of transition triples
+\newcommand{\nA}{\mbox{A}}  % nucleotides:
+\newcommand{\nC}{\mbox{C}}
+\newcommand{\nG}{\mbox{G}}
+\newcommand{\nT}{\mbox{T}}
+\newcommand{\join}{\oplus}  % matches
+\newcommand{\st}{\colon}  % such that
+
+\theoremstyle{definition}
+\newtheorem{example}{Example}[section]
 
@@
 
+The common framework of the above examples is a one-dimensional grid of $n$ sites,
+and each site is labeled with one of a finite set of states.
+These change with time -- so, write $X_i(t)$ for the state of the $i^\mathrm{th}$ site at time $t$,
+where each $X_i(t) \in \calS$, the set of possible states.
+(To avoid edge effects, we sometimes arrange these sites in a circle, so the indices are interpreted $\mod n$.)
+The time evolution is Markov, and is determined by a set of transition rates and associated patterns.  
+``Patterns'' are sequences of states, and we let $|u|$ denote the length of the pattern $u$.
+Transition rates are assumed to be local and homogeneous, so for example saying ``pattern $u$ changes to $v$ at rate $\mu$''
+means that if $(X_i(t), \ldots, X_{i+\ell}(t)) = (u_1, \ldots, u_\ell)$ (where $\ell=|u|=|v|$),
+then $\P\{(X_i(t+dt), \ldots, X_{i+\ell}(t+dt)) = (v_1, \ldots, v_\ell) = \mu dt + o(dt)$,
+regardless of the location $i$ along the sequence.
+We call such a $(\mu,u,v)$ a ``transition triple''.
+The set of all transition triples $\calT = \{ (\mu_j,u_j,v_j) \}_{j=1}^{n_T}$ 
+determines the time evolution of the process.
+
+It will often be convenient, notationally and computationally, to avoid edge effects
+by treating the $n$ sites as a circle, and treating all indices $\mod n$.
+
+\begin{example}[TASEP]
+  The \emph{Totally Asymmetric Simple Exclusion Process} labels each site as ``empty'' or ``occupied'' (0 or 1, respectively), 
+  and says that each occupied site, independently at rate $\lambda$, checks the site to the right to see if it is empty,
+  and if it is, moves there.  
+  This process therefore has only one transition triple:
+
+  \begin{center}
+    \begin{tabular}{c@{\quad$\to$\quad}c@{\quad at rate\quad }c}
+      $u$  &  $v$  &  $\mu$  \\
+      \hline
+      01  &   10   &  $\lambda$
+    \end{tabular}
+  \end{center}
+
+  This only has one parameter, the speed.  
+  Note, however, that it may not be entirely trivial to estimate the speed,
+  since (on the circle) it may not be obvious which particle moved where.
+
+\end{example}
+
+\begin{example}[CpG mutation]
+  A sequence of genome can be written using A, C, G, and T;
+  in the most general model of single-site mutation, each of the 12 possible transitions occurs at its own rate.
+  Furthermore, it is a well-known observation that in many species, and certain contexts, 
+  a C followed by a G has an additional, higher rate with either the C changing to a T or the G changing to an A.
+  This simplest model of context-dependent substitution rates in a genome sequence has rates
+
+  \begin{center}
+    \begin{tabular}{c@{\quad$\to$\quad}c@{\quad at rate\quad }cc}
+      $u$  &  $v$  &  $\mu$  & \\
+      \hline
+      $x$  &  $y$  &  $m_{xy}$ & \qquad $x \neq y \in \{\nA,\nC,\nG,\nT\}$  \\
+      CG   &  TG   &  $\gamma$ & \\
+      CG   &  CA   &  $\gamma$ &  
+    \end{tabular}
+  \end{center}
+
+  This has 13 parameters.
+  Note that there are no conserved quantities here, unlike in TASEP,
+  which we could imagine as indestructible particles moving.
+
+\end{example}
+
+\begin{example}[Ising model with Glauber dynamics]
+  In the Ising model, each site is labeled as either ``up'' or ``down'' ($+1$ or $-1$ respectively),
+  imagined as magnetic dipoles,
+  and the energy associated with a given state $x$ is $H(x) = - \beta \sum_i x_i x_{i+1} + \gamma \sum_i x_i$.
+  The associated stationary distribution on configurations is proportional to $\exp(-H(x))$.
+  One way to add temporal dynamics that preserve the distribution 
+  is to say that each site, independently at rate $\lambda$,
+  forgets its spin, 
+  and reconfigures to a state chosen with probability proportional to the stationary probability of the resulting configuration.
+  This is known as ``Glauber dynamics'', and ignoring transitions that don't change the state,
+  we get that if patterns $u$ and $v$ differ at only one site, then $u$ changes to $v$ at rate $1/(1+\exp(H(u)-H(v)))$, so
+
+  \begin{center}
+    \begin{tabular}{c@{\quad$\to$\quad}c@{\quad at rate\quad }c}
+      $u$  &  $v$  &  $\mu$  \\
+      \hline
+      $+++$  &   $+-+$   &  $\lambda/1+e^{2\beta - \gamma})$ \\
+      $++-$  &   $+--$   &  $\lambda/(1+e^{-\gamma})$ \\
+      $+-+$  &   $+++$   &  $\lambda/(1+e^{-2\beta + \gamma})$ \\
+      $+--$  &   $++-$   &  $\lambda/(1+e^{\gamma})$ \\
+      $-++$  &   $--+$   &  $\lambda/(1+e^{-\gamma})$ \\
+      $-+-$  &   $---$   &  $\lambda/(1+e^{-2\beta - \gamma})$ \\
+      $--+$  &   $-++$   &  $\lambda/(1+e^{\gamma})$ \\
+      $---$  &   $-+-$   &  $\lambda/(1+e^{2\beta + \gamma})$ 
+    \end{tabular}
+  \end{center}
+
+  This has three parameters: the speed $\lambda$, the inverse temperature $\beta$, and the strength of magnetization $\gamma$
+  (here scaled by temperature).
+
+\end{example}
+
+\paragraph{The generator matrix}
+For clarity, we should define how the set of transition rates determines the transition rate matrix,
+the $|\calS|^n \times |\calS|^n$ matrix $G(n)$ whose $(x,y)^\text{th}$ entry gives the instantaneous rate 
+with which the process in state $x$ jumps to state $y$.
+Let $x_i^{(\ell)} = (x_i, x_{i+1}, \ldots, x_{i+\ell-1})$ be the subsequence of length $\ell$ beginning at location $i$,
+and for each $1\le i \le n$, and patterns $u,v \in \calS^\ell$ define the relation
+\[
+x \xrightarrow{i,u,v} y \qquad \text{iff} \qquad \begin{cases}
+  x_j = y_j \quad &\text{for } j<i \\
+  x_{i+k} = u_k \quad &\text{for } k < \ell \\
+  y_{i+k} = v_k \quad &\text{for } k < \ell, and \\
+  x_j = y_j \quad &\text{for } j\ge i+\ell ,
+\end{cases}
+\]
+i.e.\ if $x$ and $y$ match except for in $i,i+1,\ldots,i+\ell$, where $x$ matches with $u$ and $y$ matches with $v$.
+Interpet these indices $\mod n$, even for short sequences, a point we will come back to.
+We'll also need the set of transitions that can move from $x$ to $y$ at a given location: let  $J(i,x,y) = \{ j \st x \xrightarrow{i,u_j,v_j} y \}$.
+Then, the rate $G(n)_{x,y}$ is the sum of all matching transition rates,
+namely
+\begin{align*}
+  G(n)_{x,y} = \sum_{i=1}^n \sum_{j \in J(i,x,y)}  \mu_j ,
+\end{align*}
+and if there are no triples $(\mu,u,v)$ with $x \xrightarrow{i,u,v} y$ for some $i$, then $G(n)_{x,y}=0$.
+In principle, this gives us the transition probabilities for the process:
+\begin{align} \label{eqn:full_likelihood}
+  p_n(t;x,y) := \P\{ X(t) = y \mid X(0) = x \} = (e^{tG_n})_{xy} .
+\end{align}
+
+\section{Inference}
+
+The problem at hand is to infer the parameters of the model, given the state at time 0 and later at time $t$.
+If $n$ is large enough and $t$ is not too large,
+this should be feasible.
+(If $t$ is too large, the process may ``saturate'' --
+if there are many changes at most sites, then in at least most models, 
+we will lost most of the information about the dynamics,
+retaining information only about the parameters that affect the stationary distribution.
+In TASEP, we lose all information as it approaches stationarity,
+in the Ising model we have information about $\beta$ and $\gamma$ but not $\lambda$,
+and in the CpG model it is not immediately clear.
+
+But, how can we extract the information?
+These are Markov processes, on the state space $\calS^n$, 
+so the full likelihood function is given by \eqref{eqn:full_likelihood},
+but doing anything with the $|\calS|^n \times |\calS|^n$ matrix $G_n$ is clearly infeasible.
+We can, however, compute \eqref{eqn:full_likelihood} for smaller $n$,
+so the first thing that one might think to do is to break the sequence up into many blocks of length $m$,
+and treat these as independent, taking $m$ as large as is feasible.
+Then we would define $p_m(t;x,y)$ to be the probability that a string $x$ of length $m$ evolves to string $y$ over time $t$,
+denote by $x_i^{(m)}$ the block of length $m$ beginning at index $i$,
+and obtain the approximate likelihood function
+\[
+  \prod_{k=0}^{n/m} p_m(x_{km}^{(m)},y_{km}^{(m)}) .
+\]
+Variations are possible (what do we do about the edge effects for $p_m$?),
+but in any case there is no guarentee that this is close to correct,
+as it ignore dependencies between neighboring blocks.
+
+Intuitively, we want to increase the size of the context 
+-- but increasing $m$ does not do this, as there are always edge effects.
+However, we could look at the context in the initial sequence but not in the final sequence,
+including the states of an additional $\ell$ and $r$ sites on the left and right, respectively.
+Concretely, define the marginal probabilities
+\begin{align}
+  p_{\ell,m,r}(t;x,y) &= \sum_{a \in \calS^\ell} \sum_{b \in \calS^r} p_{\ell+m+r}(t;x,a \join y \join b) , \quad \text{for}\; x \in \calS^{\ell+m+r}, \; y \in \calS^m ,
+\end{align}
+where $a \join y \join b$ is the sequence composed of concatenating $a$, $y$, and $b$ together in that order.
+This is the probability that if we begin with a sequence $x$ of length $\ell+m+r$, 
+then after time $t$ we will see at the middle $m$ positions the pattern $y$,
+ignoring what is present in the leftmost $\ell$ and rightmost $r$ positions.
+In a long sequence, $p_m(x,z)$ might not be a very good approximation for $\P\{ X_i^{(m)}(t) = z | X_i(0)^{(m)} = x \}$,
+but for long enough $\ell$ and $m$ we do expect
+\begin{align}
+  \P\{ X_i^{(m)}(t) = y | X_i(0)^{(\ell+m+r)} = x \} \approx p_{\ell,m,r}(x,y) .
+\end{align}
+
+It is easy to convince oneself that the approximation gets better as $\ell$ and $r$ increase.
+Here is a simple argument.
+Suppose the maximum length of any transition pattern is $R = \max\{ |u| : (\mu,u,v) \in \calT \}$.
+The transition rate at site $i$ at time $s$ is determined if we know $X_{i-R}^{(2R+1)}(s)$;
+this can be determined from $X_{i-R}^{2R+1}(0)$ and the outcome of all changes happening before time $s$ at sites in $(i-R-1, \ldots, i+R)$.
+
 \end{document}

### 2013-05-14 context-dependence.tex
more writeup

@@
 \documentclass{article}
+\usepackage{fullpage}
 \usepackage{amsmath,amssymb}
 \usepackage{amsthm}
+
 \newcommand{\E}{\mathbb{E}}
@@
 
+\theoremstyle{plain}
+\newtheorem{theorem}{Theorem}
 \theoremstyle{definition}
@@
 
+\section{Simulation}
+
+First, it will be useful to have a brief description of how to simulate such a process.
+Briefly, we can put down a Poisson process of times and locations of possible changes,
+at an appropriate rate $\mu_*$,
+and then resolve each possible change in temporal order.
+We situate ``changes'' at the beginning of the matching pattern,
+so $\mu_*$ can be defined as the maximum sum of rates matching  the start of a sequence:
+\[
+\mu_* = \max \left\{ \sum_{y \in \calS^n} \sum_{j\in J(1,x,y)} \mu_j \st x \in \calS^n \right\} .
+\]
+Now sample possible changes: let $s_{ij}$ denote the $j^\text{th}$ possible change at site $i$,
+so that $0 \le s_{i1} < s_{i2} < \cdots < s_{iN_i} \le t$ is a rate $\mu_*$ Poisson process.
+Then, suppose we have determined the state at time $s$ to be $X(s) = x$,
+and the next event is $s_{ij} = \min \{ s_{k\ell} : s_{k\ell}>s \}$.
+The process then jumps to a new state at times $s_{ij}$ with probability proportional to the corresponding rate.
+In other words, let
+\[
+q_k = \begin{cases}
+  \mu_k/\mu_* \qquad & \text{if } x_i^{(|u_k|)} = u_k  \\
+  0 \qquad & \text{otherwise,}
+\end{cases}
+\]
+and let $q_0 = 1-\sum_{k=1}^{n_T} q_k$.
+By construction of $\mu_*$, $q_0\ge 0$.)
+Now choose transition $k$ with probability $q_k$, and if $k>0$, replace $x_i^{|u_k|}$ with $v_k$.
+
+For instance, to simulate TASEP, one has only to let $\mu_*=\lambda$, and at each $s_{ij}$
+check if at site $i$ there is a 1 followed by a 0,
+and if so, switch them.
+
 \section{Inference}
@@
 In a long sequence, $p_m(x,z)$ might not be a very good approximation for $\P\{ X_i^{(m)}(t) = z | X_i(0)^{(m)} = x \}$,
-but for long enough $\ell$ and $m$ we do expect
-\begin{align}
+we do expect $p_{\ell,m,r}(x,y)$ to be correct, asymptotically in $\ell$ and $r$:
+\begin{align} \label{eqn:window_approx}
   \P\{ X_i^{(m)}(t) = y | X_i(0)^{(\ell+m+r)} = x \} \approx p_{\ell,m,r}(x,y) .
@@
 
-It is easy to convince oneself that the approximation gets better as $\ell$ and $r$ increase.
-Here is a simple argument.
+In fact, the approximation improves exponentially as $\ell$ and $m$ increase --
+here is a simple argument.
 Suppose the maximum length of any transition pattern is $R = \max\{ |u| : (\mu,u,v) \in \calT \}$.
@@
 this can be determined from $X_{i-R}^{2R+1}(0)$ and the outcome of all changes happening before time $s$ at sites in $(i-R-1, \ldots, i+R)$.
+These will depend in turn on the outcome of previous possible changes.
+For each site $i$ and times $s_1 < s_2$, define the set of states $H(i,s_1,s_2)$ 
+so that if one knows the possible change points $s_{ij}$,
+then changes outside of $H(i,s_1,s_2)$ do not affect the distribution of $X_i(s_2)$.
+Concretely, define $H(i,s_1,s_2)$ by the property that $i \in H(i,s_1,s_2)$, and
+\begin{gather}
+  \text{if } s_1 < s_{jk} \le s_2 \text{ for some } i-R-1 \le j \le i+R, \text{ then } H(j,s_1,s_{jk}) \subseteq H(i,s_1,s_2) .
+\end{gather}
+In other words, $H(i,s_1,s_2)$ is the union of the $R$-neighborhoods of $i$ and all possible changes occurring in $H(i,s_1,s_2)$ between $s_1$ and $s_2$.
+For a set of states $I$, let $H(I,s_1,s_2) = \bigcup_{i\in I} H(i,s_1,s_2)$.
+Formally, the state $X_i(s_2)$ is conditionally independent of $\{X_j(s_1) \st j \notin H(i,s_1,s_2)\}$.
+Informally, the approximation \eqref{eqn:window_approx} will be good if $H$ does note expand beyond the flanking context.
+Now, MAKE THIS EASIER TO READ
+\begin{multline}
+  \P\{ X_i^{(m)}(t) = y \mid X(0) \} 
+  = \P\left\{ H(\{i,\ldots,i+m\},0,t)\subseteq\{i-\ell,\ldots,i+m+r\} \right\} \\
+   \qquad \qquad {} \times \P\left\{ X_i^{(m)}(t) = y \mid X_{i-\ell}^{(\ell+m+r)}(0), \; H(\{i,\ldots,i+m\},0,t)\subseteq\{i-\ell,\ldots,i+m+r\} \right\} \\
+     \qquad {} + \P\left\{ H(\{i,\ldots,i+m\},0,t)\nsubseteq\{i-\ell,\ldots,i+m+r\} \right\}  \\
+   \qquad\qquad {} \times \P\left\{ X_i^{(m)}(t) = y \mid X_{i-\ell}^{(\ell+m+r)}(0), \; H(\{i,\ldots,i+m\},0,t)\nsubseteq\{i-\ell,\ldots,i+m+r\} \right\} ,
+\end{multline}
+and so
+\begin{align} \label{eqn:prob_approx}
+  \left| \P\{ X_i^{(m)}(t) = y \mid X(0) \} - p_{\ell,m,r}(t;x,y) \right| \le 2 \P\left\{  H(\{I,\ldots,i+m\},0,t)\subseteq\{i-\ell,\ldots,i+m+r\} \right\}.
+\end{align}
+
+So, the error is bounded by the probability that $H(I,0,t)$ is larger than the context from $I-\ell$ to $I+r$.
+However, following $H$ backwards in time, the outer edge of $H$ expands as a continuous-time random walk:
+for instance, the left edge $L(s) = \min \{ j \in H(I,t-s,t) \}$ jumps leftwards at rate $R\mu_*$ with a random jump size
+that has mean $R/2$.
+General facts about random walks imply that \eqref{eqn:prob_approx} is bounded above by $C e^{-\alpha \ell/t}$, for some constants $C$ and $\alpha$.
 

### 2013-05-15 context-dependence.tex
.

@@
 
-\section*{Notation}
+\section{Notation and Examples}
 
@@
   imagined as magnetic dipoles,
-  and the energy associated with a given state $x$ is $H(x) = - \beta \sum_i x_i x_{i+1} + \gamma \sum_i x_i$.
+  and the energy associated with a given state $x$ is $H(x) = - \beta \sum_i x_i x_{i+1} - \gamma \sum_i x_i$.
   The associated stationary distribution on configurations is proportional to $\exp(-H(x))$.
@@
 namely
-\begin{align*}
+\begin{align} \label{eqn:G_defn}
   G(n)_{x,y} = \sum_{i=1}^n \sum_{j \in J(i,x,y)}  \mu_j ,
-\end{align*}
+\end{align}
 and if there are no triples $(\mu,u,v)$ with $x \xrightarrow{i,u,v} y$ for some $i$, then $G(n)_{x,y}=0$.
@@
 \end{align}
+We provide a concrete statement of this approximation later.
+
+%%%%
+\subsection*{Proof of the approximation}
+
+\begin{proof}
 
@@
 
+\end{proof}
+
+%%%%%%%%
+\subsection{Edge effects and cyclicization}
+
+The choice to treat long sequences as cyclical is generally of little importance.
+However, we have defined in \eqref{eqn:G_defn} the basic building block of our approximation, $G(m)$,
+to be the transition matrix \emph{for a circular sequence} of length $m$, where $m$ may be short.
+This seems strange -- but, what else would we do at the edges?
+We want to use $G(\ell+m+r)$ to compute the chance that a given $(\ell+m+r)$-sequence ends up with another given $m$-sequence in the middle;
+and if $\ell$ and $r$ are big enough relative to $t$, this will be a good approximation regardless of what we do at the edges.
+One answer would be to not allow patterns hanging off of either end to match,
+resulting in a slight underestimate of the chance of change.
+A better approximation would be to use the marginal frequencies in some way to average over surroudning sequence 
+-- but, this requires marginal frequencies to be known, and these are not stationary.
+However, the leftmost site in a randomly chosen window is a sample from the marginal distribution,
+and if sites separated by distance $\ell+m+r$ are sufficiently uncorrelated,
+wrapping the sequence around is a good approximation to averaging over marginal frequencies in the surrounding sequence.
+It is easy to imagine situations where this would not be the best solution --
+but again, if $\ell$ and $r$ are large enough, it won't matter much.
+
+%%%%%%%%%%%
+\section{Computation}
+
 \end{document}

### 2013-05-15 context-dependence.tex
.

@@
 Goal: parameter inference, given before-after observations, or else observations descended from a common ancestor.
+
+Review stats lit, e.g. reviews by Friel, original by Besag (autologistic model), genetics by Larribe and Fearnead.
 
@@
 %%%%
+\subsection{Poisson approximation?}
+
+Now we know how to compute the (marginal) probability that a particular subsequence changes into another subsequence.
+Therefore, if we know the pattern $x$ (say) appears $N(x)$ times in the original sequence $X(0)$,
+the expected number of times $y$ would appear juxtaposed with $x$ is $N(x)p_{\ell,m,r}(x,y)$.
+
+\begin{flushright}
+\begin{tabular}{c|ccccccc}
+ $X(0)$ &  C  & A & C & T & C & A & C \\
+ $X(t)$ &  $\centerdot$  & $\centerdot$  & C & T & C & $\centerdot$  & $\centerdot$  
+\end{tabular}
+\end{flushright}
+
+There is correlation, of course.
+
+%%%%
 \subsection*{Proof of the approximation}

### 2013-05-16 context-dependence.tex
.

--- 2013-05-15 context-dependence.tex
@@
 \newcommand{\st}{\colon}  % such that
+\newcommand{\var}{\mathop{\mbox{var}}}
+\newcommand{\cov}{\mathop{\mbox{cov}}}
 
@@
 %%%%
+\subsection{A likelihood function}
+
+Now we know how to compute the (marginal) probability that a particular subsequence changes into another subsequence.
+For patterns $x$ and $y$, with offset $\ell$ (so that $|x|=|y|+2\ell$),
+we can count how many times $x$ appears in $X(0)$,
+and how many times $x$ appears in $X(0)$ while $y$ appears at the same place in $X(t)$, offset by $\ell$.
+For instance, if $x$ is ACTCAC, $y$ is CGC, and $\ell=3$, this looks like:
+\begin{center}
+\begin{tabular}{c|ccccccc}
+ $x$ &  C  & A & C & T & C & A & C \\
+ $y$ &  $\centerdot$  & $\centerdot$  & C & G & C & $\centerdot$  & $\centerdot$  
+\end{tabular}
+\end{center}
+where $\cdot$ could be any base.
+Let $N(x)$ and $N_\ell(x,y)$ count these events, respectively.
+
+From the definitions above, 
+the expected number of times $y$ would appear juxtaposed with $x$ is approximately $N(x)p_{\ell,m,r}(x,y)$.
+Covariances are also easy to compute.
+If we define $M_i(x,y)$ to be 1 if $X_{i-\ell}^{|x|}(0)=x$ and $X_i^({|y|})(t)=y$, and 0 otherwise,
+so that $N(x,y) = \sum_{i=1}^n M_i(x,y)$, then
+\begin{align}
+  \var[N(x,y)] &= \sum_{i=1}^n \var[M_i(x,y)] + \sum_{i\neq j} \cov[M_i(x,y), M_j(x,y)] \\
+  &= n \var[M_1(x,y)] + 2n\sum_{k=1}^n \cov[M_1(x,y),M_{k+1}(x,y)] \\
+  &\approx n p_{\ell,m}(x,y)(1-p_{\ell,m}(x,y)) + 2n\sum_{k=1}^{m+2\ell} \left( \E[M_1(x,y) M_{k+1}(x,y)] - p_{\ell,m}(x,y)^2 \right) \\
+  &= n p_{\ell,m}(x,y)(1-(1+2m+4\ell)p_{\ell,m}(x,y)) + 2n\sum_{k=1}^{m+2\ell} \E[M_1(x,y) M_{k+1}(x,y)] .
+\end{align}
+The term $\E[M_1(x,y) M_{k+1}(x,y)]$ is the probability that $x$ matches at both site 1 and site $k+1$,
+and that $y$ matches both at site $\ell+1$ and at site $\ell+k+1$.
+For this to be nonzero, the joint pattern $(x,y)$ must allow self-overlaps.
+Similarly, if $(x_1,y_1) \neq (x_2,y_2)$ then
+\begin{align}
+  \begin{split}
+    \cov[N(x_1,y_1),N(x_2,y_2)] 
+    & \approx - n (2m+4\ell) p_{\ell,m}(x_1,y_1)p_{\ell,m}(x_2,y_2) \\
+    & \qquad {} + n\sum_{k=1}^{m+2\ell} \E[M_1(x_1,y_1) M_{k+1}(x_2,y_2) + M_1(x_2,y_2) M_{k+1}(x_1,y_1)] .
+  \end{split}
+\end{align}
+
+%%%%
 \subsection*{Proof of the approximation}

### 2013-05-16 context-dependence.tex
.

@@
 %%%%
-\subsection{A likelihood function}
+\subsection{Summary statistics}
 
@@
 \end{align}
+
+%%%%
+\subsection{A likelihood function}
+
+The covariance calculations show that
+if $p_{\ell,m}(x,y)$ is small enough (i.e.\ the pattern is not common),
+then $N(x,y)$ is close to Poisson.
+
+Using this observation, we can estimate parameters by composite likelihood,
+using all $N(x,y)$ as data.
+
+To get confidence intervals, we need to either simulate, or do something different.
+If we use only a subset of patterns $\{(x_k,y_k)\}$ chosen so that $\E[M_1(x,y) M_{k+1}(x,y)]$ is small (or zero),
+the entire collection $\{N(x_k,y_k)\}$ will be close to independent Poisson.
+(using Stein's method)
+Then we'll have the correct likelihood function for (this subset of) data, and can go Bayesian.
 

### 2013-05-17 context-dependence.tex
more writing

@@
 using all $N(x,y)$ as data.
+Maximum likelihood then gets an asymptotically unbiased estimate of the parameters.
 
 To get confidence intervals, we need to either simulate, or do something different.
+
 If we use only a subset of patterns $\{(x_k,y_k)\}$ chosen so that $\E[M_1(x,y) M_{k+1}(x,y)]$ is small (or zero),
@@
 (using Stein's method)
-Then we'll have the correct likelihood function for (this subset of) data, and can go Bayesian.
+Then we'll have the correct likelihood function for (this subset of) data.
+If changes are rare, then one way to get a set of patterns with this property is to only choose ones such that going from $x$ to $y$ entails a change;
+and any overlapping patterns must require additional changes, 
+i.e.\ if $(x_1,y_1)$ matches at $i_1$, 
+then for $(x_2,y_2)$ to match at $i_2$, 
+there must be additional changes outside $(i_1,\ldots,i_1+\ell-1)$.
 
@@
 
+%%%%%%%%
+\section{Phylogenetic inference}
+
+In phylogenetic applications, rather than ``before'' and ``after'' observations,
+we get two (or more) observations evolved from a common root.
+In the simplest case of two taxa we have two processes $X$ and $Y$,
+with identical starting states $X(0)=Y(0)$,
+observed only at times $t_X$ and $t_Y$ respectively.
+
+If the approximation described above holds,
+then we can write the probability that we see (long) pattern $x$ in $X(t_X)$ juxtaposed with (short) pattern $y$ in $Y(t_Y)$
+using the pattern frequencies at the root.
+Concretely: pick a random location $I$ in the sequence and let $X_{I-\ell}^{m+2\ell}(0) = \rho$ be the (long) pattern seen there at the root,
+and write $\pi(z)$ for the frequency of pattern $z$ in $X(0)$, i.e.\ $\P\{\rho=z\}=\pi(z)$.
+% Given the corresponding pattern in $X(t_X)$, the distribution of $\rho$ is proportional to
+% \[
+%     \P\{ \rho=z \mid X_{I-\ell}^{m+2\ell}(t_X)=x\} \propto \pi(z) p_{m+2\ell}(t,z,x) .
+% \]
+The probability that we see $x$ and $y$ at the random location $I$ is
+\begin{align} \label{eqn:phylo_likelihood}
+    \P\{X_{I-\ell}^{m+2\ell}(t_X)=x \text{ and } Y_I^\ell(t_Y)=y \} = \sum_{z \in \calS^{m+2\ell}} \pi(z) p_{m+2\ell}(t_X,z,x) p_{m,\ell}(t_Y,z,y) .
+\end{align}
+This can be computed without much more effort than the simpler case above,
+but requires the frequencies at the root.
+
+\subsection{The root distribution}
+
+We clearly don't want to add an extra parameter $\pi(z)$ for all $z \in \calS^{m+2\ell}$. 
+If nothing else, it would make the inference depend on $m$ and $\ell$, which we'd like to avoid.
+One approach would be to put a Dirichlet distribution on $\pi(z)$,
+but this would destroy many of the nice computational properties of the likelihoods,
+since making the frequencies of patterns at the root random introduces correlations between counts.
+An alternative is to specify $\pi$ through lower-dimensional statistics,
+for instance to model the root state as a sample from a Markov random field
+with parameters for the marginal frequencies of each state, and pairwise (or higher) correlations between them.
+
+A Markov random field with interactions between sites no more than distance two away 
+is determined by the energy function $\phi(a,b,c)$ which gives the ``energy'' of the configuration $abc$ for each $a,b,c \in \calS$.
+We use this to define the distribution on root patterns $\pi$ by
+\begin{align}  \label{eqn:root_mrf}
+    \pi(z) = \frac{1}{Z} \exp\left( - \sum_{i=1}^{|z|} h(z_{i-1},z_i,z_{i+1}) \right),
+\end{align}
+i.e.\ the exponential of the total energy of each three adjacent sites,
+where $Z$ is the normalizing constant.
+Note that it is often difficult to compute $Z$ in applications of Markov random fields,
+but here we are working with relatively short sequences, 
+so we just compute \eqref{eqn:root_mrf} for all possible $z$, and divide by the sum.
+(However, for this same reason this is not equivalent to saying that $X(0)$ is a sample from the Markov random field
+and finding the distribution of patterns within this.)
+
 %%%%%%%%%%%
@@
 
+For inference, we need to compute the $|\calS|^{m+2\ell} \times |\calS|^m$ matrix whose $(x,y)^\text{th}$ entry is $p_{m,\ell}(t,x,y)$,
+for various $t$.
+This matrix is a projection of the $|\calS|^{m+2\ell} \times |\calS|^{m+2\ell}$ matrix whose $(x,z)^\text{th}$ entry is
+\begin{align}
+    p_{m+2\ell}(x,z) = \left( e^{t G(m+2\ell)} \right)_{x,z} ,
+\end{align}
+where $G(m+2\ell)$ is the sparse $|\calS|^{m+2\ell} \times |\calS|^{m+2\ell}$ matrix defined in \eqref{eqn:G_defn}.
+The ``projection'' we need just marginalizes over long patterns $z$ that match the shorter pattern $y$:
+if we define the matrix $U$ in this way, so that $U_{zy}=1$ if $z_\ell^m=y$ and $U_{zy}=0$ otherwise,
+then
+\[
+    p_{m,\ell}(t,x,y) = \sum_{z \in \calS^{m+2\ell}} \left( e^{t G(m+2\ell)} \right)_{xz} U_{zy} ,
+\]
+so in fact we don't need the entire matrix $e^{t G(m+2\ell)}$,
+just the product of this matrix with each of the columns of $U$.
+
+Modern techniques in sparse matrix computation (e.g.\ Krylov methods) provide efficient ways to do this
+by only multiplying $G$ and $U$ some number of times, 
+which is much more efficient because $G$ is sparse.
+
+Also note that the further calculations in the phylogenetic case can be done using these methods.
+To compute
+\[
+    \sum_z \pi(z) p_{m+2\ell}(t_X,z,x) p_{m,\ell}(t_Y,z,y) 
+    = \sum_z \left( e^{t_X G(m+2\ell)} \right)_{zx} \pi(z) \sum_w \left( e^{t_X G(m+2\ell)} \right)_{zw} U_{wy} ,
+\]
+we can first compute $\sum_w \left( e^{t_X G(m+2\ell)} \right)_{zw} U_{wy}$ as above,
+multiply by $\pi(z)$, and then apply the algorithm again using the transpose of $G(m+2\ell)$.
+
+Using the R package expm, this makes computation quite feasible:
+with four possible states and $\ell=2$ and $m=1$, so that $G(m+2\ell)$ is a $1024 \times 1024$ matrix,
+computing $e^{t G(5) }$ takes 13 seconds, while computing $e^{t G(5)} U$ takes only 0.3 seconds.
+Increasing to $\ell=4$ and $m=1$ or $\ell=3$ and $m=2$ is still feasible, taking $e^{t G(8)} U$ in 42 and 47 seconds, respectively.
+
 \end{document}

### 2013-05-18 sim-context-fns.R
tasep

@@ -6,15 +6,16 @@ rinitseq <- function (seqlen,bases,basefreqs=rep(1/length(bases),length(bases)))
     # return a random sequence.  bases can in fact be longer patterns.
     # note that if bases are longer than 1 character, 
     #  DOES NOT actually produce a sting with frequencies basefreqs.
+    stringfun <- if ( all(bases %in% c("A","C","G","T","-") ) ) { DNAString } else { BString }
     patlen <- mean( nchar(bases) )
     rpats <- c()
     while( sum(nchar(rpats))<seqlen ) {
         rpats <- c( rpats, sample( bases, floor(1.05*seqlen/patlen), replace=TRUE, prob=basefreqs ) )
     }
-    return( DNAString(substr( paste( rpats, collapse="" ), 1, seqlen )) )
+    return( stringfun(substr( paste( rpats, collapse="" ), 1, seqlen )) )
 }
 
-simseq <- function (seqlen, tlen, patlen, mutpats, mutrates, selpats, selcoef, Ne, bases=c("A","C","G","T"), count.trans=TRUE, initseq, basefreqs=rep(1/length(bases),length(bases)), ... ) {
+simseq <- function (seqlen, tlen, patlen, mutpats, mutrates, selpats, selcoef, Ne=1, bases=c("A","C","G","T"), count.trans=TRUE, initseq, basefreqs=rep(1/length(bases),length(bases)), ... ) {
     # simulate a random sequence and evolve it with genmatrix.
     #  record transition counts if count.trans it TRUE (e.g. for debugging)
     # First make transition matrix
@@ -22,13 +23,14 @@ simseq <- function (seqlen, tlen, patlen, mutpats, mutrates, selpats, selcoef, N
     #    e.g. if the rate of CG -> TG is 1.5, and pattern length is 4,
     #    then we'll match on CG.., .CG., and ..CG, so the transition rates of e.g. CGTT -> TGTT should be 1.5/(4-2+1) = 0.5
     #   So, rescale mutrates:
+    stringsetfun <- if ( all(bases %in% c("A","C","G","T","-") ) ) { DNAStringSet } else { BStringSet }
     genmatrix <- makegenmatrix( mutpats, selpats, patlen=patlen )
     mutpatlens <- lapply( mutpats, function (x) unique(nchar(unlist(x))) )
     if (! all( sapply(mutpatlens,length)==1 ) ) { stop("need each list in mutpats to have patterns of the same length") }
[…]

### 2013-05-25 context-dependence.tex
tasep
more writeup.

@@
 then $N(x,y)$ is close to Poisson.
-
 Using this observation, we can estimate parameters by composite likelihood,
@@
 Maximum likelihood then gets an asymptotically unbiased estimate of the parameters.
-
-To get confidence intervals, we need to either simulate, or do something different.
+However, since they are not independent, to get confidence intervals, we need to either simulate, or do something different.
 
@@
 there must be additional changes outside $(i_1,\ldots,i_1+\ell-1)$.
+
+Here is one way to obtain such a set.
+Take a pair of patterns $(x,y)$, that overlap at $m$ positions,
+and suppose that they differ at $1 \le i_1, \ldots, i_d \le m$.
+For example, in the example above with $m=3$ and $\ell=2$,
+\begin{center}
+\begin{tabular}{c|ccccccc}
+ $x$ &  C  & A & C & T & C & A & C \\
+ $y$ &  $\centerdot$  & $\centerdot$  & C & G & C & $\centerdot$  & $\centerdot$  
+\end{tabular}
+\end{center}
+$x$ and $y$ differ only at (relative) position $i_1=2$.
+Let $\bar i = (1/d) \sum_{j=1}^d i_j$ be the mean position of the changes,
+and take only patterns with $(m-1)/2 < \bar i \le (m+1)/2$.
+As another example,
+\begin{center}
+\begin{tabular}{c|ccccccc}
+ $x$ &  A & C & T & C & A & C & G \\
+ $y$ &  $\centerdot$  & $\centerdot$  & G & C & A & $\centerdot$  & $\centerdot$  
+\end{tabular}
+\end{center}
+has $i_1 = \bar i = 1$.
+No two such patterns can be shifted versions of each other
+without adding more changed sites,
+since shifting by $k$ without adding changed sites just adds $k$ to $\bar i$.
 
@@
 
-Modern techniques in sparse matrix computation (e.g.\ Krylov methods) provide efficient ways to do this
+Modern techniques in sparse matrix computation (Krylov methods) provide efficient ways to do this
 by only multiplying $G$ and $U$ some number of times, 
@@
 
-Also note that the further calculations in the phylogenetic case can be done using these methods.
+\subsection{Updating $G$}
+
+We can also take advantage of the sparsity of $G$ to perform efficient computation of the likelihood
+under many sets of parameters.
+First note that $G(m+2\ell)$ has, assuming only single-position changes,
+$(1+m(|\calS|-1)) |\calS|^{m+2\ell}$ nonzero entries, since each of the $|\calS|^{m+2\ell}$ can change in $m$ places.
+This will determine how the computation scales with $m$, $\ell$, and $|\calS|$,
+once we precompute a number of things.
+Let $g = (g_1, \ldots, g_d)$ be the nonzero entries of $G$, in some fixed ordering;
+sparse matrix representations of $G$ store only $g$ along with information about the rows and columns these are found in.
+Each $g_i$ is a linear combination of mutation rates $\mu_j$, say $g_i = \sum_j a_{ij} \mu_j$;
+so by precomputing $a_{ij}$ we can efficiently update $G$.
+This is true if the boundary conditions are circular, mean-value, or neither.
+
+\subsection{Phylogenetic peeling}
+
+Furthermore, note that the further calculations in the phylogenetic case can be done using these methods.
 To compute
@@
 computing $e^{t G(5) }$ takes 13 seconds, while computing $e^{t G(5)} U$ takes only 0.3 seconds.
-Increasing to $\ell=4$ and $m=1$ or $\ell=3$ and $m=2$ is still feasible, taking $e^{t G(8)} U$ in 42 and 47 seconds, respectively.
+Increasing to $\ell=4$ and $m=1$ or $\ell=3$ and $m=2$ is still feasible, taking $e^{t G(8)} U$ in 42 and 47 seconds, respectively
+(and much longer for the whole matrix $e^{t G(8)}$).
+
+%%%%%%
+\section{Bayesian things}
+
+So, suppose we've got a list of patterns $(x_i,y_i)$
+and we're happy with the idea that they're independent.
+Recall that
+\[
+\E\left[ N(x,y) \mid N(x) \right] = N(x) p_{m,\ell}(x,y) ,
+\]
+Parameters $\theta$ enter implicitly through $p_{m,\ell}$.
+
+In the case that we observe the initial sequence, $N(x,y)$ are binomial with sample size $N(x)$.
+and so the likelihood is 
+\[
+\prod_x N(x)! \prod_y \frac{ p_{m,\ell}(x,y)^{N(x,y)} }{ N(x,y)! } ,
+\]
+and so the negative log likelihood is, up to a constant,
+\[
+\sum_{x,y} N(x,y) \log p_{m,\ell}(x,y) .
+\]
 

### 2013-05-25 ising/ising-inference.R

@@ -19,6 +19,8 @@ option_list <- list(
 opt <- parse_args(OptionParser(option_list=option_list,description=usage))
 attach(opt)
 
+if (interactive()) { win <- lwin <- rwin <- 2; nbatches <- 10; blen <- 10; mmean <- 1; svar <- 1 }
+
 if (is.null(infile) | is.null(nbatches)) { cat("Run\n  ising-inference.R -h\n for help.") }
 
 scriptdir <- "../"
@@ -27,25 +29,6 @@ source(paste(scriptdir,"sim-context-fns.R",sep=''))
 
 require(mcmc)
 
-if (interactive()) {
-
-    # available simulated sequences
-    simdir <- "ising-sims"
-    simfiles <- list.files(simdir,"*.RData",full.names=TRUE)
-    siminfo <- do.call(rbind, lapply(simfiles, function (x) {
-            load(x)
-            y <- do.call( data.frame, c(list(date=now,seqlen=seqlen, tlen=tlen, file=x), as.list(mutrates), as.list(selcoef), stringsAsFactors=FALSE) )
-            colnames(y) <- c( colnames(y)[1:4], paste("mutrate",seq_along(mutrates),sep=''), paste("selcoef",seq_along(selcoef),sep='') )
-            y
-        } ) )
-    siminfo$meandist <- siminfo$tlen * colSums(siminfo[,grep("mutrate",colnames(siminfo)),drop=FALSE])
-    siminfo <- siminfo[ order(siminfo$meandist), ]
-
-    win <- 2
-
-    # pick one
-    infile <- siminfo[1,"file"]
-}
 
 load(infile)
 basedir <- gsub(".RData","",infile,fixed=TRUE)
@@ -192,13 +175,3 @@ dev.off()
 
 
 print(format(Sys.time(),"%Y-%m-%d-%H-%M"))
-
-for (x in list.files("ising-sims","*.RData",full.names=TRUE)) {
-  tmp <- load(x)
-  nmuts <- 1; nsel <- 2
-  projmatrix <- collapsepatmatrix( ipatterns=rownames(genmatrix), lwin=lwin, rwin=rwin )
-  nonoverlapping <- leftchanged(rownames(counts[[1]]),colnames(counts[[1]]),lwin=lwin,win=win)
-  nov.counts <- counts[[1]][nonoverlapping]
-  mmean <- 1; svar <- 1
-  save( nmuts, nsel, projmatrix, nonoverlapping, nov.counts, mmean, svar, list=tmp, file=x)
-}

### 2013-05-31 context-dependence.tex
cpg?

@@
 \usepackage{amsthm}
+\usepackage{graphicx}
 
@@
 
+%%%%%%%
+\section{Simulation results}
+
+The method works well to find point estimates of the parameters.
+Obtaining posterior density estimates using the approximation likelihood function on a subset of patterns
+works in some cases;
+in other cases the likelihood surface is still too peaked, underestimating the uncertainty.
+
+\subsection{TASEP}
+
+Strings of length $10^5$ were evolved under TASEP for varying amounts of time,
+and pattern occurrences counted.
+Running 10,000 MCMC iterations with window sizes of $\ell=w=3$ (total length 9) on these takes around 40 minutes.
+Apparently, the patterns used were not sufficiently independent in this case,
+as the posterior is too sharp.
+
+\begin{figure}
+  \begin{center}
+    \includegraphics[width=\textwidth]{tasep/all-mcmc-runs}
+  \end{center}
+  \caption{ Results from MCMC estimates of the parameter in TASEP (mutation rate multiplied by time).  
+  Vertical line shows the true value.
+  }
+\end{figure}
+
+\subsection{Ising model}
+
+Strings of length $10^6$ were evolved under Glauber dynamics on the Ising model with varying parameters,
+and pattern occurrences counted.
+As for TASEP, unning 10,000 MCMC iterations with window sizes of $\ell=w=3$ (total length 9) on these takes around 40 minutes.
+In this case, the true parameter values have good distribution within the estimated posterior distributions.
+
+\begin{figure}
+  \begin{center}
+    \includegraphics[width=\textwidth,page=1]{ising/all-mcmc-runs}
+  \end{center}
+  \caption{ Results from MCMC estimates of the mutation rate parameter in the Ising model (scaled by amount of time). 
+  Vertical line shows the true value.
+  }
+\end{figure}
+
+\begin{figure}
+  \begin{center}
+    \includegraphics[width=\textwidth,page=2]{ising/all-mcmc-runs}
+  \end{center}
+  \caption{ Results from MCMC estimates of the inverse temperature $\beta$ in the Ising model. 
+  Vertical line shows the true value.
+  }
+\end{figure}
+
+\begin{figure}
+  \begin{center}
+    \includegraphics[width=\textwidth,page=3]{ising/all-mcmc-runs}
+  \end{center}
+  \caption{ Results from MCMC estimates of the magenization parameter in the Ising model. 
+  Vertical line shows the true value.
+  }
+\end{figure}
+
+%%%%%%%%
+\section{What next}
+
+\begin{enumerate}
+
+  \item Do inference along a tree.
+
+  \item See if counting in nonoverlapping windows fixes the peakiness problem with TASEP.
+
+  \item See what's feasible with a larger alphabet -- 4 bases, or 20 amino acids -- and more parameters.
+
+  \item Show that taking $\ell=0$ can be positively misleading (when?).
+
+\end{enumerate}
+
 \end{document}

### 2013-06-11 context-dependence.tex
.

@@
   \end{center}
-  \caption{ Results from MCMC estimates of the parameter in TASEP (mutation rate multiplied by time).  
+  \caption{ Results from MCMC estimates for six different data sets of the parameter in TASEP (mutation rate multiplied by time).  
   Vertical line shows the true value.
@@
   \end{center}
-  \caption{ Results from MCMC estimates of the mutation rate parameter in the Ising model (scaled by amount of time). 
+  \caption{ Results from MCMC estimates of the time-scaled mutation rate parameter $\lambda t$,
+  in ten different simulated datasets under the Ising model.
   Vertical line shows the true value.
@@
   \end{center}
-  \caption{ Results from MCMC estimates of the magenization parameter in the Ising model. 
+  \caption{ Results from MCMC estimates of the magnetization parameter $\beta$ in the Ising model. 
   Vertical line shows the true value.
@@
 \end{figure}
+
+%%%
+\subsection{CpG mutation}
+
+Nucleotide sequences of length $10^6$ were simulated with various
 

### 2013-06-11 context-dependence.tex
writeup

@@
 If nothing else, it would make the inference depend on $m$ and $\ell$, which we'd like to avoid.
-One approach would be to put a Dirichlet distribution on $\pi(z)$,
-but this would destroy many of the nice computational properties of the likelihoods,
-since making the frequencies of patterns at the root random introduces correlations between counts.
+One approach would be to put a Dirichlet distribution on $\pi(z)$.
+This has the nice property that you can choose priors that are consistent across different values of $m$ and $\ell$.
+We'd have to still MCMC around in the space of $\pi(z)$, thought, so they'd be nuisance parameters from that point of view.
+Also, making the frequencies at the root random introduces correlations between counts, which we want to avoid.
 An alternative is to specify $\pi$ through lower-dimensional statistics,
@@
   \begin{center}
+    \includegraphics{writeup-plots/selsims-2013-05-28-17-12-0275615-traces}
+  \end{center}
+  \caption{ 
+  MCMC traces of the parameters in the Ising model across 100,000 iterations.
+  The units of $\lambda$ are in mean number of changes per site.
+  Vertical line shows the true value.
+  }
+\end{figure}
+
+\begin{figure}
+  \begin{center}
     \includegraphics[width=\textwidth,page=1]{ising/all-mcmc-runs}
@@
 %%%
-\subsection{CpG mutation}
+\subsection{CpG mutation, on a tree}
 
-Nucleotide sequences of length $10^6$ were simulated with various
+Nucleotide sequences of length $10^6$ were simulated down two (unequal) branches of a tree,
+with all single-base substitution rates $m_{xy}$ set equal (varying from a total of .01 to .1 across the tree)
+and an additional CpG rate $\gamma$ set to either 0 or three times the single-base rate.
+Analysis was done with $\ell=2$ and $w=1$ (a five-to-one window).
+
+The state at the root of the tree was independently chosen bases with equal frequencies, 
+and these frequencies at the root were added as parameters to the model (with a Dirichlet prior).
+
+The point estimates are good, 
+as are the MCMC estimates,
+but there seems to be mixing trouble having to do with the root frequencies.
+
+\begin{figure}
+  \begin{center}
+    \includegraphics{writeup-plots/selsims-2013-06-03-13-17-0790276-initfreqs}
+  \end{center}
+  \caption{ 
+  MCMC traces across 660,000 iterations of the frequencies at the root of the (two-taxon) tree in the CpG model.
+  Vertical line shows the true value.
+  }
+\end{figure}
+
+\begin{figure}
+  \begin{center}
+    \includegraphics{writeup-plots/selsims-2013-06-03-13-17-0790276-mutrates}
+  \end{center}
+  \caption{ 
+  MCMC traces across 660,000 iterations of several mutation rate parameters in the CpG model.
+  Units are in mean number of substitutions per base across the entire tree.
+  Vertical line shows the true value.
+  }
+\end{figure}
 
@@
 
-  \item Do inference along a tree.
+  \item Allow more than single-base distribution at the root.
+
+  \item What's up with the mixing problem in root frequencies?
 
@@
 
-  \item See what's feasible with a larger alphabet -- 4 bases, or 20 amino acids -- and more parameters.
+  \item See what's feasible with a larger alphabet -- 60 codons -- and more parameters.
 
-  \item Show that taking $\ell=0$ can be positively misleading (when?).
+  \item Show that taking no context ($\ell=0$) can be positively misleading in some context.
+
+  \item Do a three-taxa tree.
+
+  \item How to identify motifs in the residuals?
 

### 2013-06-12 context-dependence.tex
.

@@
 
+Here's a short example sequence with $\lambda=1$:
+\begin{center}
+ \setlength{\tabcolsep}{0pt}
+\begin{tabular}{cccccccccccccccccccccccccccccccccccccccccccccccccccccccccccc}
+X&X&X&O&O&X&O&O&O&O&X&O&X&O&X&X&O&O&X&O&O&X&O&X&X&X&O&X&O&O&O&X&X&O&X&O&O&O&X&X&O&O&O&X&O&O&O&O&X&O&X&X&O&X&X&O&O&X&O&O \\
+$\centerdot$&O&X&O&X&O&O&O&X&$\centerdot$&O&X&O&O&X&X&X&o&x&$\centerdot$&o&O&X&x&x&O&X&O&X&o&o&O&X&X&$\centerdot$&$\centerdot$&o&o&O&X&X&o&$\centerdot$&O&X&$\centerdot$&o&o&$\centerdot$&$\centerdot$&O&X&X&x&O&X&o&O&X&$\centerdot$
+\end{tabular}
+\end{center}
+here particles are X's, moving to the right; and spaces are O's; 
+in the second line, 
+$\centerdot$ is a position that has not considered a change; lowercase letters are positions that considered a change but did not;
+and upper case are positions that did change.
+
 \begin{figure}
@@
 In this case, the true parameter values have good distribution within the estimated posterior distributions.
+
+Here is an example, with all parameters equal to 1, for $t=.5$:
+\begin{center}
+ \setlength{\tabcolsep}{0pt}
+\begin{tabular}{cccccccccccccccccccccccccccccccccccccccccccccccccccccccccccc}
+X&X&O&X&X&O&O&X&O&X&O&X&X&O&O&X&O&O&O&O&X&X&X&O&X&O&X&O&X&O&O&O&X&X&O&X&X&X&X&X&O&X&O&X&O&X&X&X&X&O&O&X&X&X&X&X&X&O&O&O \\
+x&x&o&x&X&X&O&x&o&x&$\centerdot$&x&x&O&X&X&$\centerdot$&O&X&O&X&x&x&o&x&o&x&o&x&O&X&O&$\centerdot$&$\centerdot$&O&O&X&X&O&X&X&X&o&x&o&$\centerdot$&X&O&X&o&o&x&X&X&X&O&X&o&$\centerdot$&$\centerdot$ 
+\end{tabular}
+\end{center}
+The notation is as above; except that the entire matched pattern involved in a change is uppercase:
+to simulate, we look in windows of size 3; so if XOO changes to XXO, the whole thing is uppercase, not just the middle X.
 
@@
 
+Here is a short example, with all parameters equal to 1, with each branch of the tree having length $0.1$:
+\begin{center}
+ \setlength{\tabcolsep}{0pt}
+\begin{tabular}{cccccccccccccccccccccccccccccccccccccccccccccccccccccccccccc}
+T&G&G&A&G&A&G&G&T&C&T&G&C&C&A&T&A&G&A&C&G&G&G&C&G&C&C&G&C&C&T&A&G&G&A&G&C&C&G&C&G&T&T&T&T&G&C&G&T&G&G&T&G&T&T&G&T&G&G&C \\
+$\centerdot$&$\centerdot$&g&C&G&a&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&T&A&C&A&$\centerdot$&$\centerdot$&$\centerdot$&T&A&T&$\centerdot$&$\centerdot$&$\centerdot$&T&G&c&C&C&A&C&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&G&G&G&t&A&T&$\centerdot$&$\centerdot$&G&A&g&C&G&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&t&T&G&A \\
+$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&A&T&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&T&T&G&T&$\centerdot$&$\centerdot$&G&A&T&$\centerdot$&$\centerdot$&$\centerdot$&G&G&c&t&A&A&G&C&$\centerdot$&$\centerdot$&$\centerdot$&G&T&G&A&t&t&t&g&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&G&A&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&A&C
+\end{tabular}
+\end{center}
+Top is ancestral sequence (``unobserved''); bottom two are at the tips.
+
 The point estimates are good, 

### 2013-06-12 writeup-plots.R

@@ -1,0 +1,117 @@
+scriptdir <- "./"
+source(paste(scriptdir,"codon-inference-fns.R",sep=''))
+source(paste(scriptdir,"sim-context-fns.R",sep=''))
+require(mcmc)
+
+
+##########
+## Tree CpG, with CpG = 0
+infile <- "ising/ising-sims/selsims-2013-05-28-17-12-0275615.RData"
+
+basedir <- gsub(".RData","",infile,fixed=TRUE)
+load(infile)
+
+lwin <- rwin <- win <- 3
+basename <- paste(basedir,"/win-",lwin,"-",win,"-",rwin,sep='')
+datafile <- paste( basename ,"-results.RData",sep='')
+plotfile <- paste( "writeup-plots/", basename(basedir), sep='')
+mcmcdatafiles <- list.files(path=basedir,pattern="-mcmc.*RData",full.names=TRUE)
+mcmcnum <- 1+max(c(0,as.numeric(gsub(".*-mcmc-","",gsub(".RData","",mcmcdatafiles)))),na.rm=TRUE)
+
+load(datafile)
+
+varnames <- c("lambda","beta","gamma")
+
+all.mrun <- do.call( rbind, lapply( seq_along(mcmcdatafiles), function (k) {
+            load(mcmcdatafiles[[k]])
+            tmp <- data.frame( mrun$batch )
+            names(tmp) <- varnames
+            tmp$mcmcnum <- k
+            tmp
+        } ) )
+
+subseq <- seq(1,nrow(all.mrun),by=max(1,floor(nrow(all.mrun)/1000)))
+
+x <- rbind( all.mrun[subseq,(-1)*match("mcmcnum",colnames(all.mrun))], truth ) 
+cols <- c( adjustcolor(rainbow(64),.3)[1+floor(65*(1:(nrow(x)-1))/nrow(x))], adjustcolor("black",.5) )
+
+pdf(file=paste(plotfile,"-traces.pdf",sep=''),width=7,height=5, pointsize=10)
+plotthese <- varnames
+layout(matrix(1:6,nrow=2))
+par(mar=c(4,4,1,1)+.1)
+for (k in 1:3) {
+    thisone <- match(plotthese[k],names(x))
+    thatone <- match(plotthese[1+k%%3],names(x))
+    plot( x[,thisone], x[,thatone], xlab=plotthese[k], ylab=plotthese[1+k%%3], col=cols, cex=c(rep(.5,nrow(x)-1),2), pch=20 )
[…]

### 2013-06-17 context-dependence.tex
.

@@
 regardless of the location $i$ along the sequence.
+If more than one pattern would match to cause the same change, their rates add.
 We call such a $(\mu,u,v)$ a ``transition triple''.
-The set of all transition triples $\calT = \{ (\mu_j,u_j,v_j) \}_{j=1}^{n_T}$ 
+The set of all transition triples $\calT = \{ (\mu^j,u^j,v^j) \}_{j=1}^{n_T}$ 
 determines the time evolution of the process.
@@
 with which the process in state $x$ jumps to state $y$.
-Let $x_i^{(\ell)} = (x_i, x_{i+1}, \ldots, x_{i+\ell-1})$ be the subsequence of length $\ell$ beginning at location $i$,
-and for each $1\le i \le n$, and patterns $u,v \in \calS^\ell$ define the relation
+% Let $x_i^{(\ell)} = (x_i, x_{i+1}, \ldots, x_{i+\ell-1})$ be the subsequence of length $\ell$ beginning at location $i$, and 
+For each $1\le i \le n$, pattern length $\ell \ge 0$, and patterns $u,v \in \calS^\ell$ define the relation
 \[
 x \xrightarrow{i,u,v} y \qquad \text{iff} \qquad \begin{cases}
-  x_j = y_j \quad &\text{for } j<i \\
-  x_{i+k} = u_k \quad &\text{for } k < \ell \\
-  y_{i+k} = v_k \quad &\text{for } k < \ell, and \\
-  x_j = y_j \quad &\text{for } j\ge i+\ell ,
+  x_k = y_k \quad &\text{for } k<i \\
+  x_{i+k} = u_k \quad &\text{for } 0 \le k < \ell \\
+  y_{i+k} = v_k \quad &\text{for } 0 \le k < \ell, and \\
+  x_k = y_k \quad &\text{for } k\ge i+\ell ,
 \end{cases}
 \]
-i.e.\ if $x$ and $y$ match except for in $i,i+1,\ldots,i+\ell$, where $x$ matches with $u$ and $y$ matches with $v$.
-Interpet these indices $\mod n$, even for short sequences, a point we will come back to.
-We'll also need the set of transitions that can move from $x$ to $y$ at a given location: let  $J(i,x,y) = \{ j \st x \xrightarrow{i,u_j,v_j} y \}$.
+i.e.\ if $x$ and $y$ match except for in $i,i+1,\ldots,i+\ell-1$, where $x$ matches with $u$ and $y$ matches with $v$.
+(Interpet these indices modulo $n$, even for short sequences, a point we will come back to.)
+We'll also need the set of defined transitions in $\calT$ 
+that can move from $x$ to $y$ at a given location: let  $J(i,x,y) = \{ (\mu^j,u^j,v^j) \in \calT \st x \xrightarrow{i,u^j,v^j} y \}$.
 Then, the rate $G(n)_{x,y}$ is the sum of all matching transition rates,
@@
 \begin{align} \label{eqn:G_defn}
-  G(n)_{x,y} = \sum_{i=1}^n \sum_{j \in J(i,x,y)}  \mu_j ,
+  G(n)_{x,y} = \sum_{i=1}^n \sum_{J(i,x,y)}  \mu^j ,
 \end{align}
 and if there are no triples $(\mu,u,v)$ with $x \xrightarrow{i,u,v} y$ for some $i$, then $G(n)_{x,y}=0$.
+This is just to say that the transition rate from $x$ to $y$ is the sum of the mutation rates of all transition triples
+that could change $x$ into $y$.
 In principle, this gives us the transition probabilities for the process:
 \begin{align} \label{eqn:full_likelihood}
-  p_n(t;x,y) := \P\{ X(t) = y \mid X(0) = x \} = (e^{tG_n})_{xy} .
+    p_n(t;x,y) := \P\{ X(t) = y \mid X(0) = x \} = (e^{tG(n)})_{xy} .
 \end{align}

### 2013-07-06 sim-context-fns.R
power sims set up?

@@ -102,24 +102,31 @@ countpats <- function (patterns, simseq ) {
     return( sapply( patterns, countPattern, simseq ) )
 }
 
-counttrans <- function (ipatterns, fpatterns, initseq=simseqs[["initseq"]], finalseq=simseqs[["finalseq"]], simseqs, lwin=0) {
+counttrans <- function (ipatterns, fpatterns, initseq=simseqs[["initseq"]], finalseq=simseqs[["finalseq"]], simseqs, lwin=0, shift=0) {
     # count number of times initseq matches ipatterns while finalseq matches fpatterns
-    # again, cyclical
+    #   again, cyclical
+    # if shift is nonzero, return a list  with the counts in each of (shift) frames
     patlen <- nchar(ipatterns[1])
     seqlen <- nchar(initseq)
     stopifnot( seqlen == nchar(finalseq) )
     # cyclic-ize
     initseq <- xscat( initseq, subseq(initseq,1,patlen-1) )
     finalseq <- xscat( finalseq, subseq(finalseq,1,patlen-1) )
-    # Ok, count occurrences.  
+    # Ok, count occurrences.  uses bioconductor stuff.
     initmatches <- lapply( ipatterns, matchPattern, initseq )
     finalmatches <- lapply( fpatterns, matchPattern, finalseq )
-    counts <- Matrix( 0, nrow=length(initmatches), ncol=length(finalmatches), dimnames=list(ipatterns,fpatterns) )
+    counts <- lapply( 1:max(1,shift), function (k) Matrix( 0, nrow=length(initmatches), ncol=length(finalmatches), dimnames=list(ipatterns,fpatterns) ) )
     for (x in seq_along(initmatches))  {
         for (y in seq_along(finalmatches)) {
-            counts[x,y] <- length(intersect(start(initmatches[[x]]),(-lwin)+start(finalmatches[[y]])))
+            xycounts <- intersect(start(initmatches[[x]]),(-lwin)+start(finalmatches[[y]]))
+            if (length(xycounts)>0) {
+                for (k in 0:max(0,shift-1)) {
+                    counts[[k+1]][x,y] <- sum( ( ( xycounts+k ) %% max(shift,1) ) == 0 )
+                }
+            }
         } 
     }
+    if (shift==0) { counts <- counts[[1]] }
     return(counts)
 }
 

### 2013-07-06 tasep/tasep-mcmc.R

@@ -46,7 +46,7 @@ mcmcdatafiles <- list.files(path=basedir,pattern="-mcmc.*RData",full.names=TRUE)
 mcmcnum <- 1+max(c(0,as.numeric(gsub(".*-mcmc-","",gsub(".RData","",mcmcdatafiles)))),na.rm=TRUE)
 
 if (logfile=="" & !interactive()) { logfile <- paste(basename,"-mcmc-run-",mcmcnum,".Rout",sep='') }
-if (!is.null(logfile)) { 
+if (logfile != '' & !is.null(logfile)) { 
     logcon <- if (logfile=="-") { stdout() } else { file(logfile,open="wt") }
     sink(file=logcon, type="message") 
     sink(file=logcon, type="output") 
@@ -61,31 +61,32 @@ if (length(mcmcdatafiles)) { load(grep(paste("-mcmc-",mcmcnum-1,".RData",sep='')
 winlen <- lwin+win+rwin
 genmatrix <- meangenmatrix( lwin=1, rwin=1, patlen=winlen, mutpats=mutpats, selpats=selpats, mutrates=mutrates*tlen, selcoef=selcoef, boundary="none" )
 projmatrix <- collapsepatmatrix( ipatterns=rownames(genmatrix), lwin=lwin, rwin=rwin )
-counts <- list( counttrans( rownames(projmatrix), colnames(projmatrix), simseqs[[1]]$initseq, simseqs[[1]]$finalseq, lwin=lwin ) )
-# want only patterns with leftmost possible position changed
-nonoverlapping <- leftchanged(rownames(counts[[1]]),colnames(counts[[1]]),lwin=lwin,win=win)
-nov.counts <- counts[[1]][nonoverlapping]
-initcounts <- rowSums(counts[[1]])
+counts <- list( counttrans( rownames(projmatrix), colnames(projmatrix), simseqs[[1]]$initseq, simseqs[[1]]$finalseq, lwin=lwin, shift=win ) )
+# look only every (win)-th window
+initcounts <- lapply( counts[[1]], rowSums )
 nmuts <- length(mutpats)
 nsel <- length(selpats)
 # using only nonoverlapping counts, plus priors -- indep't poisson.
-lud <- function (mutrates) {
+lud <- function (mutrates,k=1,...) {
     # params are: mutrates*tlen, selcoef
     if (any(mutrates<0)) {
         return( -Inf )
     } else {
         genmatrix@x <- update(genmatrix,mutrates=mutrates,selcoef=numeric(0))
[…]

### 2013-07-14 context-dependence.tex
power sims set up?
began human-chimp-gorilla with cpg and gc-bias

--- 2013-06-12 context-dependence.tex
@@
 This can be computed without much more effort than the simpler case above,
-but requires the frequencies at the root.
+given the frequencies at the root, as described in greater generality below.
 
+%%%%
+\subsection{Peeling}
+
+\begin{center}
+\includegraphics{peeling-schematic}
+\end{center}
+
+To compute the likelihood on a more complicated tree, consider the following,
+with the three-taxon tree above as an example,
+where we are counting occurrences of (long) $m+2\ell$-tuples in taxon $X$,
+and (shorter) $m$-tuples in taxa $Y$ and $Z$.
+
+We can compute the likelihood in a similar way, summing over the root state (the summation over $u$)
+and the interior node ($v$):
+\begin{align} \label{eqn:three_taxa_likelihood}
+  \begin{split}
+    & \P\{X_{I-\ell}^{m+2\ell}=x \text{ and } Y_I^\ell=y \text{ and } Z_I^\ell=z \} \\
+    &\qquad = \sum_{u \in \calS^{m+2\ell}} \pi(u) p_{m,\ell}(t_C,u,z) \sum_{v \in \calS^{m+2\ell}} p_{m+2\ell}(t_{AB},u,v) p_{m+2\ell}(t_{A},v,x) p_{m,\ell}(t_B,v,y) 
+  \end{split}
+\end{align}
+The steps for computing this are then:
+\begin{enumerate}
+
+  \item Compute $M_1(u,z) = p_{m,\ell}(t_C,u,z)$ (a $|\calS|^{m+2\ell} \times |\calS|^m$ matrix)
+
+  \item Compute $M_2(v,z) = \sum_{v \in \calS^{m+2\ell}}  p_{m+2\ell}(t_{AB},u,v) \pi(u) M_1(u,z)$ (still a $|\calS|^{m+2\ell} \times |\calS|^m$ matrix)
+
+  \item Compute $M_3(v,y) = p_{m,\ell}(t_B,v,y)$ (a $|\calS|^{m+2\ell} \times |\calS|^m$ matrix)
+
+  \item Compute $M_4(x,(y,z)) = \sum_{v \in \calS^{m+2\ell}} p_{m+2\ell}(t_{A},v,x) M_3(v,z) M_4(v,y)$ (stored as a $|\calS|^{m+2\ell} \times |\calS|^{2m}$ matrix).
+
+\end{enumerate}
+Then $M_4$ gives all the likelihoods.
+This is depicted in the figure, writing $\left(e^{tG}\right)_{ab}$ for the matrix $p_{m+2\ell}(t,a,b)$,
+and $P$ for the projection matrix that gives $p_{m,\ell}(t,a,b) = \left( e^{tG} P\right)_{ab}$.
+
+In total there are $|\calS|^{3m+2\ell}$ possible data combinations.  
+However, we can reduce dimensionality by replacing the last step with
+\begin{enumerate}
+
+  \item[4'.] Compute $M_4(x,w) = \sum_{v \in \calS^{m+2\ell}} p_{m+2\ell}(t_{A},v,x) M_3(v,w) M_4(v,w)$ (stored as a $|\calS|^{m+2\ell} \times |\calS|^{m}$ matrix).
+
+\end{enumerate}
+and only computing the likelihood for cases with $y=z=w$.
+On a three-taxon tree it is not clear that this is desireable, since missing out on double transitions 
+might be a significant loss of information,
+but on larger trees it seems likely that we'd want to restrict to combinations with most sites agreeing across the tree.
+(Note, however, that one should not first look at the data, observe which combinations $(y,z)$ were most common, and then restrict to only those!)
+
+%%%%
 \subsection{The root distribution}
@@
 and finding the distribution of patterns within this.)
+
+%%%%%
+\subsubsection{Mixing}
+
+There can be significant confounding of the frequencies at the root with the other parameters,
+which results in poor mixing.
+(See CpG results, below.)
+To see why this is, consider the following back-of-the-envelope calculation.
+Take $x \neq y$ to be single bases (so, set $\ell=0$ and $w=1$),
+so that $N(x,y)$ gives the number of times an $x$ and a $y$ are homologous.
+Let $q(x,y)$ be the instantaneous rate that $x \to y$ (averaged somehow over context?),
+so that $N(x,y)$ should be, roughly,
+\begin{align*}
+  \sum_z \pi(z) q(z,x) q(z,y) &\approx t \pi(x) (1+q(x,x)) q(x,y) + t \pi(y) q(y,x) (1+q(y,y)) ,  
+\end{align*}
+where as usual $q(x,x) = (-1) \sum_{z \neq x} q(x,z)$ is (minus) the total jump rate out of $x$.
+
+So, when we propose a change to $\pi$ we might want to also adjust the rates accordingly to keep this more-or-less constant:
+if we chnage $\pi(x) \mapsto \pi(x) + \epsilon_x$, we want $\delta_{xy}$ to satisfy
+\begin{align*}
+  \sum_z \pi(z) q(z,x) q(z,y) = \sum_z (\pi(z) + \epsilon_z) ( q(z,x) + \delta_{zx} ) ( q(z,y) + \delta_{zy} ) ,
+\end{align*}
+or, to first order in $\delta$,
+\begin{align*}
+  0 &\approx \sum_z \epsilon_z q(z,x) q(z,y) + \sum_z \pi(z) \delta_{zx} q(z,y) + \sum_z \pi(z) q(z,x) \delta_{zy} \\
+   &\approx \sum_z \epsilon_z q(z,x) q(z,y) + \sum_z (\pi(z)+\epsilon_z) \delta_{zx} q(z,y) + \sum_z (\pi(z)+\epsilon_z) q(z,x) \delta_{zy} \\
+   &\approx \epsilon_x (1+q(x,x)) q(x,y) + \epsilon_y q(y,x) (1+q(y,y)) + (\pi(y)+\epsilon_y) \delta_{yx} (1+q(y,y)) + (\pi(x)+\epsilon_x) (1+q(x,x)) \delta_{xy} \\
+   &= \left( \epsilon_x q(x,y) + (\pi(x)+\epsilon_x) \delta_{xy} \right)(1+q(x,x)) + \left( \epsilon_y q(y,x) + (\pi(y)+\epsilon_y) \delta_{yx} \right)(1+q(y,y)) 
+\end{align*}
+One solution to this is to let $\delta_{xy} = - \epsilon_x q(x,y) / \pi(x)$.
+
+Suppose each change is a single-base change, so that changes $\delta \mu_i$ in $\mu$ contribute to $q$ by
+$\delta_{xy} = \sum_{i: x \to y} \delta \mu_i$.
+Given $\delta_{xy}$, this gives a linear equation for $\delta \mu_i$ that may or may not be solvable,
+but certainly is if we have all the single-base transitions as paramters.
 
@@
 This is true if the boundary conditions are circular, mean-value, or neither.
+
+We can do similar precomputations to help with the case of selective differences:
+XXX describe
 
@@
 %%%%%%%%
+\section{Power sims}
+
+Assess power and accuracy in the Ising model, by running replicates at these parameters:
+
+\begin{tabular}{|rlllrrr|}
+  \hline
+  seqlen   &  $\lambda t$   &  $\beta$  &  $\gamma$  &  $w$ &  $\ell$   & \# reps \\
+  \hline
+  $10^4$    &   .2  &   1   &   .5  &   3   &   3   &   10  \\
+  $10^3$    &   .2  &   1   &   .5  &   3   &   3   &   10  \\
+  $10^2$    &   .2  &   1   &   .5  &   3   &   3   &   10  \\
+  \hline
+  $10^4$    &   .4  &   1   &   .5  &   3   &   3   &   10  \\
+  $10^4$    &   .1  &   1   &   .5  &   3   &   3   &   10  \\
+  $10^4$    &   .05  &   1   &   .5  &   3   &   3   &   10  \\
+  \hline
+  $10^4$    &   .2  &   1   &   0  &   3   &   3   &   10  \\
+  \hline
+  $10^4$    &   .2  &   1   &   .5  &   3   &   2   &   10  \\
+  $10^4$    &   .2  &   1   &   .5  &   3   &   1   &   10  \\
+  $10^4$    &   .2  &   1   &   .5  &   3   &   0   &   10  \\
+  \hline
+\end{tabular}
+
+%%%%%%%%
 \section{What next}
@@
 
-  \item See if counting in nonoverlapping windows fixes the peakiness problem with TASEP.
+  \item See if counting in nonoverlapping windows fixes the peakiness problem with TASEP. \emph{Yes, it does.}
 

### 2013-07-14 cpg/cpg-mcmc.R

@@ -79,7 +79,7 @@ initcounts <- rowSums(counts[[1]])
 nmuts <- length(mutpats)
 # using only nonoverlapping counts, plus priors -- indep't poisson.
 lud <- function (mutrates) {
-    # params are: mutrates*tlen, selcoef
+    # params are: mutrates*tlen
     if (any(mutrates<0)) {
         return( -Inf )
     } else {

### 2013-07-14 tasep/tasep-mcmc.R

@@ -13,10 +13,12 @@ option_list <- list(
         make_option( c("-n","--nbatches"), type="integer", default=1000, help="Number of MCMC batches. [default \"%default\"]" ),
         make_option( c("-b","--blen"), type="integer", default=10, help="Length of each MCMC batch. [default \"%default\"]" ),
         make_option( c("-c","--stepscale"), type="numeric", default=3e-3, help="Scale of proposal steps for Metropolis algorithm. [default \"%default\"]" ),
+        make_option( c("-f","--shift"), type="integer", default=NULL, help="Number of bases to skip between window starting positions. [default equal to win]" ),
         make_option( c("-s","--restart"), action="store_true", default=FALSE, help="Start a whole new MCMC run?" ),
         make_option( c("-o","--logfile"), type="character", default="", help="Direct output to this file. [default appends .Rout]" )
     )
 opt <- parse_args(OptionParser(option_list=option_list,description=usage))
+if (is.null(opt$shift)) { opt$shift <- opt$win }
 attach(opt)
 
 if (interactive()) { win <- lwin <- rwin <- 3; nbatches <- 100; blen <- 10; restart <- FALSE }

### 2013-07-24 writeups/context-dependence.tex
.

--- 2013-06-17 context-dependence.tex
@@
   \begin{center}
-    \includegraphics[width=\textwidth]{tasep/all-mcmc-runs}
+    \includegraphics[width=\textwidth]{../tasep/all-mcmc-runs}
   \end{center}
@@
   \begin{center}
-    \includegraphics{writeup-plots/selsims-2013-05-28-17-12-0275615-traces}
+    \includegraphics{../writeup-plots/selsims-2013-05-28-17-12-0275615-traces}
   \end{center}
@@
   \begin{center}
-    \includegraphics[width=\textwidth,page=1]{ising/all-mcmc-runs}
+    \includegraphics[width=\textwidth,page=1]{../ising/all-mcmc-runs}
   \end{center}
@@
   \begin{center}
-    \includegraphics[width=\textwidth,page=2]{ising/all-mcmc-runs}
+    \includegraphics[width=\textwidth,page=2]{../ising/all-mcmc-runs}
   \end{center}
@@
   \begin{center}
-    \includegraphics[width=\textwidth,page=3]{ising/all-mcmc-runs}
+    \includegraphics[width=\textwidth,page=3]{../ising/all-mcmc-runs}
   \end{center}

### 2013-09-10 writeups/context-dependence.tex
.

--- 2013-07-14 context-dependence.tex
@@
 %%%%
-\subsection{Peeling}
-
-\begin{center}
-\includegraphics{peeling-schematic}
-\end{center}
-
-To compute the likelihood on a more complicated tree, consider the following,
-with the three-taxon tree above as an example,
-where we are counting occurrences of (long) $m+2\ell$-tuples in taxon $X$,
-and (shorter) $m$-tuples in taxa $Y$ and $Z$.
-
-We can compute the likelihood in a similar way, summing over the root state (the summation over $u$)
-and the interior node ($v$):
-\begin{align} \label{eqn:three_taxa_likelihood}
-  \begin{split}
-    & \P\{X_{I-\ell}^{m+2\ell}=x \text{ and } Y_I^\ell=y \text{ and } Z_I^\ell=z \} \\
-    &\qquad = \sum_{u \in \calS^{m+2\ell}} \pi(u) p_{m,\ell}(t_C,u,z) \sum_{v \in \calS^{m+2\ell}} p_{m+2\ell}(t_{AB},u,v) p_{m+2\ell}(t_{A},v,x) p_{m,\ell}(t_B,v,y) 
-  \end{split}
-\end{align}
-The steps for computing this are then:
-\begin{enumerate}
-
-  \item Compute $M_1(u,z) = p_{m,\ell}(t_C,u,z)$ (a $|\calS|^{m+2\ell} \times |\calS|^m$ matrix)
-
-  \item Compute $M_2(v,z) = \sum_{v \in \calS^{m+2\ell}}  p_{m+2\ell}(t_{AB},u,v) \pi(u) M_1(u,z)$ (still a $|\calS|^{m+2\ell} \times |\calS|^m$ matrix)
-
-  \item Compute $M_3(v,y) = p_{m,\ell}(t_B,v,y)$ (a $|\calS|^{m+2\ell} \times |\calS|^m$ matrix)
-
-  \item Compute $M_4(x,(y,z)) = \sum_{v \in \calS^{m+2\ell}} p_{m+2\ell}(t_{A},v,x) M_3(v,z) M_4(v,y)$ (stored as a $|\calS|^{m+2\ell} \times |\calS|^{2m}$ matrix).
-
-\end{enumerate}
-Then $M_4$ gives all the likelihoods.
-This is depicted in the figure, writing $\left(e^{tG}\right)_{ab}$ for the matrix $p_{m+2\ell}(t,a,b)$,
-and $P$ for the projection matrix that gives $p_{m,\ell}(t,a,b) = \left( e^{tG} P\right)_{ab}$.
-
-In total there are $|\calS|^{3m+2\ell}$ possible data combinations.  
-However, we can reduce dimensionality by replacing the last step with
-\begin{enumerate}
-
-  \item[4'.] Compute $M_4(x,w) = \sum_{v \in \calS^{m+2\ell}} p_{m+2\ell}(t_{A},v,x) M_3(v,w) M_4(v,w)$ (stored as a $|\calS|^{m+2\ell} \times |\calS|^{m}$ matrix).
-
-\end{enumerate}
-and only computing the likelihood for cases with $y=z=w$.
-On a three-taxon tree it is not clear that this is desireable, since missing out on double transitions 
-might be a significant loss of information,
-but on larger trees it seems likely that we'd want to restrict to combinations with most sites agreeing across the tree.
-(Note, however, that one should not first look at the data, observe which combinations $(y,z)$ were most common, and then restrict to only those!)
-
-%%%%
 \subsection{The root distribution}
@@
 
+More generally\dots
+
+\begin{center}
+\includegraphics{peeling-schematic}
+\end{center}
+
+To compute the likelihood on a more complicated tree, consider the following,
+with the three-taxon tree above as an example,
+where we are counting occurrences of (long) $m+2\ell$-tuples in taxon $X$,
+and (shorter) $m$-tuples in taxa $Y$ and $Z$.
+
+We can compute the likelihood in a similar way, summing over the root state (the summation over $u$)
+and the interior node ($v$):
+\begin{align} \label{eqn:three_taxa_likelihood}
+  \begin{split}
+    & \P\{X_{I-\ell}^{m+2\ell}=x \text{ and } Y_I^\ell=y \text{ and } Z_I^\ell=z \} \\
+    &\qquad = \sum_{u \in \calS^{m+2\ell}} \pi(u) p_{m,\ell}(t_C,u,z) \sum_{v \in \calS^{m+2\ell}} p_{m+2\ell}(t_{AB},u,v) p_{m+2\ell}(t_{A},v,x) p_{m,\ell}(t_B,v,y) 
+  \end{split}
+\end{align}
+The steps for computing this are then:
+\begin{enumerate}
+
+  \item Compute $M_1(u,z) = p_{m,\ell}(t_C,u,z)$ (a $|\calS|^{m+2\ell} \times |\calS|^m$ matrix)
+
+  \item Compute $M_2(v,z) = \sum_{v \in \calS^{m+2\ell}}  p_{m+2\ell}(t_{AB},u,v) \pi(u) M_1(u,z)$ (still a $|\calS|^{m+2\ell} \times |\calS|^m$ matrix)
+
+  \item Compute $M_3(v,y) = p_{m,\ell}(t_B,v,y)$ (a $|\calS|^{m+2\ell} \times |\calS|^m$ matrix)
+
+  \item Compute $M_4(x,(y,z)) = \sum_{v \in \calS^{m+2\ell}} p_{m+2\ell}(t_{A},v,x) M_3(v,z) M_4(v,y)$ (stored as a $|\calS|^{m+2\ell} \times |\calS|^{2m}$ matrix).
+
+\end{enumerate}
+Then $M_4$ gives all the likelihoods.
+This is depicted in the figure, writing $\left(e^{tG}\right)_{ab}$ for the matrix $p_{m+2\ell}(t,a,b)$,
+and $P$ for the projection matrix that gives $p_{m,\ell}(t,a,b) = \left( e^{tG} P\right)_{ab}$.
+
+In total there are $|\calS|^{3m+2\ell}$ possible data combinations.  
+However, we can reduce dimensionality by replacing the last step with
+\begin{enumerate}
+
+  \item[4'.] Compute $M_4(x,w) = \sum_{v \in \calS^{m+2\ell}} p_{m+2\ell}(t_{A},v,x) M_3(v,w) M_4(v,w)$ (stored as a $|\calS|^{m+2\ell} \times |\calS|^{m}$ matrix).
+
+\end{enumerate}
+and only computing the likelihood for cases with $y=z=w$.
+On a three-taxon tree it is not clear that this is desireable, since missing out on double transitions 
+might be a significant loss of information,
+but on larger trees it seems likely that we'd want to restrict to combinations with most sites agreeing across the tree.
+(Note, however, that one should not first look at the data, observe which combinations $(y,z)$ were most common, and then restrict to only those!)
+
 %%%%%%
@@
   \begin{center}
-    \includegraphics{writeup-plots/selsims-2013-06-03-13-17-0790276-initfreqs}
+    \includegraphics{../writeup-plots/selsims-2013-06-03-13-17-0790276-initfreqs}
   \end{center}
@@
   \begin{center}
-    \includegraphics{writeup-plots/selsims-2013-06-03-13-17-0790276-mutrates}
+    \includegraphics{../writeup-plots/selsims-2013-06-03-13-17-0790276-mutrates}
   \end{center}
@@
 
+\section{Applications}
+
+graham:
+\begin{itemize}
+
+\item estimating freq spectrum [Hernandez]
+
+\item probabilistic ancestral states for common ancestor of e.g. humans 4N generations ago [Siepel] -- would widely used for primates
+
+\end{itemize}
+
 %%%%%%%%

### 2013-09-10 mtCDNA/mtCDNA-inference.R
.

@@ -1,0 +1,85 @@
+#!/usr/bin/Rscript --vanilla
+require(optparse)
+
+usage <- "\
+Infer parameters from output of sim-tree-cpg.R .\
+"
+
+option_list <- list(
+        make_option( c("-c","--infile"), type="character", default="mtCDNApri-sub.nuc", help=".RData file containing simulation." ),
+        make_option( c("-w","--win"), type="integer", default=1, help="Size of matching window. [default \"%default\"]" ),
+        make_option( c("-l","--lwin"), type="integer", default=2, help="Size of left-hand context. [default \"%default\"]" ),
+        make_option( c("-r","--rwin"), type="integer", default=2, help="Size of left-hand context. [default \"%default\"]" ),
+        make_option( c("-n","--nbatches"), type="integer", default=20, help="Number of MCMC batches. [default \"%default\"]" ),
+        make_option( c("-b","--blen"), type="integer", default=10, help="Length of each MCMC batch. [default \"%default\"]" ),
+        make_option( c("-s","--stepscale"), type="numeric", default=1e-4, help="Scale of proposal steps for Metropolis algorithm. [default \"%default\"]" ),
+        make_option( c("-m","--mmean"), type="double", default=1, help="Prior mean on single base mutation rates. [default \"%default\"]" ),
+        make_option( c("-c","--cpgmean"), type="double", default=1, help="Prior variance on CpG rate. [default \"%default\"]" ),
+        make_option( c("-p","--pprior"), type="double", default=1, help="Parameter for Dirichlet prior on base frequencies. [default \"%default\"]" ),
+        make_option( c("-v","--tprior"), type="double", default=.5, help="Parameter for Beta prior on branch length. [default \"%default\"]" ),
+        make_option( c("-d","--boundary"), type="character", default="none", help="Boundary conditions for generator matrix. [default \"%default\"]"),
+        make_option( c("-y","--meanboundary"), type="integer", default=0, help="Average over this many neighboring bases in computing generator matrix. [default \"%default\"]" ),
[…]

### 2013-10-12 cpg/cpg-nocontext-inference.R
no context inference for CpG

@@ -1,0 +1,180 @@
+#!/usr/bin/Rscript --vanilla
+require(optparse)
+
+usage <- "\
+Infer parameters from output of sim-cpg.R ,\
+in the case where there is NO CONTEXT-DEPENDENCE.
+"
+
+option_list <- list(
+        make_option( c("-i","--infile"), type="character", default=NULL, help=".RData file containing simulation." ),
+        make_option( c("-n","--nbatches"), type="integer", default=20, help="Number of MCMC batches. [default \"%default\"]" ),
+        make_option( c("-b","--blen"), type="integer", default=10, help="Length of each MCMC batch. [default \"%default\"]" ),
+        make_option( c("-m","--mmean"), type="double", default=1, help="Prior mean on single base mutation rates. [default \"%default\"]" ),
+        make_option( c("-o","--logfile"), type="character", default="", help="Direct output to this file. [default appends .Rout]" )
+    )
+opt <- parse_args(OptionParser(option_list=option_list,description=usage))
+attach(opt)
+options(error=traceback)
+
+lwin <- rwin <- 0; win <- 1
+winlen <- lwin + rwin + win
+
+if (is.null(infile)) { cat("Run\n  cpg-inference.R -h\n for help.") }
+
+if (logfile=="" & !interactive()) { 
+    logfile <- gsub(".RData",".Rout",infile,fixed=TRUE)
+    logcon <- if (logfile=="-") { stdout() } else { file(logfile,open="wt") }
+    sink(file=logcon, type="message") 
+    sink(file=logcon, type="output")   # send both to log file
+}
+
+scriptdir <- "../"
+source(paste(scriptdir,"codon-inference-fns.R",sep=''))
+source(paste(scriptdir,"sim-context-fns.R",sep=''))
+
+require(mcmc)
+
+load(infile)
+basedir <- gsub(".RData","",infile,fixed=TRUE)
+if (!file.exists(basedir)) { dir.create(basedir) }
+basename <- paste(basedir,"/nocontext",sep='')
+datafile <- paste( basename ,"-results.RData",sep='')
+resultsfile <- paste( basename ,"-results.tsv",sep='')
+plotfile <- paste( basename ,"-plot",sep='')
+
[…]

### 2013-10-14 writeups/context-dependence.tex
.

@@
       \hline
-      $+++$  &   $+-+$   &  $\lambda/1+e^{2\beta - \gamma})$ \\
+      $+++$  &   $+-+$   &  $\lambda/(1+e^{2\beta - \gamma})$ \\
       $++-$  &   $+--$   &  $\lambda/(1+e^{-\gamma})$ \\
@@
 This has the nice property that you can choose priors that are consistent across different values of $m$ and $\ell$.
-We'd have to still MCMC around in the space of $\pi(z)$, thought, so they'd be nuisance parameters from that point of view.
+We'd have to still MCMC around in the space of $\pi(z)$, so they'd be nuisance parameters from that point of view.
 Also, making the frequencies at the root random introduces correlations between counts, which we want to avoid.
@@
 \begin{align}  \label{eqn:root_mrf}
-    \pi(z) = \frac{1}{Z} \exp\left( - \sum_{i=1}^{|z|} h(z_{i-1},z_i,z_{i+1}) \right),
+    \pi(z) = \frac{1}{Z} \exp\left( - \sum_{i=1}^{|z|} \phi(z_{i-1},z_i,z_{i+1}) \right),
 \end{align}
@@
 
-  \item Compute $M_4(x,(y,z)) = \sum_{v \in \calS^{m+2\ell}} p_{m+2\ell}(t_{A},v,x) M_3(v,z) M_4(v,y)$ (stored as a $|\calS|^{m+2\ell} \times |\calS|^{2m}$ matrix).
+  \item Compute $M_4(x,(y,z)) = \sum_{v \in \calS^{m+2\ell}} p_{m+2\ell}(t_{A},v,x) M_2(v,z) M_3(v,y)$ (stored as a $|\calS|^{m+2\ell} \times |\calS|^{2m}$ matrix).
 
@@
 and pattern occurrences counted.
-As for TASEP, unning 10,000 MCMC iterations with window sizes of $\ell=w=3$ (total length 9) on these takes around 40 minutes.
+As for TASEP, running 10,000 MCMC iterations with window sizes of $\ell=w=3$ (total length 9) on these takes around 40 minutes.
 In this case, the true parameter values have good distribution within the estimated posterior distributions.
@@
 
+sergey:
+\begin{itemize}
+
+  \item ancestral CpG
+
+  \item ancient DNA
+
+\end{itemize}
+
 %%%%%%%%

### 2013-10-14 writeup-plots.R

@@ -30,7 +30,11 @@ all.mrun <- do.call( rbind, lapply( seq_along(mcmcdatafiles), function (k) {
             tmp
         } ) )
 
-subseq <- seq(1,nrow(all.mrun),by=max(1,floor(nrow(all.mrun)/1000)))
+win.1.1 <- which( grepl("1-1-1",mcmcdatafiles) & table( all.mrun$mcmcnum ) >= 10000 )
+win.3.0 <- which( grepl("0-3-0",mcmcdatafiles) & table( all.mrun$mcmcnum ) >= 10000 )
+win.3.3 <- which( grepl("3-3-3",mcmcdatafiles) & table( all.mrun$mcmcnum ) >= 10000 )
+
+subseq <- seq( min(which(all.mrun$mcmcnum %in% win.3.3)), max(which(all.mrun$mcmcnum %in% win.3.3)), by=max(1,floor(sum(all.mrun$mcmcnum %in% win.3.3)/1000)) )
 
 x <- rbind( all.mrun[subseq,(-1)*match("mcmcnum",colnames(all.mrun))], truth ) 
 cols <- c( adjustcolor(rainbow(64),.3)[1+floor(65*(1:(nrow(x)-1))/nrow(x))], adjustcolor("black",.5) )
@@ -44,7 +48,7 @@ for (k in 1:3) {
     thatone <- match(plotthese[1+k%%3],names(x))
     plot( x[,thisone], x[,thatone], xlab=plotthese[k], ylab=plotthese[1+k%%3], col=cols, cex=c(rep(.5,nrow(x)-1),2), pch=20 )
     text(truth[thisone],truth[thatone],"truth", pos=4)
-    hist( all.mrun[,thisone], main='', xlab=paste("posterior distribution of",plotthese[k]) )
+    hist( all.mrun[all.mrun$mcmcnum %in% win.3.3,thisone], main='', xlab=paste("posterior distribution of",plotthese[k]) )
     abline(v=truth[thisone], col='red')
 }
 dev.off()
@@ -52,13 +56,17 @@ dev.off()
 # for talk
 pdf(file=paste(plotfile,"-estimate-hists.pdf",sep=''),width=3,height=3,pointsize=10)
 par(mar=c(4,3,.3,0)+.1,mgp=c(2.1,1,0))
-tmp <- sweep(all.mrun[,1:3],2,truth,"/")
+
+for (mcnums in list( win.3.3, win.3.0, win.1.1 ) ) {
[…]

### 2013-10-14 mtCDNA/mtCDNA-inference.R
.

@@ -7,6 +7,7 @@ Infer parameters from output of sim-tree-cpg.R .\
 
 option_list <- list(
         make_option( c("-i","--infile"), type="character", default="mtCDNApri-sub.nuc", help=".nuc file with data. [default \"%default\"]"),
+        make_option( c("-f","--basedir"), type="character", default="", help="Directory for output. [default infile with suffix removed]"),
         make_option( c("-w","--win"), type="integer", default=2, help="Size of matching window. [default \"%default\"]" ),
         make_option( c("-l","--lwin"), type="integer", default=1, help="Size of left-hand context. [default \"%default\"]" ),
         make_option( c("-r","--rwin"), type="integer", default=1, help="Size of left-hand context. [default \"%default\"]" ),
@@ -33,6 +34,8 @@ if (gmfile=="TRUE") { gmfile <- paste(paste("genmatrices/genmatrix",winlen,bound
 
 if (countfile=="") { countfile <- paste(paste("countdata/counts",lwin,win,rwin,sep="-"),".RData",sep="") }
 
+if (basedir=="") { basedir <- gsub(".nuc","",infile,fixed=TRUE) }
+
 
 # took mtCDNApri.nuc from paml/examples/mtCDNA
 #  and: removed extra spaces on first line (only one space between numbers)
@@ -45,7 +48,6 @@ scriptdir <- "../"
 source(paste(scriptdir,"codon-inference-fns.R",sep=''))
 source(paste(scriptdir,"sim-context-fns.R",sep=''))
 
-basedir <- gsub(".nuc","",infile,fixed=TRUE)
 if (!file.exists(basedir)) { dir.create(basedir) }
 basename <- paste(basedir,"/win-",lwin,"-",win,"-",rwin,sep='')
 datafile <- paste( basename ,"-results.RData",sep='')

### 2014-05-08 gamma-prob.tex
gamma things

\documentclass{article}
\usepackage{amsmath,amssymb,amsthm}
\newcommand{\E}{\mathbb{E}}
\renewcommand{\P}{\mathbb{P}}
\newcommand{\given}{\;\vert\;}
\newcommand{\Beta}{B}

\begin{document}

Suppose that $(X_t)_{t\ge0}$ is a continuous-time Markov process with generator matrix $G$,
i.e.\ $G_{ij}>0$ if $i\neq j$ and rows of $G$ sum to zero.
Let $\tau$ be an independent Gamma distributed time with shape parameter $\alpha$ and scale parameter $\beta$.
We would like to compute the probabilities
\[
  Q(x,y;\alpha,\beta) := \P\{ X_\tau = y \given X_0 = x \} 
\]
in an efficient manner,
focusing on the case when $G$ is large and sparse.

Consider the corresponding ``jump chain'',
which represents $X$ as $X_t = M_{N(\lambda t)}$,
where $N(t)$ is a unit Poisson process and $M$ is a discrete-time Markov chain.
Here $\lambda$ is the maximum transition rate $\lambda = \max_i \{ - G_{ii} \}$,
and the transition matrix for $M$ is given by
\begin{align}
    P_{xy} := \P\{M_{k+1} = y \given M_k = x\} \\
        &= \begin{cases}  
                G_{xy}/\lambda \qquad & x\neq y \\
                1+G_{xx}/\lambda \qquad & x=y 
            \end{cases} .
\end{align}

Since $N(\lambda \tau)$ has a negative binomial distribution,
with 
\begin{align}
    \P\{ N(\lambda \tau) = n \} = \frac{ \beta^\alpha \lambda^n }{ \Beta(n,\alpha) (\beta+\lambda)^{n+\alpha} }  \qquad \text{for } n \ge 0 ,
\end{align}
it follows that
\begin{align}
    Q(x,y;\alpha,\beta) = \sum_{n \ge 0} \P\{ N(\lambda \tau) = n \} P^n  ,
\end{align}
and the error in approximating this by a finite sum up to $n_0$ is bounded by $\P\{N(\lambda \tau)>n_0\}$.
Since $\E[\tau] = \alpha/\beta$, this tail probability will decay exponentially once $n_0$ is larger than

\end{document}

### 2014-05-08 writeups/context-dependence.tex
ericks comments
Clearing out trailing whitespace.
Tiny fix.

@@
 (To avoid edge effects, we sometimes arrange these sites in a circle, so the indices are interpreted $\mod n$.)
-The time evolution is Markov, and is determined by a set of transition rates and associated patterns.  
-``Patterns'' are sequences of states, and we let $|u|$ denote the length of the pattern $u$.
-Transition rates are assumed to be local and homogeneous, so for example saying ``pattern $u$ changes to $v$ at rate $\mu$''
-means that if $(X_i(t), \ldots, X_{i+\ell}(t)) = (u_1, \ldots, u_\ell)$ (where $\ell=|u|=|v|$),
-then $\P\{(X_i(t+dt), \ldots, X_{i+\ell}(t+dt)) = (v_1, \ldots, v_\ell) = \mu dt + o(dt)$,
+We define a \textit{pattern} to mean a contiguous sequence of states, and we let $|u|$ denote the length of the pattern $u$.
+The marginal mutation process at every context is continuous time homogeneous Markov, which are thus specified by a set of transition rates between patterns.
+The sequence-level transition process is the independent superposition of all of the individual pattern mutation processes.
+Thus given two patterns $u$ and $v$ of common length $\ell$, saying ``pattern $u$ changes to $v$ at rate $\mu$'' means that if $(X_i(t), \ldots, X_{i+\ell}(t)) = (u_1, \ldots, u_\ell)$ then
+\[
+\P\{(X_i(t+dt), \ldots, X_{i+\ell}(t+dt)) = (v_1, \ldots, v_\ell) = \mu dt + o(dt),
+\]
 regardless of the location $i$ along the sequence.
@@
 It will often be convenient, notationally and computationally, to avoid edge effects
-by treating the $n$ sites as a circle, and treating all indices $\mod n$.
+by treating the $n$ sites as a circle, and treating all indices modulo $n$.
 
@@
 Covariances are also easy to compute.
-If we define $M_i(x,y)$ to be 1 if $X_{i-\ell}^{|x|}(0)=x$ and $X_i^({|y|})(t)=y$, and 0 otherwise,
+If we define $M_i(x,y)$ to be 1 if $X_{i-\ell}^{(|x|)}(0)=x$ and $X_i^{(|y|)}(t)=y$, and 0 otherwise,
 so that $N(x,y) = \sum_{i=1}^n M_i(x,y)$, then

### 2014-05-08 writeups/context-dependence.tex
Minor fixen.

@@
 \usepackage{graphicx}
+\usepackage{bf}
+\usepackage{todonotes}
+
+\newcommand{\peter}[1]{\todo[color=green!40]{Peter #1}}
+\newcommand{\erick}[1]{\todo[color=purple!40]{Erick: #1}}
 
@@
       \hline
-      01  &   10   &  $\lambda$
+      10  &   01   &  $\lambda$
     \end{tabular}
@@
 
-\begin{example}[Ising model with Glauber dynamics]
+\begin{example}[1-D Ising model with Glauber dynamics]
   In the Ising model, each site is labeled as either ``up'' or ``down'' ($+1$ or $-1$ respectively),
@@
   and the energy associated with a given state $x$ is $H(x) = - \beta \sum_i x_i x_{i+1} - \gamma \sum_i x_i$.
+  Here $\beta$ represents inverse temperature, and $\gamma$ represents the strength of the magnetic field.
   The associated stationary distribution on configurations is proportional to $\exp(-H(x))$.
@@
       \hline
-      $+++$  &   $+-+$   &  $\lambda/(1+e^{2\beta - \gamma})$ \\
+      $+++$  &   $+-+$   &  $\lambda/\,(1+e^{2\beta - \gamma})$ \\
       $++-$  &   $+--$   &  $\lambda/(1+e^{-\gamma})$ \\

### 2014-05-08 writeups/context-dependence.tex
ising magnetization reversed

@@
     \begin{tabular}{c@{\quad$\to$\quad}c@{\quad at rate\quad }c}
-      $u$  &  $v$  &  $\mu$  \\
-      \hline
       10  &   01   &  $\lambda$
@@
     \begin{tabular}{c@{\quad$\to$\quad}c@{\quad at rate\quad }cc}
-      $u$  &  $v$  &  $\mu$  & \\
-      \hline
       $x$  &  $y$  &  $m_{xy}$ & \qquad $x \neq y \in \{\nA,\nC,\nG,\nT\}$  \\
@@
   imagined as magnetic dipoles,
-  and the energy associated with a given state $x$ is $H(x) = - \beta \sum_i x_i x_{i+1} - \gamma \sum_i x_i$.
+  and the energy associated with a given state $x$ is $H(x) = - \beta/2 \sum_i x_i x_{i+1} - \gamma/2 \sum_i x_i$.
   Here $\beta$ represents inverse temperature, and $\gamma$ represents the strength of the magnetic field.
@@
   This is known as ``Glauber dynamics'', and ignoring transitions that don't change the state,
-  we get that if patterns $u$ and $v$ differ at only one site, then $u$ changes to $v$ at rate $1/(1+\exp(H(u)-H(v)))$, so
+  we get that if patterns $u$ and $v$ differ at only one site, then $u$ changes to $v$ at rate $1/(1+\exp(H(v)-H(u)))$, so
 
@@
     \begin{tabular}{c@{\quad$\to$\quad}c@{\quad at rate\quad }c}
-      $u$  &  $v$  &  $\mu$  \\
-      \hline
-      $+++$  &   $+-+$   &  $\lambda/\,(1+e^{2\beta - \gamma})$ \\
-      $++-$  &   $+--$   &  $\lambda/(1+e^{-\gamma})$ \\
-      $+-+$  &   $+++$   &  $\lambda/(1+e^{-2\beta + \gamma})$ \\
-      $+--$  &   $++-$   &  $\lambda/(1+e^{\gamma})$ \\
-      $-++$  &   $--+$   &  $\lambda/(1+e^{-\gamma})$ \\
-      $-+-$  &   $---$   &  $\lambda/(1+e^{-2\beta - \gamma})$ \\
-      $--+$  &   $-++$   &  $\lambda/(1+e^{\gamma})$ \\
-      $---$  &   $-+-$   &  $\lambda/(1+e^{2\beta + \gamma})$
+      $+++$  &   $+-+$   &  $\lambda/\,(1+e^{2\beta + \gamma})$ \\
+      $++-$  &   $+--$   &  $\lambda/(1+e^{\gamma})$ \\
+      $+-+$  &   $+++$   &  $\lambda/(1+e^{-2\beta - \gamma})$ \\
+      $+--$  &   $++-$   &  $\lambda/(1+e^{-\gamma})$ \\
+      $-++$  &   $--+$   &  $\lambda/(1+e^{\gamma})$ \\
+      $-+-$  &   $---$   &  $\lambda/(1+e^{-2\beta + \gamma})$ \\
+      $--+$  &   $-++$   &  $\lambda/(1+e^{-\gamma})$ \\
+      $---$  &   $-+-$   &  $\lambda/(1+e^{2\beta - \gamma})$
     \end{tabular}

### 2014-05-09 writeups/context-dependence.tex
Suggesting a change to avoid letter conflict.
(With *s* for state above.)
A few notes for @petrelharp.

@@
 and then resolve each possible change in temporal order.
+\todo{OK, rad.}
 We situate ``changes'' at the beginning of the matching pattern,
@@
 \]
-Now sample possible changes: let $s_{ij}$ denote the $j^\text{th}$ possible change at site $i$,
-so that $0 \le s_{i1} < s_{i2} < \cdots < s_{iN_i} \le t$ is a rate $\mu_*$ Poisson process.
+\todo[inline]{OK, I'm getting confused now. From the above, I would have thought that $\mu_*$ was the overall rate of mutation, but now it's only happening from pattern x?}
+Now sample possible changes: let $c_{ij}$ denote the $j^\text{th}$ possible change at site $i$,
+so that
+\[
+0 \le c_{i1} < c_{i2} < \cdots < c_{iN_i} \le t
+\]
+is a rate $\mu_*$ Poisson process.
+\todo[inline]{Further confusion: now $\mu_*$ is just the rate at a single site?}
 Then, suppose we have determined the state at time $s$ to be $X(s) = x$,

### 2014-05-09 writeups/context-dependence.tex

@@
 \usepackage{graphicx}
-\usepackage{bf}
+\usepackage{bm}
 \usepackage{todonotes}
@@
     \begin{tabular}{c@{\quad$\to$\quad}c@{\quad at rate\quad }cc}
-      $x$  &  $y$  &  $m_{xy}$ & \qquad $x \neq y \in \{\nA,\nC,\nG,\nT\}$  \\
+      $x$  &  $y$  &  $m_{xy}$ & when $x \neq y \in \{\nA,\nC,\nG,\nT\}$  \\
       CG   &  TG   &  $\gamma$ & \\
@@
   \end{center}
+  in which $\gamma$ is the additional CpG rate above the base mutation rate.
 
@@
   imagined as magnetic dipoles,
-  and the energy associated with a given state $x$ is $H(x) = - \beta/2 \sum_i x_i x_{i+1} - \gamma/2 \sum_i x_i$.
-  Here $\beta$ represents inverse temperature, and $\gamma$ represents the strength of the magnetic field.
+  and the energy associated with a given state $x$ is $H(x) = - \frac{1}{2} \beta \sum_i x_i x_{i+1} - \frac{1}{2} \gamma \sum_i x_i$.
+  Here $\beta$ represents inverse temperature, and $\gamma$ represents the strength of the magnetic field (here scaled by temperature).
   The associated stationary distribution on configurations is proportional to $\exp(-H(x))$.
@@
   \end{center}
-
-  This has three parameters: the speed $\lambda$, the inverse temperature $\beta$, and the strength of magnetization $\gamma$
-  (here scaled by temperature).
 
@@
 \begin{align} \label{eqn:full_likelihood}
-    p_n(t;x,y) := \P\{ X(t) = y \mid X(0) = x \} = (e^{tG(n)})_{xy} .
+    p_n(t;x,y) := \P\{ X(t) = y \mid X(0) = x \} = \left(e^{tG(n)}\right)_{xy} ,
 \end{align}
+but in practice the size of these matrices can make a direct application of this formula inefficient.
+This paper proposes an alternate strategy.
 

### 2014-05-09 writeups/context-dependence.tex
OK, that's all for my previous comments.

@@
 this should be feasible.
-(If $t$ is too large, the process may ``saturate'' --
+If $t$ is too large, the process may ``saturate'' --
 if there are many changes at most sites, then in at least most models,
@@
 In TASEP, we lose all information as it approaches stationarity,
-in the Ising model we have information about $\beta$ and $\gamma$ but not $\lambda$,
+in the Ising model we have information about $\beta$ and $\gamma$ but not $\lambda$ from the stationary distribution,
 and in the CpG model it is not immediately clear.
+\todo[inline]{I would think that this is identifiable from the stationary distribution if we know the structure of the model.}
 
@@
 so the full likelihood function is given by \eqref{eqn:full_likelihood},
-but doing anything with the $|\calS|^n \times |\calS|^n$ matrix $G_n$ is clearly infeasible.
+but doing anything with the $|\calS|^n \times |\calS|^n$ matrix $G_n$ is clearly infeasible for even moderately sized $n$.
 We can, however, compute \eqref{eqn:full_likelihood} for smaller $n$,
@@
 \]
-Variations are possible (what do we do about the edge effects for $p_m$?),
-but in any case there is no guarentee that this is close to correct,
-as it ignore dependencies between neighboring blocks.
+\todo{Do you prefer 1-indexing or 0-indexing for the sequences? Seems like 1-indexing before}
+Variations are possible, such as modifying edge effects behavior for $p_m$.
 
+However, in any case there is no guarantee that this is close to correct,
+as it ignores dependencies between neighboring blocks.
 Intuitively, we want to increase the size of the context
 -- but increasing $m$ does not do this, as there are always edge effects.
+\todo{I'm a little confused by this use of "edge effects." Are you using it to mean non-independence between blocks generally?}
 However, we could look at the context in the initial sequence but not in the final sequence,
@@
 In a long sequence, $p_m(x,z)$ might not be a very good approximation for $\P\{ X_i^{(m)}(t) = z | X_i(0)^{(m)} = x \}$,
-we do expect $p_{\ell,m,r}(x,y)$ to be correct, asymptotically in $\ell$ and $r$:
+we do expect $p_{\ell,m,r}(x,y)$ to asymptotically correct when $\ell$ and $r$ are large:
 \begin{align} \label{eqn:window_approx}
@@
 \end{align}
-We provide a concrete statement of this approximation later.
+We provide a precise statement of this approximation later.
 
@@
 
-Now we know how to compute the (marginal) probability that a particular subsequence changes into another subsequence.
+Now we know how to compute the approximate (marginal) probability that a particular subsequence changes into another subsequence.
 For patterns $x$ and $y$, with offset $\ell$ (so that $|x|=|y|+2\ell$),

### 2014-05-09 writeups/context-dependence.tex
Swapping h for ell. Closes #1.

@@
 The sequence-level transition process is the independent superposition of all of the individual pattern mutation processes.
-Thus given two patterns $u$ and $v$ of common length $\ell$, saying ``pattern $u$ changes to $v$ at rate $\mu$'' means that if $(X_i(t), \ldots, X_{i+\ell}(t)) = (u_1, \ldots, u_\ell)$ then
+Thus given two patterns $u$ and $v$ of common length $h$, saying ``pattern $u$ changes to $v$ at rate $\mu$'' means that if $(X_i(t), \ldots, X_{i+h}(t)) = (u_1, \ldots, u_h)$ then
 \[
-\P\{(X_i(t+dt), \ldots, X_{i+\ell}(t+dt)) = (v_1, \ldots, v_\ell) = \mu dt + o(dt),
+\P\{(X_i(t+dt), \ldots, X_{i+h}(t+dt)) = (v_1, \ldots, v_h) = \mu dt + o(dt),
 \]
@@
 with which the process in state $x$ jumps to state $y$.
-% Let $x_i^{(\ell)} = (x_i, x_{i+1}, \ldots, x_{i+\ell-1})$ be the subsequence of length $\ell$ beginning at location $i$, and
-For each $1\le i \le n$, pattern length $\ell \ge 0$, and patterns $u,v \in \calS^\ell$ define the relation
+% Let $x_i^{(h)} = (x_i, x_{i+1}, \ldots, x_{i+h-1})$ be the subsequence of length $h$ beginning at location $i$, and
+For each $1\le i \le n$, pattern length $h \ge 0$, and patterns $u,v \in \calS^h$ define the relation
 \[
@@
   x_k = y_k \quad &\text{for } k<i \\
-  x_{i+k} = u_k \quad &\text{for } 0 \le k < \ell \\
-  y_{i+k} = v_k \quad &\text{for } 0 \le k < \ell, and \\
-  x_k = y_k \quad &\text{for } k\ge i+\ell ,
+  x_{i+k} = u_k \quad &\text{for } 0 \le k < h \\
+  y_{i+k} = v_k \quad &\text{for } 0 \le k < h, and \\
+  x_k = y_k \quad &\text{for } k\ge i+h ,
 \end{cases}
 \]
-i.e.\ if $x$ and $y$ match except for in $i,i+1,\ldots,i+\ell-1$, where $x$ matches with $u$ and $y$ matches with $v$.
+i.e.\ if $x$ and $y$ match except for in $i,i+1,\ldots,i+h-1$, where $x$ matches with $u$ and $y$ matches with $v$.
 (Interpet these indices modulo $n$, even for short sequences, a point we will come back to.)
@@
 Then, suppose we have determined the state at time $s$ to be $X(s) = x$,
-and the next event is $s_{ij} = \min \{ s_{k\ell} : s_{k\ell}>s \}$.
+and the next event is $s_{ij} = \min \{ s_{kh} : s_{kh}>s \}$.
 The process then jumps to a new state at times $s_{ij}$ with probability proportional to the corresponding rate.

### 2014-05-09 writeups/context-dependence.tex
"Fringe" definition.

@@
 
-\newcommand{\peter}[1]{\todo[color=green!40]{Peter #1}}
+\newcommand{\peter}[1]{\todo[color=green!40]{Peter: #1}}
 \newcommand{\erick}[1]{\todo[color=purple!40]{Erick: #1}}
@@
 \]
-\todo{Do you prefer 1-indexing or 0-indexing for the sequences? Seems like 1-indexing before}
+\todo[inline]{Do you prefer 1-indexing or 0-indexing for the sequences? Seems like 1-indexing before}
 Variations are possible, such as modifying edge effects behavior for $p_m$.
@@
 -- but increasing $m$ does not do this, as there are always edge effects.
-\todo{I'm a little confused by this use of "edge effects." Are you using it to mean non-independence between blocks generally?}
+\todo[inline]{I'm a little confused by this use of "edge effects." Are you using it to mean non-independence between blocks generally?}
 However, we could look at the context in the initial sequence but not in the final sequence,
 including the states of an additional $\ell$ and $r$ sites on the left and right, respectively.
+We will call these left and right \emph{fringes}.
 Concretely, define the marginal probabilities
@@
 Now we know how to compute the approximate (marginal) probability that a particular subsequence changes into another subsequence.
-For patterns $x$ and $y$, with offset $\ell$ (so that $|x|=|y|+2\ell$),
+For patterns $x$ and $y$, with both fringe widths equal to $\ell$ (so that $|x|=|y|+2\ell$),
 we can count how many times $x$ appears in $X(0)$,

### 2014-05-10 writeups/context-dependence.tex
A few notational tweaks.

@@
 Now we know how to compute the approximate (marginal) probability that a particular subsequence changes into another subsequence.
+Let $N(x)$ be the number of occurrences of a pattern $x$ in $X(0)$.
 For patterns $x$ and $y$, with both fringe widths equal to $\ell$ (so that $|x|=|y|+2\ell$),
-we can count how many times $x$ appears in $X(0)$,
+let $N_\ell(x,y)$ count these events, respectively be the number of times $x$ appears in $X(0)$,
 and how many times $x$ appears in $X(0)$ while $y$ appears at the same place in $X(t)$, offset by $\ell$.
-For instance, if $x$ is ACTCAC, $y$ is CGC, and $\ell=3$, this looks like:
+For instance, $N_3(\text{ACTCAC}, \text{CGC})$ counts the number of occurrences of
 \begin{center}
@@
 \end{center}
-where $\cdot$ could be any base.
-Let $N(x)$ and $N_\ell(x,y)$ count these events, respectively.
+where $\cdot$ represents any base.
 
 From the definitions above,
-the expected number of times $y$ would appear juxtaposed with $x$ is approximately $N(x)p_{\ell,m,r}(x,y)$.
+the expected number of times $y$ would appear juxtaposed with $x$ is approximately $N(x) p_{\ell,m,r}(x,y)$.
 Covariances are also easy to compute.

### 2014-05-10 writeups/context-dependence.tex
\newcommand{\lyl}{\ell,|y|,\ell}

@@
 by treating the $n$ sites as a circle, and treating all indices modulo $n$.
+\todo[inline]{So shall we commit to this? I think that would be best, and then explain any deviation from it individually.}
 
@@
 
+\newcommand{\lyl}{\ell,|y|,\ell}
+
+Now we will now calculate the mean and covariances of $N_\ell(x,y)$.
 From the definitions above,
-the expected number of times $y$ would appear juxtaposed with $x$ is approximately $N(x) p_{\ell,m,r}(x,y)$.
+the expected number of times $y$ would appear juxtaposed with $x$ is approximately $N(x) p_{\lyl}(x,y)$.
 Covariances are also easy to compute.
@@
   &= n \var[M_1(x,y)] + 2n\sum_{k=1}^n \cov[M_1(x,y),M_{k+1}(x,y)] \\
-  &\approx n p_{\ell,m}(x,y)(1-p_{\ell,m}(x,y)) + 2n\sum_{k=1}^{m+2\ell} \left( \E[M_1(x,y) M_{k+1}(x,y)] - p_{\ell,m}(x,y)^2 \right) \\
-  &= n p_{\ell,m}(x,y)(1-(1+2m+4\ell)p_{\ell,m}(x,y)) + 2n\sum_{k=1}^{m+2\ell} \E[M_1(x,y) M_{k+1}(x,y)] .
+  &\approx n p_{\lyl}(x,y)(1-p_{\lyl}(x,y)) + 2n\sum_{k=1}^{m+2\ell} \left( \E[M_1(x,y) M_{k+1}(x,y)] - p_{\lyl}(x,y)^2 \right) \\
+  &= n p_{\lyl}(x,y)(1-(1+2m+4\ell)p_{\lyl}(x,y)) + 2n\sum_{k=1}^{m+2\ell} \E[M_1(x,y) M_{k+1}(x,y)] .
 \end{align}
@@
     \cov[N(x_1,y_1),N(x_2,y_2)]
-    & \approx - n (2m+4\ell) p_{\ell,m}(x_1,y_1)p_{\ell,m}(x_2,y_2) \\
+    & \approx - n (2m+4\ell) p_{\lyl}(x_1,y_1)p_{\lyl}(x_2,y_2) \\
     & \qquad {} + n\sum_{k=1}^{m+2\ell} \E[M_1(x_1,y_1) M_{k+1}(x_2,y_2) + M_1(x_2,y_2) M_{k+1}(x_1,y_1)] .
@@
 The covariance calculations show that
-if $p_{\ell,m}(x,y)$ is small enough (i.e.\ the pattern is not common),
+if $p_{\lyl}(x,y)$ is small enough (i.e.\ the pattern is not common),
 then $N(x,y)$ is close to Poisson.
@@
 \begin{align} \label{eqn:prob_approx}
-  \left| \P\{ X_i^{(m)}(t) = y \mid X(0) \} - p_{\ell,m,r}(t;x,y) \right| \le 2 \P\left\{  H(\{I,\ldots,i+m\},0,t)\subseteq\{i-\ell,\ldots,i+m+r\} \right\}.
+  \left| \P\{ X_i^{(m)}(t) = y \mid X(0) \} - p_{\lyl}(t;x,y) \right| \le 2 \P\left\{  H(\{I,\ldots,i+m\},0,t)\subseteq\{i-\ell,\ldots,i+m+r\} \right\}.
 \end{align}

### 2014-05-10 writeups/context-dependence.tex
Minor changes.

@@
 We define a \textit{pattern} to mean a contiguous sequence of states, and we let $|u|$ denote the length of the pattern $u$.
+
 The marginal mutation process at every context is continuous time homogeneous Markov, which are thus specified by a set of transition rates between patterns.
@@
 with which the process in state $x$ jumps to state $y$.
-% Let $x_i^{(h)} = (x_i, x_{i+1}, \ldots, x_{i+h-1})$ be the subsequence of length $h$ beginning at location $i$, and
+Let $x_i^{(h)} = (x_i, x_{i+1}, \ldots, x_{i+h-1})$ be the subsequence of length $h$ beginning at location $i$.
 For each $1\le i \le n$, pattern length $h \ge 0$, and patterns $u,v \in \calS^h$ define the relation
@@
 ignoring what is present in the leftmost $\ell$ and rightmost $r$ positions.
-In a long sequence, $p_m(x,z)$ might not be a very good approximation for $\P\{ X_i^{(m)}(t) = z | X_i(0)^{(m)} = x \}$,
-we do expect $p_{\ell,m,r}(x,y)$ to asymptotically correct when $\ell$ and $r$ are large:
+\todo[inline]{Is the following notation right?
+It seems like taking it out is a viable option: "In a long sequence, $p_m(x,z)$ might not be a very good approximation for $\P\{ X_i^{(m)}(t) = z \mid X_i^{(m)}(0) = x \}$,"}
+We expect $p_{\ell,m,r}(x,y)$ to be asymptotically correct when $\ell$ and $r$ are large \todo{and $t$ is not} :
 \begin{align} \label{eqn:window_approx}
-  \P\{ X_i^{(m)}(t) = y | X_i(0)^{(\ell+m+r)} = x \} \approx p_{\ell,m,r}(x,y) .
+  p_{\ell,m,r}(x,y) \approx \P\{ X_i^{(m)}(t) = y \mid X_i^{(\ell+m+r)}(0) = x \}.
 \end{align}
@@
 
-Now we will now calculate the mean and covariances of $N_\ell(x,y)$.
+Now we will now calculate the expectation and covariances of $N_\ell(x,y)$.
 From the definitions above,
-the expected number of times $y$ would appear juxtaposed with $x$ is approximately $N(x) p_{\lyl}(x,y)$.
-Covariances are also easy to compute.
-If we define $M_i(x,y)$ to be 1 if $X_{i-\ell}^{(|x|)}(0)=x$ and $X_i^{(|y|)}(t)=y$, and 0 otherwise,
-so that $N(x,y) = \sum_{i=1}^n M_i(x,y)$, then
-\begin{align}
+\[
+\E [N_\ell(x,y)] \approx N(x) \, p_{\lyl}(x,y).
+\]
+To compute covariances, define $M_i(x,y)$ to be 1 if $X_{i-\ell}^{(|x|)}(0)=x$ and $X_i^{(|y|)}(t)=y$, and 0 otherwise,
+so that $N(x,y) = \sum_{i=1}^n M_i(x,y)$.
+\begin{align*}
   \var[N(x,y)] &= \sum_{i=1}^n \var[M_i(x,y)] + \sum_{i\neq j} \cov[M_i(x,y), M_j(x,y)] \\
-  &= n \var[M_1(x,y)] + 2n\sum_{k=1}^n \cov[M_1(x,y),M_{k+1}(x,y)] \\
+  &= n \var[M_1(x,y)] + 2n\sum_{k=1}^{n-1} \cov[M_1(x,y),M_{k+1}(x,y)] \\
   &\approx n p_{\lyl}(x,y)(1-p_{\lyl}(x,y)) + 2n\sum_{k=1}^{m+2\ell} \left( \E[M_1(x,y) M_{k+1}(x,y)] - p_{\lyl}(x,y)^2 \right) \\
@@
   \end{split}
-\end{align}
+\end{align*}
 

### 2014-05-12 writeups/context-dependence.tex
Confusion regarding the starting state.
@petrelharp-- could you please let me know if I'm off the mark with this
commit? Is X(0) drawn from a distribution or can we start it with an
arbitrary series of states?
A little clarification on my confusion.

@@
 
+The process starts with an arbitrary specified sequence of states $X(0)$.
 The marginal mutation process at every context is continuous time homogeneous Markov, which are thus specified by a set of transition rates between patterns.
@@
 so that $N(x,y) = \sum_{i=1}^n M_i(x,y)$.
+\todo[inline]{
+In the below, are you are using an equivalence
+\[
+\cov[M_1(x,y),M_{k+1}(x,y)] = \E[M_1(x,y) M_{k+1}(x,y)] - p_{\lyl}(x,y)^2
+\]
+How do you justify this?
+I don't think that $M_1(x,y)$ is equal to $p_{\lyl}(x,y)$: the former is defined in terms of observed events, whereas the latter is defined starting with "if we have a sequence $x$" without any reference to the frequency of $x$.
+For example, what if $x$ doesn't appear in $X(0)$?
+I was with you on $\E [N_\ell(x,y)] \approx N(x) \, p_{\lyl}(x,y)$.
+}
+
 \begin{align*}

### 2014-05-12 writeups/context-dependence.tex
minor edits

@@
 
-It will often be convenient, notationally and computationally, to avoid edge effects
+It will often be convenient in the notation to avoid edge effects
 by treating the $n$ sites as a circle, and treating all indices modulo $n$.
 \todo[inline]{So shall we commit to this? I think that would be best, and then explain any deviation from it individually.}
+\peter{I think it would be better to remove this, but I need to parse the remainder...}
 
@@
 \todo{OK, rad.}
-We situate ``changes'' at the beginning of the matching pattern,
-so $\mu_*$ can be defined as the maximum sum of rates matching  the start of a sequence:
+The rate $\mu_*$ should be the maximum ``local'' rate at which transitions occur at any position:
 \[
-\mu_* = \max \left\{ \sum_{y \in \calS^n} \sum_{j\in J(1,x,y)} \mu_j \st x \in \calS^n \right\} .
+\mu_* = \max \left\{ \sum_{y \in \calS^n} \sum_{j\in J(i,x,y)} \mu_j \qquad \st x \in \calS^n, \quad 1 \le i \le n \right\} .
 \]
 \todo[inline]{OK, I'm getting confused now. From the above, I would have thought that $\mu_*$ was the overall rate of mutation, but now it's only happening from pattern x?}
-Now sample possible changes: let $c_{ij}$ denote the $j^\text{th}$ possible change at site $i$,
-so that
+\peter{$\mu_*$ is the maximum transition rate out of any possible pattern $x$.  I think the notation is correct? I have tried to clarify.}
+Note that since mutation is homogeneous, the maximum rate of changes occuring at any position will be the same, except possibly near the ends of the sequence.
+Now sample possible changes: let $c_{ij}$ denote the time of the $j^\text{th}$ possible change at site $i$,
+so that for each site $i$,
 \[
@@
 \]
-is a rate $\mu_*$ Poisson process.
+is a rate $\mu_*$ Poisson process, and these are independent across sites.
 \todo[inline]{Further confusion: now $\mu_*$ is just the rate at a single site?}
+\peter{yes.}
 Then, suppose we have determined the state at time $s$ to be $X(s) = x$,
-and the next event is $s_{ij} = \min \{ s_{kh} : s_{kh}>s \}$.
-The process then jumps to a new state at times $s_{ij}$ with probability proportional to the corresponding rate.
+and the next event is $s_{ij} = \min_{k,h} \{ s_{kh} : s_{kh}>s \}$.
+The process then jumps to a new state with probability proportional to the corresponding rate.
 In other words, let
@@
 and let $q_0 = 1-\sum_{k=1}^{n_T} q_k$.
-By construction of $\mu_*$, $q_0\ge 0$.)
+(By construction of $\mu_*$, $q_0\ge 0$.)
 Now choose transition $k$ with probability $q_k$, and if $k>0$, replace $x_i^{|u_k|}$ with $v_k$.
@@
 if there are many changes at most sites, then in at least most models,
-we will lost most of the information about the dynamics,
+we will lose most of the information about the dynamics,
 retaining information only about the parameters that affect the stationary distribution.
@@
 
-\begin{align*}
+\begin{align}
   \var[N(x,y)] &= \sum_{i=1}^n \var[M_i(x,y)] + \sum_{i\neq j} \cov[M_i(x,y), M_j(x,y)] \\
@@
   \end{split}
-\end{align*}
+\end{align}
 

### 2014-05-13 writeups/context-dependence.tex
edited up until summary stats

@@
 \usepackage{graphicx}
+\usepackage{natbib}
 \usepackage{bm}
@@
 \newcommand{\cov}{\mathop{\mbox{cov}}}
+\newcommand{\given}{\;\mid\;}
 
@@
 
-The common framework of the above examples is a one-dimensional grid of $n$ sites,
-and each site is labeled with one of a finite set of states.
-These change with time -- so, write $X_i(t)$ for the state of the $i^\mathrm{th}$ site at time $t$,
-where each $X_i(t) \in \calS$, the set of possible states.
-(To avoid edge effects, we sometimes arrange these sites in a circle, so the indices are interpreted $\mod n$.)
-We define a \textit{pattern} to mean a contiguous sequence of states, and we let $|u|$ denote the length of the pattern $u$.
+Each example above describes a 1-dimensional grid of sites, each of which can take one of a finite set of states,
+and that switch randomly between states according to a local set of rules.
+Concretely, we suppose that we observe a finite collection of these sites at only a few times,
+and that the dynamics are Markov:
+with the set of possible states by $\calS$, 
+and $X_i(t)$ the state of site $i$ at time $t$,
+we assume that $\{X(t)\}_{t \ge 0}$ is a Markov process on $\calS^n$ 
+for which the probability a given site changes state in a small amount of time
+depends only on the sequence of states at nearby sites.
 
-The process starts with an arbitrary specified sequence of states $X(0)$.
-The marginal mutation process at every context is continuous time homogeneous Markov, which are thus specified by a set of transition rates between patterns.
-The sequence-level transition process is the independent superposition of all of the individual pattern mutation processes.
-Thus given two patterns $u$ and $v$ of common length $h$, saying ``pattern $u$ changes to $v$ at rate $\mu$'' means that if $(X_i(t), \ldots, X_{i+h}(t)) = (u_1, \ldots, u_h)$ then
+To make this notion of local dependence formal,
+first define a \textit{pattern} to mean a contiguous sequence of states; 
+let $|u|$ denote the length of the pattern $u$;
+and let $x_i^{(h)} = (x_i, x_{i+1}, \ldots, x_{i+h-1})$ denote the subsequence of $x$ of length $h$ beginning at location $i$.
+The dynamics are determined by the set of allowed transitions and associated rates:
+given two patterns $u$ and $v$ of common length $h$, 
+saying ``pattern $u$ changes to $v$ at rate $\mu$'' means that 
 \[
-\P\{(X_i(t+dt), \ldots, X_{i+h}(t+dt)) = (v_1, \ldots, v_h) = \mu dt + o(dt),
+    \P\{ X_i^{(h)}(t+dt) = v \given X_i^{(h)}(t) = u \} = \mu dt + o(dt),
 \]
@@
 If more than one pattern would match to cause the same change, their rates add.
-We call such a $(\mu,u,v)$ a ``transition triple''.
-The set of all transition triples $\calT = \{ (\mu^j,u^j,v^j) \}_{j=1}^{n_T}$
-determines the time evolution of the process.
+Such a rule is defined by its ``transition triple'' $(\mu,u,v)$.
+The set of all transition triples, $\calT$, determines the time evolution of the process.
 
-It will often be convenient in the notation to avoid edge effects
-by treating the $n$ sites as a circle, and treating all indices modulo $n$.
-\todo[inline]{So shall we commit to this? I think that would be best, and then explain any deviation from it individually.}
-\peter{I think it would be better to remove this, but I need to parse the remainder...}
+A final note: recall we are focusing on finite sequences.
+By default, we do not allow patterns to ``hang off'' the end of the sequence.
+This is natural in many situations, but in others we might want to other boundary conditions,
+modeling extant, but unobserved, neighbors.
+We will mention this when necessary.
 
@@
   This only has one parameter, the speed.
-  Note, however, that it may not be entirely trivial to estimate the speed,
-  since (on the circle) it may not be obvious which particle moved where.
+  If we allow particles to exit from the right, and new particles to enter from the left
+  (or, arrange them on a circle)
+  then from observing only starting and ending configurations 
+  we cannot tell which particles have moved where,
+  so it is not obvious how to estimate the speed.
 
@@
   in the most general model of single-site mutation, each of the 12 possible transitions occurs at its own rate.
-  Furthermore, it is a well-known observation that in many species, and certain contexts,
-  a C followed by a G has an additional, higher rate with either the C changing to a T or the G changing to an A.
-  This simplest model of context-dependent substitution rates in a genome sequence has rates
+  Furthermore, it is a well-known observation that in many species, 
+  adjacent, methylated CG dinucleotides (``CpG sites'') have a much higher mutation rate to TG and CA
+  than either single nucleotide change.
+  Embellishing the single-nucleotide model with this additional rate results in the model defined by
 
@@
       CG   &  CA   &  $\gamma$ &
-    \end{tabular}
+    \end{tabular} .
   \end{center}
-  in which $\gamma$ is the additional CpG rate above the base mutation rate.
+  Here $\gamma$ is the additional CpG rate above the base mutation rate,
+  and a total of 13 parameters.
+  Note that this model is very different in spirit to TASEP,
+  in which the total number of particles was conserved.
 
-  This has 13 parameters.
-  Note that there are no conserved quantities here, unlike in TASEP,
-  which we could imagine as indestructible particles moving.
+  In this model, for a given pair of sequences, there may be more than one way to mutate a given string $x$ to get $y$:
+  for instance, a change $\nA\nC\nG \to \nA\nT\nG$ could have occurred by a single $\nC \to \nT$ mutation,
+  or by a $\nC\nG \to \nT\nG$ mutation; 
+  and hence the total instantaneous rate at which ACG changes to ATG is $\gamma + m_{\nC\nT}$.
+  Although this means there are many possible parameterizations,
+  note that the model is identifiable, not overparameterized.
 
@@
 
+  Glauber dynamics have been widely studied... XXX
+
 \end{example}
@@
 \paragraph{The generator matrix}
-For clarity, we should define how the set of transition rates determines the transition rate matrix,
+For clarity, we describe how the set of transition rates determines the transition rate matrix,
 the $|\calS|^n \times |\calS|^n$ matrix $G(n)$ whose $(x,y)^\text{th}$ entry gives the instantaneous rate
 with which the process in state $x$ jumps to state $y$.
-Let $x_i^{(h)} = (x_i, x_{i+1}, \ldots, x_{i+h-1})$ be the subsequence of length $h$ beginning at location $i$.
+Recall that $x_i^{(h)} = (x_i, x_{i+1}, \ldots, x_{i+h-1})$ denotes the subsequence of length $h$ beginning at location $i$.
 For each $1\le i \le n$, pattern length $h \ge 0$, and patterns $u,v \in \calS^h$ define the relation
@@
 \]
-i.e.\ if $x$ and $y$ match except for in $i,i+1,\ldots,i+h-1$, where $x$ matches with $u$ and $y$ matches with $v$.
-(Interpet these indices modulo $n$, even for short sequences, a point we will come back to.)
-We'll also need the set of defined transitions in $\calT$
-that can move from $x$ to $y$ at a given location: let  $J(i,x,y) = \{ (\mu^j,u^j,v^j) \in \calT \st x \xrightarrow{i,u^j,v^j} y \}$.
-Then, the rate $G(n)_{x,y}$ is the sum of all matching transition rates,
+i.e.\ if $x$ and $y$ match except at positions $i,i+1,\ldots,i+h-1$, 
+and in this region $x$ matches with $u$ while $y$ matches with $v$.
+
+As noted above, there may be more than one way to mutate a sequence $x$ to get $y$:
+for each position $i$, these are those transitions in $\calT$ that can be applied at $i$
+to change $x$ into $y$, denoted  $J(i,x,y) = \{ (\mu^j,u^j,v^j) \in \calT \st x \xrightarrow{i,u^j,v^j} y \}$.
+The rate $G(n)_{x,y}$ is then the sum of all matching transition rates,
 namely
@@
 This is just to say that the transition rate from $x$ to $y$ is the sum of the mutation rates of all transition triples
-that could change $x$ into $y$.
-In principle, this gives us the transition probabilities for the process:
+that would change $x$ into $y$.
+In principle, this gives us the transition probabilities for the process by a matrix exponential:
 \begin{align} \label{eqn:full_likelihood}
@@
 \end{align}
-but in practice the size of these matrices can make a direct application of this formula inefficient.
+which would then provide a route to parameter estimation.
+In practice, the size of these matrices makes direct application obviously impractical.
 This paper proposes an alternate strategy.
@@
 
-First, it will be useful to have a brief description of how to simulate such a process.
-Briefly, we can put down a Poisson process of times and locations of possible changes,
-at an appropriate rate $\mu_*$,
-and then resolve each possible change in temporal order.
-\todo{OK, rad.}
-The rate $\mu_*$ should be the maximum ``local'' rate at which transitions occur at any position:
+First, it will be useful for later proofs to have an explicit construction of the process,
+also used for simulation,
+using a variant of the ``jump chain representation'' (also known as the Gillespie algorithm \citep{vankampen}).
+Briefly, this works by first sampling a homogeneous Poisson process of times and locations of possible changes,
+at an appropriate rate $\mu_*$ per site,
+and then resolving each possible change in temporal order.
+The rate $\mu_*$ should be the maximum ``local'' rate at which transitions occur, per site:
 \[
@@
 \]
+Note that since mutation is homogeneous, the maximum rate of changes occuring at any position will be the same, except possibly near the end of the sequence,
+so the maximization over $i$ is unnecessary.
 \todo[inline]{OK, I'm getting confused now. From the above, I would have thought that $\mu_*$ was the overall rate of mutation, but now it's only happening from pattern x?}
-\peter{$\mu_*$ is the maximum transition rate out of any possible pattern $x$.  I think the notation is correct? I have tried to clarify.}
-Note that since mutation is homogeneous, the maximum rate of changes occuring at any position will be the same, except possibly near the ends of the sequence.
+\peter{$\mu_*$ is the maximum transition rate per site.  better now?}
 Now sample possible changes: let $c_{ij}$ denote the time of the $j^\text{th}$ possible change at site $i$,
@@
 \]
-is a rate $\mu_*$ Poisson process, and these are independent across sites.
+is a rate $\mu_*$ Poisson process on $[0,t]$, and these are independent across sites.
 \todo[inline]{Further confusion: now $\mu_*$ is just the rate at a single site?}
 \peter{yes.}
-Then, suppose we have determined the state at time $s$ to be $X(s) = x$,
-and the next event is $s_{ij} = \min_{k,h} \{ s_{kh} : s_{kh}>s \}$.
-The process then jumps to a new state with probability proportional to the corresponding rate.
-In other words, let
+Now, suppose we have determined the state at time $s$ to be $X(s) = x$,
+and the next event is $c_{ij} = \min_{k,h} \{ c_{kh} : c_{kh}>s \}$.
+At $c_{ij}$, 
+a new subsequence is then chosen at position $i$
+with probability proportional to the corresponding rate:
 \[
-q_k = \begin{cases}
-  \mu_k/\mu_* \qquad & \text{if } x_i^{(|u_k|)} = u_k  \\
+q_j = \begin{cases}
+  \mu^j/\mu_* \qquad & \text{if } x_i^{(|u^j|)} = u^j  \\
   0 \qquad & \text{otherwise,}
@@
 \]
-and let $q_0 = 1-\sum_{k=1}^{n_T} q_k$.
+is the probability that $x_i^{|u^j|}$ is replaced with $v^j$,
+and no chance is made with the remaining probability $q_0 = 1-\sum_{j=1}^{n_T} q_j$.
 (By construction of $\mu_*$, $q_0\ge 0$.)
-Now choose transition $k$ with probability $q_k$, and if $k>0$, replace $x_i^{|u_k|}$ with $v_k$.
 
-For instance, to simulate TASEP, one has only to let $\mu_*=\lambda$, and at each $s_{ij}$
+For instance, to simulate TASEP, one has only to let $\mu_*=\lambda$, and at each $c_{ij}$
 check if at site $i$ there is a 1 followed by a 0,
@@
 If $n$ is large enough and $t$ is not too large,
-this should be feasible.
+this should be feasible -- there should be enough information in our observations to do this reliably.
 If $t$ is too large, the process may ``saturate'' --
 if there are many changes at most sites, then in at least most models,
-we will lose most of the information about the dynamics,
+we will lose most information about the dynamics,
 retaining information only about the parameters that affect the stationary distribution.
-In TASEP, we lose all information as it approaches stationarity,
-in the Ising model we have information about $\beta$ and $\gamma$ but not $\lambda$ from the stationary distribution,
-and in the CpG model it is not immediately clear.
+In TASEP with any boundary conditions, we lose all information as it approaches stationarity,
+while in the Ising model we have information about the temperature $\beta$ and magnetic field $\gamma$ but not the speed of changes, $\lambda$.
+In the CpG model it is not immediately clear what information is retained.
 \todo[inline]{I would think that this is identifiable from the stationary distribution if we know the structure of the model.}
+\peter{What is identifiable?  The entire model, up to a time change?  I go back and forth on this...}
 
@@
 so the first thing that one might think to do is to break the sequence up into many blocks of length $m$,
-and treat these as independent, taking $m$ as large as is feasible.
-Then we would define $p_m(t;x,y)$ to be the probability that a string $x$ of length $m$ evolves to string $y$ over time $t$,
-denote by $x_i^{(m)}$ the block of length $m$ beginning at index $i$,
-and obtain the approximate likelihood function
+and treat these as independent.
+Then, defining $p_m(t;x,y)$ to be the probability that a string $x$ of length $m$ evolves to string $y$ over time $t$,
+we would obtain the approximate likelihood function
 \[
-  \prod_{k=0}^{n/m} p_m(x_{km}^{(m)},y_{km}^{(m)}) .
+  \prod_{k=0}^{n/m} p_m(x_{km+1}^{(m)},y_{km+1}^{(m)}) .
 \]
-\todo[inline]{Do you prefer 1-indexing or 0-indexing for the sequences? Seems like 1-indexing before}
-Variations are possible, such as modifying edge effects behavior for $p_m$.
-
-However, in any case there is no guarantee that this is close to correct,
+However, there is no guarantee that this is even close to correct,
 as it ignores dependencies between neighboring blocks.
 Intuitively, we want to increase the size of the context
--- but increasing $m$ does not do this, as there are always edge effects.
-\todo[inline]{I'm a little confused by this use of "edge effects." Are you using it to mean non-independence between blocks generally?}
-However, we could look at the context in the initial sequence but not in the final sequence,
-including the states of an additional $\ell$ and $r$ sites on the left and right, respectively.
-We will call these left and right \emph{fringes}.
-Concretely, define the marginal probabilities
+-- and the likelihood thus obtained is asymptotically correct for $m$ large,
+but computationally, we are stuck with small values of $m$.
+
+Alternatively, we could take into account context in the initial sequence only,
+by including the states of an additional $\ell$ sites on both the left and right.
+The probability that an initial sequence $x$ of length $m+2\ell$ after time $t$ 
+is found to match a smaller, middle sequence $y$ of length $m$ is found by marginalizing:
 \begin{align}
-  p_{\ell,m,r}(t;x,y) &= \sum_{a \in \calS^\ell} \sum_{b \in \calS^r} p_{\ell+m+r}(t;x,a \join y \join b) , \quad \text{for}\; x \in \calS^{\ell+m+r}, \; y \in \calS^m ,
+  p_{\ell,m}(t;x,y) &= \sum_{a \in \calS^\ell} \sum_{b \in \calS^\ell} p_{2\ell+m}(t;x,a \join y \join b) , \quad \text{for}\; x \in \calS^{2\ell+m}, \; y \in \calS^m ,
 \end{align}
 where $a \join y \join b$ is the sequence composed of concatenating $a$, $y$, and $b$ together in that order.
-This is the probability that if we begin with a sequence $x$ of length $\ell+m+r$,
+This is the probability that if we begin with a sequence $x$ of length $2\ell+m$,
 then after time $t$ we will see at the middle $m$ positions the pattern $y$,
-ignoring what is present in the leftmost $\ell$ and rightmost $r$ positions.
-\todo[inline]{Is the following notation right?
-It seems like taking it out is a viable option: "In a long sequence, $p_m(x,z)$ might not be a very good approximation for $\P\{ X_i^{(m)}(t) = z \mid X_i^{(m)}(0) = x \}$,"}
-We expect $p_{\ell,m,r}(x,y)$ to be asymptotically correct when $\ell$ and $r$ are large \todo{and $t$ is not} :
+ignoring what is present in the leftmost and rightmost $\ell$ positions.
+We expect $p_{\ell,m}(x,y)$ to be asymptotically correct when $\ell$ is large (and $t$ is fixed),
+i.e.\ that the probability a given subsequence matching $x$ is later observed to have its center matching $y$
+is close to $p_{\ell,m}(x,y)$, regardless of context:
 \begin{align} \label{eqn:window_approx}
-  p_{\ell,m,r}(x,y) \approx \P\{ X_i^{(m)}(t) = y \mid X_i^{(\ell+m+r)}(0) = x \}.
+  p_{\ell,m}(x,y) \approx \P\{ X_i^{(m)}(t) = y \mid X_i^{(2\ell+m)}(0) = x \}.
 \end{align}
-We provide a precise statement of this approximation later.
+We provide a precise statement of the error in this approximation later.
 
@@
 \subsection{Summary statistics}
+
+Equipped with marginal transition probabilities from length-$m+2\ell$ patterns to length-$m$ patterns
+that correctly account for local dependencies,
+we can now consider inference procedures.
 

### 2014-05-14 writeups/context-dependence.tex
Zapping trailing whitespace.
A collection of trivialities.
@petrelharp: thank you for your edits! This all makes a lot more sense
now, and I feel silly for not realizing that you were talking about a
jump chain.

@@
 i.e.\ if $x$ and $y$ match except at positions $i,i+1,\ldots,i+h-1$,
-and in this region $x$ matches with $u$ while $y$ matches with $v$.
+and for those positions $x$ matches with $u$ while $y$ matches with $v$.
 
@@
 and then resolving each possible change in temporal order.
-The rate $\mu_*$ should be the maximum ``local'' rate at which transitions occur, per site:
+The rate $\mu_*$ should be the maximum ``local'' rate at which transitions occur, across sites and transition outcome states:
 \[
@@
 \]
-Note that since mutation is homogeneous, the maximum rate of changes occuring at any position will be the same, except possibly near the end of the sequence,
+Note that since mutation is homogeneous, the maximum rate of changes occurring at any position will be the same, except possibly near the end of the sequence,
 so the maximization over $i$ is unnecessary.
-\todo[inline]{OK, I'm getting confused now. From the above, I would have thought that $\mu_*$ was the overall rate of mutation, but now it's only happening from pattern x?}
-\peter{$\mu_*$ is the maximum transition rate per site.  better now?}
 Now sample possible changes: let $c_{ij}$ denote the time of the $j^\text{th}$ possible change at site $i$,
@@
 \]
-is a rate $\mu_*$ Poisson process on $[0,t]$, and these are independent across sites.
-\todo[inline]{Further confusion: now $\mu_*$ is just the rate at a single site?}
-\peter{yes.}
+is a draw from a rate $\mu_*$ Poisson process on $[0,t]$.
+These processes are independent across sites.
 Now, suppose we have determined the state at time $s$ to be $X(s) = x$,
@@
 At $c_{ij}$,
-a new subsequence is then chosen at position $i$
-with probability proportional to the corresponding rate:
+a new state is then chosen at position $i$
+with probability proportional to the corresponding rate, i.e.\
 \[
 q_j = \begin{cases}
-  \mu^j/\mu_* \qquad & \text{if } x_i^{(|u^j|)} = u^j  \\
+  \mu_j/\mu_* \qquad & \text{if } x_i^{(|u^j|)} = u^j  \\
   0 \qquad & \text{otherwise,}
@@
 \]
-is the probability that $x_i^{|u^j|}$ is replaced with $v^j$,
-and no chance is made with the remaining probability $q_0 = 1-\sum_{j=1}^{n_T} q_j$.
+is the probability that $x_i^{(|u^j|)}$ is replaced with $v^j$.
+The remaining probability $q_0 = 1-\sum_{j=1}^{|\calT|} q_j$ gives the probability that the state remains the same.
 (By construction of $\mu_*$, $q_0\ge 0$.)
@@
 In the CpG model it is not immediately clear what information is retained.
-\todo[inline]{I would think that this is identifiable from the stationary distribution if we know the structure of the model.}
 \peter{What is identifiable?  The entire model, up to a time change?  I go back and forth on this...}
+\todo[inline]{Not sure what I was smoking when I said that.
+I think that the stationary distribution of the single-nucleotide model should be identifiable for sure.
+Then, I would guess the two CpG rates are identifiable modulo a rate factor that compares the single-nucleotide mutation rate to the CpG mutation rate.}
 

### 2014-05-14 writeups/context-dependence.tex
writeup edits

--- 2014-05-13 writeups/context-dependence.tex
@@
 
-Equipped with marginal transition probabilities from length-$m+2\ell$ patterns to length-$m$ patterns
-that correctly account for local dependencies,
-we can now consider inference procedures.
-
-Now we know how to compute the approximate (marginal) probability that a particular subsequence changes into another subsequence.
-Let $N(x)$ be the number of occurrences of a pattern $x$ in $X(0)$.
-For patterns $x$ and $y$, with both fringe widths equal to $\ell$ (so that $|x|=|y|+2\ell$),
-let $N_\ell(x,y)$ count these events, respectively be the number of times $x$ appears in $X(0)$,
-and how many times $x$ appears in $X(0)$ while $y$ appears at the same place in $X(t)$, offset by $\ell$.
-For instance, $N_3(\text{ACTCAC}, \text{CGC})$ counts the number of occurrences of
+Since we cannot compute reasonably the full likelihood for the data,
+we will need to choose appropriate summary statistics
+that contain as much information about the process as possible,
+and have a tractable expression for their marginal likelihood.
+Assume here that $\ell$ has been chosen to be sufficiently large that approximation \eqref{eqn:window_approx} holds exactly.
+If $i_1 < \ldots < i_{n(x)}$ are a set of nonoverlapping sites where the initial sequence matches $x$
+-- i.e.\ $X_{i_k}^{(2\ell+m)}(0)=x$ and $i_{k+1} > i_k + 2\ell+m$ for all $k$ --
+then the counts $N(x,y)=\#\{ i_k \st X_{i_k+\ell}^{(m)}(t)=y\}$ (the number of these sites that match the smaller pattern $y$ at time $t$)
+are multinomial, with probabilities $\{p_{\ell,m}(x,y)\st y \in \calS^m\}$.
+For instance, if $\ell=3$ then $N(\text{ACTCAC}, \text{CGC})$ counts the number of occurrences of
 \begin{center}
@@
 where $\cdot$ represents any base.
+More generally, if we partition the sequence into windows of length $2\ell+m$,
+and let $N(x,y)$ be the number of times we see a window in which $x$ matches $X(0)$ and $y$ matches $X(t)$,
+then these counts are independently multinomial across each $x$.
+% so the marginal likelihood of the statistics $N(x,y)$ is
+% \begin{align}
+%     \prod_{x \in \calS^{2\ell+m}} N(x)! \prod_{y \in \calS^m} \frac{p_{\ell,m}(x,y)^{N(x,y)}}{N(x,y)!} ,
+% \end{align}
+% where $N(x) = \sum_y N(x,y)$.
 
-\newcommand{\lyl}{\ell,|y|,\ell}
+Of course, a drawback to this approach is that we throw away a good fraction of the data.
+In fact, we do not even look at a fraction $2\ell/(2\ell+m)$ of the second sequence, $X(t)$,
+that lies in the fringes of each window.
+As before, let $N(x)$ be the number of occurrences of a pattern $x$ in $X(0)$,
+and $N(x,y)$ the number of times $x$ is seen juxtaposed with $y$ in $X(t)$,
+but this time, count over all positions,
+so that e.g.\ $N(x) = \#\{ 1 \le i \le n - 2\ell - m + 1 \st X_i^{(m+2\ell)}(0)=x\}$.
 
-Now we will now calculate the expectation and covariances of $N_\ell(x,y)$.
-From the definitions above,
+\peter{Ooops, yes.  Previously, I was semi-inconsistently thinking of the initial sequence $X(0)$ as random as well. ,
+    which is why I was talking about Poisson instead of Multinomial counts.}
+
+Now we will now calculate the expectation and covariances of $N(x,y)$.
+\peter{The following is Not Right, but also we don't really use it at the moment.}
+If we assume that \eqref{eqn:window_approx} is exact, then
 \[
-\E [N_\ell(x,y)] \approx N(x) \, p_{\lyl}(x,y).
+    \E [N(x,y)] = N(x) \, p_{\ell,m}(x,y).
 \]
-To compute covariances, define $M_i(x,y)$ to be 1 if $X_{i-\ell}^{(|x|)}(0)=x$ and $X_i^{(|y|)}(t)=y$, and 0 otherwise,
-so that $N(x,y) = \sum_{i=1}^n M_i(x,y)$.
-\todo[inline]{
-In the below, are you are using an equivalence
-\[
-\cov[M_1(x,y),M_{k+1}(x,y)] = \E[M_1(x,y) M_{k+1}(x,y)] - p_{\lyl}(x,y)^2
-\]
-How do you justify this?
-I don't think that $M_1(x,y)$ is equal to $p_{\lyl}(x,y)$: the former is defined in terms of observed events, whereas the latter is defined starting with "if we have a sequence $x$" without any reference to the frequency of $x$.
-For example, what if $x$ doesn't appear in $X(0)$?
-I was with you on $\E [N_\ell(x,y)] \approx N(x) \, p_{\lyl}(x,y)$.
-}
-
+To compute covariances, define $\chi_i(x,y)$ to be 1 if $X_{i-\ell}^{(|x|)}(0)=x$ and $X_i^{(|y|)}(t)=y$, and 0 otherwise,
+so that $N(x,y) = \sum_{i=1}^n \chi_i(x,y)$.
+Overloading notation, also define $\chi_i(x)$ to be 1 if $X_{i-\ell}^{(|x|)}(0)=x$,
 \begin{align}
-  \var[N(x,y)] &= \sum_{i=1}^n \var[M_i(x,y)] + \sum_{i\neq j} \cov[M_i(x,y), M_j(x,y)] \\
-  &= n \var[M_1(x,y)] + 2n\sum_{k=1}^{n-1} \cov[M_1(x,y),M_{k+1}(x,y)] \\
-  &\approx n p_{\lyl}(x,y)(1-p_{\lyl}(x,y)) + 2n\sum_{k=1}^{m+2\ell} \left( \E[M_1(x,y) M_{k+1}(x,y)] - p_{\lyl}(x,y)^2 \right) \\
-  &= n p_{\lyl}(x,y)(1-(1+2m+4\ell)p_{\lyl}(x,y)) + 2n\sum_{k=1}^{m+2\ell} \E[M_1(x,y) M_{k+1}(x,y)] .
+  \var[N(x,y)] &= \sum_{i=1}^n \var[\chi_i(x,y)] + \sum_{i\neq j} \cov[\chi_i(x,y), \chi_j(x,y)] \\
+               &\approx \sum_{i=1}^n \chi_i(x) p_{\ell,m}(x,y)(1-p_{\ell,m}(x,y)) 
+            + 2 \sum_{i=1}^{n-2\ell-m} \sum_{j=i+1}^{i+2\ell+m-1} \cov[ \chi_i(x,y) \chi_j(x,y) ] \\
+            &= N(x) p_{\ell,m}(x,y)(1-p_{\ell,m}(x,y)) + 
 \end{align}
-The term $\E[M_1(x,y) M_{k+1}(x,y)]$ is the probability that $x$ matches at both site 1 and site $k+1$,
+The term $\E[\chi_1(x,y) \chi_{k+1}(x,y)]$ is the probability that $x$ matches at both site 1 and site $k+1$,
 and that $y$ matches both at site $\ell+1$ and at site $\ell+k+1$.
@@
     \cov[N(x_1,y_1),N(x_2,y_2)]
-    & \approx - n (2m+4\ell) p_{\lyl}(x_1,y_1)p_{\lyl}(x_2,y_2) \\
-    & \qquad {} + n\sum_{k=1}^{m+2\ell} \E[M_1(x_1,y_1) M_{k+1}(x_2,y_2) + M_1(x_2,y_2) M_{k+1}(x_1,y_1)] .
+    & \approx - n (2m+4\ell) p_{\ell,m}(x_1,y_1)p_{\ell,m}(x_2,y_2) \\
+    & \qquad {} + n\sum_{k=1}^{m+2\ell} \E[\chi_1(x_1,y_1) \chi_{k+1}(x_2,y_2) + \chi_1(x_2,y_2) \chi_{k+1}(x_1,y_1)] .
   \end{split}
@@
 The covariance calculations show that
-if $p_{\lyl}(x,y)$ is small enough (i.e.\ the pattern is not common),
+if $p_{\ell,m}(x,y)$ is small enough (i.e.\ the pattern is not common),
 then $N(x,y)$ is close to Poisson.
@@
 \begin{align} \label{eqn:prob_approx}
-  \left| \P\{ X_i^{(m)}(t) = y \mid X(0) \} - p_{\lyl}(t;x,y) \right| \le 2 \P\left\{  H(\{I,\ldots,i+m\},0,t)\subseteq\{i-\ell,\ldots,i+m+r\} \right\}.
+  \left| \P\{ X_i^{(m)}(t) = y \mid X(0) \} - p_{\ell,m}(t;x,y) \right| \le 2 \P\left\{  H(\{I,\ldots,i+m\},0,t)\subseteq\{i-\ell,\ldots,i+m+r\} \right\}.
 \end{align}

### 2014-05-28 writeups/context-dependence.tex
.
intro/lit review

@@
 
-Interacting particle systems with neighborhood structure: transition probabilities generally not expressable (except TASEP).
-Markov random fields with Glauber dynamics a case of this.
+It not uncommon to encounter models 
+consisting of a lattice of sites,
+each of which taking one of a finite collection of possible states,
+and whose stochastic, temporal evolution is Markov and governed by a set of local rules.
+For instance, at each site may sit 
+a nucelotide whose mutation rate depends on physical properties determined by the nearby DNA sequence;
+a cell whose infection status depends on the state of its neighbors;
+or a particle whose spin is perturbed by external noise to a state whose distribution depends on the local energy configuration.
+Practical use of such models often requires
+inferring transition rules 
+based on observations of the system at several time points,
+or of several states evolved along a tree from a single starting point.
+In this paper,
+we propose one solution to this problem,
+motivated by the problem of inferring context-dependent mutation rates from diverged nucleotide sequences.
 
-Example: Ising model.
+The general class of models we consider are known in the probability literature as
+interacting particle systems \citep{liggett2005ips} with neighborhood structure
+-- continuous-time Markov chains on lattice-indexed collections of states
+whose transition probabilities are \emph{local}
+in the sense that any instantaneous change only affects a small number of nearby sites,
+and the rates of such instantaneous changes depend only on the states in some bounded neighborhood of those sites to change.
+As finite-state Markov chains,
+transition probabilities are in principle simply expressable as a matrix exponential,
+but this is impractical as the size of the matrix is equal to the number of possible configurations.
+As an alternative, we show that the conditional likelihood of local pattern count statistics
+can be well-approximated by marginalizing over a finite amount of surrounding context.
 
-Context-dependent substitution processes are an example of interest for phylogenetics and evolutionary inference.
+The situation is similar to the widely-studied problem of inference based on observation 
+of a single instance of a Markov random field,
+and in many cases reduces to this if we only observe the system at one time point at stationarity.
+These are used, for instance, in spatial statistics \citep{besag1972nearestneighbour,gelfand2010handbook}
+and image reconstruction \citep{geman1984stochastic,besag1986statistical}.
+The conditioning method we consider here
+is simliar to the ``coding'' scheme introduced by \citet{besag1974spatial},
+that conditions on a set of sites that makes the remaining observations independent (thanks to the Gibbs property).
+A good review of recent statistical techniques is given by \citet{friel2012bayesian}.
+
+In the context of genomics,
+it is well-known that certain short nucleotide sequences 
+are in many organisms much more, or less, abundant than expected by chance \citep{karlin},
+in part due to context dependence of the nucleotide mutation and substitution processes.
+Phylogenetic and population genetic methods usually ignore such dependencies
+in the interest of computational efficiency and lack of understanding of the underlying process,
+but some progress has been made.
+XXX
 Usually dealt with either by using larger and larger windows (but, only up to four bp),
@@
 Also, Markov-along-the-genome method.
+Larribe and Fearnhead.
 
-The difficulty comes from the fact that the rates of change at any one site depend in principle on arbitrarily far away sites,
-so transition matrix is ginormous.
+Another canonical example is the Ising model of statistical physics
+with time evolution given by Glauber dynamics \citep{glauber1963timedependent},
+in which a lattice of up/down spins are perturbed by thermal noise, 
+relaxing into states dependent on the energy of the resulting configuration.
+Parameter estimation for Ising model without temporal dynamics
+is relatively well-undersood \citep{pickard1982inference,frigessi1990parameter},
+but there does not seem to be similar work on the case when more than one observation is made of the dynamical system.
 
-Goal: parameter inference, given before-after observations, or else observations descended from a common ancestor.
+The general framework also fits certain cellular automata models,
+e.g.\ modeling wildfire \citep{clarke1994cellular}, 
+the spread of HIV \citep{zorzenondossantos2001dynamics}, 
+or land use patterns \citep{clarke1996methods}.
+Complex models may introduce long-range dependencies beyond the scope of this paper,
+but these methods may still prove useful in the modeling process.
 
-Review stats lit, e.g. reviews by Friel, original by Besag (autologistic model), genetics by Larribe and Fearnead.
-
-Here restrict to one-dimensional neighborhood structure,
-to make it easier.
-
+%%%%%%% %%%%%%%%%%
 \section{Notation and Examples}
@@
 and have a tractable expression for their marginal likelihood.
-Assume here that $\ell$ has been chosen to be sufficiently large that approximation \eqref{eqn:window_approx} holds exactly.
+For simplicity, 
+we assume in this section that approximation \eqref{eqn:window_approx} holds exactly.
 If $i_1 < \ldots < i_{n(x)}$ are a set of nonoverlapping sites where the initial sequence matches $x$

### 2014-05-31 writeups/context-dependence.tex

@@
 \usepackage{natbib}
+\usepackage{hyperref}
 \usepackage{bm}
@@
 \newtheorem{example}{Example}[section]
+
+\bibliographystyle{plainnat}
 
@@
 is simliar to the ``coding'' scheme introduced by \citet{besag1974spatial},
-that conditions on a set of sites that makes the remaining observations independent (thanks to the Gibbs property).
+that conditions on a set of sites that makes the remaining observations independent (thanks to the Gibbs property);
+consistency of such methods has been shown by \citet{comets1992consistency} and reviewed by \citet{larribe2011composite}.
 A good review of recent statistical techniques is given by \citet{friel2012bayesian}.
@@
 it is well-known that certain short nucleotide sequences 
-are in many organisms much more, or less, abundant than expected by chance \citep{karlin},
-in part due to context dependence of the nucleotide mutation and substitution processes.
-Phylogenetic and population genetic methods usually ignore such dependencies
+are in many organisms much more, or less, abundant than expected by chance \citep{burge1992underrepresentation},
+due to the combined effects of context dependence of the nucleotide mutation process,
+selective constraints on the function of the sequence,
+and other processes such as biased gene conversion \citep{duret2009biased}.
+Indeed, molecular studies have demonstrated that the spectrum of new mutations in humans
+is highly context dependent \citep{schaibley2013influence} 
+and that multinucleotide substitutions are relatively common \citep{schrider2011pervasive,terekhanova2013prevalence}.
+(We are concerned with effects that are \emph{homogeneous} across the sequence,
+so do not consider further the many constraints on protein-coding sequence \citep[reviewed in][]{thorne2007protein}.)
+However, phylogenetic and population genetic methods usually ignore such dependencies
 in the interest of computational efficiency and lack of understanding of the underlying process,
 but some progress has been made.
-XXX
-Usually dealt with either by using larger and larger windows (but, only up to four bp),
-or by doing MCMC over additional information (entire history of changes).
-Also, Markov-along-the-genome method.
-Larribe and Fearnhead.
+Often, the effects of a complex mutational spectrum are explored using summary statistics of larger blocks or other sensible but \textit{ad hoc} methods.
+For instance, \citet{arndt2003sequence} studied dinucleotide transitions,
+and \citet{yaari2013models} displayed strong heterogeneity in the probability of synonymous mutations across 5-mers
+in B-cell immunoglobin genes.
+Others have made progress using simplifying assumptions \citep{berard2012accurate},
+or by other approximations \citep{christensen2005pseudolikelihood}.
+\citet{siepel2004phylogenetic} introduced a method to compute likelihoods under a context-dependent mutation model
+by moving iteratively along the genome,
+while
+\citet{lunter2004nucleotide} and \citet{hobolth2008markov} used data augmentation and an MCMC algorithm to do Bayesian inference;
+both these methods are promising but very computationally intensive.
 
@@
 the spread of HIV \citep{zorzenondossantos2001dynamics}, 
-or land use patterns \citep{clarke1996methods}.
+or land use patterns \citep{wu2002calibration}.
 Complex models may introduce long-range dependencies beyond the scope of this paper,
@@
 also used for simulation,
-using a variant of the ``jump chain representation'' (also known as the Gillespie algorithm \citep{vankampen}).
+using a variant of the ``jump chain representation''.
 Briefly, this works by first sampling a homogeneous Poisson process of times and locations of possible changes,
@@
 
+\bibliography{context-dependence}
+
 \end{document}

### 2014-06-03 writeups/context-dependence.tex
poster

@@
 is highly context dependent \citep{schaibley2013influence} 
-and that multinucleotide substitutions are relatively common \citep{schrider2011pervasive,terekhanova2013prevalence}.
+and that multinucleotide substitutions are relatively common \citep{schrider2011pervasive,terekhanova2013prevalence,harris2013errorprone}.
 (We are concerned with effects that are \emph{homogeneous} across the sequence,
@@
 or by other approximations \citep{christensen2005pseudolikelihood}.
-\citet{siepel2004phylogenetic} introduced a method to compute likelihoods under a context-dependent mutation model
-by moving iteratively along the genome,
-while
-\citet{lunter2004nucleotide} and \citet{hobolth2008markov} used data augmentation and an MCMC algorithm to do Bayesian inference;
-both these methods are promising but very computationally intensive.
+\citet{pedersen2000dependent}, later extended by 
+\citet{hobolth2008markov,baele2010using}, used data augmentation and an MCMC algorithm to do Bayesian inference;
+\citet{lunter2004nucleotide} used an approximate matrix decomposition;
+while \citet{siepel2004phylogenetic} and later \citep{baele2010modelling} computed likelihoods by assuming a model that is Markov along the genome;
+however, all these methods quite computationally intensive.
 

### 2014-06-03 sim-context-fns.R

@@ -206,7 +206,7 @@ show.simseq <- function (x,printit=FALSE,maxchar=min(nchar(x$initseq),200),latex
                 "\\end{tabular} \\end{center} \n" 
             ), collapse="" )
     }
-    if (printit) { lapply( outstrings, cat ) }
+    if (printit | latex) { lapply( outstrings, cat ) }
     return(invisible(outstrings))
 } 
 

### 2014-06-04 mus/adjust-resids.R
.

@@ -1,0 +1,61 @@
+#!/usr/bin/Rscript
+# usage:
+#  quick-residuals.R (length of window) (files to load)
+
+gmfile <- paste(paste("genmatrices/genmatrix",commandArgs[1],'none',0,sep="-"),".RData",sep='') 
+load(gmfile)
+big.genmat <- genmatrix
+
+for (x in commandArgs(TRUE)[-1]) { load(x) }
+
+if (mrun$nbatch * mrun$blen > 1e4) {
+    estimates <- rbind(estimates, 
+           final=c(mrun$final,1-sum(mrun$final[length(mrun$final)-0:2]),NA),
+           mean=c(colMeans(mrun$batch),1-sum(colMeans(mrun$batch[,length(mrun$final)-0:2])),NA))
+}
+
+counts <- lapply( list(infile,revfile), function (ifile) {
+        count.table <- read.table(ifile,header=TRUE,stringsAsFactors=FALSE)
+        counts <- Matrix(0,nrow=nrow(genmatrix),ncol=ncol(projmatrix))
+        rownames(counts) <- rownames(genmatrix)
+        colnames(counts) <- colnames(projmatrix)
+        stopifnot( all( count.table$reference %in% rownames(genmatrix) ) & all(count.table$derived %in% colnames(projmatrix)) ) 
+        counts[cbind( match(count.table$reference,rownames(genmatrix)), match(count.table$derived,colnames(projmatrix)) )] <- count.table$count
+        return(counts)
+    } )
+
+expected <- function (x) {
+            x <- as.numeric(x)
+            branchlens <- c(x[1],1-x[1])
+            mutrates <- x[1+(1:nmuts)]
+            initfreqs <- x[1+nmuts+(1:nfreqs)]
+            initfreqs <- initfreqs/sum(initfreqs)
+            names(initfreqs) <- bases
+            list( 
+                    predicttreecounts( win, lwin, rwin, initcounts=rowSums(counts[[1]]), mutrates=list(mutrates,mutrates), selcoef=list(numeric(0),numeric(0)), 
+                            genmatrix=genmatrix, projmatrix=projmatrix, initfreqs=initfreqs, tlens=rev(branchlens) ),
+                    predicttreecounts( win, lwin, rwin, initcounts=rowSums(counts[[2]]), mutrates=list(mutrates,mutrates), selcoef=list(numeric(0),numeric(0)), 
[…]

### 2014-06-04 mus/quick-residuals.R

@@ -1,0 +1,124 @@
+#!/usr/bin/R
+
+source("../context-inference-fns.R")
+
+long.win <- list( lwin=2, rwin=2, win=3, winlen=7 )
+# long.win <- list( lwin=1, rwin=1, win=5, winlen=7 )
+
+bases <- c("A","T","C","G")
+mutpats <- c(
+        apply(combn(bases,2),2,list),  # single-base rates
+        apply(combn(bases,2)[2:1,],2,list),  # single-base rates
+        list( list( c("CG","TG"), c("CG","CA") ) )  # CpG rate
+    ) 
+selpats <- list()
+fixfn <- function (...) { 1 }
+boundary <- 'none'
+meanboundary <- 0
+
+long.win$genmatrix <- with(long.win, makegenmatrix( patlen=winlen, mutpats=mutpats, selpats=list(), boundary=boundary ) )
+long.win$projmatrix <- with(long.win, collapsepatmatrix( ipatterns=rownames(genmatrix), lwin=lwin, rwin=rwin ) )
+
+# all kmer -> kmer changes
+long.mutpats <- lapply(1:3, function (k) {
+    kmers <- getpatterns(k)
+    c( apply(combn(kmers,2),2,list),  
+            apply(combn(kmers,2)[2:1,],2,list)
+        )
+} )
+
+
+# gmfile <- paste(paste("genmatrices/genmatrix",commandArgs[1],'none',0,sep="-"),".RData",sep='') 
+# load(gmfile)
+# long.genmat <- genmatrix
+
+setwd("mm9rn5")
+
+getcounts <- function (winlen,win,genmatrix,projmatrix) {
+    infile <- paste(winlen,win,"counts",sep='.')
+    revfile <- paste("rev.",infile,sep='')
+    counts <- lapply( list(infile,revfile), function (ifile) {
+            count.table <- read.table(ifile,header=TRUE,stringsAsFactors=FALSE)
+            counts <- Matrix(0,nrow=nrow(genmatrix),ncol=ncol(projmatrix))
+            rownames(counts) <- rownames(genmatrix)
+            colnames(counts) <- colnames(projmatrix)
+            stopifnot( all( count.table$reference %in% rownames(genmatrix) ) & all(count.table$derived %in% colnames(projmatrix)) ) 
[…]

### 2014-06-05 context-inference-fns.R
poster

@@ -33,7 +33,7 @@ divergence <- function (counts, lwin) {
     return( sum(nchanges*counts)/(sum(counts)*patlen) )
 }
 
-countmuts <- function (counts, mutpats, lwin) {
+countmuts <- function (counts, mutpats, lwin, ...) {
     # given a table of matched tuples,
     # return counts of how many of these could be produced by each of mutpats
     #   and the total possible
@@ -45,7 +45,7 @@ countmuts <- function (counts, mutpats, lwin) {
     #    divergence(...)
     counts <- as.matrix(counts)
     win <- nchar(colnames(counts)[1])
-    stopifnot( len(win)>0 & win>0 )
+    stopifnot( length(win)>0 & win>0 )
     # restrict to same-length seqs
     xx <- substr(rownames(counts),lwin+1,lwin+win)
     yy <- colnames(counts)
@@ -55,14 +55,14 @@ countmuts <- function (counts, mutpats, lwin) {
             mutpat <- mutpats[[j]]
             mx <- sapply(mutpat, function (mp) {
                     patlen <- nchar(mp[1])
-                    sum( counts[ ( substr(xx,k,k+patlen-1) == mp[1] ), ( substr(yy,k,k+patlen-1) == mp[2] ) ] )
+                    sum( counts[ ( substr(xx,k,k+patlen-1) == mp[1] ), ( substr(yy,k,k+patlen-1) == mp[2] ) ], ... )
                 } )
-            sum.changed[j] <- sum.changed[j] + sum( mx )
+            sum.changed[j] <- sum.changed[j] + sum( mx, ... )
             px <- sapply(mutpat, function (mp) {
                     patlen <- nchar(mp[1])
-                    sum( counts[ ( substr(xx,k,k+patlen-1) == mp[1] ), ] )
+                    sum( counts[ ( substr(xx,k,k+patlen-1) == mp[1] ), ], ... )
                 } )
-            possible[j] <- possible[j] + sum( px )
+            possible[j] <- possible[j] + sum( px, ... )
         }
     }
     x <- rbind(sum.changed,possible)

### 2014-06-05 mus/flex-genmat.R

@@ -1,0 +1,73 @@
+#!/usr/bin/Rscript --vanilla
+require(optparse)
+require(jsonlite)
+
+usage <- '\
+Pre-compute a generator matrix and associated structures,  \
+reading list of patterns from a configuration file, \
+formatted in json as follows (see singlebase.mut): \
+ \
+{  \
+    "bases" : [ "A", "C", "G", "T" ], \
+    "mutpats" : [ \
+    [ [ "A", "T" ], [ "T", "A" ] ], \
+    [ [ "C", "G" ], [ "G", "C" ] ], \
+    [ [ "A", "C" ], [ "T", "G" ] ], \
+    [ [ "A", "G" ], [ "T", "C" ] ], \
+    [ [ "C", "A" ], [ "G", "T" ] ], \
+    [ [ "C", "T" ], [ "G", "A" ] ] \
+    ], \
+    "selpats" : [ \
+    ] \
+} \
+ \
+Entries in the same sub-list have the same associated parameter (reverse-complement, above). \
+'
+
+option_list <- list(
+        make_option( c("-c","--configfile"), type="character", default="-", help="Input file of mutation and selection motifs." ),
+        make_option( c("-s","--outfile"), type="character", default="", help="Save resulting matrix in this file.  [default: genmatrices/genmatrix-(configfile).RData]" ),
+        make_option( c("-w","--winlen"), type="integer", default=2, help="Size of matching window. [default \"%default\"]" ),
+        make_option( c("-b","--boundary"), type="character", default="none", help="Boundary conditions. [default \"%default\"]" ),
+        make_option( c("-m","--meanboundary"), type="integer", default=0, help="Average over this many neighboring bases. [default \"%default\"]" ),
+        make_option( c("-k","--patlen"), type="integer", default=1, help="Include mutation rates for all tuples of this length. [default \"%default\"]" ),
+        make_option( c("-o","--logfile"), type="character", default="", help="Direct logging output to this file. [default appends .Rout]" )
+    )
+opt <- parse_args(OptionParser(option_list=option_list,description=usage))
+if (is.null(opt$configfile)) { stop("Need a config file.") }
+options(error=traceback)
+
+config <- fromJSON(configfile)
+attach(c(opt,config))
+
+if (outfile=="") { 
[…]

### 2014-07-10 writeups/context-dependence.tex
Clearing trailing whitespace.
Goodbye fringe! Closes gh-3.
Made it to the is-X(0)-random note.

@@
 consisting of a lattice of sites,
-each of which taking one of a finite collection of possible states,
+each site taking one of a finite collection of possible states,
 and whose stochastic, temporal evolution is Markov and governed by a set of local rules.
@@
 transition probabilities are in principle simply expressable as a matrix exponential,
-but this is impractical as the size of the matrix is equal to the number of possible configurations.
+but this is impractical because the size of the matrix is equal to the number of possible configurations.
+\todo{I suggest pulling this sentence together with your "In this paper" sentence to form a separate paragraph.}
 As an alternative, we show that the conditional likelihood of local pattern count statistics
@@
 so do not consider further the many constraints on protein-coding sequence \citep[reviewed in][]{thorne2007protein}.)
+
 However, phylogenetic and population genetic methods usually ignore such dependencies
@@
 while \citet{siepel2004phylogenetic} and later \citep{baele2010modelling} computed likelihoods by assuming a model that is Markov along the genome;
-however, all these methods quite computationally intensive.
+however, all these methods are quite computationally intensive.
 
-Another canonical example is the Ising model of statistical physics
+Another canonical example of context-dependent transition is the Ising model of statistical physics
 with time evolution given by Glauber dynamics \citep{glauber1963timedependent},
@@
 
-Each example above describes a 1-dimensional grid of sites, each of which can take one of a finite set of states,
+Each example above describes \todo{Has a formulation as? Glauber can be greater than 1 D} a 1-dimensional grid of sites, each of which can take one of a finite set of states,
 and that switch randomly between states according to a local set of rules.
@@
 and $X_i(t)$ the state of site $i$ at time $t$,
-we assume that $\{X(t)\}_{t \ge 0}$ is a Markov process on $\calS^n$
+we assume that $\{X(t)\}_{t \ge 0}$ is a Markov process on sequences of $n$ states $\calS^n$
 for which the probability a given site changes state in a small amount of time
@@
 
-To make this notion of local dependence formal,
+To formalize this notion of local dependence,
 first define a \textit{pattern} to mean a contiguous sequence of states;
-let $|u|$ denote the length of the pattern $u$;
-and let $x_i^{(h)} = (x_i, x_{i+1}, \ldots, x_{i+h-1})$ denote the subsequence of $x$ of length $h$ beginning at location $i$.
-The dynamics are determined by the set of allowed transitions and associated rates:
+let $|u|$ denote the length of the pattern $u$.
+Given $x \in \calS$,
+let $x_i^{(h)} = (x_i, x_{i+1}, \ldots, x_{i+h-1})$ denote the pattern of $x$ of length $h$ beginning at location $i$.
+Recall we are focusing on finite sequences, and by default, we do not allow patterns to ``hang off'' the end of the sequence.
+This is natural in many situations, but in others we might want to other boundary conditions,
+modeling extant, but unobserved, neighbors.
+We will mention this when necessary.
+
+The dynamics of our stochastic process are determined by the set of allowed transitions and associated rates:
 given two patterns $u$ and $v$ of common length $h$,
@@
 The set of all transition triples, $\calT$, determines the time evolution of the process.
-
-A final note: recall we are focusing on finite sequences.
-By default, we do not allow patterns to ``hang off'' the end of the sequence.
-This is natural in many situations, but in others we might want to other boundary conditions,
-modeling extant, but unobserved, neighbors.
-We will mention this when necessary.
+\todo{I'd suggest being more formal and defining a name for the class of processes you describe, and saying what it consists of.}
 
@@
   and if it is, moves there.
-  This process therefore has only one transition triple:
+  This process therefore has only one transition triple \todo{Not sure if reviewers will like having a different notation for triples so soon after you defined the parens notation}:
 
@@
   A sequence of genome can be written using A, C, G, and T;
-  in the most general model of single-site mutation, each of the 12 possible transitions occurs at its own rate.
+  in the most general model of independent mutation across sites, each of the 12 possible transitions occurs at its own rate.\todo{Given the importance of the independent model, I think it deserves a more complete presentation.}
   Furthermore, it is a well-known observation that in many species,
-  adjacent, methylated CG dinucleotides (``CpG sites'') have a much higher mutation rate to TG and CA
-  than either single nucleotide change.
+  adjacent, methylated CG dinucleotides (``CpG sites'') \todo{not all CpG sites are methylated} have a much higher mutation rate to TG and CA
+  than either single nucleotide change \todo{Those *are* single nucleotide changes}.
   Embellishing the single-nucleotide model with this additional rate results in the model defined by
@@
   The associated stationary distribution on configurations is proportional to $\exp(-H(x))$.
-  One way to add temporal dynamics that preserve the distribution
-  is to say that each site, independently at rate $\lambda$,
+  The following ``Glauber dynamics'' preserve the stationary distribution:
+  each site, independently at rate $\lambda$,
   forgets its spin,
   and reconfigures to a state chosen with probability proportional to the stationary probability of the resulting configuration.
-  This is known as ``Glauber dynamics'', and ignoring transitions that don't change the state,
-  we get that if patterns $u$ and $v$ differ at only one site, then $u$ changes to $v$ at rate $1/(1+\exp(H(v)-H(u)))$, so
+  Ignoring transitions that don't change the state,
+  if patterns $u$ and $v$ differ at only one site then $u$ changes to $v$ at rate $1/(1+\exp(H(v)-H(u)))$ \todo{Hum, this isn't immediately obvious to me-- quick explanation please?
+Also, shouldn't there be a $\lambda$ up top?}, so
 
@@
 Recall that $x_i^{(h)} = (x_i, x_{i+1}, \ldots, x_{i+h-1})$ denotes the subsequence of length $h$ beginning at location $i$.
-For each $1\le i \le n$, pattern length $h \ge 0$, and patterns $u,v \in \calS^h$ define the relation
+For each $1\le i \le n$, pattern length $h > 0$, and patterns $u,v \in \calS^h$ define the relation
 \[
@@
   x_{i+k} = u_k \quad &\text{for } 0 \le k < h \\
-  y_{i+k} = v_k \quad &\text{for } 0 \le k < h, and \\
+  y_{i+k} = v_k \quad &\text{for } 0 \le k < h,\ \text{and} \\
   x_k = y_k \quad &\text{for } k\ge i+h ,
@@
 As noted above, there may be more than one way to mutate a sequence $x$ to get $y$:
-for each position $i$, these are those transitions in $\calT$ that can be applied at $i$
-to change $x$ into $y$, denoted  $J(i,x,y) = \{ (\mu^j,u^j,v^j) \in \calT \st x \xrightarrow{i,u^j,v^j} y \}$.
+for each position $i$, let $J(i,x,y)$ be those transitions in $\calT$ that can be applied at $i$
+to change $x$ into $y$, i.e.
+\[
+J(i,x,y) = \{ (\mu^j,u^j,v^j) \in \calT \st x \xrightarrow{i,u^j,v^j} y \}.
+\]
 The rate $G(n)_{x,y}$ is then the sum of all matching transition rates,
@@
 and if there are no triples $(\mu,u,v)$ with $x \xrightarrow{i,u,v} y$ for some $i$, then $G(n)_{x,y}=0$.
-This is just to say that the transition rate from $x$ to $y$ is the sum of the mutation rates of all transition triples
+This formalizes the statement that the transition rate from $x$ to $y$ is the sum of the mutation rates of all transition triples
 that would change $x$ into $y$.
@@
 using a variant of the ``jump chain representation''.
-Briefly, this works by first sampling a homogeneous Poisson process of times and locations of possible changes,
+Briefly, this \todo{This is called the Gillespie algorithm, right?} works by first sampling a homogeneous Poisson process of times and locations of possible changes,
 at an appropriate rate $\mu_*$ per site,
@@
 with probability proportional to the corresponding rate, i.e.\
+\todo{A little confused here. Previously, the $j$ in $u^j$ was used to index the possible patterns, right? Now it's indexing possible changes at a site.}
 \[
@@
 In the CpG model it is not immediately clear what information is retained.
-\peter{What is identifiable?  The entire model, up to a time change?  I go back and forth on this...}
-\todo[inline]{Not sure what I was smoking when I said that.
+\todo[inline]{
 I think that the stationary distribution of the single-nucleotide model should be identifiable for sure.
-Then, I would guess the two CpG rates are identifiable modulo a rate factor that compares the single-nucleotide mutation rate to the CpG mutation rate.}
+Then, I would guess the two CpG rates are identifiable modulo a rate factor that compares the single-nucleotide mutation rate to the CpG mutation rate.
+}
 
@@
 i.e.\ that the probability a given subsequence matching $x$ is later observed to have its center matching $y$
-is close to $p_{\ell,m}(x,y)$, regardless of context:
+is close to $p_{\ell,m}(x,y)$, regardless of context wider than $2k+l$:
 \begin{align} \label{eqn:window_approx}
-  p_{\ell,m}(x,y) \approx \P\{ X_i^{(m)}(t) = y \mid X_i^{(2\ell+m)}(0) = x \}.
+  p_{\ell,m}(x,y) \approx \P\{ X_{i+\ell}^{(m)}(t) = y \mid X_i^{(2\ell+m)}(0) = x \}.
 \end{align}
@@
 
-Since we cannot compute reasonably the full likelihood for the data,
+Since we cannot compute reasonably the full likelihood for the data \todo{Well, can we prove that it's not possible? Perhaps say that we can't using straightforward exponentiation?},
 we will need to choose appropriate summary statistics
@@
 If $i_1 < \ldots < i_{n(x)}$ are a set of nonoverlapping sites where the initial sequence matches $x$
--- i.e.\ $X_{i_k}^{(2\ell+m)}(0)=x$ and $i_{k+1} > i_k + 2\ell+m$ for all $k$ --
+-- i.e.\ $X_{i_k}^{(2\ell+m)}(0)=x$ and $i_{k+1} > i_k + 2\ell+m$ for all $k$ -- \todo{En dash close to math confuses me! This sentence is a doozy.}
 then the counts $N(x,y)=\#\{ i_k \st X_{i_k+\ell}^{(m)}(t)=y\}$ (the number of these sites that match the smaller pattern $y$ at time $t$)
@@
 In fact, we do not even look at a fraction $2\ell/(2\ell+m)$ of the second sequence, $X(t)$,
-that lies in the fringes of each window.
+that lies in the outside edges of each window.
 As before, let $N(x)$ be the number of occurrences of a pattern $x$ in $X(0)$,
@@
 so that e.g.\ $N(x) = \#\{ 1 \le i \le n - 2\ell - m + 1 \st X_i^{(m+2\ell)}(0)=x\}$.
-
-\peter{Ooops, yes.  Previously, I was semi-inconsistently thinking of the initial sequence $X(0)$ as random as well. ,
-    which is why I was talking about Poisson instead of Multinomial counts.}
 

### 2014-07-10 writeups/context-dependence.tex
A few more changes to make things clearer IMHO.

@@
 This is the probability that if we begin with a sequence $x$ of length $2\ell+m$,
-then after time $t$ we will see at the middle $m$ positions the pattern $y$,
+then after time $t$ we will see the pattern $y$ at the middle $m$ positions,
 ignoring what is present in the leftmost and rightmost $\ell$ positions.
@@
 Then we'll have the correct likelihood function for (this subset of) data.
+
 If changes are rare, then one way to get a set of patterns with this property is to only choose ones such that going from $x$ to $y$ entails a change;
@@
 and suppose that they differ at $1 \le i_1, \ldots, i_d \le m$.
+Let $\bar i = (1/d) \sum_{j=1}^d i_j$ be the mean position of the changes,
+and take only patterns with $(m-1)/2 < \bar i \le (m+1)/2$.
 For example, in the example above with $m=3$ and $\ell=2$,
@@
 $x$ and $y$ differ only at (relative) position $i_1=2$.
-Let $\bar i = (1/d) \sum_{j=1}^d i_j$ be the mean position of the changes,
-and take only patterns with $(m-1)/2 < \bar i \le (m+1)/2$.
-As another example,
+
+Another way to achieve the same goal of having all overlapping patterns require additional changes is to require $i_1 = \bar i = 1$.
+No two such patterns can be shifted versions of each other
+without adding more changed sites,
+since shifting by $k$ without adding changed sites just adds $k$ to $\bar i$.
+For example, we might take
 \begin{center}
@@
 \end{center}
-has $i_1 = \bar i = 1$.
-No two such patterns can be shifted versions of each other
-without adding more changed sites,
-since shifting by $k$ without adding changed sites just adds $k$ to $\bar i$.
 

### 2014-07-10 writeups/context-dependence.tex
Some questions for @petrelharp.
This shows the depth of my ignorance. :)

I'd like to better understand how you are approaching likelihood
calculation that has some per-site component but also a nontrivial
motif component. For example, in the CpG model, mutations can arise
because of the "usual" per-nucleotide mutation, or arise because of the
actual CpG motif mutation as you describe. Do we think of this as being
a model with m=1 and l=1? If that's the case then we have many
overlapping motifs, and so then do we only calculate the likelihood for
part of the data via nonoverlapping windows? Or

@@
 
+\todo[inline]{In this section I believe that we are talking about the full likelihood, rather than the likelihood of just non-overlapping windows of the data.
+I'm also a little confused about "choosing" a subset of patterns.
+These pattern choices are determining our model?}
+
 The covariance calculations show that
@@
 If we use only a subset of patterns $\{(x_k,y_k)\}$ chosen so that $\E[M_1(x,y) M_{k+1}(x,y)]$ is small (or zero),
+\todo{M not defined, but I'm guessing that this is effectively saying that the probability of having two different patterns apply to the same sequences is small?}
 the entire collection $\{N(x_k,y_k)\}$ will be close to independent Poisson.

### 2014-07-31 bcells/point-estimates.R
hutch talk done

@@ -70,6 +70,7 @@ initcounts <- rowSums(counts)
 # ad-hoc estimate
 adhoc <- countmuts(counts=counts,mutpats=mutpats,lwin=lwin)
 adhoc <- adhoc[1,]/adhoc[2,]
+stopifnot( all( is.finite( adhoc ) )
 
 nmuts <- length(genmatrix@mutpats)
 nsel <- length(genmatrix@selpats)

### 2014-08-07 writeups/context-dependence.tex
hutch talk done
A few comments and fixes.

@@
       $x$  &  $y$  &  $m_{xy}$ & when $x \neq y \in \{\nA,\nC,\nG,\nT\}$  \\
-      CG   &  TG   &  $\gamma$ & \\
-      CG   &  CA   &  $\gamma$ &
+      \nC\nG   &  \nT\nG   &  $\gamma$ & \\
+      \nC\nG   &  \nC\nA   &  $\gamma$ &
     \end{tabular} .
@@
 also used for simulation,
-using a variant of the ``jump chain representation''.
-Briefly, this \todo{This is called the Gillespie algorithm, right?} works by first sampling a homogeneous Poisson process of times and locations of possible changes,
+using a variant of the ``jump chain representation'', also known as the ``Gillespie algorithm.''
+.
+Briefly, this  works by first sampling a homogeneous Poisson process of times and locations of possible changes,
 at an appropriate rate $\mu_*$ per site,
@@
 \subsection{A likelihood function}
-
-\todo[inline]{In this section I believe that we are talking about the full likelihood, rather than the likelihood of just non-overlapping windows of the data.
-I'm also a little confused about "choosing" a subset of patterns.
-These pattern choices are determining our model?}
 

### 2014-08-08 writeups/context-dependence.tex
More comments!
@petrelharp could you please look at the YYY comments, and decide
whether you like them or not? I think that heading in the direction
of more consistency will keep me sane.

@@
 then
-\[
+\begin{align} \label{eqn:Tmer_trans}
     p_{m,\ell}(t,x,y) = \sum_{z \in \calS^{m+2\ell}} \left( e^{t G(m+2\ell)} \right)_{xz} U_{zy} ,
-\]
+\end{align}
 so in fact we don't need the entire matrix $e^{t G(m+2\ell)}$,

### 2014-08-09 writeups/context-dependence.tex
EM finally gets an inkling of how things work.
More notes for @petrelharp marked with YYYs.

@@
 using all $N(x,y)$ as data.
-Maximum likelihood then gets an asymptotically unbiased estimate of the parameters.
+Specifically, this composite likelihood is
+\begin{align} \label{eqn:comp_like}
+  \prod_{x,y} F_{(x,y)}^{N(x,y)}
+\end{align}
+where $F$ is the $|\calS|^{m+2\ell} \times |\calS|^m$ matrix whose $(x,y)^\text{th}$ entry is $p_{m,\ell}(t,x,y)$ as described in the "Computation" section.
+Maximum likelihood then gets an asymptotically unbiased estimate of the parameters. (cite)
 However, since they are not independent, to get confidence intervals, we need to either simulate, or do something different.
@@
 
-For inference, we need to compute the $|\calS|^{m+2\ell} \times |\calS|^m$ matrix whose $(x,y)^\text{th}$ entry is $p_{m,\ell}(t,x,y)$,
+For inference, we need to compute the $|\calS|^{m+2\ell} \times |\calS|^m$ matrix $F$ whose $(x,y)^\text{th}$ entry is $p_{m,\ell}(t,x,y)$,
 for various $t$.

### 2014-09-02 writeups/context-dependence.tex

@@
 Such a rule is defined by its ``transition triple'' $(\mu,u,v)$.
-The set of all transition triples, $\calT$, determines the time evolution of the process.
-\todo{I'd suggest being more formal and defining a name for the class of processes you describe, and saying what it consists of.}
+The set of all transition triples, $\calT$, determines the time evolution of this process.
 
@@
 \paragraph{The generator matrix}
-For clarity, we describe how the set of transition rates determines the transition rate matrix,
+For clarity, we describe how the set of transition rates on patterns determines the transition rate matrix for complete sequences, that is,
 the $|\calS|^n \times |\calS|^n$ matrix $G(n)$ whose $(x,y)^\text{th}$ entry gives the instantaneous rate
-with which the process in state $x$ jumps to state $y$.
+with which the process in state $x \in \calS^n$ jumps to state $y \in \calS^n$.
 Recall that $x_i^{(h)} = (x_i, x_{i+1}, \ldots, x_{i+h-1})$ denotes the subsequence of length $h$ beginning at location $i$.

### 2014-09-04 generic.Rmd
Start of an Rmd report.

# Results from fitting procedure

## Invocation

```{r}
cat(model@invocation)
```

## fitted parameter values

```{r}
model@results[["par"]]
```

## window sizes and amount of data used to fit

```{r}
counts <- model@counts@counts
head(counts) # for some reason the code doesn't work if we don't print out some of the counts
cat("left window size:", model@counts@leftwin,
    "\ntotal window size:", nchar(rownames(counts)[[1]]))
cat("total counts:", sum(counts))
```

## result of fitting operation

```{r}
cat("convergence: ", model@results$convergence,
    "\nmessage: ", model@results$message)
```

## residuals for sufficiently short T-mers you can look at all of them (2-2? 3-1?)
## the top and bottom residuals for longer T-mers (9-5? 9-9?)
## if present, MCMC traces
## ... and posterior marginal distributions.

### 2014-09-05 generic.Rmd
Factoring out the residuals computation.
@petrelharp -- this works great for the command line RScript, but when I try to
have it do a thing in the RMarkdown world, I get this error:

```
Quitting from lines 37-38 (generic.Rmd)
Error in ifelse(ds == 0, 1, Ne * expm1(-2 * ds)/expm1(-2 * Ne * ds)) :
  argument "Ne" is missing, with no default
Calls: render ... predictcounts.context -> predictcounts -> update -> <Anonymous> -> ifelse
```

please help!

@@
 
-## Invocation
-
 ```{r}
+source("context-inference-fns.R")
 cat(model@invocation)
@@
 cat("left window size:", model@counts@leftwin,
-    "\ntotal window size:", nchar(rownames(counts)[[1]]))
-cat("total counts:", sum(counts))
+    "\ncomplete window size:", nchar(rownames(counts)[[1]]),
+    "\ntotal counts:", sum(counts))
 ```
@@
 ## residuals for sufficiently short T-mers you can look at all of them (2-2? 3-1?)
+
+```{r}
+computeresids(model)
+```
+
 ## the top and bottom residuals for longer T-mers (9-5? 9-9?)

### 2014-09-05 compute-resids.R

@@ -28,62 +28,17 @@ options(error = print.and.dump)
 stopifnot(file.exists(opt$infile))
 load(opt$infile) # provides 'model'
 
-if (is.null(opt$longwin)){ opt$longwin <- longwin(model) }
-if (is.null(opt$shortwin)){ opt$shortwin <- shortwin(model) }
-if (is.null(opt$leftwin)){ opt$leftwin <- leftwin(model) }
-
 # set outfile name
-if (is.null(opt$outfile)) { 
-    opt$outfile <- paste( opt$basedir, "/", gsub("\\.[^.]*","",basename(opt$infile) ), "-resids", ".", opt$longwin, ".", opt$shortwin, ".l", opt$leftwin, ".tsv", sep='' ) 
+if (is.null(opt$outfile)) {
+    opt$outfile <- paste( opt$basedir, "/", gsub("\\.[^.]*","",basename(opt$infile) ), "-resids", ".", opt$longwin, ".", opt$shortwin, ".l", opt$leftwin, ".tsv", sep='' )
     print(opt$outfile)
 }
 
-# load generator matrix, if needed
-if (!is.null(opt$gmfile)) {
-    stopifnot(file.exists(opt$gmfile))
-    load(opt$gmfile)  # provides 'genmatrix'
-} else {
-    genmatrix <- model@genmatrix
-}
-
-# get counts
-if (is.null(opt$countfile) && ( opt$longwin > longwin(model) || opt$shortwin > shortwin(model) ) ) {
-    stop("If window lengths are longer than fitted model, then need to supply counts.")
-} else if (!is.null(opt$countfile)) {
-    counts <- read.counts(opt$countfile,opt$leftwin)
-    if ( ( opt$longwin > longwin(counts) || opt$shortwin > shortwin(counts) ) ) {
-        stop("Supplied counts use a window that is too short.")
-    }
-} else {
-    counts <- model@counts
-}
-if ( (opt$longwin < longwin(counts)) || (opt$shortwin < shortwin(counts)) ) {
-    counts <- projectcounts( counts, opt$leftwin, opt$shortwin, opt$longwin-opt$leftwin-opt$shortwin )
-}
-stopifnot( all( rownames(counts) == rownames(genmatrix) ) )
-
-expected <- fitted( model, longwin=opt$longwin, shortwin=opt$shortwin, leftwin=opt$leftwin, initcounts=rowSums(counts), genmatrix=genmatrix )
-
-if (opt$pretty) {
[…]

### 2014-09-08 writeups/context-dependence.tex
computing transition matrix across a general tree does not error (but not right yet? see tree-test.R)

@@
 
+\subsection*{Algorithm}
+
+Fix one tip of the tree, $a$, to be the taxa where we count ``long'' sequences, let $\rho$ be the root, and let $b_1, \ldots, b_n$ be the remaining tips,
+where we observe ``short'' sequences.
+Let $v_0=\rho, v_1, \ldots, v_\ell = a$ be the path from the root down to $a$,
+let $u_1, u_2, \ldots, u_m$ be the remaining internal nodes,
+and for each $v_i$ let $u(v_i)$ be the offspring of $v_i$ that is not one of the other $v$.
+For each node in the tree,
+we will compute the matrix of probabilities that a ``long'' sequence observed at that node
+is matched by the ``short'' sequences seen on all the tips of the tree
+further away from $a$ (i.e.\ below that node if it was rooted at $a$).
+For a node $w$ this will be stored as a $|\calS|^{m+2\ell} \times |\calS|^{cm}$ matrix, for some $c$,
+and denoted $M(w)$.
+
+\begin{enumerate}
+
+  \item Set $M(b_i) = P$ for each $b_i$.
+
+  \item While there are $u_j$ whose children $w_1$ and $w_2$ both have matrices computed,
+    \begin{enumerate}
+      \item let
+        \begin{align}
+          M(u_j)_{x,(y,z)} = \left( e^{t(w_1) G} M(w_1) \right)_{x,y} \left( e^{t(w_2) G} M(w_2)_{x,z} \right)_{x,z} ,
+        \end{align}
+        where $t(w_1)$ and $t(w_2)$ are the lengths of the respective branches.
+    \end{enumerate}
+
+  \item Let 
+    \begin{align}
+      M(\rho)_{x,y} = \pi(x) \left( e^{t(w(\rho)) G} M(w(\rho)) \right)_{x,y}
+    \end{align}
+
+  \item For each $1 \le k < \ell$,
+    \begin{enumerate}
+      \item let
+        \begin{align}
+          M(v_k)_{x,(y,z)} = \left( e^{t(v_k) G} M(v_{k-1}) \right)_{x,y} \left( e^{t(u(v_k)) G} M(u(v_{k-1})) \right)_{x,z}  .
+        \end{align}
+    \end{enumerate}
+
+  \item Finally, let
+    \begin{align}
+      M(a)_{x,y} = \left( e^{t(a) G} M(v_{\ell-1}) \right)_{x,y}  .
+    \end{align}
+
+\end{enumerate}
+
 %%%%%%

### 2014-09-08 testing-code/unit-test-treeness.sh

@@ -2,7 +2,7 @@
 set -eu
 set -o pipefail
 
-Rscript ../sim-seq.R -c big-tree-model.json -s 100 -o TEMP-tree-test.RData
+Rscript ../sim-seq.R -c fwds-rev-tree-model.json -s 100 -o TEMP-tree-test.RData
 Rscript -e 'load("TEMP-tree-test.RData");suppressMessages({require(Biostrings);require(IRanges)})' \
     -e 'x <- lapply(simseqs,function (z) strsplit(as.character(z$finalseq),"")[[1]])' \
     -e 'stopifnot(unique(x$sp1)=="O")' \

### 2014-09-09 writeups/context-dependence.tex
tree inference working IF columns are in the right order

@@
 
-  \item Compute $M_2(v,z) = \sum_{v \in \calS^{m+2\ell}}  p_{m+2\ell}(t_{AB},u,v) \pi(u) M_1(u,z)$ (still a $|\calS|^{m+2\ell} \times |\calS|^m$ matrix)
+  \item Compute $M_2(v,z) = \sum_{u \in \calS^{m+2\ell}}  p_{m+2\ell}(t_{AB},u,v) \pi(u) M_1(u,z)$ (still a $|\calS|^{m+2\ell} \times |\calS|^m$ matrix)
 
@@
 For each node in the tree,
-we will compute the matrix of probabilities that a ``long'' sequence observed at that node
-is matched by the ``short'' sequences seen on all the tips of the tree
-further away from $a$ (i.e.\ below that node if it was rooted at $a$).
+we will compute a matrix of probabilities:
+for nodes not on the path from $a$ to the root,
+we compute the chance that that each ``long'' sequence observed at that node
+is matched by each possible combination of ``short'' sequences on the tips of the tree below it.
+For nodes on the path from $a$ to the root, we compute the probability of observing
+each ``long'' sequence at that node, along with the combinations of ``short'' sequences
+on all tips further away from $a$ (i.e.\ below that node if it was rooted at $a$).
 For a node $w$ this will be stored as a $|\calS|^{m+2\ell} \times |\calS|^{cm}$ matrix, for some $c$,
 and denoted $M(w)$.
+Rows of $M(u_j)$ sum to 1,
+while each entire matrix $M(v_i)$ sums to 1;
+the difference being that $M(v_i)$ sums over the prior at the root,
+and $M(u_j)$ takes the value at $u_j$ as given.
 
@@
         \begin{align}
-          M(v_k)_{x,(y,z)} = \left( e^{t(v_k) G} M(v_{k-1}) \right)_{x,y} \left( e^{t(u(v_k)) G} M(u(v_{k-1})) \right)_{x,z}  .
+          M(v_k)_{x,(y,z)} = \left( e^{t(v_k) G} M(v_{k-1}) \right)_{y,x} \left( e^{t(u(v_k)) G} M(u(v_{k-1})) \right)_{x,z}  .
         \end{align}

### 2014-09-09 input-output.R

@@ -127,7 +127,7 @@ parse.fixfn <- function (fixfn,fixfn.params) {
 
 config.dereference <- function (config, x) {
     # follow pointers in config: if config[[x]] is a string, it refers to another entry.
-    sapply(x, function (xx) { while (is.character(config[[xx]]) & (length(config[[xx]])==1) ) { xx <- config[[xx]] }; xx } )
+    sapply(x, function (xx) { n <- 1; while (n < 20 && is.character(config[[xx]]) & (length(config[[xx]])==1) ) { xx <- config[[xx]]; n<-n+1 }; xx } )
 }
 
 parse.models <- function (config,do.fixfns=TRUE) {

### 2014-09-09 json-tree-cpg/workflow.sh

@@ -17,13 +17,12 @@ Rscript ../count-seq.R -i $BASEDIR/sim-tree-cpg.RData -c sp2 -w $LONGWIN -s $SHO
 
 # precompute generator matrices:
 #   width-3
-MODEL=tree-cpg-model.json
 GENMAT=genmatrices/genmatrix-${LONGWIN}-cpg.RData
-Rscript ../make-genmat.R -c $MODEL -w ${LONGWIN} -o ${GENMAT} -n sp1
+Rscript ../make-genmat.R -c tree-cpg-model.json -w ${LONGWIN} -o ${GENMAT} -n sp1
 
 # fit a model
-Rscript ../fit-model.R -i $BASEDIR/sim-tree-cpg-3-sp1-1-sp2-l1.counts -l ${LEFTWIN} -m $GENMAT -j 001
-Rscript ../fit-model.R -i $BASEDIR/sim-tree-cpg-3-sp2-1-sp1-l1.counts -l ${LEFTWIN} -m $GENMAT -j 001
+Rscript ../fit-model.R -i $BASEDIR/sim-tree-cpg-3-sp1-1-sp2-l1.counts -c tree-cpg-model.json -w ${LONGWIN} -l ${LEFTWIN} -j 001
+Rscript ../fit-model.R -i $BASEDIR/sim-tree-cpg-3-sp2-1-sp1-l1.counts -c tree-cpg-model.json -w ${LONGWIN} -l ${LEFTWIN} -j 001
 
 # compute residuals
 Rscript ../compute-resids.R -i $BASE-123456-genmatrix-${LONGWIN}-cpg-54321.RData -w 3 -s 1 -l 1 -m ${GENMAT}

### 2014-09-09 testing-code/tree-tests.R

@@ -2,16 +2,45 @@
 
 source("../context-inference-fns.R")
 
-config <- treeify.config( read.config("big-tree-model.json") )
+
+###
+# Little tree
+
+
+config <- treeify.config( read.config("little-tree-model.json") )
 
 genmatrices <- lapply( selfname(c("forward","reverse")), function (x) {
         makegenmatrix( mutpats=config[[x]]$mutpats, bases=config$bases, mutrates=config[[x]]$mutrates, patlen=2, fixfn=null.fixfn )
     } )
-genmatrices <- genmatrices[ unlist(config[nodenames(config$tree)]) ]
+models <- unlist( config[ nodenames(config$tree) ] )
 
 projmatrix <- collapsepatmatrix( ipatterns=rownames(genmatrices[[1]]), leftwin=0, fpatterns=getpatterns(1,config$bases) )
 root.distrn <- runif( nrow(projmatrix) )
 root.distrn <- root.distrn / sum(root.distrn)
 
-transmat <- peel.transmat( config$tree, rowtaxon="sp1", coltaxa=c("sp2","sp3","sp4"), genmatrices=genmatrices, projmatrix=projmatrix, root.distrn=root.distrn, return.list=TRUE )
+transmat <- peel.transmat( config$tree, rowtaxon="sp1", coltaxa=c("sp2"), models=models, genmatrices=genmatrices, projmatrix=projmatrix, root.distrn=root.distrn, return.list=TRUE )
+
+# check if probabilities sum to 1:
+stopifnot( all( ( abs( sapply( transmat, sum ) - 1 ) < sqrt(.Machine$double.eps) ) | ( sapply( transmat, function (x) all( abs(rowSums(x)-1) < sqrt(.Machine$double.eps) ) ) ) ) )
+
+
+###
+# Bigger tree
+
+config <- treeify.config( read.config("big-tree-model.json") )
+
+genmatrices <- lapply( selfname(c("forward","reverse")), function (x) {
+        makegenmatrix( mutpats=config[[x]]$mutpats, bases=config$bases, mutrates=config[[x]]$mutrates, patlen=2, fixfn=null.fixfn )
+    } )
+models <- unlist( config[ nodenames(config$tree) ] )
+
+projmatrix <- collapsepatmatrix( ipatterns=rownames(genmatrices[[1]]), leftwin=0, fpatterns=getpatterns(1,config$bases) )
[…]

### 2014-09-23 generic.Rmd
RMarkdown now has residual tables.

@@
 ```{r}
+library(pander)
+panderOptions("table.split.table", Inf)  # Let tables become arbitrarily wide rather than wrapping.
 source("context-inference-fns.R")
@@
 
-```{r}
-computeresids(model)
+```{r, results="asis"}
+resids = computeresids(model)
+pandoc.table(resids, style="rmarkdown")
 ```

### 2014-09-23 compute-resids.R

@@ -38,11 +38,12 @@ if (is.null(opt$outfile)) {
     print(opt$outfile)
 }
 
-computeresids (model,
+residframe = computeresids (model,
     pretty            = opt$pretty,
-    outfile           = opt$outfile,
     in_longwin        = opt$longwin,
     in_shortwin       = opt$shortwin,
     in_leftwin        = opt$leftwin,
     in_countfile      = opt$countfile,
     in_genmatrixfile  = opt$genmatrixfile)
+
+write.table(file=opt$outfile, x=residframe, sep='\t', quote=FALSE, row.names=TRUE )

### 2014-09-30 generic.Rmd
Limited size tables.

--- 2014-09-04 generic.Rmd
@@
 # Results from fitting procedure
-
-## Invocation
-
 ```{r}
+library(pander)
+library(ggplot2)
+panderOptions("table.split.table", Inf)  # Let tables become arbitrarily wide rather than wrapping.
+source("context-inference-fns.R")
+top_n_bottom <- function(df, k) { # Take the top and bottom k rows
+    n = nrow(df)
+    if(2*k < n) {
+        df[c(1:k,(n-(k-1)):n), ]
+    }
+    else {
+        df
+    }
+}
+counts_table <- function(in_longwin, in_shortwin, half_n_show) {
+    if ( in_longwin > longwin(model@counts) || in_shortwin > shortwin(model@counts) ) {
+        cat("Window size of counts too small.")
+    }
+    else {
+        resids = computeresids(model,
+                in_longwin=in_longwin,
+                in_shortwin=in_shortwin,
+                in_leftwin=in_shortwin,
+                counts=model@counts)
+        pandoc.table(top_n_bottom(resids, half_n_show), style="rmarkdown")
+    }
+}
 cat(model@invocation)
@@
 
-## fitted parameter values
-
-```{r}
-model@results[["par"]]
+## Fitted parameter values
+```{r, results="asis"}
+par_df = data.frame(value = model@results[["par"]], row.names = names(model@results[["par"]]))
+pandoc.table(par_df, style="rmarkdown")
+par_df$names = row.names(par_df)
+par_df = par_df[grep("->", par_df$names),]
+ggplot(par_df, aes(x = names, y = value)) +
+    geom_bar(stat = "identity", fill="white", color="gray") +
+    theme(axis.text.x = element_text(angle = 90))
 ```
 
-## window sizes and amount of data used to fit
-
+## Window sizes and amount of data used to fit
 ```{r}
@@
 cat("left window size:", model@counts@leftwin,
-    "\ntotal window size:", nchar(rownames(counts)[[1]]))
-cat("total counts:", sum(counts))
+    "\ncomplete window size:", nchar(rownames(counts)[[1]]),
+    "\ntotal counts:", sum(counts))
 ```
 
-## result of fitting operation
-
+## Result of fitting operation
 ```{r}
@@
 
-## residuals for sufficiently short T-mers you can look at all of them (2-2? 3-1?)
-## the top and bottom residuals for longer T-mers (9-5? 9-9?)
+## residuals for 2-2 Tmers
+```{r, results="asis"}
+counts_table(in_longwin=2, in_shortwin=2, half_n_show=5)
+```
+
+## residuals for 3-1 Tmers
+```{r, results="asis"}
+counts_table(in_longwin=3, in_shortwin=1, half_n_show=5)
+```
+
+## residuals for 9-5 Tmers
+```{r, results="asis"}
+counts_table(in_longwin=9, in_shortwin=5, half_n_show=5)
+```
+
 ## if present, MCMC traces
+
 ## ... and posterior marginal distributions.

### 2014-10-08 generic.Rmd
Making the .Rmd output tidier.

@@
 # Results from fitting procedure
-```{r}
+
+*Invocation:*
+```{r, echo=FALSE}
 library(pander)
@@
 ## Fitted parameter values
-```{r, results="asis"}
+```{r, echo=FALSE, results="asis"}
 par_df = data.frame(value = model@results[["par"]], row.names = names(model@results[["par"]]))
@@
 ## Window sizes and amount of data used to fit
-```{r}
+```{r, echo=FALSE, results="hide"}
+# For some reason the code doesn't work if we don't print out some of the
+# counts, but we can do it in a hidden block.
 counts <- model@counts@counts
-head(counts) # for some reason the code doesn't work if we don't print out some of the counts
+head(counts)
+```
+
+```{r, echo=FALSE}
 cat("left window size:", model@counts@leftwin,
@@
 ## Result of fitting operation
-```{r}
+```{r, echo=FALSE}
 cat("convergence: ", model@results$convergence,
@@
 ## residuals for 2-2 Tmers
-```{r, results="asis"}
+```{r, echo=FALSE, results="asis"}
 counts_table(in_longwin=2, in_shortwin=2, half_n_show=5)

### 2014-10-13 generic.Rmd
alternative to ggplot (looks nice, but grr, dependencies)

@@
 library(pander)
-library(ggplot2)
+# library(ggplot2)
 panderOptions("table.split.table", Inf)  # Let tables become arbitrarily wide rather than wrapping.
@@
 par_df = par_df[grep("->", par_df$names),]
-ggplot(par_df, aes(x = names, y = value)) +
-    geom_bar(stat = "identity", fill="white", color="gray") +
-    theme(axis.text.x = element_text(angle = 90))
+plot(1:2,type='n',xaxt='n',yaxt='n',bty='n',xlab='',ylab='')
+par(mar=c(max(strwidth(par_df$names))/(0.75*strheight("A")),3,1,1)+.1, new=TRUE)
+barplot( par_df$value, space=0.2 )
+axis(1, at=(1:nrow(par_df))*1.2-0.6, las=3, labels=par_df$names)
+plot(1:2,type='n',xaxt='n',yaxt='n',bty='n',xlab='',ylab='')
+par(mar=c(max(strwidth(par_df$names))/(0.75*strheight("A")),3,1,1)+.1, new=TRUE)
+barplot( par_df$value, log='y', space=0.2 )
+axis(1, at=(1:nrow(par_df))*1.2-0.6, las=3, labels=par_df$names)
+# ggplot(par_df, aes(x = names, y = value)) +
+#     geom_bar(stat = "identity", fill="white", color="gray") +
+#     theme(axis.text.x = element_text(angle = 90))
 ```

### 2014-10-17 simulation.Rmd
gh-27: Adding a results-combining rscript step.
Comparing results to simulation.

--- 2014-10-08 generic.Rmd
@@
 ```{r, echo=FALSE, results="asis"}
-par_df = data.frame(value = model@results[["par"]], row.names = names(model@results[["par"]]))
-pandoc.table(par_df, style="rmarkdown")
-par_df$names = row.names(par_df)
-par_df = par_df[grep("->", par_df$names),]
-ggplot(par_df, aes(x = names, y = value)) +
-    geom_bar(stat = "identity", fill="white", color="gray") +
-    theme(axis.text.x = element_text(angle = 90))
+mr_compare <- data.frame(fit=model@mutrates / time)
+mr_compare$names <- rownames(mr_compare)
+simvalues= data.frame(names=mutnames(simseq.config$tip$mutpats), 
+                      simulated=simseq.config$tip$mutrates)
+mr_compare <- merge(simvalues, mr_compare)
+pandoc.table(mr_compare, style="rmarkdown")
+```
+
+The corresponding plot, where bars are the simulated values, and points are the fit values:
+```{r, echo=FALSE, results="asis"}
+ggplot(mr_compare, aes(x=names)) +
+    geom_bar(aes(y=fit), stat = "identity", fill="white", color="gray") + 
+    geom_point(aes(y=simulated)) +
+    theme(axis.text.x = element_text(angle = 90),
+          axis.title.y = element_blank())
 ```
@@
 ```
-
-## if present, MCMC traces
-
-## ... and posterior marginal distributions.

### 2014-10-17 simulation.Rmd
Adding `scons seed` that runs things w/ diff seeds.

--- 2014-10-08 generic.Rmd
@@
 
-## Fitted parameter values
-```{r, echo=FALSE, results="asis"}
-par_df = data.frame(value = model@results[["par"]], row.names = names(model@results[["par"]]))
-pandoc.table(par_df, style="rmarkdown")
-par_df$names = row.names(par_df)
-par_df = par_df[grep("->", par_df$names),]
-ggplot(par_df, aes(x = names, y = value)) +
-    geom_bar(stat = "identity", fill="white", color="gray") +
-    theme(axis.text.x = element_text(angle = 90))
-```
-
-## Window sizes and amount of data used to fit
+## Details
 ```{r, echo=FALSE, results="hide"}
@@
     "\ncomplete window size:", nchar(rownames(counts)[[1]]),
-    "\ntotal counts:", sum(counts))
+    "\ntotal counts:", sum(counts),
+    "\ntime:", simseq.opt$tlen,
+    "\nseed:", simseq.opt$seed
+    )
 ```
@@
     "\nmessage: ", model@results$message)
+```
+
+## Fitted parameter values
+```{r, echo=FALSE, results="asis"}
+time <- as.numeric(simseq.opt$tlen)
+mr_compare <- data.frame(fit=model@mutrates / time)
+mr_compare$names <- rownames(mr_compare)
+simvalues= data.frame(names=mutnames(simseq.config$tip$mutpats),
+                      simulated=simseq.config$tip$mutrates)
+mr_compare <- merge(simvalues, mr_compare)
+pandoc.table(mr_compare, style="rmarkdown")
+```
+
+The corresponding plot, where bars are the fit values, and points are the simulated values:
+```{r, echo=FALSE, results="asis"}
+ggplot(mr_compare, aes(x=names)) +
+    geom_bar(aes(y=fit), stat = "identity", fill="white", color="gray") +
+    geom_point(aes(y=simulated)) +
+    theme(axis.text.x = element_text(angle = 90),
+          axis.title.y = element_blank())
 ```

### 2014-10-19 generic.Rmd
debugging json-tasep

--- 2014-10-08 generic.Rmd
@@
                 in_shortwin=in_shortwin,
-                in_leftwin=in_shortwin,
+                in_leftwin=floor((in_longwin-in_shortwin)/2),
                 counts=model@counts)
@@
 }
-cat(model@invocation)
+cat(model@invocation,"\n")
 ```
@@
 ```{r, echo=FALSE, results="asis"}
-par_df = data.frame(value = model@results[["par"]], row.names = names(model@results[["par"]]))
+par_df = coef(model)
 pandoc.table(par_df, style="rmarkdown")
-par_df$names = row.names(par_df)
-par_df = par_df[grep("->", par_df$names),]
-ggplot(par_df, aes(x = names, y = value)) +
+par_df = par_df[grep("->", names(par_df))]
+ggplot(data.frame(names=names(par_df),value=par_df), aes(x = names, y = value)) +
     geom_bar(stat = "identity", fill="white", color="gray") +
@@
     "\ncomplete window size:", nchar(rownames(counts)[[1]]),
-    "\ntotal counts:", sum(counts))
+    "\ntotal counts:", sum(counts), "\n")
 ```
@@
 cat("convergence: ", model@results$convergence,
-    "\nmessage: ", model@results$message)
+    "\nmessage: ", model@results$message, "\n")
 ```

### 2014-10-19 input-output.R

@@ -52,8 +52,6 @@ read.counts <- function (infile,leftwin,bases,longpats,shortpats) {
 
 read.config <- function (configfile,quiet=FALSE) {
     # read in JSON config file
-    #   and put in default parameters if missing
-    #   e.g. treeify it if not already
     con <- openread(configfile)
     json <- paste(readLines(con, warn = FALSE), collapse = "\n")
     close(con)
@@ -78,6 +76,20 @@ treeify.config <- function (config,tlen=NULL) {
     }
     if (is.null(config$tree$tip.label) | is.null(config$tree$node.label)) { stop("Please label tips and nodes on the tree.") }
     return(config)
+}
+
+fill.default.config <- function (config, defaults=NULL) {
+    # fill in default values in a config list
+    # for selpats, bases, fixfn, fixfn.params
+    for (x in c("selpats","bases","fixfn.params")) {
+        if (is.null(config[[x]])) {
+            config[[x]] <- if (is.null(defaults[[x]])) { list() } else { defaults[[x]] }
+        }
+    }
+    if (is.null(config[["fixfn"]])) { 
+        config[["fixfn"]] <- if(is.null(defaults[["fixfn"]])) { null.fixfn } else { defaults[["fixfn"]] }
+    }
+    return( config )
 }
 
 # tree helper functions
@@ -107,6 +119,7 @@ parse.fixfn <- function (fixfn,fixfn.params) {
     #   either by looking it up as a name
     #   or parsing it directly
     # also, check the arguments match fixfn.params
+    if (is.null(fixfn)) { fixfn <- null.fixfn }
     if (is.character(fixfn)) {
         if (exists(fixfn,mode="function")) {
             fixfn <- get(fixfn,mode="function")

### 2014-10-19 json-tasep/workflow.sh

@@ -9,7 +9,7 @@ Rscript ../sim-seq.R -c tasep-model.json -t 0.1 -s 10000 -d simseqs -o sim-tasep
 
 # and count the Tmers
 echo "counting"
-Rscript ../count-seq.R -i sim-tasep-123456.RData -w 4 -s 2 -l 1
+Rscript ../count-seq.R -i simseqs/sim-tasep-123456.RData -w 4 -s 2 -l 1
 
 # precompute generator matrices:
 #   width-4 with all selection
@@ -17,7 +17,10 @@ echo "making genmatrix, width 4"
 Rscript ../make-genmat.R -c genmatrices/complete.json -w 4 --meanboundary 1
 
 echo "fitting, width 4"
-Rscript ../fit-model.R -i sim-tasep-123456.4.2.l1.counts -l 1 -m genmatrices/genmatrix-4-complete.RData -j 54321
+Rscript ../fit-model.R -i simseqs/sim-tasep-123456-4-root-2-tip-l1.counts -l 1 -m genmatrices/genmatrix-4-complete.RData -j 54321
 
 echo "computing residuals"
-Rscript ../compute-resids.R -i sim-tasep-123456-genmatrix-4-complete-54321.RData
+Rscript ../compute-resids.R -i simseqs/sim-tasep-123456-4-root-2-tip-l1-genmatrix-4-complete-54321.RData
+
+echo "look at results"
+../templated-Rmd.sh ../simulation.Rmd simseqs/sim-tasep-123456-4-root-2-tip-l1-genmatrix-4-complete-54321.RData

### 2014-10-19 templated-Rmd.sh

@@ -3,7 +3,7 @@
 set -e
 set -u
 
-usage="Usage: $0 [-o OUTFILE] TEMPLATE RDATA"
+usage="Usage: $0 TEMPLATE RDATA"
 
 test "$#" -eq 2 || {
   echo $usage >&2

### 2014-10-22 simulation.Rmd
json-tasep works with Rmd

@@
 # counts, but we can do it in a hidden block.
-counts <- model@counts@counts
-head(counts)
+counttab <- counts(model)
+head(counttab)
 ```
@@
 cat("left window size:", model@counts@leftwin,
-    "\ncomplete window size:", nchar(rownames(counts)[[1]]),
-    "\ntotal counts:", sum(counts),
+    "\ncomplete window size:", nchar(rownames(counttab)[[1]]),
+    "\ntotal counts:", sum(counttab),
     "\ntime:", simseq.opt$tlen,
-    "\nseed:", simseq.opt$seed
+    "\nseed:", simseq.opt$seed,
+    "\n"
     )
@@
 cat("convergence: ", model@results$convergence,
-    "\nmessage: ", model@results$message)
+    "\nmessage: ", model@results$message,
+    "\n" )
 ```

### 2014-10-22 context-inference-fns.R

@@ -322,6 +322,7 @@ setClass("contextMCMC", representation(
          contains="context")
 
 setMethod("dimnames", signature=c(x="context"), definition=function (x) { dimnames(x@counts) } )
+setMethod("counts", signature=c(x="context"), definition=function (x) { counts(x@counts) } )
 
 # We would like the likfun function to automatically have access to the stuff in the context object
 #  ... where is 'self'?!?
@@ -821,12 +822,13 @@ computeresids <- function (model, pretty=TRUE, in_longwin=longwin(model), in_sho
     } else {
         counts <- model@counts
     }
+
+    expected <- fitted( model, longwin=longwin(model), shortwin=shortwin(model), leftwin=leftwin(model), initcounts=rowSums(counts), genmatrix=genmatrix )
+
     if ( (in_longwin < longwin(counts)) || (in_shortwin < shortwin(counts)) ) {
         counts <- projectcounts( counts, in_leftwin, in_shortwin, in_longwin-in_leftwin-in_shortwin )
+        expected <- projectcounts( expected, in_leftwin, in_shortwin, in_longwin-in_leftwin-in_shortwin )
     }
-    stopifnot( all( rownames(counts) == rownames(genmatrix) ) )
-
-    expected <- fitted( model, longwin=in_longwin, shortwin=in_shortwin, leftwin=in_leftwin, initcounts=rowSums(counts), genmatrix=genmatrix )
 
     if (pretty) {
         # data frame with columns for long pattern, short pattern, observed, expected, residual, z-score

### 2014-10-22 fit-model.R

@@ -64,6 +64,7 @@ likfun <- function (params){
 }
 
 initpar <- c( adhoc.mutrates, adhoc.selcoef, adhoc.fixparams )
+names(initpar) <- c( mutnames(genmatrix@mutpats), selnames(genmatrix@selpats), fixparams(genmatrix) )
 stopifnot( length(initpar) == nmuts(genmatrix)+nsel(genmatrix)+length(fixparams(genmatrix)) )
 lbs <- c( rep(1e-6,nmuts(genmatrix)), rep(-5,nsel(genmatrix)), rep(-Inf,length(fixparams(genmatrix))) )
 ubs <- c( rep(2,nmuts(genmatrix)), rep(5,nsel(genmatrix)), rep(Inf,length(fixparams(genmatrix))) )

### 2014-10-22 testing-code/check-sim.Rmd
can check simulated sequences to see they agree with expected

# Compare simulation to expected counts under true parameters

```{r, echo=FALSE}
# Should load: simulation results; generator matrix
library(pander)
library(ggplot2)
panderOptions("table.split.table", Inf)  # Let tables become arbitrarily wide rather than wrapping.
source("../context-inference-fns.R")
source("../sim-context-fns.R")
top_n_bottom <- function(df, k) { # Take the top and bottom k rows
    n = nrow(df)
    if(2*k < n) {
        df[c(1:k,(n-(k-1)):n), ]
    }
    else {
        df
    }
}
counts_table <- function(in_longwin, in_shortwin, half_n_show) {
    if ( in_longwin > longwin(genmatrix) || in_shortwin > longwin(genmatrix) ) {
        cat("Window size of counts too small.")
    }
    else {
        observed <- projectcounts( counttab, lcountwin=floor((in_longwin-in_shortwin)/2), countwin=in_shortwin, rcountwin=ceiling((in_longwin-in_shortwin)/2) )
        expected <- projectcounts( expected, lcountwin=floor((in_longwin-in_shortwin)/2), countwin=in_shortwin, rcountwin=ceiling((in_longwin-in_shortwin)/2) )
        obsexp <- cbind( as.matrix(observed@counts), as.matrix(expected@counts) )
        colnames(obsexp) <- c( paste("obs:",colnames(observed),sep=''), paste("exp:",colnames(observed),sep='') )
        obsexp <- obsexp[, rbind(1:ncol(observed),ncol(observed)+(1:ncol(observed)))]
        pandoc.table(top_n_bottom(obsexp, half_n_show), style="rmarkdown")
    }
}
longpats <- getpatterns(longwin(genmatrix),simseq.config$bases)
counttab <- counttrans.list( list(longpats,longpats), simseqs=simseqs, leftwin=0, bases=simseq.config$bases )
initcounts <- rowSums(counttab)
time <- as.numeric(simseq.opt$tlen)
expected <- predictcounts(longwin=longwin(genmatrix), shortwin=longwin(genmatrix), leftwin=0, initcounts=initcounts, 
        mutrates=time*simseq.config$tip$mutrates, selcoef=simseq.config$tip$selcoef, genmatrix=genmatrix, params=simseq.config$tip$fixfn.params )
```

# Simulated parameter values
```{r, echo=FALSE, results="asis"}
params <- as.data.frame( matrix( c(
        time,
        simseq.config$tip$mutrates,
        simseq.config$tip$selcoef,
        simseq.config$tip$fixfn.params
    ), nrow=1 ) )
colnames(params) <- c(
        "time",
        mutnames(simseq.config$tip$mutpats),
        selnames(simseq.config$tip$selpats),
        names(simseq.config$tip$fixfn.params)
    )
pandoc.table(params, style="rmarkdown")
```

```{r, echo=FALSE}
cat("window size:", longwin(genmatrix),
    "\ntotal counts:", sum(counttab@counts),
    "\ntime:", simseq.opt$tlen,
    "\nseed:", simseq.opt$seed,
    "\n"
    )
```

## residuals for 2-2 Tmers
```{r, echo=FALSE, results="asis"}
counts_table(in_longwin=2, in_shortwin=2, half_n_show=5)
```

## residuals for 3-1 Tmers
```{r, results="asis"}
counts_table(in_longwin=3, in_shortwin=1, half_n_show=5)
```

## residuals for 9-5 Tmers
```{r, results="asis"}
counts_table(in_longwin=9, in_shortwin=5, half_n_show=5)
```

### 2014-10-22 json-ising/workflow.sh

@@ -4,43 +4,50 @@ set -eu
 set -o pipefail
 
 # simulate up some sequence for testing
-Rscript ../sim-seq.R -c ising-model.json -t .1 -s 100000 -o test-ising-123456.RData
+Rscript ../sim-seq.R -c ising-model.json -t .1 -s 100000 -d simseqs -o test-ising-123456.RData
 
 # and count the Tmers
-Rscript ../count-seq.R -i test-ising-123456.RData -w 3 -s 1 -l 1
+Rscript ../count-seq.R -i simseqs/test-ising-123456.RData -w 3 -s 1 -l 1
 
 # precompute generator matrices:
 #   width-3
 Rscript ../make-genmat.R -c genmatrices/complete.json -w 3
 
-# fit a model
-Rscript ../fit-model.R -i test-ising-123456.3.1.l1.counts -l 1 -m genmatrices/genmatrix-3-complete.RData -j 54321
+echo "fitting the model"
+Rscript ../fit-model.R -i simseqs/test-ising-123456-3-root-1-tip-l1.counts -l 1 -m genmatrices/genmatrix-3-complete.RData -j 54321
 
-# mcmc also
-Rscript ../mcmc-model.R -i test-ising-123456-genmatrix-3-complete-54321.RData -c ising-model.json -b 3 -j 1111
-Rscript ../mcmc-model.R -i test-ising-123456-genmatrix-3-complete-54321-mcmc-1111.RData -b 100 -j 2222
-Rscript ../mcmc-model.R -i test-ising-123456-genmatrix-3-complete-54321-mcmc-2222.RData -b 1000 -j 3333
+echo "computing residuals"
+Rscript ../compute-resids.R -i simseqs/test-ising-123456-3-root-1-tip-l1-genmatrix-3-complete-54321.RData
 
+echo "look at results"
+../templated-Rmd.sh ../simulation.Rmd simseqs/test-ising-123456-3-root-1-tip-l1-genmatrix-3-complete-54321.RData simseqs/test-ising-123456.RData
 
-# OK, now do this on more Tmer sizes:
-Rscript ../count-seq.R -i test-ising-123456.RData -w 4 -s 2 -l 1
-Rscript ../make-genmat.R -c genmatrices/complete.json -w 4
-Rscript ../fit-model.R -i test-ising-123456.4.2.l1.counts -l 1 -m genmatrices/genmatrix-4-complete.RData -j 54321
-
[…]

### 2014-10-22 json-tasep/workflow.sh

@@ -16,6 +16,9 @@ Rscript ../count-seq.R -i simseqs/sim-tasep-123456.RData -w 4 -s 2 -l 1
 echo "making genmatrix, width 4"
 Rscript ../make-genmat.R -c genmatrices/complete.json -w 4 --meanboundary 1
 
+echo "check simulated model matches expected"
+../templated-Rmd.sh ../testing-code/check-sim.Rmd simseqs/sim-tasep-123456.RData genmatrices/genmatrix-4-complete.RData
+
 echo "fitting, width 4"
 Rscript ../fit-model.R -i simseqs/sim-tasep-123456-4-root-2-tip-l1.counts -l 1 -m genmatrices/genmatrix-4-complete.RData -j 54321
 

### 2014-10-22 sim-context-fns.R

@@ -157,7 +157,7 @@ counttrans.list <- function (lpatterns, seqlist=lapply(simseqs,"[[","finalseq"),
         for (j in rev(seq_along(ii)[-1])) { ii[j-1] <- ii[j-1] + ( ii[j] > npats[j] ) }
         ii[-1] <- 1 + ( (ii[-1]-1) %% npats[-1] )
     }
-    colpatterns <- do.call( expand.grid, list( shortpats )[ rep.int(1,length(seqlist)-1) ] )
+    colpatterns <- do.call( expand.grid, lpatterns[-1] )
     colnames(colpatterns) <- names(seqlist)[-1]
     counts <- lapply( counts, function (x) {
             dim(x) <- c(dim(x)[1],prod(dim(x)[-1]))

### 2014-10-22 simulation.Rmd
.

@@
         cat("Window size of counts too small.")
+        return(NULL)
     }
@@
         pandoc.table(top_n_bottom(resids, half_n_show), style="rmarkdown")
+        return(invisible(resids))
+    }
+}
+counts_plot <- function (resids) {
+    if (NROW(resids)>0) { 
+        hist(resids$z, xlab="z-score", ylab='') 
+        outfac <- factor(resids$outpat)
+        with(resids, plot(observed, expected, xlab='observed', ylab='expected', col=rainbow(nlevels(outfac))[as.numeric(outfac)], pch=20 ) )
+        abline(0,1)
+        with(resids, plot(observed, expected, xlab='observed', ylab='expected', log='xy', col=rainbow(nlevels(outfac))[as.numeric(outfac)], pch=20 ) )
+        abline(0,1)
     }
@@
     "\ntime:", simseq.opt$tlen,
-    "\nseed:", simseq.opt$seed,
+    "\nseed:", formatC(simseq.opt$seed,format="d"),
     "\n"
@@
 time <- as.numeric(simseq.opt$tlen)
-mr_compare <- data.frame(fit=model@mutrates / time)
-mr_compare$names <- rownames(mr_compare)
-simvalues= data.frame(names=mutnames(simseq.config$tip$mutpats),
-                      simulated=simseq.config$tip$mutrates)
-mr_compare <- merge(simvalues, mr_compare)
+mr_compare <- data.frame( 
+    fit=c( model@mutrates / time, model@selcoef, model@params), 
+    stringsAsFactors=FALSE )
+simvalues <- data.frame(
+        names=c(
+            mutnames(simseq.config$tip$mutpats), 
+            selnames(simseq.config$tip$selpats), 
+            names(simseq.config$tip$fixfn.params) ),
+        simulated=unlist(c(
+            simseq.config$tip$mutrates,
+            simseq.config$tip$selcoef,
+            simseq.config$tip$fixfn.params ) )
+)
+mr_compare$simulated <- simvalues$simulated[match(rownames(mr_compare),simvalues$names)]
 pandoc.table(mr_compare, style="rmarkdown")
@@
 ```{r, echo=FALSE, results="asis"}
+mr_compare$names <- factor(rownames(mr_compare), levels=rownames(mr_compare))
 ggplot(mr_compare, aes(x=names)) +
@@
 ```{r, echo=FALSE, results="asis"}
-counts_table(in_longwin=2, in_shortwin=2, half_n_show=5)
+resids <- counts_table(in_longwin=2, in_shortwin=2, half_n_show=20)
+counts_plot(resids)
 ```
@@
 ```{r, results="asis"}
-counts_table(in_longwin=3, in_shortwin=1, half_n_show=5)
+resids <- counts_table(in_longwin=3, in_shortwin=1, half_n_show=20)
+counts_plot(resids)
 ```
@@
 ```{r, results="asis"}
-counts_table(in_longwin=9, in_shortwin=5, half_n_show=5)
+resids <- counts_table(in_longwin=9, in_shortwin=5, half_n_show=20)
+counts_plot(resids)
 ```

### 2014-10-22 testing-code/sim-short-ising.R

@@ -1,0 +1,53 @@
+#!/usr/bin/Rscript
+
+require(expm)
+require(parallel)
+numcores<-as.numeric(scan(pipe("cat /proc/cpuinfo | grep processor | tail -n 1 | awk '{print $3}'")))+1
+
+# Run some tests!!
+source("../sim-context-fns.R")
+source("../context-inference-fns.R")
+
+bases <- c("X","O")
+
+patlen <- 2
+mutpats <- list( 
+    list( c("O","X") ),
+    list( c("X","O") )
+    ) 
+selpats <- list(
+        c("OX","XO"),
+        c("X")
+    )
+
+
+fixfn <- function (ds,...) { 1/(1+exp(-ds)) }
+
+# Short sequences:
+tlen <- 2
+seqlen <- 5
+mutrates <- c(1,1)
+selcoef <- c(-.5,.5)
+
+initseq <- rinitseq(seqlen,bases)
+simseqs <- mclapply( 1:1000, function (k) 
+        simseq( seqlen, tlen, intiseq=initseq, mutpats=mutpats, selpats=selpats, mutrates=mutrates, selcoef=selcoef, bases=bases, fixfn=fixfn ) 
+    , mc.cores=numcores)
+full.genmatrix <- makegenmatrix( mutpats, selpats, patlen=seqlen, boundary="none", bases=bases, fixfn=fixfn )
+require(expm)
+full.transmatrix <- expm( tlen*(as.matrix(full.genmatrix)-diag(rowSums(full.genmatrix))) )
+stopifnot( all( abs( rowSums(full.transmatrix) - rep(1.0,nrow(full.transmatrix)) ) < 1e-8 ) )
+expected.freqs <- full.transmatrix[match(as.character(initseq),rownames(full.transmatrix)),]
+fseqs <- as.numeric(table( factor( sapply( simseqs, function (x) as.character(x$finalseq) ), levels=names(expected.freqs) ) ))
+# plot( length(simseqs)*expected.freqs, fseqs ); abline(0,1)
+resids <- data.frame( observed=fseqs, expected=expected.freqs*length(simseqs) )
+resids$z <- (resids$observed-resids$expected)/sqrt(resids$expected)
+resids$p <- pbinom( resids$observed, size=length(simseqs), prob=expected.freqs )
+
+cat("Beginning from ", as.character(initseq), ":\n")
+print(resids)
+
[…]

### 2014-10-29 simulation.Rmd
fixed bug in check-sim.Rmd (ordering of counttrans.list); inference checks out in ising

@@
 mr_compare$names <- factor(rownames(mr_compare), levels=rownames(mr_compare))
-ggplot(mr_compare, aes(x=names)) +
+param_sizes <- floor(log10(abs(mr_compare$simulated)+1e-4)/2)
+for (k in unique(param_sizes)) {
+    print( ggplot(mr_compare[param_sizes==k,], aes(x=names)) + xlab("") +
         geom_bar(aes(y=fit), stat = "identity", fill="white", color="gray") +
@@
               axis.title.y = element_blank())
+      )
+}
 ```

### 2014-10-29 testing-code/check-sim.Rmd
fixed bug in check-sim.Rmd (ordering of counttrans.list); inference checks out in ising

@@
 longpats <- getpatterns(longwin(genmatrix),simseq.config$bases)
-counttab <- counttrans.list( list(longpats,longpats), simseqs=simseqs, leftwin=0, bases=simseq.config$bases )
+counttab <- counttrans.list( list(longpats,longpats), simseqs=simseqs[c("root","tip")], leftwin=0, bases=simseq.config$bases )
 initcounts <- rowSums(counttab)
@@
     ), nrow=1 ) )
-colnames(params) <- c(
+colnames(params) <- gsub("|","\\|",c(
         "time",
@@
         names(simseq.config$tip$fixfn.params)
-    )
+    ), fixed=TRUE)
 pandoc.table(params, style="rmarkdown")

### 2014-10-29 fit-model.R

@@ -16,7 +16,6 @@ option_list <- list(
         make_option( c("-l","--leftwin"), type="integer", help="Size of left-hand context." ),
         make_option( c("-m","--gmfile"), type="character", default="TRUE", help="File with precomputed generator matrix."),
         make_option( c("-x","--maxit"), type="integer", default=100, help="Number of iterations of optimization to run for. [default=%default]"),
-        make_option( c("-z","--dummy"), type="character", help="If this is the name of a config file, skip optimization and put the parameters into the resulting model object."),
         make_option( c("-j","--jobid"), type="character", default=formatC(1e6*runif(1),width=6,format="d",flag="0"), help="Unique job id. [default random]")
     )
 opt <- parse_args(OptionParser(option_list=option_list,description=usage))
@@ -43,18 +42,19 @@ projmatrix <- collapsepatmatrix( ipatterns=rownames(genmatrix), leftwin=leftwin(
 adhoc <- countmuts(counts=counts,mutpats=genmatrix@mutpats,leftwin=leftwin(counts))
 adhoc.mutrates <- adhoc[1,]/adhoc[2,]
 adhoc.mutrates <- ifelse( is.finite(adhoc.mutrates) & adhoc.mutrates > 0, adhoc.mutrates, 1e-4 )
-# start selection parameters at zero, why not
+# start selection parameters at zero, why not?
 adhoc.selcoef <- rep(1e-6,nsel(genmatrix))
 names(adhoc.selcoef) <- selnames(genmatrix@selpats)
-# ditto for additional parameters
+# how to choose additional parameters?? Need guidance here.
 adhoc.fixparams <- rep(1,length(fixparams(genmatrix)))
 names(adhoc.fixparams) <- fixparams(genmatrix)
+adhoc.fixparams[names(adhoc.fixparams)=="Ne"] <- 200 + 2000*runif(1)
 
 # Compute (quasi)-likelihood function using all counts -- multinomial as described in eqn:comp_like.
 likfun <- function (params){
     # params are: mutrates, selcoef, fixparams
     fparams <- params[seq( 1+nmuts(genmatrix)+nsel(genmatrix), length.out=length(fixparams(genmatrix)) )]
-    names(fparams) <- names(fixparams(genmatrix))
[…]

### 2014-10-31 simulation.Rmd

@@
 ```
+
+# Pairwise marginal posteriors from MCMC, colored by time in MCMC run
+```{r, results="asis"}
+mcmcrun <- model@results$batch
+subseq <- seq(1,nrow(mcmcrun),length.out=1000)
+x <- rbind( mcmcrun[subseq,], mr_compare$simulated ) 
+# color by time
+cols <- c( adjustcolor(rainbow(64),.2)[1+floor(65*(1:(nrow(x)-1))/nrow(x))], adjustcolor("black",.5) )
+pairs( x, col=cols, pch=20, cex=c(rep(.5,nrow(x)-1),2) )
+```

### 2014-10-31 simulation.Rmd
.

--- 2014-10-29 simulation.Rmd
@@
 ```
+
+# Pairwise marginal posteriors from MCMC, colored by time in MCMC run
+```{r, results="asis"}
+if (class(model)=="contextMCMC") {
+    mcmcrun <- model@results$batch
+    subseq <- seq(1,nrow(mcmcrun),length.out=1000)
+    x <- rbind( mcmcrun[subseq,], mr_compare$simulated ) 
+    # color by time
+    cols <- c( adjustcolor(rainbow(64),.2)[1+floor(65*(1:(nrow(x)-1))/nrow(x))], adjustcolor("black",.5) )
+    pairs( x, col=cols, pch=20, cex=c(rep(.5,nrow(x)-1),2) )
+} else {
+    cat("Not an MCMC run.")
+}
+```

### 2014-11-02 simulation.Rmd
.

--- 2014-10-29 simulation.Rmd
@@
 ```{r, echo=FALSE}
+if (class(model)=="context") {
     cat("convergence: ", model@results$convergence,
@@
         "\n" )
+} else if (class(model)=="contextMCMC") {
+    cat(
+        "number of batches: ", model@results$nbatch,
+        "\nbatch length: ", model@results$blen,
+        "\nacceptance rate: ", model@results$accept,
+        "\nrun time: ", model@results$time['elapsed'],
+        "\n" )
+}
 ```
@@
 ```
+
+# Pairwise marginal posteriors from MCMC, colored by time in MCMC run
+```{r, results="asis", echo=FALSE, fig.align='center', }
+if (class(model)=="contextMCMC") {
+    subseq <- seq(1,nrow(model@results$batch),length.out=min(1000,nrow(model@results$batch)))
+    x <- rbind( model@results$batch[subseq,], mr_compare$simulated[model@results$use.par] ) 
+    colnames(x) <- rownames(mr_compare)[model@results$use.par]
+    # color by time
+    cols <- c( adjustcolor(rainbow(64),.2)[1+floor(65*(1:(nrow(x)-1))/nrow(x))], adjustcolor("black",.5) )
+    pairs( x, col=cols, pch=20, cex=c(rep(.5,nrow(x)-1),2) )
+} else {
+    cat("Not an MCMC run.")
+}
+```

### 2014-11-02 simulation.Rmd
.

@@
 ## Details
-```{r, echo=FALSE, results="hide"}
+```{r, include=FALSE}
 # For some reason the code doesn't work if we don't print out some of the
@@
 counttab <- counts(model)
-head(counttab)
 ```
@@
 The corresponding plot, where bars are the fit values, and points are the simulated values:
-```{r, echo=FALSE, results="asis"}
+```{r, echo=FALSE, results="asis", fig.align='center'}
 mr_compare$names <- factor(rownames(mr_compare), levels=rownames(mr_compare))
@@
 ## residuals for 2-2 Tmers
-```{r, echo=FALSE, results="asis"}
+```{r, echo=FALSE, results="asis", fig.align='center'}
 resids <- counts_table(in_longwin=2, in_shortwin=2, half_n_show=20)
@@
 ## residuals for 3-1 Tmers
-```{r, results="asis"}
+```{r, results="asis", fig.align='center'}
 resids <- counts_table(in_longwin=3, in_shortwin=1, half_n_show=20)
@@
 ## residuals for 9-5 Tmers
-```{r, results="asis"}
+```{r, results="asis", fig.align='center'}
 resids <- counts_table(in_longwin=9, in_shortwin=5, half_n_show=20)
@@
 
+# Trace plot of the MCMC run:
+```{r, results="asis", echo=FALSE, fig.width=12, fig.height=7, fig.align='center'}
+if (class(model)=="contextMCMC") {
+    subseq <- seq(1,nrow(model@results$batch),length.out=min(1000,nrow(model@results$batch)))
+    x <- model@results$batch[subseq,]
+    colnames(x) <- rownames(mr_compare)[model@results$use.par]
+    matplot( model@results$blen*subseq, x, type='l', col=1:5, lty=1:6, xlab='number of MCMC iterations' )
+    abline(h=mr_compare$simulated[model@results$use.par], col=1:5, lty=1:6)
+    legend("topright",legend=colnames(x),col=1:5,lty=1:6)
+} else {
+    cat("Not an MCMC run.")
+}
+```
+
 # Pairwise marginal posteriors from MCMC, colored by time in MCMC run
-```{r, results="asis", echo=FALSE, fig.align='center', }
+```{r, results="asis", echo=FALSE, fig.width=12, fig.height=12, fig.align='center'}
 if (class(model)=="contextMCMC") {
@@
     cols <- c( adjustcolor(rainbow(64),.2)[1+floor(65*(1:(nrow(x)-1))/nrow(x))], adjustcolor("black",.5) )
-    pairs( x, col=cols, pch=20, cex=c(rep(.5,nrow(x)-1),2) )
+    pairs( x, col=cols, pch=20, cex=c(rep(.5,nrow(x)-1),2), gap=0 )
 } else {

### 2014-11-03 json-cpg/test-workflow.sh
added plots of MCMC runs to simulation.Rmd

@@ -22,7 +22,6 @@ echo "do mcmc"
 Rscript ../mcmc-model.R -i testsim/test-cpg-01-fit.RData -o testsim/test-cpg-01-mcmc-1.RData -c cpg-model-constrained.json --blen 3
 Rscript ../mcmc-model.R -i testsim/test-cpg-01-mcmc-1.RData -o testsim/test-cpg-01-mcmc-2.RData -c cpg-model-constrained.json --blen 100 -b 1000
 
-Rscript ../mcmc-model.R -i testsim/test-cpg-01-mcmc-1.RData -o testsim/test-cpg-01-mcmc-3.RData -c cpg-model-constrained.json --blen 100 -b 10
 
 
 echo "look at results: MLE"

### 2014-11-03 simulation.Rmd
.

@@
     colnames(x) <- rownames(mr_compare)[model@results$use.par]
-    matplot( model@results$blen*subseq, x, type='l', col=1:5, lty=1:6, xlab='number of MCMC iterations' )
+    param_sizes <- floor(log10(abs(mr_compare$simulated)+1e-4)/2)
+    param_sizes <- param_sizes[model@results$use.par]
+    for (k in unique(param_sizes)) {
+        print( 
+            matplot( model@results$blen*subseq, x[,param_sizes==k], type='l', col=1:5, lty=1:6, xlab='number of MCMC iterations' )
             abline(h=mr_compare$simulated[model@results$use.par], col=1:5, lty=1:6)
-    legend("topright",legend=colnames(x),col=1:5,lty=1:6)
+            legend("topright",legend=colnames(x)[param_sizes==k],col=1:5,lty=1:6)
+          )
+    }
 } else {

### 2014-11-16 simulation.Rmd
Adding a gather-results script.
The idea is to make summaries in this script so that we can use them
downstream in places other than the .Rmd.

cc @petrelharp

--- 2014-11-02 simulation.Rmd
@@
 ```{r, echo=FALSE, results="asis"}
-time <- as.numeric(simseq.opt$tlen)
-mr_compare <- data.frame( 
-    fit=c( model@mutrates / time, model@selcoef, model@params), 
-    stringsAsFactors=FALSE )
-simvalues <- data.frame(
-        names=c(
-            mutnames(simseq.config$tip$mutpats), 
-            selnames(simseq.config$tip$selpats), 
-            names(simseq.config$tip$fixfn.params) ),
-        simulated=unlist(c(
-            simseq.config$tip$mutrates,
-            simseq.config$tip$selcoef,
-            simseq.config$tip$fixfn.params ) )
-)
-mr_compare$simulated <- simvalues$simulated[match(rownames(mr_compare),simvalues$names)]
 pandoc.table(mr_compare, style="rmarkdown")
@@
     colnames(x) <- rownames(mr_compare)[model@results$use.par]
-    matplot( model@results$blen*subseq, x, type='l', col=1:5, lty=1:6, xlab='number of MCMC iterations' )
+    param_sizes <- floor(log10(abs(mr_compare$simulated)+1e-4)/2)
+    param_sizes <- param_sizes[model@results$use.par]
+    for (k in unique(param_sizes)) {
+        print( {
+            matplot( model@results$blen*subseq, x[,param_sizes==k], type='l', col=1:5, lty=1:6, xlab='number of MCMC iterations' )
             abline(h=mr_compare$simulated[model@results$use.par], col=1:5, lty=1:6)
-    legend("topright",legend=colnames(x),col=1:5,lty=1:6)
+            legend("topright",legend=colnames(x)[param_sizes==k],col=1:5,lty=1:6)
+        } )
+    }
 } else {

### 2014-11-16 gather-results.R

@@ -1,0 +1,37 @@
+#!/usr/bin/Rscript --vanilla
+require(optparse)
+
+usage <- "Gather and summarize results of analysis into a single Rmd file."
+
+option_list <- list(
+        make_option( c("-s","--sim"), type="character", help=".RData file containing simulation." ),
+        make_option( c("-f","--fit"), type="character", help=".RData file containing fit." ),
+        make_option( c("-o","--outfile"), type="character", help="File to save things in.")
+        )
+opt <- parse_args(OptionParser(option_list=option_list,description=usage))
+
+source("../context-inference-fns.R")
+
+load(opt$sim)
+load(opt$fit)
+
+time <- as.numeric(simseq.opt$tlen);
+mr_compare <- data.frame(
+    fit=c( model@mutrates / time, model@selcoef, model@params),
+    stringsAsFactors=FALSE );
+simvalues <- data.frame(
+        names=c(
+            mutnames(simseq.config$tip$mutpats),
+            selnames(simseq.config$tip$selpats),
+            names(simseq.config$tip$fixfn.params) ),
+        simulated=unlist(c(
+            simseq.config$tip$mutrates,
+            simseq.config$tip$selcoef,
+            simseq.config$tip$fixfn.params ) )
+);
+mr_compare$simulated <-
+    simvalues$simulated[match(rownames(mr_compare),simvalues$names)];
+
+save(simseq.config, simseq.opt, model, mr_compare,
+    file=opt$outfile);
+

### 2014-12-05 simulation.Rmd

@@
 ```{r, echo=FALSE, results="asis"}
+time <- as.numeric(simseq.opt$tlen)
+mr_compare <- data.frame( 
+    fit=c( model@mutrates / time, model@selcoef, model@params), 
+    stringsAsFactors=FALSE )
+simvalues <- data.frame(
+        names=c(
+            mutnames(simseq.config$tip$mutpats), 
+            selnames(simseq.config$tip$selpats), 
+            names(simseq.config$tip$fixfn.params) ),
+        simulated=unlist(c(
+            simseq.config$tip$mutrates,
+            simseq.config$tip$selcoef,
+            simseq.config$tip$fixfn.params ) )
+)
+mr_compare$simulated <- simvalues$simulated[match(rownames(mr_compare),simvalues$names)]
 pandoc.table(mr_compare, style="rmarkdown")

### 2014-12-05 testing-code/check-sim.Rmd
syntax for projectcounts changed to (left,long,short) to agree with everything else

@@
         cat("Window size of counts too small.")
+        return(NULL)
     }
     else {
-        observed <- projectcounts( counttab, lcountwin=floor((in_longwin-in_shortwin)/2), countwin=in_shortwin, rcountwin=ceiling((in_longwin-in_shortwin)/2) )
-        expected <- projectcounts( expected, lcountwin=floor((in_longwin-in_shortwin)/2), countwin=in_shortwin, rcountwin=ceiling((in_longwin-in_shortwin)/2) )
+        observed <- projectcounts( counttab, new.leftwin=floor((in_longwin-in_shortwin)/2), new.shortwin=in_shortwin, new.longwin=in_longwin, overlap=TRUE )
+        expected <- projectcounts( expected, new.leftwin=floor((in_longwin-in_shortwin)/2), new.shortwin=in_shortwin, new.longwin=in_longwin, overlap=TRUE )
         obsexp <- cbind( as.matrix(observed@counts), as.matrix(expected@counts) )
@@
     }
+    return(invisible(obsexp))
 }
@@
 ```{r, echo=FALSE, results="asis"}
-counts_table(in_longwin=2, in_shortwin=2, half_n_show=5)
+counts_table(in_longwin=2, in_shortwin=2, half_n_show=20)
 ```
@@
 ```{r, results="asis"}
-counts_table(in_longwin=3, in_shortwin=1, half_n_show=5)
+counts_table(in_longwin=3, in_shortwin=1, half_n_show=20)
 ```
@@
 ```{r, results="asis"}
-counts_table(in_longwin=9, in_shortwin=5, half_n_show=5)
+counts_table(in_longwin=9, in_shortwin=5, half_n_show=20)
 ```

### 2014-12-05 testing-code/check-sim.Rmd
predictcounts now optionally adjusts for overcounting

@@
     else {
-        observed <- projectcounts( counttab, new.leftwin=floor((in_longwin-in_shortwin)/2), new.shortwin=in_shortwin, new.longwin=in_longwin, overlap=TRUE )
-        expected <- projectcounts( expected, new.leftwin=floor((in_longwin-in_shortwin)/2), new.shortwin=in_shortwin, new.longwin=in_longwin, overlap=TRUE )
+        observed <- projectcounts( counttab, new.leftwin=floor((in_longwin-in_shortwin)/2), new.shortwin=in_shortwin, new.longwin=in_longwin, overlapping=TRUE )
+        expected <- projectcounts( expected, new.leftwin=floor((in_longwin-in_shortwin)/2), new.shortwin=in_shortwin, new.longwin=in_longwin, overlapping=TRUE )
         obsexp <- cbind( as.matrix(observed@counts), as.matrix(expected@counts) )

### 2014-12-05 context-inference-fns.R

@@ -760,7 +760,7 @@ predictcounts <- function (longwin, shortwin, leftwin, initcounts, mutrates, sel
     return( new("tuplecounts", leftwin=leftwin, counts=Matrix(fullcounts), bases=genmatrix@bases ) )
 }
 
-projectcounts <- function( counts, new.leftwin, new.shortwin, new.longwin, spacing=1 ) {
+projectcounts <- function( counts, new.leftwin, new.shortwin, new.longwin, overlapping=FALSE ) {
     # compute counts for shorter Tmers.
     #   valid ranges for parameters are
     #    (l-lc)^+ <= k < (l+w)-(lc+wc)+(r-rc)^-
@@ -769,15 +769,16 @@ projectcounts <- function( counts, new.leftwin, new.shortwin, new.longwin, spaci
     #       w = shortwin, wc = new.shortwin
     #       r = rightwin, rc = new.rightwin
     # if the original counts were from overlapping windows, then this will overcount the resulting patterns:
-    #    if you slide a window of length L in steps of size S then a subwindow of size W
-    #      will be seen in floor( (L-W)/S ) big windows;
-    #    so we need to divide the counts by min( floor( (longwin-new.longwin)/spacing ), floor( (shortwin-new.shortwin)/spacing ) )
-    #      ... but patterns at the boundary of the sequence will not be overcounted; take the ceiling of the resulting counts to fix these.
+    #    if you slide a window of length L in steps of size 1 then a subwindow of size W
+    #      will be seen in ( (L-W) ) big windows;
+    #    so we need to divide the counts by the factor 'overcount' below
+    #      ... but patterns at the boundary of the sequence will not be overcounted. 
+    #     Take the ceiling of the resulting counts to fix these.
     leftwin <- leftwin(counts)
     longwin <- longwin(counts)
     shortwin <- shortwin(counts)
     rightwin <- longwin-shortwin-leftwin
-    new.rightwin <- new.longwin-new.shortwin-new.rightwin
+    new.rightwin <- new.longwin-new.shortwin-new.leftwin
     if ( max(0L,leftwin-new.leftwin) > (leftwin+shortwin)-(new.leftwin+new.shortwin)+min(0L,rightwin-new.rightwin) ) {
         stop("unreconcilable windows specified.")
     }
[…]

### 2014-12-06 simulation.Rmd
added marginal likelihood surfaces to simulation.Rmd

--- 2014-11-16 simulation.Rmd
@@
 }
+counttab <- counts(model)
 cat(model@invocation,"\n")
@@
 
-## Details
-```{r, include=FALSE}
-# For some reason the code doesn't work if we don't print out some of the
-# counts, but we can do it in a hidden block.
-counttab <- counts(model)
-```
-
 ```{r, echo=FALSE}
-cat("left window size:", model@counts@leftwin,
-    "\ncomplete window size:", nchar(rownames(counttab)[[1]]),
+cat(
+    "long window size:", longwin(model@counts),
+    "short window size:", shortwin(model@counts),
+    "left window size:", leftwin(model@counts),
     "\ntotal counts:", sum(counttab),
@@
 ```{r, echo=FALSE, results="asis"}
+time <- as.numeric(simseq.opt$tlen)
+mr_compare <- data.frame( 
+    fit=c( model@mutrates / time, model@selcoef, model@params), 
+    stringsAsFactors=FALSE )
+simvalues <- data.frame(
+        names=c(
+            mutnames(simseq.config$tip$mutpats), 
+            selnames(simseq.config$tip$selpats), 
+            names(simseq.config$tip$fixfn.params) ),
+        simulated=unlist(c(
+            simseq.config$tip$mutrates,
+            simseq.config$tip$selcoef,
+            simseq.config$tip$fixfn.params ) )
+)
+mr_compare$simulated <- simvalues$simulated[match(rownames(mr_compare),simvalues$names)]
+mr_compare$names <- factor(rownames(mr_compare), levels=rownames(mr_compare))
 pandoc.table(mr_compare, style="rmarkdown")
@@
 ```{r, echo=FALSE, results="asis", fig.align='center'}
-mr_compare$names <- factor(rownames(mr_compare), levels=rownames(mr_compare))
-param_sizes <- floor(log10(abs(mr_compare$simulated)+1e-4)/2)
+param_log <- log(abs(mr_compare$simulated)+1e-4)
+param_sizes <- cut( param_log, breaks=seq(min(param_log), by=2, length.out=1+ceiling(diff(range(param_log))/2)), include.lowest=TRUE )
 for (k in unique(param_sizes)) {
@@
       )
+}
+```
+
+The likelihood surface about the fitted and simulated values:
+```{r, echo=FALSE, results="asis", fig.align='center'}
+timevec <- c( rep(as.numeric(simseq.opt$tlen),nmuts(model)), rep(1,nrow(mr_compare)-nmuts(model)) )
+f <- function (x) { model@likfun(x*timevec) }
+for (k in 1:nrow(mr_compare)) {
+    pvals <- seq( .9*min(mr_compare[k,c("fit","simulated")]), 1.1*max(mr_compare[k,c("fit","simulated")]), length.out=20 )
+    inds <- 1:nrow(mr_compare)
+    inds[k] <- 1+nrow(mr_compare)
+    lvals <- sapply( pvals, function (x) { f( c(mr_compare$fit,x)[inds] ) } )
+    l2vals <- sapply( pvals, function (x) { f( c(mr_compare$simulated,x)[inds] ) } )
+    plot( pvals, lvals, type='b', main=paste("likelihood surface for", mr_compare$name[k]), ylab="log-likelihood", xlab='value', ylim=range(lvals,l2vals) )
+    lines( pvals, l2vals, col='red', type='b' )
+    abline(v=mr_compare[k,c("fit","simulated")], col=c("black","red"))
 }
@@
         print( { 
-            matplot( model@results$blen*subseq, x[,param_sizes==k], type='l', col=1:5, lty=1:6, xlab='number of MCMC iterations' )
+            matplot( model@results$blen*subseq, x[,param_sizes==k], type='l', col=1:5, lty=1:6, xlab='number of MCMC iterations', ylab='parameter values' )
             abline(h=mr_compare$simulated[model@results$use.par], col=1:5, lty=1:6)
@@
     # color by time
-    cols <- c( adjustcolor(rainbow(64),.2)[1+floor(65*(1:(nrow(x)-1))/nrow(x))], adjustcolor("black",.5) )
+    cols <- c( adjustcolor(rainbow(64),.4)[1+floor(65*(1:(nrow(x)-1))/nrow(x))], adjustcolor("black",.8) )
     pairs( x, col=cols, pch=20, cex=c(rep(.5,nrow(x)-1),2), gap=0 )

### 2014-12-06 mcmc-model.R

@@ -120,6 +120,17 @@ model <- new( "contextMCMC",
              invocation=invocation
          )
 
+# set this up so that we can call likfun again in the future, directly
+likfun.env <- new.env()
+assign("use.par",use.par,envir=likfun.env)
+assign("params",params,envir=likfun.env)
+assign("prior.config",prior.config,envir=likfun.env)
+assign("genmatrix",model@genmatrix,envir=likfun.env)
+assign("projmatrix",model@projmatrix,envir=likfun.env)
+assign("counts",model@counts,envir=likfun.env)
+environment(model@likfun) <- likfun.env
+
+
 
 save(model,file=opt$outfile)
 

### 2014-12-07 simulation.Rmd
.

@@
                 counts=model@counts)
+        true.resids = computeresids(true.model,
+                in_longwin=in_longwin,
+                in_shortwin=in_shortwin,
+                in_leftwin=floor((in_longwin-in_shortwin)/2),
+                counts=model@counts)
+        resids$sim.expected <- true.resids$expected
         pandoc.table(top_n_bottom(resids, half_n_show), style="rmarkdown")
@@
         outfac <- factor(resids$outpat)
-        with(resids, plot(observed, expected, xlab='observed', ylab='expected', col=rainbow(nlevels(outfac))[as.numeric(outfac)], pch=20 ) )
+        with(resids, {
+                plot(observed, expected, xlab='observed', ylab='expected', col=rainbow(nlevels(outfac))[as.numeric(outfac)], pch=20, ylim=range(expected,sim.expected) ) 
+                # points(observed, sim.expected, xlab='observed', ylab='expected', col=adjustcolor(rainbow(nlevels(outfac))[as.numeric(outfac)],.5), pch=1, cex=.75 ) 
+                segments(x0=observed, y0=expected, y1=sim.expected, col=adjustcolor(rainbow(nlevels(outfac))[as.numeric(outfac)],.5) )
+            } )
         abline(0,1)
-        with(resids, plot(observed, expected, xlab='observed', ylab='expected', log='xy', col=rainbow(nlevels(outfac))[as.numeric(outfac)], pch=20 ) )
+        with(resids, {
+                plot(observed, expected, xlab='observed', ylab='expected', log='xy', col=rainbow(nlevels(outfac))[as.numeric(outfac)], pch=20, ylim=pmax(1e-4,range(expected,sim.expected)) ) 
+                segments(x0=observed, y0=expected, y1=sim.expected, col=adjustcolor(rainbow(nlevels(outfac))[as.numeric(outfac)],.5) )
+            } )
         abline(0,1)
@@
 counttab <- counts(model)
+true.model <- model
+true.model@mutrates <- simseq.config$tip$mutrates
+true.model@selcoef <- simseq.config$tip$selcoef
+if (length(true.model@params)>0) { true.model@params <- simseq.config$tip$fixfn.params }
 cat(model@invocation,"\n")
@@
     cat("convergence: ", model@results$convergence,
+        "\nnumber of iterations: ", model@results$counts,
         "\nmessage: ", model@results$message,
@@
 time <- as.numeric(simseq.opt$tlen)
+timevec <- c( rep(as.numeric(simseq.opt$tlen),nmuts(model)), rep(1,length(coef(model))-nmuts(model)) )
 mr_compare <- data.frame( 
-    fit=c( model@mutrates / time, model@selcoef, model@params), 
+    fit=coef(model)/timevec,
+    simulated=coef(true.model)/timevec,
     stringsAsFactors=FALSE )
-simvalues <- data.frame(
-        names=c(
-            mutnames(simseq.config$tip$mutpats), 
-            selnames(simseq.config$tip$selpats), 
-            names(simseq.config$tip$fixfn.params) ),
-        simulated=unlist(c(
-            simseq.config$tip$mutrates,
-            simseq.config$tip$selcoef,
-            simseq.config$tip$fixfn.params ) )
-)
-mr_compare$simulated <- simvalues$simulated[match(rownames(mr_compare),simvalues$names)]
-mr_compare$names <- factor(rownames(mr_compare), levels=rownames(mr_compare))
 pandoc.table(mr_compare, style="rmarkdown")
+mr_compare$names <- names(coef(model))
 ```
@@
 ```{r, echo=FALSE, results="asis", fig.align='center'}
-timevec <- c( rep(as.numeric(simseq.opt$tlen),nmuts(model)), rep(1,nrow(mr_compare)-nmuts(model)) )
 f <- function (x) { model@likfun(x*timevec) }

### 2014-12-07 simulation.Rmd
added JSON line to top of '.counts' file that records the left offset ('leftwin') and added more tests to testing-code; run-Rscript-test.sh runs these.

@@
     else {
-        resids = computeresids(model,
+        in_leftwin <- floor((in_longwin-in_shortwin)/2)
+        resids <- computeresids(model,
                 in_longwin=in_longwin,
                 in_shortwin=in_shortwin,
-                in_leftwin=floor((in_longwin-in_shortwin)/2),
-                counts=model@counts)
-        true.resids = computeresids(true.model,
-                in_longwin=in_longwin,
-                in_shortwin=in_shortwin,
-                in_leftwin=floor((in_longwin-in_shortwin)/2),
-                counts=model@counts)
-        resids$sim.expected <- true.resids$expected
+                in_leftwin=in_leftwin,
+                counts=model@counts, 
+                overlapping=TRUE)
+        true.expected <- projectcounts(
+                fitted( true.model, longwin=longwin(model), shortwin=shortwin(model), leftwin=leftwin(model) ),
+                in_leftwin, in_shortwin, in_longwin, 
+                overlapping=TRUE )
+        resids$sim.expected <- true.expected@counts[ cbind(resids$inpat,resids$outpat) ]
         pandoc.table(top_n_bottom(resids, half_n_show), style="rmarkdown")
@@
 counttab <- counts(model)
+time <- as.numeric(simseq.opt$tlen)
+timevec <- c( rep(as.numeric(simseq.opt$tlen),nmuts(model)), rep(1,length(coef(model))-nmuts(model)) )
 true.model <- model
-true.model@mutrates <- simseq.config$tip$mutrates
+true.model@mutrates <- simseq.config$tip$mutrates*time
 true.model@selcoef <- simseq.config$tip$selcoef
+mr_compare <- data.frame( 
+    fit=coef(model)/timevec,
+    simulated=coef(true.model)/timevec,
+    stringsAsFactors=FALSE )
 if (length(true.model@params)>0) { true.model@params <- simseq.config$tip$fixfn.params }
@@
 ```{r, echo=FALSE, results="asis"}
-time <- as.numeric(simseq.opt$tlen)
-timevec <- c( rep(as.numeric(simseq.opt$tlen),nmuts(model)), rep(1,length(coef(model))-nmuts(model)) )
-mr_compare <- data.frame( 
-    fit=coef(model)/timevec,
-    simulated=coef(true.model)/timevec,
-    stringsAsFactors=FALSE )
 pandoc.table(mr_compare, style="rmarkdown")

### 2014-12-07 input-output.R

@@ -10,16 +10,24 @@ source(paste(.PATH,"/helper-fns.R",sep=''))
 
 require(jsonlite)
 
-read.counts <- function (infile,leftwin,bases,longpats,shortpats) {
+read.counts <- function (infile,leftwin,bases,longpats,shortpats,skip=0) {
     # read in a file of counts of the following form:
+    #     # leftwin=1
     #        taxon1  taxon2 ... count
-    #   1      AAAA      AA ...    19
-    #   2      CAAA      AA ...     2
-    #   3      GAAA      AA ...     6
-    #   4      TAAA      AA ...     3
+    #         AAAA      AA ...    19
+    #         CAAA      AA ...     2
+    #         GAAA      AA ...     6
+    #         TAAA      AA ...     3
     # ... and convert it to a 'tuplecounts' object
     # optionally passing in the orderings of the rows and columns
-    count.table <- read.table(infile,header=TRUE,stringsAsFactors=FALSE)
+    count.paramstring <- scan(infile,what='char',nlines=1,sep="\n")
+    count.params <- if (substr(count.paramstring,1,1)=="#") {
+            skip <- 1
+            fromJSON(gsub("^#*","",count.paramstring),simplifyMatrix=FALSE)
+        } else { NULL }
+    if (missing(leftwin)) { leftwin <- count.params$leftwin }
+    if (is.null(leftwin)) { stop("leftwin not specified.") }
+    count.table <- read.table(infile,header=TRUE,stringsAsFactors=FALSE,skip=skip)
     longwin <- nchar( count.table[1,1] )
     shortwin <- nchar( count.table[1,2] )
     taxa <- colnames(count.table)[-ncol(count.table)]

### 2014-12-07 testing-code/sim-short-ising.R

@@ -2,11 +2,11 @@
 
 require(expm)
 require(parallel)
-numcores<-as.numeric(scan(pipe("cat /proc/cpuinfo | grep processor | tail -n 1 | awk '{print $3}'")))+1
 
 # Run some tests!!
 source("../sim-context-fns.R")
 source("../context-inference-fns.R")
+numcores<-getcores()
 
 bases <- c("X","O")
 
@@ -31,9 +31,10 @@ selcoef <- c(-.5,.5)
 
 initseq <- rinitseq(seqlen,bases)
 simseqs <- mclapply( 1:1000, function (k) 
-        simseq( seqlen, tlen, intiseq=initseq, mutpats=mutpats, selpats=selpats, mutrates=mutrates, selcoef=selcoef, bases=bases, fixfn=fixfn ) 
+        simseq( seqlen, tlen, initseq=initseq, mutpats=mutpats, selpats=selpats, mutrates=mutrates, selcoef=selcoef, bases=bases, fixfn=fixfn, count.trans=TRUE ) 
     , mc.cores=numcores)
-full.genmatrix <- makegenmatrix( mutpats, selpats, patlen=seqlen, boundary="none", bases=bases, fixfn=fixfn )
+# note that simseq simulates WRAPPED sequences
+full.genmatrix <- makegenmatrix( mutpats, selpats, murates=mutrates, selcoef=selcoef, patlen=seqlen, boundary="wrap", bases=bases, fixfn=fixfn )
 require(expm)
 full.transmatrix <- expm( tlen*(as.matrix(full.genmatrix)-diag(rowSums(full.genmatrix))) )
 stopifnot( all( abs( rowSums(full.transmatrix) - rep(1.0,nrow(full.transmatrix)) ) < 1e-8 ) )
@@ -47,7 +48,7 @@ resids$p <- pbinom( resids$observed, size=length(simseqs), prob=expected.freqs )
 cat("Beginning from ", as.character(initseq), ":\n")
 print(resids)
 
-if( !( all( abs(abs(resids$p-1/2)-1/2)>1/2^(seqlen+1) ) ) ) {
-    stop("Some high residuals there.")
+if ( min( pbeta( min(resids$p), 1,nrow(resids) ) , pbeta( min(1-resids$p), 1,nrow(resids) ) ) < .01 ) {
+    stop("Error: some high residuals there.")
 }
 

### 2014-12-08 simulation.Rmd
removed --leftwin option from fit-model, since that is stored in .counts file now

@@
     subseq <- seq(1,nrow(model@results$batch),length.out=min(1000,nrow(model@results$batch)))
-    x <- model@results$batch[subseq,]
+    x <- sweep( model@results$batch[subseq,], 2, timevec[model@results$use.par], "/" )
     colnames(x) <- rownames(mr_compare)[model@results$use.par]
@@
     for (k in unique(param_sizes)) {
-        print( { 
-            matplot( model@results$blen*subseq, x[,param_sizes==k], type='l', col=1:5, lty=1:6, xlab='number of MCMC iterations', ylab='parameter values' )
+        matplot( model@results$blen*subseq, x[,param_sizes==k], type='l', col=1:5, lty=1:6, 
+            xlab='number of MCMC iterations', ylab='parameter values' )
         abline(h=mr_compare$simulated[model@results$use.par], col=1:5, lty=1:6)
         legend("topright",legend=colnames(x)[param_sizes==k],col=1:5,lty=1:6)
-        } )
     }
@@
     subseq <- seq(1,nrow(model@results$batch),length.out=min(1000,nrow(model@results$batch)))
-    x <- rbind( model@results$batch[subseq,], mr_compare$simulated[model@results$use.par] ) 
-    colnames(x) <- rownames(mr_compare)[model@results$use.par]
+    x <- rbind( x, mr_compare$simulated[model@results$use.par] ) 
     # color by time

### 2014-12-08 json-ising/test-workflow.sh

@@ -1,0 +1,82 @@
+#!/bin/bash
+
+set -eu
+set -o pipefail
+
+# simulate up some sequence for testing
+Rscript ../sim-seq.R -c ising-model.json -t .1 -s 100000 -d simseqs -o test-ising-123456.RData
+
+# and count the Tmers
+Rscript ../count-seq.R -i simseqs/test-ising-123456.RData -w 3 -s 1 -l 1
+Rscript ../count-seq.R -i simseqs/test-ising-123456.RData -w 9 -s 5 -l 2
+
+# precompute generator matrices:
+#   width-3
+Rscript ../make-genmat.R -c genmatrices/complete.json -w 3
+Rscript ../make-genmat.R -c genmatrices/complete.json -w 4
+Rscript ../make-genmat.R -c genmatrices/complete.json -w 9
+
+echo "check simulated model matches expected"
+../templated-Rmd.sh ../testing-code/check-sim.Rmd simseqs/test-ising-123456.RData genmatrices/genmatrix-3-complete.RData 
+
+echo "fitting the model"
+Rscript ../fit-model.R -i simseqs/test-ising-123456-3-root-1-tip-l1.counts -m genmatrices/genmatrix-3-complete.RData -j 54321
+
+echo "computing residuals"
+Rscript ../compute-resids.R -i simseqs/test-ising-123456-3-root-1-tip-l1-genmatrix-3-complete-54321.RData
+
+echo "look at results"
+../templated-Rmd.sh ../simulation.Rmd simseqs/test-ising-123456-3-root-1-tip-l1-genmatrix-3-complete-54321.RData simseqs/test-ising-123456.RData
+
+ 
+echo "mcmc also"
+Rscript ../mcmc-model.R -i simseqs/test-ising-123456-3-root-1-tip-l1-genmatrix-3-complete-54321.RData -c ising-model.json -b 3 -j 1111
+Rscript ../mcmc-model.R -i simseqs/test-ising-123456-3-root-1-tip-l1-genmatrix-3-complete-54321.RData -c ising-model.json -b 100 -j 2222
+Rscript ../mcmc-model.R -i simseqs/test-ising-123456-3-root-1-tip-l1-genmatrix-3-complete-54321.RData -c ising-model.json -b 1000 -j 3333
+
+echo "look at results"
[…]

### 2014-12-08 model-fitting/workflow.sh

@@ -19,9 +19,9 @@ Rscript ../make-genmat.R -c genmatrices/singlebase.json -w 2 &
 Rscript ../make-genmat.R -c genmatrices/dualbases.json -w 4
 
 # fit a simple model: width 4 with all one-base transitions on 1+2+1 Tmers
-Rscript ../fit-model.R -i sim-cpg-123456.4.2.l1.counts -l 1 -m genmatrices/genmatrix-4-singlebase.RData -o simple-fit.RData
+Rscript ../fit-model.R -i sim-cpg-123456.4.2.l1.counts -m genmatrices/genmatrix-4-singlebase.RData -o simple-fit.RData
 # and a more complex one: width 4 with all two-base transitions
-Rscript ../fit-model.R -i sim-cpg-123456.4.2.l1.counts -l 1 -m genmatrices/genmatrix-4-dualbases.RData -o complex-fit.RData
+Rscript ../fit-model.R -i sim-cpg-123456.4.2.l1.counts -m genmatrices/genmatrix-4-dualbases.RData -o complex-fit.RData
 
 # first look at simple model residuals on 1+2+1 Tmers (what it was fit with)
 Rscript ../compute-resids.R -i simple-fit.RData

### 2014-12-08 simulation.Rmd
.

@@
     if (NROW(resids)>0) { 
-        hist(resids$z, xlab="z-score", ylab='') 
+        hist(resids$z, xlab="z-score", ylab='', main="z score") 
         outfac <- factor(resids$outpat)
@@
     "long window size:", longwin(model@counts),
-    "short window size:", shortwin(model@counts),
-    "left window size:", leftwin(model@counts),
+    "\nshort window size:", shortwin(model@counts),
+    "\nleft window size:", leftwin(model@counts),
     "\ntotal counts:", sum(counttab),
@@
 
-The likelihood surface about the fitted and simulated values:
+The likelihood surface about the fitted and simulated values (red are the values simulated under, black is the fitted values):
 ```{r, echo=FALSE, results="asis", fig.align='center'}

### 2014-12-08 count-seq.R

@@ -76,10 +76,10 @@ counts <- counttrans.list( list(longpats,shortpats)[c(1,rep.int(2,length(opt$sho
    shift=opt$shift )
 if (opt$RData) {
     outfile <- gsub("\\.counts$",".RData",opt$outfile)
-    cat("Writing to:" outfile, "\n")
+    cat("Writing to:", outfile, "\n")
     save( counts, file=outfile )
 } else if (opt$shift==0) {
-    cat("Writing to:" opt$outfile, "\n")
+    cat("Writing to:", opt$outfile, "\n")
     cframe <- countframe( counts )
     cat( paste('# { "leftwin" : ', opt$leftwin, '}\n', sep=''), file=opt$outfile )
     write.table(cframe, file=opt$outfile, row.names=FALSE, sep='\t', quote=FALSE, append=TRUE)
@@ -87,7 +87,7 @@ if (opt$RData) {
     for (k in 1:opt$shift) {
         cframe <- countframe( counts[[k]] )
         outfile <- paste(opt$outfile,k,sep='.')
-        cat("Writing to:" outfile, "\n")
+        cat("Writing to:", outfile, "\n")
         cat( paste('#', toJSON( list( leftwin=opt$leftwin, shift=opt$shift, offset=k ), auto_unbox=TRUE ), "\n" ), file=outfile )
         # cat( paste('# { "leftwin" : ', opt$leftwin, '}\n', sep=''), file=outfile )
         write.table(cframe, file=outfile, row.names=FALSE, sep='\t', quote=FALSE, append=TRUE)

### 2014-12-08 json-ising/test-workflow.sh

@@ -42,6 +42,7 @@ echo "mcmc on offset counts"
 Rscript ../count-seq.R -i simseqs/test-ising-123456.RData -w 4 -s 2 -l 1 --shift 4
 Rscript ../fit-model.R -i simseqs/test-ising-123456-4-root-2-tip-l1-shift4.counts.1 -m genmatrices/genmatrix-4-complete.RData -j 10101
 Rscript ../mcmc-model.R -i simseqs/test-ising-123456-4-root-2-tip-l1-shift4-genmatrix-4-complete-10101.RData -c ising-model.json -b 1000 -j 20202
+../templated-Rmd.sh ../simulation.Rmd simseqs/test-ising-123456-4-root-2-tip-l1-shift4-genmatrix-4-complete-10101-mcmc-20202.RData simseqs/test-ising-123456.RData
 
 echo "now on a larger T-mer (4/2)"
 Rscript ../count-seq.R -i simseqs/test-ising-123456.RData -w 4 -s 2 -l 1
@@ -56,8 +57,6 @@ Rscript ../fit-model.R -i simseqs/test-ising-123456-4-root-2-tip-l1.counts -m ge
 # Rscript ../count-seq.R -i test-ising-123456.RData -w 6 -s 2 -l 2
 # Rscript ../make-genmat.R -c genmatrices/complete.json -w 6
 # Rscript ../fit-model.R -i test-ising-123456.6.2.l2.counts -m genmatrices/genmatrix-6-complete.RData -j 54321
- 
-exit 0;
  
 # ###
 # # and for the constrained model

### 2014-12-09 simulation.Rmd
simulation.Rmd deals with fixed params correctly

@@
 true.model@selcoef <- simseq.config$tip$selcoef
+if (length(true.model@params)>0) { true.model@params <- unlist(simseq.config$tip$fixfn.params) }
 mr_compare <- data.frame( 
@@
     simulated=coef(true.model)/timevec,
+    fixed=!model@results$use.par,
     stringsAsFactors=FALSE )
-if (length(true.model@params)>0) { true.model@params <- simseq.config$tip$fixfn.params }
 cat(model@invocation,"\n")
@@
 ```{r, echo=FALSE, results="asis", fig.align='center'}
-f <- function (x) { model@likfun(x*timevec) }
-for (k in 1:nrow(mr_compare)) {
-    pvals <- seq( .9*min(mr_compare[k,c("fit","simulated")]), 1.1*max(mr_compare[k,c("fit","simulated")]), length.out=20 )
+f <- function (x) { model@likfun((x*timevec)[model@results$use.par]) }
+for (k in (1:nrow(mr_compare))[model@results$use.par]) {
+    pvals <- seq( .9*min(mr_compare[k,c("fit","simulated")]), 1.1*max(mr_compare[k,c("fit","simulated")]), length.out=10 )
     inds <- 1:nrow(mr_compare)

### 2014-12-09 simulation.Rmd

@@
 pandoc.table(mr_compare, style="rmarkdown")
-mr_compare$names <- names(coef(model))
+mr_compare$names <- factor(names(coef(model)),levels=names(coef(model)))  # factor explicitly to get ggplot to put bars in order
 ```

### 2014-12-10 simulation.Rmd
added tlen scaling to initial parameters for model-fit and mcmc

@@
 }
+finite_range <- function (...) { x <- unlist( list(...) ); range(x[is.finite(x)&!is.na(x)]) }
 counts_plot <- function (resids) {
@@
         with(resids, {
-                plot(observed, expected, xlab='observed', ylab='expected', col=rainbow(nlevels(outfac))[as.numeric(outfac)], pch=20, ylim=range(expected,sim.expected) ) 
+                plot(observed, expected, xlab='observed', ylab='expected', col=rainbow(nlevels(outfac))[as.numeric(outfac)], pch=20, ylim=finite_range(expected,sim.expected) ) 
                 # points(observed, sim.expected, xlab='observed', ylab='expected', col=adjustcolor(rainbow(nlevels(outfac))[as.numeric(outfac)],.5), pch=1, cex=.75 ) 
@@
         with(resids, {
-                plot(observed, expected, xlab='observed', ylab='expected', log='xy', col=rainbow(nlevels(outfac))[as.numeric(outfac)], pch=20, ylim=pmax(1e-4,range(expected,sim.expected)) ) 
+                plot(observed, expected, xlab='observed', ylab='expected', log='xy', col=rainbow(nlevels(outfac))[as.numeric(outfac)], pch=20, ylim=pmax(1e-4,finite_range(expected,sim.expected)) ) 
                 segments(x0=observed, y0=expected, y1=sim.expected, col=adjustcolor(rainbow(nlevels(outfac))[as.numeric(outfac)],.5) )
@@
     l2vals <- sapply( pvals, function (x) { f( c(mr_compare$simulated,x)[inds] ) } )
-    plot( pvals, lvals, type='b', main=paste("likelihood surface for", mr_compare$name[k]), ylab="log-likelihood", xlab='value', ylim=range(lvals,l2vals) )
+    plot( pvals, lvals, type='b', main=paste("likelihood surface for", mr_compare$name[k]), ylab="log-likelihood", xlab='value', ylim=finite_range(lvals,l2vals) )
     lines( pvals, l2vals, col='red', type='b' )

### 2014-12-10 count-seq.R

@@ -79,16 +79,17 @@ if (opt$RData) {
     cat("Writing to:", outfile, "\n")
     save( counts, file=outfile )
 } else if (opt$shift==0) {
-    cat("Writing to:", opt$outfile, "\n")
+    outfile <- opt$outfile
+    cat("Writing to:", outfile, "\n")
     cframe <- countframe( counts )
-    cat( paste('# { "leftwin" : ', opt$leftwin, '}\n', sep=''), file=opt$outfile )
+    cat( paste('#', toJSON( list( leftwin=opt$leftwin, longwin=opt$longwin, shortwin=opt$shortwin, shift=opt$shift, offset=k ), auto_unbox=TRUE ), "\n" ), file=outfile )
     write.table(cframe, file=opt$outfile, row.names=FALSE, sep='\t', quote=FALSE, append=TRUE)
 } else {
     for (k in 1:opt$shift) {
         cframe <- countframe( counts[[k]] )
         outfile <- paste(opt$outfile,k,sep='.')
         cat("Writing to:", outfile, "\n")
-        cat( paste('#', toJSON( list( leftwin=opt$leftwin, shift=opt$shift, offset=k ), auto_unbox=TRUE ), "\n" ), file=outfile )
+        cat( paste('#', toJSON( list( leftwin=opt$leftwin, longwin=opt$longwin, shortwin=opt$shortwin, shift=opt$shift, offset=k ), auto_unbox=TRUE ), "\n" ), file=outfile )
         # cat( paste('# { "leftwin" : ', opt$leftwin, '}\n', sep=''), file=outfile )
         write.table(cframe, file=outfile, row.names=FALSE, sep='\t', quote=FALSE, append=TRUE)
     }

### 2014-12-10 fit-tree-model.R

@@ -14,18 +14,17 @@ Parameters whose scale is set to zero *will be regarded as fixed.* \
 option_list <- list(
     # input/output
         make_option( c("-i","--infile"), type="character", help="Input file with tuple counts, tab-separated, with header 'reference', 'derived', 'count'. [default, looks in basedir]" ),
-        make_option( c("-l","--leftwin"), type="integer", help="Size of left-hand context for count file." ),
+        make_option( c("-c","--configfile"), type="character", help="File with model configuration (optional if genmatrix provided, not on a tree)."),
         make_option( c("-o","--outfile"), type="character", help="File to save results to.  [default: base of infile + base of genmatrix + jobid + .RData]"),
         make_option( c("-u","--basedir"), type="character", default=NULL, help="Directory to put output in. [default: same as infile]"),
-        make_option( c("-c","--configfile"), type="character", help="File with model configuration (optional if genmatrix provided, not on a tree)."),
-        make_option( c("-w","--longwin"), type="numeric", help="Size of long window. (optional if genmatrix specified, not on a tree)"),
-        make_option( c("-s","--shortwin"), type="integer", help="Size of short window. (optional if genmatrix specified, not on a tree)" ),
         make_option( c("-x","--maxit"), type="integer", default=100, help="Number of iterations of optimization to run for. [default=%default]"),
-        make_option( c("-j","--jobid"), type="character", default=formatC(1e6*runif(1),width=6,format="d",flag="0"), help="Unique job id. [default random]")
+        make_option( c("-j","--jobid"), type="character", default=formatC(1e6*runif(1),width=6,format="d",flag="0"), help="Unique job id. [default random]"),
+        make_option( c("-z","--seed"), type="integer", help="Seed for pseudorandom number generator; an integer. [default: does not meddle]")
     )
 opt <- parse_args(OptionParser(option_list=option_list,description=usage))
 if (is.null(opt$infile)) { stop("No input file.  Run\n  fit-tree-model.R -h\n for help.\n") }
[…]

### 2014-12-10 json-ising/plus-workflow.sh

@@ -4,6 +4,7 @@ set -eu
 set -o pipefail
 
 # simulate up some sequence for testing
+mkdir -p simplus
 Rscript ../sim-seq.R -c ising-model-plus.json -t .1 -s 100000 -o simplus/plus-ising-01.RData
 
 # and count the Tmers
@@ -14,16 +15,14 @@ Rscript ../count-seq.R -i simplus/plus-ising-01.RData -w 4 -s 2 -l 1
 Rscript ../make-genmat.R -c ising-model-plus.json -w 4 -o genmatrices/genmatrix-4-plus.RData
 
 echo "fitting the model"
-Rscript ../fit-model.R -i simplus/plus-ising-01-4-root-2-tip-l1-shift0.counts -m genmatrices/genmatrix-4-plus.RData -j 456
-
-echo "computing residuals"
-Rscript ../compute-resids.R -i simplus/plus-ising-01-4-root-2-tip-l1-shift0-genmatrix-4-plus-456.RData
+Rscript ../fit-model.R -i simplus/plus-ising-01-4-root-2-tip-l1-shift0.counts -t .1 -m genmatrices/genmatrix-4-plus.RData -j 456 -c ising-model-plus.json
 
 echo "look at results"
 ../templated-Rmd.sh ../simulation.Rmd simplus/plus-ising-01-4-root-2-tip-l1-shift0-genmatrix-4-plus-456.RData simplus/plus-ising-01.RData
 
 echo "mcmc also"
-Rscript ../mcmc-model.R -i simplus/plus-ising-01-4-root-2-tip-l1-shift0-genmatrix-4-plus-456.RData -c ising-model-plus.json -b 1000 -j 789
+Rscript ../mcmc-model.R -i simplus/plus-ising-01-4-root-2-tip-l1-shift0-genmatrix-4-plus-456.RData -c ising-model-plus.json -t .1 -b 10
+Rscript ../mcmc-model.R -i simplus/plus-ising-01-4-root-2-tip-l1-shift0-genmatrix-4-plus-456.RData -c ising-model-plus.json -t .1 -b 1000 -j 789
 
 echo "look at results"
 ../templated-Rmd.sh ../simulation.Rmd simplus/plus-ising-01-4-root-2-tip-l1-shift0-genmatrix-4-plus-456-mcmc-789.RData simplus/plus-ising-01.RData

### 2014-12-10 run-analysis.sh

@@ -41,7 +41,7 @@ Rscript ../count-seq.R -i $SIMFILE -w $LONGWIN -s $SHORTWIN -l $LEFTWIN -o $COUN
 
 FITFILE="$BASEDIR/$BASE-fit-${LONGWIN}-${SHORTWIN}-l${LEFTWIN}-shift0.RData"
 echo "Fitting a model, to $FITFILE ."
-Rscript ../fit-model.R -i $COUNTFILE -m $GENMAT -c $MODEL -o $FITFILE
+Rscript ../fit-model.R -i $COUNTFILE -m $GENMAT -c $MODEL -t $TLEN -o $FITFILE
 
 echo "Look at results."
 ../templated-Rmd.sh ../simulation.Rmd $FITFILE $SIMFILE
@@ -50,4 +50,4 @@ echo "Look at results."
 exit 0
 
 echo "mcmc also"
-Rscript ../mcmc-model.R -i $FITFILE -c $MODEL -b 1000 -j 3333
+Rscript ../mcmc-model.R -i $FITFILE -c $MODEL -t $TLEN -b 1000 -j 3333

### 2014-12-11 simulation.Rmd
informative message if missing starting parameters

@@
 param_log <- log(abs(mr_compare$simulated)+1e-4)
-param_sizes <- cut( param_log, breaks=seq(min(param_log), by=2, length.out=1+ceiling(diff(range(param_log))/2)), include.lowest=TRUE )
+param_sizes <- cut( param_log, breaks=seq(min(param_log), by=2, length.out=max(2,1+ceiling(diff(range(param_log))/2))), include.lowest=TRUE )
 for (k in unique(param_sizes)) {

### 2014-12-11 context-inference-fns.R

@@ -300,26 +300,45 @@ setMethod("%*%", signature=c(x="tuplecounts",y="dgTMatrix"), definition=function
 setMethod("%*%", signature=c(x="TsparseMatrix",y="tuplecounts"), definition=function (x,y) { x %*% y@counts } )
 setMethod("%*%", signature=c(x="tuplecounts",y="TsparseMatrix"), definition=function (x,y) { x@counts %*% y } )
 
+setClass("contextModel",
+        # A model of mutation; one for each branch.
+         representation(
+                       genmatrix="genmatrix",
+                       projmatrix="Matrix",
+                       mutrates="numeric",
+                       selcoef="numeric",
+                       params="numeric"
+                    )
+            )
 
 setClass("context",
-         representation(counts="tuplecounts",
-                        genmatrix="genmatrix",
-                        mutrates="numeric",
-                        selcoef="numeric",
-                        params="numeric",
-                        projmatrix="Matrix",
-                        likfun="function",
-                        results="list",
-                        invocation="character"
-                    )
+         representation(
+                       counts="tuplecounts",
+                       likfun="function",
+                       results="list",
+                       invocation="character"
+                    ),
+            contains="contextModel"
          )
 
 setClass("contextMCMC", representation(
-                                       mutrates.prior="numeric",
-                                       selcoef.prior="numeric",
-                                       fixfn.params.prior="numeric"
-                                       ),
+                       mutrates.prior="numeric",
+                       selcoef.prior="numeric",
+                       fixfn.params.prior="numeric"
+                   ),
          contains="context")
+
+setClass("contextTree", representation(
+                       counts="tuplecounts",
+                       tree="list",  # as in 'ape'
+                       initfreqs="numeric",
+                       models="list",  # list of 'contextModel's
+                       likfun="function",
+                       results="list",
+                       invocation="character"
+                   )
+         )
+
 
 setMethod("dimnames", signature=c(x="context"), definition=function (x) { dimnames(x@counts) } )
 setMethod("counts", signature=c(x="context"), definition=function (x) { counts(x@counts) } )

### 2014-12-11 fit-model.R

@@ -49,6 +49,14 @@ options(error = print.and.dump)
 # read configuration
 init.config <- read.config(opt$configfile)  # returns NULL if it is NULL
 if (!is.null(init.config$tree)) { warning("Appears to be a tree-config file; should be using fit-tree-model.R?") }
+if ( is.null(init.config$mutrates) || 
+        is.null(init.config$mutrates.scale) || 
+        is.null(init.config$selcoef) || 
+        is.null(init.config$selcoef.scale) || 
+        is.null(init.config$fixfn.params) || 
+        is.null(init.config$fixfn.params.scale) ) {
+    stop("Need to specify starting parameter values and scalings in config file.")
+}
 
 # load generator matrix
 stopifnot(file.exists(opt$gmfile))

### 2014-12-11 json-tasep/workflow.sh

@@ -20,7 +20,7 @@ echo "check simulated model matches expected"
 ../templated-Rmd.sh ../testing-code/check-sim.Rmd simseqs/sim-tasep-123456.RData genmatrices/genmatrix-4-complete.RData
 
 echo "fitting, width 4"
-Rscript ../fit-model.R -i simseqs/sim-tasep-123456-4-root-2-tip-l1.counts -m genmatrices/genmatrix-4-complete.RData -j 54321
+Rscript ../fit-model.R -i simseqs/sim-tasep-123456-4-root-2-tip-l1.counts -m genmatrices/genmatrix-4-complete.RData -c tasep-model.json -t 0.1 -j 54321
 
 echo "computing residuals"
 Rscript ../compute-resids.R -i simseqs/sim-tasep-123456-4-root-2-tip-l1-genmatrix-4-complete-54321.RData

### 2015-05-22 writeups/context-dependence.tex
tests for tree transition probabilities
Adding Saunders 2007 reference
.

@@
 \usepackage{natbib}
-\usepackage{hyperref}
+\usepackage[colorlinks=false]{hyperref}
 \usepackage{bm}
@@
 \begin{document}
+
+\title{Inference using context-dependent models of mutation}
+\author{}
+
+\maketitle
+
+\begin{abstract}
+\end{abstract}
 
@@
 however, all these methods are quite computationally intensive.
+
+Model with fixed flanking \citet{saunders2007insights}.
 
@@
 
+Let $t(w)$ be the length of the branch above node $w$, for $w \neq \rho$,
+and denote by $G^T$ the transpose of $G$.
+
 \begin{enumerate}
@@
         \begin{align}
-          M(u_j)_{x,(y,z)} = \left( e^{t(w_1) G} M(w_1) \right)_{x,y} \left( e^{t(w_2) G} M(w_2)_{x,z} \right)_{x,z} ,
-        \end{align}
-        where $t(w_1)$ and $t(w_2)$ are the lengths of the respective branches.
+          M(u_j)_{x,(y,z)} = \left( e^{t(w_1) G} M(w_1) \right)_{x,y} \left( e^{t(w_2) G} M(w_2)_{x,z} \right)_{x,z} .
+        \end{align}
     \end{enumerate}
@@
     \begin{align}
-      M(\rho)_{x,y} = \pi(x) \left( e^{t(w(\rho)) G} M(w(\rho)) \right)_{x,y}
+      M(\rho)_{x,y} = \pi(x) \left( e^{t(u(\rho)) G} M(w(\rho)) \right)_{x,y}
     \end{align}
@@
         \begin{align}
-          M(v_k)_{x,(y,z)} = \left( e^{t(v_k) G} M(v_{k-1}) \right)_{y,x} \left( e^{t(u(v_k)) G} M(u(v_{k-1})) \right)_{x,z}  .
+          M(v_k)_{x,(y,z)} = \left( e^{t(v_k) G^T} M(v_{k-1}) \right)_{y,x} \left( e^{t(u(v_k)) G} M(u(v_{k-1})) \right)_{x,z}  .
         \end{align}
@@
     \begin{align}
-      M(a)_{x,y} = \left( e^{t(a) G} M(v_{\ell-1}) \right)_{x,y}  .
+      M(a)_{x,y} = \left( e^{t(a) G^T} M(v_{\ell-1}) \right)_{x,y}  .
     \end{align}

### 2015-06-11 testing-code/check-sim.Rmd
.

@@
         "time",
-        mutnames(simseq.config$tip$mutpats),
-        selnames(simseq.config$tip$selpats),
+        mutnames(simseq.config$tip$mutpats,pretty=TRUE),
+        selnames(simseq.config$tip$selpats,pretty=TRUE),
         names(simseq.config$tip$fixfn.params)

### 2015-06-11 shape/sim-test-sequence.sh

@@ -15,8 +15,8 @@ MODELFILE=${MODEL}.json
 SEED=$(printf "%06.0f" $RANDOM)
 
 BASEDIR="testseq"
-mkdir -p $BASEDIR
 GMDIR="genmatrices"
+mkdir -p $BASEDIR
 mkdir -p $GMDIR
 
 echo "Simulating from ${MODEL} ."
@@ -44,5 +44,12 @@ echo "check simulated model matches expected"
 echo "fit a model"
 FITFILE=$(echo $SIMFILE | sed -e "s/.RData/-${LONGWIN}-${SHORTWIN}-l${LEFTWIN}.fit.RData/")
 ls $COUNTFILE $MODELFILE $GENMAT && \
-    Rscript ../fit-model.R -i $COUNTFILE -t .01 --maxit 5 -c $MODELFILE -m $GENMAT -o $FITFILE
+    Rscript ../fit-model.R -i $COUNTFILE -t .2 --maxit 500 -c $MODELFILE -m $GENMAT -o $FITFILE
 
+echo "computing residuals"
+RESIDFILE=$(echo $SIMFILE | sed -e "s/.RData/-${LONGWIN}-${SHORTWIN}-l${LEFTWIN}.resid.tsv/")
+ls $FITFILE && \
+    Rscript ../compute-resids.R -i $FITFILE -o $RESIDFILE
+
+echo "look at results"
+../templated-Rmd.sh ../simulation.Rmd $FITFILE $SIMFILE

### 2015-06-11 simulation.Rmd
switched mutnames, selnames to use "names" attribute if present

--- 2014-12-09 simulation.Rmd
@@
 panderOptions("table.split.table", Inf)  # Let tables become arbitrarily wide rather than wrapping.
-source("context-inference-fns.R")
+source("context-inference-fns.R",chdir=TRUE)
 top_n_bottom <- function(df, k) { # Take the top and bottom k rows
@@
 
-```{r, echo=FALSE}
+```{r summary, echo=FALSE, results="asis"}
 cat(
-    "long window size:", longwin(model@counts),
-    "\nshort window size:", shortwin(model@counts),
-    "\nleft window size:", leftwin(model@counts),
-    "\ntotal counts:", sum(counttab),
-    "\ntime:", simseq.opt$tlen,
-    "\nseed:", formatC(simseq.opt$seed,format="d"),
+    "\n*  long window size:", longwin(model@counts),
+    "\n*  short window size:", shortwin(model@counts),
+    "\n*  left window size:", leftwin(model@counts),
+    "\n*  total counts:", sum(counttab),
+    "\n*  time:", simseq.opt$tlen,
+    "\n*  seed:", formatC(simseq.opt$seed,format="d"),
     "\n"
@@
 ## Result of fitting operation
-```{r, echo=FALSE}
+```{r convergence, echo=FALSE, results="asis"}
 if (class(model)=="context") {
-    cat("convergence: ", model@results$convergence,
-        "\nnumber of iterations: ", model@results$counts,
-        "\nmessage: ", model@results$message,
+    cat("\n* convergence: ", model@results$convergence,
+        "\n* number of iterations: ", model@results$counts,
+        "\n* message: ", model@results$message,
         "\n" )
@@
     cat(
-        "number of batches: ", model@results$nbatch,
-        "\nbatch length: ", model@results$blen,
-        "\nacceptance rate: ", model@results$accept,
-        "\nrun time: ", model@results$time['elapsed'],
+        "\n* number of batches: ", model@results$nbatch,
+        "\n* batch length: ", model@results$blen,
+        "\n* acceptance rate: ", model@results$accept,
+        "\n* run time: ", model@results$time['elapsed'],
         "\n" )
@@
 ## Fitted parameter values
-```{r, echo=FALSE, results="asis"}
-pandoc.table(mr_compare, style="rmarkdown")
+```{r fitted, echo=FALSE}
+pander(mr_compare)
 mr_compare$names <- factor(names(coef(model)),levels=names(coef(model)))  # factor explicitly to get ggplot to put bars in order
@@
 The corresponding plot, where bars are the fit values, and points are the simulated values:
-```{r, echo=FALSE, results="asis", fig.align='center'}
+```{r fit_plot, echo=FALSE, results="asis", fig.align='center'}
 param_log <- log(abs(mr_compare$simulated)+1e-4)
-param_sizes <- cut( param_log, breaks=seq(min(param_log), by=2, length.out=1+ceiling(diff(range(param_log))/2)), include.lowest=TRUE )
+param_sizes <- cut( param_log, breaks=seq(min(param_log), by=2, length.out=max(2,1+ceiling(diff(range(param_log))/2))), include.lowest=TRUE )
 for (k in unique(param_sizes)) {
@@
 The likelihood surface about the fitted and simulated values (red are the values simulated under, black is the fitted values):
-```{r, echo=FALSE, results="asis", fig.align='center'}
+```{r likelihood_surface, echo=FALSE, results="asis", fig.align='center'}
 f <- function (x) { model@likfun((x*timevec)[model@results$use.par]) }
@@
 ## residuals for 2-2 Tmers
-```{r, echo=FALSE, results="asis", fig.align='center'}
+```{r resids_2_2, echo=FALSE, results="asis", fig.align='center'}
 resids <- counts_table(in_longwin=2, in_shortwin=2, half_n_show=20)
@@
 ## residuals for 3-1 Tmers
-```{r, results="asis", fig.align='center'}
+```{r resids_3_1, results="asis", fig.align='center'}
 resids <- counts_table(in_longwin=3, in_shortwin=1, half_n_show=20)
@@
 ## residuals for 9-5 Tmers
-```{r, results="asis", fig.align='center'}
+```{r resids_9_5, results="asis", fig.align='center'}
 resids <- counts_table(in_longwin=9, in_shortwin=5, half_n_show=20)
@@
 # Trace plot of the MCMC run:
-```{r, results="asis", echo=FALSE, fig.width=12, fig.height=7, fig.align='center'}
+```{r mcmc_trace, results="asis", echo=FALSE, fig.width=12, fig.height=7, fig.align='center'}
 if (class(model)=="contextMCMC") {
@@
 # Pairwise marginal posteriors from MCMC, colored by time in MCMC run
-```{r, results="asis", echo=FALSE, fig.width=12, fig.height=12, fig.align='center'}
+```{r mcmc_posterior, results="asis", echo=FALSE, fig.width=12, fig.height=12, fig.align='center'}
 if (class(model)=="contextMCMC") {

### 2015-06-11 shape/sim-test-sequence.sh

@@ -25,8 +25,8 @@ SIMFILE="$BASEDIR/simseq-${MODEL}-seed-${SEED}.RData"
 Rscript ../sim-seq.R -c $MODELFILE -t .2 -s 10000 -m $SIMGENMAT -z $SEED -o $SIMFILE
 
 LONGWIN=5
-SHORTWIN=1
-LEFTWIN=2
+SHORTWIN=3
+LEFTWIN=1
 
 GENMAT="$GMDIR/${MODEL}-genmatrix-${LONGWIN}.RData"
 COUNTFILE=$(echo $SIMFILE | sed -e "s/.RData/-${LONGWIN}-${SHORTWIN}-l${LEFTWIN}.counts/")

### 2015-06-12 testing-code/check-sim.Rmd
moved all source()s to source(,chdir=TRUE)

@@
 panderOptions("table.split.table", Inf)  # Let tables become arbitrarily wide rather than wrapping.
-source("../context-inference-fns.R")
-source("../sim-context-fns.R")
+source("../context-inference-fns.R",chdir=TRUE)
+source("../sim-context-fns.R",chdir=TRUE)
 top_n_bottom <- function(df, k) { # Take the top and bottom k rows

### 2015-06-12 bcells/bcells-mcmc.R

@@ -30,8 +30,8 @@ if (!file.exists(infile)) { stop("Cannot read file ", infile) }
 basename <- gsub("-results.RData","",infile)
 
 scriptdir <- "../"
-source(paste(scriptdir,"context-inference-fns.R",sep=''))
-# source(paste(scriptdir,"sim-context-fns.R",sep=''))
+source(paste(scriptdir,"context-inference-fns.R",sep=''),chdir=TRUE)
+# source(paste(scriptdir,"sim-context-fns.R",sep=''),chdir=TRUE)
 require(mcmc)
 
 load(infile)  # has mrun and previous things (called 'datafile' in -inference.R)

### 2015-06-12 bcells/view-resids.R

@@ -1,5 +1,5 @@
 #!/usr/bin/Rscript
-source("../context-inference-fns.R")
+source("../context-inference-fns.R",chdir=TRUE)
 
 resultsfile <- commandArgs(TRUE)[1]
 if (length(commandArgs(TRUE))<2) {

### 2015-06-12 compute-resids.R

@@ -21,7 +21,7 @@ if (is.null(opt$infile)) { stop("No input file.  Run\n  compute-resids.R -h\n fo
 if (is.null(opt$basedir)) { opt$basedir <- dirname(opt$infile) }
 print(opt) # this will go in the pbs log
 
-source("../context-inference-fns.R")
+source("../context-inference-fns.R",chdir=TRUE)
 options(error = print.and.dump)
 
 # and fitted model

### 2015-06-12 gather-results.R

@@ -11,7 +11,7 @@ option_list <- list(
         )
 opt <- parse_args(OptionParser(option_list=option_list,description=usage))
 
-source("../context-inference-fns.R")
+source("../context-inference-fns.R",chdir=TRUE)
 
 load(opt$sim)  # provides "simseq.opt"    "simseq.config" "simseqs"
 load(opt$fit)  # provides "model"

### 2015-06-12 ising/ising-mcmc.R

@@ -35,8 +35,8 @@ if (!file.exists(basedir)) {
 }
 
 scriptdir <- "../"
-source(paste(scriptdir,"context-inference-fns.R",sep=''))
-source(paste(scriptdir,"sim-context-fns.R",sep=''))
+source(paste(scriptdir,"context-inference-fns.R",sep=''),chdir=TRUE)
+source(paste(scriptdir,"sim-context-fns.R",sep=''),chdir=TRUE)
 require(mcmc)
 
 # set-up

### 2015-06-12 model-fitting/interactive-noodling.R

@@ -10,8 +10,8 @@ sink(file=logcon, type="message")
 sink(file=logcon, type="output")   # send both to log file
 
 scriptdir <- "../"
-source(paste(scriptdir,"context-inference-fns.R",sep=''))
-source(paste(scriptdir,"sim-context-fns.R",sep=''))
+source(paste(scriptdir,"context-inference-fns.R",sep=''),chdir=TRUE)
+source(paste(scriptdir,"sim-context-fns.R",sep=''),chdir=TRUE)
 
 lwin <- 1
 

### 2015-06-12 mtCDNA/mtCDNA-setup.R

@@ -1,7 +1,7 @@
 #!/usr/bin/Rscript --vanilla
 # setup for mtCDNA inference
 
-source("../context.R")
+source("../context.R",chdir=TRUE)
 codonpairs <- combn( as.character(codons$codon), 2 )
 ndiffs <- apply( codonpairs, 2, function (x) { sum( do.call( "!=",  strsplit(x,"") ) ) } )
 codonpairs <- codonpairs[,ndiffs==1]

### 2015-06-12 mus/quick-residuals.R

@@ -1,6 +1,6 @@
 #!/usr/bin/R
 
-source("../context-inference-fns.R")
+source("../context-inference-fns.R",chdir=TRUE)
 
 # long.win <- list( leftwin=2, rightwin=2, shortwin=3, longwin=7 )
 long.win <- list( leftwin=1, rightwin=1, shortwin=5, longwin=7 )

### 2015-06-12 sim-seq.R

@@ -37,8 +37,8 @@ if ( is.null(opt$configfile) | is.null(opt$seqlen) )  { stop("Rscript sim-seq.R 
 if ( !file.exists(opt$configfile) ) { stop("Could not find config file `", opt$configfile, "`.") }
 if ( !is.null(opt$seed) ) { set.seed(opt$seed) }
 
-source("../context-inference-fns.R")
-source("../sim-context-fns.R")
+source("../context-inference-fns.R",chdir=TRUE)
+source("../sim-context-fns.R",chdir=TRUE)
 
 # identifiers
 if (is.null(opt$outfile)) {

### 2015-06-12 sims-selection/sim-selection-cpg.R

@@ -46,8 +46,8 @@ if (!is.null(logfile)) {
 
 bases <- c("A","T","C","G")
 
-source("../sim-context-fns-dual.R")
-source("../context-inference-fns.R")
+source("../sim-context-fns-dual.R",chdir=TRUE)
+source("../context-inference-fns.R",chdir=TRUE)
 
 # maximum size of pattern (for simulation)
 mutpats <- c(

### 2015-06-12 tasep/tasep-inference.R

@@ -31,8 +31,8 @@ if (logfile!="" & !interactive()) {
 }
 
 scriptdir <- "../"
-source(paste(scriptdir,"context-inference-fns.R",sep=''))
-source(paste(scriptdir,"sim-context-fns.R",sep=''))
+source(paste(scriptdir,"context-inference-fns.R",sep=''),chdir=TRUE)
+source(paste(scriptdir,"sim-context-fns.R",sep=''),chdir=TRUE)
 
 require(mcmc)
 

### 2015-06-12 templated-Rmd.sh

@@ -19,7 +19,7 @@ outfile=$(pwd)/$(dirname $RData)/$(basename $RData .RData)-$(basename $template 
 R --vanilla --slave << EOF
 library("rmarkdown")
 
-source("$(dirname $0)/context-inference-fns.R")
+source("$(dirname $0)/context-inference-fns.R",chdir=TRUE)
 for (rdata in ${RDataList}) { cat("Loading ", rdata, "\n"); load(rdata) }
 render("$template", output_file="$outfile")
 EOF

### 2015-06-12 testing-code/check-longmuts.R

@@ -2,8 +2,8 @@
 
 # a contrived example to test longer mutpats
 
-source("../sim-context-fns.R")
-source("../context-inference-fns.R")
+source("../sim-context-fns.R",chdir=TRUE)
+source("../context-inference-fns.R",chdir=TRUE)
 
 bases <- c("O","X")
 

### 2015-06-12 testing-code/parse-tree-config-test.R

@@ -1,8 +1,8 @@
 #!/usr/bin/Rscript
 
-source("../context-inference-fns.R")
-source("../sim-context-fns.R")
-source("../input-output.R")
+source("../context-inference-fns.R",chdir=TRUE)
+source("../sim-context-fns.R",chdir=TRUE)
+source("../input-output.R",chdir=TRUE)
 
 
 ## Pattern should be:

### 2015-06-12 testing-code/varying-selcoef-test.R

@@ -1,8 +1,8 @@
 #!/usr/bin/Rscript
 
-source("../context-inference-fns.R")
-source("../sim-context-fns.R")
-source("../input-output.R")
+source("../context-inference-fns.R",chdir=TRUE)
+source("../sim-context-fns.R",chdir=TRUE)
+source("../input-output.R",chdir=TRUE)
 
 # this has X's as VERY deleterious; O's as VERY beneficial
 #  and  sp2 on a LONG branch

### 2015-06-12 writeup-plots.R

@@ -1,6 +1,6 @@
 scriptdir <- "./"
-source(paste(scriptdir,"context-inference-fns.R",sep=''))
-source(paste(scriptdir,"sim-context-fns.R",sep=''))
+source(paste(scriptdir,"context-inference-fns.R",sep=''),chdir=TRUE)
+source(paste(scriptdir,"sim-context-fns.R",sep=''),chdir=TRUE)
 require(mcmc)
 
 

### 2015-07-25 simulation.Rmd
testing shape

@@
 
-## residuals for 9-5 Tmers
-```{r resids_9_5, results="asis", fig.align='center'}
-resids <- counts_table(in_longwin=9, in_shortwin=5, half_n_show=20)
+## residuals for 3-3 Tmers
+```{r resids_3_3, results="asis", fig.align='center'}
+resids <- counts_table(in_longwin=3, in_shortwin=3, half_n_show=20)
+counts_plot(resids)
+```
+
+## residuals for 5-5 Tmers
+```{r resids_5_5, results="asis", fig.align='center'}
+resids <- counts_table(in_longwin=5, in_shortwin=5, half_n_show=20)
 counts_plot(resids)

### 2015-07-29 simulation.Rmd
added hessian to output

@@
 
-The likelihood surface about the fitted and simulated values (red are the values simulated under, black is the fitted values):
+The profile likelihood surface about the fitted and simulated values (red are the values simulated under, black is the fitted values):
 ```{r likelihood_surface, echo=FALSE, results="asis", fig.align='center'}
@@
 }
+```
+
+Here is the Hessian of the likelihood (obtained by numerical differentiation)
+```{r likelihood_hessian, echo=FALSE}
+# Hessian of the likelihood surface
+likfun.hess <- hessian(function(x)fit.model@likfun(x*timevec[fit.model@results$use.par]),method.args=list(eps=min(coef(model)/10)),x=coef(model)[fit.model@results$use.par])
+colnames( likfun.hess ) <- rownames( likfun.hess ) <- names(coef(fit.model))[fit.model@results$use.par]
+pander(likfun.hess)
+```
+and its inverse (the asymptotic covariance matrix)
+```{r likelihood_covariance, echo=FALSE}
+# asymptotic covariance matrix
+est.covmat <- ginv(likfun.hess)
+pander(est.covmat)
+```
+and the corresponding correlation matrix
+```{r likelihood_covariance, echo=FALSE}
+# asymptotic covariance matrix
+pander(cov2cor(est.covmat))
 ```

### 2015-07-29 future-vignette.R

@@ -11,6 +11,8 @@
 
 source("../context-inference-fns.R",chdir=TRUE)
 source("../sim-context-fns.R",chdir=TRUE)
+require(MASS)
+require(numDeriv)
 
 modelfile <- "less-simple-shape-model.json"
 
@@ -136,6 +138,12 @@ compare.params <- data.frame(
         fixed=!fit.model@results$use.par
     )
 
+# Hessian of the likelihood surface
+likfun.hess <- hessian(function(x)fit.model@likfun(x*timevec[fit.model@results$use.par]),method.args=list(eps=min(coef(model)/10)),x=coef(model)[fit.model@results$use.par])
+colnames( likfun.hess ) <- rownames( likfun.hess ) <- names(coef(fit.model))[fit.model@results$use.par]
+# asymptotic covariance matrix
+est.covmat <- ginv(likfun.hess)
+
 f <- function (x) { fit.model@likfun((x*timevec)[fit.model@results$use.par]) }
 
 # profiles
@@ -165,67 +173,3 @@ for (j in (1:nrow(compare.params))[fit.model@results$use.par]) {
     }
 }
 
-
-if (FALSE) {
-
-
-## check likelihood is at maximum at truth for predicted counts
-
-pred.counts <- fitted(fit.model)
-pred.env <- new.env()
-assign( "counts", pred.counts, envir=pred.env )
-pred.likfun <- likfun
-environment(pred.likfun) <- pred.env
-
-pred.optim.results <- optim( par=initpar[use.par], fn=pred.likfun, method="L-BFGS-B", lower=lbs[use.par], upper=ubs[use.par], control=list(fnscale=(-1)*abs(baseval), parscale=parscale[use.par] ) )
-
-pred.optim.results$use.par <- use.par
-pred.optim.results$parscale <- parscale
-pred.optim.results$initpar <- initpar
-
-pred.optim.par <- initpar
-pred.optim.par[use.par] <- pred.optim.results$par
-pred.optim.results$par <- pred.optim.par
-
-pred.fit.model <- new( "context",
-             counts=pred.counts,
-             genmatrix=genmatrix,
-             projmatrix=projmatrix,
-             mutrates=pred.optim.results$par[1:nmuts(genmatrix)],
-             selcoef=pred.optim.results$par[seq(nmuts(genmatrix)+1,length.out=nsel(genmatrix))],
[…]

### 2015-07-29 simulation.Rmd
.

@@
 and the corresponding correlation matrix
-```{r likelihood_covariance, echo=FALSE}
+```{r likelihood_correlation, echo=FALSE}
 # asymptotic covariance matrix

### 2015-07-29 simulation.Rmd
.

@@
 library(ggplot2)
+library(numDeriv)
+library(MASS)
 panderOptions("table.split.table", Inf)  # Let tables become arbitrarily wide rather than wrapping.

### 2015-08-02 simulation.Rmd
updated Ne to proper value in shape/

@@
 # Hessian of the likelihood surface
-likfun.hess <- hessian(function(x)fit.model@likfun(x*timevec[fit.model@results$use.par]),method.args=list(eps=min(coef(model)/10)),x=coef(model)[fit.model@results$use.par])
-colnames( likfun.hess ) <- rownames( likfun.hess ) <- names(coef(fit.model))[fit.model@results$use.par]
+likfun.hess <- hessian(function(x)model@likfun(x*timevec[model@results$use.par]),method.args=list(eps=min(coef(model)/10)),x=coef(model)[model@results$use.par])
+colnames( likfun.hess ) <- rownames( likfun.hess ) <- names(coef(model))[model@results$use.par]
 pander(likfun.hess)

### 2015-08-04 likelihood-surface.Rmd
likelihood surface

---
title: "The likelihood surface"
date: "`r date()`"
---

```{r setup, include=FALSE}
required.input <- c("modelfile","longwin","shortwin","leftwin","tlen")
if (!all(sapply(required.input,exists))) {
    stop("Not all required input (", paste(required.input[!sapply(required.input,exists)],collapse=", "), ") has been given.")
}
names(required.input) <- required.input
opt <- lapply( required.input, get )
opt$boundary <- "none"
source("context-inference-fns.R",chdir=TRUE)
library(MASS)
library(numDeriv)
library(pander)
if (!file.exists(modelfile)) { stop("Can't find ", modelfile) }
model.config <- fill.default.config( read.config(modelfile) )
model.config$fixfn <- parse.fixfn( model.config$fixfn, model.config$fixfn.params )
if (!exists("gmfile")) {
    gmfile <- file.path(dirname(modelfile),gsub("%",opt$longwin,model.config$genmatrix,fixed=TRUE))
}
if (file.exists(gmfile)) {
    load(gmfile)
} else { 
    cat("Can't find generator matrix ", gmfile, " -- making a new one.\n")
    genmatrix <- do.call( makegenmatrix,
                    c( list(
                        patlen=opt$longwin, 
                        mutpats=model.config$mutpats, 
                        selpats=model.config$selpats, 
                        selfactors=model.config$selfactors, 
                        boundary=opt$boundary, 
                        bases=model.config$bases, 
                        fixfn=model.config$fixfn ),
                        model.config$fixfn.params
                    )
                )
}
genmatrix@x <- do.call( update, c( list(G=genmatrix, mutrates=model.config$mutrates, selcoef=model.config$selcoef), model.config$fixfn.params ) )
projmatrix <- collapsepatmatrix( ipatterns=rownames(genmatrix), leftwin=opt$leftwin, rightwin=opt$longwin-opt$shortwin-opt$leftwin, bases=genmatrix@bases )
```

```{r expected_counts, include=FALSE}
total.counts <- 1e8
expected.counts <- predictcounts(opt$longwin, opt$shortwin, opt$leftwin, initcounts=rep(total.counts/nrow(genmatrix),nrow(genmatrix)), genmatrix=genmatrix, projmatrix=projmatrix, tlen=opt$tlen)

initpar <- with(model.config, unlist(c(mutrates*opt$tlen,selcoef,fixfn.params)) )
parscale <- with(model.config, unlist( c(mutrates.scale*opt$tlen, selcoef.scale, fixfn.params.scale) ) )
names(initpar) <- names(parscale) <- c( mutnames(genmatrix@mutpats), selnames(genmatrix@selpats), fixparams(genmatrix) )

# skip these parameters
use.par <- ( parscale!=0 )
params <- initpar

likfun <- function (sub.params){
    # params are: mutrates, selcoef, fixparams
    params[use.par] <- sub.params
    fparams <- params[seq( 1+nmuts(genmatrix)+nsel(genmatrix), length.out=length(fixparams(genmatrix)) )]
    names(fparams) <- fixparams(genmatrix)
    genmatrix@x <- do.call( update, c( list( G=genmatrix,mutrates=params[1:nmuts(genmatrix)],selcoef=params[seq(1+nmuts(genmatrix),length.out=nsel(genmatrix))]), as.list(fparams) ) )
    # this is collapsed transition matrix
    subtransmatrix <- computetransmatrix( genmatrix, projmatrix, tlen=1, time="fixed") # shape=params[length(params)], time="gamma" )
    # return POSITIVE log-likelihood
    ans <- sum( counts@counts * log(subtransmatrix) )
    if (!is.finite(ans)) { print(paste("Warning: non-finite likelihood with params:",paste(params,collapse=", "))) }
    return(ans)
}

model <- new( "context",
             counts=expected.counts,
             genmatrix=genmatrix,
             projmatrix=projmatrix,
             mutrates=model.config$mutrates*opt$tlen,
             selcoef=model.config$selcoef,
             params=unlist(model.config$fixfn.params),
             results=list(use.par=use.par),
             likfun=likfun,
             invocation="by hand"
         )
expected.env <- new.env()
assign("counts",model@counts,envir=expected.env)
environment(model@likfun) <- expected.env

```

Model configuration (`r modelfile`)
===================================

```{r show_config, echo=FALSE}
pander(model.config)
```

Hessian of the likelihood
=========================

Here are the square roots of the diagonal of the inverse of the Hessian:

```{r hessian, echo=FALSE}
time <- as.numeric(opt$tlen)
timevec <- c( rep(as.numeric(opt$tlen),nmuts(model)), rep(1,length(coef(model))-nmuts(model)) )
h <- function(x)model@likfun(x*timevec[model@results$use.par])
likfun.hess <- hessian(h,method.args=list(eps=min(abs(coef(model))[1:nmuts(model)]/10)),x=coef(model)[model@results$use.par])
colnames( likfun.hess ) <- rownames( likfun.hess ) <- names(coef(model))[model@results$use.par]
# asymptotic covariance matrix
est.covmat <- (-1)*ginv(likfun.hess)
dimnames(est.covmat) <- dimnames(likfun.hess)
pander( sqrt(diag(est.covmat)) )
```
and the Hessian itself:
```{r hessian_itself, echo=FALSE}
pander(likfun.hess)
```

Profiles of the likelihood
==========================

And, the pairwise profiles of the likelihood:
```{r pairwise_prof, echo=FALSE}
likfun.grids <- likelihood.surface( model, tlen=opt$tlen )
```

### 2015-08-04 likelihood-surface.Rmd
.

@@
 ```{r setup, include=FALSE}
-required.input <- c("modelfile","longwin","shortwin","leftwin","tlen")
+required.input <- c("modelfile","longwin","shortwin","leftwin","tlen","ncounts")
 if (!all(sapply(required.input,exists))) {
@@
 opt$boundary <- "none"
+if (!exists("lik.grid")) { lik.grid <- 10 }
 source("context-inference-fns.R",chdir=TRUE)
@@
 
+Here is an estimate of the likelihood surface near the true values for the model in `r modelfile`,
+at tuple size:
+
+*  *longwin* : `r opt$longwin`
+*  *shortwin* : `r opt$shortwin`
+*  *leftwin* : `r opt$leftwin`
+and other parameters
+*  *total time* : `r opt$tlen`  (`tlen`)
+*  *number of counts* : `r opt$ncounts`  (`ncounts`)
+*  *number of grid points for likelihood* : `r lik.grid` (`lik.grid`)
+
 ```{r expected_counts, include=FALSE}
-total.counts <- 1e8
-expected.counts <- predictcounts(opt$longwin, opt$shortwin, opt$leftwin, initcounts=rep(total.counts/nrow(genmatrix),nrow(genmatrix)), genmatrix=genmatrix, projmatrix=projmatrix, tlen=opt$tlen)
+expected.counts <- predictcounts(opt$longwin, opt$shortwin, opt$leftwin, initcounts=rep(ncounts/nrow(genmatrix),nrow(genmatrix)), genmatrix=genmatrix, projmatrix=projmatrix, tlen=opt$tlen)
 
@@
 ```{r pairwise_prof, echo=FALSE}
-likfun.grids <- likelihood.surface( model, tlen=opt$tlen )
+likfun.grids <- likelihood.surface( model, tlen=opt$tlen, ngrid=lik.grid )
 ```

### 2015-11-15 writeups/gamma-prob.Rmd
gamma updating

---
title: "Checking the extension of an exponential time."
date: "`r date()`"
---

\newcommand{\deq}{\stackrel{\scriptscriptstyle{d}}{=}}
\renewcommand{\P}{\mathbb{P}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\R}{\mathbb{R}}

```{r doc_setup, include=FALSE}
fig.dim <- 5
knitr::opts_chunk$set(fig.width=2*fig.dim,fig.height=fig.dim,fig.align='center')
```

Let $N$ be a Poisson process on $\R$,
let $L$ be a Poisson process on $\R^2$,
and let $T_t = \inf\{s>0 : L([0,s]\times[0,1/t])>0\}$.
The claim is that if we let
$$ Z(t) = N(T_t) $$
then this process has increments
$$
Z(t+u) - Z(t) \deq X_t + Y_t ,
$$
where $X_t$ is Bernoulli($t/(t+u)$)
and $Y_t$ is Geometric($(u+t)/(u+t+ut)$).

Here's the process:
```{r sim_process}
N <- cumsum(rexp(1e3))
min.t <- 1
max.x <- 1000
nL <- rpois(1,1/min.t*max.x) 
L <- cbind( runif(nL)*max.x, runif(nL)/min.t )
L <- L[order(L[,1],decreasing=FALSE),]
Tt <- function (t,L) {
    useL <- outer(L[,2],1/t,"<")
    L[ apply(useL,2,function(u)min(which(u))), 1 ]
}
Z <- function (t,L) {
    colSums(outer(N,Tt(t,L),"<"))
}
tt <- seq(min.t,500,length.out=1000)
layout(t(1:2))
plot( tt, Tt(tt,L), type='l', main="T(t)" )
plot( tt, Z(tt,L), type='l', main="Z(t)" )
```

Clearly, $Z(t+u)=Z(t)$ if the $y$-coordinate of the point of $L$
whose $x$-coordinate is $T_t$ is less than $1/(t+u)$;
this occurs with probability $t/(t+u)$.
If this is not the case,
then $T_t$ increments by an Exponential($1/(t+u)$) time,
which increments $Z$ by a Geometric($1/(1+(t+u)^{-1})=(t+u)/(1+t+u)$) number,
so we should have
$$ Z(t+u)-Z(t) \deq \text{Geom}\left(\frac{t+u}{1+t+u}\right) . $$

Increments of $T(t)$
------------------

We claim that
$$
T(t+u) \deq T(t) + \sum_{k=0}^M T_k,
$$
where the $T_k$ are independent copies of $T(t)$
and $M$ is Geometric($t/(t+u)$).

Here is the predicted distribution:
```{r geoexponential}
rgeoexp <- function (n,t,dt) {
    M <- rgeom(n,t/(t+dt))
    dT <- numeric(n)
    while ( any(M>0) ) {
        dT[M>0] <- dT[M>0] + rexp(n,1/t)[M>0]
        M <- M-1
    }
    return(dT)
}
rgs <- rgeoexp(1e4,t=10,dt=5)
mean(rgs==0)
```

And, the simulated distribution:
```{r sim_dT}
sim_dT <- function (t,dt) {
    N <- cumsum( rexp(5*(t+dt)) )
    max.x <- 100*t
    nL <- rpois(1,max.x/t)
    L <- cbind( runif(nL)*max.x, runif(nL)/t )
    L <- L[order(L[,1],decreasing=FALSE),]
    Tt(t+dt,L)-Tt(t,L)
}
sim.rgs <- replicate(1e4,sim_dT(t=10,dt=5))
mean(sim.rgs==0)
```

And the comparison for nonzero probabilities:
```{r compare_dT}
breaks <- seq(0,max(rgs,sim.rgs),length.out=50)
hist(rgs[rgs>0],breaks=breaks,freq=FALSE,col=adjustcolor("black",0.5))
hist(sim.rgs[sim.rgs>0],breaks=breaks,freq=FALSE,col=adjustcolor("red",0.5),add=TRUE)
```

Increments of $Z(t)$
------------------

Here's the predicted distribution:
```{r bernougeo}
rbernougeo <- function (n,t,dt) {
    ifelse( rbinom(n,1,t/(t+dt))>0, 0, rgeom(n,1/(1+t+dt)) )
}
bgs <- rbernougeo(1e6,t=10,dt=5)
mean(bgs==0)
```

And, the simulated distribution:
```{r sim_distrn}
sim_ppp <- function (t,dt) {
    N <- cumsum( rexp(5*(t+dt)) )
    max.x <- 100*t
    nL <- rpois(1,max.x/t)
    L <- cbind( runif(nL)*max.x, runif(nL)/t )
    L <- L[order(L[,1],decreasing=FALSE),]
    Z(t+dt,L)-Z(t,L)
}
sim.bgs <- replicate(1e4,sim_ppp(t=10,dt=5))
mean(sim.bgs==0)
```

```{r compare_Z}
breaks <- seq(0,max(bgs,sim.bgs,length.out=30))
hist(bgs[bgs>0],breaks=breaks,freq=FALSE,col=adjustcolor("black",0.5))
hist(sim.bgs[sim.bgs>0],breaks=breaks,freq=FALSE,col=adjustcolor("red",0.5),add=TRUE)
```

### 2015-11-15 testing-code/check-gammaAtv.R

@@ -45,3 +45,31 @@ extend.fac <- 0.5
 egM <- extend.gammam( gM, eps=1-extend.fac, tol=1e-8 )
 stopifnot( all( abs( egM - matrix( c(1/(1+scale/extend.fac*M[1,2]), 0, 1-1/(1+scale/extend.fac*M[1,2]), 1.0), nrow=2) ) < 1e-8 ) )
 stopifnot( all( abs( egM - gammam( M, scale=scale/extend.fac, shape=1, tol=1e-8 ) ) < 2e-8 ) )
+
+# bigger matrix
+M <- matrix( c(
+    0,1,1,1,1,0,0,0,1,0,0,0,1,0,0,0,#AA
+    1,0,1,1,0,1,0,0,0,1,0,0,0,1,0,0,#CA
+    1,1,0,1,0,0,1,0,0,0,1,0,0,0,1,0,#GA
+    1,1,1,0,0,0,0,1,0,0,0,1,0,0,0,1,#TA
+    1,0,0,0,0,1,1,1,1,0,0,0,1,0,0,0,#AC
+    0,1,0,0,1,0,1,1,0,1,0,0,0,1,0,0,#CC
+    0,0,1,0,1,1,0,1,0,0,1,0,0,0,1,0,#GC
+    0,0,0,1,1,1,1,0,0,0,0,1,0,0,0,1,#TC
+    1,0,0,0,1,0,0,0,0,1,1,1,1,0,0,0,#AG
+    0,1,0,0,0,1,0,0,1,0,1,1,0,1,0,0,#CG
+    0,0,1,0,0,0,1,0,1,1,0,1,0,0,1,0,#GG
+    0,0,0,1,0,0,0,1,1,1,1,0,0,0,0,1,#TG
+    1,0,0,0,1,0,0,0,1,0,0,0,0,1,1,1,#AT
+    0,1,0,0,0,1,0,0,0,1,0,0,1,0,1,1,#CT
+    0,0,1,0,0,0,1,0,0,0,1,0,1,1,0,1,#GT
+    0,0,0,1,0,0,0,1,0,0,0,1,1,1,1,0 #TT
+    ), nrow=16, byrow=TRUE )
+diag(M) <- (-1)*rowSums(M)
+
+scale <- 0.3
+gM <- gammam(M, scale=scale, shape=1, tol=1e-8)
+extend.fac <- 0.8
+egM <- extend.gammam( gM, eps=1-extend.fac, tol=1e-8 )
[…]

### 2016-07-04 writeups/context-dependence.tex
adjusted nestly
Clearing out some of my silly comments.

@@
 For instance, \citet{arndt2003sequence} studied dinucleotide transitions,
+%EM suggest going back to http://www.sciencedirect.com/science/article/pii/016747819290134L
 and \citet{yaari2013models} displayed strong heterogeneity in the probability of synonymous mutations across 5-mers
@@
   A sequence of genome can be written using A, C, G, and T;
-  in the most general model of independent mutation across sites, each of the 12 possible transitions occurs at its own rate.\todo{Given the importance of the independent model, I think it deserves a more complete presentation.}
+  in the most general model of independent mutation across sites, each of the 12 possible transitions occurs at its own rate.
   Furthermore, it is a well-known observation that in many species,
-  adjacent, methylated CG dinucleotides (``CpG sites'') \todo{not all CpG sites are methylated} have a much higher mutation rate to TG and CA
-  than either single nucleotide change \todo{Those *are* single nucleotide changes}.
+  adjacent, methylated CG dinucleotides (``CpG sites'') have a much higher mutation rate to TG and CA
+  than either single nucleotide change under the independent model.
   Embellishing the single-nucleotide model with this additional rate results in the model defined by
@@
   Ignoring transitions that don't change the state,
-  if patterns $u$ and $v$ differ at only one site then $u$ changes to $v$ at rate $1/(1+\exp(H(v)-H(u)))$ \todo{Hum, this isn't immediately obvious to me-- quick explanation please?
-Also, shouldn't there be a $\lambda$ up top?}, so
+  if patterns $u$ and $v$ differ at only one site then $u$ changes to $v$ at rate $1/(1+\exp(H(v)-H(u)))$ \todo{Shouldn't there be a $\lambda$ up top?}, so
 
@@
 with probability proportional to the corresponding rate, i.e.\
-\todo{A little confused here. Previously, the $j$ in $u^j$ was used to index the possible patterns, right? Now it's indexing possible changes at a site.}
+\erick{Previously, the $j$ in $\mu^j$ was used to index the possible patterns. Now it's indexing possible changes at a site.}
 \[
@@
 In the CpG model it is not immediately clear what information is retained.
-\todo[inline]{
-I think that the stationary distribution of the single-nucleotide model should be identifiable for sure.
-Then, I would guess the two CpG rates are identifiable modulo a rate factor that compares the single-nucleotide mutation rate to the CpG mutation rate.
-}
 
@@
 
-Since we cannot compute reasonably the full likelihood for the data \todo{Well, can we prove that it's not possible? Perhaps say that we can't using straightforward exponentiation?},
+Since we cannot compute reasonably the full likelihood for the data,
 we will need to choose appropriate summary statistics
@@
 If $i_1 < \ldots < i_{n(x)}$ are a set of nonoverlapping sites where the initial sequence matches $x$
--- i.e.\ $X_{i_k}^{(2\ell+m)}(0)=x$ and $i_{k+1} > i_k + 2\ell+m$ for all $k$ -- \todo{En dash close to math confuses me! This sentence is a doozy.}
+-- i.e.\ $X_{i_k}^{(2\ell+m)}(0)=x$ and $i_{k+1} > i_k + 2\ell+m$ for all $k$ %En dash close to math could be confusing. This sentence is a doozy.
 then the counts $N(x,y)=\#\{ i_k \st X_{i_k+\ell}^{(m)}(t)=y\}$ (the number of these sites that match the smaller pattern $y$ at time $t$)
@@
 If we use only a subset of patterns $\{(x_k,y_k)\}$ chosen so that $\E[M_1(x,y) M_{k+1}(x,y)]$ is small (or zero),
-\todo{M not defined, but I'm guessing that this is effectively saying that the probability of having two different patterns apply to the same sequences is small?}
+\erick{M not defined, but I'm guessing that this is effectively saying that the probability of having two different patterns apply to the same sequences is small?}
 the entire collection $\{N(x_k,y_k)\}$ will be close to independent Poisson.

### 2016-07-17 writeups/context-dependence.tex
added writeup of computing full likelihoods from abphylo

@@
 \newcommand{\given}{\;\mid\;}
+\newcommand{\like}{\mathcal L}
+\newcommand{\loglike}{\ell}
+\newcommand{\alike}{\widetilde{\like}}
+\newcommand{\aloglike}{\widetilde{\loglike}}
 
@@
 
+%%%%%%
+\section{Full likelihood}
+
+\emph{The following is copied over from work on the abphylo project.}
+
+The s5f model is defined as follows:
+for each $(a,b,c,d,e,x) \subset \{\nA,\nC,\nG,\nT\}$,
+it gives a (relative) mutation rate
+\begin{align}
+    \gamma(c,x;a,b,d,e) = \text{( mutation rate of $c \to x$ in context $abcde$ )} .
+\end{align}
+Given these rates, the only unknown parameter is the total amount of time
+that the sequence has been mutated for.
+Suppose that $X(t)$ is a sequence of length $n$ evolving under this model,
+and let $G_n$ be the corresponding generator matrix,
+a $4^n \times 4^n$ matrix.
+(In the s5f model, the mutation rates are only specified in the middle of 5-mers,
+so the two bases on each end never mutate.)
+The transition probabilities are $e^{tG_n}$,
+and by marginalization,
+we can compute quantities like
+the probability for observing the middle set of basis
+given the flanking $\ell$ on each end:
+\begin{align}
+    p_{\ell,n-2\ell}(t;a,b)
+    &=
+    \P\{
+        (X_{\ell+1}(t), X_{\ell+2}(t), \ldots, X_{n-\ell}(t))=(b_1,b_2,\ldots,b_{n-2\ell}) \\
+    &\qquad \qquad
+        \given (X_{1}(0), X_{2}(0), \ldots, X_{n}(0))=(a_1,a_2,\ldots,a_{n})
+    \} .
+\end{align}
+If the size of the context, $\ell$, is big enough,
+then this probability will not depend significantly if the sequence $a$ is in fact embedded in a longer sequence
+(the error decreases exponentially with $\ell$).
+
+The sequence $X(t)$ can be constructed by beginning with $X(0)$,
+putting down a Poisson process of potential mutations of rate $\gamma = \max\{\gamma(c,x;a,b,d,e)\}$,
+and thinning appropriately.
+The process depends on the flanking 2bp,
+so any other mutation within 2bp of a site affects the process at that site,
+and the chance that this occurs depends on the flanking 4bp on either side.
+Concretely, the number of flanking positions in each direction
+that must be kept track of to decide what happens at a given site
+grows by 1 at rate $\gamma$, and by 2 at rate $\gamma$;
+so the amount of flanking sequence on each side affecting a given site at time $t$
+is the sum of a Poisson($\gamma t$) and twice an independent Poisson($\gamma t$).
+This has mean $3\gamma t$ and variance $5 \gamma t$;
+with high probability it will not exceed, say,
+$3\gamma t + 6 \sqrt{\gamma t}$;
+if $\gamma t < 1$ this is less than 9.
+
+How can we use this to compute the likelihood of a full sequence?
+Actually, we know more.
+For two positions $(i,s)$ and $(j,t)$ in sequence $\times$ time,
+let $r(i,s;j,t) = |j-i|/|t-s|$ be the slope of the line between them,
+and if $I$ and $J$ are subsets of sites, let $r(I,s;J,t) = \min\{r(i,s;j,t) : i\in I, j \in J\}$
+be the minimum.
+Then, we know that there is a speed of dependency propagation $r_*$
+such that if $r(I,s;J,t) > r_*$,
+then $X_I(s)$ and $X_J(t)$ are nearly independent.
+Now, fix $t$ and for each site $i$ let $N_i$ denote its $r_*$-neighborhood,
+i.e.,
+\begin{align}
+    N_i = \{ (j,s) : r(i,t; j,s) \le r_*, \; 0 \le s \le t \} .
+\end{align}
+These are depicted in figure~\ref{fig:neighborhoods}A.
+We will write $i \sim j$ if $|i-j| \le 2 r_* t$,
+or equivalently, if $N_i \cap N_j \neq \emptyset$,
+and $i \nsim j$ otherwise.
+We can think of $N_i$ as the parts of ancestral sequences in which a mutation might have affected the state at $(i,t)$.
+so if $i \nsim j$, then $i$ and $j$ are (approximately) independent at time $t$, given the sequence at time 0.
+Concretely,
+we have a small $\epsilon$ such that for any $i \nsim j$,
+\begin{align}
+    &\big|
+    \log( \P\{ X_i(t) = x_i \and X_j(t) = x_j \} ) \\
+    &\qquad {} -
+    \log( \P\{ X_i(t) = x_i \} \P\{ X_j(t) = x_j \} )
+    ) \big|
+    \le \epsilon .
+\end{align}
+
+\begin{figure}
+    \begin{center}
+        \includegraphics{context-neighborhoods-defns}
+
+        \vspace{2em}
+
+        \includegraphics{context-neighborhoods-widths}
+    \end{center}
+    \caption{
+        Diagram of things in the text:
+        (top) Dependency neighboorhods.
+        (bottom) What we need to compute $\P\{X_i(t)=x \given X(0), X_{<i}(t)\}$:
+        site $i$ depends on the blue squares in $N_i$;
+        the sites to the left of this in $L(i)$ are those other sites in $X(t)$
+        whose dependency neighborhoods overlap $N_i$;
+        any further sites to the left of this are basically independent of $X_i(t)$;
+        the red squares give the dependency neighborhood of $L(i)$,
+        so we need to condition on the row of red and blue sites at the top.
+        On the bottom, we have $2w+1$ sites; at the top we have $4w+1$;
+        here $w=2$.
+        \label{fig:neighborhoods}
+    }
+\end{figure}
+
+What we'd like to know is the likelihood, i.e.,
+\begin{align}
+    \like(y|x)
+    &:=
+    \P\{ X(t) = y \given X(0) = x \} .
+\end{align}
+If we define $X_{<i}=(X_1, \ldots, X_{i-1})$, then
+\begin{align}
+    \like(y|x)
+    &=
+    \prod_{i=1}^n
+    \P\{ X_i(t) = y_i \given X(0) = x \and X_{<i}(t) = y_{<i} \} .
+\end{align}
+Now, defining $L(i) = \{ j < i \st j \sim i \}$
+to be the sites to the left of $i$ whose dependency neighborhoods overlap with $i$'s,
+and $A(I) = \{ j \st |j-i| \le r_*t \text{ for some } i \in I\}$
+to be the ancestral sites in the dependency neighborhood of $I$,
+\begin{align}
+&\big|
+\log( \P\{ X_i(t) = y_i \given X(0) = x \and X_{<i}(t) = y_{<i} \} ) \\
+& \qquad {} -
+\log( \P\{ X_i(t) = y_i \given X_{A(L(i))}(0) = x_{A(L(i))} \and X_{L(i)}(t) = y_{L(i)} \} )
+ \big| \\
+&\qquad \qquad \le
+    \epsilon ,
+\end{align}
+so that if we define
+\begin{align}
+    \alike(y|x)
+    &=
+    \prod_{i=1}^n
+    \P\{ X_i(t) = y_i \given X_{A(L(i))}(0) = x_{A(L(i))} \and X_{L(i)}(t) = y_{L(i)} \} ,
+\end{align}
+then (with $\loglike = \log \like$),
+\begin{align}
+    \left| \aloglike(y|x) - \loglike(y|x) \right|
+    \le n \epsilon .
+\end{align}
+
+This is good, because we can compute $\aloglike$:
+let $w=r_*t$ be radius of a dependency neighborhood,
+so that the width of $L(i)$ is $2w$,
+and for a sequence $x$ define $x^{(m)}_i = (x_{i-m+1}, \ldots, x_i)$:
+\begin{align}
+    \alike(y|x)
+    &=
+    \prod_{i=1}^n
+    \frac{
+        p_{w,2w+1}(y^{(2w+1)}_i|x^{(4w+1)}_{i+w})
+    }{
+        p_{w,2w}(y^{(2w)}_{i-1}|x^{(4w)}_{i-1+w})
+    }
+\end{align}
+This is the ratio of the likelihoods of all the $(w,2w+1)$ T-mers,
+divided by the likelihoods of all the $(w,2w)$ T-mers.
+
 %%%%%%%

### 2017-04-19 writeups/check-likelihood.Rmd
tidied up

---
title: "Computing sequence likelihoods"
date: "`r date()`"
---

\newcommand{\deq}{\stackrel{\scriptscriptstyle{d}}{=}}
\renewcommand{\P}{\mathbb{P}}
\newcommand{\E}{\mathbb{E}}

```{r doc_setup, include=FALSE}
fig.dim <- 5
knitr::opts_chunk$set(fig.width=2*fig.dim,fig.height=fig.dim,fig.align='center')
context.dir <- "context"
source(file.path(context.dir,"context-inference-fns.R"),chdir=TRUE)
source(file.path(context.dir,"sim-context-fns.R"),chdir=TRUE)
```

The basic result is that if $L_k$ is the log-likelihood of all the $(k,w)$ T-mers in a sequence,
and $w$ and $k$ are long enough,
then $L_{k+1}-L_k$ should equal the likelihood of the entire sequence.

Let's check this.

Computing the likelihood
========================

Here's a function to compute the likelihood.
Note that the full likelihood of the *counts* includes a combinatorial factor
for where the T-mers occur,
but we *don't want* that factor.
```{r lhood_fun, cache=TRUE}
# opt will be a list with 'longwin', 'shortwin', and 'leftwin' defined.

likelihood_fun <- function (opt,config,seq) {
    ## count-seq: count T-mers
    longpats <- getpatterns(opt$longwin,config$bases)
    shortpats <- getpatterns(opt$shortwin,config$bases)
    counts <- counttrans.list( list(longpats,shortpats), 
        seqlist=seq[c("initseq","finalseq")],
        leftwin=opt$leftwin, bases=config$bases, 
        shift=0, cyclic=TRUE )

    ## make-genmatrix
    genmatrix <- do.call( makegenmatrix, c( list( patlen=opt$longwin ), boundary="none", config ) )
    projmatrix <- collapsepatmatrix( ipatterns=rownames(genmatrix), leftwin=leftwin(counts), fpatterns=colnames(counts) )
    # this is collapsed transition matrix
    subtransmatrix <- computetransmatrix( genmatrix, projmatrix, tlen=seq$tlen, time="fixed")
    # return log-likelihood (*not* negative log-likelihood)
    ans <- sum( counts@counts * log(subtransmatrix) )
    dummy <- function() { 1 } # this will allow us access back to this environment

    return(list(loglik=ans,dummy=dummy))
}
```

The simplest model
==================

Let's start with a model without context-dependence.
```{r acgt_model, cache=TRUE}
acgt.config.json <- '{
    "bases" : [ "A", "C", "G", "T" ],
    "initfreqs" : [ 0.25, 0.25, 0.25, 0.25 ],
    "mutpats" : [
        [ [ "A", "T" ] ],
        [ [ "T", "A" ] ],
        [ [ "C", "G" ] ],
        [ [ "G", "C" ] ],
        [ [ "A", "C" ] ],
        [ [ "T", "G" ] ],
        [ [ "A", "G" ] ],
        [ [ "T", "C" ] ],
        [ [ "C", "A" ] ],
        [ [ "G", "T" ] ],
        [ [ "C", "T" ] ],
        [ [ "G", "A" ] ]
    ],
    "mutrates" : [ 1.0, 1.0, 1.0, 1.0, 1.0, 1.0, 1.0, 1.0, 1.0, 1.0, 1.0, 1.0 ]
}'
acgt.config <- fill.default.config(read.config(json=acgt.config.json))
```

Long time
---------

First, we'll do it for a long time, so the distribution should be uniform:
```{r first_likelihood, cache=TRUE, dependson=c("acgt_model","lhood_fun")}
seq <- simseq.config( seqlen=100, tlen=10.0, config=acgt.config )
opts.numerator <- list( 
            list(longwin=4, shortwin=2, leftwin=1),
            list(longwin=5, shortwin=3, leftwin=1),
            list(longwin=6, shortwin=4, leftwin=1)
        )
opts.denominator <- lapply( opts.numerator, function (x) { x$shortwin <- x$shortwin-1; x } )
loglik.numerators <- loglik.denominators <- list()
for (k in seq_along(opts.numerator)) {
    loglik.numerators[[k]] <- likelihood_fun( opts.numerator[[k]], config=acgt.config, seq=seq )
    loglik.denominators[[k]] <- likelihood_fun( opts.denominator[[k]], config=acgt.config, seq=seq )
}
results <- data.frame( 
      longwin=sapply(opts.numerator,"[[","longwin"),
      numerators=sapply(loglik.numerators,"[[","loglik"),
      denominators=sapply(loglik.denominators,"[[","loglik"),
      theory=(-1)*length(seq[["initseq"]])*log(4)
      )
results$est_loglik <- results$numerators-results$denominators
results
```
The likelihood differences should be close to the theory.
Note that we're treating the string as cyclic, so we don't need to worry about edge effects.

Less long time
--------------

For a shorter time, since the total time is `r seq$tlen` and the mutation rate is 1.0 to each of the possible three other bases,
a base is randomized after an Exponential(4) time;
so the likelihood, where $m$ bases differ from the original and $n-m$ remain the same,
$$\begin{aligned}
\left( e^{-4t} + (1-e^{-4t}) \frac{1}{4} \right)^{n-m}
\left( (1-e^{-4t}) \frac{1}{4} \right)^m  .
\end{aligned}$$
```{r shorter_likelihood, cache=TRUE, dependson=c("acgt_model","lhood_fun")}
seq <- simseq.config( seqlen=100, tlen=0.1, config=acgt.config )
theoretical <- function (seq) {
    n <- length(seq$initseq)
    m <- stringDist( DNAStringSet( seq[c("initseq","finalseq")]), method="hamming" )
    m * ( log((1-exp(-4*seq$tlen))/4) ) + (n-m) * log( exp(-4*seq$tlen) - expm1(-4*seq$tlen)/4)
}
opts.numerator <- list( 
            list(longwin=4, shortwin=2, leftwin=1),
            list(longwin=5, shortwin=3, leftwin=1),
            list(longwin=6, shortwin=4, leftwin=1)
        )
opts.denominator <- lapply( opts.numerator, function (x) { x$shortwin <- x$shortwin-1; x } )
loglik.numerators <- loglik.denominators <- list()
for (k in seq_along(opts.numerator)) {
    loglik.numerators[[k]] <- likelihood_fun( opts.numerator[[k]], config=acgt.config, seq=seq )
    loglik.denominators[[k]] <- likelihood_fun( opts.denominator[[k]], config=acgt.config, seq=seq )
}
results <- data.frame( 
      longwin=sapply(opts.numerator,"[[","longwin"),
      numerators=sapply(loglik.numerators,"[[","loglik"),
      denominators=sapply(loglik.denominators,"[[","loglik"),
      theory=as.numeric(theoretical(seq))
      )
results$est_loglik <- results$numerators-results$denominators
results
```
The "diff" column should match the "theory" column.

Simple, context-dependent model
===============================

Now, let's check this looks sensible for a simple context-dependent one.
```{r binary_model, cache=TRUE, dependson=c("lhood_fun")}
xo.config.json <- '{
    "bases" : [ "O", "X" ], 
    "initfreqs" : [ 0.5, 0.5 ],
    "mutpats" : [
        [ [ "O", "X" ], [ "X", "O" ] ],
        [ [ "OO", "XX" ], [ "XX", "OO" ] ]
    ],
    "mutrates" : [ 1.0, 1.0 ]
}'
xo.config <- fill.default.config(read.config(json=xo.config.json))
xo.seq <- simseq.config( seqlen=10000, tlen=0.1, config=xo.config )
opts.numerator <- list( 
            list(longwin=4, shortwin=2, leftwin=1),
            list(longwin=6, shortwin=2, leftwin=2),
            list(longwin=8, shortwin=2, leftwin=3),
            list(longwin=5, shortwin=3, leftwin=1),
            list(longwin=7, shortwin=3, leftwin=2),
            list(longwin=8, shortwin=3, leftwin=3),
            list(longwin=9, shortwin=3, leftwin=3),
            list(longwin=6, shortwin=4, leftwin=1),
            list(longwin=8, shortwin=4, leftwin=2)
        )
opts.denominator <- lapply( opts.numerator, function (x) { x$shortwin <- x$shortwin-1; x } )
loglik.numerators <- loglik.denominators <- list()
for (k in seq_along(opts.numerator)) {
    loglik.numerators[[k]] <- likelihood_fun( opts.numerator[[k]], config=xo.config, seq=xo.seq )
    loglik.denominators[[k]] <- likelihood_fun( opts.denominator[[k]], config=xo.config, seq=xo.seq )
}
results <- data.frame( 
      longwin=sapply(opts.numerator,"[[","longwin"),
      shortwin=sapply(opts.numerator,"[[","shortwin"),
      numerators=sapply(loglik.numerators,"[[","loglik"),
      denominators=sapply(loglik.denominators,"[[","loglik")
      )
results$est_loglik <- results$numerators-results$denominators
results
```

With a prior on the branch length
========================

Now, let's modify the function to allow for a Gamma prior on the branch length,
with shape $\alpha$ and scale $\theta$.
```{r lhood_fun_gamma, cache=TRUE}
# opt will be a list with 'longwin', 'shortwin', and 'leftwin' defined.
likelihood_fun_gamma <- function (opt,config,seq,shape,scale) {
    ## count-seq: count T-mers
    longpats <- getpatterns(opt$longwin,config$bases)
    shortpats <- getpatterns(opt$shortwin,config$bases)
    counts <- counttrans.list( list(longpats,shortpats), 
        seqlist=seq[c("initseq","finalseq")],
        leftwin=opt$leftwin, bases=config$bases, 
        shift=0, cyclic=TRUE )
    ## make-genmatrix
    genmatrix <- do.call( makegenmatrix, c( list( patlen=opt$longwin ), boundary="none", config ) )
    projmatrix <- collapsepatmatrix( ipatterns=rownames(genmatrix), leftwin=leftwin(counts), fpatterns=colnames(counts) )
    # this is collapsed transition matrix
    subtransmatrix <- computetransmatrix( genmatrix, projmatrix, tlen=scale, shape=shape, time="gamma")
    # return log-likelihood (*not* negative log-likelihood)
    ans <- sum( counts@counts * log(subtransmatrix) )
    dummy <- function() { 1 } # this will allow us access back to this environment
    return(list(loglik=ans,dummy=dummy))
}
```

The first model
--------------

In the first model, in the expression for the likelihood
we just need to adjust the probability of having mutated or not;
the probability that a Gamma($k$,$\theta$) occurs before a Exponential(4)
is
$$\begin{aligned}
\int_0^\infty \frac{1}{\Gamma(k)\theta^k} t^{k-1} e^{-t/\theta} e^{-4t} dt
&=
\int_0^\infty \frac{1}{\Gamma(k)\theta^k} t^{k-1} e^{-t\frac{4\theta+1}{\theta}} dt \\
&=
\frac{1}{\Gamma(k)\theta^k} \left(\frac{\theta}{4\theta+1}\right)^k \int_0^\infty u^{k-1} e^{-u} du \\
&=
\frac{1}{(4\theta+1)^k}
\end{aligned}$$
(also the Laplace transform for a Gamma distribution)
and so the likelihood of a sequence with $m$ sites differing is, as above,
$$\begin{aligned}
\left( \frac{1}{(4θ+1)ᵏ} + \left(1-\frac{1}{(4\theta+1)ᵏ}\right) \frac{1}{4} \right)^{n-m}
\left( \left(1-\frac{1}{(4\theta+1)ᵏ}\right) \frac{1}{4} \right)^m  .
\end{aligned}$$
```{r gamma_likelihood, cache=TRUE, dependson=c("acgt_model","lhood_fun_gamma")}
shape <- 3
scale <- 0.1/3

seq <- simseq.config( seqlen=100, tlen=0.1, config=acgt.config )
theoretical_gamma <- function (seq,shape,scale) {
    n <- length(seq$initseq)
    m <- stringDist( DNAStringSet( seq[c("initseq","finalseq")]), method="hamming" )
    p <- 1/(4*scale+1)^shape
    m * ( log((1-p)/4) ) + (n-m) * log( p + (1-p)/4 )
}
opts.numerator <- list( 
            list(longwin=4, shortwin=2, leftwin=1),
            list(longwin=5, shortwin=3, leftwin=1),
            list(longwin=6, shortwin=4, leftwin=1)
        )
opts.denominator <- lapply( opts.numerator, function (x) { x$shortwin <- x$shortwin-1; x } )
loglik.numerators <- loglik.denominators <- list()
for (k in seq_along(opts.numerator)) {
    loglik.numerators[[k]] <- likelihood_fun_gamma( opts.numerator[[k]], config=acgt.config, seq=seq, shape=shape, scale=scale )
    loglik.denominators[[k]] <- likelihood_fun_gamma( opts.denominator[[k]], config=acgt.config, seq=seq, shape=shape, scale=scale )
}
results <- data.frame( 
      longwin=sapply(opts.numerator,"[[","longwin"),
      shortwin=sapply(opts.numerator,"[[","shortwin"),
      numerators=sapply(loglik.numerators,"[[","loglik"),
      denominators=sapply(loglik.denominators,"[[","loglik"),
      theory=as.numeric(theoretical_gamma(seq, shape=shape, scale=scale))
      )
results$est_loglik <- results$numerators-results$denominators
results
```
The "diff" column should match the "theory" column.

Finding the MLE for the time
============================

We'll do this with a lightly context-dependent model.
```{r cpg_model, cache=TRUE}
cpg.config.json <- '{
    "bases" : [ "A", "C", "G", "T" ],
    "initfreqs" : [ 0.25, 0.25, 0.25, 0.25 ],
    "mutpats" : [
        [ [ "A", "T" ] ],
        [ [ "T", "A" ] ],
        [ [ "C", "G" ] ],
        [ [ "G", "C" ] ],
        [ [ "A", "C" ] ],
        [ [ "T", "G" ] ],
        [ [ "A", "G" ] ],
        [ [ "T", "C" ] ],
        [ [ "C", "A" ] ],
        [ [ "G", "T" ] ],
        [ [ "C", "T" ] ],
        [ [ "G", "A" ] ],
        [ [ "CG", "CT" ], [ "CG", "AG" ] ]
    ],
    "mutrates" : [ 1.0, 1.0, 1.0, 1.0, 1.0, 1.0, 1.0, 1.0, 1.0, 1.0, 1.0, 1.0, 0.8 ]
}'
cpg.config <- fill.default.config(read.config(json=cpg.config.json))
```

Fixed times
-----------

Let $P_t = e^{tG}$.
This involves computing $P_t V$, where $V$ is a projection matrix.
If we've already computed this for $t$,
and want to check the likelihood for $t+s$, 
then we need $P_{t+s} V = P_s P_t V$ --
so, the previous sub-transition matrix $P_t V$ can be re-used for larger times.

```{r infer_fixed_time_fun, cache=TRUE}
t_lhoods <- function (opt, config, seq, dt, ndt) {
    ans <- numeric(ndt)
    longpats <- getpatterns(opt$longwin,config$bases)
    shortpats <- getpatterns(opt$shortwin,config$bases)
    counts <- counttrans.list( list(longpats,shortpats), 
        seqlist=seq[c("initseq","finalseq")],
        leftwin=opt$leftwin, bases=config$bases, 
        shift=0, cyclic=TRUE )
    genmatrix <- do.call( makegenmatrix, c( list( patlen=opt$longwin ), boundary="none", config ) )
    projmatrix <- collapsepatmatrix( ipatterns=rownames(genmatrix), leftwin=leftwin(counts), fpatterns=colnames(counts) )
    subtransmatrix <- projmatrix
    for (k in 1:ndt) {
        subtransmatrix <- computetransmatrix( genmatrix, subtransmatrix, tlen=dt, time="fixed")  # add time dt
        # return log-likelihood (*not* negative log-likelihood)
        ans[k] <- sum( counts@counts * log(subtransmatrix) )
    }
    return(ans)
}
lower_opt <- function (x) { x$shortwin <- (x$shortwin-1); x }
```

OK, here's the likelihood profile:
```{r infer_fixed_time, cache=TRUE, dependson=c("cpg_model","infer_fixed_time_fun")}
cpg.seq <- simseq.config( seqlen=10000, tlen=0.1, config=cpg.config )
opts <- list( 
            list(longwin=4, shortwin=2, leftwin=1)
        )
dt <- .005
ndt <- ceiling(.2/dt)
tt <- (1:ndt)*dt
numerators <- t_lhoods(opts[[1]], cpg.config, cpg.seq, dt, ndt )
denominators <- t_lhoods(lower_opt(opts[[1]]), cpg.config, cpg.seq, dt, ndt )
cbind(tlen=seq$tlen,num=numerators,den=denominators,diff=numerators-denominators)
plot( tt, numerators-denominators )
abline(v=seq$tlen)
```

Exponential times
-----------------

Consider the case $k=1$, i.e., Exponential.
Let $\lambda = \max_i -G_{ii}$ and $Q_{ij} = G_{ij}/\lambda$ and $Q_{ii}=-\sum_{j \neq i} G_{ij}$,
so that the process jumps as $Q$ at the times of a Poisson($\lambda$) process.
As described in [context/writeups/gamma-prob.html](the context project),
to move from an Exponential($1/\theta$) time to an Exponential($1/(\theta+u)$) time,
we need to add either zero jumps (with probability $\theta/(\theta+u)$)
or a Geometric($\lambda(\theta+u)/(1+\lambda(\theta+u))$) number of jumps.
This is done by `extend.expAtv`;
a similar thing could be done for Gamma times with shape not equal to 1.
```{r infer_expl_time_fun, cache=TRUE}
e_lhoods <- function (opt, config, seq, dscale, ndscale) {
    ans <- numeric(ndscale)
    longpats <- getpatterns(opt$longwin,config$bases)
    shortpats <- getpatterns(opt$shortwin,config$bases)
    counts <- counttrans.list( list(longpats,shortpats), 
        seqlist=seq[c("initseq","finalseq")],
        leftwin=opt$leftwin, bases=config$bases, 
        shift=0, cyclic=TRUE )
    genmatrix <- do.call( makegenmatrix, c( list( patlen=opt$longwin ), boundary="none", config ) )
    projmatrix <- collapsepatmatrix( ipatterns=rownames(genmatrix), leftwin=leftwin(counts), fpatterns=colnames(counts) )
    # subtransmatrix <- computetransmatrix( genmatrix, projmatrix, tlen=dscale, time="gamma")  # add scale dscale
    subtransmatrix <- projmatrix
    scale <- 0
    for (k in 1:ndscale) {
        subtransmatrix <- extend.expAtv( A=genmatrix, scale=scale, dscale=dscale, gAtv=subtransmatrix )  # add scale dscale
        # return log-likelihood (*not* negative log-likelihood)
        ans[k] <- sum( counts@counts * log(subtransmatrix) )
    }
    return(ans)
}
lower_opt <- function (x) { x$shortwin <- (x$shortwin-1); x }
```

OK, here's the likelihood profile:
```{r infer_expl_time, cache=TRUE, dependson=c("cpg_model","infer_expl_time_fun")}
scale <- 0.1
shape <- 1
nseqs <- 1000
seqlen <- 100
random.tlen <- rgamma(nseqs,scale=scale,shape=shape)
expl.seqs <- simseq.config( seqlen=seqlen, tlen=random.tlen, config=cpg.config )
expl.seq <- list(
        initseq = do.call( xscat, lapply(expl.seqs,"[[","initseq") ),
        finalseq = do.call( xscat, lapply(expl.seqs,"[[","finalseq") ),
        scale=scale,
        shape=shape
    )
opts <- list( 
            list(longwin=4, shortwin=2, leftwin=1)
        )
dscale <- .005
ndscale <- ceiling(.2/dscale)
scalevals <- (1:ndscale)*dscale
numerators <- e_lhoods(opts[[1]], cpg.config, expl.seq, dscale, ndscale )
denominators <- e_lhoods(lower_opt(opts[[1]]), cpg.config, expl.seq, dscale, ndscale )
cbind(scale=scalevals,num=numerators,den=denominators,diff=numerators-denominators)
plot( scalevals, numerators-denominators, main=paste( names(opts[[1]]), opts[[1]], sep="=", collapse=", ") )
abline(v=expl.seq$scale)
```

Hm, let's try that again with a longer window?
```{r infer_expl_longer, cache=TRUE, dependson="infer_expl_time"}
longer.opt <- list( longwin=7, shortwin=3, leftwin=2 )
numerators <- e_lhoods(longer.opt, cpg.config, expl.seq, dscale, ndscale )
denominators <- e_lhoods(lower_opt(longer.opt), cpg.config, expl.seq, dscale, ndscale )
cbind(scale=scalevals,num=numerators,den=denominators,diff=numerators-denominators)
plot( scalevals, numerators-denominators, main=paste( names(longer.opt), longer.opt, sep="=", collapse=", ") )
abline(v=expl.seq$scale)
```

Not useful or not used
======================

Gamma inference
---------------

Let $f(t;k,\theta)$ be the density function for a Gamma distribution
with shape $k$ and scale $\theta$ (and hence mean $k \theta$).
Define
$$  P_{k,\theta} = \int_0^\infty f(t;k,\theta) P_t dt ,$$
i.e., the transition matrix after a Gamma($k$,$\theta$)-distributed time.
If $T_i$ are independent, Gamma($k_i$,$\theta$), then $\sum_i T_i$ is Gamma($\sum_i k_i$,$\theta$),
and so
$$\begin{aligned}
    P_{k_1,\theta} P_{k_2,\theta} &= P_{k_1+k_2,\theta} .
\end{aligned}$$

For $k$ an integer,
since a Gamma($k$,$\theta$) is the sum of $k$ independent Exponential($1/\theta$) times,
and by Poisson thinning,
a Gamma($k$,$\theta/(1-\epsilon)$) is the sum of a NegativeBinomial($k$,$1-\epsilon$) number of Exponential($1/\theta$) times,
i.e., the number of times to flip a $1-\epsilon$ coin until getting $k$ heads,
so
$$\begin{aligned}
    P_{k,\theta/(1-\epsilon)}
    &=
    \sum_{n \ge 0} 
    \binom{n+k-1}{n} \epsilon^{n} (1-\epsilon)^k
    P_{1,\theta}^{n+k}  .
\end{aligned}$$
Or, to get a Gamma($k$,$\theta/(1-\epsilon)$),
we could start with a Gamma($k$,$\theta$),
and add on $M$ indepenent Exponential($1/\theta$) times,
where $M$ is NegativeBinomial($N$,$1-\epsilon$)
and $N$ is Binomial($k$,$\epsilon$).

Extending gamma times
---------------------

Now consider the case $k=1$.
By Poisson thinning,
a Geometric($p$) number of independent Exponential($\lambda$) times
is Exponential($\lambda p$);
and therefore
$$\begin{aligned}
    P_{1,\theta/(1-\epsilon)}
    &=
    \sum_{n \ge 1} 
    \epsilon^{n-1} (1-\epsilon)
    P_{1,\theta}^n \\
    &=
    \sum_{n=1}^k
    \epsilon^{n-1} (1-\epsilon)
    P_{1,\theta}^n
    + O(\epsilon^k) .
\end{aligned}$$

```{r build_exponential_fun, cache=TRUE}
extend_transmatrix <- function (P,a,da,eps=da/(a+da),tol=1e-8) {
    # if P is e^{TG} where T is Exp(a)
    #   return e^{SG} where S is Exp(a+da), with da>0.
    # as above, a+da = a/(1-eps), so eps=(da/a)/(1+da/a)
    ans <- Pm <- (1-eps)*P
    m <- ceiling(log(tol)/log(eps))
    # cat("m=",m,"\n")
    for (k in seq(2,length.out=(m-1))) {
        Pm <- eps * P %*% Pm
        ans <- ans + Pm
    }
    return(ans)
}
# prob an exponential(theta) is greater than an exponential(1) is 1/(1+theta), so:
stopifnot( abs( 1/(1+0.25/.8) - extend_transmatrix( 1/(1+0.25), eps=0.2 ) ) <= 1e-8 )
stopifnot( abs( 1/(1+0.25/.8) - extend_transmatrix( 1/(1+0.25), eps=0.2, tol=1e-15 ) ) <= 1e-15 )
```

Here's a quick check of this:
```{r check_exponential_additivity, cache=TRUE, dependson="build_exponential_fun"}
## one step
scale <- 0.3
genmatrix <- do.call( makegenmatrix, c( list( patlen=2 ), boundary="none", acgt.config ) )
subtransmatrix <- computetransmatrix( genmatrix, Diagonal(nrow(genmatrix)), tlen=scale, shape=1, time="gamma", tol=1e-12 )
stm1 <- extend_transmatrix( subtransmatrix, eps=.2, tol=1e-12 )
stm2 <- computetransmatrix( genmatrix, Diagonal(nrow(genmatrix)), tlen=scale/0.8, shape=1, time="gamma", tol=1e-12 )
range(stm1-stm2)
stopifnot(all(abs(stm1-stm2)<1e-12))
## more steps
tt <- seq(0,scale,length.out=11)[-1]
stm.list <- vector(length(tt),mode='list')
stm.list[[1]] <- computetransmatrix( genmatrix, Diagonal(nrow(genmatrix)), tlen=tt[1], shape=1, time="gamma", tol=1e-12 ) 
for (k in seq_along(tt)[-1]) {
    stm.list[[k]] <- extend_transmatrix( stm.list[[k-1]], a=tt[k-1], da=diff(tt)[k-1], tol=1e-12 )
}
range(subtransmatrix-stm.list[[length(tt)]])
stopifnot(all(abs(subtransmatrix-stm.list[[length(tt)]])<1e-12))
```

Now let's look at inference.
Hm, since we're actually dealing with $P_{k,\theta} v$ instead of $P_{k,\theta}$ itself,
maybe that's not too useful.

### 2017-04-19 json-tree-cpg/workflow.sh

@@ -4,7 +4,7 @@ set -eu
 set -o pipefail
 
 
-BASEDIR=sim-01
+BASEDIR=test-sims
 LONGWIN=4
 SHORTWIN=2
 LEFTWIN=1

### 2017-04-19 mtCDNA/mtCDNA-writeup-analysis.R

@@ -1,0 +1,19 @@
+source("../context.R")
+
+
+load(gmfile)
+
+mutpatinfo <- data.frame(
+        mutpat = sapply( lapply( mutpats, lapply, paste, collapse="->" ), paste, collapse=" | " ),
+        codon1 = factor( sapply( lapply( mutpats, lapply, "[", 1 ), paste, collapse=" | " ), levels=levels(codons$codon) ),
+        codon2 = factor( sapply( lapply( mutpats, lapply, "[", 2 ), paste, collapse=" | " ), levels=levels(codons$codon) )
+    )
+mutpatinfo$aa1 <- codons$aa[ match(mutpatinfo$codon1,codons$codon) ]
+mutpatinfo$aa2 <- codons$aa[ match(mutpatinfo$codon2,codons$codon) ]
+mutpatinfo$synon <- with(mutpatinfo, aa1==aa2 )
+
+for (k in seq_along(frame.mrun)) {
+    mutpatinfo[ paste("frame",k,sep='') ] <- colMeans( frame.mrun[[k]]$batch[100:nrow(frame.mrun[[k]]$batch),1+1:length(mutpats)] )
+}
+
+with( mutpatinfo, cbind( levels(mutpat)[mutpat][order(frame1)], levels(mutpat)[mutpat][order(frame2)], levels(mutpat)[mutpat][order(frame3)] ) )

### 2017-04-19 mus/get-regions.sh

@@ -1,0 +1,5 @@
+#!/bin/bash
+
+python ../tuple-counting/get-regions-from-alignment.py --axtfile mm9rn5/chrX.mm9.rn5.net.axt.gz --posfile <(tail -n +2 10_filter_regions.pos | cut -f 1 | sed -e 's/\(.*\):\([0-9]*\)-\([0-9]*\).*/\1 \2 \3/') -o mm9rn5/chrX.mm9.rn5.net.axt.sub.gz
+
+python ../tuple-counting/count-paired-tuples.py --strict --infile mm9rn5/chrX.mm9.rn5.net.axt.sub.gz --informat axt -w 5 -l 1 -r 1 --outfile mm9rn5/chrX.mm9.rn5.5.1.counts

### 2017-04-19 generic.Rmd
switched over to library in scripts

--- 2014-10-19 generic.Rmd
@@
 panderOptions("table.split.table", Inf)  # Let tables become arbitrarily wide rather than wrapping.
-source("context-inference-fns.R")
+
+library(contextual)
+library(contextutils)
+
 top_n_bottom <- function(df, k) { # Take the top and bottom k rows

### 2017-04-19 likelihood-surface.Rmd
switched over to library in scripts

@@
 if (!exists("lik.grid")) { lik.grid <- 10 }
-source("context-inference-fns.R",chdir=TRUE)
+
+library(contextual)
+library(contextutils)
+
 library(MASS)

### 2017-04-19 testing-code/check-sim.Rmd
testing shape
switched over to library in scripts

--- 2015-06-11 testing-code/check-sim.Rmd
@@
 panderOptions("table.split.table", Inf)  # Let tables become arbitrarily wide rather than wrapping.
-source("../context-inference-fns.R")
-source("../sim-context-fns.R")
+
+library(contextual)
+library(contextutils)
+library(simcontext)
+
 top_n_bottom <- function(df, k) { # Take the top and bottom k rows
@@
 
+## residuals for 3-3 Tmers
+```{r, results="asis"}
+counts_table(in_longwin=3, in_shortwin=3, half_n_show=20)
+```
+
+## residuals for 5-5 Tmers
+```{r, results="asis"}
+counts_table(in_longwin=5, in_shortwin=5, half_n_show=20)
+```
+
 ## residuals for 9-5 Tmers

### 2017-04-19 writeups/check-likelihood.Rmd
switched over to library in scripts

@@
 context.dir <- "context"
-source(file.path(context.dir,"context-inference-fns.R"),chdir=TRUE)
-source(file.path(context.dir,"sim-context-fns.R"),chdir=TRUE)
+
+library(contextual)
+library(contextutils)
+library(simcontext)
 ```

### 2017-04-19 RData-to-md.R

@@ -5,8 +5,8 @@ library(pander)
 library(ape)
 
 
-scriptdir <-  dirname(sub("--file=","",commandArgs()[grep("--file",commandArgs())]))
-source(file.path(scriptdir,"sim-context-fns.R"),chdir=TRUE)
+library(contextual)
+library(contextutils)
 
 usage <- '\
 Convert an RData file to a Markdown file. \

### 2017-04-19 bcells/bcells-mcmc.R

@@ -29,10 +29,10 @@ if (mcmcopt$debug & !interactive()) { options(error = quote({dump.frames(to.file
 if (!file.exists(infile)) { stop("Cannot read file ", infile) }
 basename <- gsub("-results.RData","",infile)
 
-scriptdir <- "../"
-source(paste(scriptdir,"context-inference-fns.R",sep=''),chdir=TRUE)
-# source(paste(scriptdir,"sim-context-fns.R",sep=''),chdir=TRUE)
-require(mcmc)
+library(contextual)
+library(contextutils)
+
+library(mcmc)
 
 load(infile)  # has mrun and previous things (called 'datafile' in -inference.R)
 

### 2017-04-19 bcells/make-genmat.R

@@ -33,8 +33,9 @@ if (interactive()) { shortwin <- 2; boundary <- "none"; meanboundary <- 0 }
 
 bases <- c("A","T","C","G")
 
-source("../sim-context-fns.R",chdir=TRUE)
-source("../context-inference-fns.R",chdir=TRUE)
+library(contextual)
+library(contextutils)
+library(simcontext)
 
 mutpats <- getmutpats(patlen)
 # # DO NOT allow GC-bias

### 2017-04-19 compute-resids.R

@@ -22,7 +22,7 @@ if (is.null(opt$basedir)) { opt$basedir <- dirname(opt$infile) }
 print(opt) # this will go in the pbs log
 
 library(contextual)
-library(helpers)
+library(contextutils)
 
 options(error = print.and.dump)
 

### 2017-04-19 cpg/cpg-mcmc.R

@@ -26,11 +26,11 @@ if (interactive()) { nbatches <- 100; blen <- 10; restart <- FALSE; gmfile <- TR
 
 if (!'infile'%in%names(opt)) { stop("Run\n  cpg-mcmc.R -h\n for help.") }
 
-# options(error=traceback)
-scriptdir <- "../"
-source(paste(scriptdir,"context-inference-fns.R",sep=''),chdir=TRUE)
-source(paste(scriptdir,"sim-context-fns.R",sep=''),chdir=TRUE)
-require(mcmc)
+library(contextual)
+library(contextutils)
+library(simcontext)
+
+library(mcmc)
 
 basedir <- gsub(".RData","",infile,fixed=TRUE)
 load(infile)

### 2017-04-19 cpg/cpg-nocontext-inference.R

@@ -29,11 +29,11 @@ if (logfile=="" & !interactive()) {
     sink(file=logcon, type="output")   # send both to log file
 }
 
-scriptdir <- "../"
-source(paste(scriptdir,"context-inference-fns.R",sep=''),chdir=TRUE)
-source(paste(scriptdir,"sim-context-fns.R",sep=''),chdir=TRUE)
+library(contextual)
+library(contextutils)
+library(simcontext)
 
-require(mcmc)
+library(mcmc)
 
 load(infile)
 basedir <- gsub(".RData","",infile,fixed=TRUE)

### 2017-04-19 cpg/sim-cpg.R

@@ -45,8 +45,9 @@ if (!is.null(logfile)) {
 
 bases <- c("A","T","C","G")
 
-source("../sim-context-fns.R",chdir=TRUE)
-source("../context-inference-fns.R",chdir=TRUE)
+library(contextual)
+library(contextutils)
+library(simcontext)
 
 # maximum size of pattern (for simulation)
 mutpats <- c(

### 2017-04-19 fit-tree-model.R

@@ -33,8 +33,8 @@ if (is.null(opt$outfile)) {
 }
 print(opt) # this will go in the pbs log
 
-scriptdir <-  dirname(sub("--file=","",commandArgs()[grep("--file",commandArgs())]))
-source(file.path(scriptdir,"context-inference-fns.R"),chdir=TRUE)
+library(contextual)
+library(contextutils)
 
 options(error = print.and.dump)
 

### 2017-04-19 make-genmat.R

@@ -57,8 +57,8 @@ if (is.null(opt$outfile)) {
 basename <- gsub(".RData",'',opt$outfile)
 if (!file.exists(dirname(opt$outfile))) { dir.create(dirname(opt$outfile)) }
 
-scriptdir <-  dirname(sub("--file=","",commandArgs()[grep("--file",commandArgs())]))
-source(file.path(scriptdir,"context-inference-fns.R"),chdir=TRUE)
+library(contextual)
+library(contextutils)
 
 config <- read.config(opt$configfile)
 if (!is.null(config$tree)) {

### 2017-04-19 mtCDNA/mtCDNA-inference.R

@@ -44,9 +44,9 @@ if (basedir=="") { basedir <- gsub(".nuc","",infile,fixed=TRUE) }
 
 require(Biostrings)
 
-scriptdir <- "../"
-source(paste(scriptdir,"context-inference-fns.R",sep=''),chdir=TRUE)
-source(paste(scriptdir,"sim-context-fns.R",sep=''),chdir=TRUE)
+library(contextual)
+library(contextutils)
+library(simcontext)
 
 if (!file.exists(basedir)) { dir.create(basedir) }
 basename <- paste(basedir,"/win-",leftwin,"-",shortwin,"-",rightwin,sep='')

### 2017-04-19 mus/simple-stats.R

@@ -24,8 +24,8 @@ longwin <- shortwin + leftwin + rightwin
 
 if (opt$outfile=="") { outfile <- with(opt, paste(basedir,"/simple-stats",shortwin,"-",leftwin,"-",rightwin,".pdf",sep='')) }
 
-scriptdir <- "../"
-source(paste(scriptdir,"context-inference-fns.R",sep=''),chdir=TRUE)
+library(contextual)
+library(contextutils)
 
 countfiles <- list.files(basedir,paste("*",longwin,lrwin,"counts",sep="."), full.names=TRUE)
 names(countfiles) <- gsub(".*chr([0-9XY]*).*","chr\\1",countfiles)

### 2017-04-19 tasep/tasep-inference.R

@@ -30,9 +30,9 @@ if (logfile!="" & !interactive()) {
     sink(file=logcon, type="output")   # send both to log file
 }
 
-scriptdir <- "../"
-source(paste(scriptdir,"context-inference-fns.R",sep=''),chdir=TRUE)
-source(paste(scriptdir,"sim-context-fns.R",sep=''),chdir=TRUE)
+library(contextual)
+library(contextutils)
+library(simcontext)
 
 require(mcmc)
 

### 2017-04-19 testing-code/check-ising-genmats.R

@@ -1,8 +1,10 @@
 #!/usr/bin/Rscript
 
 # Run some tests!!
-source("../sim-context-fns.R",chdir=TRUE)
-source("../context-inference-fns.R",chdir=TRUE)
+
+library(contextual)
+library(contextutils)
+library(simcontext)
 
 bases <- c("X","O")
 

### 2017-04-19 testing-code/check-longmuts.R

@@ -2,8 +2,9 @@
 
 # a contrived example to test longer mutpats
 
-source("../sim-context-fns.R",chdir=TRUE)
-source("../context-inference-fns.R",chdir=TRUE)
+library(contextual)
+library(contextutils)
+library(simcontext)
 
 bases <- c("O","X")
 

### 2017-04-19 testing-code/tree-tests.R

@@ -1,6 +1,7 @@
 #!/usr/bin/Rscript 
 
-source("../context-inference-fns.R",chdir=TRUE)
+library(contextual)
+library(contextutils)
 
 
 ####################################

### 2017-04-19 testing-code/varying-selcoef-test.R

@@ -1,8 +1,8 @@
 #!/usr/bin/Rscript
 
-source("../context-inference-fns.R",chdir=TRUE)
-source("../sim-context-fns.R",chdir=TRUE)
-source("../input-output.R",chdir=TRUE)
+library(contextual)
+library(contextutils)
+library(simcontext)
 
 # this has X's as VERY deleterious; O's as VERY beneficial
 #  and  sp2 on a LONG branch

### 2017-04-19 tree-cpg/cpg-tree-inference.R

@@ -39,9 +39,9 @@ if (logfile!="") {
     sink(file=logcon, type="output", split=interactive())   # send both to log file
 }
 
-scriptdir <- "../"
-source(paste(scriptdir,"context-inference-fns.R",sep=''),chdir=TRUE)
-source(paste(scriptdir,"sim-context-fns.R",sep=''),chdir=TRUE)
+library(contextual)
+library(contextutils)
+library(simcontext)
 
 require(mcmc)
 

### 2017-04-19 writeup-plots/make-examples.R

@@ -1,7 +1,8 @@
 #/usr/bin/R --vanilla
 
-source("../sim-context-fns.R",chdir=TRUE)
-source("../context-inference-fns.R",chdir=TRUE)
+library(contextual)
+library(contextutils)
+library(simcontext)
 
 
 # EXAMPLE 1 in the talk

### 2017-04-21 shape/likelihood-surface.Rmd
more documentation changed update() to update_x() added a likelihood function factory

--- 2017-04-19 likelihood-surface.Rmd
@@
 }
-genmatrix@x <- do.call( update, c( list(G=genmatrix, mutrates=model.config$mutrates, selcoef=model.config$selcoef), model.config$fixfn.params ) )
+genmatrix@x <- do.call( update_x, c( list(G=genmatrix, mutrates=model.config$mutrates, selcoef=model.config$selcoef), model.config$fixfn.params ) )
 projmatrix <- collapsepatmatrix( ipatterns=rownames(genmatrix), leftwin=opt$leftwin, rightwin=opt$longwin-opt$shortwin-opt$leftwin, bases=genmatrix@bases )
@@
     names(fparams) <- fixparams(genmatrix)
-    genmatrix@x <- do.call( update, c( list( G=genmatrix,mutrates=params[1:nmuts(genmatrix)],selcoef=params[seq(1+nmuts(genmatrix),length.out=nsel(genmatrix))]), as.list(fparams) ) )
+    genmatrix@x <- do.call( update_x, c( list( G=genmatrix,mutrates=params[1:nmuts(genmatrix)],selcoef=params[seq(1+nmuts(genmatrix),length.out=nsel(genmatrix))]), as.list(fparams) ) )
     # this is collapsed transition matrix

### 2017-04-21 bcells/bcells-mcmc.R

@@ -89,7 +89,7 @@ lud <- function (params) {
     if (any(mutrates<0)) {
         return( -Inf )
     } else {
-        genmatrix@x <- update(genmatrix,mutrates=mutrates,selcoef=numeric(0))
+        genmatrix@x <- update_x(genmatrix,mutrates=mutrates,selcoef=numeric(0))
         subtransmatrix <- computetransmatrix( genmatrix, projmatrix, shape=shape, time="gamma" )
         # return POSITIVE log-posterior
         return( sum( counts * log(subtransmatrix) ) - sum(mutrates/priormeans) - shape/shapemean )

### 2017-04-21 contextual/R/patterns.R

@@ -82,12 +82,30 @@ selnames <- function (selpats,pretty=TRUE) {
 #' ipatterns are the "input" patterns, while fpatterns are the "final" projected patterns
 #' This function assumes that all input patterns are the same length.
 #'
-#' @return A (nbases)^k by (nbases)^{k-leftwin-rightwin} matrix projection matrix that
-#'    maps patterns onto the shorter patterns obtained by deleting leftwin
-#'    characters at the start and rightwin characters at the end.  
+#' @param ipatterns Long patterns index the rows of the result.
+#' @param leftwin Left offset of short patterns from long patterns.
+#' @param shortwin Length of short patterns.
+#' @param rightwin Right offset of short patterns from long patterns.
+#' @param fpatterns Short patterns, which index the columns of the result.
+#' @param bases Alphabet of bases.
+#' @return A (nbases)^k by (nbases)^{k-leftwin-rightwin} sparse matrix
+#' projection matrix that maps patterns onto the shorter patterns obtained by
+#' deleting leftwin characters at the start and rightwin characters at the end.  
+#'
+#' @examples
+#' bases <- c("X","O")
+#' # columns are the right-hand character
+#' collapsepatmatrix(ipatterns=getpatterns(2,bases), leftwin=1, fpatterns=bases)
+#' # columns are the middle two characters
+#' collapsepatmatrix(ipatterns=getpatterns(4,bases), leftwin=1, rightwin=1, bases=bases)
 #'
 #' @export
-collapsepatmatrix <- function (ipatterns, leftwin, shortwin=nchar(fpatterns[1]), rightwin=nchar(ipatterns[1])-shortwin-leftwin, fpatterns=getpatterns(nchar(ipatterns[1])-leftwin-rightwin,bases), bases ) {
+collapsepatmatrix <- function (ipatterns, 
+                               leftwin, 
+                               shortwin=nchar(fpatterns[1]), 
+                               rightwin=nchar(ipatterns[1])-shortwin-leftwin, 
+                               fpatterns=getpatterns(nchar(ipatterns[1])-leftwin-rightwin,bases), 
+                               bases ) {
     patlen <- nchar(ipatterns[1])
     shortwin <- patlen - leftwin - rightwin
     stopifnot(shortwin>0)

### 2017-04-22 writeups/check-likelihood.Rmd
a bit of reordering; un-exported a few things

@@
     for (k in 1:ndscale) {
-        subtransmatrix <- extend.expAtv( A=genmatrix, scale=scale, dscale=dscale, gAtv=subtransmatrix )  # add scale dscale
+        subtransmatrix <- contextual:::extend.expAtv( A=genmatrix, scale=scale, dscale=dscale, gAtv=subtransmatrix )  # add scale dscale
         # return log-likelihood (*not* negative log-likelihood)

### 2017-04-22 contextual/R/predict_counts.R

@@ -1,5 +1,5 @@
 
-#' @export
+#' Predict counts for a 'context' model
 predictcounts.context <- function (model, longwin=NULL, shortwin=NULL, leftwin=NULL, 
                                    initcounts=rowSums(model), mutrates=model@mutrates, 
                                    selcoef=model@selcoef, genmatrix=model@genmatrix, 
@@ -83,6 +83,7 @@ projectcounts <- function( counts, new.leftwin, new.shortwin, new.longwin, overl
 
 #' Compute expected counts of paired patterns
 #'
+#' @export
 predicttreecounts <- function (shortwin, leftwin=0, rightwin=0, initcounts, mutrates, selcoef, tlens, genmatrix, projmatrix, initfreqs, patcomp, ... ) {
     longwin <- leftwin+shortwin+rightwin
     if (missing(genmatrix)) { genmatrix <- makegenmatrix(patlen=leftwin+shortwin+rightwin,...) }

### 2017-04-22 contextual/tests/testthat/test_gammaAtv.R

@@ -17,7 +17,7 @@ expect_true(all(eM$values <= 0))
 scale <- 2
 shape <- 3
 scale.t <- max((-1)*diag(M))
-gM <- gammam(M,scale=scale,shape=shape)
+gM <- contextual:::gammam(M,scale=scale,shape=shape)
 
 true.gM <- matrix(0,nrow=nrow(M),ncol=ncol(M))
 for (k in 1:nrow(M)) { true.gM <- true.gM + outer(eM$vectors[,k],eM$vectors[,k]) * (1/scale)^shape / ((1/scale)-eM$values[k])^shape }
@@ -25,7 +25,7 @@ for (k in 1:nrow(M)) { true.gM <- true.gM + outer(eM$vectors[,k],eM$vectors[,k])
 test_that("gammam() computations agree", {
     expect_equal( rowSums(true.gM), rep(1,nrow(true.gM)) )
     ee( (true.gM), (gM) )
-    ee( (true.gM), (gammam(M,scale=scale,shape=shape,tol=1e-12)))
+    ee( (true.gM), (contextual:::gammam(M,scale=scale,shape=shape,tol=1e-12)))
 })
 
 # construct an answer:
@@ -33,28 +33,28 @@ test_that("gammam() computations agree", {
 #    which is 1/(1+scale*M[1,2])
 M <- matrix( c(-1,0,1,0), nrow=2 )
 scale <- 2
-gM <- gammam(M, scale=scale, shape=1, tol=1e-8)
+gM <- contextual:::gammam(M, scale=scale, shape=1, tol=1e-8)
 ee( gM, matrix( c(1/(1+scale*M[1,2]), 0, 1-1/(1+scale*M[1,2]), 1.0), nrow=2))
 
-gM <- gammam(M, scale=scale, shape=1, tol=1e-14)
+gM <- contextual:::gammam(M, scale=scale, shape=1, tol=1e-14)
 ee( gM, matrix( c(1/(1+scale*M[1,2]), 0, 1-1/(1+scale*M[1,2]), 1.0), nrow=2) )
 
 # check extend.gammam:
 # prob an exponential(theta) is greater than an exponential(1) is 1/(1+theta), so:
-expect_equal( 1/(1+0.25/.8), as.numeric(extend.gammam( 1/(1+0.25), eps=0.2 )) )
[…]

### 2017-04-25 scripts/simulation.Rmd
moved to scripts dir

--- 2015-08-02 simulation.Rmd
@@
 panderOptions("table.split.table", Inf)  # Let tables become arbitrarily wide rather than wrapping.
-source("context-inference-fns.R",chdir=TRUE)
+
+library(contextual)
+library(contextutils)
+
 top_n_bottom <- function(df, k) { # Take the top and bottom k rows

### 2017-04-26 writeups/context-dependence.tex
moved writeup-plots/
re-found figures for tex writeup

@@
   \begin{center}
-    \includegraphics[width=\textwidth]{../tasep/all-mcmc-runs}
+    \includegraphics[width=\textwidth]{writeup-plots/tasep_all-mcmc-runs}
   \end{center}
@@
   \begin{center}
-    \includegraphics{../writeup-plots/selsims-2013-05-28-17-12-0275615-traces}
+    \includegraphics{writeup-plots/selsims-2013-05-28-17-12-0275615-traces}
   \end{center}
@@
   \begin{center}
-    \includegraphics[width=\textwidth,page=1]{../ising/all-mcmc-runs}
+    \includegraphics[width=\textwidth,page=1]{writeup-plots/ising_all-mcmc-runs}
   \end{center}
@@
   \begin{center}
-    \includegraphics[width=\textwidth,page=2]{../ising/all-mcmc-runs}
+    \includegraphics[width=\textwidth,page=2]{writeup-plots/ising_all-mcmc-runs}
   \end{center}
@@
   \begin{center}
-    \includegraphics[width=\textwidth,page=3]{../ising/all-mcmc-runs}
+    \includegraphics[width=\textwidth,page=3]{writeup-plots/ising_all-mcmc-runs}
   \end{center}
@@
   \begin{center}
-    \includegraphics{../writeup-plots/selsims-2013-06-03-13-17-0790276-initfreqs}
+    \includegraphics{writeup-plots/selsims-2013-06-03-13-17-0790276-initfreqs}
   \end{center}
@@
   \begin{center}
-    \includegraphics{../writeup-plots/selsims-2013-06-03-13-17-0790276-mutrates}
+    \includegraphics{writeup-plots/selsims-2013-06-03-13-17-0790276-mutrates}
   \end{center}

### 2017-04-26 writeups/context-dependence.tex
rearranged writeup

@@
 or of several states evolved along a tree from a single starting point.
-In this paper,
-we propose one solution to this problem,
-motivated by the problem of inferring context-dependent mutation rates from diverged nucleotide sequences.
 
@@
 but this is impractical because the size of the matrix is equal to the number of possible configurations.
-\todo{I suggest pulling this sentence together with your "In this paper" sentence to form a separate paragraph.}
-As an alternative, we show that the conditional likelihood of local pattern count statistics
+
+In this paper, we propose a solution to this problem,
+showing that the conditional likelihood of local pattern count statistics
 can be well-approximated by marginalizing over a finite amount of surrounding context.
+Motivated by the problem of inferring context-dependent mutation rates from diverged nucleotide sequences,
+we extend the inference framework to observations on trees.
+\peter{something about efficient computation and sparse matrices}
 
@@
 %%%%%%% %%%%%%%%%%
-\section{Notation and Examples}
-
-Each example above describes \todo{Has a formulation as? Glauber can be greater than 1 D} a 1-dimensional grid of sites, each of which can take one of a finite set of states,
-and that switch randomly between states according to a local set of rules.
-Concretely, we suppose that we observe a finite collection of these sites at only a few times,
-and that the dynamics are Markov:
+\section{A general context-dependent model}
+
+Consider a 1-dimensional grid of sites, each of which can take one of a finite set of states,
+and that switch randomly between states according to a local set of rules;
+we then observe a finite collection of these sites at only a few times,
+Suppose that the dynamics are Markov:
 with the set of possible states by $\calS$,
@@
 If more than one pattern would match to cause the same change, their rates add.
-Such a rule is defined by its ``transition triple'' $(\mu,u,v)$.
+Such a rule is defined by its ``transition triple'' $(\mu,u,v)$,
+written as ``$u \to v$ at rate $\mu$''.
 The set of all transition triples, $\calT$, determines the time evolution of this process.
+
+\peter{INCLUDE SELECTION}
 
@@
   and if it is, moves there.
-  This process therefore has only one transition triple \todo{Not sure if reviewers will like having a different notation for triples so soon after you defined the parens notation}:
+  This process therefore has only one transition triple:
 
@@
   Ignoring transitions that don't change the state,
-  if patterns $u$ and $v$ differ at only one site then $u$ changes to $v$ at rate $1/(1+\exp(H(v)-H(u)))$ \todo{Shouldn't there be a $\lambda$ up top?}, so
+  if patterns $u$ and $v$ differ at only one site then $u$ changes to $v$ at rate $\lambda/(1+\exp(H(v)-H(u)))$,
+  so
+
+    \peter{reformulate with a potential}
 
@@
 also used for simulation,
-using a variant of the ``jump chain representation'', also known as the ``Gillespie algorithm.''
-.
+using a variant of the ``jump chain representation'', also known as the ``Gillespie algorithm''.
 Briefly, this  works by first sampling a homogeneous Poisson process of times and locations of possible changes,
@@
 
+\peter{shorten, refer to figure}
+
 The problem at hand is to infer the parameters of the model, given the state at time 0 and later at time $t$.
@@
 
+\peter{how to choose window size}
+
+%%%%%%
+\section{Full likelihood}
+
+\peter{shorten/edit this to follow on from the above}
+
+The s5f model is defined as follows:
+for each $(a,b,c,d,e,x) \subset \{\nA,\nC,\nG,\nT\}$,
+it gives a (relative) mutation rate
+\begin{align}
+    \gamma(c,x;a,b,d,e) = \text{( mutation rate of $c \to x$ in context $abcde$ )} .
+\end{align}
+Given these rates, the only unknown parameter is the total amount of time
+that the sequence has been mutated for.
+Suppose that $X(t)$ is a sequence of length $n$ evolving under this model,
+and let $G_n$ be the corresponding generator matrix,
+a $4^n \times 4^n$ matrix.
+(In the s5f model, the mutation rates are only specified in the middle of 5-mers,
+so the two bases on each end never mutate.)
+The transition probabilities are $e^{tG_n}$,
+and by marginalization,
+we can compute quantities like
+the probability for observing the middle set of basis
+given the flanking $\ell$ on each end:
+\begin{align}
+    p_{\ell,n-2\ell}(t;a,b)
+    &=
+    \P\{
+        (X_{\ell+1}(t), X_{\ell+2}(t), \ldots, X_{n-\ell}(t))=(b_1,b_2,\ldots,b_{n-2\ell}) \\
+    &\qquad \qquad
+        \given (X_{1}(0), X_{2}(0), \ldots, X_{n}(0))=(a_1,a_2,\ldots,a_{n})
+    \} .
+\end{align}
+If the size of the context, $\ell$, is big enough,
+then this probability will not depend significantly if the sequence $a$ is in fact embedded in a longer sequence
+(the error decreases exponentially with $\ell$).
+
+The sequence $X(t)$ can be constructed by beginning with $X(0)$,
+putting down a Poisson process of potential mutations of rate $\gamma = \max\{\gamma(c,x;a,b,d,e)\}$,
+and thinning appropriately.
+The process depends on the flanking 2bp,
+so any other mutation within 2bp of a site affects the process at that site,
+and the chance that this occurs depends on the flanking 4bp on either side.
+Concretely, the number of flanking positions in each direction
+that must be kept track of to decide what happens at a given site
+grows by 1 at rate $\gamma$, and by 2 at rate $\gamma$;
+so the amount of flanking sequence on each side affecting a given site at time $t$
+is the sum of a Poisson($\gamma t$) and twice an independent Poisson($\gamma t$).
+This has mean $3\gamma t$ and variance $5 \gamma t$;
+with high probability it will not exceed, say,
+$3\gamma t + 6 \sqrt{\gamma t}$;
+if $\gamma t < 1$ this is less than 9.
+
+How can we use this to compute the likelihood of a full sequence?
+Actually, we know more.
+For two positions $(i,s)$ and $(j,t)$ in sequence $\times$ time,
+let $r(i,s;j,t) = |j-i|/|t-s|$ be the slope of the line between them,
+and if $I$ and $J$ are subsets of sites, let $r(I,s;J,t) = \min\{r(i,s;j,t) : i\in I, j \in J\}$
+be the minimum.
+Then, we know that there is a speed of dependency propagation $r_*$
+such that if $r(I,s;J,t) > r_*$,
+then $X_I(s)$ and $X_J(t)$ are nearly independent.
+Now, fix $t$ and for each site $i$ let $N_i$ denote its $r_*$-neighborhood,
+i.e.,
+\begin{align}
+    N_i = \{ (j,s) : r(i,t; j,s) \le r_*, \; 0 \le s \le t \} .
+\end{align}
+These are depicted in figure~\ref{fig:neighborhoods}A.
+We will write $i \sim j$ if $|i-j| \le 2 r_* t$,
+or equivalently, if $N_i \cap N_j \neq \emptyset$,
+and $i \nsim j$ otherwise.
+We can think of $N_i$ as the parts of ancestral sequences in which a mutation might have affected the state at $(i,t)$.
+so if $i \nsim j$, then $i$ and $j$ are (approximately) independent at time $t$, given the sequence at time 0.
+Concretely,
+we have a small $\epsilon$ such that for any $i \nsim j$,
+\begin{align}
+    &\big|
+    \log( \P\{ X_i(t) = x_i \and X_j(t) = x_j \} ) \\
+    &\qquad {} -
+    \log( \P\{ X_i(t) = x_i \} \P\{ X_j(t) = x_j \} )
+    ) \big|
+    \le \epsilon .
+\end{align}
+
+\begin{figure}
+    \begin{center}
+        \includegraphics{context-neighborhoods-defns}
+
+        \vspace{2em}
+
+        \includegraphics{context-neighborhoods-widths}
+    \end{center}
+    \caption{
+        Diagram of things in the text:
+        (top) Dependency neighboorhods.
+        (bottom) What we need to compute $\P\{X_i(t)=x \given X(0), X_{<i}(t)\}$:
+        site $i$ depends on the blue squares in $N_i$;
+        the sites to the left of this in $L(i)$ are those other sites in $X(t)$
+        whose dependency neighborhoods overlap $N_i$;
+        any further sites to the left of this are basically independent of $X_i(t)$;
+        the red squares give the dependency neighborhood of $L(i)$,
+        so we need to condition on the row of red and blue sites at the top.
+        On the bottom, we have $2w+1$ sites; at the top we have $4w+1$;
+        here $w=2$.
+        \label{fig:neighborhoods}
+    }
+\end{figure}
+
+What we'd like to know is the likelihood, i.e.,
+\begin{align}
+    \like(y|x)
+    &:=
+    \P\{ X(t) = y \given X(0) = x \} .
+\end{align}
+If we define $X_{<i}=(X_1, \ldots, X_{i-1})$, then
+\begin{align}
+    \like(y|x)
+    &=
+    \prod_{i=1}^n
+    \P\{ X_i(t) = y_i \given X(0) = x \and X_{<i}(t) = y_{<i} \} .
+\end{align}
+Now, defining $L(i) = \{ j < i \st j \sim i \}$
+to be the sites to the left of $i$ whose dependency neighborhoods overlap with $i$'s,
+and $A(I) = \{ j \st |j-i| \le r_*t \text{ for some } i \in I\}$
+to be the ancestral sites in the dependency neighborhood of $I$,
+\begin{align}
+&\big|
+\log( \P\{ X_i(t) = y_i \given X(0) = x \and X_{<i}(t) = y_{<i} \} ) \\
+& \qquad {} -
+\log( \P\{ X_i(t) = y_i \given X_{A(L(i))}(0) = x_{A(L(i))} \and X_{L(i)}(t) = y_{L(i)} \} )
+ \big| \\
+&\qquad \qquad \le
+    \epsilon ,
+\end{align}
+so that if we define
+\begin{align}
+    \alike(y|x)
+    &=
+    \prod_{i=1}^n
+    \P\{ X_i(t) = y_i \given X_{A(L(i))}(0) = x_{A(L(i))} \and X_{L(i)}(t) = y_{L(i)} \} ,
+\end{align}
+then (with $\loglike = \log \like$),
+\begin{align}
+    \left| \aloglike(y|x) - \loglike(y|x) \right|
+    \le n \epsilon .
+\end{align}
+
+This is good, because we can compute $\aloglike$:
+let $w=r_*t$ be radius of a dependency neighborhood,
+so that the width of $L(i)$ is $2w$,
+and for a sequence $x$ define $x^{(m)}_i = (x_{i-m+1}, \ldots, x_i)$:
+\begin{align}
+    \alike(y|x)
+    &=
+    \prod_{i=1}^n
+    \frac{
+        p_{w,2w+1}(y^{(2w+1)}_i|x^{(4w+1)}_{i+w})
+    }{
+        p_{w,2w}(y^{(2w)}_{i-1}|x^{(4w)}_{i-1+w})
+    }
+\end{align}
+This is the ratio of the likelihoods of all the $(w,2w+1)$ T-mers,
+divided by the likelihoods of all the $(w,2w)$ T-mers.
+
+%%%%%%%%%%%
+\section{Computation}
+
+For inference, we need to compute the $|\calS|^{m+2\ell} \times |\calS|^m$ matrix $F$ whose $(x,y)^\text{th}$ entry is $p_{m,\ell}(t,x,y)$,
+for various $t$.
+This matrix is a projection of the $|\calS|^{m+2\ell} \times |\calS|^{m+2\ell}$ matrix whose $(x,z)^\text{th}$ entry is
+\begin{align}
+    p_{m+2\ell}(x,z) = \left( e^{t G(m+2\ell)} \right)_{x,z} ,
+\end{align}
+where $G(m+2\ell)$ is the sparse $|\calS|^{m+2\ell} \times |\calS|^{m+2\ell}$ matrix defined in \eqref{eqn:G_defn}.
+The ``projection'' we need just marginalizes over long patterns $z$ that match the shorter pattern $y$:
+if we define the matrix $U$ in this way, so that $U_{zy}=1$ if $z_\ell^m=y$ and $U_{zy}=0$ otherwise,
+then
+\begin{align} \label{eqn:Tmer_trans}
+    p_{m,\ell}(t,x,y) = \sum_{z \in \calS^{m+2\ell}} \left( e^{t G(m+2\ell)} \right)_{xz} U_{zy} ,
+\end{align}
+so in fact we don't need the entire matrix $e^{t G(m+2\ell)}$,
+just the product of this matrix with each of the columns of $U$.
+
+Modern techniques in sparse matrix computation (Krylov methods) provide efficient ways to do this
+by only multiplying $G$ and $U$ some number of times,
+which is much more efficient because $G$ is sparse.
+
+\subsection{Updating $G$}
+
+We can also take advantage of the sparsity of $G$ to perform efficient computation of the likelihood
+under many sets of parameters.
+First note that $G(m+2\ell)$ has, assuming only single-position changes,
+$(1+m(|\calS|-1)) |\calS|^{m+2\ell}$ nonzero entries, since each of the $|\calS|^{m+2\ell}$ can change in $m$ places.
+This will determine how the computation scales with $m$, $\ell$, and $|\calS|$,
+once we precompute a number of things.
+Let $g = (g_1, \ldots, g_d)$ be the nonzero entries of $G$, in some fixed ordering;
+sparse matrix representations of $G$ store only $g$ along with information about the rows and columns these are found in.
+Each $g_i$ is a linear combination of mutation rates $\mu_j$, say $g_i = \sum_j a_{ij} \mu_j$;
+so by precomputing $a_{ij}$ we can efficiently update $G$.
+This is true if the boundary conditions are circular, mean-value, or neither.
+
+We can do similar precomputations to help with the case of selective differences:
+XXX describe
+
+%%%%%%%%
+\section{Phylogenetic inference}
+
+In phylogenetic applications, rather than ``before'' and ``after'' observations,
+we get two (or more) observations evolved from a common root.
+In the simplest case of two taxa we have two processes $X$ and $Y$,
+with identical starting states $X(0)=Y(0)$,
+observed only at times $t_X$ and $t_Y$ respectively.
+
+If the approximation described above holds,
+then we can write the probability that we see (long) pattern $x$ in $X(t_X)$ juxtaposed with (short) pattern $y$ in $Y(t_Y)$
+using the pattern frequencies at the root.
+Concretely: pick a random location $I$ in the sequence and let $X_{I-\ell}^{m+2\ell}(0) = \rho$ be the (long) pattern seen there at the root,
+and write $\pi(z)$ for the frequency of pattern $z$ in $X(0)$, i.e.\ $\P\{\rho=z\}=\pi(z)$.
+% Given the corresponding pattern in $X(t_X)$, the distribution of $\rho$ is proportional to
+% \[
+%     \P\{ \rho=z \mid X_{I-\ell}^{m+2\ell}(t_X)=x\} \propto \pi(z) p_{m+2\ell}(t,z,x) .
+% \]
+The probability that we see $x$ and $y$ at the random location $I$ is
+\begin{align} \label{eqn:phylo_likelihood}
+    \P\{X_{I-\ell}^{m+2\ell}(t_X)=x \text{ and } Y_I^\ell(t_Y)=y \} = \sum_{z \in \calS^{m+2\ell}} \pi(z) p_{m+2\ell}(t_X,z,x) p_{m,\ell}(t_Y,z,y) .
+\end{align}
+This can be computed without much more effort than the simpler case above,
+given the frequencies at the root, as described in greater generality below.
+
+\subsection{Phylogenetic peeling}
+
+Furthermore, note that the further calculations in the phylogenetic case can be done using these methods.
+To compute
+\[
+    \sum_z \pi(z) p_{m+2\ell}(t_X,z,x) p_{m,\ell}(t_Y,z,y)
+    = \sum_z \left( e^{t_X G(m+2\ell)} \right)_{zx} \pi(z) \sum_w \left( e^{t_X G(m+2\ell)} \right)_{zw} U_{wy} ,
+\]
+we can first compute $\sum_w \left( e^{t_X G(m+2\ell)} \right)_{zw} U_{wy}$ as above,
+multiply by $\pi(z)$, and then apply the algorithm again using the transpose of $G(m+2\ell)$.
+
+Using the R package expm, this makes computation quite feasible:
+with four possible states and $\ell=2$ and $m=1$, so that $G(m+2\ell)$ is a $1024 \times 1024$ matrix,
+computing $e^{t G(5) }$ takes 13 seconds, while computing $e^{t G(5)} U$ takes only 0.3 seconds.
+Increasing to $\ell=4$ and $m=1$ or $\ell=3$ and $m=2$ is still feasible, taking $e^{t G(8)} U$ in 42 and 47 seconds, respectively
+(and much longer for the whole matrix $e^{t G(8)}$).
+
+More generally\dots
+
+\begin{center}
+\includegraphics{peeling-schematic}
+\end{center}
+
+To compute the likelihood on a more complicated tree, consider the following,
+with the three-taxon tree above as an example,
+where we are counting occurrences of (long) $m+2\ell$-tuples in taxon $X$,
+and (shorter) $m$-tuples in taxa $Y$ and $Z$.
+
+We can compute the likelihood in a similar way, summing over the root state (the summation over $u$)
+and the interior node ($v$):
+\begin{align} \label{eqn:three_taxa_likelihood}
+  \begin{split}
+    & \P\{X_{I-\ell}^{m+2\ell}=x \text{ and } Y_I^\ell=y \text{ and } Z_I^\ell=z \} \\
+    &\qquad = \sum_{u \in \calS^{m+2\ell}} \pi(u) p_{m,\ell}(t_C,u,z) \sum_{v \in \calS^{m+2\ell}} p_{m+2\ell}(t_{AB},u,v) p_{m+2\ell}(t_{A},v,x) p_{m,\ell}(t_B,v,y)
+  \end{split}
+\end{align}
+The steps for computing this are then:
+\begin{enumerate}
+
+  \item Compute $M_1(u,z) = p_{m,\ell}(t_C,u,z)$ (a $|\calS|^{m+2\ell} \times |\calS|^m$ matrix)
+
+  \item Compute $M_2(v,z) = \sum_{u \in \calS^{m+2\ell}}  p_{m+2\ell}(t_{AB},u,v) \pi(u) M_1(u,z)$ (still a $|\calS|^{m+2\ell} \times |\calS|^m$ matrix)
+
+  \item Compute $M_3(v,y) = p_{m,\ell}(t_B,v,y)$ (a $|\calS|^{m+2\ell} \times |\calS|^m$ matrix)
+
+  \item Compute $M_4(x,(y,z)) = \sum_{v \in \calS^{m+2\ell}} p_{m+2\ell}(t_{A},v,x) M_2(v,z) M_3(v,y)$ (stored as a $|\calS|^{m+2\ell} \times |\calS|^{2m}$ matrix).
+
+\end{enumerate}
+Then $M_4$ gives all the likelihoods.
+This is depicted in the figure, writing $\left(e^{tG}\right)_{ab}$ for the matrix $p_{m+2\ell}(t,a,b)$,
+and $P$ for the projection matrix that gives $p_{m,\ell}(t,a,b) = \left( e^{tG} P\right)_{ab}$.
+
+In total there are $|\calS|^{3m+2\ell}$ possible data combinations.
+However, we can reduce dimensionality by replacing the last step with
+\begin{enumerate}
+
+  \item[4'.] Compute $M_4(x,w) = \sum_{v \in \calS^{m+2\ell}} p_{m+2\ell}(t_{A},v,x) M_3(v,w) M_4(v,w)$ (stored as a $|\calS|^{m+2\ell} \times |\calS|^{m}$ matrix).
+
+\end{enumerate}
+and only computing the likelihood for cases with $y=z=w$.
+On a three-taxon tree it is not clear that this is desireable, since missing out on double transitions
+might be a significant loss of information,
+but on larger trees it seems likely that we'd want to restrict to combinations with most sites agreeing across the tree.
+(Note, however, that one should not first look at the data, observe which combinations $(y,z)$ were most common, and then restrict to only those!)
+
+The algorithm in general is written out in Appendix \peter{ref it}.
+
+%%%%%%%
+\section{Results}
+
+\peter{from simulation: parameter estimation; model selection}
+
+\peter{do a three-taxon tree? infer with shape data?}
+
+The method works well to find point estimates of the parameters.
+Obtaining posterior density estimates using the approximation likelihood function on a subset of patterns
+works in some cases;
+in other cases the likelihood surface is still too peaked, underestimating the uncertainty.
+(but counting in nonoverlapping windows fixes the peakiness problem)
+
+\subsection{TASEP}
+
+Strings of length $10^5$ were evolved under TASEP for varying amounts of time,
+and pattern occurrences counted.
+Running 10,000 MCMC iterations with window sizes of $\ell=w=3$ (total length 9) on these takes around 40 minutes.
+Apparently, the patterns used were not sufficiently independent in this case,
+as the posterior is too sharp.
+
+Here's a short example sequence with $\lambda=1$:
+\begin{center}
+ \setlength{\tabcolsep}{0pt}
+\begin{tabular}{cccccccccccccccccccccccccccccccccccccccccccccccccccccccccccc}
+X&X&X&O&O&X&O&O&O&O&X&O&X&O&X&X&O&O&X&O&O&X&O&X&X&X&O&X&O&O&O&X&X&O&X&O&O&O&X&X&O&O&O&X&O&O&O&O&X&O&X&X&O&X&X&O&O&X&O&O \\
+$\centerdot$&O&X&O&X&O&O&O&X&$\centerdot$&O&X&O&O&X&X&X&o&x&$\centerdot$&o&O&X&x&x&O&X&O&X&o&o&O&X&X&$\centerdot$&$\centerdot$&o&o&O&X&X&o&$\centerdot$&O&X&$\centerdot$&o&o&$\centerdot$&$\centerdot$&O&X&X&x&O&X&o&O&X&$\centerdot$
+\end{tabular}
+\end{center}
+here particles are X's, moving to the right; and spaces are O's;
+in the second line,
+$\centerdot$ is a position that has not considered a change; lowercase letters are positions that considered a change but did not;
+and upper case are positions that did change.
+
+\begin{figure}
+  \begin{center}
+    \includegraphics[width=\textwidth]{writeup-plots/tasep_all-mcmc-runs}
+  \end{center}
+  \caption{ Results from MCMC estimates for six different data sets of the parameter in TASEP (mutation rate multiplied by time).
+  Vertical line shows the true value.
+  }
+\end{figure}
+
+\subsection{Ising model}
+
+Strings of length $10^6$ were evolved under Glauber dynamics on the Ising model with varying parameters,
+and pattern occurrences counted.
+As for TASEP, running 10,000 MCMC iterations with window sizes of $\ell=w=3$ (total length 9) on these takes around 40 minutes.
+In this case, the true parameter values have good distribution within the estimated posterior distributions.
+
+Here is an example, with all parameters equal to 1, for $t=.5$:
+\begin{center}
+ \setlength{\tabcolsep}{0pt}
+\begin{tabular}{cccccccccccccccccccccccccccccccccccccccccccccccccccccccccccc}
+X&X&O&X&X&O&O&X&O&X&O&X&X&O&O&X&O&O&O&O&X&X&X&O&X&O&X&O&X&O&O&O&X&X&O&X&X&X&X&X&O&X&O&X&O&X&X&X&X&O&O&X&X&X&X&X&X&O&O&O \\
+x&x&o&x&X&X&O&x&o&x&$\centerdot$&x&x&O&X&X&$\centerdot$&O&X&O&X&x&x&o&x&o&x&o&x&O&X&O&$\centerdot$&$\centerdot$&O&O&X&X&O&X&X&X&o&x&o&$\centerdot$&X&O&X&o&o&x&X&X&X&O&X&o&$\centerdot$&$\centerdot$
+\end{tabular}
+\end{center}
+The notation is as above; except that the entire matched pattern involved in a change is uppercase:
+to simulate, we look in windows of size 3; so if XOO changes to XXO, the whole thing is uppercase, not just the middle X.
+
+\begin{figure}
+  \begin{center}
+    \includegraphics{writeup-plots/selsims-2013-05-28-17-12-0275615-traces}
+  \end{center}
+  \caption{
+  MCMC traces of the parameters in the Ising model across 100,000 iterations.
+  The units of $\lambda$ are in mean number of changes per site.
+  Vertical line shows the true value.
+  }
+\end{figure}
+
+\begin{figure}
+  \begin{center}
+    \includegraphics[width=\textwidth,page=1]{writeup-plots/ising_all-mcmc-runs}
+  \end{center}
+  \caption{ Results from MCMC estimates of the time-scaled mutation rate parameter $\lambda t$,
+  in ten different simulated datasets under the Ising model.
+  Vertical line shows the true value.
+  }
+\end{figure}
+
+\begin{figure}
+  \begin{center}
+    \includegraphics[width=\textwidth,page=2]{writeup-plots/ising_all-mcmc-runs}
+  \end{center}
+  \caption{ Results from MCMC estimates of the inverse temperature $\beta$ in the Ising model.
+  Vertical line shows the true value.
+  }
+\end{figure}
+
+\begin{figure}
+  \begin{center}
+    \includegraphics[width=\textwidth,page=3]{writeup-plots/ising_all-mcmc-runs}
+  \end{center}
+  \caption{ Results from MCMC estimates of the magnetization parameter $\beta$ in the Ising model.
+  Vertical line shows the true value.
+  }
+\end{figure}
+
+%%%
+\subsection{CpG mutation, on a tree}
+
+Nucleotide sequences of length $10^6$ were simulated down two (unequal) branches of a tree,
+with all single-base substitution rates $m_{xy}$ set equal (varying from a total of .01 to .1 across the tree)
+and an additional CpG rate $\gamma$ set to either 0 or three times the single-base rate.
+Analysis was done with $\ell=2$ and $w=1$ (a five-to-one window).
+
+The state at the root of the tree was independently chosen bases with equal frequencies,
+and these frequencies at the root were added as parameters to the model (with a Dirichlet prior).
+
+Here is a short example, with all parameters equal to 1, with each branch of the tree having length $0.1$:
+\begin{center}
+ \setlength{\tabcolsep}{0pt}
+\begin{tabular}{cccccccccccccccccccccccccccccccccccccccccccccccccccccccccccc}
+T&G&G&A&G&A&G&G&T&C&T&G&C&C&A&T&A&G&A&C&G&G&G&C&G&C&C&G&C&C&T&A&G&G&A&G&C&C&G&C&G&T&T&T&T&G&C&G&T&G&G&T&G&T&T&G&T&G&G&C \\
+$\centerdot$&$\centerdot$&g&C&G&a&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&T&A&C&A&$\centerdot$&$\centerdot$&$\centerdot$&T&A&T&$\centerdot$&$\centerdot$&$\centerdot$&T&G&c&C&C&A&C&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&G&G&G&t&A&T&$\centerdot$&$\centerdot$&G&A&g&C&G&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&t&T&G&A \\
+$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&A&T&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&T&T&G&T&$\centerdot$&$\centerdot$&G&A&T&$\centerdot$&$\centerdot$&$\centerdot$&G&G&c&t&A&A&G&C&$\centerdot$&$\centerdot$&$\centerdot$&G&T&G&A&t&t&t&g&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&G&A&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&A&C
+\end{tabular}
+\end{center}
+Top is ancestral sequence (``unobserved''); bottom two are at the tips.
+
+The point estimates are good,
+as are the MCMC estimates,
+but there seems to be mixing trouble having to do with the root frequencies.
+
+\begin{figure}
+  \begin{center}
+    \includegraphics{writeup-plots/selsims-2013-06-03-13-17-0790276-initfreqs}
+  \end{center}
+  \caption{
+  MCMC traces across 660,000 iterations of the frequencies at the root of the (two-taxon) tree in the CpG model.
+  Vertical line shows the true value.
+  }
+\end{figure}
+
+\begin{figure}
+  \begin{center}
+    \includegraphics{writeup-plots/selsims-2013-06-03-13-17-0790276-mutrates}
+  \end{center}
+  \caption{
+  MCMC traces across 660,000 iterations of several mutation rate parameters in the CpG model.
+  Units are in mean number of substitutions per base across the entire tree.
+  Vertical line shows the true value.
+  }
+\end{figure}
+
+%%%%%%%%
+\section{Power sims}
+
+Assess power and accuracy in the Ising model, by running replicates at these parameters:
+
+\begin{tabular}{|rlllrrr|}
+  \hline
+  seqlen   &  $\lambda t$   &  $\beta$  &  $\gamma$  &  $w$ &  $\ell$   & \# reps \\
+  \hline
+  $10^4$    &   .2  &   1   &   .5  &   3   &   3   &   10  \\
+  $10^3$    &   .2  &   1   &   .5  &   3   &   3   &   10  \\
+  $10^2$    &   .2  &   1   &   .5  &   3   &   3   &   10  \\
+  \hline
+  $10^4$    &   .4  &   1   &   .5  &   3   &   3   &   10  \\
+  $10^4$    &   .1  &   1   &   .5  &   3   &   3   &   10  \\
+  $10^4$    &   .05  &   1   &   .5  &   3   &   3   &   10  \\
+  \hline
+  $10^4$    &   .2  &   1   &   0  &   3   &   3   &   10  \\
+  \hline
+  $10^4$    &   .2  &   1   &   .5  &   3   &   2   &   10  \\
+  $10^4$    &   .2  &   1   &   .5  &   3   &   1   &   10  \\
+  $10^4$    &   .2  &   1   &   .5  &   3   &   0   &   10  \\
+  \hline
+\end{tabular}
+
+\section{Discussion}
+
+\begin{itemize}
+
+  \item estimating freq spectrum [Hernandez]
+
+  \item probabilistic ancestral states for common ancestor of e.g. humans 4N generations ago [Siepel] -- would widely used for primates
+
+  \item ancestral CpG
+
+  \item ancient DNA
+
+  \item noisy image transmission denoising: first learn model; then reduce redundancy
+
+  \item root distribution and confounding/mixing
+
+  \item shrinkage prior on highly parameterized model
+
+  \item Show that taking no context ($\ell=0$) can be positively misleading in some context.
+
+  \item How to identify motifs in the residuals?
+
+\end{itemize}
+
+\subsection{Acknowledgements}
+Erick, Jessica, Graham, Yaniv, Sergey, Simon, Matt, Rasmus
+
+\bibliography{context-dependence}
+
+\appendix
+
 %%%%
-\subsection{Summary statistics}
-
-Since we cannot compute reasonably the full likelihood for the data,
-we will need to choose appropriate summary statistics
-that contain as much information about the process as possible,
-and have a tractable expression for their marginal likelihood.
-For simplicity,
-we assume in this section that approximation \eqref{eqn:window_approx} holds exactly.
-If $i_1 < \ldots < i_{n(x)}$ are a set of nonoverlapping sites where the initial sequence matches $x$
--- i.e.\ $X_{i_k}^{(2\ell+m)}(0)=x$ and $i_{k+1} > i_k + 2\ell+m$ for all $k$ %En dash close to math could be confusing. This sentence is a doozy.
-then the counts $N(x,y)=\#\{ i_k \st X_{i_k+\ell}^{(m)}(t)=y\}$ (the number of these sites that match the smaller pattern $y$ at time $t$)
-are multinomial, with probabilities $\{p_{\ell,m}(x,y)\st y \in \calS^m\}$.
-For instance, if $\ell=3$ then $N(\text{ACTCAC}, \text{CGC})$ counts the number of occurrences of
-\begin{center}
-\begin{tabular}{c|ccccccc}
- $x$ &  C  & A & C & T & C & A & C \\
- $y$ &  $\centerdot$  & $\centerdot$  & C & G & C & $\centerdot$  & $\centerdot$
-\end{tabular}
-\end{center}
-where $\cdot$ represents any base.
-More generally, if we partition the sequence into windows of length $2\ell+m$,
-and let $N(x,y)$ be the number of times we see a window in which $x$ matches $X(0)$ and $y$ matches $X(t)$,
-then these counts are independently multinomial across each $x$.
-% so the marginal likelihood of the statistics $N(x,y)$ is
-% \begin{align}
-%     \prod_{x \in \calS^{2\ell+m}} N(x)! \prod_{y \in \calS^m} \frac{p_{\ell,m}(x,y)^{N(x,y)}}{N(x,y)!} ,
-% \end{align}
-% where $N(x) = \sum_y N(x,y)$.
-
-Of course, a drawback to this approach is that we throw away a good fraction of the data.
-In fact, we do not even look at a fraction $2\ell/(2\ell+m)$ of the second sequence, $X(t)$,
-that lies in the outside edges of each window.
-As before, let $N(x)$ be the number of occurrences of a pattern $x$ in $X(0)$,
-and $N(x,y)$ the number of times $x$ is seen juxtaposed with $y$ in $X(t)$,
-but this time, count over all positions,
-so that e.g.\ $N(x) = \#\{ 1 \le i \le n - 2\ell - m + 1 \st X_i^{(m+2\ell)}(0)=x\}$.
-
-Now we will now calculate the expectation and covariances of $N(x,y)$.
-\peter{The following is Not Right, but also we don't really use it at the moment.}
-If we assume that \eqref{eqn:window_approx} is exact, then
-\[
-    \E [N(x,y)] = N(x) \, p_{\ell,m}(x,y).
-\]
-To compute covariances, define $\chi_i(x,y)$ to be 1 if $X_{i-\ell}^{(|x|)}(0)=x$ and $X_i^{(|y|)}(t)=y$, and 0 otherwise,
-so that $N(x,y) = \sum_{i=1}^n \chi_i(x,y)$.
-Overloading notation, also define $\chi_i(x)$ to be 1 if $X_{i-\ell}^{(|x|)}(0)=x$,
-\begin{align}
-  \var[N(x,y)] &= \sum_{i=1}^n \var[\chi_i(x,y)] + \sum_{i\neq j} \cov[\chi_i(x,y), \chi_j(x,y)] \\
-               &\approx \sum_{i=1}^n \chi_i(x) p_{\ell,m}(x,y)(1-p_{\ell,m}(x,y))
-            + 2 \sum_{i=1}^{n-2\ell-m} \sum_{j=i+1}^{i+2\ell+m-1} \cov[ \chi_i(x,y) \chi_j(x,y) ] \\
-            &= N(x) p_{\ell,m}(x,y)(1-p_{\ell,m}(x,y)) +
-\end{align}
-The term $\E[\chi_1(x,y) \chi_{k+1}(x,y)]$ is the probability that $x$ matches at both site 1 and site $k+1$,
-and that $y$ matches both at site $\ell+1$ and at site $\ell+k+1$.
-For this to be nonzero, the joint pattern $(x,y)$ must allow self-overlaps.
-Similarly, if $(x_1,y_1) \neq (x_2,y_2)$ then
-\begin{align}
-  \begin{split}
-    \cov[N(x_1,y_1),N(x_2,y_2)]
-    & \approx - n (2m+4\ell) p_{\ell,m}(x_1,y_1)p_{\ell,m}(x_2,y_2) \\
-    & \qquad {} + n\sum_{k=1}^{m+2\ell} \E[\chi_1(x_1,y_1) \chi_{k+1}(x_2,y_2) + \chi_1(x_2,y_2) \chi_{k+1}(x_1,y_1)] .
-  \end{split}
-\end{align}
-
-%%%%
-\subsection{A likelihood function}
-
-The covariance calculations show that
-if $p_{\ell,m}(x,y)$ is small enough (i.e.\ the pattern is not common),
-then $N(x,y)$ is close to Poisson.
-Using this observation, we can estimate parameters by composite likelihood,
-using all $N(x,y)$ as data.
-Specifically, this composite likelihood is
-\begin{align} \label{eqn:comp_like}
-  \prod_{x,y} F_{(x,y)}^{N(x,y)}
-\end{align}
-where $F$ is the $|\calS|^{m+2\ell} \times |\calS|^m$ matrix whose $(x,y)^\text{th}$ entry is $p_{m,\ell}(t,x,y)$ as described in the "Computation" section.
-Maximum likelihood then gets an asymptotically unbiased estimate of the parameters. (cite)
-However, since they are not independent, to get confidence intervals, we need to either simulate, or do something different.
-
-If we use only a subset of patterns $\{(x_k,y_k)\}$ chosen so that $\E[M_1(x,y) M_{k+1}(x,y)]$ is small (or zero),
-\erick{M not defined, but I'm guessing that this is effectively saying that the probability of having two different patterns apply to the same sequences is small?}
-the entire collection $\{N(x_k,y_k)\}$ will be close to independent Poisson.
-(using Stein's method)
-Then we'll have the correct likelihood function for (this subset of) data.
-
-If changes are rare, then one way to get a set of patterns with this property is to only choose ones such that going from $x$ to $y$ entails a change;
-and any overlapping patterns must require additional changes,
-i.e.\ if $(x_1,y_1)$ matches at $i_1$,
-then for $(x_2,y_2)$ to match at $i_2$,
-there must be additional changes outside $(i_1,\ldots,i_1+\ell-1)$.
-
-Here is one way to obtain such a set.
-Take a pair of patterns $(x,y)$, that overlap at $m$ positions,
-and suppose that they differ at $1 \le i_1, \ldots, i_d \le m$.
-Let $\bar i = (1/d) \sum_{j=1}^d i_j$ be the mean position of the changes,
-and take only patterns with $(m-1)/2 < \bar i \le (m+1)/2$.
-For example, in the example above with $m=3$ and $\ell=2$,
-\begin{center}
-\begin{tabular}{c|ccccccc}
- $x$ &  C  & A & C & T & C & A & C \\
- $y$ &  $\centerdot$  & $\centerdot$  & C & G & C & $\centerdot$  & $\centerdot$
-\end{tabular}
-\end{center}
-$x$ and $y$ differ only at (relative) position $i_1=2$.
-
-Another way to achieve the same goal of having all overlapping patterns require additional changes is to require $i_1 = \bar i = 1$.
-No two such patterns can be shifted versions of each other
-without adding more changed sites,
-since shifting by $k$ without adding changed sites just adds $k$ to $\bar i$.
-For example, we might take
-\begin{center}
-\begin{tabular}{c|ccccccc}
- $x$ &  A & C & T & C & A & C & G \\
- $y$ &  $\centerdot$  & $\centerdot$  & G & C & A & $\centerdot$  & $\centerdot$
-\end{tabular}
-\end{center}
-
-%%%%
-\subsection*{Proof of the approximation}
+\section*{Proof of the approximation}
 
@@
 
-%%%%%%%%
-\subsection{Edge effects and cyclicization}
-
-The choice to treat long sequences as cyclical is generally of little importance.
-However, we have defined in \eqref{eqn:G_defn} the basic building block of our approximation, $G(m)$,
-to be the transition matrix \emph{for a circular sequence} of length $m$, where $m$ may be short.
-This seems strange -- but, what else would we do at the edges?
-We want to use $G(\ell+m+r)$ to compute the chance that a given $(\ell+m+r)$-sequence ends up with another given $m$-sequence in the middle;
-and if $\ell$ and $r$ are big enough relative to $t$, this will be a good approximation regardless of what we do at the edges.
-One answer would be to not allow patterns hanging off of either end to match,
-resulting in a slight underestimate of the chance of change.
-A better approximation would be to use the marginal frequencies in some way to average over surroudning sequence
--- but, this requires marginal frequencies to be known, and these are not stationary.
-However, the leftmost site in a randomly chosen window is a sample from the marginal distribution,
-and if sites separated by distance $\ell+m+r$ are sufficiently uncorrelated,
-wrapping the sequence around is a good approximation to averaging over marginal frequencies in the surrounding sequence.
-It is easy to imagine situations where this would not be the best solution --
-but again, if $\ell$ and $r$ are large enough, it won't matter much.
-
-%%%%%%%%
-\section{Phylogenetic inference}
-
-In phylogenetic applications, rather than ``before'' and ``after'' observations,
-we get two (or more) observations evolved from a common root.
-In the simplest case of two taxa we have two processes $X$ and $Y$,
-with identical starting states $X(0)=Y(0)$,
-observed only at times $t_X$ and $t_Y$ respectively.
-
-If the approximation described above holds,
-then we can write the probability that we see (long) pattern $x$ in $X(t_X)$ juxtaposed with (short) pattern $y$ in $Y(t_Y)$
-using the pattern frequencies at the root.
-Concretely: pick a random location $I$ in the sequence and let $X_{I-\ell}^{m+2\ell}(0) = \rho$ be the (long) pattern seen there at the root,
-and write $\pi(z)$ for the frequency of pattern $z$ in $X(0)$, i.e.\ $\P\{\rho=z\}=\pi(z)$.
-% Given the corresponding pattern in $X(t_X)$, the distribution of $\rho$ is proportional to
-% \[
-%     \P\{ \rho=z \mid X_{I-\ell}^{m+2\ell}(t_X)=x\} \propto \pi(z) p_{m+2\ell}(t,z,x) .
-% \]
-The probability that we see $x$ and $y$ at the random location $I$ is
-\begin{align} \label{eqn:phylo_likelihood}
-    \P\{X_{I-\ell}^{m+2\ell}(t_X)=x \text{ and } Y_I^\ell(t_Y)=y \} = \sum_{z \in \calS^{m+2\ell}} \pi(z) p_{m+2\ell}(t_X,z,x) p_{m,\ell}(t_Y,z,y) .
-\end{align}
-This can be computed without much more effort than the simpler case above,
-given the frequencies at the root, as described in greater generality below.
-
-%%%%
-\subsection{The root distribution}
-
-We clearly don't want to add an extra parameter $\pi(z)$ for all $z \in \calS^{m+2\ell}$.
-If nothing else, it would make the inference depend on $m$ and $\ell$, which we'd like to avoid.
-One approach would be to put a Dirichlet distribution on $\pi(z)$.
-This has the nice property that you can choose priors that are consistent across different values of $m$ and $\ell$.
-We'd have to still MCMC around in the space of $\pi(z)$, so they'd be nuisance parameters from that point of view.
-Also, making the frequencies at the root random introduces correlations between counts, which we want to avoid.
-An alternative is to specify $\pi$ through lower-dimensional statistics,
-for instance to model the root state as a sample from a Markov random field
-with parameters for the marginal frequencies of each state, and pairwise (or higher) correlations between them.
-
-A Markov random field with interactions between sites no more than distance two away
-is determined by the energy function $\phi(a,b,c)$ which gives the ``energy'' of the configuration $abc$ for each $a,b,c \in \calS$.
-We use this to define the distribution on root patterns $\pi$ by
-\begin{align}  \label{eqn:root_mrf}
-    \pi(z) = \frac{1}{Z} \exp\left( - \sum_{i=1}^{|z|} \phi(z_{i-1},z_i,z_{i+1}) \right),
-\end{align}
-i.e.\ the exponential of the total energy of each three adjacent sites,
-where $Z$ is the normalizing constant.
-Note that it is often difficult to compute $Z$ in applications of Markov random fields,
-but here we are working with relatively short sequences,
-so we just compute \eqref{eqn:root_mrf} for all possible $z$, and divide by the sum.
-(However, for this same reason this is not equivalent to saying that $X(0)$ is a sample from the Markov random field
-and finding the distribution of patterns within this.)
-
-%%%%%
-\subsubsection{Mixing}
-
-There can be significant confounding of the frequencies at the root with the other parameters,
-which results in poor mixing.
-(See CpG results, below.)
-To see why this is, consider the following back-of-the-envelope calculation.
-Take $x \neq y$ to be single bases (so, set $\ell=0$ and $w=1$),
-so that $N(x,y)$ gives the number of times an $x$ and a $y$ are homologous.
-Let $q(x,y)$ be the instantaneous rate that $x \to y$ (averaged somehow over context?),
-so that $N(x,y)$ should be, roughly,
-\begin{align*}
-  \sum_z \pi(z) q(z,x) q(z,y) &\approx t \pi(x) (1+q(x,x)) q(x,y) + t \pi(y) q(y,x) (1+q(y,y)) ,
-\end{align*}
-where as usual $q(x,x) = (-1) \sum_{z \neq x} q(x,z)$ is (minus) the total jump rate out of $x$.
-
-So, when we propose a change to $\pi$ we might want to also adjust the rates accordingly to keep this more-or-less constant:
-if we chnage $\pi(x) \mapsto \pi(x) + \epsilon_x$, we want $\delta_{xy}$ to satisfy
-\begin{align*}
-  \sum_z \pi(z) q(z,x) q(z,y) = \sum_z (\pi(z) + \epsilon_z) ( q(z,x) + \delta_{zx} ) ( q(z,y) + \delta_{zy} ) ,
-\end{align*}
-or, to first order in $\delta$,
-\begin{align*}
-  0 &\approx \sum_z \epsilon_z q(z,x) q(z,y) + \sum_z \pi(z) \delta_{zx} q(z,y) + \sum_z \pi(z) q(z,x) \delta_{zy} \\
-   &\approx \sum_z \epsilon_z q(z,x) q(z,y) + \sum_z (\pi(z)+\epsilon_z) \delta_{zx} q(z,y) + \sum_z (\pi(z)+\epsilon_z) q(z,x) \delta_{zy} \\
-   &\approx \epsilon_x (1+q(x,x)) q(x,y) + \epsilon_y q(y,x) (1+q(y,y)) + (\pi(y)+\epsilon_y) \delta_{yx} (1+q(y,y)) + (\pi(x)+\epsilon_x) (1+q(x,x)) \delta_{xy} \\
-   &= \left( \epsilon_x q(x,y) + (\pi(x)+\epsilon_x) \delta_{xy} \right)(1+q(x,x)) + \left( \epsilon_y q(y,x) + (\pi(y)+\epsilon_y) \delta_{yx} \right)(1+q(y,y))
-\end{align*}
-One solution to this is to let $\delta_{xy} = - \epsilon_x q(x,y) / \pi(x)$.
-
-Suppose each change is a single-base change, so that changes $\delta \mu_i$ in $\mu$ contribute to $q$ by
-$\delta_{xy} = \sum_{i: x \to y} \delta \mu_i$.
-Given $\delta_{xy}$, this gives a linear equation for $\delta \mu_i$ that may or may not be solvable,
-but certainly is if we have all the single-base transitions as paramters.
-
-%%%%%%%%%%%
-\section{Computation}
-
-For inference, we need to compute the $|\calS|^{m+2\ell} \times |\calS|^m$ matrix $F$ whose $(x,y)^\text{th}$ entry is $p_{m,\ell}(t,x,y)$,
-for various $t$.
-This matrix is a projection of the $|\calS|^{m+2\ell} \times |\calS|^{m+2\ell}$ matrix whose $(x,z)^\text{th}$ entry is
-\begin{align}
-    p_{m+2\ell}(x,z) = \left( e^{t G(m+2\ell)} \right)_{x,z} ,
-\end{align}
-where $G(m+2\ell)$ is the sparse $|\calS|^{m+2\ell} \times |\calS|^{m+2\ell}$ matrix defined in \eqref{eqn:G_defn}.
-The ``projection'' we need just marginalizes over long patterns $z$ that match the shorter pattern $y$:
-if we define the matrix $U$ in this way, so that $U_{zy}=1$ if $z_\ell^m=y$ and $U_{zy}=0$ otherwise,
-then
-\begin{align} \label{eqn:Tmer_trans}
-    p_{m,\ell}(t,x,y) = \sum_{z \in \calS^{m+2\ell}} \left( e^{t G(m+2\ell)} \right)_{xz} U_{zy} ,
-\end{align}
-so in fact we don't need the entire matrix $e^{t G(m+2\ell)}$,
-just the product of this matrix with each of the columns of $U$.
-
-Modern techniques in sparse matrix computation (Krylov methods) provide efficient ways to do this
-by only multiplying $G$ and $U$ some number of times,
-which is much more efficient because $G$ is sparse.
-
-\subsection{Updating $G$}
-
-We can also take advantage of the sparsity of $G$ to perform efficient computation of the likelihood
-under many sets of parameters.
-First note that $G(m+2\ell)$ has, assuming only single-position changes,
-$(1+m(|\calS|-1)) |\calS|^{m+2\ell}$ nonzero entries, since each of the $|\calS|^{m+2\ell}$ can change in $m$ places.
-This will determine how the computation scales with $m$, $\ell$, and $|\calS|$,
-once we precompute a number of things.
-Let $g = (g_1, \ldots, g_d)$ be the nonzero entries of $G$, in some fixed ordering;
-sparse matrix representations of $G$ store only $g$ along with information about the rows and columns these are found in.
-Each $g_i$ is a linear combination of mutation rates $\mu_j$, say $g_i = \sum_j a_{ij} \mu_j$;
-so by precomputing $a_{ij}$ we can efficiently update $G$.
-This is true if the boundary conditions are circular, mean-value, or neither.
-
-We can do similar precomputations to help with the case of selective differences:
-XXX describe
-
-\subsection{Phylogenetic peeling}
-
-Furthermore, note that the further calculations in the phylogenetic case can be done using these methods.
-To compute
-\[
-    \sum_z \pi(z) p_{m+2\ell}(t_X,z,x) p_{m,\ell}(t_Y,z,y)
-    = \sum_z \left( e^{t_X G(m+2\ell)} \right)_{zx} \pi(z) \sum_w \left( e^{t_X G(m+2\ell)} \right)_{zw} U_{wy} ,
-\]
-we can first compute $\sum_w \left( e^{t_X G(m+2\ell)} \right)_{zw} U_{wy}$ as above,
-multiply by $\pi(z)$, and then apply the algorithm again using the transpose of $G(m+2\ell)$.
-
-Using the R package expm, this makes computation quite feasible:
-with four possible states and $\ell=2$ and $m=1$, so that $G(m+2\ell)$ is a $1024 \times 1024$ matrix,
-computing $e^{t G(5) }$ takes 13 seconds, while computing $e^{t G(5)} U$ takes only 0.3 seconds.
-Increasing to $\ell=4$ and $m=1$ or $\ell=3$ and $m=2$ is still feasible, taking $e^{t G(8)} U$ in 42 and 47 seconds, respectively
-(and much longer for the whole matrix $e^{t G(8)}$).
-
-More generally\dots
-
-\begin{center}
-\includegraphics{peeling-schematic}
-\end{center}
-
-To compute the likelihood on a more complicated tree, consider the following,
-with the three-taxon tree above as an example,
-where we are counting occurrences of (long) $m+2\ell$-tuples in taxon $X$,
-and (shorter) $m$-tuples in taxa $Y$ and $Z$.
-
-We can compute the likelihood in a similar way, summing over the root state (the summation over $u$)
-and the interior node ($v$):
-\begin{align} \label{eqn:three_taxa_likelihood}
-  \begin{split}
-    & \P\{X_{I-\ell}^{m+2\ell}=x \text{ and } Y_I^\ell=y \text{ and } Z_I^\ell=z \} \\
-    &\qquad = \sum_{u \in \calS^{m+2\ell}} \pi(u) p_{m,\ell}(t_C,u,z) \sum_{v \in \calS^{m+2\ell}} p_{m+2\ell}(t_{AB},u,v) p_{m+2\ell}(t_{A},v,x) p_{m,\ell}(t_B,v,y)
-  \end{split}
-\end{align}
-The steps for computing this are then:
-\begin{enumerate}
-
-  \item Compute $M_1(u,z) = p_{m,\ell}(t_C,u,z)$ (a $|\calS|^{m+2\ell} \times |\calS|^m$ matrix)
-
-  \item Compute $M_2(v,z) = \sum_{u \in \calS^{m+2\ell}}  p_{m+2\ell}(t_{AB},u,v) \pi(u) M_1(u,z)$ (still a $|\calS|^{m+2\ell} \times |\calS|^m$ matrix)
-
-  \item Compute $M_3(v,y) = p_{m,\ell}(t_B,v,y)$ (a $|\calS|^{m+2\ell} \times |\calS|^m$ matrix)
-
-  \item Compute $M_4(x,(y,z)) = \sum_{v \in \calS^{m+2\ell}} p_{m+2\ell}(t_{A},v,x) M_2(v,z) M_3(v,y)$ (stored as a $|\calS|^{m+2\ell} \times |\calS|^{2m}$ matrix).
-
-\end{enumerate}
-Then $M_4$ gives all the likelihoods.
-This is depicted in the figure, writing $\left(e^{tG}\right)_{ab}$ for the matrix $p_{m+2\ell}(t,a,b)$,
-and $P$ for the projection matrix that gives $p_{m,\ell}(t,a,b) = \left( e^{tG} P\right)_{ab}$.
-
-In total there are $|\calS|^{3m+2\ell}$ possible data combinations.
-However, we can reduce dimensionality by replacing the last step with
-\begin{enumerate}
-
-  \item[4'.] Compute $M_4(x,w) = \sum_{v \in \calS^{m+2\ell}} p_{m+2\ell}(t_{A},v,x) M_3(v,w) M_4(v,w)$ (stored as a $|\calS|^{m+2\ell} \times |\calS|^{m}$ matrix).
-
-\end{enumerate}
-and only computing the likelihood for cases with $y=z=w$.
-On a three-taxon tree it is not clear that this is desireable, since missing out on double transitions
-might be a significant loss of information,
-but on larger trees it seems likely that we'd want to restrict to combinations with most sites agreeing across the tree.
-(Note, however, that one should not first look at the data, observe which combinations $(y,z)$ were most common, and then restrict to only those!)
-
-\subsection*{Algorithm}
+\section*{Peeling algorithm}
 
@@
 
-%%%%%%
-\section{Bayesian things}
-
-So, suppose we've got a list of patterns $(x_i,y_i)$
-and we're happy with the idea that they're independent.
-Recall that
+%%%%
+\section{Other likelihood functions}
+
+\subsection{Summary statistics}
+
+Since we cannot compute reasonably the full likelihood for the data,
+we will need to choose appropriate summary statistics
+that contain as much information about the process as possible,
+and have a tractable expression for their marginal likelihood.
+For simplicity,
+we assume in this section that approximation \eqref{eqn:window_approx} holds exactly.
+If $i_1 < \ldots < i_{n(x)}$ are a set of nonoverlapping sites where the initial sequence matches $x$
+-- i.e.\ $X_{i_k}^{(2\ell+m)}(0)=x$ and $i_{k+1} > i_k + 2\ell+m$ for all $k$ %En dash close to math could be confusing. This sentence is a doozy.
+then the counts $N(x,y)=\#\{ i_k \st X_{i_k+\ell}^{(m)}(t)=y\}$ (the number of these sites that match the smaller pattern $y$ at time $t$)
+are multinomial, with probabilities $\{p_{\ell,m}(x,y)\st y \in \calS^m\}$.
+For instance, if $\ell=3$ then $N(\text{ACTCAC}, \text{CGC})$ counts the number of occurrences of
+\begin{center}
+\begin{tabular}{c|ccccccc}
+ $x$ &  C  & A & C & T & C & A & C \\
+ $y$ &  $\centerdot$  & $\centerdot$  & C & G & C & $\centerdot$  & $\centerdot$
+\end{tabular}
+\end{center}
+where $\cdot$ represents any base.
+More generally, if we partition the sequence into windows of length $2\ell+m$,
+and let $N(x,y)$ be the number of times we see a window in which $x$ matches $X(0)$ and $y$ matches $X(t)$,
+then these counts are independently multinomial across each $x$.
+% so the marginal likelihood of the statistics $N(x,y)$ is
+% \begin{align}
+%     \prod_{x \in \calS^{2\ell+m}} N(x)! \prod_{y \in \calS^m} \frac{p_{\ell,m}(x,y)^{N(x,y)}}{N(x,y)!} ,
+% \end{align}
+% where $N(x) = \sum_y N(x,y)$.
+
+Of course, a drawback to this approach is that we throw away a good fraction of the data.
+In fact, we do not even look at a fraction $2\ell/(2\ell+m)$ of the second sequence, $X(t)$,
+that lies in the outside edges of each window.
+As before, let $N(x)$ be the number of occurrences of a pattern $x$ in $X(0)$,
+and $N(x,y)$ the number of times $x$ is seen juxtaposed with $y$ in $X(t)$,
+but this time, count over all positions,
+so that e.g.\ $N(x) = \#\{ 1 \le i \le n - 2\ell - m + 1 \st X_i^{(m+2\ell)}(0)=x\}$.
+
+Now we will now calculate the expectation and covariances of $N(x,y)$.
+\peter{The following is Not Right, but also we don't really use it at the moment.}
+If we assume that \eqref{eqn:window_approx} is exact, then
 \[
-\E\left[ N(x,y) \mid N(x) \right] = N(x) p_{m,\ell}(x,y) ,
+    \E [N(x,y)] = N(x) \, p_{\ell,m}(x,y).
 \]
-Parameters $\theta$ enter implicitly through $p_{m,\ell}$.
-
-In the case that we observe the initial sequence, $N(x,y)$ are binomial with sample size $N(x)$.
-and so the likelihood is
-\[
-\prod_x N(x)! \prod_y \frac{ p_{m,\ell}(x,y)^{N(x,y)} }{ N(x,y)! } ,
-\]
-and so the negative log likelihood is, up to a constant,
-\[
-\sum_{x,y} N(x,y) \log p_{m,\ell}(x,y) .
-\]
-
-%%%%%%
-\section{Full likelihood}
-
-\emph{The following is copied over from work on the abphylo project.}
-
-The s5f model is defined as follows:
-for each $(a,b,c,d,e,x) \subset \{\nA,\nC,\nG,\nT\}$,
-it gives a (relative) mutation rate
-\begin{align}
-    \gamma(c,x;a,b,d,e) = \text{( mutation rate of $c \to x$ in context $abcde$ )} .
-\end{align}
-Given these rates, the only unknown parameter is the total amount of time
-that the sequence has been mutated for.
-Suppose that $X(t)$ is a sequence of length $n$ evolving under this model,
-and let $G_n$ be the corresponding generator matrix,
-a $4^n \times 4^n$ matrix.
-(In the s5f model, the mutation rates are only specified in the middle of 5-mers,
-so the two bases on each end never mutate.)
-The transition probabilities are $e^{tG_n}$,
-and by marginalization,
-we can compute quantities like
-the probability for observing the middle set of basis
-given the flanking $\ell$ on each end:
-\begin{align}
-    p_{\ell,n-2\ell}(t;a,b)
-    &=
-    \P\{
-        (X_{\ell+1}(t), X_{\ell+2}(t), \ldots, X_{n-\ell}(t))=(b_1,b_2,\ldots,b_{n-2\ell}) \\
-    &\qquad \qquad
-        \given (X_{1}(0), X_{2}(0), \ldots, X_{n}(0))=(a_1,a_2,\ldots,a_{n})
-    \} .
-\end{align}
-If the size of the context, $\ell$, is big enough,
-then this probability will not depend significantly if the sequence $a$ is in fact embedded in a longer sequence
-(the error decreases exponentially with $\ell$).
-
-The sequence $X(t)$ can be constructed by beginning with $X(0)$,
-putting down a Poisson process of potential mutations of rate $\gamma = \max\{\gamma(c,x;a,b,d,e)\}$,
-and thinning appropriately.
-The process depends on the flanking 2bp,
-so any other mutation within 2bp of a site affects the process at that site,
-and the chance that this occurs depends on the flanking 4bp on either side.
-Concretely, the number of flanking positions in each direction
-that must be kept track of to decide what happens at a given site
-grows by 1 at rate $\gamma$, and by 2 at rate $\gamma$;
-so the amount of flanking sequence on each side affecting a given site at time $t$
-is the sum of a Poisson($\gamma t$) and twice an independent Poisson($\gamma t$).
-This has mean $3\gamma t$ and variance $5 \gamma t$;
-with high probability it will not exceed, say,
-$3\gamma t + 6 \sqrt{\gamma t}$;
-if $\gamma t < 1$ this is less than 9.
-
-How can we use this to compute the likelihood of a full sequence?
-Actually, we know more.
-For two positions $(i,s)$ and $(j,t)$ in sequence $\times$ time,
-let $r(i,s;j,t) = |j-i|/|t-s|$ be the slope of the line between them,
-and if $I$ and $J$ are subsets of sites, let $r(I,s;J,t) = \min\{r(i,s;j,t) : i\in I, j \in J\}$
-be the minimum.
-Then, we know that there is a speed of dependency propagation $r_*$
-such that if $r(I,s;J,t) > r_*$,
-then $X_I(s)$ and $X_J(t)$ are nearly independent.
-Now, fix $t$ and for each site $i$ let $N_i$ denote its $r_*$-neighborhood,
-i.e.,
-\begin{align}
-    N_i = \{ (j,s) : r(i,t; j,s) \le r_*, \; 0 \le s \le t \} .
-\end{align}
-These are depicted in figure~\ref{fig:neighborhoods}A.
-We will write $i \sim j$ if $|i-j| \le 2 r_* t$,
-or equivalently, if $N_i \cap N_j \neq \emptyset$,
-and $i \nsim j$ otherwise.
-We can think of $N_i$ as the parts of ancestral sequences in which a mutation might have affected the state at $(i,t)$.
-so if $i \nsim j$, then $i$ and $j$ are (approximately) independent at time $t$, given the sequence at time 0.
-Concretely,
-we have a small $\epsilon$ such that for any $i \nsim j$,
-\begin{align}
-    &\big|
-    \log( \P\{ X_i(t) = x_i \and X_j(t) = x_j \} ) \\
-    &\qquad {} -
-    \log( \P\{ X_i(t) = x_i \} \P\{ X_j(t) = x_j \} )
-    ) \big|
-    \le \epsilon .
-\end{align}
-
-\begin{figure}
-    \begin{center}
-        \includegraphics{context-neighborhoods-defns}
-
-        \vspace{2em}
-
-        \includegraphics{context-neighborhoods-widths}
-    \end{center}
-    \caption{
-        Diagram of things in the text:
-        (top) Dependency neighboorhods.
-        (bottom) What we need to compute $\P\{X_i(t)=x \given X(0), X_{<i}(t)\}$:
-        site $i$ depends on the blue squares in $N_i$;
-        the sites to the left of this in $L(i)$ are those other sites in $X(t)$
-        whose dependency neighborhoods overlap $N_i$;
-        any further sites to the left of this are basically independent of $X_i(t)$;
-        the red squares give the dependency neighborhood of $L(i)$,
-        so we need to condition on the row of red and blue sites at the top.
-        On the bottom, we have $2w+1$ sites; at the top we have $4w+1$;
-        here $w=2$.
-        \label{fig:neighborhoods}
-    }
-\end{figure}
-
-What we'd like to know is the likelihood, i.e.,
-\begin{align}
-    \like(y|x)
-    &:=
-    \P\{ X(t) = y \given X(0) = x \} .
-\end{align}
-If we define $X_{<i}=(X_1, \ldots, X_{i-1})$, then
-\begin{align}
-    \like(y|x)
-    &=
-    \prod_{i=1}^n
-    \P\{ X_i(t) = y_i \given X(0) = x \and X_{<i}(t) = y_{<i} \} .
-\end{align}
-Now, defining $L(i) = \{ j < i \st j \sim i \}$
-to be the sites to the left of $i$ whose dependency neighborhoods overlap with $i$'s,
-and $A(I) = \{ j \st |j-i| \le r_*t \text{ for some } i \in I\}$
-to be the ancestral sites in the dependency neighborhood of $I$,
-\begin{align}
-&\big|
-\log( \P\{ X_i(t) = y_i \given X(0) = x \and X_{<i}(t) = y_{<i} \} ) \\
-& \qquad {} -
-\log( \P\{ X_i(t) = y_i \given X_{A(L(i))}(0) = x_{A(L(i))} \and X_{L(i)}(t) = y_{L(i)} \} )
- \big| \\
-&\qquad \qquad \le
-    \epsilon ,
-\end{align}
-so that if we define
-\begin{align}
-    \alike(y|x)
-    &=
-    \prod_{i=1}^n
-    \P\{ X_i(t) = y_i \given X_{A(L(i))}(0) = x_{A(L(i))} \and X_{L(i)}(t) = y_{L(i)} \} ,
-\end{align}
-then (with $\loglike = \log \like$),
-\begin{align}
-    \left| \aloglike(y|x) - \loglike(y|x) \right|
-    \le n \epsilon .
-\end{align}
-
-This is good, because we can compute $\aloglike$:
-let $w=r_*t$ be radius of a dependency neighborhood,
-so that the width of $L(i)$ is $2w$,
-and for a sequence $x$ define $x^{(m)}_i = (x_{i-m+1}, \ldots, x_i)$:
-\begin{align}
-    \alike(y|x)
-    &=
-    \prod_{i=1}^n
-    \frac{
-        p_{w,2w+1}(y^{(2w+1)}_i|x^{(4w+1)}_{i+w})
-    }{
-        p_{w,2w}(y^{(2w)}_{i-1}|x^{(4w)}_{i-1+w})
-    }
-\end{align}
-This is the ratio of the likelihoods of all the $(w,2w+1)$ T-mers,
-divided by the likelihoods of all the $(w,2w)$ T-mers.
-
-%%%%%%%
-\section{Simulation results}
-
-The method works well to find point estimates of the parameters.
-Obtaining posterior density estimates using the approximation likelihood function on a subset of patterns
-works in some cases;
-in other cases the likelihood surface is still too peaked, underestimating the uncertainty.
-
-\subsection{TASEP}
-
-Strings of length $10^5$ were evolved under TASEP for varying amounts of time,
-and pattern occurrences counted.
-Running 10,000 MCMC iterations with window sizes of $\ell=w=3$ (total length 9) on these takes around 40 minutes.
-Apparently, the patterns used were not sufficiently independent in this case,
-as the posterior is too sharp.
-
-Here's a short example sequence with $\lambda=1$:
-\begin{center}
- \setlength{\tabcolsep}{0pt}
-\begin{tabular}{cccccccccccccccccccccccccccccccccccccccccccccccccccccccccccc}
-X&X&X&O&O&X&O&O&O&O&X&O&X&O&X&X&O&O&X&O&O&X&O&X&X&X&O&X&O&O&O&X&X&O&X&O&O&O&X&X&O&O&O&X&O&O&O&O&X&O&X&X&O&X&X&O&O&X&O&O \\
-$\centerdot$&O&X&O&X&O&O&O&X&$\centerdot$&O&X&O&O&X&X&X&o&x&$\centerdot$&o&O&X&x&x&O&X&O&X&o&o&O&X&X&$\centerdot$&$\centerdot$&o&o&O&X&X&o&$\centerdot$&O&X&$\centerdot$&o&o&$\centerdot$&$\centerdot$&O&X&X&x&O&X&o&O&X&$\centerdot$
+To compute covariances, define $\chi_i(x,y)$ to be 1 if $X_{i-\ell}^{(|x|)}(0)=x$ and $X_i^{(|y|)}(t)=y$, and 0 otherwise,
+so that $N(x,y) = \sum_{i=1}^n \chi_i(x,y)$.
+Overloading notation, also define $\chi_i(x)$ to be 1 if $X_{i-\ell}^{(|x|)}(0)=x$,
+\begin{align}
+  \var[N(x,y)] &= \sum_{i=1}^n \var[\chi_i(x,y)] + \sum_{i\neq j} \cov[\chi_i(x,y), \chi_j(x,y)] \\
+               &\approx \sum_{i=1}^n \chi_i(x) p_{\ell,m}(x,y)(1-p_{\ell,m}(x,y))
+            + 2 \sum_{i=1}^{n-2\ell-m} \sum_{j=i+1}^{i+2\ell+m-1} \cov[ \chi_i(x,y) \chi_j(x,y) ] \\
+            &= N(x) p_{\ell,m}(x,y)(1-p_{\ell,m}(x,y)) +
+\end{align}
+The term $\E[\chi_1(x,y) \chi_{k+1}(x,y)]$ is the probability that $x$ matches at both site 1 and site $k+1$,
+and that $y$ matches both at site $\ell+1$ and at site $\ell+k+1$.
+For this to be nonzero, the joint pattern $(x,y)$ must allow self-overlaps.
+Similarly, if $(x_1,y_1) \neq (x_2,y_2)$ then
+\begin{align}
+  \begin{split}
+    \cov[N(x_1,y_1),N(x_2,y_2)]
+    & \approx - n (2m+4\ell) p_{\ell,m}(x_1,y_1)p_{\ell,m}(x_2,y_2) \\
+    & \qquad {} + n\sum_{k=1}^{m+2\ell} \E[\chi_1(x_1,y_1) \chi_{k+1}(x_2,y_2) + \chi_1(x_2,y_2) \chi_{k+1}(x_1,y_1)] .
+  \end{split}
+\end{align}
+
+%%%%
+\subsection{A likelihood function}
+
+The covariance calculations show that
+if $p_{\ell,m}(x,y)$ is small enough (i.e.\ the pattern is not common),
+then $N(x,y)$ is close to Poisson.
+Using this observation, we can estimate parameters by composite likelihood,
+using all $N(x,y)$ as data.
+Specifically, this composite likelihood is
+\begin{align} \label{eqn:comp_like}
+  \prod_{x,y} F_{(x,y)}^{N(x,y)}
+\end{align}
+where $F$ is the $|\calS|^{m+2\ell} \times |\calS|^m$ matrix whose $(x,y)^\text{th}$ entry is $p_{m,\ell}(t,x,y)$ as described in the "Computation" section.
+Maximum likelihood then gets an asymptotically unbiased estimate of the parameters. (cite)
+However, since they are not independent, to get confidence intervals, we need to either simulate, or do something different.
+
+If we use only a subset of patterns $\{(x_k,y_k)\}$ chosen so that $\E[M_1(x,y) M_{k+1}(x,y)]$ is small (or zero),
+\erick{M not defined, but I'm guessing that this is effectively saying that the probability of having two different patterns apply to the same sequences is small?}
+the entire collection $\{N(x_k,y_k)\}$ will be close to independent Poisson.
+(using Stein's method)
+Then we'll have the correct likelihood function for (this subset of) data.
+
+If changes are rare, then one way to get a set of patterns with this property is to only choose ones such that going from $x$ to $y$ entails a change;
+and any overlapping patterns must require additional changes,
+i.e.\ if $(x_1,y_1)$ matches at $i_1$,
+then for $(x_2,y_2)$ to match at $i_2$,
+there must be additional changes outside $(i_1,\ldots,i_1+\ell-1)$.
+
+Here is one way to obtain such a set.
+Take a pair of patterns $(x,y)$, that overlap at $m$ positions,
+and suppose that they differ at $1 \le i_1, \ldots, i_d \le m$.
+Let $\bar i = (1/d) \sum_{j=1}^d i_j$ be the mean position of the changes,
+and take only patterns with $(m-1)/2 < \bar i \le (m+1)/2$.
+For example, in the example above with $m=3$ and $\ell=2$,
+\begin{center}
+\begin{tabular}{c|ccccccc}
+ $x$ &  C  & A & C & T & C & A & C \\
+ $y$ &  $\centerdot$  & $\centerdot$  & C & G & C & $\centerdot$  & $\centerdot$
 \end{tabular}
 \end{center}
-here particles are X's, moving to the right; and spaces are O's;
-in the second line,
-$\centerdot$ is a position that has not considered a change; lowercase letters are positions that considered a change but did not;
-and upper case are positions that did change.
-
-\begin{figure}
-  \begin{center}
-    \includegraphics[width=\textwidth]{writeup-plots/tasep_all-mcmc-runs}
-  \end{center}
-  \caption{ Results from MCMC estimates for six different data sets of the parameter in TASEP (mutation rate multiplied by time).
-  Vertical line shows the true value.
-  }
-\end{figure}
-
-\subsection{Ising model}
-
-Strings of length $10^6$ were evolved under Glauber dynamics on the Ising model with varying parameters,
-and pattern occurrences counted.
-As for TASEP, running 10,000 MCMC iterations with window sizes of $\ell=w=3$ (total length 9) on these takes around 40 minutes.
-In this case, the true parameter values have good distribution within the estimated posterior distributions.
-
-Here is an example, with all parameters equal to 1, for $t=.5$:
-\begin{center}
- \setlength{\tabcolsep}{0pt}
-\begin{tabular}{cccccccccccccccccccccccccccccccccccccccccccccccccccccccccccc}
-X&X&O&X&X&O&O&X&O&X&O&X&X&O&O&X&O&O&O&O&X&X&X&O&X&O&X&O&X&O&O&O&X&X&O&X&X&X&X&X&O&X&O&X&O&X&X&X&X&O&O&X&X&X&X&X&X&O&O&O \\
-x&x&o&x&X&X&O&x&o&x&$\centerdot$&x&x&O&X&X&$\centerdot$&O&X&O&X&x&x&o&x&o&x&o&x&O&X&O&$\centerdot$&$\centerdot$&O&O&X&X&O&X&X&X&o&x&o&$\centerdot$&X&O&X&o&o&x&X&X&X&O&X&o&$\centerdot$&$\centerdot$
+$x$ and $y$ differ only at (relative) position $i_1=2$.
+
+Another way to achieve the same goal of having all overlapping patterns require additional changes is to require $i_1 = \bar i = 1$.
+No two such patterns can be shifted versions of each other
+without adding more changed sites,
+since shifting by $k$ without adding changed sites just adds $k$ to $\bar i$.
+For example, we might take
+\begin{center}
+\begin{tabular}{c|ccccccc}
+ $x$ &  A & C & T & C & A & C & G \\
+ $y$ &  $\centerdot$  & $\centerdot$  & G & C & A & $\centerdot$  & $\centerdot$
 \end{tabular}
 \end{center}
-The notation is as above; except that the entire matched pattern involved in a change is uppercase:
-to simulate, we look in windows of size 3; so if XOO changes to XXO, the whole thing is uppercase, not just the middle X.
-
-\begin{figure}
-  \begin{center}
-    \includegraphics{writeup-plots/selsims-2013-05-28-17-12-0275615-traces}
-  \end{center}
-  \caption{
-  MCMC traces of the parameters in the Ising model across 100,000 iterations.
-  The units of $\lambda$ are in mean number of changes per site.
-  Vertical line shows the true value.
-  }
-\end{figure}
-
-\begin{figure}
-  \begin{center}
-    \includegraphics[width=\textwidth,page=1]{writeup-plots/ising_all-mcmc-runs}
-  \end{center}
-  \caption{ Results from MCMC estimates of the time-scaled mutation rate parameter $\lambda t$,
-  in ten different simulated datasets under the Ising model.
-  Vertical line shows the true value.
-  }
-\end{figure}
-
-\begin{figure}
-  \begin{center}
-    \includegraphics[width=\textwidth,page=2]{writeup-plots/ising_all-mcmc-runs}
-  \end{center}
-  \caption{ Results from MCMC estimates of the inverse temperature $\beta$ in the Ising model.
-  Vertical line shows the true value.
-  }
-\end{figure}
-
-\begin{figure}
-  \begin{center}
-    \includegraphics[width=\textwidth,page=3]{writeup-plots/ising_all-mcmc-runs}
-  \end{center}
-  \caption{ Results from MCMC estimates of the magnetization parameter $\beta$ in the Ising model.
-  Vertical line shows the true value.
-  }
-\end{figure}
-
-%%%
-\subsection{CpG mutation, on a tree}
-
-Nucleotide sequences of length $10^6$ were simulated down two (unequal) branches of a tree,
-with all single-base substitution rates $m_{xy}$ set equal (varying from a total of .01 to .1 across the tree)
-and an additional CpG rate $\gamma$ set to either 0 or three times the single-base rate.
-Analysis was done with $\ell=2$ and $w=1$ (a five-to-one window).
-
-The state at the root of the tree was independently chosen bases with equal frequencies,
-and these frequencies at the root were added as parameters to the model (with a Dirichlet prior).
-
-Here is a short example, with all parameters equal to 1, with each branch of the tree having length $0.1$:
-\begin{center}
- \setlength{\tabcolsep}{0pt}
-\begin{tabular}{cccccccccccccccccccccccccccccccccccccccccccccccccccccccccccc}
-T&G&G&A&G&A&G&G&T&C&T&G&C&C&A&T&A&G&A&C&G&G&G&C&G&C&C&G&C&C&T&A&G&G&A&G&C&C&G&C&G&T&T&T&T&G&C&G&T&G&G&T&G&T&T&G&T&G&G&C \\
-$\centerdot$&$\centerdot$&g&C&G&a&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&T&A&C&A&$\centerdot$&$\centerdot$&$\centerdot$&T&A&T&$\centerdot$&$\centerdot$&$\centerdot$&T&G&c&C&C&A&C&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&G&G&G&t&A&T&$\centerdot$&$\centerdot$&G&A&g&C&G&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&t&T&G&A \\
-$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&A&T&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&T&T&G&T&$\centerdot$&$\centerdot$&G&A&T&$\centerdot$&$\centerdot$&$\centerdot$&G&G&c&t&A&A&G&C&$\centerdot$&$\centerdot$&$\centerdot$&G&T&G&A&t&t&t&g&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&G&A&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&A&C
-\end{tabular}
-\end{center}
-Top is ancestral sequence (``unobserved''); bottom two are at the tips.
-
-The point estimates are good,
-as are the MCMC estimates,
-but there seems to be mixing trouble having to do with the root frequencies.
-
-\begin{figure}
-  \begin{center}
-    \includegraphics{writeup-plots/selsims-2013-06-03-13-17-0790276-initfreqs}
-  \end{center}
-  \caption{
-  MCMC traces across 660,000 iterations of the frequencies at the root of the (two-taxon) tree in the CpG model.
-  Vertical line shows the true value.
-  }
-\end{figure}
-
-\begin{figure}
-  \begin{center}
-    \includegraphics{writeup-plots/selsims-2013-06-03-13-17-0790276-mutrates}
-  \end{center}
-  \caption{
-  MCMC traces across 660,000 iterations of several mutation rate parameters in the CpG model.
-  Units are in mean number of substitutions per base across the entire tree.
-  Vertical line shows the true value.
-  }
-\end{figure}
+
+\section{Removed bits that we probably don't want}
 
 %%%%%%%%
-\section{Power sims}
-
-Assess power and accuracy in the Ising model, by running replicates at these parameters:
-
-\begin{tabular}{|rlllrrr|}
-  \hline
-  seqlen   &  $\lambda t$   &  $\beta$  &  $\gamma$  &  $w$ &  $\ell$   & \# reps \\
-  \hline
-  $10^4$    &   .2  &   1   &   .5  &   3   &   3   &   10  \\
-  $10^3$    &   .2  &   1   &   .5  &   3   &   3   &   10  \\
-  $10^2$    &   .2  &   1   &   .5  &   3   &   3   &   10  \\
-  \hline
-  $10^4$    &   .4  &   1   &   .5  &   3   &   3   &   10  \\
-  $10^4$    &   .1  &   1   &   .5  &   3   &   3   &   10  \\
-  $10^4$    &   .05  &   1   &   .5  &   3   &   3   &   10  \\
-  \hline
-  $10^4$    &   .2  &   1   &   0  &   3   &   3   &   10  \\
-  \hline
-  $10^4$    &   .2  &   1   &   .5  &   3   &   2   &   10  \\
-  $10^4$    &   .2  &   1   &   .5  &   3   &   1   &   10  \\
-  $10^4$    &   .2  &   1   &   .5  &   3   &   0   &   10  \\
-  \hline
-\end{tabular}
-
-\section{Applications}
-
-graham:
-\begin{itemize}
-
-\item estimating freq spectrum [Hernandez]
-
-\item probabilistic ancestral states for common ancestor of e.g. humans 4N generations ago [Siepel] -- would widely used for primates
-
-\end{itemize}
-
-sergey:
-\begin{itemize}
-
-  \item ancestral CpG
-
-  \item ancient DNA
-
-\end{itemize}
-
-%%%%%%%%
-\section{What next}
-
-\begin{enumerate}
-
-  \item Allow more than single-base distribution at the root.
-
-  \item What's up with the mixing problem in root frequencies?
-
-  \item See if counting in nonoverlapping windows fixes the peakiness problem with TASEP. \emph{Yes, it does.}
-
-  \item See what's feasible with a larger alphabet -- 60 codons -- and more parameters.
-
-  \item Show that taking no context ($\ell=0$) can be positively misleading in some context.
-
-  \item Do a three-taxa tree.
-
-  \item How to identify motifs in the residuals?
-
-\end{enumerate}
-
-\bibliography{context-dependence}
+\subsection{Edge effects and cyclicization}
+
+The choice to treat long sequences as cyclical is generally of little importance.
+However, we have defined in \eqref{eqn:G_defn} the basic building block of our approximation, $G(m)$,
+to be the transition matrix \emph{for a circular sequence} of length $m$, where $m$ may be short.
+\peter{no we haven't}
+This seems strange -- but, what else would we do at the edges?
+We want to use $G(\ell+m+r)$ to compute the chance that a given $(\ell+m+r)$-sequence ends up with another given $m$-sequence in the middle;
+and if $\ell$ and $r$ are big enough relative to $t$, this will be a good approximation regardless of what we do at the edges.
+One answer would be to not allow patterns hanging off of either end to match,
+resulting in a slight underestimate of the chance of change.
+A better approximation would be to use the marginal frequencies in some way to average over surroudning sequence
+-- but, this requires marginal frequencies to be known, and these are not stationary.
+However, the leftmost site in a randomly chosen window is a sample from the marginal distribution,
+and if sites separated by distance $\ell+m+r$ are sufficiently uncorrelated,
+wrapping the sequence around is a good approximation to averaging over marginal frequencies in the surrounding sequence.
+It is easy to imagine situations where this would not be the best solution --
+but again, if $\ell$ and $r$ are large enough, it won't matter much.
+
+%%%%
+\subsection{The root distribution}
+
+We clearly don't want to add an extra parameter $\pi(z)$ for all $z \in \calS^{m+2\ell}$.
+If nothing else, it would make the inference depend on $m$ and $\ell$, which we'd like to avoid.
+One approach would be to put a Dirichlet distribution on $\pi(z)$.
+This has the nice property that you can choose priors that are consistent across different values of $m$ and $\ell$.
+We'd have to still MCMC around in the space of $\pi(z)$, so they'd be nuisance parameters from that point of view.
+Also, making the frequencies at the root random introduces correlations between counts, which we want to avoid.
+An alternative is to specify $\pi$ through lower-dimensional statistics,
+for instance to model the root state as a sample from a Markov random field
+with parameters for the marginal frequencies of each state, and pairwise (or higher) correlations between them.
+
+A Markov random field with interactions between sites no more than distance two away
+is determined by the energy function $\phi(a,b,c)$ which gives the ``energy'' of the configuration $abc$ for each $a,b,c \in \calS$.
+We use this to define the distribution on root patterns $\pi$ by
+\begin{align}  \label{eqn:root_mrf}
+    \pi(z) = \frac{1}{Z} \exp\left( - \sum_{i=1}^{|z|} \phi(z_{i-1},z_i,z_{i+1}) \right),
+\end{align}
+i.e.\ the exponential of the total energy of each three adjacent sites,
+where $Z$ is the normalizing constant.
+Note that it is often difficult to compute $Z$ in applications of Markov random fields,
+but here we are working with relatively short sequences,
+so we just compute \eqref{eqn:root_mrf} for all possible $z$, and divide by the sum.
+(However, for this same reason this is not equivalent to saying that $X(0)$ is a sample from the Markov random field
+and finding the distribution of patterns within this.)
+
+%%%%%
+\subsubsection{Mixing}
+
+There can be significant confounding of the frequencies at the root with the other parameters,
+which results in poor mixing.
+(See CpG results, below.)
+To see why this is, consider the following back-of-the-envelope calculation.
+Take $x \neq y$ to be single bases (so, set $\ell=0$ and $w=1$),
+so that $N(x,y)$ gives the number of times an $x$ and a $y$ are homologous.
+Let $q(x,y)$ be the instantaneous rate that $x \to y$ (averaged somehow over context?),
+so that $N(x,y)$ should be, roughly,
+\begin{align*}
+  \sum_z \pi(z) q(z,x) q(z,y) &\approx t \pi(x) (1+q(x,x)) q(x,y) + t \pi(y) q(y,x) (1+q(y,y)) ,
+\end{align*}
+where as usual $q(x,x) = (-1) \sum_{z \neq x} q(x,z)$ is (minus) the total jump rate out of $x$.
+
+So, when we propose a change to $\pi$ we might want to also adjust the rates accordingly to keep this more-or-less constant:
+if we chnage $\pi(x) \mapsto \pi(x) + \epsilon_x$, we want $\delta_{xy}$ to satisfy
+\begin{align*}
+  \sum_z \pi(z) q(z,x) q(z,y) = \sum_z (\pi(z) + \epsilon_z) ( q(z,x) + \delta_{zx} ) ( q(z,y) + \delta_{zy} ) ,
+\end{align*}
+or, to first order in $\delta$,
+\begin{align*}
+  0 &\approx \sum_z \epsilon_z q(z,x) q(z,y) + \sum_z \pi(z) \delta_{zx} q(z,y) + \sum_z \pi(z) q(z,x) \delta_{zy} \\
+   &\approx \sum_z \epsilon_z q(z,x) q(z,y) + \sum_z (\pi(z)+\epsilon_z) \delta_{zx} q(z,y) + \sum_z (\pi(z)+\epsilon_z) q(z,x) \delta_{zy} \\
+   &\approx \epsilon_x (1+q(x,x)) q(x,y) + \epsilon_y q(y,x) (1+q(y,y)) + (\pi(y)+\epsilon_y) \delta_{yx} (1+q(y,y)) + (\pi(x)+\epsilon_x) (1+q(x,x)) \delta_{xy} \\
+   &= \left( \epsilon_x q(x,y) + (\pi(x)+\epsilon_x) \delta_{xy} \right)(1+q(x,x)) + \left( \epsilon_y q(y,x) + (\pi(y)+\epsilon_y) \delta_{yx} \right)(1+q(y,y))
+\end{align*}
+One solution to this is to let $\delta_{xy} = - \epsilon_x q(x,y) / \pi(x)$.
+
+Suppose each change is a single-base change, so that changes $\delta \mu_i$ in $\mu$ contribute to $q$ by
+$\delta_{xy} = \sum_{i: x \to y} \delta \mu_i$.
+Given $\delta_{xy}$, this gives a linear equation for $\delta \mu_i$ that may or may not be solvable,
+but certainly is if we have all the single-base transitions as paramters.
 

### 2017-04-26 scripts/fit-model.R
switched fit-model to full likelihood function

@@ -68,7 +68,11 @@ check.genmatrix(init.config,genmatrix)
 counts <- read.counts(opt$infile, bases=genmatrix@bases, longpats=rownames(genmatrix) )
 stopifnot( all( rownames(counts) == rownames(genmatrix) ) )
 
+# shorter counts for full likelihood
+counts.0 <- projectcounts(counts, new.shortwin=shortwin(counts) - 1)
+
 projmatrix <- collapsepatmatrix( ipatterns=rownames(genmatrix), leftwin=leftwin(counts), fpatterns=colnames(counts) )
+projmatrix.0 <- collapsepatmatrix( ipatterns=rownames(genmatrix), leftwin=leftwin(counts), fpatterns=colnames(counts.0) )
 
 if (FALSE) {  ## DO NOT DO THIS; move elsewhere
     if (is.null(init.config$mutrates)) {
@@ -117,8 +121,11 @@ likfun <- function (sub.params){
     genmatrix@x <- do.call( update_x, c( list( G=genmatrix,mutrates=params[1:nmuts(genmatrix)],selcoef=params[seq(1+nmuts(genmatrix),length.out=nsel(genmatrix))]), as.list(fparams) ) )
     # this is collapsed transition matrix
     subtransmatrix <- computetransmatrix( genmatrix, projmatrix, tlen=1, time="fixed") # shape=params[length(params)], time="gamma" )
+    subtransmatrix.0 <- computetransmatrix( genmatrix, projmatrix.0, tlen=1, time="fixed") # shape=params[length(params)], time="gamma" )
     # return POSITIVE log-likelihood
-    ans <- sum( counts@counts * log(subtransmatrix) )
+    num <- sum( counts@counts * log(subtransmatrix) )
+    den <- sum( counts.0@counts * log(subtransmatrix.0) )
+    ans <- num - den
     if (!is.finite(ans)) { print(paste("Warning: non-finite likelihood with params:",paste(params,collapse=", "))) }
     return(ans)
 }

### 2017-04-27 writeups/context-dependence.tex
switched fit-model to full likelihood function
some more words

@@
 \usepackage{natbib}
-\usepackage[colorlinks=false]{hyperref}
+\usepackage[hidelinks]{hyperref}
 \usepackage{bm}
@@
 \newcommand{\calT}{\mathcal{T}}  % set of transition triples
+\newcommand{\calP}{\mathcal{P}}  % set of selection pairs
 \newcommand{\nA}{\mbox{A}}  % nucleotides:
@@
 let $x_i^{(h)} = (x_i, x_{i+1}, \ldots, x_{i+h-1})$ denote the pattern of $x$ of length $h$ beginning at location $i$.
-Recall we are focusing on finite sequences, and by default, we do not allow patterns to ``hang off'' the end of the sequence.
-This is natural in many situations, but in others we might want to other boundary conditions,
-modeling extant, but unobserved, neighbors.
-We will mention this when necessary.
-
+% Recall we are focusing on finite sequences, and by default, we do not allow patterns to ``hang off'' the end of the sequence.
+% This is natural in many situations, but in others we might want to other boundary conditions,
+% modeling extant, but unobserved, neighbors.
+% We will mention this when necessary.
 The dynamics of our stochastic process are determined by the set of allowed transitions and associated rates:
@@
 written as ``$u \to v$ at rate $\mu$''.
-The set of all transition triples, $\calT$, determines the time evolution of this process.
-
-\peter{INCLUDE SELECTION}
+The set of all transition triples is denoted $\calT$.
 
@@
 
-\begin{example}[CpG mutation]
+\peter{Should call this ``selection'' instead of a ``potential''? Or what?}
+
+To allow more compact specification of models,
+we also introduce a ``potential'': 
+suppose that there is a collection of patterns $\{p_i\}$
+such that each occurrence of $p_i$ adds a quantity $e_i$ to the total ``energy'' of a sequence,
+and that the rate at which each possible change occurs is modulated by a function of the energy difference the change would produce.
+This is natural if that transition triples describe \emph{proposed} changes,
+and that the probability a proposed change actually occurs depends on how much it affects the energy.
+Concretely, let $\calP = \{(p_i, e_i\}$ be a set of (pattern, energy change) pairs,
+and for any sequence $x$ define
+$E(x) = \sum_i e_i n(x,p_i)$,
+where $n(x,p_i)$ is the number of times the pattern $p_i$ occurs in $x$.
+
+\begin{example}[Genomic GC content]
     A sequence of genome can be written using A, C, G, and T;
@@
     Embellishing the single-nucleotide model with this additional rate results in the model defined by
-
     \begin{center}
@@
         \nC\nG   &  \nC\nA   &  $\gamma$ &
-    \end{tabular} .
-  \end{center}
-  Here $\gamma$ is the additional CpG rate above the base mutation rate,
-  and a total of 13 parameters.
-  Note that this model is very different in spirit to TASEP,
-  in which the total number of particles was conserved.
-
-  In this model, for a given pair of sequences, there may be more than one way to mutate a given string $x$ to get $y$:
+      \end{tabular} ,
+    \end{center}
+    where $\gamma$ is the additional CpG rate above the base mutation rate.
+    Note that for a given pair of sequences, there may be more than one way to mutate a given string $x$ to get $y$:
     for instance, a change $\nA\nC\nG \to \nA\nT\nG$ could have occurred by a single $\nC \to \nT$ mutation,
@@
     and hence the total instantaneous rate at which ACG changes to ATG is $\gamma + m_{\nC\nT}$.
-  Although this means there are many possible parameterizations,
-  note that the model is identifiable, not overparameterized.
+    % Although this means there are many possible parameterizations,
+    % note that the model is identifiable, not overparameterized.
+
+    Counteracting this trend towards increased G/C is GC-biased gene conversion \citep{gcbiased},
+    which is effectively a selective pressure against G or C bases.
+    We model the evolution of a \emph{single} sequence, not a population of sequences,
+    imagining this sequence to be the consensus sequence of the population.
+    A new mutation with an effect $s$ on fitness that occurs
+    in a population of effective size $N$ becomes ubiquitous, rather than dying out,
+    with probability approximately $(1-\exp(-2 s))/(1-\exp(-2 s N))$ \citep{fixation_prob}.
+    If the mutation occurs at rate $\mu$ per individual, the total rate it occurs at in the population is $\mu N$.
+    GC-biased gene conversion effectively means that sequences with more G/Cs are selected against.
+    This is incorporated as follows: set
+    \begin{center}
+        \begin{tabular}{cc}
+        pattern  &  energy \\
+        \nC{} or \nG   &  $s$   
+      \end{tabular} ,
+    \end{center}
+    so that $E(y)-E(x)$ is the net change in number of G's and C's multiplied by $s$.
+    Also, let
+    \begin{align*}
+        \phi(e) = N (1-\exp(-2e))/(1-\exp(-2eN))
+    \end{align*}
+    so that if a given change that occurs with rate $\mu$ would change a sequence $x$ to $y$,
+    then $\mu \phi(E(y)-E(x))$ is the rate at which the mutation appears and successfully takes over in the population.
 
@@
     imagined as magnetic dipoles,
-  and the energy associated with a given state $x$ is $H(x) = - \frac{1}{2} \beta \sum_i x_i x_{i+1} - \frac{1}{2} \gamma \sum_i x_i$.
+    and the energy associated with a given state $x$ is $E(x) = - \frac{1}{2} \beta \sum_i x_i x_{i+1} - \frac{1}{2} \gamma \sum_i x_i$.
     Here $\beta$ represents inverse temperature, and $\gamma$ represents the strength of the magnetic field (here scaled by temperature).
-  The associated stationary distribution on configurations is proportional to $\exp(-H(x))$.
-  The following ``Glauber dynamics'' preserve the stationary distribution:
+    The associated stationary distribution on configurations is proportional to $\exp(-E(x))$.
+    The following dynamics preserve the stationary distribution:
     each site, independently at rate $\lambda$,
     forgets its spin,
-  and reconfigures to a state chosen with probability proportional to the stationary probability of the resulting configuration.
-  Ignoring transitions that don't change the state,
-  if patterns $u$ and $v$ differ at only one site then $u$ changes to $v$ at rate $\lambda/(1+\exp(H(v)-H(u)))$,
-  so
-
-    \peter{reformulate with a potential}
-
+    and reconfigures to a state chosen with probability proportional to the stationary probability of the resulting configuration:
+    Since $\exp(-E(x))/(\exp(-E(x))+\exp(-E(y))) = 1/(1+exp(-(E(x)-E(y)))$,
+    this is:
     \begin{center}
         \begin{tabular}{c@{\quad$\to$\quad}c@{\quad at rate\quad }c}
-      $+++$  &   $+-+$   &  $\lambda/\,(1+e^{2\beta + \gamma})$ \\
-      $++-$  &   $+--$   &  $\lambda/(1+e^{\gamma})$ \\
-      $+-+$  &   $+++$   &  $\lambda/(1+e^{-2\beta - \gamma})$ \\
-      $+--$  &   $++-$   &  $\lambda/(1+e^{-\gamma})$ \\
-      $-++$  &   $--+$   &  $\lambda/(1+e^{\gamma})$ \\
-      $-+-$  &   $---$   &  $\lambda/(1+e^{-2\beta + \gamma})$ \\
-      $--+$  &   $-++$   &  $\lambda/(1+e^{-\gamma})$ \\
-      $---$  &   $-+-$   &  $\lambda/(1+e^{2\beta - \gamma})$
+          $+$  &   $-$   &  $\lambda$ \\
+          $-$  &   $+$   &  $\lambda$ 
         \end{tabular} 
-  \end{center}
-
-  Glauber dynamics have been widely studied... XXX
-
+        \qquad and \qquad
+        \begin{tabular}{cc}
+        pattern  &  energy \\
+        $+-$ or $-+$  &   $\beta$ \\
+        $+$ &   $\gamma$
+        \end{tabular} 
+    \end{center}
+    and 
+    \begin{align*}
+        \phi(e) = \frac{1}{1+\exp(-e)} .
+    \end{align*}
+
+    This process has been studied by XXX.
 \end{example}
@@
 \begin{align} \label{eqn:G_defn}
-  G(n)_{x,y} = \sum_{i=1}^n \sum_{J(i,x,y)}  \mu^j ,
+    G(n)_{x,y} = \sum_{i=1}^n \phi(E(y)-E(x)) \sum_{J(i,x,y)}  \mu^j ,
 \end{align}
 and if there are no triples $(\mu,u,v)$ with $x \xrightarrow{i,u,v} y$ for some $i$, then $G(n)_{x,y}=0$.
-This formalizes the statement that the transition rate from $x$ to $y$ is the sum of the mutation rates of all transition triples
+This formalizes the statement that the transition rate from $x$ to $y$ is the sum of the rates of all transition triples
 that would change $x$ into $y$.
@@
 \]
-Note that since mutation is homogeneous, the maximum rate of changes occurring at any position will be the same, except possibly near the end of the sequence,
+Note that since the process is homogeneous, the maximum rate of changes occurring at any position will be the same, except possibly near the end of the sequence,
 so the maximization over $i$ is unnecessary.
@@
 
-The problem at hand is to infer the parameters of the model, given the state at time 0 and later at time $t$.
-If $n$ is large enough and $t$ is not too large,
-this should be feasible -- there should be enough information in our observations to do this reliably.
-If $t$ is too large, the process may ``saturate'' --
-if there are many changes at most sites, then in at least most models,
-we will lose most information about the dynamics,
-retaining information only about the parameters that affect the stationary distribution.
-In TASEP with any boundary conditions, we lose all information as it approaches stationarity,
-while in the Ising model we have information about the temperature $\beta$ and magnetic field $\gamma$ but not the speed of changes, $\lambda$.
-In the CpG model it is not immediately clear what information is retained.
-
-But, how can we extract the information?
+The problem at hand is to infer the parameters of the model based on the final state of the model after it hsa evolved from some known initial state.
+% If the sequence is long enough and not too much time has gone by
+% this should be feasible -- there should be enough information in our observations to do this reliably.
+% If the time is too large, the process may ``saturate'' --
+% if there are many changes at most sites, then in at least most models,
+% we will lose most information about the dynamics,
+% retaining information only about the parameters that affect the stationary distribution.
+% In TASEP with any boundary conditions, we lose all information as it approaches stationarity,
+% while in the Ising model we have information about the temperature $\beta$ and magnetic field $\gamma$ but not the speed of changes, $\lambda$.
+% In the CpG model it is not immediately clear what information is retained.
+How can we extract the information?
 These are Markov processes, on the state space $\calS^n$,
@@
 as it ignores dependencies between neighboring blocks.
-Intuitively, we want to increase the size of the context
--- and the likelihood thus obtained is asymptotically correct for $m$ large,
-but computationally, we are stuck with small values of $m$.
-
-Alternatively, we could take into account context in the initial sequence only,
-by including the states of an additional $\ell$ sites on both the left and right.
-The probability that an initial sequence $x$ of length $m+2\ell$ after time $t$
-is found to match a smaller, middle sequence $y$ of length $m$ is found by marginalizing:
-\begin{align}
-  p_{\ell,m}(t;x,y) &= \sum_{a \in \calS^\ell} \sum_{b \in \calS^\ell} p_{2\ell+m}(t;x,a \join y \join b) , \quad \text{for}\; x \in \calS^{2\ell+m}, \; y \in \calS^m ,
+% Intuitively, we want to increase the size of the context
+% -- and the likelihood thus obtained is asymptotically correct for $m$ large,
+% but computationally, we are stuck with small values of $m$.
+
+\begin{figure}
+    \begin{center}
+        \includegraphics{Tmer-dependency}
+    \end{center}
+    \caption{
+        \textbf{(A)} A $(6,2,2)$ T-mer.
+        Nonmutated sites are shown in lower case letters.
+        \textbf{(B)}
+        The three mutations have extended the influence of the circled \nA{} 
+        in the initial sequence (top)
+        to the six positions colored red in the final sequence.
+        Red dotted arrows show the propagation of dependency 
+        between the circled \nA{}
+        and the third site in the final sequence (bottom).
+        The maximum window width, $w$, is equal to 1.
+        \label{fig:Tmers}
+    }
+\end{figure}
+
+The key insight we use is that by including neighboring sequence context in the \emph{initial} sequence only,
+we can compute \emph{something} correctly,
+and that something will be sufficient to compute the full likelihood.
+To do this, let $p_{n,\ell,r}(x,y)$ for $r+\ell < n$ denote
+the probability that an initial sequence $x$ of length $n$, after time $t$,
+is found to match a smaller sequence $y$ of length $n-\ell-r$, offset by $\ell$ sites.
+We refer to such patterns as ``T-mers'', as depicted in Figure \ref{fig:Tmers}.
+This quantity is found by marginalizing $p_m$:
+\begin{align}
+    p_{n,\ell,r}(t;x,y) &= \sum_{a \in \calS^\ell} \sum_{b \in \calS^r} p_n(t;x,a \join y \join b) , \quad \text{for}\; x \in \calS^n, \; y \in \calS^{n-\ell-r} ,
 \end{align}
 where $a \join y \join b$ is the sequence composed of concatenating $a$, $y$, and $b$ together in that order.
-This is the probability that if we begin with a sequence $x$ of length $2\ell+m$,
-then after time $t$ we will see the pattern $y$ at the middle $m$ positions,
-ignoring what is present in the leftmost and rightmost $\ell$ positions.
-We expect $p_{\ell,m}(x,y)$ to be asymptotically correct when $\ell$ is large (and $t$ is fixed),
+We expect $p_{n,\ell,m}(x,y)$ to be asymptotically correct when $\ell$ is large (and $t$ is fixed),
 i.e.\ that the probability a given subsequence matching $x$ is later observed to have its center matching $y$
@@
 \begin{align} \label{eqn:window_approx}
-  p_{\ell,m}(x,y) \approx \P\{ X_{i+\ell}^{(m)}(t) = y \mid X_i^{(2\ell+m)}(0) = x \}.
-\end{align}
-We provide a precise statement of the error in this approximation later.
-
-\peter{how to choose window size}
+    p_{n,\ell,r}(x,y) \approx \P\{ X_{i+\ell}^{(n-\ell-r)}(t) = y \mid X_i^{(n)}(0) = x \}.
+\end{align}
+
+%%%%%
+\subsection{Choice of overhang}
+
+The intuition behind this fact is straightforward, and provides a way to pick the overhang lengths $\ell$ and $r$;
+a precise statement of the approximation is given in Section \ref{ss:approx_pf}.
+Suppose that each instantaneous change depends on positions at most $w$ sites away.
+Then, the mutation process at any particular site
+depends on the initial sequence at that site and within distance $w$ on either side, 
+and the outcomes of any mutations that have occurred within that window.
+
+Then, the sequence at a particular site may directly affect the process at the neighboring $w$ sites on either side.
+If another change happens, say, at the end of this window,
+then the indirect effects of this change may extend out to $2w-1$ sites.
+In this way, the influence of the sequence at each site extends outwards, as shown in Figure \ref{fig:Tmers} --
+so if we take the overhangs long enough that this dependency is unlikely to extend to the base of the Tmer,
+the approximation will be good.
+As we show in Section \ref{ss:approx_pf}, if the overhangs are $\ell = r = kw$,
+then the error in approximation \eqref{eqn:window_approx}
+is bounded by the probability that there are more than $k$ changes in a given window of length $w$.
+
+For instance, in the CpG model the maximum dependency is $w=1$ as the process only depends on neighboring sites;
+and if the sequence has evolved for time $t$ then we expect no more than a Poisson($\lambda t$) number of changes at each site.
+If $\lambda t = 0.75$ (three-quarters of the sequence will have changed!),
+then taking $\ell = r = 3$ allows us to compute Tmer probabilities to within an error of 0.007,
+since this is the probability that a Poisson with mean 0.75 will be larger than 3.
+If $\lambda t = 0.25$ we would do even better with $\ell = r = 2$.
 
@@
 
-\peter{shorten/edit this to follow on from the above}
-
-The s5f model is defined as follows:
-for each $(a,b,c,d,e,x) \subset \{\nA,\nC,\nG,\nT\}$,
-it gives a (relative) mutation rate
-\begin{align}
-    \gamma(c,x;a,b,d,e) = \text{( mutation rate of $c \to x$ in context $abcde$ )} .
-\end{align}
-Given these rates, the only unknown parameter is the total amount of time
-that the sequence has been mutated for.
-Suppose that $X(t)$ is a sequence of length $n$ evolving under this model,
-and let $G_n$ be the corresponding generator matrix,
-a $4^n \times 4^n$ matrix.
-(In the s5f model, the mutation rates are only specified in the middle of 5-mers,
-so the two bases on each end never mutate.)
-The transition probabilities are $e^{tG_n}$,
-and by marginalization,
-we can compute quantities like
-the probability for observing the middle set of basis
-given the flanking $\ell$ on each end:
-\begin{align}
-    p_{\ell,n-2\ell}(t;a,b)
+What we actually want is the \emph{full likelihood} $\like(y|x)$,
+the probability that a given sequence $x$ evolves into another, $y$, over a given period of time.
+This can be computed as the product of the probability of each site in $y$ given $x$ and previous sites:
+defining $x_{<i}=(x_1, \ldots, x_{i-1})$, this is:
+\begin{align*}
+    \like(y|x) &= \prod_i \P\{ Y_i = y_i \given Y_{<i} = y_{<i} \;\&\; X=x \} .
+\end{align*}
+We can compute these term by discarding sites outside the window of influence:
+\begin{align*}
+    \P\{ Y_i = y_i \given Y_{<i} = y_{<i} \;\&\; X=x \} 
     &=
-    \P\{
-        (X_{\ell+1}(t), X_{\ell+2}(t), \ldots, X_{n-\ell}(t))=(b_1,b_2,\ldots,b_{n-2\ell}) \\
-    &\qquad \qquad
-        \given (X_{1}(0), X_{2}(0), \ldots, X_{n}(0))=(a_1,a_2,\ldots,a_{n})
-    \} .
-\end{align}
-If the size of the context, $\ell$, is big enough,
-then this probability will not depend significantly if the sequence $a$ is in fact embedded in a longer sequence
-(the error decreases exponentially with $\ell$).
-
-The sequence $X(t)$ can be constructed by beginning with $X(0)$,
-putting down a Poisson process of potential mutations of rate $\gamma = \max\{\gamma(c,x;a,b,d,e)\}$,
-and thinning appropriately.
-The process depends on the flanking 2bp,
-so any other mutation within 2bp of a site affects the process at that site,
-and the chance that this occurs depends on the flanking 4bp on either side.
-Concretely, the number of flanking positions in each direction
-that must be kept track of to decide what happens at a given site
-grows by 1 at rate $\gamma$, and by 2 at rate $\gamma$;
-so the amount of flanking sequence on each side affecting a given site at time $t$
-is the sum of a Poisson($\gamma t$) and twice an independent Poisson($\gamma t$).
-This has mean $3\gamma t$ and variance $5 \gamma t$;
-with high probability it will not exceed, say,
-$3\gamma t + 6 \sqrt{\gamma t}$;
-if $\gamma t < 1$ this is less than 9.
-
-How can we use this to compute the likelihood of a full sequence?
-Actually, we know more.
-For two positions $(i,s)$ and $(j,t)$ in sequence $\times$ time,
-let $r(i,s;j,t) = |j-i|/|t-s|$ be the slope of the line between them,
-and if $I$ and $J$ are subsets of sites, let $r(I,s;J,t) = \min\{r(i,s;j,t) : i\in I, j \in J\}$
-be the minimum.
-Then, we know that there is a speed of dependency propagation $r_*$
-such that if $r(I,s;J,t) > r_*$,
-then $X_I(s)$ and $X_J(t)$ are nearly independent.
-Now, fix $t$ and for each site $i$ let $N_i$ denote its $r_*$-neighborhood,
-i.e.,
-\begin{align}
-    N_i = \{ (j,s) : r(i,t; j,s) \le r_*, \; 0 \le s \le t \} .
-\end{align}
-These are depicted in figure~\ref{fig:neighborhoods}A.
-We will write $i \sim j$ if $|i-j| \le 2 r_* t$,
-or equivalently, if $N_i \cap N_j \neq \emptyset$,
-and $i \nsim j$ otherwise.
-We can think of $N_i$ as the parts of ancestral sequences in which a mutation might have affected the state at $(i,t)$.
-so if $i \nsim j$, then $i$ and $j$ are (approximately) independent at time $t$, given the sequence at time 0.
-Concretely,
-we have a small $\epsilon$ such that for any $i \nsim j$,
-\begin{align}
-    &\big|
-    \log( \P\{ X_i(t) = x_i \and X_j(t) = x_j \} ) \\
-    &\qquad {} -
-    \log( \P\{ X_i(t) = x_i \} \P\{ X_j(t) = x_j \} )
-    ) \big|
-    \le \epsilon .
-\end{align}
-
-\begin{figure}
-    \begin{center}
-        \includegraphics{context-neighborhoods-defns}
-
-        \vspace{2em}
-
-        \includegraphics{context-neighborhoods-widths}
-    \end{center}
-    \caption{
-        Diagram of things in the text:
-        (top) Dependency neighboorhods.
-        (bottom) What we need to compute $\P\{X_i(t)=x \given X(0), X_{<i}(t)\}$:
-        site $i$ depends on the blue squares in $N_i$;
-        the sites to the left of this in $L(i)$ are those other sites in $X(t)$
-        whose dependency neighborhoods overlap $N_i$;
-        any further sites to the left of this are basically independent of $X_i(t)$;
-        the red squares give the dependency neighborhood of $L(i)$,
-        so we need to condition on the row of red and blue sites at the top.
-        On the bottom, we have $2w+1$ sites; at the top we have $4w+1$;
-        here $w=2$.
-        \label{fig:neighborhoods}
-    }
-\end{figure}
-
-What we'd like to know is the likelihood, i.e.,
-\begin{align}
-    \like(y|x)
-    &:=
-    \P\{ X(t) = y \given X(0) = x \} .
-\end{align}
-If we define $X_{<i}=(X_1, \ldots, X_{i-1})$, then
-\begin{align}
-    \like(y|x)
-    &=
-    \prod_{i=1}^n
-    \P\{ X_i(t) = y_i \given X(0) = x \and X_{<i}(t) = y_{<i} \} .
-\end{align}
-Now, defining $L(i) = \{ j < i \st j \sim i \}$
-to be the sites to the left of $i$ whose dependency neighborhoods overlap with $i$'s,
-and $A(I) = \{ j \st |j-i| \le r_*t \text{ for some } i \in I\}$
-to be the ancestral sites in the dependency neighborhood of $I$,
-\begin{align}
-&\big|
-\log( \P\{ X_i(t) = y_i \given X(0) = x \and X_{<i}(t) = y_{<i} \} ) \\
-& \qquad {} -
-\log( \P\{ X_i(t) = y_i \given X_{A(L(i))}(0) = x_{A(L(i))} \and X_{L(i)}(t) = y_{L(i)} \} )
- \big| \\
-&\qquad \qquad \le
-    \epsilon ,
-\end{align}
-so that if we define
-\begin{align}
-    \alike(y|x)
-    &=
-    \prod_{i=1}^n
-    \P\{ X_i(t) = y_i \given X_{A(L(i))}(0) = x_{A(L(i))} \and X_{L(i)}(t) = y_{L(i)} \} ,
-\end{align}
-then (with $\loglike = \log \like$),
-\begin{align}
-    \left| \aloglike(y|x) - \loglike(y|x) \right|
-    \le n \epsilon .
-\end{align}
-
-This is good, because we can compute $\aloglike$:
-let $w=r_*t$ be radius of a dependency neighborhood,
-so that the width of $L(i)$ is $2w$,
-and for a sequence $x$ define $x^{(m)}_i = (x_{i-m+1}, \ldots, x_i)$:
-\begin{align}
-    \alike(y|x)
-    &=
-    \prod_{i=1}^n
         \frac{
-        p_{w,2w+1}(y^{(2w+1)}_i|x^{(4w+1)}_{i+w})
+            \P\{ Y_i = y_i \;\&\; Y_{< i} = y_{<i} \;\&\; X=x \} 
         }{
-        p_{w,2w}(y^{(2w)}_{i-1}|x^{(4w)}_{i-1+w})
-    }
-\end{align}
-This is the ratio of the likelihoods of all the $(w,2w+1)$ T-mers,
-divided by the likelihoods of all the $(w,2w)$ T-mers.
+            \P\{ Y_{<i} = y_{<i} \;\&\; X=x \} 
+        } \\
+    &\approx
+        \frac{
+            p_{n,\ell,\ell}(x_{i-n+\ell}^{(n)}, y_{i-n+2\ell}^{(n-2\ell)})
+        }{
+            p_{n,\ell,\ell+1}(x_{i-n+\ell}^{(n)}, y_{i-n+2\ell}^{(n-2\ell-1)})
+        } .
+\end{align*}
+The approximation depends on $\ell$ being large enough, as discussed above, and $n \ge 2\ell+1$ --
+this is because (a) to compute the likelihood in the numerator of the $i^\text{th}$ site, we need the flanking $\ell$ sites (as discussed above),
+and (b) to compute the denominator we need to include all sites to the left affected by those $2\ell+1$ sites.
+
+By rearranging the product, 
+this gives us an expression for the full likelihood as the ratio of the likelihoods of all $(n,\ell,\ell)$ T-mers
+and the likelihood of all $(n,\ell,\ell+1)$ T-mers.
 
@@
 
+%%%%%%%%%%%%%%%%%%%%
 \section{Discussion}
@@
 \section*{Proof of the approximation}
+\label{ss:approx_pf}
 
@@
 Informally, the approximation \eqref{eqn:window_approx} will be good if $H$ does note expand beyond the flanking context.
-Now, MAKE THIS EASIER TO READ
+Now,
 \begin{multline}
@@
 
+\subsection{s5f likelihood}
+
+The s5f model is defined as follows:
+for each $(a,b,c,d,e,x) \subset \{\nA,\nC,\nG,\nT\}$,
+it gives a (relative) mutation rate
+\begin{align}
+    \gamma(c,x;a,b,d,e) = \text{( mutation rate of $c \to x$ in context $abcde$ )} .
+\end{align}
+Given these rates, the only unknown parameter is the total amount of time
+that the sequence has been mutated for.
+Suppose that $X(t)$ is a sequence of length $n$ evolving under this model,
+and let $G_n$ be the corresponding generator matrix,
+a $4^n \times 4^n$ matrix.
+(In the s5f model, the mutation rates are only specified in the middle of 5-mers,
+so the two bases on each end never mutate.)
+The transition probabilities are $e^{tG_n}$,
+and by marginalization,
+we can compute quantities like
+the probability for observing the middle set of basis
+given the flanking $\ell$ on each end:
+\begin{align}
+    p_{\ell,n-2\ell}(t;a,b)
+    &=
+    \P\{
+        (X_{\ell+1}(t), X_{\ell+2}(t), \ldots, X_{n-\ell}(t))=(b_1,b_2,\ldots,b_{n-2\ell}) \\
+    &\qquad \qquad
+        \given (X_{1}(0), X_{2}(0), \ldots, X_{n}(0))=(a_1,a_2,\ldots,a_{n})
+    \} .
+\end{align}
+If the size of the context, $\ell$, is big enough,
+then this probability will not depend significantly if the sequence $a$ is in fact embedded in a longer sequence
+(the error decreases exponentially with $\ell$).
+
+The sequence $X(t)$ can be constructed by beginning with $X(0)$,
+putting down a Poisson process of potential mutations of rate $\gamma = \max\{\gamma(c,x;a,b,d,e)\}$,
+and thinning appropriately.
+The process depends on the flanking 2bp,
+so any other mutation within 2bp of a site affects the process at that site,
+and the chance that this occurs depends on the flanking 4bp on either side.
+Concretely, the number of flanking positions in each direction
+that must be kept track of to decide what happens at a given site
+grows by 1 at rate $\gamma$, and by 2 at rate $\gamma$;
+so the amount of flanking sequence on each side affecting a given site at time $t$
+is the sum of a Poisson($\gamma t$) and twice an independent Poisson($\gamma t$).
+This has mean $3\gamma t$ and variance $5 \gamma t$;
+with high probability it will not exceed, say,
+$3\gamma t + 6 \sqrt{\gamma t}$;
+if $\gamma t < 1$ this is less than 9.
+
+How can we use this to compute the likelihood of a full sequence?
+Actually, we know more.
+For two positions $(i,s)$ and $(j,t)$ in sequence $\times$ time,
+let $r(i,s;j,t) = |j-i|/|t-s|$ be the slope of the line between them,
+and if $I$ and $J$ are subsets of sites, let $r(I,s;J,t) = \min\{r(i,s;j,t) : i\in I, j \in J\}$
+be the minimum.
+Then, we know that there is a speed of dependency propagation $r_*$
+such that if $r(I,s;J,t) > r_*$,
+then $X_I(s)$ and $X_J(t)$ are nearly independent.
+Now, fix $t$ and for each site $i$ let $N_i$ denote its $r_*$-neighborhood,
+i.e.,
+\begin{align}
+    N_i = \{ (j,s) : r(i,t; j,s) \le r_*, \; 0 \le s \le t \} .
+\end{align}
+These are depicted in figure~\ref{fig:neighborhoods}A.
+We will write $i \sim j$ if $|i-j| \le 2 r_* t$,
+or equivalently, if $N_i \cap N_j \neq \emptyset$,
+and $i \nsim j$ otherwise.
+We can think of $N_i$ as the parts of ancestral sequences in which a mutation might have affected the state at $(i,t)$.
+so if $i \nsim j$, then $i$ and $j$ are (approximately) independent at time $t$, given the sequence at time 0.
+Concretely,
+we have a small $\epsilon$ such that for any $i \nsim j$,
+\begin{align}
+    &\big|
+    \log( \P\{ X_i(t) = x_i \and X_j(t) = x_j \} ) \\
+    &\qquad {} -
+    \log( \P\{ X_i(t) = x_i \} \P\{ X_j(t) = x_j \} )
+    ) \big|
+    \le \epsilon .
+\end{align}
+
+\begin{figure}
+    \begin{center}
+        \includegraphics{context-neighborhoods-defns}
+
+        \vspace{2em}
+
+        \includegraphics{context-neighborhoods-widths}
+    \end{center}
+    \caption{
+        Diagram of things in the text:
+        (top) Dependency neighboorhods.
+        (bottom) What we need to compute $\P\{X_i(t)=x \given X(0), X_{<i}(t)\}$:
+        site $i$ depends on the blue squares in $N_i$;
+        the sites to the left of this in $L(i)$ are those other sites in $X(t)$
+        whose dependency neighborhoods overlap $N_i$;
+        any further sites to the left of this are basically independent of $X_i(t)$;
+        the red squares give the dependency neighborhood of $L(i)$,
+        so we need to condition on the row of red and blue sites at the top.
+        On the bottom, we have $2w+1$ sites; at the top we have $4w+1$;
+        here $w=2$.
+        \label{fig:neighborhoods}
+    }
+\end{figure}
+
+What we'd like to know is the likelihood, i.e.,
+\begin{align}
+    \like(y|x)
+    &:=
+    \P\{ X(t) = y \given X(0) = x \} .
+\end{align}
+If we define $X_{<i}=(X_1, \ldots, X_{i-1})$, then
+\begin{align}
+    \like(y|x)
+    &=
+    \prod_{i=1}^n
+    \P\{ X_i(t) = y_i \given X(0) = x \and X_{<i}(t) = y_{<i} \} .
+\end{align}
+Now, defining $L(i) = \{ j < i \st j \sim i \}$
+to be the sites to the left of $i$ whose dependency neighborhoods overlap with $i$'s,
+and $A(I) = \{ j \st |j-i| \le r_*t \text{ for some } i \in I\}$
+to be the ancestral sites in the dependency neighborhood of $I$,
+\begin{align}
+&\big|
+\log( \P\{ X_i(t) = y_i \given X(0) = x \and X_{<i}(t) = y_{<i} \} ) \\
+& \qquad {} -
+\log( \P\{ X_i(t) = y_i \given X_{A(L(i))}(0) = x_{A(L(i))} \and X_{L(i)}(t) = y_{L(i)} \} )
+ \big| \\
+&\qquad \qquad \le
+    \epsilon ,
+\end{align}
+so that if we define
+\begin{align}
+    \alike(y|x)
+    &=
+    \prod_{i=1}^n
+    \P\{ X_i(t) = y_i \given X_{A(L(i))}(0) = x_{A(L(i))} \and X_{L(i)}(t) = y_{L(i)} \} ,
+\end{align}
+then (with $\loglike = \log \like$),
+\begin{align}
+    \left| \aloglike(y|x) - \loglike(y|x) \right|
+    \le n \epsilon .
+\end{align}
+
+This is good, because we can compute $\aloglike$:
+let $w=r_*t$ be radius of a dependency neighborhood,
+so that the width of $L(i)$ is $2w$,
+and for a sequence $x$ define $x^{(m)}_i = (x_{i-m+1}, \ldots, x_i)$:
+\begin{align}
+    \alike(y|x)
+    &=
+    \prod_{i=1}^n
+    \frac{
+        p_{w,2w+1}(y^{(2w+1)}_i|x^{(4w+1)}_{i+w})
+    }{
+        p_{w,2w}(y^{(2w)}_{i-1}|x^{(4w)}_{i-1+w})
+    }
+\end{align}
+This is the ratio of the likelihoods of all the $(w,2w+1)$ T-mers,
+divided by the likelihoods of all the $(w,2w)$ T-mers.
+
 %%%%%%%%
@@
 
+\subsection{Models without using potential}
+
+Here is a reformulation of the CpG and Ising models without using a potential.
+
+\begin{example}[CpG mutation]
+    \peter{version without selection}
+  A sequence of genome can be written using A, C, G, and T;
+  in the most general model of independent mutation across sites, each of the 12 possible transitions occurs at its own rate.
+  Furthermore, it is a well-known observation that in many species,
+  adjacent, methylated CG dinucleotides (``CpG sites'') have a much higher mutation rate to TG and CA
+  than either single nucleotide change under the independent model.
+  Embellishing the single-nucleotide model with this additional rate results in the model defined by
+
+  \begin{center}
+    \begin{tabular}{c@{\quad$\to$\quad}c@{\quad at rate\quad }cc}
+      $x$  &  $y$  &  $m_{xy}$ & when $x \neq y \in \{\nA,\nC,\nG,\nT\}$  \\
+      \nC\nG   &  \nT\nG   &  $\gamma$ & \\
+      \nC\nG   &  \nC\nA   &  $\gamma$ &
+    \end{tabular} .
+  \end{center}
+  Here $\gamma$ is the additional CpG rate above the base mutation rate,
+  and a total of 13 parameters.
+  Note that this model is very different in spirit to TASEP,
+  in which the total number of particles was conserved.
+
+  In this model, for a given pair of sequences, there may be more than one way to mutate a given string $x$ to get $y$:
+  for instance, a change $\nA\nC\nG \to \nA\nT\nG$ could have occurred by a single $\nC \to \nT$ mutation,
+  or by a $\nC\nG \to \nT\nG$ mutation;
+  and hence the total instantaneous rate at which ACG changes to ATG is $\gamma + m_{\nC\nT}$.
+  Although this means there are many possible parameterizations,
+  note that the model is identifiable, not overparameterized.
+
+\end{example}
+
+\begin{example}[1-D Ising model with Glauber dynamics]
+    \peter{version without selection}
+  In the Ising model, each site is labeled as either ``up'' or ``down'' ($+1$ or $-1$ respectively),
+  imagined as magnetic dipoles,
+  and the energy associated with a given state $x$ is $H(x) = - \frac{1}{2} \beta \sum_i x_i x_{i+1} - \frac{1}{2} \gamma \sum_i x_i$.
+  Here $\beta$ represents inverse temperature, and $\gamma$ represents the strength of the magnetic field (here scaled by temperature).
+  The associated stationary distribution on configurations is proportional to $\exp(-H(x))$.
+  The following ``Glauber dynamics'' preserve the stationary distribution:
+  each site, independently at rate $\lambda$,
+  forgets its spin,
+  and reconfigures to a state chosen with probability proportional to the stationary probability of the resulting configuration.
+  Ignoring transitions that don't change the state,
+  if patterns $u$ and $v$ differ at only one site then $u$ changes to $v$ at rate $\lambda/(1+\exp(H(v)-H(u)))$,
+  so
+
+    \peter{reformulate with a potential}
+
+  \begin{center}
+    \begin{tabular}{c@{\quad$\to$\quad}c@{\quad at rate\quad }c}
+      $+++$  &   $+-+$   &  $\lambda/\,(1+e^{2\beta + \gamma})$ \\
+      $++-$  &   $+--$   &  $\lambda/(1+e^{\gamma})$ \\
+      $+-+$  &   $+++$   &  $\lambda/(1+e^{-2\beta - \gamma})$ \\
+      $+--$  &   $++-$   &  $\lambda/(1+e^{-\gamma})$ \\
+      $-++$  &   $--+$   &  $\lambda/(1+e^{\gamma})$ \\
+      $-+-$  &   $---$   &  $\lambda/(1+e^{-2\beta + \gamma})$ \\
+      $--+$  &   $-++$   &  $\lambda/(1+e^{-\gamma})$ \\
+      $---$  &   $-+-$   &  $\lambda/(1+e^{2\beta - \gamma})$
+    \end{tabular}
+  \end{center}
+
+\end{example}
+
 \end{document}

### 2017-04-27 writeups/context-dependence.tex
more words and rearrangement

@@
 
+%%%%%%
+\subsection{Full likelihood}
+
+What we actually want is the \emph{full likelihood} $\like(y|x)$,
+the probability that a given sequence $x$ evolves into another, $y$, over a given period of time.
+This can be computed as the product of the probability of each site in $y$ given $x$ and previous sites:
+defining $x_{<i}=(x_1, \ldots, x_{i-1})$, this is:
+\begin{align*}
+    \like(y|x) &= \prod_i \P\{ Y_i = y_i \given Y_{<i} = y_{<i} \;\&\; X=x \} .
+\end{align*}
+Using the same methods, we can approximate the terms in this sum
+by discarding approximately independent sites from the conditioning.
+If $\ell$ is long enough that approximation \eqref{eqn:window_approx} holds, then:
+\begin{align} \label{eqn:full_approx}
+    \P\{ Y_i = y_i \given Y_{<i} \;\&\; X \} 
+    &\approx
+        \P\{ Y_i = y_i \given Y_{i-2\ell}^{(2\ell+1)} \;\&\; X_{i-3\ell}^{(4\ell+1)} \}  \\
+    &=
+        \frac{
+            p_{2\ell+1,\ell,\ell}(x_{i-3\ell}^{(4\ell+1)}, y_{i-2\ell}^{(2\ell+1)})
+        }{
+            p_{2\ell+1,\ell,\ell+1}(x_{i-3\ell}^{(4\ell+1)}, y_{i-2\ell}^{(2\ell)})
+        } .
+\end{align}
+By rearranging the product, 
+this gives us an expression for the full likelihood as the ratio of the likelihoods of all $(n,\ell,\ell)$ T-mers in $(x,y)$
+to that of of all $(n,\ell,\ell+1)$ T-mers.
+% Define $n_\ell(x,y;a,b)$ to be the number of times pattern $a$ is seen in sequence $x$
+% matched to pattern $b$ in $y$ at a position offset by $\ell$.
+% Then
+% \begin{align}
+%     \log \like(y|x) 
+%     &\approx 
+%         \sum_{a \in \calS^n} \sum_{b \in \calS^{n-2\ell}} n_\ell(x,y;a,b) \log p_{n,\ell,\ell}(a,b)
+%         -
+%         \sum_{c \in \calS^n} \sum_{d \in \calS^{n-2\ell-1}} n_\ell(x,y;c,d) \log p_{n,\ell,\ell+1}(c,d)
+% \end{align}
+
 %%%%%
@@
 
-The intuition behind this fact is straightforward, and provides a way to pick the overhang lengths $\ell$ and $r$;
-a precise statement of the approximation is given in Section \ref{ss:approx_pf}.
+The approximations in \eqref{eqn:window_approx} and \eqref{eqn:full_approx} hold if the overhang, $\ell$, is long enough.
+The intuition is straightforward, and provides a way to pick the overhang lengths $\ell$ and $r$;
+a precise statement is given in Section \ref{ss:approx_pf}.
 Suppose that each instantaneous change depends on positions at most $w$ sites away.
@@
 
-Then, the sequence at a particular site may directly affect the process at the neighboring $w$ sites on either side.
+Therefore, the sequence at a particular site may directly affect the process at the neighboring $w$ sites on either side.
 If another change happens, say, at the end of this window,
@@
 is bounded by the probability that there are more than $k$ changes in a given window of length $w$.
+Concretely, sites further away than $kw$ only matter if there is a chain of at least $k$ intervening mutations
+-- which is unlikely if $k$ is larger than the expected number.
 
@@
 
-%%%%%%
-\section{Full likelihood}
-
-What we actually want is the \emph{full likelihood} $\like(y|x)$,
-the probability that a given sequence $x$ evolves into another, $y$, over a given period of time.
-This can be computed as the product of the probability of each site in $y$ given $x$ and previous sites:
-defining $x_{<i}=(x_1, \ldots, x_{i-1})$, this is:
-\begin{align*}
-    \like(y|x) &= \prod_i \P\{ Y_i = y_i \given Y_{<i} = y_{<i} \;\&\; X=x \} .
-\end{align*}
-We can compute these term by discarding sites outside the window of influence:
-\begin{align*}
-    \P\{ Y_i = y_i \given Y_{<i} = y_{<i} \;\&\; X=x \} 
-    &=
-        \frac{
-            \P\{ Y_i = y_i \;\&\; Y_{< i} = y_{<i} \;\&\; X=x \} 
-        }{
-            \P\{ Y_{<i} = y_{<i} \;\&\; X=x \} 
-        } \\
-    &\approx
-        \frac{
-            p_{n,\ell,\ell}(x_{i-n+\ell}^{(n)}, y_{i-n+2\ell}^{(n-2\ell)})
-        }{
-            p_{n,\ell,\ell+1}(x_{i-n+\ell}^{(n)}, y_{i-n+2\ell}^{(n-2\ell-1)})
-        } .
-\end{align*}
-The approximation depends on $\ell$ being large enough, as discussed above, and $n \ge 2\ell+1$ --
-this is because (a) to compute the likelihood in the numerator of the $i^\text{th}$ site, we need the flanking $\ell$ sites (as discussed above),
-and (b) to compute the denominator we need to include all sites to the left affected by those $2\ell+1$ sites.
-
-By rearranging the product, 
-this gives us an expression for the full likelihood as the ratio of the likelihoods of all $(n,\ell,\ell)$ T-mers
-and the likelihood of all $(n,\ell,\ell+1)$ T-mers.
+To obtain the full likelihood, in \eqref{eqn:full_approx}, in addition to an overhang of $\ell$
+we needed the base of the T-mer to be of length $2\ell+1$.
+This is because we have also conditioned on the final sequence to the left of the focal site, $Y_{<i}$,
+and although the process at $i$ only depends on events within $\pm \ell$,
+these events may affect other sites in $Y$ going back a distance $2\ell$.
 
@@
 just the product of this matrix with each of the columns of $U$.
-
-Modern techniques in sparse matrix computation (Krylov methods) provide efficient ways to do this
-by only multiplying $G$ and $U$ some number of times,
-which is much more efficient because $G$ is sparse.
-
-\subsection{Updating $G$}
-
+Modern techniques in sparse matrix computation (e.g., Krylov methods) provide efficient ways to do this.
+
+Using the R package expm \citep{R_expm}, this makes computation quite feasible:
+with four possible states and $\ell=2$ and $m=1$, so that $G(m+2\ell)$ is a $1024 \times 1024$ matrix,
+computing $e^{t G(5) }$ takes 13 seconds, while computing $e^{t G(5)} U$ takes only 0.3 seconds.
+Increasing to $\ell=4$ and $m=1$ or $\ell=3$ and $m=2$ is still feasible, taking $e^{t G(8)} U$ in 42 and 47 seconds, respectively
+(and much longer for the whole matrix $e^{t G(8)}$).
+
+\paragraph{Sparsity and updating $G$}
 We can also take advantage of the sparsity of $G$ to perform efficient computation of the likelihood
@@
 sparse matrix representations of $G$ store only $g$ along with information about the rows and columns these are found in.
-Each $g_i$ is a linear combination of mutation rates $\mu_j$, say $g_i = \sum_j a_{ij} \mu_j$;
-so by precomputing $a_{ij}$ we can efficiently update $G$.
-This is true if the boundary conditions are circular, mean-value, or neither.
-
-We can do similar precomputations to help with the case of selective differences:
-XXX describe
+Each $g_i$ is a linear combination of mutation rates $\mu_j$,
+multiplied by a function ($\phi$) of a linear combination of energy coefficents $e_j$, 
+say $g_i = \sum_j A_{ij} \mu_j \phi(\sum_k B_{ik} e_k)$,
+so by precomputing the matrices $A$ and $B$ we can update $G$ with new parameter values
+using only two matrix multiplications and evaluation of $\phi()$ over a vector.
+This remains efficient to perform in interpreted languages such as R,
+since each step is carried out by lower-level compiled code (e.g., optimized linear algebra libraries).
+% This is true if the boundary conditions are circular, mean-value, or neither.
 
@@
 we get two (or more) observations evolved from a common root.
-In the simplest case of two taxa we have two processes $X$ and $Y$,
+In the simplest case of two tips we have two processes $X$ and $Y$,
 with identical starting states $X(0)=Y(0)$,
@@
 given the frequencies at the root, as described in greater generality below.
-
-\subsection{Phylogenetic peeling}
-
-Furthermore, note that the further calculations in the phylogenetic case can be done using these methods.
-To compute
+These calculations can be done using the the same sparse matrix methods as above.
+For instance, to compute
 \[
@@
 we can first compute $\sum_w \left( e^{t_X G(m+2\ell)} \right)_{zw} U_{wy}$ as above,
-multiply by $\pi(z)$, and then apply the algorithm again using the transpose of $G(m+2\ell)$.
-
-Using the R package expm, this makes computation quite feasible:
-with four possible states and $\ell=2$ and $m=1$, so that $G(m+2\ell)$ is a $1024 \times 1024$ matrix,
-computing $e^{t G(5) }$ takes 13 seconds, while computing $e^{t G(5)} U$ takes only 0.3 seconds.
-Increasing to $\ell=4$ and $m=1$ or $\ell=3$ and $m=2$ is still feasible, taking $e^{t G(8)} U$ in 42 and 47 seconds, respectively
-(and much longer for the whole matrix $e^{t G(8)}$).
-
-More generally\dots
-
+multiply rows by $\pi(z)$, and then matrix multiply by the transpose of $G(m+2\ell)$.
+
+\subsection{Phylogenetic peeling}
+
+The general case is a modification of Felsenstein's ``peeling'' algorithm.
+
+\begin{figure}
     \begin{center}
@@
     \end{center}
+    \caption{
+        A depiction of the four steps in computing the likelihood of finding particular ``short'' subsequence at $Y$ and $Z$
+        given the ``long'' subsequence at $X$: see text for details.
+        \label{fig:peeling}
+    }
+\end{figure}
 
 To compute the likelihood on a more complicated tree, consider the following,
-with the three-taxon tree above as an example,
+with the three-taxon tree of figure \ref{fig:peeling} as an example,
 where we are counting occurrences of (long) $m+2\ell$-tuples in taxon $X$,
 and (shorter) $m$-tuples in taxa $Y$ and $Z$.
-
-We can compute the likelihood in a similar way, summing over the root state (the summation over $u$)
+The likelihood in this example is found by summing over the root state (the summation over $u$)
 and the interior node ($v$):
@@
 \end{enumerate}
-and only computing the likelihood for cases with $y=z=w$.
+i.e., only computing the likelihood for cases with $y=z=w$.
 On a three-taxon tree it is not clear that this is desireable, since missing out on double transitions
@@
 
-The algorithm in general is written out in Appendix \peter{ref it}.
+The algorithm for a general tree is written out in Appendix \ref{ss:peeling_algorithm}.
 
@@
 %%%%
-\section*{Proof of the approximation}
+\section{Proof of the approximation}
 \label{ss:approx_pf}
@@
 
-In fact, the approximation improves exponentially as $\ell$ and $m$ increase --
+First we quantify and prove the approximation \eqref{eqn:window_approx}.
+In fact, the approximation improves exponentially as $\ell$ and $r$ increase --
 here is a simple argument.
@@
 
+\paragraph{Full likelihood} 
+    Now we use these ideas to quantify and prove approximation \eqref{eqn:full_approx}.
+For two positions $(i,s)$ and $(j,t)$ in sequence $\times$ time,
+let $r(i,s;j,t) = |j-i|/|t-s|$ be the slope of the line between them,
+and if $I$ and $J$ are subsets of sites, let $r(I,s;J,t) = \min\{r(i,s;j,t) : i\in I, j \in J\}$
+be the minimum.
+Above we have shown that there is a speed of dependency propagation $r_*$
+such that if $r(I,s;J,t) > r_*$,
+then $X_I(s)$ and $X_J(t)$ are nearly independent.
+Now, fix $t$ and for each site $i$ let $N_i$ denote its $r_*$-neighborhood,
+i.e.,
+\begin{align}
+    N_i = \{ (j,s) : r(i,t; j,s) \le r_*, \; 0 \le s \le t \} .
+\end{align}
+These are depicted in figure~\ref{fig:neighborhoods}A.
+We will write $i \sim j$ if $|i-j| \le 2 r_* t$,
+or equivalently, if $N_i \cap N_j \neq \emptyset$,
+and $i \nsim j$ otherwise.
+We can think of $N_i$ as the parts of ancestral sequences in which a mutation might have affected the state at $(i,t)$.
+so if $i \nsim j$, then $i$ and $j$ are (approximately) independent at time $t$, given the sequence at time 0.
+Concretely,
+we have a small $\epsilon$ such that for any $i \nsim j$,
+\begin{align}
+    &\big|
+    \log( \P\{ X_i(t) = x_i \and X_j(t) = x_j \} ) \\
+    &\qquad {} -
+    \log( \P\{ X_i(t) = x_i \} \P\{ X_j(t) = x_j \} )
+    ) \big|
+    \le \epsilon .
+\end{align}
+
+\begin{figure}
+    \begin{center}
+        \includegraphics{context-neighborhoods-defns}
+
+        \vspace{2em}
+
+        \includegraphics{context-neighborhoods-widths}
+    \end{center}
+    \caption{
+        Diagram of things in the text:
+        (top) Dependency neighboorhods.
+        (bottom) What we need to compute $\P\{X_i(t)=x \given X(0), X_{<i}(t)\}$:
+        site $i$ depends on the blue squares in $N_i$;
+        the sites to the left of this in $L(i)$ are those other sites in $X(t)$
+        whose dependency neighborhoods overlap $N_i$;
+        any further sites to the left of this are basically independent of $X_i(t)$;
+        the red squares give the dependency neighborhood of $L(i)$,
+        so we need to condition on the row of red and blue sites at the top.
+        On the bottom, we have $2w+1$ sites; at the top we have $4w+1$;
+        here $w=2$.
+        \label{fig:neighborhoods}
+    }
+\end{figure}
+
+What we'd like to know is the likelihood, i.e.,
+\begin{align}
+    \like(y|x)
+    &:=
+    \P\{ X(t) = y \given X(0) = x \} .
+\end{align}
+If we define $X_{<i}=(X_1, \ldots, X_{i-1})$, then
+\begin{align}
+    \like(y|x)
+    &=
+    \prod_{i=1}^n
+    \P\{ X_i(t) = y_i \given X(0) = x \and X_{<i}(t) = y_{<i} \} .
+\end{align}
+Now, defining $L(i) = \{ j < i \st j \sim i \}$
+to be the sites to the left of $i$ whose dependency neighborhoods overlap with $i$'s,
+and $A(I) = \{ j \st |j-i| \le r_*t \text{ for some } i \in I\}$
+to be the ancestral sites in the dependency neighborhood of $I$,
+\begin{align}
+&\big|
+\log( \P\{ X_i(t) = y_i \given X(0) = x \and X_{<i}(t) = y_{<i} \} ) \\
+& \qquad {} -
+\log( \P\{ X_i(t) = y_i \given X_{A(L(i))}(0) = x_{A(L(i))} \and X_{L(i)}(t) = y_{L(i)} \} )
+ \big| \\
+&\qquad \qquad \le
+    \epsilon ,
+\end{align}
+so that if we define
+\begin{align}
+    \alike(y|x)
+    &=
+    \prod_{i=1}^n
+    \P\{ X_i(t) = y_i \given X_{A(L(i))}(0) = x_{A(L(i))} \and X_{L(i)}(t) = y_{L(i)} \} ,
+\end{align}
+then (with $\loglike = \log \like$),
+\begin{align}
+    \left| \aloglike(y|x) - \loglike(y|x) \right|
+    \le n \epsilon .
+\end{align}
+
+This is good, because we can compute $\aloglike$:
+let $w=r_*t$ be radius of a dependency neighborhood,
+so that the width of $L(i)$ is $2w$,
+and for a sequence $x$ define $x^{(m)}_i = (x_{i-m+1}, \ldots, x_i)$:
+\begin{align}
+    \alike(y|x)
+    &=
+    \prod_{i=1}^n
+    \frac{
+        p_{w,2w+1}(y^{(2w+1)}_i|x^{(4w+1)}_{i+w})
+    }{
+        p_{w,2w}(y^{(2w)}_{i-1}|x^{(4w)}_{i-1+w})
+    }
+\end{align}
+This is the ratio of the likelihoods of all the $(w,2w+1)$ T-mers,
+divided by the likelihoods of all the $(w,2w)$ T-mers.
+
 \end{proof}
 
-\section*{Peeling algorithm}
+\section{Peeling algorithm}
+\label{ss:peeling_algorithm}
 
@@
 
-How can we use this to compute the likelihood of a full sequence?
-Actually, we know more.
-For two positions $(i,s)$ and $(j,t)$ in sequence $\times$ time,
-let $r(i,s;j,t) = |j-i|/|t-s|$ be the slope of the line between them,
-and if $I$ and $J$ are subsets of sites, let $r(I,s;J,t) = \min\{r(i,s;j,t) : i\in I, j \in J\}$
-be the minimum.
-Then, we know that there is a speed of dependency propagation $r_*$
-such that if $r(I,s;J,t) > r_*$,
-then $X_I(s)$ and $X_J(t)$ are nearly independent.
-Now, fix $t$ and for each site $i$ let $N_i$ denote its $r_*$-neighborhood,
-i.e.,
-\begin{align}
-    N_i = \{ (j,s) : r(i,t; j,s) \le r_*, \; 0 \le s \le t \} .
-\end{align}
-These are depicted in figure~\ref{fig:neighborhoods}A.
-We will write $i \sim j$ if $|i-j| \le 2 r_* t$,
-or equivalently, if $N_i \cap N_j \neq \emptyset$,
-and $i \nsim j$ otherwise.
-We can think of $N_i$ as the parts of ancestral sequences in which a mutation might have affected the state at $(i,t)$.
-so if $i \nsim j$, then $i$ and $j$ are (approximately) independent at time $t$, given the sequence at time 0.
-Concretely,
-we have a small $\epsilon$ such that for any $i \nsim j$,
-\begin{align}
-    &\big|
-    \log( \P\{ X_i(t) = x_i \and X_j(t) = x_j \} ) \\
-    &\qquad {} -
-    \log( \P\{ X_i(t) = x_i \} \P\{ X_j(t) = x_j \} )
-    ) \big|
-    \le \epsilon .
-\end{align}
-
-\begin{figure}
-    \begin{center}
-        \includegraphics{context-neighborhoods-defns}
-
-        \vspace{2em}
-
-        \includegraphics{context-neighborhoods-widths}
-    \end{center}
-    \caption{
-        Diagram of things in the text:
-        (top) Dependency neighboorhods.
-        (bottom) What we need to compute $\P\{X_i(t)=x \given X(0), X_{<i}(t)\}$:
-        site $i$ depends on the blue squares in $N_i$;
-        the sites to the left of this in $L(i)$ are those other sites in $X(t)$
-        whose dependency neighborhoods overlap $N_i$;
-        any further sites to the left of this are basically independent of $X_i(t)$;
-        the red squares give the dependency neighborhood of $L(i)$,
-        so we need to condition on the row of red and blue sites at the top.
-        On the bottom, we have $2w+1$ sites; at the top we have $4w+1$;
-        here $w=2$.
-        \label{fig:neighborhoods}
-    }
-\end{figure}
-
-What we'd like to know is the likelihood, i.e.,
-\begin{align}
-    \like(y|x)
-    &:=
-    \P\{ X(t) = y \given X(0) = x \} .
-\end{align}
-If we define $X_{<i}=(X_1, \ldots, X_{i-1})$, then
-\begin{align}
-    \like(y|x)
-    &=
-    \prod_{i=1}^n
-    \P\{ X_i(t) = y_i \given X(0) = x \and X_{<i}(t) = y_{<i} \} .
-\end{align}
-Now, defining $L(i) = \{ j < i \st j \sim i \}$
-to be the sites to the left of $i$ whose dependency neighborhoods overlap with $i$'s,
-and $A(I) = \{ j \st |j-i| \le r_*t \text{ for some } i \in I\}$
-to be the ancestral sites in the dependency neighborhood of $I$,
-\begin{align}
-&\big|
-\log( \P\{ X_i(t) = y_i \given X(0) = x \and X_{<i}(t) = y_{<i} \} ) \\
-& \qquad {} -
-\log( \P\{ X_i(t) = y_i \given X_{A(L(i))}(0) = x_{A(L(i))} \and X_{L(i)}(t) = y_{L(i)} \} )
- \big| \\
-&\qquad \qquad \le
-    \epsilon ,
-\end{align}
-so that if we define
-\begin{align}
-    \alike(y|x)
-    &=
-    \prod_{i=1}^n
-    \P\{ X_i(t) = y_i \given X_{A(L(i))}(0) = x_{A(L(i))} \and X_{L(i)}(t) = y_{L(i)} \} ,
-\end{align}
-then (with $\loglike = \log \like$),
-\begin{align}
-    \left| \aloglike(y|x) - \loglike(y|x) \right|
-    \le n \epsilon .
-\end{align}
-
-This is good, because we can compute $\aloglike$:
-let $w=r_*t$ be radius of a dependency neighborhood,
-so that the width of $L(i)$ is $2w$,
-and for a sequence $x$ define $x^{(m)}_i = (x_{i-m+1}, \ldots, x_i)$:
-\begin{align}
-    \alike(y|x)
-    &=
-    \prod_{i=1}^n
-    \frac{
-        p_{w,2w+1}(y^{(2w+1)}_i|x^{(4w+1)}_{i+w})
-    }{
-        p_{w,2w}(y^{(2w)}_{i-1}|x^{(4w)}_{i-1+w})
-    }
-\end{align}
-This is the ratio of the likelihoods of all the $(w,2w+1)$ T-mers,
-divided by the likelihoods of all the $(w,2w)$ T-mers.
-
 %%%%%%%%

### 2017-05-03 writeups/context-dependence.tex
begun ising testing
various adjustments

@@
 
+%%%%%%%%
+\subsection{Model fit and model selection}
+
+The difference between observed and expected T-mer counts provide a good assessment of how and whether a model can be improved.
+These \emph{residual} counts are usually most usefully computed for T-mers shorter than those used to fit the model.
+For each T-mer $(x,y)$, write $O_{x,y}$ for the number of this T-mer observed in the data, 
+and $E_{x,y}$ for the number expected under the model.
+(The expected number is the number of occurrences of $x$ in the initial sequence 
+multiplied by the probability $p_{n,\ell,r}(t;x,y)$, where $n$ is the length of $x$,
+and $\ell$, $r$ are the appropriate overhangs.)
+If the T-mers were nonoverlapping, this would form a standard contingency table,
+and standard practice would be to divide the residuals by the square root of the expected counts
+and compare these to a standard Normal distribution.
+However, there are are $n$ possible such tables --
+so, we compute normalized residuals as
+\begin{align}
+    Z_{x,y} = \frac{ O_{x,y} - E_{x,y} }{ \sqrt{n E_{x,y}} },
+\end{align}
+which is equivalent to averaging the $z$-scores of the $n$ possible tables of nonoverlapping counts.
+These can then be compared to a standard Normal distribution.
+
 %%%%%%%%%%%
@@
 
-\peter{from simulation: parameter estimation; model selection}
-
-\peter{do a three-taxon tree? infer with shape data?}
+\textbf{To-do:}
+
+1. Posterior coverage of parameters for Ising in 20 runs (more?).
+
+2. With a CpG model with two 3mer motifs added, 1. fit single-base model; 2. add all 2mers; 3. add top one if significant by LR test and repeat; 4. do same for 3mers
+
+3. Analyze human data with shape model, Bayesian.
 
@@
 
+%%%%%%%%%%%
+\subsection{Stepwise model selection}
+
+In exploratory application to real data, how does one decide when to stop adading mutational motifs?
+Perhaps the simplest method conceptually is to 
+(a) fit a simple model,
+(b) choose new motifs to add based on statistically significant residuals,
+(c) repeat until no residuals are statisically significant.
+To test this method,
+we simulated a single sequence of $10^6$ nucleotides
+for one units of time
+under the model
+    \begin{center}
+      \begin{tabular}{c@{\quad$\to$\quad}c@{\quad at rate\quad }cc}
+        $x$  &    $y$  &  $m_{xy}$ & when $x \neq y \in \{\nA,\nC,\nG,\nT\}$  \\
+        \nC\nG     &  \nT\nG     &  $0.4$ & \\
+        \nC\nG     &  \nC\nA     &  $0.4$ & \\
+        \nT\nC\nC  &  \nT\nT\nC  &  $0.1$ & \\
+        \nA\nG\nG  &  \nA\nA\nG  &  $0.1$ &
+      \end{tabular} ,
+    \end{center}
+with $m_{AT}=m_{TA}=0.1$, $m_{CG}=m_{GC}=0.15$, $m_{AC}=m_{TG}=m_{AG}=m_{TC}=0.08$, and $m_{CA}=m_{GT}=m_{CT}=m_{GA}=0.12$.
+% Note the model is symmetric under reverse complementation.
+The resulting sequences differed at 28.5\% of the sites.
+We then fit an unconstrained model of single-nucleotide substitution using (5,1,1) T-mers
+(three sites at the base with an overhang of one on each side).
+The top and bottom of the table of (2,1,0) and (2,0,1) T-mer residuals of this model were as follows:
+    \begin{center}
+        XXX
+    \end{center}
+The situation clearly calls for addition of the CpG mutation tuple.
+After adding these patterns into the model and re-fitting,
+some length 2 residuals had statistically significant residuals,
+but none nearly so large as those seen in (3,1,1) T-mer residuals:
+    \begin{center}
+        XXX
+    \end{center}
+\ldots which again clearly call for adding the appropriate 3mer mutation tuples.
+After adding these in, no further (3,1,1) T-mer residuals achieved
+statistical significance after a Bonferroni correction;
+the final parameter values are shown in Table XXX;
+parameter estimates differ from the truth by less than X\%.
+
+\begin{table}
+  \begin{center}
+    \includegraphics[width=\textwidth]{writeup-plots/tasep_all-mcmc-runs}
+  \end{center}
+  \caption{ Results from iterative fitting to a $10^6$ sequence evolved to have 28.5\% sequence divergence
+    with the above mutational motifs.
+    \label{tab:cpg_results} }
+\end{table}
+
+%%%%%%%%%%%
+\subsection{Bayesian inference}
+
+The usage of fast sparse matrix methods described above make likelihood computation fast enough
+to use a Metropolis-Hasting-based Markov chain Monte Carlo method
+to sample from the posterior distribution on parameters.
+To test this,
+we evolved $10^6$ sites from the Ising model (described above) with $\lambda=0.15$, $\beta=0.3$, and $\gamma=0.15$
+for 1 unit of time.
+We then placed an Exponential prior with mean 1 on both transitions $+ \to -$ and $- \to +$
+(without constraining these to be equal)
+and half-Gaussian priors with mean 0 and scale 3 on both $\beta$ and $\gamma$.
+
+%%%%%%%%%%%
 \subsection{TASEP}

### 2017-05-03 writeups/context-dependence.tex
scaled resids

@@
 We then fit an unconstrained model of single-nucleotide substitution using (5,1,1) T-mers
-(three sites at the base with an overhang of one on each side).
+(three sites at the base with an overhang of one on each side)
+using a constrained quasi-Newton method (``optim(.., method='L-BFGS-B')'' in R).
 The top and bottom of the table of (2,1,0) and (2,0,1) T-mer residuals of this model were as follows:
-    \begin{center}
-        XXX
-    \end{center}
-The situation clearly calls for addition of the CpG mutation tuple.
+% cpg-plus-epsilon/32140/base-model-fit-5-3-l1-resids.2.1.l0.tsv
+% cpg-plus-epsilon/32140/base-model-fit-5-3-l1-resids.2.1.l1.tsv
+    \begin{center}
+        \begin{tabular}{ccrrrr}
+                $x$ & $y$ & observed &   expected &    residual &  $Z$ \\
+                \hline
+                \nC\nG  & \_\nC   & 100410   &  120303    & -19892.50   &	-40.55  \\
+                \nC\nA  & \_\nT   &  19884   &   28350    &  -8466.43   &	-35.55  \\
+                \nC\nT  & \_\nT   &  19998   &   28061    &  -8063.46   &	-34.03  \\
+                \nC\nC  & \_\nT   &  23673   &   28177    &  -4503.96   &	-18.97  \\
+                \nG\nG  & \_\nC   &  17742   &   19418    &  -1675.92   &	 -8.50  \\
+                \nC\nG  & \_\nG   &  18450   &   19478    &  -1028.34   &	 -5.21  \\
+                \nA\nG  & \_\nC   &  10968   &   11760    &   -791.73   &	 -5.16  \\
+                \nG\nG  & \_\nG   & 118805   &  120975    &  -2170.25   &    \\
+                \hline
+                \nA\nG  & \_\nT   &  17280   &   16515    &    765.20   &  4.21   \\
+                \nG\nT  & \_\nC   &  20490   &   19507    &    983.44   &  4.97   \\
+                \nG\nG  & \_\nT   &  20929   &   19278    &   1650.83   &  8.40   \\
+                \nC\nC  & \_\nC   & 124149   &  119880    &   4268.52   &  8.71   \\
+                \nG\nG  & \_\nA   &  30483   &   28288    &   2195.34   &  9.22   \\
+                \nC\nT  & \_\nC   & 126957   &  119389    &   7567.92   & 15.48   \\
+                \nC\nA  & \_\nC   & 128673   &  120619    &   8054.46   & 16.39   \\
+                \nC\nG  & \_\nT   &  49311   &   28276    &  21034.85   & 88.45   \\
+                \hline
+                \hline
+                \nC\nG  &  \nG\_  &  100803  &  120541  &  -19737.80  &  -40.19 \\
+                \nT\nG  &  \nA\_  &   20040  &   28179  &   -8139.45  &  -34.28 \\
+                \nG\nG  &  \nA\_  &   20985  &   28288  &   -7302.96  &  -30.70 \\
+                \nA\nG  &  \nA\_  &   22929  &   28399  &   -5469.73  &  -22.95 \\
+                \nC\nC  &  \nG\_  &   17805  &   19410  &   -1605.01  &   -8.14 \\
+                \nC\nT  &  \nG\_  &   10470  &   11684  &   -1214.31  &   -7.94 \\
+                \nC\nA  &  \nG\_  &   10839  &   11882  &   -1043.39  &   -6.76 \\
+                \nC\nG  &  \nC\_  &   18354  &   19348  &    -994.19  &   -5.05 \\
+                \hline
+                \nC\nT  &  \nA\_  &   17007  &   16078  &     929.33  &    5.18 \\
+                \nT\nC  &  \nG\_  &   20691  &   19509  &    1182.40  &    5.98 \\
+                \nC\nC  &  \nA\_  &   20640  &   19160  &    1480.44  &    7.56 \\
+                \nT\nC  &  \nT\_  &   30309  &   28320  &    1988.91  &    8.35 \\
+                \nA\nG  &  \nG\_  &  127191  &  121450  &    5740.76  &   11.64 \\
+                \nG\nG  &  \nG\_  &  127946  &  120977  &    6969.46  &   14.16 \\
+                \nT\nG  &  \nG\_  &  127539  &  120512  &    7026.52  &   14.31 \\
+                \nC\nG  &  \nA\_  &   49098  &   28186  &   20911.93  &   88.07 \\
+                \hline
+        \end{tabular}
+    \end{center}
+The situation clearly calls for addition of the CpG mutation motif.
 After adding these patterns into the model and re-fitting,
+% starting at the parameters inferred from the first run and a CpG rate of 0.0,
 some length 2 residuals had statistically significant residuals,
-but none nearly so large as those seen in (3,1,1) T-mer residuals:
-    \begin{center}
-        XXX
-    \end{center}
-\ldots which again clearly call for adding the appropriate 3mer mutation tuples.
-After adding these in, no further (3,1,1) T-mer residuals achieved
+but none so large as those seen in (3,1,1) T-mer residuals, shown here:
+    \begin{center}
+        \begin{tabular}{ccrrrr}
+                $x$ & $y$ & observed &   expected &    residual &  $Z$ \\
+                \hline
+                \nA\nG\nG  &  \_\nG\_  &  30024  &  32167  &  -2143.48  &  -6.90 \\
+                \nT\nC\nC  &  \_\nC\_  &  29910  &  31748  &  -1838.13  &  -5.95 \\
+                \nG\nG\nT  &  \_\nA\_  &   5064  &   5549  &   -484.61  &  -3.75 \\
+                \nT\nG\nC  &  \_\nA\_  &   4830  &   5172  &   -342.42  &  -2.74 \\
+                \nG\nG\nA  &  \_\nA\_  &   5115  &   5462  &   -346.97  &  -2.71 \\
+                \nC\nT\nT  &  \_\nG\_  &   2472  &   2717  &   -244.56  &  -2.70 \\
+                \nG\nC\nT  &  \_\nT\_  &   4863  &   5192  &   -328.56  &  -2.63 \\
+                \hline
+                \nA\nT\nG  &  \_\nG\_  &   3225  &   3028  &    196.78  &   2.06  \\
+                \nA\nC\nG  &  \_\nA\_  &   4830  &   4588  &    242.29  &   2.06  \\
+                \nT\nG\nT  &  \_\nC\_  &   5364  &   5105  &    258.74  &   2.09  \\
+                \nG\nA\nG  &  \_\nG\_  &   3285  &   3055  &    229.78  &   2.40  \\
+                \nT\nT\nG  &  \_\nC\_  &   2973  &   2752  &    220.59  &   2.42  \\
+                \nT\nC\nT  &  \_\nG\_  &   5421  &   5056  &    364.54  &   2.95  \\
+                \nT\nC\nC  &  \_\nT\_  &   7590  &   5500  &   2089.79  &  16.26  \\
+                \nA\nG\nG  &  \_\nA\_  &   7566  &   5263  &   2302.74  &  18.32  \\
+                \hline                                                     
+        \end{tabular}
+    \end{center}
+\ldots which again clearly call for adding the mutation tuples $TCC \to TTC$ and $AGG \to AAG$.
+After adding these in, 
+% and initializing all parameters to 0.1 
+no further (3,1,1) T-mer residuals achieved
 statistical significance after a Bonferroni correction;
-the final parameter values are shown in Table XXX;
+the final parameter values are shown in Table \ref{tab:cpg_results};
 parameter estimates differ from the truth by less than X\%.
@@
 (without constraining these to be equal)
-and half-Gaussian priors with mean 0 and scale 3 on both $\beta$ and $\gamma$.
+and half-Gaussian priors with mean 0 and scale 3 on both $\beta$ and $\gamma$,
+and 
+We repeated this procedure separately on XX independently simulated sequences,
+and show in Figure \ref{fig:ising_coverage}
+the 95\% credible intervals for each parameter.
+
+\begin{figure}
+    \begin{center} 
+    \end{center} 
+    \caption{
+        \label{fig:ising_coverage}}
+\end{figure}
 

### 2017-05-03 contextual/R/residuals.R

@@ -55,7 +55,7 @@ computeresids <- function (model, pretty=TRUE, in_longwin=longwin(model), in_sho
                             resid=as.vector(counts)-as.vector(expected),
                             stringsAsFactors=FALSE
                         )
-        residframe$z <- residframe$resid/sqrt(as.vector(expected))
+        residframe$z <- residframe$resid/sqrt(as.vector(expected)*in_longwin)
         residframe <- residframe[order(residframe$z),]
     } else {
         # concise matrix

### 2017-05-03 cpg/cpg-plus-epsilon-workflow.sh

@@ -18,11 +18,17 @@ mkdir -p $BASEDIR
 TRUE_MODEL="$BASEDIR/cpg-plus-epsilon.json"
 # single-base-only model initialized for mean sequence divergence
 BASEMODEL="$BASEDIR/base-model.json"
+# single-base-only with CpG, initialized to the previous model
+MODEL_A="$BASEDIR/cpg-only.json"
+# as above but with Harris's mutations added, initialized to previous model
+# note this is structurally the same as TRUE_MODEL
+MODEL_B="$BASEDIR/cpg-and-harris.json"
+
+## --------
 
 MODEL2="$BASEDIR/base-plus-2mers.json"
 MODEL3="$BASEDIR/base-plus-3mers.json"
 
-MODEL_A="$BASEDIR/cpg-only.json"
 MODEL_C="cpg-plus-epsilon/cpg-plus-1.json"
 MODEL_D="cpg-plus-epsilon/cpg-plus-2.json"
 
@@ -32,7 +38,7 @@ TLEN=1.0
 # precompute generator matrices:
 GMDIR=$BASEDIR/genmatrices
 mkdir -p $GMDIR
-for THISMODEL in $TRUE_MODEL $BASEMODEL
+for THISMODEL in $TRUE_MODEL $BASEMODEL $MODEL_A $MODEL_B
 do
     MODELNAME=$(basename ${THISMODEL%.json})
     for LONGWIN in 4 5
@@ -53,6 +59,17 @@ SEQLEN=1000000
 #     (
 #        DIR=$BASEDIR/$(printf "%05g" $RANDOM);
 
+fitmodel ()
+{
+    THISMODEL=$1
+    MODELNAME=$(basename ${THISMODEL%.json})
+    FITFILE="$DIR/${MODELNAME}-fit-${LONGWIN}-${SHORTWIN}-l${LEFTWIN}.RData";
+    LOGFILE=${FITFILE%RData}log
+    Rscript ../scripts/fit-model.R -c $THISMODEL -t $TLEN -i "$DIR/sim-${LONGWIN}-root-${SHORTWIN}-tip-l${LEFTWIN}-shift0.counts" -m "$GMDIR/genmatrix-${LONGWIN}-${MODELNAME}.RData" -o $FITFILE &>$LOGFILE
[…]

### 2017-05-04 writeups/context-dependence.tex
cpg results

@@
 for one units of time
-under the model
-    \begin{center}
-      \begin{tabular}{c@{\quad$\to$\quad}c@{\quad at rate\quad }cc}
-        $x$  &    $y$  &  $m_{xy}$ & when $x \neq y \in \{\nA,\nC,\nG,\nT\}$  \\
-        \nC\nG     &  \nT\nG     &  $0.4$ & \\
-        \nC\nG     &  \nC\nA     &  $0.4$ & \\
-        \nT\nC\nC  &  \nT\nT\nC  &  $0.1$ & \\
-        \nA\nG\nG  &  \nA\nA\nG  &  $0.1$ &
-      \end{tabular} ,
-    \end{center}
-with $m_{AT}=m_{TA}=0.1$, $m_{CG}=m_{GC}=0.15$, $m_{AC}=m_{TG}=m_{AG}=m_{TC}=0.08$, and $m_{CA}=m_{GT}=m_{CT}=m_{GA}=0.12$.
+under the model given in Table \ref{tab:cpg_results},
+obtained by embellishing the basic CpG hypermutability
+with the longer mutational motif identified by \citet{harris}.
+%     \begin{center}
+%       \begin{tabular}{c@{\quad$\to$\quad}c@{\quad at rate\quad }cc}
+%         $x$  &    $y$  &  $m_{xy}$ & when $x \neq y \in \{\nA,\nC,\nG,\nT\}$  \\
+%         \nC\nG     &  \nT\nG     &  $0.4$ & \\
+%         \nC\nG     &  \nC\nA     &  $0.4$ & \\
+%         \nT\nC\nC  &  \nT\nT\nC  &  $0.1$ & \\
+%         \nA\nG\nG  &  \nA\nA\nG  &  $0.1$ &
+%       \end{tabular} ,
+%     \end{center}
+% with $m_{AT}=m_{TA}=0.1$, $m_{CG}=m_{GC}=0.15$, $m_{AC}=m_{TG}=m_{AG}=m_{TC}=0.08$, and $m_{CA}=m_{GT}=m_{CT}=m_{GA}=0.12$.
 % Note the model is symmetric under reverse complementation.
@@
 the final parameter values are shown in Table \ref{tab:cpg_results};
-parameter estimates differ from the truth by less than X\%.
+parameter estimates differ from the truth by less than 3\% (mostly within 1\%).
+
+% df <- structure(list(
+%     sim = c(0.1, 0.1, 0.15, 0.15, 0.08, 0.08, 0.08, 0.08, 0.12, 0.12, 0.12, 0.12, 0.4, 0.1), 
+%     fit = c(0.0996, 0.0982, 0.1505, 0.1497, 0.0784, 0.0792, 0.0798, 0.0795, 0.1202, 0.1196, 0.1198, 0.1193, 0.3978, 0.1033)), 
+%     .Names = c("sim", "fit"), 
+%     row.names = c("A->T", "T->A", "C->G", "G->C", "A->C", "T->G", "A->G", "T->C", "C->A", "G->T", "C->T", "G->A", "CG->TG|CG->CA", "TCC->TTC|AGG->AAG"), 
+%     class = "data.frame")
 
@@
   \begin{center}
-    \includegraphics[width=\textwidth]{writeup-plots/tasep_all-mcmc-runs}
+        \begin{tabular}{c@{\quad$\to$\quad}c@{\quad at rate\quad }ll}
+          \hline
+            x & y & truth & fitted \\ 
+          \hline
+          \nA  &   \nT           & 0.10 & 0.0996 \\ 
+          \nT  &   \nA           & 0.10 & 0.0982 \\ 
+          \nC  &   \nG           & 0.15 & 0.1505 \\ 
+          \nG  &   \nC           & 0.15 & 0.1497 \\ 
+          \nA  &   \nC           & 0.08 & 0.0784 \\ 
+          \nT  &   \nG           & 0.08 & 0.0792 \\ 
+          \nA  &   \nG           & 0.08 & 0.0798 \\ 
+          \nT  &   \nC           & 0.08 & 0.0795 \\ 
+          \nC  &   \nA           & 0.12 & 0.1202 \\ 
+          \nG  &   \nT           & 0.12 & 0.1196 \\ 
+          \nC  &   \nT           & 0.12 & 0.1198 \\ 
+          \nG  &   \nA           & 0.12 & 0.1193 \\ 
+       \nC\nG  &  \nT\nG         & 0.40 & 0.3978 \\ 
+       \nC\nG  &  \nC\nA         & 0.40 & 0.3978 \\ 
+    \nT\nC\nC  &  \nT\nT\nC      & 0.10 & 0.1033 \\ 
+    \nA\nG\nG  &  \nA\nA\nG      & 0.10 & 0.1033 \\ 
+           \hline
+        \end{tabular}
   \end{center}
   \caption{ Results from iterative fitting to a $10^6$ sequence evolved to have 28.5\% sequence divergence
-    with the above mutational motifs.
+    with the above mutational motifs. The transitions $CG \to TG$ and $CG \to CA$ 
+    were constrained to be equal (DNA strand symmetry),
+    as were the transitions $TCC \to TTC$ and $AGG \to AAG$.
     \label{tab:cpg_results} }

### 2017-05-08 writeups/context-dependence.tex
.

@@
 
-1. Posterior coverage of parameters for Ising in 20 runs (more?).
-
-2. With a CpG model with two 3mer motifs added, 1. fit single-base model; 2. add all 2mers; 3. add top one if significant by LR test and repeat; 4. do same for 3mers
-
 3. Analyze human data with shape model, Bayesian.
@@
 To test this,
-we evolved $10^6$ sites from the Ising model (described above) with $\lambda=0.15$, $\beta=0.3$, and $\gamma=0.15$
-for 1 unit of time.
+we evolved $10^6$ sites from the Ising model (described above) with $\lambda=0.5$, $\beta=1.0$, and $\gamma=0.5$
+for 1 unit of time, 
+resulting in around 460,000 possible mutations
+and differences at around 186,000 sites between initial and final sequence.
 We then placed an Exponential prior with mean 1 on both transitions $+ \to -$ and $- \to +$
@@
 and half-Gaussian priors with mean 0 and scale 3 on both $\beta$ and $\gamma$,
-and 
-We repeated this procedure separately on XX independently simulated sequences,
-and show in Figure \ref{fig:ising_coverage}
-the 95\% credible intervals for each parameter.
+counted $(5,3,1)$ T-mers,
+and used the mcmc package \citep{mcmcR} to run a random walk sampler for 10,000 steps
+(judged sufficient by examination of diagnostic plots).
+We repeated this procedure separately on 60 independently simulated sequences,
+and show the resulting credible intervals in Figure \ref{fig:ising_coverage}.
+Poseterior medians are within half a percent of the truth for $+ \to -$ and $- \to +$ (corresponding to $\lambda$)
+and the temperature ($\beta$), but showed around 2\% error for the magnetic field $\gamma$.
+Credible intervals were somewhat too narrow, as the true value fell within 95\% credible intervals
+for between 55\% and 90\% of the simulations;
+which is likely due to remaining nonindependence at the window sizes used.
 
@@
     \begin{center} 
+        \includegraphics{writeup-plots/coverage_results}
     \end{center} 
     \caption{
+        Credible intervals for the four parameters of the dynamic Ising model,
+        relative to the true value,
+        from 60 MCMC runs on independently simulated sequences of $10^6$ bases
+        differing at around 18.6\% of the sites.
         \label{fig:ising_coverage}}

### 2017-05-09 writeups/context-dependence.tex
added coverage results to paper

@@
 and the temperature ($\beta$), but showed around 2\% error for the magnetic field $\gamma$.
-Credible intervals were somewhat too narrow, as the true value fell within 95\% credible intervals
-for between 55\% and 90\% of the simulations;
-which is likely due to remaining nonindependence at the window sizes used.
+As seen in Table \ref{tab:ising_coverage}, 
+credible intervals were somewhat too narrow for shorter T-mers,
+but $(6,2,2)$ T-mers were long enough with sufficient overhang to produce well-calibrated posteriors.
+
+\begin{table}[ht]
+\centering
+\begin{tabular}{rrrr}
+  \hline
+    & (4,1,1) & (5,1,1) & (6,2,2) \\ 
+  \hline
+  $\lambda_-$ & 0.61 & 0.63 & 0.96 \\ 
+  $\lambda_+$ & 0.87 & 0.86 & 0.85 \\ 
+  $\beta$     & 0.76 & 0.81 & 0.94 \\ 
+  $\gamma$    & 0.69 & 0.76 & 0.85 \\ 
+   \hline
+\end{tabular}
+    \caption{
+        Posterior coverage of the four parameters of the dynamic Ising model at four different T-mer sizes.
+        Shown are the proportion of runs in which the true value fell within the 95\% credible interval,
+        for T-mers of length 4, 5, and 6 with overhangs of length 1, 1, and 2 respectively.
+        The total number of runs was 54 for $(6,2,2)$ T-mers and 114 for the others.
+        \label{tab:ising_coverage}
+    }
+\end{table}
 
@@
         Credible intervals for the four parameters of the dynamic Ising model,
-        relative to the true value,
-        from 60 MCMC runs on independently simulated sequences of $10^6$ bases
+        calculated using $(6,2,2)$ T-mers,
+        from 54 MCMC runs on independently simulated sequences of $10^6$ bases
         differing at around 18.6\% of the sites.
+        Shown are relative intervals, obtained by dividing by the true value.
         \label{fig:ising_coverage}}

### 2017-05-09 ising/coverage-experiment.sh

@@ -52,9 +52,7 @@ do
         # and mcmc
         MCMCID=$RANDOM
         Rscript ../scripts/mcmc-model.R -i $DIR/ising-fit-4-2-1.RData -c ising-model.json -b $MCMCITER -l 1 -j $MCMCID
-        # too long
         Rscript ../scripts/mcmc-model.R -i $DIR/ising-fit-5-3-1.RData -c ising-model.json -b $MCMCITER -l 1 -j $MCMCID
-        # waaay too long
         Rscript ../scripts/mcmc-model.R -i $DIR/ising-fit-6-2-2.RData -c ising-model.json -b $MCMCITER -l 1 -j $MCMCID
 
         # gather results into json
@@ -71,14 +69,26 @@ wait;
 Rscript ../scripts/collect-params-results.R $(ls -t $BASEDIR/*/ising-fit*.json) | tail -n +2 >> coverage_results.tsv
 
 PLOTSCRIPT="
-res <- read.table('coverage_results.tsv', header=TRUE, check.names=FALSE)
-res <- res[grepl('mcmc', res[['file']]) & res[['longwin']]==5,]
+all.res <- read.table('coverage_results.tsv', header=TRUE, check.names=FALSE)
+all.res <- all.res[grepl('mcmc', all.res[['file']]),]
 
 paramnames <- c('O->X', 'X->O', 'OX|XO', 'X')
-true.cols <- match(paste0('sim:', paramnames), colnames(res))
-est.cols <- match(paste0('q50%.', paramnames), colnames(res))
+paramlabels <- c(expression(lambda['+']), expression(lambda['-']), expression(beta), expression(gamma))
+true.cols <- match(paste0('sim:', paramnames), colnames(all.res))
+est.cols <- match(paste0('q50%.', paramnames), colnames(all.res))
 q.cols <- outer(c('q2.5%', 'q25%', 'q75%', 'q97.5%'), paramnames, 
-                function(x,y) { match(paste(x,y,sep='.'), colnames(res)) } )
+                function(x,y) { match(paste(x,y,sep='.'), colnames(all.res)) } )
+
+coverages <- tapply( 1:nrow(all.res), all.res[['longwin']], function (kk) { res <- all.res[kk,]; 
+    sapply( paramnames, function (pn) {
+        k <- match(pn, paramnames)
+        ut <- TRUE
[…]

### 2017-05-10 writeups/context-dependence.tex
updated ising results in writeup

@@
         for T-mers of length 4, 5, and 6 with overhangs of length 1, 1, and 2 respectively.
-        The total number of runs was 54 for $(6,2,2)$ T-mers and 114 for the others.
+        The total number of runs was 61 for $(6,2,2)$ T-mers and 121 for the others.
         \label{tab:ising_coverage}
@@
         calculated using $(6,2,2)$ T-mers,
-        from 54 MCMC runs on independently simulated sequences of $10^6$ bases
+        from 61 MCMC runs on independently simulated sequences of $10^6$ bases
         differing at around 18.6\% of the sites.
@@
 \appendix
+
+\section{Supplementary figures}
+
+\begin{figure}
+    \begin{center} 
+        \includegraphics{writeup-plots/coverage_results_all}
+    \end{center} 
+    \caption{
+        As in \ref{fig:ising_coverage}:
+        Credible intervals for the four parameters of the dynamic Ising model,
+        from MCMC runs on independently simulated sequences of $10^6$ bases
+        differing at around 18.6\% of the sites.
+        Shown are relative intervals, obtained by dividing by the true value.
+        The three groups are left to right, credible intervals obtained from $(4,2,1)$, $(5,3,1)$, and $(6,2,2)$ T-mers:
+        note that only the longest $(6,2,2)$ T-mers provide well-calibrated coverage (Table \ref{tab:ising_coverage})
+        and do not show bias.
+        \label{fig:all_ising_coverage}}
+\end{figure}
 

### 2017-05-12 writeups/context-dependence.tex

@@
 
-\begin{example}[1-D Ising model with Glauber dynamics]
+\begin{example}[Gibbs sampling of the Ising model]
     In the Ising model, each site is labeled as either ``up'' or ``down'' ($+1$ or $-1$ respectively),
@@
     The associated stationary distribution on configurations is proportional to $\exp(-E(x))$.
-    The following dynamics preserve the stationary distribution:
+    The following process, also known as ``Glauber dynamics'', preserves the stationary distribution:
     each site, independently at rate $\lambda$,
@@
         \begin{tabular}{ccrrrr}
+            \hline
                 $x$ & $y$ & observed &   expected &    residual &  $Z$ \\
@@
           \hline
-            x & y & truth & fitted \\ 
+            x & y & truth & MLE \\ 
           \hline
@@
 
-%%%%%%%%%%%
-\subsection{TASEP}
-
-Strings of length $10^5$ were evolved under TASEP for varying amounts of time,
-and pattern occurrences counted.
-Running 10,000 MCMC iterations with window sizes of $\ell=w=3$ (total length 9) on these takes around 40 minutes.
-Apparently, the patterns used were not sufficiently independent in this case,
-as the posterior is too sharp.
-
-Here's a short example sequence with $\lambda=1$:
-\begin{center}
- \setlength{\tabcolsep}{0pt}
-\begin{tabular}{cccccccccccccccccccccccccccccccccccccccccccccccccccccccccccc}
-X&X&X&O&O&X&O&O&O&O&X&O&X&O&X&X&O&O&X&O&O&X&O&X&X&X&O&X&O&O&O&X&X&O&X&O&O&O&X&X&O&O&O&X&O&O&O&O&X&O&X&X&O&X&X&O&O&X&O&O \\
-$\centerdot$&O&X&O&X&O&O&O&X&$\centerdot$&O&X&O&O&X&X&X&o&x&$\centerdot$&o&O&X&x&x&O&X&O&X&o&o&O&X&X&$\centerdot$&$\centerdot$&o&o&O&X&X&o&$\centerdot$&O&X&$\centerdot$&o&o&$\centerdot$&$\centerdot$&O&X&X&x&O&X&o&O&X&$\centerdot$
-\end{tabular}
-\end{center}
-here particles are X's, moving to the right; and spaces are O's;
-in the second line,
-$\centerdot$ is a position that has not considered a change; lowercase letters are positions that considered a change but did not;
-and upper case are positions that did change.
-
-\begin{figure}
-  \begin{center}
-    \includegraphics[width=\textwidth]{writeup-plots/tasep_all-mcmc-runs}
-  \end{center}
-  \caption{ Results from MCMC estimates for six different data sets of the parameter in TASEP (mutation rate multiplied by time).
-  Vertical line shows the true value.
-  }
-\end{figure}
-
-\subsection{Ising model}
-
-Strings of length $10^6$ were evolved under Glauber dynamics on the Ising model with varying parameters,
-and pattern occurrences counted.
-As for TASEP, running 10,000 MCMC iterations with window sizes of $\ell=w=3$ (total length 9) on these takes around 40 minutes.
-In this case, the true parameter values have good distribution within the estimated posterior distributions.
-
-Here is an example, with all parameters equal to 1, for $t=.5$:
-\begin{center}
- \setlength{\tabcolsep}{0pt}
-\begin{tabular}{cccccccccccccccccccccccccccccccccccccccccccccccccccccccccccc}
-X&X&O&X&X&O&O&X&O&X&O&X&X&O&O&X&O&O&O&O&X&X&X&O&X&O&X&O&X&O&O&O&X&X&O&X&X&X&X&X&O&X&O&X&O&X&X&X&X&O&O&X&X&X&X&X&X&O&O&O \\
-x&x&o&x&X&X&O&x&o&x&$\centerdot$&x&x&O&X&X&$\centerdot$&O&X&O&X&x&x&o&x&o&x&o&x&O&X&O&$\centerdot$&$\centerdot$&O&O&X&X&O&X&X&X&o&x&o&$\centerdot$&X&O&X&o&o&x&X&X&X&O&X&o&$\centerdot$&$\centerdot$
-\end{tabular}
-\end{center}
-The notation is as above; except that the entire matched pattern involved in a change is uppercase:
-to simulate, we look in windows of size 3; so if XOO changes to XXO, the whole thing is uppercase, not just the middle X.
-
-\begin{figure}
-  \begin{center}
-    \includegraphics{writeup-plots/selsims-2013-05-28-17-12-0275615-traces}
-  \end{center}
-  \caption{
-  MCMC traces of the parameters in the Ising model across 100,000 iterations.
-  The units of $\lambda$ are in mean number of changes per site.
-  Vertical line shows the true value.
-  }
-\end{figure}
-
-\begin{figure}
-  \begin{center}
-    \includegraphics[width=\textwidth,page=1]{writeup-plots/ising_all-mcmc-runs}
-  \end{center}
-  \caption{ Results from MCMC estimates of the time-scaled mutation rate parameter $\lambda t$,
-  in ten different simulated datasets under the Ising model.
-  Vertical line shows the true value.
-  }
-\end{figure}
-
-\begin{figure}
-  \begin{center}
-    \includegraphics[width=\textwidth,page=2]{writeup-plots/ising_all-mcmc-runs}
-  \end{center}
-  \caption{ Results from MCMC estimates of the inverse temperature $\beta$ in the Ising model.
-  Vertical line shows the true value.
-  }
-\end{figure}
-
-\begin{figure}
-  \begin{center}
-    \includegraphics[width=\textwidth,page=3]{writeup-plots/ising_all-mcmc-runs}
-  \end{center}
-  \caption{ Results from MCMC estimates of the magnetization parameter $\beta$ in the Ising model.
-  Vertical line shows the true value.
-  }
-\end{figure}
-
-%%%
-\subsection{CpG mutation, on a tree}
-
-Nucleotide sequences of length $10^6$ were simulated down two (unequal) branches of a tree,
-with all single-base substitution rates $m_{xy}$ set equal (varying from a total of .01 to .1 across the tree)
-and an additional CpG rate $\gamma$ set to either 0 or three times the single-base rate.
-Analysis was done with $\ell=2$ and $w=1$ (a five-to-one window).
-
-The state at the root of the tree was independently chosen bases with equal frequencies,
-and these frequencies at the root were added as parameters to the model (with a Dirichlet prior).
-
-Here is a short example, with all parameters equal to 1, with each branch of the tree having length $0.1$:
-\begin{center}
- \setlength{\tabcolsep}{0pt}
-\begin{tabular}{cccccccccccccccccccccccccccccccccccccccccccccccccccccccccccc}
-T&G&G&A&G&A&G&G&T&C&T&G&C&C&A&T&A&G&A&C&G&G&G&C&G&C&C&G&C&C&T&A&G&G&A&G&C&C&G&C&G&T&T&T&T&G&C&G&T&G&G&T&G&T&T&G&T&G&G&C \\
-$\centerdot$&$\centerdot$&g&C&G&a&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&T&A&C&A&$\centerdot$&$\centerdot$&$\centerdot$&T&A&T&$\centerdot$&$\centerdot$&$\centerdot$&T&G&c&C&C&A&C&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&G&G&G&t&A&T&$\centerdot$&$\centerdot$&G&A&g&C&G&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&t&T&G&A \\
-$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&A&T&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&T&T&G&T&$\centerdot$&$\centerdot$&G&A&T&$\centerdot$&$\centerdot$&$\centerdot$&G&G&c&t&A&A&G&C&$\centerdot$&$\centerdot$&$\centerdot$&G&T&G&A&t&t&t&g&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&G&A&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&A&C
-\end{tabular}
-\end{center}
-Top is ancestral sequence (``unobserved''); bottom two are at the tips.
-
-The point estimates are good,
-as are the MCMC estimates,
-but there seems to be mixing trouble having to do with the root frequencies.
-
-\begin{figure}
-  \begin{center}
-    \includegraphics{writeup-plots/selsims-2013-06-03-13-17-0790276-initfreqs}
-  \end{center}
-  \caption{
-  MCMC traces across 660,000 iterations of the frequencies at the root of the (two-taxon) tree in the CpG model.
-  Vertical line shows the true value.
-  }
-\end{figure}
-
-\begin{figure}
-  \begin{center}
-    \includegraphics{writeup-plots/selsims-2013-06-03-13-17-0790276-mutrates}
-  \end{center}
-  \caption{
-  MCMC traces across 660,000 iterations of several mutation rate parameters in the CpG model.
-  Units are in mean number of substitutions per base across the entire tree.
-  Vertical line shows the true value.
-  }
-\end{figure}
-
-%%%%%%%%
-\section{Power sims}
-
-Assess power and accuracy in the Ising model, by running replicates at these parameters:
-
-\begin{tabular}{|rlllrrr|}
-  \hline
-  seqlen   &  $\lambda t$   &  $\beta$  &  $\gamma$  &  $w$ &  $\ell$   & \# reps \\
-  \hline
-  $10^4$    &   .2  &   1   &   .5  &   3   &   3   &   10  \\
-  $10^3$    &   .2  &   1   &   .5  &   3   &   3   &   10  \\
-  $10^2$    &   .2  &   1   &   .5  &   3   &   3   &   10  \\
-  \hline
-  $10^4$    &   .4  &   1   &   .5  &   3   &   3   &   10  \\
-  $10^4$    &   .1  &   1   &   .5  &   3   &   3   &   10  \\
-  $10^4$    &   .05  &   1   &   .5  &   3   &   3   &   10  \\
-  \hline
-  $10^4$    &   .2  &   1   &   0  &   3   &   3   &   10  \\
-  \hline
-  $10^4$    &   .2  &   1   &   .5  &   3   &   2   &   10  \\
-  $10^4$    &   .2  &   1   &   .5  &   3   &   1   &   10  \\
-  $10^4$    &   .2  &   1   &   .5  &   3   &   0   &   10  \\
-  \hline
-\end{tabular}
-
 %%%%%%%%%%%%%%%%%%%%
@@
 
-\begin{itemize}
-
-  \item estimating freq spectrum [Hernandez]
-
-  \item probabilistic ancestral states for common ancestor of e.g. humans 4N generations ago [Siepel] -- would widely used for primates
-
-  \item ancestral CpG
-
-  \item ancient DNA
-
-  \item noisy image transmission denoising: first learn model; then reduce redundancy
-
-  \item root distribution and confounding/mixing
-
-  \item shrinkage prior on highly parameterized model
-
-  \item Show that taking no context ($\ell=0$) can be positively misleading in some context.
-
-  \item How to identify motifs in the residuals?
-
-\end{itemize}
+This paper presents a tractable, provably correct method
+for doing statistical inference using observations of a particle model
+in which probabilities of change are determined only by nearby sites.
+Such models appear in many fields, for instance,
+as the dynamic counterparts to Markov random fields;
+many spin glass models;
+and models of DNA sequence evolution.
+
+The methods described here provide a way to efficiently calculate full likelihoods,
+which in many circumstances provide the optimal basis for inference \citep{neymanpearson},
+fast enough to allow Bayesian inference from Monte Carlo sampling.
+The method is fast thanks to its dependence only only local neighborhoods
+(so the time does not scale with sequence length once T-mer abundances are counted)
+and its reliance on sparse matrices whose structure can be precomputed and cached.
+It avoids approximations made by previous methods thanks to the use of T-mers
+and explict quantification of the necessary length of the ``arms'' of the T-mers.
+We first discuss possible uses of the method,
+followed by challenges (and possible solutions).
+
+\paragraph{Other applications}
+One class of applications stems from \emph{inference of process} --
+where the numerical values of various mutation rates are directly of interest.
+For instance, CpG hypermutability depends on the \nC\nG pair being methylated,
+an epigenetic modification that has important consequences for gene regulation;
+therefore, inference of CpG mutation rates in ancient taxa provides an indirect estimate
+of methylation status in extinct organisms.
+Similarly, many mutational motifs are thought to be caused by particular enzymes;
+inference of relative rates of these motifs
+may provide clues as to the activity and evolution of those enzymes.
+Furthermore, iterative model building by examining residual motif counts
+could provide a powerful way to discover unknown processes
+whose action had been obscured by other sources of noise.
+
+In other applications, reconstruction of an unknown sequence may of interest.
+For instance, training data might be used to estimate parameters of a noisy transmission line,
+facilitating the later reconstruction of noisily transmitted signals.
+Alternatively, multiple observations of independently transmitted copies of a single sequence
+(e.g., down a phylogenetic tree)
+might be used to estimate the process, and from this infer the original (ancestral) sequence
+(or sample from its posterior distribution).
+
+\paragraph{Limitations}
+The model presented assumes that the same process occurs at every location along the sequence,
+and that the length of the sequnce does not change (no indels).
+Neither of these assumptions are likely to be true for any data set of DNA sequence;
+however, both assumptions are also made the state of the art in phylogenetics.
+Some relaxation of the assumptions is possible -- like in phylogenetics \citep{ratemixtures},
+we can easily allow heterogeneity between sites by scaling rates by an overall gamma-distributed random factor.
+Short deletions might be accomodated by introducing a ``gap'' letter to the alphabet,
+but insertions may be problematic.
+
+\paragraph{Computation}
+The main factor that determines computation time
+is the lengths of the T-mers used for inference,
+since all computations scale with the number of possible patterns at the long end of the T-mer.
+Although the model is efficent enough to deal with fairly long T-mers,
+this exponential growth puts hard constraints on the application of the method.
+The length of T-mers required to accurately compute the likelihood in most models grows at most linearly
+with the mean density of substitution between the initial and final sequence.
+This does not in practice greatly increase the length of required T-mer 
+since substitution densities approaching 100\% are unlikely to be usable in practice
+especially if how the sequences align with each other is unknown.
+A more serious determining factor of T-mer length is the degree of dependency in the model itself.
+For instance, inserting the 13-base pair motif selected against by meiotic drive in primates \citep{myers2010drive}
+would require T-mers of width at least 25, and hence vectors of length $4^25 \approx 1.13 \times 10^15$.
+
+\paragraph{Modeling}
+Careful building of complex models in this framework may take care,
+but no more than in any other class of models (e.g., multivariate linear regression).
+It is easy to specify nonidentifiable models -- 
+for instance, all 2mer mutation motifs include the 1mer motifs.
+The large number of possible mutation motifs also makes multiple testing an issue;
+an unguided approach might include all possible motifs but place a shrinkage prior on their values \citep{shrinkage,horseshoe}.
+The converse approach of adding motifs that are overrepresented in the residuals
+may be quicker, but entails some degree of arbitrariness in choosing the motifs to add (and in which order).
+Finally, inference on trees may be difficult unless the root placement and distribution are strongly constrained:
+otherwise, initial investigations suggest the likelihood surface is often strongly ridged --
+for instance, increasing the abundance of \nG nucleotides at the root while also increasing the mutation rate away from \nG
+can achieve roughly the same result.
+However, in phylogenetic applications outgroups can be used to constrain these quantities.
 
 \subsection{Acknowledgements}
-Erick, Jessica, Graham, Yaniv, Sergey, Simon, Matt, Rasmus
+The initial impetus for this project came from Matt Dean; 
+along the way substantial encouragement and useful suggestions were provided by
+Erick Matsen, Simon Tavar\'e, Rasmus Nielsen, Graham Coop, Yaniv Brandvain, and Sergey Nuzhdin.
+In particular, Erick Matsen helped improve the code and suggested the name ``T-mer''.
+Many thanks are also due Jessica Crisci for curating the human-chimp data.
 

### 2017-06-02 writeups/context-dependence.tex
more CG

@@
 \newcommand{\nT}{\mbox{T}}
+\newcommand{\nW}{\mbox{W}}
+\newcommand{\nR}{\mbox{R}}
+\newcommand{\nY}{\mbox{Y}}
 \newcommand{\join}{\oplus}  % matches
@@
 The algorithm for a general tree is written out in Appendix \ref{ss:peeling_algorithm}.
+
+%%%%%%%%%%%
+\subsection{Homonid data}
+
+We also applied the model to human--chimpanzee divergence data, 
+restricting the comparison to putative regulatory sequences,
+avoiding the additional constraints of coding sequences.
+To do this, we extracted regions in the human (hg38)--chimpanzee (panTro4) alignment \citep[from the UCSC Genome Browser][]{pantroalign}
+according to four types of regulatory feature in the Ensembl Regulatory Build \citep[release 81][]{ensebml}:
+``CTCF binding site'', ``enhancer'', ``open chromatin region'', and ``promoter flanking region''.
+These were furthermore divided into regions that either did or did not overlap a transcription start or end site of a known gene
+from the hg38 Annotations Database \citep{hg_annotationsdb}.
+We then counted Tmers of appropriate length in each of the eight resulting sets of aligned regions, omitting any Tmers that included gaps or repeat-masked positions,
+and fit models separately to each of the eight data sets.
+
+We constructed models from three, nested, sets of mutation motifs.
+All models are strand-symmetric.
+\begin{itemize}
+    \item[(basic)] 
+        Single-base changes, with separate rates for transitions ($\mu_v$),
+        \nA$\leftrightarrow$\nT ($\mu_{A}T$), \nC$\leftrightarrow$\nG ($\mu_{CG}$),
+        and other transversions ($\mu_{r}$):
+          \begin{center}
+            \begin{tabular}{c@{\quad$\to$\quad}c@{\quad at rate\quad }c|c@{\quad$\to$\quad}c@{\quad at rate\quad }c|c@{\quad$\to$\quad}c@{\quad at rate\quad }c}
+                \nT  &   \nA   &  $\mu_{AT}$ & \nA  &   \nG   &  $\mu_{r}$ & \nA  &   \nG   &  $\mu_{v}$ \\
+                \nA  &   \nT   &  $\mu_{AT}$ & \nG  &   \nA   &  $\mu_{r}$ & \nG  &   \nA   &  $\mu_{v}$ \\
+                \nC  &   \nG   &  $\mu_{CG}$ & \nC  &   \nT   &  $\mu_{r}$ & \nT  &   \nC   &  $\mu_{v}$ \\
+                \nG  &   \nC   &  $\mu_{CG}$ & \nT  &   \nC   &  $\mu_{r}$ & \nC  &   \nT   &  $\mu_{v}$ \\
+            \end{tabular}
+          \end{center}
+
+      \item[(CpG)] In addition to the above, an addition CpG rate:
+          \begin{center}
+            \begin{tabular}{c@{\quad$\to$\quad}c@{\quad at rate\quad }c}
+                CG  &   TG   &  $\nu_{CpG}$ \\
+                CG  &   CA   &  $\nu_{CpG}$ \\
+            \end{tabular}
+          \end{center}
+
+      \item[(repair)] In addition to the above, a collection of motifs associated with known biochemical causes
+          curated from the literature:
+          deamination due to repair of UV-induced pyramidine dimers ((\nT/\nC)\nC $\to$ (\nT/\nC)\nT and \nC\nC$\to$\nT\nT, \citet{uv_damage});
+          an interaction between CpG and pyramidine dimers (\nT\nC\nG$\to$\nT\nT\nG and \nC\nC\nG$\to$\nC\nT\nG, \citet{cpg_review});
+          repair of guanine adducts at CpG sites (\nC\nG$\to$\nC\nT, \citet{tobacco});
+          deamination by AID (\nW\nR\nC\nY$\to$\nW\nR\nT\nY, \nW is \nA or \nT; \nR is \nA or \nG; \nY is \nT or \nC, \citet{AID});
+          deamination by members of the APOBEC family (\nT\nC\nW$\to$\nT\nT\nW or \nT\nC\nW$\to$\nT\nG\nW, \citet{APOBEC});
+          and repair by error-prone polymerases pol $\eta$ (\nW\nA$\to$\nW\nG, \citet{pol_eta});
+          and pol $\iota$ (\nA\nA$\to$\nG\nT and \nA\nA$\to$\nA\nG \citet{pol_iota}):
+          \begin{center}
+            \begin{tabular}{c@{\quad$\to$\quad}c@{\quad at rate\quad }c|c@{\quad$\to$\quad}c@{\quad at rate\quad }c|c@{\quad$\to$\quad}c@{\quad at rate\quad }c}
+                \nT\nC     &   \nT\nT      &  $\mu_{UV1}$ & \nA  &   \nG   &  $\mu_{r}$ & \nA  &   \nG   &  $\mu_{v}$ \\
+                \nG\nA     &   \nA\nA      &  $\mu_{UV1}$ & \nA  &   \nG   &  $\mu_{r}$ & \nA  &   \nG   &  $\mu_{v}$ \\
+                \nC\nC     &   \nC\nT      &  $\mu_{UV2}$ & \nA  &   \nG   &  $\mu_{r}$ & \nA  &   \nG   &  $\mu_{v}$ \\
+                \nG\nG     &   \nA\nG      &  $\mu_{UV2}$ & \nA  &   \nG   &  $\mu_{r}$ & \nA  &   \nG   &  $\mu_{v}$ \\
+                \nC\nC     &   \nT\nT      &  $\mu_{UV3}$ & \nA  &   \nG   &  $\mu_{r}$ & \nA  &   \nG   &  $\mu_{v}$ \\
+                \nG\nG     &   \nA\nA      &  $\mu_{UV3}$ & \nA  &   \nG   &  $\mu_{r}$ & \nA  &   \nG   &  $\mu_{v}$ \\
+                \nT\nC\nG  &   \nT\nT\nG   &  $\mu_{mUV}$ & \nA  &   \nG   &  $\mu_{r}$ & \nA  &   \nG   &  $\mu_{v}$ \\
+                \nC\nC\nG  &   \nC\nT\nG   &  $\mu_{mUV}$ & \nA  &   \nG   &  $\mu_{r}$ & \nA  &   \nG   &  $\mu_{v}$ \\
+                \nC\nG\nA  &   \nC\nA\nA   &  $\mu_{mUV}$ & \nA  &   \nG   &  $\mu_{r}$ & \nA  &   \nG   &  $\mu_{v}$ \\
+                \nC\nG\nG  &   \nC\nA\nG   &  $\mu_{mUV}$ & \nA  &   \nG   &  $\mu_{r}$ & \nA  &   \nG   &  $\mu_{v}$ \\
+            \end{tabular}
+          \end{center}
+
+% [ [ "TC", "TT" ], [ "GA", "AA" ] ], UV1
+% [ [ "CC", "CT" ], [ "GG", "AG" ] ], UV2
+% [ [ "CC", "TT" ], [ "GG", "AA" ] ], UV3
+% [ [ "TCG", "TTG" ], [ "CCG", "CTG" ], [ "CGA", "CAA" ], [ "CGG", "CAG" ] ], mUV
+% [ [ "CG", "CT" ], [ "CG", "AG" ] ],
+% [ [ "AACT", "AATT" ], [ "AACC", "AATC" ], [ "AGCT", "AGTT" ], [ "AGCC", "AGTC" ], [ "TACT", "TATT" ], [ "TACC", "TATC" ], [ "TGCT", "TGTT" ], [ "TGCC", "TGTC" ],
+%   [ "AGTT", "AATT" ], [ "GGTT", "GATT" ], [ "AGCT", "AACT" ], [ "GGCT", "GACT" ], [ "AGTA", "AATA" ], [ "GGTA", "GATA" ], [ "AGCA", "AACA" ], [ "GGCA", "GACA" ] ],
+% [ [ "TCA", "TTA" ], [ "TCT", "TTT" ], [ "TCA", "TGA" ], [ "TCT", "TGT" ],
+%   [ "TGA", "TAA" ], [ "AGA", "AAA" ], [ "TGA", "TCA" ], [ "AGA", "ACA" ] ],
+% [ [ "TA", "TG" ], [ "TA", "CA" ] ],
+% [ [ "AA", "GT" ], [ "TT", "AC" ] ],
+% [ [ "AA", "AG" ], [ "CT", "TT" ] ]
+
+\end{itemize}
 
@@
 
+%%%%%%%%%%%
+\subsection{Hominid mutation spectrum}
+
+We also applied the method to changes between the human and chimpanzee reference sequences.
+
 %%%%%%%%%%%%%%%%%%%%

### 2017-06-03 writeups/context-dependence.tex
mcmc tree models

@@
 and that multinucleotide substitutions are relatively common \citep{schrider2011pervasive,terekhanova2013prevalence,harris2013errorprone}.
+In humans, the molecular processes underlying mutations are best studied in contexts of intense somatic mutation:
+cancer and the hypervariable regions that generate immune system diversity \citep{evolution_within},
+which provide quite different cellular contexts but share commonalities due to shared molecular machinery.
+In general, substitution rates depend on sequence context due to the interaction of a great many factors including:
+(a) likelihood of DNA damage or errors in cell division (e.g., UV damage \citep{UV} or non-homologous end joining \citep{NHEJ});
+(b) activity levels and structural properties of the enzymes responsible for repair (e.g., base excision repar \citep{BER} and Y-family polymerases \citep{polymerase_review});
+and (c) any functional effects of an error (e.g., lethal mutations will not be seen in a population census).
 (We are concerned with effects that are \emph{homogeneous} across the sequence,
 so do not consider further the many constraints on protein-coding sequence \citep[reviewed in][]{thorne2007protein}.)
+Some of these effects -- particularly, activity levels of different enzymes in the germline --
+will likely vary over evolutionary time,
+and so comparisons between widely separated species will identify time-averaged mutation rates.
+Recent studies have used nonnegative matrix factorization to identify sets of mutational patterns
+co-occurring in comparisons (a) between different cancer types \citep{cancer_scan} and (b) between different human populations \citep{harris,matthieson},
+to disentangle distinct signals presumably coming from the action of distinct sources of mutation.
 
@@
 %%%%%%%%%%%
-\subsection{Homonid data}
+\subsection{Hominid data}
 
@@
           curated from the literature:
-          deamination due to repair of UV-induced pyramidine dimers ((\nT/\nC)\nC $\to$ (\nT/\nC)\nT and \nC\nC$\to$\nT\nT, \citet{uv_damage});
-          an interaction between CpG and pyramidine dimers (\nT\nC\nG$\to$\nT\nT\nG and \nC\nC\nG$\to$\nC\nT\nG, \citet{cpg_review});
-          repair of guanine adducts at CpG sites (\nC\nG$\to$\nC\nT, \citet{tobacco});
-          deamination by AID (\nW\nR\nC\nY$\to$\nW\nR\nT\nY, \nW is \nA or \nT; \nR is \nA or \nG; \nY is \nT or \nC, \citet{AID});
-          deamination by members of the APOBEC family (\nT\nC\nW$\to$\nT\nT\nW or \nT\nC\nW$\to$\nT\nG\nW, \citet{APOBEC});
-          and repair by error-prone polymerases pol $\eta$ (\nW\nA$\to$\nW\nG, \citet{pol_eta});
-          and pol $\iota$ (\nA\nA$\to$\nG\nT and \nA\nA$\to$\nA\nG \citet{pol_iota}):
+          deamination due to repair of UV-induced pyramidine dimers ($\mu_{UV}$: (\nT/\nC)\nC $\to$ (\nT/\nC)\nT and \nC\nC$\to$\nT\nT, \citet{uv_damage});
+          an interaction between CpG and pyramidine dimers ($\mu_{mUV}$: \nT\nC\nG$\to$\nT\nT\nG and \nC\nC\nG$\to$\nC\nT\nG, \citet{cpg_review});
+          repair of guanine adducts at CpG sites ($\mu_g$: \nC\nG$\to$\nC\nT, \citet{tobacco});
+          deamination by AID ($\mu_{AID}$: \nW\nR\nC\nY$\to$\nW\nR\nT\nY, \nW is \nA or \nT; \nR is \nA or \nG; \nY is \nT or \nC, \citet{AID});
+          deamination by members of the APOBEC family ($\mu_{APOBEC}$: \nT\nC\nW$\to$\nT\nT\nW or \nT\nC\nW$\to$\nT\nG\nW, \citet{APOBEC});
+          and repair by error-prone polymerases pol $\eta$ ($\mu_\eta$: \nT\nA$\to$\nT\nG, \citet{pol_eta});
+          and pol $\iota$ ($\mu_{\iota}$: \nA\nA$\to$\nG\nT and \nA\nA$\to$\nA\nG \citet{pol_iota}):
           \begin{center}
             \begin{tabular}{c@{\quad$\to$\quad}c@{\quad at rate\quad }c|c@{\quad$\to$\quad}c@{\quad at rate\quad }c|c@{\quad$\to$\quad}c@{\quad at rate\quad }c}
-                \nT\nC     &   \nT\nT      &  $\mu_{UV1}$ & \nA  &   \nG   &  $\mu_{r}$ & \nA  &   \nG   &  $\mu_{v}$ \\
-                \nG\nA     &   \nA\nA      &  $\mu_{UV1}$ & \nA  &   \nG   &  $\mu_{r}$ & \nA  &   \nG   &  $\mu_{v}$ \\
-                \nC\nC     &   \nC\nT      &  $\mu_{UV2}$ & \nA  &   \nG   &  $\mu_{r}$ & \nA  &   \nG   &  $\mu_{v}$ \\
-                \nG\nG     &   \nA\nG      &  $\mu_{UV2}$ & \nA  &   \nG   &  $\mu_{r}$ & \nA  &   \nG   &  $\mu_{v}$ \\
-                \nC\nC     &   \nT\nT      &  $\mu_{UV3}$ & \nA  &   \nG   &  $\mu_{r}$ & \nA  &   \nG   &  $\mu_{v}$ \\
-                \nG\nG     &   \nA\nA      &  $\mu_{UV3}$ & \nA  &   \nG   &  $\mu_{r}$ & \nA  &   \nG   &  $\mu_{v}$ \\
-                \nT\nC\nG  &   \nT\nT\nG   &  $\mu_{mUV}$ & \nA  &   \nG   &  $\mu_{r}$ & \nA  &   \nG   &  $\mu_{v}$ \\
-                \nC\nC\nG  &   \nC\nT\nG   &  $\mu_{mUV}$ & \nA  &   \nG   &  $\mu_{r}$ & \nA  &   \nG   &  $\mu_{v}$ \\
-                \nC\nG\nA  &   \nC\nA\nA   &  $\mu_{mUV}$ & \nA  &   \nG   &  $\mu_{r}$ & \nA  &   \nG   &  $\mu_{v}$ \\
-                \nC\nG\nG  &   \nC\nA\nG   &  $\mu_{mUV}$ & \nA  &   \nG   &  $\mu_{r}$ & \nA  &   \nG   &  $\mu_{v}$ \\
+                \nT\nC     &   \nT\nT      &  $\mu_{UV1}$    & \nC\nG     &   \nC\nT      &  $\mu_{g}$      & \nA\nA\nC\nT  &   \nA\nA\nT\nT   &  $\mu_{AID}$ \\
+                \nG\nA     &   \nA\nA      &  $\mu_{UV1}$    & \nC\nG     &   \nA\nG      &  $\mu_{g}$      & \nA\nA\nC\nC  &   \nA\nA\nT\nC   &  $\mu_{AID}$ \\
+                \nC\nC     &   \nC\nT      &  $\mu_{UV2}$    & \nT\nC\nA  &   \nT\nT\nA   &  $\mu_{APOBEC}$ & \nA\nG\nC\nT  &   \nA\nG\nT\nT   &  $\mu_{AID}$ \\
+                \nG\nG     &   \nA\nG      &  $\mu_{UV2}$    & \nT\nC\nT  &   \nT\nT\nT   &  $\mu_{APOBEC}$ & \nA\nG\nC\nC  &   \nA\nG\nT\nC   &  $\mu_{AID}$ \\
+                \nC\nC     &   \nT\nT      &  $\mu_{UV3}$    & \nT\nC\nA  &   \nT\nG\nA   &  $\mu_{APOBEC}$ & \nA\nA\nC\nT  &   \nA\nA\nT\nT   &  $\mu_{AID}$ \\
+                \nG\nG     &   \nA\nA      &  $\mu_{UV3}$    & \nT\nC\nT  &   \nT\nG\nT   &  $\mu_{APOBEC}$ & \nA\nA\nC\nC  &   \nA\nA\nT\nC   &  $\mu_{AID}$ \\
+                \nT\nC\nG  &   \nT\nT\nG   &  $\mu_{mUV}$    & \nT\nG\nA  &   \nT\nA\nA   &  $\mu_{APOBEC}$ & \nA\nG\nC\nT  &   \nA\nG\nT\nT   &  $\mu_{AID}$ \\
+                \nC\nC\nG  &   \nC\nT\nG   &  $\mu_{mUV}$    & \nA\nG\nA  &   \nA\nA\nA   &  $\mu_{APOBEC}$ & \nA\nG\nC\nC  &   \nA\nG\nT\nC   &  $\mu_{AID}$ \\
+                \nC\nG\nA  &   \nC\nA\nA   &  $\mu_{mUV}$    & \nT\nG\nA  &   \nT\nC\nA   &  $\mu_{APOBEC}$ & \nT\nA\nC\nT  &   \nT\nA\nT\nT   &  $\mu_{AID}$ \\
+                \nC\nG\nG  &   \nC\nA\nG   &  $\mu_{mUV}$    & \nA\nC\nA  &   \nA\nC\nA   &  $\mu_{APOBEC}$ & \nT\nA\nC\nC  &   \nT\nA\nT\nC   &  $\mu_{AID}$ \\
+                \nT\nA     &   \nT\nG      &  $\mu_{\eta}$   & \nA\nA    &    \nA\nG      &  $\mu_{\iota1}$ & \nT\nG\nC\nT  &   \nT\nG\nT\nT   &  $\mu_{AID}$ \\
+                \nT\nA     &   \nC\nA      &  $\mu_{\eta}$   & \nT\nT    &    \nC\nT      &  $\mu_{\iota1}$ & \nT\nG\nC\nC  &   \nT\nG\nT\nC   &  $\mu_{AID}$ \\
+                \nA\nA     &   \nG\nT      &  $\mu_{\iota2}$ &           &                &                 & \nT\nA\nC\nT  &   \nT\nA\nT\nT   &  $\mu_{AID}$ \\
+                \nT\nT     &   \nA\nC      &  $\mu_{\iota2}$ &           &                &                 & \nT\nA\nC\nC  &   \nT\nA\nT\nC   &  $\mu_{AID}$ \\
+                           &               &                 &           &                &                 & \nT\nG\nC\nT  &   \nT\nG\nT\nT   &  $\mu_{AID}$ \\
+                           &               &                 &           &                &                 & \nT\nG\nC\nC  &   \nT\nG\nT\nC   &  $\mu_{AID}$ \\
             \end{tabular}
@@
 
+In the genomic context,
+if we knew the specific mutational spectrum of each possible repair or copying pathways,
+then this framework could be used to rigorously estimate their average usages and impacts.
+Despite substantial progress over the past 15 years \citep{polymerase_reviews},
+this is still far from known.
+A promising alternative would be to use the distinct signatures obtained by dimension reduction techniques as in
+\citet{cancer_scan}, \citep{matthieson} or \citep{pritchard} as a proxy for unknown but distinct pathways.
+
 \subsection{Acknowledgements}

### 2017-06-04 writeups/context-dependence.tex
.

@@
 These were furthermore divided into regions that either did or did not overlap a transcription start or end site of a known gene
-from the hg38 Annotations Database \citep{hg_annotationsdb}.
+from the hg38 Annotations Database \citep{hg_annotationsdb}
+and filtered for length:
+aligned regions below either 3000bp (promoter flanking regions) or 1000bp (all other types).
 We then counted Tmers of appropriate length in each of the eight resulting sets of aligned regions, omitting any Tmers that included gaps or repeat-masked positions,

### 2017-06-04 writeups/context-dependence.tex
various

@@
 \newcommand{\aloglike}{\widetilde{\loglike}}
+
+\newcommand{\APOBEC}{\text{\tiny APOBEC}}
+\newcommand{\AID}{\text{\tiny AID}}
 
@@
           deamination by AID ($\mu_{AID}$: \nW\nR\nC\nY$\to$\nW\nR\nT\nY, \nW is \nA or \nT; \nR is \nA or \nG; \nY is \nT or \nC, \citet{AID});
-          deamination by members of the APOBEC family ($\mu_{APOBEC}$: \nT\nC\nW$\to$\nT\nT\nW or \nT\nC\nW$\to$\nT\nG\nW, \citet{APOBEC});
+          deamination by members of the APOBEC family ($\mu_{\APOBEC}$: \nT\nC\nW$\to$\nT\nT\nW or \nT\nC\nW$\to$\nT\nG\nW, \citet{APOBEC});
           and repair by error-prone polymerases pol $\eta$ ($\mu_\eta$: \nT\nA$\to$\nT\nG, \citet{pol_eta});
           and pol $\iota$ ($\mu_{\iota}$: \nA\nA$\to$\nG\nT and \nA\nA$\to$\nA\nG \citet{pol_iota}):
-          \begin{center}
-            \begin{tabular}{c@{\quad$\to$\quad}c@{\quad at rate\quad }c|c@{\quad$\to$\quad}c@{\quad at rate\quad }c|c@{\quad$\to$\quad}c@{\quad at rate\quad }c}
-                \nT\nC     &   \nT\nT      &  $\mu_{UV1}$    & \nC\nG     &   \nC\nT      &  $\mu_{g}$      & \nA\nA\nC\nT  &   \nA\nA\nT\nT   &  $\mu_{AID}$ \\
-                \nG\nA     &   \nA\nA      &  $\mu_{UV1}$    & \nC\nG     &   \nA\nG      &  $\mu_{g}$      & \nA\nA\nC\nC  &   \nA\nA\nT\nC   &  $\mu_{AID}$ \\
-                \nC\nC     &   \nC\nT      &  $\mu_{UV2}$    & \nT\nC\nA  &   \nT\nT\nA   &  $\mu_{APOBEC}$ & \nA\nG\nC\nT  &   \nA\nG\nT\nT   &  $\mu_{AID}$ \\
-                \nG\nG     &   \nA\nG      &  $\mu_{UV2}$    & \nT\nC\nT  &   \nT\nT\nT   &  $\mu_{APOBEC}$ & \nA\nG\nC\nC  &   \nA\nG\nT\nC   &  $\mu_{AID}$ \\
-                \nC\nC     &   \nT\nT      &  $\mu_{UV3}$    & \nT\nC\nA  &   \nT\nG\nA   &  $\mu_{APOBEC}$ & \nA\nA\nC\nT  &   \nA\nA\nT\nT   &  $\mu_{AID}$ \\
-                \nG\nG     &   \nA\nA      &  $\mu_{UV3}$    & \nT\nC\nT  &   \nT\nG\nT   &  $\mu_{APOBEC}$ & \nA\nA\nC\nC  &   \nA\nA\nT\nC   &  $\mu_{AID}$ \\
-                \nT\nC\nG  &   \nT\nT\nG   &  $\mu_{mUV}$    & \nT\nG\nA  &   \nT\nA\nA   &  $\mu_{APOBEC}$ & \nA\nG\nC\nT  &   \nA\nG\nT\nT   &  $\mu_{AID}$ \\
-                \nC\nC\nG  &   \nC\nT\nG   &  $\mu_{mUV}$    & \nA\nG\nA  &   \nA\nA\nA   &  $\mu_{APOBEC}$ & \nA\nG\nC\nC  &   \nA\nG\nT\nC   &  $\mu_{AID}$ \\
-                \nC\nG\nA  &   \nC\nA\nA   &  $\mu_{mUV}$    & \nT\nG\nA  &   \nT\nC\nA   &  $\mu_{APOBEC}$ & \nT\nA\nC\nT  &   \nT\nA\nT\nT   &  $\mu_{AID}$ \\
-                \nC\nG\nG  &   \nC\nA\nG   &  $\mu_{mUV}$    & \nA\nC\nA  &   \nA\nC\nA   &  $\mu_{APOBEC}$ & \nT\nA\nC\nC  &   \nT\nA\nT\nC   &  $\mu_{AID}$ \\
-                \nT\nA     &   \nT\nG      &  $\mu_{\eta}$   & \nA\nA    &    \nA\nG      &  $\mu_{\iota1}$ & \nT\nG\nC\nT  &   \nT\nG\nT\nT   &  $\mu_{AID}$ \\
-                \nT\nA     &   \nC\nA      &  $\mu_{\eta}$   & \nT\nT    &    \nC\nT      &  $\mu_{\iota1}$ & \nT\nG\nC\nC  &   \nT\nG\nT\nC   &  $\mu_{AID}$ \\
-                \nA\nA     &   \nG\nT      &  $\mu_{\iota2}$ &           &                &                 & \nT\nA\nC\nT  &   \nT\nA\nT\nT   &  $\mu_{AID}$ \\
-                \nT\nT     &   \nA\nC      &  $\mu_{\iota2}$ &           &                &                 & \nT\nA\nC\nC  &   \nT\nA\nT\nC   &  $\mu_{AID}$ \\
-                           &               &                 &           &                &                 & \nT\nG\nC\nT  &   \nT\nG\nT\nT   &  $\mu_{AID}$ \\
-                           &               &                 &           &                &                 & \nT\nG\nC\nC  &   \nT\nG\nT\nC   &  $\mu_{AID}$ \\
+
+          \noindent
+      \begin{center}
+          \begin{tabular}{c@{\;\;$\to$\;\;}c@{\;\; {\small at rate}\;\; }c|c@{\;\;$\to$\;\;}c@{\;\; {\small at rate}\;\; }c|c@{\;\;$\to$\;\;}c@{\;\; {\small at rate}\;\; }c}
+            \nT\nC     &   \nT\nT      &  $\mu_{UV1}$    & \nC\nG       &   \nC\nT       &  $\mu_{g}$       & \nA\nA\nC\nT  &   \nA\nA\nT\nT   &  $\mu_{\AID}$ \\
+            \nG\nA     &   \nA\nA      &  $\mu_{UV1}$    & \nC\nG       &   \nA\nG       &  $\mu_{g}$       & \nA\nA\nC\nC  &   \nA\nA\nT\nC   &  $\mu_{\AID}$ \\
+            \nC\nC     &   \nC\nT      &  $\mu_{UV2}$    & \nT\nC\nA    &   \nT\nT\nA    &  $\mu_{\APOBEC}$ & \nA\nG\nC\nT  &   \nA\nG\nT\nT   &  $\mu_{\AID}$ \\
+            \nG\nG     &   \nA\nG      &  $\mu_{UV2}$    & \nT\nC\nT    &   \nT\nT\nT    &  $\mu_{\APOBEC}$ & \nA\nG\nC\nC  &   \nA\nG\nT\nC   &  $\mu_{\AID}$ \\
+            \nC\nC     &   \nT\nT      &  $\mu_{UV3}$    & \nT\nC\nA    &   \nT\nG\nA    &  $\mu_{\APOBEC}$ & \nA\nA\nC\nT  &   \nA\nA\nT\nT   &  $\mu_{\AID}$ \\
+            \nG\nG     &   \nA\nA      &  $\mu_{UV3}$    & \nT\nC\nT    &   \nT\nG\nT    &  $\mu_{\APOBEC}$ & \nA\nA\nC\nC  &   \nA\nA\nT\nC   &  $\mu_{\AID}$ \\
+            \nT\nC\nG  &   \nT\nT\nG   &  $\mu_{mUV}$    & \nT\nG\nA    &   \nT\nA\nA    &  $\mu_{\APOBEC}$ & \nA\nG\nC\nT  &   \nA\nG\nT\nT   &  $\mu_{\AID}$ \\
+            \nC\nC\nG  &   \nC\nT\nG   &  $\mu_{mUV}$    & \nA\nG\nA    &   \nA\nA\nA    &  $\mu_{\APOBEC}$ & \nA\nG\nC\nC  &   \nA\nG\nT\nC   &  $\mu_{\AID}$ \\
+            \nC\nG\nA  &   \nC\nA\nA   &  $\mu_{mUV}$    & \nT\nG\nA    &   \nT\nC\nA    &  $\mu_{\APOBEC}$ & \nT\nA\nC\nT  &   \nT\nA\nT\nT   &  $\mu_{\AID}$ \\
+            \nC\nG\nG  &   \nC\nA\nG   &  $\mu_{mUV}$    & \nA\nC\nA    &   \nA\nC\nA    &  $\mu_{\APOBEC}$ & \nT\nA\nC\nC  &   \nT\nA\nT\nC   &  $\mu_{\AID}$ \\
+            \nT\nA     &   \nT\nG      &  $\mu_{\eta}$   & \nA\nA       &    \nA\nG      &  $\mu_{\iota1}$  & \nT\nG\nC\nT  &   \nT\nG\nT\nT   &  $\mu_{\AID}$ \\
+            \nT\nA     &   \nC\nA      &  $\mu_{\eta}$   & \nT\nT       &    \nC\nT      &  $\mu_{\iota1}$  & \nT\nG\nC\nC  &   \nT\nG\nT\nC   &  $\mu_{\AID}$ \\
+            \nA\nA     &   \nG\nT      &  $\mu_{\iota2}$ & \nT\nG\nC\nT &  \nT\nG\nT\nT  &  $\mu_{\AID}$    & \nT\nA\nC\nT  &   \nT\nA\nT\nT   &  $\mu_{\AID}$ \\
+            \nT\nT     &   \nA\nC      &  $\mu_{\iota2}$ & \nT\nG\nC\nC &  \nT\nG\nT\nC  &  $\mu_{\AID}$    & \nT\nA\nC\nC  &   \nT\nA\nT\nC   &  $\mu_{\AID}$ 
         \end{tabular}
@@
 
+\begin{figure}
+    \begin{center} 
+        \includegraphics[width=\textwidth]{writeup-plots/biochem_mcmc_posteriors}
+    \end{center} 
+    \caption{
+        Estimated average mutation rates
+        for the motifs of Table \ref{tab:hominid_model}
+        since the common ancestor of humans and chimpanzee.
+        Rates are scaled so that the mean time to common ancestor is 0.001 units --
+        if we take this to be 6 million years \citep{scally2012insights}
+        then rates are in units of mutations per 6,000 years.
+        Shown are posterior medians (points) and 95\% credible intervals (lines, which are quite small in all cases).
+        Figure \ref{sfig:biochem_results_zoom} shows the smaller rates in more detail.
+        \label{fig:biochem_results}}
+\end{figure}
+
 %%%%%%%%%%%%%%%%%%%%
@@
         \label{fig:all_ising_coverage}}
+\end{figure}
+
+\begin{figure}
+    \begin{center} 
+        \includegraphics[width=\textwidth]{writeup-plots/biochem_mcmc_posteriors_zoomed}
+    \end{center} 
+    \caption{
+        As in Figure \ref{sfig:biochem_results}, but with reduce vertical axis range
+        to see in more detail:
+        estimated average mutation rates
+        for the motifs of Table \ref{tab:hominid_model}
+        since the common ancestor of humans and chimpanzee.
+        Rates are scaled so that the mean time to common ancestor is 0.001 units --
+        if we take this to be 6 million years \citep{scally2012insights}
+        then rates are in units of mutations per 6,000 years.
+        Shown are posterior medians (points) and 95\% credible intervals (lines, which are quite small in all cases).
+        \label{fig:biochem_results_zoom}}
 \end{figure}

### 2017-06-05 writeups/context-dependence.tex
writing

@@
 \newcommand{\nT}{\mbox{T}}
-\newcommand{\nW}{\mbox{W}}
-\newcommand{\nR}{\mbox{R}}
-\newcommand{\nY}{\mbox{Y}}
+\newcommand{\nS}{\mbox{S}} % C or G
+\newcommand{\nW}{\mbox{W}} % A or T
+\newcommand{\nR}{\mbox{R}} % purine, A or G
+\newcommand{\nY}{\mbox{Y}} % pyramidine, C or T
 \newcommand{\join}{\oplus}  % matches
@@
 selective constraints on the function of the sequence,
-and other processes such as biased gene conversion \citep{duret2009biased}.
+and other processes such as biased gene conversion \citep{duret2009biased,arbeithuber2015crossovers}.
 Indeed, molecular studies have demonstrated that the spectrum of new mutations in humans
@@
 In humans, the molecular processes underlying mutations are best studied in contexts of intense somatic mutation:
-cancer and the hypervariable regions that generate immune system diversity \citep{evolution_within},
+cancer and the hypervariable regions that generate immune system diversity \citep{cobey2015evolution,},
 which provide quite different cellular contexts but share commonalities due to shared molecular machinery.
 In general, substitution rates depend on sequence context due to the interaction of a great many factors including:
-(a) likelihood of DNA damage or errors in cell division (e.g., UV damage \citep{UV} or non-homologous end joining \citep{NHEJ});
+(a) likelihood of DNA damage or errors in cell division (e.g., UV damage \citep{goodman2002errorprone} or crossing over in meiosis \citep{arbeithuber2015crossovers});
 (b) activity levels and structural properties of the enzymes responsible for repair (e.g., base excision repar \citep{BER} and Y-family polymerases \citep{polymerase_review});
@@
 
-      \item[(repair)] In addition to the above, a collection of motifs associated with known biochemical causes
-          curated from the literature:
-          deamination due to repair of UV-induced pyramidine dimers ($\mu_{UV}$: (\nT/\nC)\nC $\to$ (\nT/\nC)\nT and \nC\nC$\to$\nT\nT, \citet{uv_damage});
-          an interaction between CpG and pyramidine dimers ($\mu_{mUV}$: \nT\nC\nG$\to$\nT\nT\nG and \nC\nC\nG$\to$\nC\nT\nG, \citet{cpg_review});
-          repair of guanine adducts at CpG sites ($\mu_g$: \nC\nG$\to$\nC\nT, \citet{tobacco});
-          deamination by AID ($\mu_{AID}$: \nW\nR\nC\nY$\to$\nW\nR\nT\nY, \nW is \nA or \nT; \nR is \nA or \nG; \nY is \nT or \nC, \citet{AID});
-          deamination by members of the APOBEC family ($\mu_{\APOBEC}$: \nT\nC\nW$\to$\nT\nT\nW or \nT\nC\nW$\to$\nT\nG\nW, \citet{APOBEC});
-          and repair by error-prone polymerases pol $\eta$ ($\mu_\eta$: \nT\nA$\to$\nT\nG, \citet{pol_eta});
-          and pol $\iota$ ($\mu_{\iota}$: \nA\nA$\to$\nG\nT and \nA\nA$\to$\nA\nG \citet{pol_iota}):
-
+      \item[(repair)] In addition to the above, we included a collection of mutation motifs associated with known DNA lesion/repair pathways
+          curated from the literature \citep[reviewed in][]{sale2012yfamily,goodman2013translesion,roberts2014hypermutation,cobey2015evolution,goodman2002errorprone}:
+          deamination due to repair of UV-induced pyramidine dimers \citep[$\mu_{UV}$][]{sale2012yfamily,sinha2002uvinduced};
+          repair of guanine adducts at CpG sites \citet[$\mu_g$][]{pfeifer2006mutagenesis};
+          deamination by AID \citep[$\mu_{AID}$][]{teng2007immunoglobulin,kasar2015wholegenome};
+          deamination by members of the APOBEC family \citep[$\mu_\APOBEC$][]{alexandrov2013deciphering};
+          and repair by error-prone polymerases pol $\eta$ \citep[$\mu_\eta$][]{alexandrov2013deciphering};
+          and pol $\iota$ \citep[$\mu_{\iota}$][]{roberts2014hypermutation,maul2016polymerase}.
+          These are denoted here using ambiguity codes 
+          \nR{}: \nA{} or \nG{},
+          \nY{}: \nC{} or \nT{},
+          \nW{}: \nA{} or \nT{}, and
+          \nS{}: \nC{} or \nG{};
+          so for instance 
+          ``\nT\nC\nW$\to$\nT\nT\nW'' 
+          means that both 
+          \nT\nC\nA$\to$\nT\nT\nA
+          and
+          \nT\nC\nT$\to$\nT\nT\nT
+          share the same mutation rate.
           \noindent
       \begin{center}
-          \begin{tabular}{c@{\;\;$\to$\;\;}c@{\;\; {\small at rate}\;\; }c|c@{\;\;$\to$\;\;}c@{\;\; {\small at rate}\;\; }c|c@{\;\;$\to$\;\;}c@{\;\; {\small at rate}\;\; }c}
-            \nT\nC     &   \nT\nT      &  $\mu_{UV1}$    & \nC\nG       &   \nC\nT       &  $\mu_{g}$       & \nA\nA\nC\nT  &   \nA\nA\nT\nT   &  $\mu_{\AID}$ \\
-            \nG\nA     &   \nA\nA      &  $\mu_{UV1}$    & \nC\nG       &   \nA\nG       &  $\mu_{g}$       & \nA\nA\nC\nC  &   \nA\nA\nT\nC   &  $\mu_{\AID}$ \\
-            \nC\nC     &   \nC\nT      &  $\mu_{UV2}$    & \nT\nC\nA    &   \nT\nT\nA    &  $\mu_{\APOBEC}$ & \nA\nG\nC\nT  &   \nA\nG\nT\nT   &  $\mu_{\AID}$ \\
-            \nG\nG     &   \nA\nG      &  $\mu_{UV2}$    & \nT\nC\nT    &   \nT\nT\nT    &  $\mu_{\APOBEC}$ & \nA\nG\nC\nC  &   \nA\nG\nT\nC   &  $\mu_{\AID}$ \\
-            \nC\nC     &   \nT\nT      &  $\mu_{UV3}$    & \nT\nC\nA    &   \nT\nG\nA    &  $\mu_{\APOBEC}$ & \nA\nA\nC\nT  &   \nA\nA\nT\nT   &  $\mu_{\AID}$ \\
-            \nG\nG     &   \nA\nA      &  $\mu_{UV3}$    & \nT\nC\nT    &   \nT\nG\nT    &  $\mu_{\APOBEC}$ & \nA\nA\nC\nC  &   \nA\nA\nT\nC   &  $\mu_{\AID}$ \\
-            \nT\nC\nG  &   \nT\nT\nG   &  $\mu_{mUV}$    & \nT\nG\nA    &   \nT\nA\nA    &  $\mu_{\APOBEC}$ & \nA\nG\nC\nT  &   \nA\nG\nT\nT   &  $\mu_{\AID}$ \\
-            \nC\nC\nG  &   \nC\nT\nG   &  $\mu_{mUV}$    & \nA\nG\nA    &   \nA\nA\nA    &  $\mu_{\APOBEC}$ & \nA\nG\nC\nC  &   \nA\nG\nT\nC   &  $\mu_{\AID}$ \\
-            \nC\nG\nA  &   \nC\nA\nA   &  $\mu_{mUV}$    & \nT\nG\nA    &   \nT\nC\nA    &  $\mu_{\APOBEC}$ & \nT\nA\nC\nT  &   \nT\nA\nT\nT   &  $\mu_{\AID}$ \\
-            \nC\nG\nG  &   \nC\nA\nG   &  $\mu_{mUV}$    & \nA\nC\nA    &   \nA\nC\nA    &  $\mu_{\APOBEC}$ & \nT\nA\nC\nC  &   \nT\nA\nT\nC   &  $\mu_{\AID}$ \\
-            \nT\nA     &   \nT\nG      &  $\mu_{\eta}$   & \nA\nA       &    \nA\nG      &  $\mu_{\iota1}$  & \nT\nG\nC\nT  &   \nT\nG\nT\nT   &  $\mu_{\AID}$ \\
-            \nT\nA     &   \nC\nA      &  $\mu_{\eta}$   & \nT\nT       &    \nC\nT      &  $\mu_{\iota1}$  & \nT\nG\nC\nC  &   \nT\nG\nT\nC   &  $\mu_{\AID}$ \\
-            \nA\nA     &   \nG\nT      &  $\mu_{\iota2}$ & \nT\nG\nC\nT &  \nT\nG\nT\nT  &  $\mu_{\AID}$    & \nT\nA\nC\nT  &   \nT\nA\nT\nT   &  $\mu_{\AID}$ \\
-            \nT\nT     &   \nA\nC      &  $\mu_{\iota2}$ & \nT\nG\nC\nC &  \nT\nG\nT\nC  &  $\mu_{\AID}$    & \nT\nA\nC\nC  &   \nT\nA\nT\nC   &  $\mu_{\AID}$ 
+          \begin{tabular}{c@{\;\;$\to$\;\;}c@{\;\; {\small at rate}\;\; }c|c@{\;\;$\to$\;\;}c@{\;\; {\small at rate}\;\; }c}
+            \nC\nG     &   \nC\nT      &  $\mu_{g}$      & \nA\nA       &   \nA\nG       &  $\mu_{\iota1}$   \\
+            \nC\nG     &   \nA\nG      &  $\mu_{g}$      & \nT\nT       &   \nC\nT       &  $\mu_{\iota1}$   \\
+            \nT\nC     &   \nT\nT      &  $\mu_{UV}$     & \nA\nA       &   \nG\nT       &  $\mu_{\iota2}$   \\
+            \nG\nA     &   \nA\nA      &  $\mu_{UV}$     & \nT\nT       &   \nA\nC       &  $\mu_{\iota2}$   \\
+            \nC\nC     &   \nC\nT      &  $\mu_{UV}$     & \nT\nC\nW    &   \nT\nT\nW    &  $\mu_{\APOBEC}$  \\
+            \nG\nG     &   \nA\nG      &  $\mu_{UV}$     & \nT\nC\nW    &   \nT\nG\nW    &  $\mu_{\APOBEC}$  \\
+            \nC\nC     &   \nT\nT      &  $\mu_{UVd}$    & \nS\nG\nA    &   \nS\nA\nA    &  $\mu_{\APOBEC}$  \\
+            \nG\nG     &   \nA\nA      &  $\mu_{UVd}$    & \nS\nG\nA    &   \nS\nC\nA    &  $\mu_{\APOBEC}$  \\
+            \nT\nA     &   \nT\nG      &  $\mu_{\eta}$   & \nW\nR\nC\nY &  \nW\nR\nT\nY  &  $\mu_{\AID}$     \\
+            \nT\nA     &   \nC\nA      &  $\mu_{\eta}$   & \nR\nG\nY\nS &  \nR\nA\nY\nS  &  $\mu_{\AID}$     
         \end{tabular}
@@
 
-% [ [ "TC", "TT" ], [ "GA", "AA" ] ], UV1
-% [ [ "CC", "CT" ], [ "GG", "AG" ] ], UV2
-% [ [ "CC", "TT" ], [ "GG", "AA" ] ], UV3
-% [ [ "TCG", "TTG" ], [ "CCG", "CTG" ], [ "CGA", "CAA" ], [ "CGG", "CAG" ] ], mUV
-% [ [ "CG", "CT" ], [ "CG", "AG" ] ],
-% [ [ "AACT", "AATT" ], [ "AACC", "AATC" ], [ "AGCT", "AGTT" ], [ "AGCC", "AGTC" ], [ "TACT", "TATT" ], [ "TACC", "TATC" ], [ "TGCT", "TGTT" ], [ "TGCC", "TGTC" ],
-%   [ "AGTT", "AATT" ], [ "GGTT", "GATT" ], [ "AGCT", "AACT" ], [ "GGCT", "GACT" ], [ "AGTA", "AATA" ], [ "GGTA", "GATA" ], [ "AGCA", "AACA" ], [ "GGCA", "GACA" ] ],
-% [ [ "TCA", "TTA" ], [ "TCT", "TTT" ], [ "TCA", "TGA" ], [ "TCT", "TGT" ],
-%   [ "TGA", "TAA" ], [ "AGA", "AAA" ], [ "TGA", "TCA" ], [ "AGA", "ACA" ] ],
-% [ [ "TA", "TG" ], [ "TA", "CA" ] ],
-% [ [ "AA", "GT" ], [ "TT", "AC" ] ],
-% [ [ "AA", "AG" ], [ "CT", "TT" ] ]
-
 \end{itemize}
+
+The mutational motifs included above are far from complete 
+-- context-specific affinities and error rates \textit{in vivo} 
+are not known for most DNA repair pathways --
+but neither do they exhaust the hypotheses available in the literature.
+For instance, many enzymes act more than once locally when they act,
+such as AID \citep{senavirathne2015activationinduced} or error-prone polymerases involved in repair \citep{maul2016polymerase}.
 
@@
 \section{Results}
-
-\textbf{To-do:}
-
-3. Analyze human data with shape model, Bayesian.
-
-The method works well to find point estimates of the parameters.
-Obtaining posterior density estimates using the approximation likelihood function on a subset of patterns
-works in some cases;
-in other cases the likelihood surface is still too peaked, underestimating the uncertainty.
-(but counting in nonoverlapping windows fixes the peakiness problem)
 
@@
         Rates are scaled so that the mean time to common ancestor is 0.001 units --
-        if we take this to be 6 million years \citep{scally2012insights}
+        if we take this to be 6 million years \citep{scally2012insights,langergraber2012generation}
         then rates are in units of mutations per 6,000 years.
@@
 \end{figure}
+
+It may seem tenous to look for mutational signatures associated with cancer in the germline,
+or absurd that UV damage could play a significant role.
+This is in part due to the ``streetlight effect'' --
+we can't look for signatures that are not characterized --
+but it is also reasonable that shared repair machinery might induce similar context-dependent errors.
+More likely sources of error in the germline include the effects of recombination 
+\citep{arbeithuber2015crossovers,myers2010drive}.
 
@@
 
+\subsection{Longer models}
+      \begin{center}
+          \begin{tabular}{c@{\;\;$\to$\;\;}c@{\;\; {\small at rate}\;\; }c|c@{\;\;$\to$\;\;}c@{\;\; {\small at rate}\;\; }c|c@{\;\;$\to$\;\;}c@{\;\; {\small at rate}\;\; }c}
+            \nT\nC     &   \nT\nT      &  $\mu_{UV1}$    & \nC\nG       &   \nC\nT       &  $\mu_{g}$       & \nA\nA\nC\nT  &   \nA\nA\nT\nT   &  $\mu_{\AID}$ \\
+            \nG\nA     &   \nA\nA      &  $\mu_{UV1}$    & \nC\nG       &   \nA\nG       &  $\mu_{g}$       & \nA\nA\nC\nC  &   \nA\nA\nT\nC   &  $\mu_{\AID}$ \\
+            \nC\nC     &   \nC\nT      &  $\mu_{UV2}$    & \nT\nC\nA    &   \nT\nT\nA    &  $\mu_{\APOBEC}$ & \nA\nG\nC\nT  &   \nA\nG\nT\nT   &  $\mu_{\AID}$ \\
+            \nG\nG     &   \nA\nG      &  $\mu_{UV2}$    & \nT\nC\nT    &   \nT\nT\nT    &  $\mu_{\APOBEC}$ & \nA\nG\nC\nC  &   \nA\nG\nT\nC   &  $\mu_{\AID}$ \\
+            \nC\nC     &   \nT\nT      &  $\mu_{UV3}$    & \nT\nC\nA    &   \nT\nG\nA    &  $\mu_{\APOBEC}$ & \nA\nA\nC\nT  &   \nA\nA\nT\nT   &  $\mu_{\AID}$ \\
+            \nG\nG     &   \nA\nA      &  $\mu_{UV3}$    & \nT\nC\nT    &   \nT\nG\nT    &  $\mu_{\APOBEC}$ & \nA\nA\nC\nC  &   \nA\nA\nT\nC   &  $\mu_{\AID}$ \\
+            \nT\nC\nG  &   \nT\nT\nG   &  $\mu_{mUV}$    & \nT\nG\nA    &   \nT\nA\nA    &  $\mu_{\APOBEC}$ & \nA\nG\nC\nT  &   \nA\nG\nT\nT   &  $\mu_{\AID}$ \\
+            \nC\nC\nG  &   \nC\nT\nG   &  $\mu_{mUV}$    & \nA\nG\nA    &   \nA\nA\nA    &  $\mu_{\APOBEC}$ & \nA\nG\nC\nC  &   \nA\nG\nT\nC   &  $\mu_{\AID}$ \\
+            \nC\nG\nA  &   \nC\nA\nA   &  $\mu_{mUV}$    & \nT\nG\nA    &   \nT\nC\nA    &  $\mu_{\APOBEC}$ & \nT\nA\nC\nT  &   \nT\nA\nT\nT   &  $\mu_{\AID}$ \\
+            \nC\nG\nG  &   \nC\nA\nG   &  $\mu_{mUV}$    & \nA\nC\nA    &   \nA\nC\nA    &  $\mu_{\APOBEC}$ & \nT\nA\nC\nC  &   \nT\nA\nT\nC   &  $\mu_{\AID}$ \\
+            \nT\nA     &   \nT\nG      &  $\mu_{\eta}$   & \nA\nA       &    \nA\nG      &  $\mu_{\iota1}$  & \nT\nG\nC\nT  &   \nT\nG\nT\nT   &  $\mu_{\AID}$ \\
+            \nT\nA     &   \nC\nA      &  $\mu_{\eta}$   & \nT\nT       &    \nC\nT      &  $\mu_{\iota1}$  & \nT\nG\nC\nC  &   \nT\nG\nT\nC   &  $\mu_{\AID}$ \\
+            \nA\nA     &   \nG\nT      &  $\mu_{\iota2}$ & \nT\nG\nC\nT &  \nT\nG\nT\nT  &  $\mu_{\AID}$    & \nT\nA\nC\nT  &   \nT\nA\nT\nT   &  $\mu_{\AID}$ \\
+            \nT\nT     &   \nA\nC      &  $\mu_{\iota2}$ & \nT\nG\nC\nC &  \nT\nG\nT\nC  &  $\mu_{\AID}$    & \nT\nA\nC\nC  &   \nT\nA\nT\nC   &  $\mu_{\AID}$ 
+        \end{tabular}
+      \end{center}
+
 \end{document}

### 2017-06-06 writeups/context-dependence.tex
got all references in

@@
 In humans, the molecular processes underlying mutations are best studied in contexts of intense somatic mutation:
-cancer and the hypervariable regions that generate immune system diversity \citep{cobey2015evolution,},
+cancer and the hypervariable regions that generate immune system diversity \citep{cobey2015evolution},
 which provide quite different cellular contexts but share commonalities due to shared molecular machinery.
@@
 (a) likelihood of DNA damage or errors in cell division (e.g., UV damage \citep{goodman2002errorprone} or crossing over in meiosis \citep{arbeithuber2015crossovers});
-(b) activity levels and structural properties of the enzymes responsible for repair (e.g., base excision repar \citep{BER} and Y-family polymerases \citep{polymerase_review});
+(b) activity levels and structural properties of the enzymes responsible for repair 
+(e.g., Y-family polymerases \citep{goodman2013translesion,sale2012yfamily});
 and (c) any functional effects of an error (e.g., lethal mutations will not be seen in a population census).
@@
 and so comparisons between widely separated species will identify time-averaged mutation rates.
-Recent studies have used nonnegative matrix factorization to identify sets of mutational patterns
-co-occurring in comparisons (a) between different cancer types \citep{cancer_scan} and (b) between different human populations \citep{harris,matthieson},
+Recent studies have used a variety of exploratory techniques to identify sets of mutational patterns
+co-occurring in comparisons (a) between different cancer types \citep{alexandrov2013signatures,shiraishi2015simple} and 
+(b) between different human populations \citep{harris2015evidence,harris2016rapid,mathieson2017differences},
 to disentangle distinct signals presumably coming from the action of distinct sources of mutation.
@@
 
-    Counteracting this trend towards increased G/C is GC-biased gene conversion \citep{gcbiased},
+    Counteracting this trend towards increased G/C is GC-biased gene conversion \citep{glemin2015quantification},
     which is effectively a selective pressure against G or C bases.
@@
     in a population of effective size $N$ becomes ubiquitous, rather than dying out,
-    with probability approximately $(1-\exp(-2 s))/(1-\exp(-2 s N))$ \citep{fixation_prob}.
+    with probability approximately $(1-\exp(-2 s))/(1-\exp(-2 s N))$ \citep{kimura1962probability}.
     If the mutation occurs at rate $\mu$ per individual, the total rate it occurs at in the population is $\mu N$.
@@
 avoiding the additional constraints of coding sequences.
-To do this, we extracted regions in the human (hg38)--chimpanzee (panTro4) alignment \citep[from the UCSC Genome Browser][]{pantroalign}
-according to four types of regulatory feature in the Ensembl Regulatory Build \citep[release 81][]{ensebml}:
+To do this, we extracted regions in the human (hg38)--chimpanzee (panTro4) alignment % \citep[from the UCSC Genome Browser][]{pantroalign}
+according to four types of regulatory feature in the Ensembl Regulatory Build \citep[release 81][]{zerbino2015ensembl}:
 ``CTCF binding site'', ``enhancer'', ``open chromatin region'', and ``promoter flanking region''.
 These were furthermore divided into regions that either did or did not overlap a transcription start or end site of a known gene
-from the hg38 Annotations Database \citep{hg_annotationsdb}
+from the hg38 Annotations Database \citep{kent2002human}
 and filtered for length:
@@
 we simulated a single sequence of $10^6$ nucleotides
-for one units of time
+for one unit of time
 under the model given in Table \ref{tab:cpg_results},
 obtained by embellishing the basic CpG hypermutability
-with the longer mutational motif identified by \citet{harris}.
+with the longer mutational motif identified by \citet{harris2015evidence}.
 %     \begin{center}
@@
 counted $(5,3,1)$ T-mers,
-and used the mcmc package \citep{mcmcR} to run a random walk sampler for 10,000 steps
+and used the mcmc package \citep{geyer2017mcmc} to run a random walk sampler for 10,000 steps
 (judged sufficient by examination of diagnostic plots).
@@
 The methods described here provide a way to efficiently calculate full likelihoods,
-which in many circumstances provide the optimal basis for inference \citep{neymanpearson},
+which in many circumstances provide the optimal basis for inference \citep{neyman1933problem},
 fast enough to allow Bayesian inference from Monte Carlo sampling.
@@
 however, both assumptions are also made the state of the art in phylogenetics.
-Some relaxation of the assumptions is possible -- like in phylogenetics \citep{ratemixtures},
-we can easily allow heterogeneity between sites by scaling rates by an overall gamma-distributed random factor.
+Some relaxation of the assumptions is possible -- for instnace, 
+we can easily allow heterogeneity between sites by scaling rates by an overall gamma-distributed random factor \citep{yang1994maximum}.
 Short deletions might be accomodated by introducing a ``gap'' letter to the alphabet,
@@
 The large number of possible mutation motifs also makes multiple testing an issue;
-an unguided approach might include all possible motifs but place a shrinkage prior on their values \citep{shrinkage,horseshoe}.
+an unguided approach might include all possible motifs but place a shrinkage prior on their values \citep{bhadra2015horseshoe}.
 The converse approach of adding motifs that are overrepresented in the residuals
@@
 then this framework could be used to rigorously estimate their average usages and impacts.
-Despite substantial progress over the past 15 years \citep{polymerase_reviews},
+Despite substantial progress over the past two decades \citep{goodman2013translesion,sale2012yfamily},
 this is still far from known.
 A promising alternative would be to use the distinct signatures obtained by dimension reduction techniques as in
-\citet{cancer_scan}, \citep{matthieson} or \citep{pritchard} as a proxy for unknown but distinct pathways.
+\citet{alexandrov2013signatures}, \citep{mathieson2017differences} or \citep{shiraishi2015simple} as a proxy for unknown but distinct pathways.
+
+More things to do: cancer; include gaps; more taxa; leading/lagging strand...
 

### 2017-06-06 writeups/context-dependence.tex
relative rates

@@
         Shown are posterior medians (points) and 95\% credible intervals (lines, which are quite small in all cases).
-        Figure \ref{sfig:biochem_results_zoom} shows the smaller rates in more detail.
+        Figure \ref{sfig:biochem_results_zoom} shows the smaller rates in more detail,
+        and figure \ref{sfig:biochem_results_relative} shows the rates between categories relative to each other.
         \label{fig:biochem_results}}
@@
     \caption{
-        As in Figure \ref{sfig:biochem_results}, but with reduce vertical axis range
+        As in Figure \ref{fig:biochem_results}, but with reduce vertical axis range
         to see in more detail:
@@
         Shown are posterior medians (points) and 95\% credible intervals (lines, which are quite small in all cases).
-        \label{fig:biochem_results_zoom}}
+        \label{sfig:biochem_results_zoom}}
+\end{figure}
+
+\begin{figure}
+    \begin{center} 
+        \includegraphics[width=\textwidth]{writeup-plots/biochem_mcmc_posteriors_relative}
+    \end{center} 
+    \caption{
+        As in Figures \ref{fig:biochem_results} and \ref{sfig:biochem_results_zoom}, 
+        but where each rate is divided by the mean rate across all eight categories
+        (as well as the 95\% credible interval endpoints).
+        \label{sfig:biochem_results_relative}}
 \end{figure}

### 2021-09-23 writeups/context-dependence.tex
minor edits found hanging around uncommitted

@@
 
+% Model-based inference neighborhood dependence
+% Nearby sites affecting the likelihood
+% Rigorous inference with neighborhood dependency
+% Statistical inference with local/neighborhood interdependency
+% It depends on the context: 
+%
+
 \title{Inference using context-dependent models of mutation}
@@
 \begin{abstract}
+
 \end{abstract}
@@
         Single-base changes, with separate rates for transitions ($\mu_v$),
-        \nA$\leftrightarrow$\nT ($\mu_{A}T$), \nC$\leftrightarrow$\nG ($\mu_{CG}$),
+        \nA$\leftrightarrow$\nT{} ($\mu_{AT}$), \nC$\leftrightarrow$\nG{} ($\mu_{CG}$),
         and other transversions ($\mu_{r}$):
@@
 
-      \item[(CpG)] In addition to the above, an addition CpG rate:
+      \item[(CpG)] In addition to the above, an additional CpG rate:
           \begin{center}
@@
           curated from the literature \citep[reviewed in][]{sale2012yfamily,goodman2013translesion,roberts2014hypermutation,cobey2015evolution,goodman2002errorprone}:
-          deamination due to repair of UV-induced pyramidine dimers \citep[$\mu_{UV}$][]{sale2012yfamily,sinha2002uvinduced};
-          repair of guanine adducts at CpG sites \citet[$\mu_g$][]{pfeifer2006mutagenesis};
-          deamination by AID \citep[$\mu_{AID}$][]{teng2007immunoglobulin,kasar2015wholegenome};
-          deamination by members of the APOBEC family \citep[$\mu_\APOBEC$][]{alexandrov2013deciphering};
-          and repair by error-prone polymerases pol $\eta$ \citep[$\mu_\eta$][]{alexandrov2013deciphering};
-          and pol $\iota$ \citep[$\mu_{\iota}$][]{roberts2014hypermutation,maul2016polymerase}.
+          deamination due to repair of UV-induced pyramidine dimers \citep[$\mu_{UV}$,][]{sale2012yfamily,sinha2002uvinduced};
+          repair of guanine adducts at CpG sites \citet[$\mu_g$,][]{pfeifer2006mutagenesis};
+          deamination by AID \citep[$\mu_{AID}$,][]{teng2007immunoglobulin,kasar2015wholegenome};
+          deamination by members of the APOBEC family \citep[$\mu_\APOBEC$,][]{alexandrov2013deciphering};
+          and repair by error-prone polymerases pol $\eta$ \citep[$\mu_\eta$,][]{alexandrov2013deciphering};
+          and pol $\iota$ \citep[$\mu_{\iota}$,][]{roberts2014hypermutation,maul2016polymerase}.
           These are denoted here using ambiguity codes 
@@
           means that both 
-          \nT\nC\nA$\to$\nT\nT\nA
+          \nT\nC\nA$\to$\nT\nT\nA{}
           and
-          \nT\nC\nT$\to$\nT\nT\nT
+          \nT\nC\nT$\to$\nT\nT\nT{}
           share the same mutation rate.
@@
                 \hline
-                \nC\nG  & \_\nC   & 100410   &  120303    & -19892.50   &	-40.55  \\
-                \nC\nA  & \_\nT   &  19884   &   28350    &  -8466.43   &	-35.55  \\
-                \nC\nT  & \_\nT   &  19998   &   28061    &  -8063.46   &	-34.03  \\
-                \nC\nC  & \_\nT   &  23673   &   28177    &  -4503.96   &	-18.97  \\
-                \nG\nG  & \_\nC   &  17742   &   19418    &  -1675.92   &	 -8.50  \\
-                \nC\nG  & \_\nG   &  18450   &   19478    &  -1028.34   &	 -5.21  \\
-                \nA\nG  & \_\nC   &  10968   &   11760    &   -791.73   &	 -5.16  \\
-                \nG\nG  & \_\nG   & 118805   &  120975    &  -2170.25   &    \\
+                \nC\nG  & \nC\_   & 100410   &  120303    & -19892.50   &	-40.55  \\
+                \nC\nA  & \nT\_   &  19884   &   28350    &  -8466.43   &	-35.55  \\
+                \nC\nT  & \nT\_   &  19998   &   28061    &  -8063.46   &	-34.03  \\
+                \nC\nC  & \nT\_   &  23673   &   28177    &  -4503.96   &	-18.97  \\
+                \nG\nG  & \nC\_   &  17742   &   19418    &  -1675.92   &	 -8.50  \\
+                \nC\nG  & \nG\_   &  18450   &   19478    &  -1028.34   &	 -5.21  \\
+                \nA\nG  & \nC\_   &  10968   &   11760    &   -791.73   &	 -5.16  \\
+                \nG\nG  & \nG\_   & 118805   &  120975    &  -2170.25   &    \\
                 \hline         
-                \nA\nG  & \_\nT   &  17280   &   16515    &    765.20   &  4.21   \\
-                \nG\nT  & \_\nC   &  20490   &   19507    &    983.44   &  4.97   \\
-                \nG\nG  & \_\nT   &  20929   &   19278    &   1650.83   &  8.40   \\
-                \nC\nC  & \_\nC   & 124149   &  119880    &   4268.52   &  8.71   \\
-                \nG\nG  & \_\nA   &  30483   &   28288    &   2195.34   &  9.22   \\
-                \nC\nT  & \_\nC   & 126957   &  119389    &   7567.92   & 15.48   \\
-                \nC\nA  & \_\nC   & 128673   &  120619    &   8054.46   & 16.39   \\
-                \nC\nG  & \_\nT   &  49311   &   28276    &  21034.85   & 88.45   \\
+                \nA\nG  & \nT\_   &  17280   &   16515    &    765.20   &  4.21   \\
+                \nG\nT  & \nC\_   &  20490   &   19507    &    983.44   &  4.97   \\
+                \nG\nG  & \nT\_   &  20929   &   19278    &   1650.83   &  8.40   \\
+                \nC\nC  & \nC\_   & 124149   &  119880    &   4268.52   &  8.71   \\
+                \nG\nG  & \nA\_   &  30483   &   28288    &   2195.34   &  9.22   \\
+                \nC\nT  & \nC\_   & 126957   &  119389    &   7567.92   & 15.48   \\
+                \nC\nA  & \nC\_   & 128673   &  120619    &   8054.46   & 16.39   \\
+                \nC\nG  & \nT\_   &  49311   &   28276    &  21034.85   & 88.45   \\
                 \hline
                 \hline
-                \nC\nG  &  \nG\_  &  100803  &  120541  &  -19737.80  &  -40.19 \\
-                \nT\nG  &  \nA\_  &   20040  &   28179  &   -8139.45  &  -34.28 \\
-                \nG\nG  &  \nA\_  &   20985  &   28288  &   -7302.96  &  -30.70 \\
-                \nA\nG  &  \nA\_  &   22929  &   28399  &   -5469.73  &  -22.95 \\
-                \nC\nC  &  \nG\_  &   17805  &   19410  &   -1605.01  &   -8.14 \\
-                \nC\nT  &  \nG\_  &   10470  &   11684  &   -1214.31  &   -7.94 \\
-                \nC\nA  &  \nG\_  &   10839  &   11882  &   -1043.39  &   -6.76 \\
-                \nC\nG  &  \nC\_  &   18354  &   19348  &    -994.19  &   -5.05 \\
+                \nC\nG  &  \_\nG  &  100803  &  120541  &  -19737.80  &  -40.19 \\
+                \nT\nG  &  \_\nA  &   20040  &   28179  &   -8139.45  &  -34.28 \\
+                \nG\nG  &  \_\nA  &   20985  &   28288  &   -7302.96  &  -30.70 \\
+                \nA\nG  &  \_\nA  &   22929  &   28399  &   -5469.73  &  -22.95 \\
+                \nC\nC  &  \_\nG  &   17805  &   19410  &   -1605.01  &   -8.14 \\
+                \nC\nT  &  \_\nG  &   10470  &   11684  &   -1214.31  &   -7.94 \\
+                \nC\nA  &  \_\nG  &   10839  &   11882  &   -1043.39  &   -6.76 \\
+                \nC\nG  &  \_\nC  &   18354  &   19348  &    -994.19  &   -5.05 \\
                 \hline       
-                \nC\nT  &  \nA\_  &   17007  &   16078  &     929.33  &    5.18 \\
-                \nT\nC  &  \nG\_  &   20691  &   19509  &    1182.40  &    5.98 \\
-                \nC\nC  &  \nA\_  &   20640  &   19160  &    1480.44  &    7.56 \\
-                \nT\nC  &  \nT\_  &   30309  &   28320  &    1988.91  &    8.35 \\
-                \nA\nG  &  \nG\_  &  127191  &  121450  &    5740.76  &   11.64 \\
-                \nG\nG  &  \nG\_  &  127946  &  120977  &    6969.46  &   14.16 \\
-                \nT\nG  &  \nG\_  &  127539  &  120512  &    7026.52  &   14.31 \\
-                \nC\nG  &  \nA\_  &   49098  &   28186  &   20911.93  &   88.07 \\
+                \nC\nT  &  \_\nA  &   17007  &   16078  &     929.33  &    5.18 \\
+                \nT\nC  &  \_\nG  &   20691  &   19509  &    1182.40  &    5.98 \\
+                \nC\nC  &  \_\nA  &   20640  &   19160  &    1480.44  &    7.56 \\
+                \nT\nC  &  \_\nT  &   30309  &   28320  &    1988.91  &    8.35 \\
+                \nA\nG  &  \_\nG  &  127191  &  121450  &    5740.76  &   11.64 \\
+                \nG\nG  &  \_\nG  &  127946  &  120977  &    6969.46  &   14.16 \\
+                \nT\nG  &  \_\nG  &  127539  &  120512  &    7026.52  &   14.31 \\
+                \nC\nG  &  \_\nA  &   49098  &   28186  &   20911.93  &   88.07 \\
                 \hline

### 2021-10-05 writeups/context-dependence.tex
Mostly trimming whitespace and Makefile additions

@@
 
-It not uncommon to encounter models
+It now common to encounter models
 consisting of a lattice of sites,

### 2021-10-05 writeups/context-dependence.tex
halfway through

@@
 \peter{something about efficient computation and sparse matrices}
+
+%EM It feels like we go back into introduction mode here after the "In this paper" paragraph.
 
@@
 \citet{pedersen2000dependent}, later extended by
-\citet{hobolth2008markov,baele2010using}, used data augmentation and an MCMC algorithm to do Bayesian inference;
+\citet{Hwang2004-pj,hobolth2008markov,baele2010using}, used data augmentation and an MCMC algorithm to do Bayesian inference;
 \citet{lunter2004nucleotide} used an approximate matrix decomposition;
@@
   This only has one parameter, the speed.
+  %EM I can't actually tell if what is developed below addresses this difficulty!
   If we allow particles to exit from the right, and new particles to enter from the left
@@
 
+%EM I think we could have both, but certainly this biological interpretation. Then we would have "selection" and "fitness" rather than "potential" and "energy"?
 \peter{Should call this ``selection'' instead of a ``potential''? Or what?}
@@
 where $n(x,p_i)$ is the number of times the pattern $p_i$ occurs in $x$.
+%EM Can we introduce the phi notation here, and say something a little more formal? Namely that if the mutation rate is mu then the actual rate of change is mu times phi of the difference of energies?
 
@@
     where $\gamma$ is the additional CpG rate above the base mutation rate.
+    %EM This "more than one way" isn't as clear as "more than one pattern may match" used above, IMHO. Because it's the same underlying substitutions, just different patterns getting us there, right?
     Note that for a given pair of sequences, there may be more than one way to mutate a given string $x$ to get $y$:
@@
     and reconfigures to a state chosen with probability proportional to the stationary probability of the resulting configuration:
-    Since $\exp(-E(x))/(\exp(-E(x))+\exp(-E(y))) = 1/(1+exp(-(E(x)-E(y)))$,
+    Since $\exp(-E(x))/(\exp(-E(x))+\exp(-E(y))) = 1/(1+\exp(-(E(x)-E(y)))$,
     this is:
@@
 also used for simulation,
+%EM My understanding is that this is also called uniformization http://dx.doi.org/10.1214/09-AOAS247
 using a variant of the ``jump chain representation'', also known as the ``Gillespie algorithm''.
@@
 \[
+%EM Here we're moving to lower-j.
 \mu_* = \max \left\{ \sum_{y \in \calS^n} \sum_{j\in J(i,x,y)} \mu_j \qquad \st x \in \calS^n, \quad 1 \le i \le n \right\} .
@@
 Note that since the process is homogeneous, the maximum rate of changes occurring at any position will be the same, except possibly near the end of the sequence,
+%EM I wonder if we could do something like introduce J(x), which is the set of all possible transitions starting at x, so we don't have to do this "unnecessary" thing. Or perhaps it's fine as is.
 so the maximization over $i$ is unnecessary.
@@
 with probability proportional to the corresponding rate, i.e.\
-\erick{Previously, the $j$ in $\mu^j$ was used to index the possible patterns. Now it's indexing possible changes at a site.}
+%EM Not sure what I was on about here.
+% \erick{Previously, the $j$ in $\mu^j$ was used to index the possible patterns. Now it's indexing possible changes at a site.}
 \[
@@
 
-For instance, to simulate TASEP, one has only to let $\mu_*=\lambda$, and at each $c_{ij}$
+For instance, to simulate the TASEP, one has only to let $\mu_*=\lambda$, and at each $c_{ij}$
 check if at site $i$ there is a 1 followed by a 0,
@@
 
+%EM I don't see this needing shortening!
 \peter{shorten, refer to figure}
@@
 
+%EM I feel like it needs emphasis that this is different than saying that we assume that the rates stay the same as in the initial sequence.
+
 %%%%%%
@@
 by discarding approximately independent sites from the conditioning.
+%EM The following sentence may get us in trouble with the formalization police.
 If $\ell$ is long enough that approximation \eqref{eqn:window_approx} holds, then:
+%EM This casual reader is a little confused about the left to right nature here.
 \begin{align} \label{eqn:full_approx}

### 2021-10-05 writeups/context-dependence.tex
More comments

@@
 
-As noted above, there may be more than one way to mutate a sequence $x$ to get $y$:
+As noted above, there may be more than one way to mutate a such a sequence $x$ to get such a $y$:
 for each position $i$, let $J(i,x,y)$ be those transitions in $\calT$ that can be applied at $i$
@@
         and the third site in the final sequence (bottom).
+        %EM This "window" terminology... should we have another quick definition here?
         The maximum window width, $w$, is equal to 1.
@@
 we can compute \emph{something} correctly,
+%EM The full exact likelihood? Isn't it an approximation?
 and that something will be sufficient to compute the full likelihood.
 To do this, let $p_{n,\ell,r}(x,y)$ for $r+\ell < n$ denote
+%EM Can we say initial subsequence or something?
 the probability that an initial sequence $x$ of length $n$, after time $t$,
 is found to match a smaller sequence $y$ of length $n-\ell-r$, offset by $\ell$ sites.
-We refer to such patterns as ``T-mers'', as depicted in Figure \ref{fig:Tmers}.
+%EM Say $a$ is the first $\ell$ bases of $x$ and $b$ is the last $r$ bases of $x$.
+We refer to such patterns as ``T-mers'', as depicted in Figure~\ref{fig:Tmers}.
+%EM Later we use the terminology "base of the T-mer." Let's define that here.
+%EM Shall we define w here?
 This quantity is found by marginalizing $p_m$:
 \begin{align}
+%EM Are we parameterizing in terms of n,\ell,r or n,\ell,m? We see both.
     p_{n,\ell,r}(t;x,y) &= \sum_{a \in \calS^\ell} \sum_{b \in \calS^r} p_n(t;x,a \join y \join b) , \quad \text{for}\; x \in \calS^n, \; y \in \calS^{n-\ell-r} ,
@@
 i.e.\ that the probability a given subsequence matching $x$ is later observed to have its center matching $y$
+%EM should this be p_{n,\ell,m} below?
+%EM What is k here? And do you mean \ell?
 is close to $p_{\ell,m}(x,y)$, regardless of context wider than $2k+l$:
 \begin{align} \label{eqn:window_approx}
+%EM I think I know what you're doing here but the $i$ seems rather lonely here. Do we think of it as something for marginalization or summation or something?
+%EM Also, x on the left hand side appears to be the full sequence, while on the right hand side it appears to be a subsequence. Same for y.
     p_{n,\ell,r}(x,y) \approx \P\{ X_{i+\ell}^{(n-\ell-r)}(t) = y \mid X_i^{(n)}(0) = x \}.
@@
 \begin{align*}
+%EM I'm not accustomed to seeing a big chunky ampersand in prob statements, rather I typically see commas. But you be you if this is how you do it.
     \like(y|x) &= \prod_i \P\{ Y_i = y_i \given Y_{<i} = y_{<i} \;\&\; X=x \} .
@@
 If $\ell$ is long enough that approximation \eqref{eqn:window_approx} holds, then:
-%EM This casual reader is a little confused about the left to right nature here.
 \begin{align} \label{eqn:full_approx}
@@
 then the indirect effects of this change may extend out to $2w-1$ sites.
-In this way, the influence of the sequence at each site extends outwards, as shown in Figure \ref{fig:Tmers} --
+In this way, the influence of the sequence at each site extends outwards, as shown in Figure~\ref{fig:Tmers} --
 so if we take the overhangs long enough that this dependency is unlikely to extend to the base of the Tmer,

### 2021-10-06 writeups/context-dependence.tex
dropping addressed comment

@@
 To do this, let $p_{n,\ell,r}(x,y)$ for $r+\ell < n$ denote
-%EM Can we say initial subsequence or something?
 the probability that an initial sequence $x$ of length $n$, after time $t$,

### 2021-10-07 writeups/context-dependence.tex
Oh, this is getting very cool!

@@
 % \end{align}
+
+This statement is made more rigorous and proven in the Appendix.
 
@@
 here is a simple argument.
+%EM Different notation for transition triples here it seems.
 Suppose the maximum length of any transition pattern is $R = \max\{ |u| : (\mu,u,v) \in \calT \}$.
@@
 These will depend in turn on the outcome of previous possible changes.
+%EM I'm going to suggest a separate sentence defining the si and sij
 For each site $i$ and times $s_1 < s_2$, define the set of states $H(i,s_1,s_2)$
 so that if one knows the possible change points $s_{ij}$,
+%EM Can I suggest "changes to the sequence outside..."
 then changes outside of $H(i,s_1,s_2)$ do not affect the distribution of $X_i(s_2)$.
@@
 \end{gather}
+%EM I'm having some difficulty following... so far H has been described as states, neighborhoods, and changes. I guess it all comes back to subsequences.
 In other words, $H(i,s_1,s_2)$ is the union of the $R$-neighborhoods of $i$ and all possible changes occurring in $H(i,s_1,s_2)$ between $s_1$ and $s_2$.
@@
 Formally, the state $X_i(s_2)$ is conditionally independent of $\{X_j(s_1) \st j \notin H(i,s_1,s_2)\}$.
-Informally, the approximation \eqref{eqn:window_approx} will be good if $H$ does note expand beyond the flanking context.
+Informally, the approximation \eqref{eqn:window_approx} will be good if $H$ does not expand beyond the flanking context.
 Now,
@@
 \begin{align} \label{eqn:prob_approx}
+  %EM You want _not_ contained in for the RHS here, right?
+  %EM I think I need this written out in a little more detail.
   \left| \P\{ X_i^{(m)}(t) = y \mid X(0) \} - p_{\ell,m}(t;x,y) \right| \le 2 \P\left\{  H(\{I,\ldots,i+m\},0,t)\subseteq\{i-\ell,\ldots,i+m+r\} \right\}.
@@
 So, the error is bounded by the probability that $H(I,0,t)$ is larger than the context from $I-\ell$ to $I+r$.
+%EM Oh, this is VERY cool.
+%EM But does this random walk ever jump to the right? If not, does left-only jumping still considered a random walk?
 However, following $H$ backwards in time, the outer edge of $H$ expands as a continuous-time random walk:
@@
 that has mean $R/2$.
-General facts about random walks imply that \eqref{eqn:prob_approx} is bounded above by $C e^{-\alpha \ell/t}$, for some constants $C$ and $\alpha$.
+General facts about random walks imply that~\eqref{eqn:prob_approx} is bounded above by $C e^{-\alpha \ell/t}$, for some constants $C$ and $\alpha$.
 
 \paragraph{Full likelihood}
-    Now we use these ideas to quantify and prove approximation \eqref{eqn:full_approx}.
+Now we use these ideas to quantify and prove approximation~\eqref{eqn:full_approx}.
 For two positions $(i,s)$ and $(j,t)$ in sequence $\times$ time,
@@
 \begin{align}
+%EM Before we had i,s,j,t, and now we have i,t,j,s. It doesn't make a difference, but should we choose one?
+%EM Do we want to specify a max allowable range for j as well on the RHS?
     N_i = \{ (j,s) : r(i,t; j,s) \le r_*, \; 0 \le s \le t \} .
 \end{align}
-These are depicted in figure~\ref{fig:neighborhoods}A.
+These are depicted in Figure~\ref{fig:neighborhoods}A.
 We will write $i \sim j$ if $|i-j| \le 2 r_* t$,
@@
 and $i \nsim j$ otherwise.
-We can think of $N_i$ as the parts of ancestral sequences in which a mutation might have affected the state at $(i,t)$.
+We can think of $N_i$ as the parts of ancestral sequences in which a mutation might have affected the state at $(i,t)$,.
 so if $i \nsim j$, then $i$ and $j$ are (approximately) independent at time $t$, given the sequence at time 0.
@@
 we have a small $\epsilon$ such that for any $i \nsim j$,
-\begin{align}
+\begin{align*}
     &\big|
@@
     \le \epsilon .
-\end{align}
+\end{align*}
 
@@
     \begin{center}
+    %EM This looks like N12 to me. Am I miscounting?
+    %EM I suggest we include t and s in the diagram too.
         \includegraphics{context-neighborhoods-defns}
@@
     \caption{
+        %EM TODO make this A and B so we can refer to them in the text.
         Diagram of things in the text:

### 2021-10-10 writeups/context-dependence.tex
edits until full likelihood bit

@@
 \newcommand{\calP}{\mathcal{P}}  % set of selection pairs
+\newcommand{\calM}{\mathcal{M}}  % set of potential mutations
 \newcommand{\nA}{\mbox{A}}  % nucleotides:
@@
 \newtheorem{theorem}{Theorem}
+\newtheorem{lemma}{Lemma}
 \theoremstyle{definition}
@@
 or of several states evolved along a tree from a single starting point.
-
-The general class of models we consider are known in the probability literature as
-interacting particle systems \citep{liggett2005ips} with neighborhood structure
--- continuous-time Markov chains on lattice-indexed collections of states
-whose transition probabilities are \emph{local}
-in the sense that any instantaneous change only affects a small number of nearby sites,
-and the rates of such instantaneous changes depend only on the states in some bounded neighborhood of those sites to change.
-As finite-state Markov chains,
-transition probabilities are in principle simply expressable as a matrix exponential,
-but this is impractical because the size of the matrix is equal to the number of possible configurations.
-
-In this paper, we propose a solution to this problem,
-showing that the conditional likelihood of local pattern count statistics
-can be well-approximated by marginalizing over a finite amount of surrounding context.
-Motivated by the problem of inferring context-dependent mutation rates from diverged nucleotide sequences,
-we extend the inference framework to observations on trees.
-\peter{something about efficient computation and sparse matrices}
-
-%EM It feels like we go back into introduction mode here after the "In this paper" paragraph.
 
@@
 The general framework also fits certain cellular automata models,
-e.g.\ modeling wildfire \citep{clarke1994cellular},
+e.g., modeling wildfire \citep{clarke1994cellular},
 the spread of HIV \citep{zorzenondossantos2001dynamics},
@@
 
+This general class of models are known in the probability literature as
+interacting particle systems \citep{liggett2005ips} with neighborhood structure
+-- continuous-time Markov chains on lattice-indexed collections of states
+whose transition probabilities are \emph{local}
+in the sense that any instantaneous change only affects a small number of nearby sites,
+and the rates of such instantaneous changes depend only on the states in some bounded neighborhood of those sites to change.
+As finite-state Markov chains,
+transition probabilities are in principle simply expressable as a matrix exponential,
+but this is impractical because the size of the matrix is equal to the number of possible configurations.
+
+In this paper, we propose a solution to this problem,
+showing that the conditional likelihood of local pattern count statistics
+can be well-approximated by marginalizing over a finite amount of surrounding context.
+Motivated by the problem of inferring context-dependent mutation rates from diverged nucleotide sequences,
+we extend the inference framework to observations on trees.
+The tools are available as an R package,
+which is fairly efficient thanks to 
+computational techniques using sparse matrices.
+
 %%%%%%% %%%%%%%%%%
@@
 Consider a 1-dimensional grid of sites, each of which can take one of a finite set of states,
-and that switch randomly between states according to a local set of rules;
-we then observe a finite collection of these sites at only a few times,
+and that switch randomly between states according to a local set of rules.
+We then observe a finite collection of these sites at only a few times,
 Suppose that the dynamics are Markov:
-with the set of possible states by $\calS$,
-and $X_i(t)$ the state of site $i$ at time $t$,
+writing $\calS$ for the set of possible states 
+and $X_i(t)$ for the state of site $i$ at time $t$,
 we assume that $\{X(t)\}_{t \ge 0}$ is a Markov process on sequences of $n$ states $\calS^n$
@@
   This only has one parameter, the speed.
-  %EM I can't actually tell if what is developed below addresses this difficulty!
   If we allow particles to exit from the right, and new particles to enter from the left
-  (or, arrange them on a circle)
   then from observing only starting and ending configurations
-  we cannot tell which particles have moved where,
+  we cannot tell with certainty which particles have moved where,
   so it is not obvious how to estimate the speed.
+  However, summing over possible particle movements would allow us to
+  compute the likelihood of a given configuration change,
+  and so estimate the speed by maximum likelihood.
 
@@
 
-%EM I think we could have both, but certainly this biological interpretation. Then we would have "selection" and "fitness" rather than "potential" and "energy"?
+%EM I think we could have both, but certainly this biological interpretation.
+%Then we would have "selection" and "fitness" rather than "potential" and
+%"energy"?
 \peter{Should call this ``selection'' instead of a ``potential''? Or what?}
@@
 where $n(x,p_i)$ is the number of times the pattern $p_i$ occurs in $x$.
-%EM Can we introduce the phi notation here, and say something a little more formal? Namely that if the mutation rate is mu then the actual rate of change is mu times phi of the difference of energies?
+These will affect the rates through a nonnegative function $\phi$:
+we declare that if the rate at which $x$ changes to $y$ as computed
+from transition triples is $\mu$,
+then the actual rate is
+$$
+    \mu \phi(E(y) - E(x)) .
+$$
+If $\phi(e)$ is not greater than 1, one way to think about this is that 
+the transition triples give ``proposed changes'',
+but these changes only take effect with probability $\phi(E(y) - E(x))$.
 
@@
     where $\gamma$ is the additional CpG rate above the base mutation rate.
-    %EM This "more than one way" isn't as clear as "more than one pattern may match" used above, IMHO. Because it's the same underlying substitutions, just different patterns getting us there, right?
-    Note that for a given pair of sequences, there may be more than one way to mutate a given string $x$ to get $y$:
-    for instance, a change $\nA\nC\nG \to \nA\nT\nG$ could have occurred by a single $\nC \to \nT$ mutation,
+    Note that for a given pair of sequences,
+    a change $\nA\nC\nG \to \nA\nT\nG$ could have occurred by a single $\nC \to \nT$ mutation,
     or by a $\nC\nG \to \nT\nG$ mutation;
-    and hence the total instantaneous rate at which ACG changes to ATG is $\gamma + m_{\nC\nT}$.
+    in such cases
+    the total instantaneous rate at which ACG changes to ATG is the sum of the rates
+    (in this example, $\gamma + m_{\nC\nT}$).
     % Although this means there are many possible parameterizations,
@@
 \paragraph{The generator matrix}
-For clarity, we describe how the set of transition rates on patterns determines the transition rate matrix for complete sequences, that is,
+We now describe concretely how the set of transition rates on patterns determines the transition rate matrix for complete sequences, that is,
 the $|\calS|^n \times |\calS|^n$ matrix $G(n)$ whose $(x,y)^\text{th}$ entry gives the instantaneous rate
@@
 \]
-i.e.\ if $x$ and $y$ match except at positions $i,i+1,\ldots,i+h-1$,
+i.e., if $x$ and $y$ match except at positions $i,i+1,\ldots,i+h-1$,
 and for those positions $x$ matches with $u$ while $y$ matches with $v$.
 
-As noted above, there may be more than one way to mutate a such a sequence $x$ to get such a $y$:
+As noted above, there may be more than one way to mutate a sequence $x$ to get $y$:
 for each position $i$, let $J(i,x,y)$ be those transitions in $\calT$ that can be applied at $i$
-to change $x$ into $y$, i.e.
+to change $x$ into $y$, i.e.,
 \[
-J(i,x,y) = \{ (\mu^j,u^j,v^j) \in \calT \st x \xrightarrow{i,u^j,v^j} y \}.
+    J(i,x,y) = \{ (\mu^{(j)},u^{(j)},v^{(j)}) \in \calT \st x \xrightarrow{i,u^{(j)},v^{(j)}} y \}.
 \]
 The rate $G(n)_{x,y}$ is then the sum of all matching transition rates,
+multiplied by the potential difference,
 namely
 \begin{align} \label{eqn:G_defn}
-    G(n)_{x,y} = \sum_{i=1}^n \phi(E(y)-E(x)) \sum_{J(i,x,y)}  \mu^j ,
+    G(n)_{x,y} = \phi(E(y)-E(x)) \sum_{i=1}^n \sum_{J(i,x,y)}  \mu^j ,
 \end{align}
@@
 also used for simulation,
-%EM My understanding is that this is also called uniformization http://dx.doi.org/10.1214/09-AOAS247
-using a variant of the ``jump chain representation'', also known as the ``Gillespie algorithm''.
+using a variant of the ``jump chain representation'', also known as the ``Gillespie algorithm''
+or ``uniformization'' \citep{hobolth2009simulation}.
 Briefly, this  works by first sampling a homogeneous Poisson process of times and locations of possible changes,
-at an appropriate rate $\mu_*$ per site,
+at an appropriate rate $\mu^*$ per site,
 and then resolving each possible change in temporal order.
-The rate $\mu_*$ should be the maximum ``local'' rate at which transitions occur, across sites and transition outcome states:
+The rate $\mu^*$ should be the maximum ``local'' rate at which transitions occur, across sites and transition outcome states:
 \[
-%EM Here we're moving to lower-j.
-\mu_* = \max \left\{ \sum_{y \in \calS^n} \sum_{j\in J(i,x,y)} \mu_j \qquad \st x \in \calS^n, \quad 1 \le i \le n \right\} .
+    \mu^* = \max \left\{ \sum_{y \in \calS^n} \sum_{j\in J(i,x,y)} \mu^{(j)} \qquad \st x \in \calS^n, \quad 1 \le i \le n \right\} .
 \]
-Note that since the process is homogeneous, the maximum rate of changes occurring at any position will be the same, except possibly near the end of the sequence,
-%EM I wonder if we could do something like introduce J(x), which is the set of all possible transitions starting at x, so we don't have to do this "unnecessary" thing. Or perhaps it's fine as is.
-so the maximization over $i$ is unnecessary.
-Now sample possible changes: let $c_{ij}$ denote the time of the $j^\text{th}$ possible change at site $i$,
-so that for each site $i$,
+Now sample possible changes:
+in a sequence of length $L$, over a time period $t$,
+first draw the number of possible changes, $N$,
+from a Poisson distribution with mean $\mu^* t L$,
+and then choose the times $s_k$ and locations $i_k$
+at which possible changes occur uniformly
+over $[0, t]$ and $\{1, 2, \ldots, L\}$ respectively.
+Order the possible change times, so that
 \[
-0 \le c_{i1} < c_{i2} < \cdots < c_{iN_i} \le t
+    0 < s_1 < \cdots < s_N < t .
 \]
-is a draw from a rate $\mu_*$ Poisson process on $[0,t]$.
-These processes are independent across sites.
-Now, suppose we have determined the state at time $s$ to be $X(s) = x$,
-and the next event is $c_{ij} = \min_{k,h} \{ c_{kh} : c_{kh}>s \}$.
-At $c_{ij}$,
-a new state is then chosen at position $i$
-with probability proportional to the corresponding rate, i.e.\
-%EM Not sure what I was on about here.
-% \erick{Previously, the $j$ in $\mu^j$ was used to index the possible patterns. Now it's indexing possible changes at a site.}
+(Note that this is equivalent to saying that possible
+changes occur at rate $\mu^*$ independently at each of the $L$ sites.)
+Now, suppose we have determined the state just before time $s_k$ to be $X(s_k) = x$.
+Then, a new state is then chosen at position $i_k$
+with probability proportional to the corresponding rate, i.e.,
+out of the possible transitions $\calT$, the $j^\text{th}$ one is chosen
+with probability
 \[
 q_j = \begin{cases}
-  \mu_j/\mu_* \qquad & \text{if } x_i^{(|u^j|)} = u^j  \\
-  0 \qquad & \text{otherwise,}
+    \mu^{(j)}/\mu^* \qquad & \text{if } x_i^{(|u^{(j)}|)} = u^{(j)}  \\
+    0 \qquad & \text{otherwise.}
 \end{cases}
 \]
-is the probability that $x_i^{(|u^j|)}$ is replaced with $v^j$.
-The remaining probability $q_0 = 1-\sum_{j=1}^{|\calT|} q_j$ gives the probability that the state remains the same.
+The remaining probability $q_0 = 1-\sum_{j=1}^{|\calT|} q_j$
+gives the probability that the state remains the same.
 (By construction of $\mu_*$, $q_0\ge 0$.)
-
-For instance, to simulate the TASEP, one has only to let $\mu_*=\lambda$, and at each $c_{ij}$
+If triple $(\mu^{(j)}, u^{(j)}, v^{(j)})$ is chosen, then
+the substring of $x$ matching $u^{(j)}$ that begins at position $i$
+is replaced with $v^j$ to produce $X(s_k)$.
+
+For instance, to simulate the TASEP, one has only to let $\mu^*=\lambda$, and at each $c_{ij}$
 check if at site $i$ there is a 1 followed by a 0,
@@
 
-%EM I don't see this needing shortening!
-\peter{shorten, refer to figure}
+\peter{refer to figure}
 
@@
 \]
-However, there is no guarantee that this is even close to correct,
-as it ignores dependencies between neighboring blocks.
+However, this ignores dependencies between neighboring blocks,
+so it is not clear how good this approximation might be.
 % Intuitively, we want to increase the size of the context
@@
         Nonmutated sites are shown in lower case letters.
+        The ``base'' of the T-mer is the lower central
+        string of length $m = n - \ell - r$.
         \textbf{(B)}
@@
         to the six positions colored red in the final sequence.
+        The rate of each mutation depends on its' neighboring sites,
+        i.e., with a window of width $w=1$.
         Red dotted arrows show the propagation of dependency
@@
         and the third site in the final sequence (bottom).
-        %EM This "window" terminology... should we have another quick definition here?
-        The maximum window width, $w$, is equal to 1.
         \label{fig:Tmers}
@@
 we can compute \emph{something} correctly,
-%EM The full exact likelihood? Isn't it an approximation?
-and that something will be sufficient to compute the full likelihood.
-To do this, let $p_{n,\ell,r}(x,y)$ for $r+\ell < n$ denote
-the probability that an initial sequence $x$ of length $n$, after time $t$,
-is found to match a smaller sequence $y$ of length $n-\ell-r$, offset by $\ell$ sites.
-%EM Say $a$ is the first $\ell$ bases of $x$ and $b$ is the last $r$ bases of $x$.
+and that something will be sufficient to compute the full likelihood
+to an arbitrarily good approximation.
+To do this, let $p_{n,\ell,r}(x,y)$ for $\ell + r < n$ denote
+the probability that an initial subsequence $x$ of length $n$, after time $t$,
+is found to match a smaller subsequence $y$ of length $n-\ell-r$, offset by $\ell$ sites.
+This central, smaller subsequence is called the ``base'' of the T-mer.
 We refer to such patterns as ``T-mers'', as depicted in Figure~\ref{fig:Tmers}.
-%EM Later we use the terminology "base of the T-mer." Let's define that here.
 %EM Shall we define w here?
-This quantity is found by marginalizing $p_m$:
+%PR No, I think that'll muddy the waters.
+This quantity is found by marginalizing $p_m$,
+summing over the possible inital and final substrings of $x$
+(denoted $a$ and $b$):
 \begin{align}
-%EM Are we parameterizing in terms of n,\ell,r or n,\ell,m? We see both.
-    p_{n,\ell,r}(t;x,y) &= \sum_{a \in \calS^\ell} \sum_{b \in \calS^r} p_n(t;x,a \join y \join b) , \quad \text{for}\; x \in \calS^n, \; y \in \calS^{n-\ell-r} ,
+    p_{n,\ell,r}(t;x,y)
+    &=
+    \sum_{a \in \calS^\ell} \sum_{b \in \calS^r} p_n(t;x,a \join y \join b) ,
+    \quad
+    \text{for}\; x \in \calS^n, \; y \in \calS^{n-\ell-r} ,
 \end{align}
 where $a \join y \join b$ is the sequence composed of concatenating $a$, $y$, and $b$ together in that order.
-We expect $p_{n,\ell,m}(x,y)$ to be asymptotically correct when $\ell$ is large (and $t$ is fixed),
-i.e.\ that the probability a given subsequence matching $x$ is later observed to have its center matching $y$
-%EM should this be p_{n,\ell,m} below?
-%EM What is k here? And do you mean \ell?
-is close to $p_{\ell,m}(x,y)$, regardless of context wider than $2k+l$:
+We expect $p_{n,\ell,r}(x,y)$ to be asymptotically correct when $\ell$ and $r$ are large (and $t$ is fixed).
+In Appendix~\ref{ss:approx_pf}, we show that this is so, namely that
+the probability a given subsequence of length $n$ matching $x$
+is later observed to have its center matching $y$
+is close to $p_{n,\ell,m}(x,y)$, regardless of wider context.
+Formally, for each $i$,
 \begin{align} \label{eqn:window_approx}
-%EM I think I know what you're doing here but the $i$ seems rather lonely here. Do we think of it as something for marginalization or summation or something?
-%EM Also, x on the left hand side appears to be the full sequence, while on the right hand side it appears to be a subsequence. Same for y.
-    p_{n,\ell,r}(x,y) \approx \P\{ X_{i+\ell}^{(n-\ell-r)}(t) = y \mid X_i^{(n)}(0) = x \}.
+    \P\{ X_{i+\ell}^{(n-\ell-r)}(t) = y \mid X_i^{(n)}(0) = x \}
+    \approx
+    p_{n,\ell,r}(x,y)
+    .
 \end{align}
@@
 %EM I feel like it needs emphasis that this is different than saying that we assume that the rates stay the same as in the initial sequence.
+%PR I'm not sure what you mean here?
 
@@
 the probability that a given sequence $x$ evolves into another, $y$, over a given period of time.
-This can be computed as the product of the probability of each site in $y$ given $x$ and previous sites:
-defining $x_{<i}=(x_1, \ldots, x_{i-1})$, this is:
+This can be computed as the product of the probability of each site in $y$
+given $x$ and all sites in $y$ further to the left.
+Defining $x_{<i}=(x_1, \ldots, x_{i-1})$ to be the subsequence to the left of $i$, this is:
 \begin{align*}
-%EM I'm not accustomed to seeing a big chunky ampersand in prob statements, rather I typically see commas. But you be you if this is how you do it.
-    \like(y|x) &= \prod_i \P\{ Y_i = y_i \given Y_{<i} = y_{<i} \;\&\; X=x \} .
+    \like(y|x) &= \prod_i \P\{ Y_i = y_i \given Y_{<i} = y_{<i} \;\text{and}\; X=x \} .
 \end{align*}
-Using the same methods, we can approximate the terms in this sum
+It turns out that
+we can approximate the terms in this sum
 by discarding approximately independent sites from the conditioning.
-%EM The following sentence may get us in trouble with the formalization police.
 If $\ell$ is long enough that approximation \eqref{eqn:window_approx} holds, then:
 \begin{align} \label{eqn:full_approx}
-    \P\{ Y_i = y_i \given Y_{<i} \;\&\; X \}
+    \begin{split}
+    \P\{ Y_i = y_i \given Y_{<i} \;\text{and}\; X \}
     &\approx
-        \P\{ Y_i = y_i \given Y_{i-2\ell}^{(2\ell+1)} \;\&\; X_{i-3\ell}^{(4\ell+1)} \}  \\
+        \P\{ Y_i = y_i \given Y_{i-2\ell}^{(2\ell+1)} \;\text{and}\; X_{i-3\ell}^{(4\ell+1)} \}  \\
     &=
@@
             p_{2\ell+1,\ell,\ell+1}(x_{i-3\ell}^{(4\ell+1)}, y_{i-2\ell}^{(2\ell)})
-        } .
+        } 
+    \end{split} .
 \end{align}
@@
 % \end{align}
-
-This statement is made more rigorous and proven in the Appendix.
+See Appendix~\ref{ss:approx_pf} for a more rigorous statement and proof,
+and the next section for some intuition.
 
@@
 
-The approximations in \eqref{eqn:window_approx} and \eqref{eqn:full_approx} hold if the overhang, $\ell$, is long enough.
+The approximations in \eqref{eqn:window_approx} and \eqref{eqn:full_approx} hold
+if the overhang, $\ell$, is long enough.
 The intuition is straightforward, and provides a way to pick the overhang lengths $\ell$ and $r$;
-a precise statement is given in Section \ref{ss:approx_pf}.
-Suppose that each instantaneous change depends on positions at most $w$ sites away.
+a precise statement is given in Appendix \ref{ss:approx_pf}.
+Suppose that each instantaneous change depends on positions at most $w$ sites away,
+and refer to $w$ as the ``window width''.
 Then, the mutation process at any particular site
-depends on the initial sequence at that site and within distance $w$ on either side,
-and the outcomes of any mutations that have occurred within that window.
+depends on the initial sequence at that site and within distance $w$ on either side\ldots
+\emph{and} the outcomes of any mutations that have occurred within that window.
 
@@
 If another change happens, say, at the end of this window,
-then the indirect effects of this change may extend out to $2w-1$ sites.
+then the indirect effects of this change may extend out to $2w$ sites.
 In this way, the influence of the sequence at each site extends outwards, as shown in Figure~\ref{fig:Tmers} --
@@
 Concretely, sites further away than $kw$ only matter if there is a chain of at least $k$ intervening mutations
--- which is unlikely if $k$ is larger than the expected number.
-
-For instance, in the CpG model the maximum dependency is $w=1$ as the process only depends on neighboring sites;
+-- which is unlikely if $k$ is much larger than the expected number.
+
+For instance, in the CpG model the window width is $w=1$
+because the process only depends on neighboring sites;
 and if the sequence has evolved for time $t$ then we expect no more than a Poisson($\lambda t$) number of changes at each site.
@@
 
+Fitting a model to nucleotide data would likely require iterative modification of the model.
 The difference between observed and expected T-mer counts provide a good assessment of how and whether a model can be improved.
@@
 Concretely: pick a random location $I$ in the sequence and let $X_{I-\ell}^{m+2\ell}(0) = \rho$ be the (long) pattern seen there at the root,
-and write $\pi(z)$ for the frequency of pattern $z$ in $X(0)$, i.e.\ $\P\{\rho=z\}=\pi(z)$.
+and write $\pi(z)$ for the frequency of pattern $z$ in $X(0)$, i.e., $\P\{\rho=z\}=\pi(z)$.
 % Given the corresponding pattern in $X(t_X)$, the distribution of $\rho$ is proportional to
@@
   \end{center}
-  \caption{ Results from iterative fitting to a $10^6$ sequence evolved to have 28.5\% sequence divergence
+  \caption{ Results from iterative fitting to a $10^6$bp sequence
+    evolved to have 28.5\% sequence divergence
     with the above mutational motifs. The transitions $CG \to TG$ and $CG \to CA$
@@
 along the way substantial encouragement and useful suggestions were provided by
-Erick Matsen, Simon Tavar\'e, Rasmus Nielsen, Graham Coop, Yaniv Brandvain, and Sergey Nuzhdin.
-In particular, Erick Matsen helped improve the code and suggested the name ``T-mer''.
+Simon Tavar\'e, Rasmus Nielsen, Graham Coop, Yaniv Brandvain, and Sergey Nuzhdin.
 Many thanks are also due Jessica Crisci for curating the human-chimp data.
@@
 
+First we quantify and prove the approximation \eqref{eqn:window_approx}.
+The basic idea is that the probability is only in error if there is a chain of mutations
+that create a dependency between the base of the T-mer and the flanking sequence.
+To simplify the argument, first assume that all transition triples
+change only a single, central site in the pattern,
+taking into account context on each side of maximum length $w$.
+(For instance, a trinucleotide model in which the flanking sites on each side
+affect the mutation rate of the central site has $w=1$.)
+Recall that in our construction of the process we first chose the times and sites
+at which potential changes might occur.
+Suppose a potential change occurs at site $i-w$, at time $t$,
+which therefore might result in a change in state at site $i$.
+To resolve this, we need to know the outcome of any other potential changes
+that might change the sequence between $i-w$ and $i + w$.
+Consider those that might happen only to the right of $i$,
+thus extending the sequence on which the change depends to the right.
+The number of such potential changes is Poisson with mean $\mu^* w t$,
+and will extend the amount of sequence on which the change at site $i$ depends
+by a uniform random number of sites between 1 and $w$.
+If one occurs, say at time $0 < s < t$,
+then any further potential changes that occur between time 0 and time $s$
+in the $w$ sites to the right of this older potential change
+will again extend the range of dependency.
+Putting this together, the amount of sequence to the right of a site
+on which a mutation at that site depends
+is bounded by the compound Poisson process $\sum_{k=1}^N W_k$,
+where $N$ is Poisson with mean $\mu^* w t$
+and each $W_k$ is independent and uniform on $\{1, 2, \ldots, w\}$.
+
+Now consider the probability of interest,
+\begin{align*}
+    \P\{ X_{i+\ell}^{(n-\ell-r)}(t) = y \mid X_i^{(n)}(0) = x \} .
+\end{align*}
+This probability can be written as a sum over all possible arrangements
+and resolutions of potential changes.
+Our approximation, $p_{n,\ell,r}(x,y)$ is the probability that would be obtained
+by ignoring any potential changes that rely on context outside of the $n$-sequence $x$.
+So, if $A$ is the event that the final subsequence $y$
+depends on seequence outside of $x$ due to potential changes,
+then
+\begin{align*}
+    \left|
+        \P\{ X_{i+\ell}^{(n-\ell-r)}(t) = y \mid X_i^{(n)}(0) = x \}
+        -
+        p_{n,\ell,r}(x,y)
+    \right|
+    \le \P\{A\} .
+\end{align*}
+Since the size of the dependency window is a compound Poisson process,
+we can bound $\P\{A\}$ using the following lemma:
+
+\begin{lemma}
+    Let $X = \sum_{j=1}^N W_j$,
+    where $N$ is Poisson with mean $\lambda$
+    and each $W_j$ is independent and bounded between 0 and $w$.
+    Then for $k \ge 1$,
+    \begin{align}
+        \P\{ X > k w \}
+        \le
+        \frac{\lambda^{k+1}}{(k+1)!} .
+    \end{align}
+\end{lemma}
+
+% # what's this bound look like and how's it compare to a Chebyshev bound
+% # that uses the variance of W
+%
+% ub1 <- function (w, lam, ell=1:10) {
+%   k1 <- ceiling(ell/w)
+%   return( lam^(k1+1) / factorial(k1+1) )
+% }
+% # the Chebyshev bound never does better
+% ub2 <- function (w, lam, ell=1:10) {
+%   k2 <- (ell - lam * (w+1) / 2) / w
+%   return( lam * (w+1) * (2*w+1) / (6 * w^2 * k2^2) )
+% }
+% wvals <- c(1, 2, 3)
+% ub1mat <- sapply(wvals, ub1, lam=0.75)
+% ub2mat <- sapply(wvals, ub2, lam=0.75)
+% matplot(ub1mat, type='l', lty=1)
+% matlines(ub2mat, type='l', lty=2)
+% legend('topright', lty=1, col=seq_along(wvals), legend=sprintf("w=%d", wvals))
+
 \begin{proof}
-
-First we quantify and prove the approximation \eqref{eqn:window_approx}.
-In fact, the approximation improves exponentially as $\ell$ and $r$ increase --
-here is a simple argument.
-%EM Different notation for transition triples here it seems.
-Suppose the maximum length of any transition pattern is $R = \max\{ |u| : (\mu,u,v) \in \calT \}$.
-The transition rate at site $i$ at time $s$ is determined if we know $X_{i-R}^{(2R+1)}(s)$;
-this can be determined from $X_{i-R}^{2R+1}(0)$ and the outcome of all changes happening before time $s$ at sites in $(i-R-1, \ldots, i+R)$.
-These will depend in turn on the outcome of previous possible changes.
-%EM I'm going to suggest a separate sentence defining the si and sij
-For each site $i$ and times $s_1 < s_2$, define the set of states $H(i,s_1,s_2)$
-so that if one knows the possible change points $s_{ij}$,
-%EM Can I suggest "changes to the sequence outside..."
-then changes outside of $H(i,s_1,s_2)$ do not affect the distribution of $X_i(s_2)$.
-Concretely, define $H(i,s_1,s_2)$ by the property that $i \in H(i,s_1,s_2)$, and
-\begin{gather}
-  \text{if } s_1 < s_{jk} \le s_2 \text{ for some } i-R-1 \le j \le i+R, \text{ then } H(j,s_1,s_{jk}) \subseteq H(i,s_1,s_2) .
-\end{gather}
-%EM I'm having some difficulty following... so far H has been described as states, neighborhoods, and changes. I guess it all comes back to subsequences.
-In other words, $H(i,s_1,s_2)$ is the union of the $R$-neighborhoods of $i$ and all possible changes occurring in $H(i,s_1,s_2)$ between $s_1$ and $s_2$.
-For a set of states $I$, let $H(I,s_1,s_2) = \bigcup_{i\in I} H(i,s_1,s_2)$.
-Formally, the state $X_i(s_2)$ is conditionally independent of $\{X_j(s_1) \st j \notin H(i,s_1,s_2)\}$.
-Informally, the approximation \eqref{eqn:window_approx} will be good if $H$ does not expand beyond the flanking context.
-Now,
-\begin{multline}
-  \P\{ X_i^{(m)}(t) = y \mid X(0) \}
-  = \P\left\{ H(\{i,\ldots,i+m\},0,t)\subseteq\{i-\ell,\ldots,i+m+r\} \right\} \\
-   \qquad \qquad {} \times \P\left\{ X_i^{(m)}(t) = y \mid X_{i-\ell}^{(\ell+m+r)}(0), \; H(\{i,\ldots,i+m\},0,t)\subseteq\{i-\ell,\ldots,i+m+r\} \right\} \\
-     \qquad {} + \P\left\{ H(\{i,\ldots,i+m\},0,t)\nsubseteq\{i-\ell,\ldots,i+m+r\} \right\}  \\
-   \qquad\qquad {} \times \P\left\{ X_i^{(m)}(t) = y \mid X_{i-\ell}^{(\ell+m+r)}(0), \; H(\{i,\ldots,i+m\},0,t)\nsubseteq\{i-\ell,\ldots,i+m+r\} \right\} ,
-\end{multline}
-and so
-\begin{align} \label{eqn:prob_approx}
-  %EM You want _not_ contained in for the RHS here, right?
-  %EM I think I need this written out in a little more detail.
-  \left| \P\{ X_i^{(m)}(t) = y \mid X(0) \} - p_{\ell,m}(t;x,y) \right| \le 2 \P\left\{  H(\{I,\ldots,i+m\},0,t)\subseteq\{i-\ell,\ldots,i+m+r\} \right\}.
-\end{align}
-
-So, the error is bounded by the probability that $H(I,0,t)$ is larger than the context from $I-\ell$ to $I+r$.
-%EM Oh, this is VERY cool.
-%EM But does this random walk ever jump to the right? If not, does left-only jumping still considered a random walk?
-However, following $H$ backwards in time, the outer edge of $H$ expands as a continuous-time random walk:
-for instance, the left edge $L(s) = \min \{ j \in H(I,t-s,t) \}$ jumps leftwards at rate $R\mu_*$ with a random jump size
-that has mean $R/2$.
-General facts about random walks imply that~\eqref{eqn:prob_approx} is bounded above by $C e^{-\alpha \ell/t}$, for some constants $C$ and $\alpha$.
+    First note that $\P\{X > kw\} \le \P\{N > k\}$.
+    Now, use the fact that if $f(n) = n (n-1) \cdots (n-k) / (k+1)!$
+    then $f(n) = 0$ for integers $0 \le n \le k$ and $f(n) \ge 1$ for $n > k$,
+    and so $\P\{N > k\} \le \E[f(N)] = \lambda^{k+1} / k!$.
+\end{proof}
+
+Putting this together,
+we find that 
+\begin{align*}
+    \left|
+        \P\{ X_{i+\ell}^{(n-\ell-r)}(t) = y \mid X_i^{(n)}(0) = x \}
+        -
+        p_{n,\ell,r}(x,y)
+    \right|
+    \le
+    \frac{
+        2 (\mu^* t)^{k+1}
+    }{
+        (k+1)!
+    } .
+\end{align*}
+Here, the factor of 2 comes from the two sides (to the left and right) of the base of the T-mer.
+
+The argument applies to more general transition triples,
+with $w$ set equal to the maximum distance
+from a position that is changed to the edge of the context
+across all transition triples.
 
@@
     &\big|
-    \log( \P\{ X_i(t) = x_i \and X_j(t) = x_j \} ) \\
+    \log( \P\{ X_i(t) = x_i \text{ and } X_j(t) = x_j \} ) \\
     &\qquad {} -
@@
     \prod_{i=1}^n
-    \P\{ X_i(t) = y_i \given X(0) = x \and X_{<i}(t) = y_{<i} \} .
+    \P\{ X_i(t) = y_i \given X(0) = x \text{ and } X_{<i}(t) = y_{<i} \} .
 \end{align}
@@
 &\big|
-\log( \P\{ X_i(t) = y_i \given X(0) = x \and X_{<i}(t) = y_{<i} \} ) \\
+    \log( \P\{ X_i(t) = y_i \given X(0) = x \text{ and } X_{<i}(t) = y_{<i} \} ) \\
 & \qquad {} -
-\log( \P\{ X_i(t) = y_i \given X_{A(L(i))}(0) = x_{A(L(i))} \and X_{L(i)}(t) = y_{L(i)} \} )
+    \log( \P\{ X_i(t) = y_i \given X_{A(L(i))}(0) = x_{A(L(i))} \text{ and } X_{L(i)}(t) = y_{L(i)} \} )
  \big| \\
@@
     \prod_{i=1}^n
-    \P\{ X_i(t) = y_i \given X_{A(L(i))}(0) = x_{A(L(i))} \and X_{L(i)}(t) = y_{L(i)} \} ,
+    \P\{ X_i(t) = y_i \given X_{A(L(i))}(0) = x_{A(L(i))} \text{ and } X_{L(i)}(t) = y_{L(i)} \} ,
 \end{align}
@@
 
-\end{proof}
-
 \section{Peeling algorithm}
@@
 each ``long'' sequence at that node, along with the combinations of ``short'' sequences
-on all tips further away from $a$ (i.e.\ below that node if it was rooted at $a$).
+on all tips further away from $a$ (i.e., below that node if it was rooted at $a$).
 For a node $w$ this will be stored as a $|\calS|^{m+2\ell} \times |\calS|^{cm}$ matrix, for some $c$,
@@
 If $i_1 < \ldots < i_{n(x)}$ are a set of nonoverlapping sites where the initial sequence matches $x$
--- i.e.\ $X_{i_k}^{(2\ell+m)}(0)=x$ and $i_{k+1} > i_k + 2\ell+m$ for all $k$ %En dash close to math could be confusing. This sentence is a doozy.
+-- i.e., $X_{i_k}^{(2\ell+m)}(0)=x$ and $i_{k+1} > i_k + 2\ell+m$ for all $k$ %En dash close to math could be confusing. This sentence is a doozy.
 then the counts $N(x,y)=\#\{ i_k \st X_{i_k+\ell}^{(m)}(t)=y\}$ (the number of these sites that match the smaller pattern $y$ at time $t$)
@@
 but this time, count over all positions,
-so that e.g.\ $N(x) = \#\{ 1 \le i \le n - 2\ell - m + 1 \st X_i^{(m+2\ell)}(0)=x\}$.
+so that e.g., $N(x) = \#\{ 1 \le i \le n - 2\ell - m + 1 \st X_i^{(m+2\ell)}(0)=x\}$.
 
@@
 The covariance calculations show that
-if $p_{\ell,m}(x,y)$ is small enough (i.e.\ the pattern is not common),
+if $p_{\ell,m}(x,y)$ is small enough (i.e., the pattern is not common),
 then $N(x,y)$ is close to Poisson.
@@
 and any overlapping patterns must require additional changes,
-i.e.\ if $(x_1,y_1)$ matches at $i_1$,
+i.e., if $(x_1,y_1)$ matches at $i_1$,
 then for $(x_2,y_2)$ to match at $i_2$,
@@
 \end{align}
-i.e.\ the exponential of the total energy of each three adjacent sites,
+i.e., the exponential of the total energy of each three adjacent sites,
 where $Z$ is the normalizing constant.

### 2021-10-11 writeups/context-dependence.tex
Quick pass through the edits

@@
     Note that for a given pair of sequences,
-    a change $\nA\nC\nG \to \nA\nT\nG$ could have occurred by a single $\nC \to \nT$ mutation,
+    a change $\nA\nC\nG \to \nA\nT\nG$ could have occurred by a $\nC \to \nT$ mutation
+    via the independent-sites model,
     or by a $\nC\nG \to \nT\nG$ mutation;

### 2021-10-11 writeups/context-dependence.tex
Nit-picking the shit out of section 1

@@
 depends only on the sequence of states at nearby sites.
+%EM If we think that this can be generalized to the setting of position-specific local models, then let's mention that here and then put in a little section.
 
@@
 first define a \textit{pattern} to mean a contiguous sequence of states;
+%EM Do we use this length notation? I couldn't easily find an instance.
 let $|u|$ denote the length of the pattern $u$.
@@
     \begin{tabular}{c@{\quad$\to$\quad}c@{\quad at rate\quad }c}
-      10  &   01   &  $\lambda$
+      10  &   01   &  $\lambda$.
     \end{tabular}
@@
 
-  This only has one parameter, the speed.
+  \noindent
+  This only has one parameter.
   If we allow particles to exit from the right, and new particles to enter from the left
@@
 %"energy"?
+%EM Yes. Seems nice unless you think that this "fitness" model would just annoy a pop gen person.
 \peter{Should call this ``selection'' instead of a ``potential''? Or what?}
@@
 and that the probability a proposed change actually occurs depends on how much it affects the energy.
-Concretely, let $\calP = \{(p_i, e_i\}$ be a set of (pattern, energy change) pairs,
+%EM I'm trying to understand why I should think of e_i as an energy *change*. Why not just an energy of pattern i? We already take differences of energies.
+Concretely, let $\calP = \{(p_i, e_i)\}$ be a set of (pattern, energy change) pairs,
 and for any sequence $x$ define
-$E(x) = \sum_i e_i n(x,p_i)$,
+$E(x) = \sum_i e_i\, n(x,p_i)$,
 where $n(x,p_i)$ is the number of times the pattern $p_i$ occurs in $x$.
@@
 then the actual rate is
-$$
-    \mu \phi(E(y) - E(x)) .
-$$
+\[
+    \mu \, \phi\left(E(y) - E(x)\right) .
+\]
 If $\phi(e)$ is not greater than 1, one way to think about this is that
@@
       \begin{tabular}{c@{\quad$\to$\quad}c@{\quad at rate\quad }cc}
-        $x$  &  $y$  &  $m_{xy}$ & when $x \neq y \in \{\nA,\nC,\nG,\nT\}$  \\
+        $x$  &  $y$  &  $m_{xy}$ & for $x, y \in \{\nA,\nC,\nG,\nT\}$ and $x \neq y$ \\
         \nC\nG   &  \nT\nG   &  $\gamma$ & \\
         \nC\nG   &  \nC\nA   &  $\gamma$ &
-      \end{tabular} ,
+      \end{tabular},
     \end{center}
@@
     This is incorporated as follows: set
+    %EM Is this table clearer than just saying "set the energy of pattern C or G to be $s$...."
     \begin{center}
@@
     \begin{align*}
+    %EM Does it make more sense to have the N here in the selection component, or in the mutation component with \mu?
+    %EM Also, could we stick with $s$ for this equation? I don't see a compelling reason to switch to e, which doesn't appear outside of the context of this equation as far as I can see.
         \phi(e) = N (1-\exp(-2e))/(1-\exp(-2eN))
@@
     so that if a given change that occurs with rate $\mu$ would change a sequence $x$ to $y$,
-    then $\mu \phi(E(y)-E(x))$ is the rate at which the mutation appears and successfully takes over in the population.
+    then $\mu \, \phi(E(y)-E(x))$ is the rate at which the mutation appears and successfully takes over in the population.
 
@@
     imagined as magnetic dipoles,
+    %EM re i+1 in the following equation: as soon as I read this I'm already wondering what happens at the end of the sequence. Could we specify that here? Are we doing a loop?
     and the energy associated with a given state $x$ is $E(x) = - \frac{1}{2} \beta \sum_i x_i x_{i+1} - \frac{1}{2} \gamma \sum_i x_i$.
@@
 and for those positions $x$ matches with $u$ while $y$ matches with $v$.
+Let $\calT$ be the set of these transition triples.
 
@@
 \[
+%EM I think we want h and not j here in the parenthetical superscript.
+%EM Perhaps we want \mu^{j} rather than \mu^{(j)}? I think the latter is more for subsequences.
+%EM And do we want to define that j is some meta-index for all triples?
     J(i,x,y) = \{ (\mu^{(j)},u^{(j)},v^{(j)}) \in \calT \st x \xrightarrow{i,u^{(j)},v^{(j)}} y \}.
 \]
-The rate $G(n)_{x,y}$ is then the sum of all matching transition rates,
-multiplied by the potential difference,
+The rate ${G(n)}_{x,y}$ is then the sum of all matching transition rates,
+multiplied by the fitness term,
 namely
 \begin{align} \label{eqn:G_defn}
-    G(n)_{x,y} = \phi(E(y)-E(x)) \sum_{i=1}^n \sum_{J(i,x,y)}  \mu^j ,
-\end{align}
-and if there are no triples $(\mu,u,v)$ with $x \xrightarrow{i,u,v} y$ for some $i$, then $G(n)_{x,y}=0$.
+%EM Are we happy with leaving the \sum_{J(i,x,y)} a little vague here? I think it's "j" for which there is a triple?
+    {G(n)}_{x,y} = \phi\left(E(y)-E(x)\right) \sum_{i=1}^n \sum_{J(i,x,y)}  \mu^j ,
+\end{align}
+and if there are no triples $(\mu,u,v)$ with $x \xrightarrow{i,u,v} y$ for some $i$, then ${G(n)}_{x,y}=0$.
+%EM Do we need this following sentence? If so, we need to update it to also say something about the fitness term.
 This formalizes the statement that the transition rate from $x$ to $y$ is the sum of the rates of all transition triples
@@
 \begin{align} \label{eqn:full_likelihood}
-    p_n(t;x,y) := \P\{ X(t) = y \mid X(0) = x \} = \left(e^{tG(n)}\right)_{xy} ,
+    p_n(t;x,y) := \P\{ X(t) = y \mid X(0) = x \} = {\left(e^{tG(n)}\right)}_{xy} ,
 \end{align}
 which would then provide a route to parameter estimation.
-In practice, the size of these matrices makes direct application obviously impractical.
+%EM Here or somewhere else do we need to say something about setting the overall rate of the process? In phylogenetics it is common to scale the rate matrix so that substitutions happen with rate 1.
+In practice, the size of these matrices makes direct application obviously impractical for anything but very small $n$.
 This paper proposes an alternate strategy.

### 2021-10-11 writeups/context-dependence.tex
The nit-picking will not stop until morale improves

@@
   compute the likelihood of a given configuration change,
-  and so estimate the speed by maximum likelihood.
+  and so estimate the speed by maximum likelihood
+  given an initial sequence, a final sequence and an elapsed time.
 
@@
 %EM I think we want h and not j here in the parenthetical superscript.
-%EM Perhaps we want \mu^{j} rather than \mu^{(j)}? I think the latter is more for subsequences.
+%EM Perhaps we want \mu^{j} rather than \mu^{(j)}? I think it could be clearer to keep the latter just for subsequences. How about lower-j?
 %EM And do we want to define that j is some meta-index for all triples?
@@
 or ``uniformization'' \citep{hobolth2009simulation}.
+%EM I note that here we are re-using the "possible change" terminology, which we used above for something that seems slightly different, namely the fitness setup.
+%EM It seems like either we should fit these possible changes into that framework, or call these something else.
 Briefly, this  works by first sampling a homogeneous Poisson process of times and locations of possible changes,
@@
 \[
+    %EM If we change anything about j and J above, propagate that change here.
     \mu^* = \max \left\{ \sum_{y \in \calS^n} \sum_{j\in J(i,x,y)} \mu^{(j)} \qquad \st x \in \calS^n, \quad 1 \le i \le n \right\} .
@@
 Now sample possible changes:
+%EM Do we want to have sequence length become L here? It has been $n$ so far.
 in a sequence of length $L$, over a time period $t$,
@@
 q_j = \begin{cases}
+    %EM Perhaps we could look at this notation together? I would think that u^{(j)} is the length-j subsequence of u, but I don't think that's what's desired here.
     \mu^{(j)}/\mu^* \qquad & \text{if } x_i^{(|u^{(j)}|)} = u^{(j)}  \\

### 2021-10-12 writeups/context-dependence.tex
Finishing pass through section 3

@@
 
-The problem at hand is to infer the parameters of the model based on the final state of the model after it hsa evolved from some known initial state.
+The problem at hand is to infer the parameters of the model based on the final state of the model after it has evolved from some known initial state.
 % If the sequence is long enough and not too much time has gone by
@@
 so the full likelihood function is given by \eqref{eqn:full_likelihood},
-but doing anything with the $|\calS|^n \times |\calS|^n$ matrix $G_n$ is clearly infeasible for even moderately sized $n$.
+but doing anything with the $|\calS|^n \times |\calS|^n$ matrix $G_n$ is clearly infeasible even for moderately sized $n$.
 We can, however, compute \eqref{eqn:full_likelihood} for smaller $n$,
@@
 
+%EM I suggest that we have something like (B) in the introduction.
+%EM We need to make it clear that the window width of the underlying model, and the T-mer width, are two different things.
+%EM Yes, this is in the last paragraph, but I think this point gets lost when we start with "a general context-dependent model."
 \begin{figure}
@@
 is later observed to have its center matching $y$
-is close to $p_{n,\ell,m}(x,y)$, regardless of wider context.
+is close to $p_{n,\ell,r}(x,y)$, regardless of wider context.
 Formally, for each $i$,
@@
 
-%EM I feel like it needs emphasis that this is different than saying that we assume that the rates stay the same as in the initial sequence.
-%PR I'm not sure what you mean here?
-
 %%%%%%
@@
 \begin{align*}
-    \like(y|x) &= \prod_i \P\{ Y_i = y_i \given Y_{<i} = y_{<i} \;\text{and}\; X=x \} .
+    \like(y|x) &= \prod_i \P\{ Y_i = y_i \given Y_{<i} = y_{<i} \ \text{and}\; X=x \} .
 \end{align*}
@@
 by discarding approximately independent sites from the conditioning.
-If $\ell$ is long enough that approximation \eqref{eqn:window_approx} holds, then:
+These approximately independent sites are exactly those that are too far away from a given site to impact its evolution.
+Recall that the last base of $x_i^{(h)}$ is $x_{i+h-1}$.
+So, $x_{i-3\ell}^{(4\ell+1)}$ runs from $i-3\ell$ to $i + \ell$, inclusive, while $y_{i-2\ell}^{(2\ell+1)}$ runs from $i-2\ell$ to $i$, inclusive.
+If $\ell$ is long enough that approximation~\eqref{eqn:window_approx} holds, then:
 \begin{align} \label{eqn:full_approx}
     \begin{split}
-    \P\{ Y_i = y_i \given Y_{<i} \;\text{and}\; X \}
+    \P\{ Y_i = y_i \given Y_{<i} \ \text{and}\; X=x \}
     &\approx
@@
 By rearranging the product,
-this gives us an expression for the full likelihood as the ratio of the likelihoods of all $(n,\ell,\ell)$ T-mers in $(x,y)$
-to that of of all $(n,\ell,\ell+1)$ T-mers.
+this gives us an approximation for the full likelihood as the ratio of the likelihoods of all $(2\ell+1,\ell,\ell)$ T-mers in $(x,y)$
+to that of of all $(2\ell+1,\ell,\ell+1)$ T-mers.
 % Define $n_\ell(x,y;a,b)$ to be the number of times pattern $a$ is seen in sequence $x$
@@
 and the next section for some intuition.
+%EM I know that it's all written out in the appendix, but I think that having a little more development here would be nice.
+%EM Can we re-instate the material just above, that's commented out?
+%EM Also, what about the edge cases? When you say "all the t-mers", I think that includes t-mers where some of the "arms" are sticking off the sequence? Or are we making the sequence into a circle?
 
@@
 
-The approximations in \eqref{eqn:window_approx} and \eqref{eqn:full_approx} hold
+The approximations in~\eqref{eqn:window_approx} and~\eqref{eqn:full_approx} hold
 if the overhang, $\ell$, is long enough.
-The intuition is straightforward, and provides a way to pick the overhang lengths $\ell$ and $r$;
-a precise statement is given in Appendix \ref{ss:approx_pf}.
+We present intuition about these formulas here, which provides a way to pick the overhang lengths $\ell$ and $r$;
+a precise statement is given in Appendix~\ref{ss:approx_pf}.
 Suppose that each instantaneous change depends on positions at most $w$ sites away,

### 2021-10-13 writeups/context-dependence.tex
Finishing pass through main text

@@
 and compare these to a standard Normal distribution.
+%EM I think n is used here in a different sense than just above.
 However, there are are $n$ possible such tables --
@@
 
+%EM This is beautifully simple but I'm not used to seeing p-values as a means of doing model selection (this is my lacking stats background showing up again).
+%EM Can you point me to other examples where this is done?
+
 %%%%%%%%%%%
@@
 
+%EM This needs some notation updating. I think we want to commit to the n, \ell, r notation just above? I can do this.
 For inference, we need to compute the $|\calS|^{m+2\ell} \times |\calS|^m$ matrix $F$ whose $(x,y)^\text{th}$ entry is $p_{m,\ell}(t,x,y)$,
@@
 
+%EM Is it worth noting that (IIUC) we could have additional sparsity based on the underlying (not T-mer) model?
+E.g.
+when you are are doing stepwise addition of parts to your model, you aren't using all possible 2m+1 - mers?
 Using the R package expm \citep{R_expm}, this makes computation quite feasible:
@@
 (and much longer for the whole matrix $e^{t G(8)}$).
+
+%EM What if we want to do derivative-based optimization of t? Then we'll have sparse times exp(sparse) times U. Still seems fast?
 
@@
 
+%EM I skipped for now.
 In phylogenetic applications, rather than ``before'' and ``after'' observations,
@@
 
+%EM Peeling vs pruning?
 The general case is a modification of Felsenstein's ``peeling'' algorithm.
@@
 i.e., only computing the likelihood for cases with $y=z=w$.
-On a three-taxon tree it is not clear that this is desireable, since missing out on double transitions
+On a three-taxon tree it is not clear that this is desirable, since missing out on double transitions
 might be a significant loss of information,
@@
 We repeated this procedure separately on 60 independently simulated sequences,
-and show the resulting credible intervals in Figure \ref{fig:ising_coverage}.
-Poseterior medians are within half a percent of the truth for $+ \to -$ and $- \to +$ (corresponding to $\lambda$)
+and show the resulting credible intervals in Figure~\ref{fig:ising_coverage}.
+
+Posterior medians were within half a percent of the truth for $+ \to -$ and $- \to +$ (corresponding to $\lambda$)
 and the temperature ($\beta$), but showed around 2\% error for the magnetic field $\gamma$.
-As seen in Table \ref{tab:ising_coverage},
+As seen in Table~\ref{tab:ising_coverage},
+%EM I think this credible interval statement is interesting. So it's definitely this narrowness vs just having the estimate be off?
 credible intervals were somewhat too narrow for shorter T-mers,
@@
 \paragraph{Limitations}
+%EM This is true for the implementation, but it's not clear that we couldn't do some of the same tricks for calculating likelihoods if we were handed a model.
 The model presented assumes that the same process occurs at every location along the sequence,
@@
 For instance, inserting the 13-base pair motif selected against by meiotic drive in primates \citep{myers2010drive}
-would require T-mers of width at least 25, and hence vectors of length $4^25 \approx 1.13 \times 10^15$.
+would require T-mers of width at least 25, and hence vectors of length $4^{25} \approx 1.13 \times 10^{15}$.
 
@@
 otherwise, initial investigations suggest the likelihood surface is often strongly ridged --
-for instance, increasing the abundance of \nG nucleotides at the root while also increasing the mutation rate away from \nG
+for instance, increasing the abundance of \nG\ nucleotides at the root while also increasing the mutation rate away from \nG\
 can achieve roughly the same result.

### 2021-10-13 writeups/context-dependence.tex
Finishing pass to the full likelihood

@@
 affect the mutation rate of the central site has $w=1$.)
+%EM I suggest that we could re-introduce the mu^* notation again and remind people that these potential changes are uniform.
 Recall that in our construction of the process we first chose the times and sites
@@
 The number of such potential changes is Poisson with mean $\mu^* w t$,
+%EM Suggest "and their location is uniform..."
 and will extend the amount of sequence on which the change at site $i$ depends
@@
 on which a mutation at that site depends
+%EM Do we care that N is used for different things elsewhere?
 is bounded by the compound Poisson process $\sum_{k=1}^N W_k$,
@@
 So, if $A$ is the event that the final subsequence $y$
-depends on seequence outside of $x$ due to potential changes,
+depends on sequence outside of $x$ due to potential changes,
 then

### 2021-10-15 writeups/context-dependence.tex
a little fix and a hint

@@
     then $f(n) = 0$ for integers $0 \le n \le k$ and $f(n) \ge 1$ for $n > k$,
-    and so $\P\{N > k\} \le \E[f(N)] = \lambda^{k+1} / k!$.
+    and so $\P\{N > k\} \le \E[f(N)]$.
+    Now, $\E[f(N)] = \lambda^{k+1} / (k+1)!$ via direct calculation or the
+    formula for the $k+1$st factorial moment for the Poisson distribution.
 \end{proof}

### 2021-10-19 writeups/context-dependence.tex
After-meeting pass

@@
 
+% TODO after submission: think about the BCR parameter regime.
+
 % Model-based inference neighborhood dependence
@@
 
+%EM My suggestion is "Inference using context-dependent models of mutation via bounds on interaction propagation"
 \title{Inference using context-dependent models of mutation}
@@
 The conditioning method we consider here
-is simliar to the ``coding'' scheme introduced by \citet{besag1974spatial},
+is similar to the ``coding'' scheme introduced by \citet{besag1974spatial},
 that conditions on a set of sites that makes the remaining observations independent (thanks to the Gibbs property);
@@
 For instance, \citet{arndt2003sequence} studied dinucleotide transitions,
-%EM suggest going back to http://www.sciencedirect.com/science/article/pii/016747819290134L
 and \citet{yaari2013models} displayed strong heterogeneity in the probability of synonymous mutations across 5-mers
-in B-cell immunoglobin genes.
+in B-cell immunoglobulin genes.
 Others have made progress using simplifying assumptions \citep{berard2012accurate},
@@
 
+\begin{figure}
+    \begin{center}
+        \includegraphics{overview}
+    \end{center}
+    \caption{
+        An overview of the method for a neighbor-dependent model.
+        The red triangle represents the ``range of influence'' of mutations on a site of interest (the site at the bottom corner of the triangle).
+        The blue T shape is called a \emph{T-mer}, which is how our approximation to the process is parameterized.
+        The bottom is called the \emph{base} of the T-mer, which is considered the outcome of a probabilistic process parameterized by additional \emph{overhang} sequence extending to the left and right of the base.%
+        \label{fig:overview}
+    }
+\end{figure}
+
+Before diving into a formal problem specification, we offer an intuitive explanation of how the method works.
+The difficulty of context-sensitive models is that, in principle, one may have a sequence of events with cascading consequences that involves a lot of sequence.
+In the extreme case, one may have a sequence of mutations from, say, the left boundary of the sequence to the right boundary of the sequence, such that the probability of each mutation has been changed by that of the previous one.
+
+However, the point of the work presented here is that for typical models such a sequence of events is improbable.
+For that reason, we can perform computation using a local context which, although bigger than the context specified in the context-sensitive model, is much smaller than the entire sequence.
+We call the building blocks of this computation \emph{T-mers} (Figure~\ref{fig:overview}) which parameterize transition probabilities for a small section of sequence (the \emph{base} of the T-mer) based on additional sequence (the \emph{overhang} of the T-mer).
+We can bound the probability that the influence of a mutation outside the overhang of the T-mer will impact the sites at the base of the T-mer, which translates into an estimate of approximation error across the sequence when using a composite likelihood calculation based on T-mers.
+
 %%%%%%% %%%%%%%%%%
@@
 and $X_i(t)$ for the state of site $i$ at time $t$,
-we assume that $\{X(t)\}_{t \ge 0}$ is a Markov process on sequences of $n$ states $\calS^n$
+we assume that $\{X(t)\}_{t \ge 0}$ is a Markov process on sequences of $L$ states $\calS^L$
 for which the probability a given site changes state in a small amount of time
 depends only on the sequence of states at nearby sites.
-%EM If we think that this can be generalized to the setting of position-specific local models, then let's mention that here and then put in a little section.
 
@@
 first define a \textit{pattern} to mean a contiguous sequence of states;
-%EM Do we use this length notation? I couldn't easily find an instance.
 let $|u|$ denote the length of the pattern $u$.
@@
 
-%EM I think we could have both, but certainly this biological interpretation.
-%Then we would have "selection" and "fitness" rather than "potential" and
-%"energy"?
-%EM Yes. Seems nice unless you think that this "fitness" model would just annoy a pop gen person.
-\peter{Should call this ``selection'' instead of a ``potential''? Or what?}
-
 To allow more compact specification of models,
@@
 and that the probability a proposed change actually occurs depends on how much it affects the energy.
-%EM I'm trying to understand why I should think of e_i as an energy *change*. Why not just an energy of pattern i? We already take differences of energies.
-Concretely, let $\calP = \{(p_i, e_i)\}$ be a set of (pattern, energy change) pairs,
+
+Concretely, let $\calP = \{(p_i, e_i)\}$ be a set of (pattern, energy difference) pairs,
 and for any sequence $x$ define
@@
 but these changes only take effect with probability $\phi(E(y) - E(x))$.
+We have described this idea in terms natural for a physics model such as an Ising model in which high energy configurations are disfavored, however it is also suitable for a population genetics model in which $\phi$ describes the probability of a mutant allele given a certain fitness change from the parent.
 
@@
     GC-biased gene conversion effectively means that sequences with more G/Cs are selected against.
-    This is incorporated as follows: set
-    %EM Is this table clearer than just saying "set the energy of pattern C or G to be $s$...."
-    \begin{center}
-        \begin{tabular}{cc}
-        pattern  &  energy \\
-        \nC{} or \nG   &  $s$
-      \end{tabular} ,
-    \end{center}
+    This is incorporated as follows: set the energy of pattern of pattern \nC{} or \nG{} to be $s$,
     so that $E(y)-E(x)$ is the net change in number of G's and C's multiplied by $s$.
-    Also, let
+    Also, assume a $\phi$ function expressed in terms of an energy difference $e$:
     \begin{align*}
-    %EM Does it make more sense to have the N here in the selection component, or in the mutation component with \mu?
-    %EM Also, could we stick with $s$ for this equation? I don't see a compelling reason to switch to e, which doesn't appear outside of the context of this equation as far as I can see.
-        \phi(e) = N (1-\exp(-2e))/(1-\exp(-2eN))
+        \phi(e) = (1-\exp(-2e))/(1-\exp(-2eN)).
     \end{align*}
-    so that if a given change that occurs with rate $\mu$ would change a sequence $x$ to $y$,
-    then $\mu \, \phi(E(y)-E(x))$ is the rate at which the mutation appears and successfully takes over in the population.
+    Thus, if a given change that occurs with rate $\mu$ would change a sequence $x$ to $y$,
+    then $\mu \, N \phi(E(y)-E(x))$ is the rate at which the mutation
+    appears and successfully takes over in the population.
 
@@
     In the Ising model, each site is labeled as either ``up'' or ``down'' ($+1$ or $-1$ respectively),
-    imagined as magnetic dipoles,
-    %EM re i+1 in the following equation: as soon as I read this I'm already wondering what happens at the end of the sequence. Could we specify that here? Are we doing a loop?
-    and the energy associated with a given state $x$ is $E(x) = - \frac{1}{2} \beta \sum_i x_i x_{i+1} - \frac{1}{2} \gamma \sum_i x_i$.
+    imagined as a string of $n$ magnetic dipoles,
+    and the energy associated with a given state $x$ is $E(x) = - \frac{1}{2} \beta \sum_{i=1}^{n-1} x_i x_{i+1} - \frac{1}{2} \gamma \sum_{i=1}^n x_i$.
     Here $\beta$ represents inverse temperature, and $\gamma$ represents the strength of the magnetic field (here scaled by temperature).
@@
 \[
-%EM I think we want h and not j here in the parenthetical superscript.
-%EM Perhaps we want \mu^{j} rather than \mu^{(j)}? I think it could be clearer to keep the latter just for subsequences. How about lower-j?
-%EM And do we want to define that j is some meta-index for all triples?
-    J(i,x,y) = \{ (\mu^{(j)},u^{(j)},v^{(j)}) \in \calT \st x \xrightarrow{i,u^{(j)},v^{(j)}} y \}.
+    J(i,x,y) = \{ (\mu,u,v) \in \calT \st x \xrightarrow{i,u,v} y \}.
 \]
@@
 \begin{align} \label{eqn:G_defn}
-%EM Are we happy with leaving the \sum_{J(i,x,y)} a little vague here? I think it's "j" for which there is a triple?
-    {G(n)}_{x,y} = \phi\left(E(y)-E(x)\right) \sum_{i=1}^n \sum_{J(i,x,y)}  \mu^j ,
+    {G(n)}_{x,y} = \phi\left(E(y)-E(x)\right) \, \sum_{i=1}^n \sum_{(\mu, u, v) \in J(i,x,y)}  \mu ,
 \end{align}
 and if there are no triples $(\mu,u,v)$ with $x \xrightarrow{i,u,v} y$ for some $i$, then ${G(n)}_{x,y}=0$.
-%EM Do we need this following sentence? If so, we need to update it to also say something about the fitness term.
-This formalizes the statement that the transition rate from $x$ to $y$ is the sum of the rates of all transition triples
-that would change $x$ into $y$.
 In principle, this gives us the transition probabilities for the process by a matrix exponential:
@@
 which would then provide a route to parameter estimation.
-%EM Here or somewhere else do we need to say something about setting the overall rate of the process? In phylogenetics it is common to scale the rate matrix so that substitutions happen with rate 1.
 In practice, the size of these matrices makes direct application obviously impractical for anything but very small $n$.
@@
 or ``uniformization'' \citep{hobolth2009simulation}.
-%EM I note that here we are re-using the "possible change" terminology, which we used above for something that seems slightly different, namely the fitness setup.
-%EM It seems like either we should fit these possible changes into that framework, or call these something else.
 Briefly, this  works by first sampling a homogeneous Poisson process of times and locations of possible changes,
@@
 and then resolving each possible change in temporal order.
+(Note that these ``possible changes'' are different than the ``proposed changes'' described above
+in terms of the energy model.)
 The rate $\mu^*$ should be the maximum ``local'' rate at which transitions occur, across sites and transition outcome states:
 \[
-    %EM If we change anything about j and J above, propagate that change here.
-    \mu^* = \max \left\{ \sum_{y \in \calS^n} \sum_{j\in J(i,x,y)} \mu^{(j)} \qquad \st x \in \calS^n, \quad 1 \le i \le n \right\} .
+    %EM I think this n is supposed to be L.
+    \mu^* = \max \left\{ \sum_{y \in \calS^n} \sum_{(\mu, u, v) \in J(i,x,y)} \mu \qquad \st x \in \calS^n, \quad 1 \le i \le n \right\} .
 \]
 Now sample possible changes:
-%EM Do we want to have sequence length become L here? It has been $n$ so far.
 in a sequence of length $L$, over a time period $t$,
@@
 q_j = \begin{cases}
-    %EM Perhaps we could look at this notation together? I would think that u^{(j)} is the length-j subsequence of u, but I don't think that's what's desired here.
+    % TODO peter, fix notation, please.
     \mu^{(j)}/\mu^* \qquad & \text{if } x_i^{(|u^{(j)}|)} = u^{(j)}  \\
@@
 
-%EM I suggest that we have something like (B) in the introduction.
-%EM We need to make it clear that the window width of the underlying model, and the T-mer width, are two different things.
-%EM Yes, this is in the last paragraph, but I think this point gets lost when we start with "a general context-dependent model."
 \begin{figure}
@@
 %PR No, I think that'll muddy the waters.
+%EM My counter-argument is that we need to make it clear that the overhang of the T-mers is different than than the context-sensitivity width.
 This quantity is found by marginalizing $p_m$,
@@
 In this way, the influence of the sequence at each site extends outwards, as shown in Figure~\ref{fig:Tmers} --
-so if we take the overhangs long enough that this dependency is unlikely to extend to the base of the Tmer,
+so if we take the overhangs long enough that this dependency is unlikely to extend to the base of the T-mer,
 the approximation will be good.

### 2021-10-25 writeups/context-dependence.tex
Notes from meeting

@@
 
-%EM My suggestion is "Inference using context-dependent models of mutation via bounds on interaction propagation"
-\title{Inference using context-dependent models of mutation}
+\title{Inference using context-dependent models of mutation via bounds on information propagation}
 \author{}
@@
     A new mutation with an effect $s$ on fitness that occurs
-    in a population of effective size $N$ becomes ubiquitous, rather than dying out,
-    with probability approximately $(1-\exp(-2 s))/(1-\exp(-2 s N))$ \citep{kimura1962probability}.
-    If the mutation occurs at rate $\mu$ per individual, the total rate it occurs at in the population is $\mu N$.
+    in a population of effective size $N_e$ becomes ubiquitous, rather than dying out,
+    with probability approximately $(1-\exp(-2 s))/(1-\exp(-2 s N_e))$ \citep{kimura1962probability}.
+    If the mutation occurs at rate $\mu$ per individual, the total rate it occurs at in the population is $\mu N_e$.
     GC-biased gene conversion effectively means that sequences with more G/Cs are selected against.
@@
     Thus, if a given change that occurs with rate $\mu$ would change a sequence $x$ to $y$,
-    then $\mu \, N \phi(E(y)-E(x))$ is the rate at which the mutation
+    then $\mu \, N_e \phi(E(y)-E(x))$ is the rate at which the mutation
     appears and successfully takes over in the population.
@@
     %EM I think this n is supposed to be L.
+    %EM TODO fix
     \mu^* = \max \left\{ \sum_{y \in \calS^n} \sum_{(\mu, u, v) \in J(i,x,y)} \mu \qquad \st x \in \calS^n, \quad 1 \le i \le n \right\} .
@@
 and the next section for some intuition.
-%EM I know that it's all written out in the appendix, but I think that having a little more development here would be nice.
-%EM Can we re-instate the material just above, that's commented out?
+%EM Intro notation for the big product.
 %EM Also, what about the edge cases? When you say "all the t-mers", I think that includes t-mers where some of the "arms" are sticking off the sequence? Or are we making the sequence into a circle?
@@
 These \emph{residual} counts are usually most usefully computed for T-mers shorter than those used to fit the model.
+%EM Let's use a and b
 For each T-mer $(x,y)$, write $O_{x,y}$ for the number of this T-mer observed in the data,
@@
 and compare these to a standard Normal distribution.
-%EM I think n is used here in a different sense than just above.
 However, there are are $n$ possible such tables --
@@
 
-%EM This is beautifully simple but I'm not used to seeing p-values as a means of doing model selection (this is my lacking stats background showing up again).
-%EM Can you point me to other examples where this is done?
+%EM This is stepwise model addition.
 
@@
 when you are are doing stepwise addition of parts to your model, you aren't using all possible 2m+1 - mers?
+%EM E think
 Using the R package expm \citep{R_expm}, this makes computation quite feasible:
@@
 %EM What if we want to do derivative-based optimization of t? Then we'll have sparse times exp(sparse) times U. Still seems fast?
+%EM Add simple statement
 
@@
 on which a mutation at that site depends
-%EM Do we care that N is used for different things elsewhere?
 is bounded by the compound Poisson process $\sum_{k=1}^N W_k$,

### 2021-10-30 writeups/context-dependence.tex
Making a pass through; part-way through phylo.

@@
 or of several states evolved along a tree from a single starting point.
+This paper concerns scalable methods for doing inference under such models using observations.
 
@@
 Before diving into a formal problem specification, we offer an intuitive explanation of how the method works.
-The difficulty of context-sensitive models is that, in principle, one may have a sequence of events with cascading consequences that involves a lot of sequence.
+The difficulty of context-sensitive models is that, in principle, one may have a sequence of events with cascading consequences that involves a lot of the sequence.
 In the extreme case, one may have a sequence of mutations from, say, the left boundary of the sequence to the right boundary of the sequence, such that the probability of each mutation has been changed by that of the previous one.
 
-However, the point of the work presented here is that for typical models such a sequence of events is improbable.
+However, the starting point of the work presented here is that for typical models such a sequence of events is improbable.
 For that reason, we can perform computation using a local context which, although bigger than the context specified in the context-sensitive model, is much smaller than the entire sequence.
@@
 and if there are no triples $(\mu,u,v)$ with $x \xrightarrow{i,u,v} y$ for some $i$, then ${G(n)}_{x,y}=0$.
+Indeed, we anticipate this matrix to be sparse because for most pairs $(x,y)$ there are no transition triples that can modify $x$ to make $y$.
+
 In principle, this gives us the transition probabilities for the process by a matrix exponential:
@@
 \[
-    %EM I think this n is supposed to be L.
-    %EM TODO fix
-    \mu^* = \max \left\{ \sum_{y \in \calS^n} \sum_{(\mu, u, v) \in J(i,x,y)} \mu \qquad \st x \in \calS^n, \quad 1 \le i \le n \right\} .
+    \mu^* = \max \left\{ \sum_{y \in \calS^L} \sum_{(\mu, u, v) \in J(i,x,y)} \mu \qquad \st x \in \calS^L, \quad 1 \le i \le L \right\} .
 \]
@@
 to an arbitrarily good approximation.
+%EM do we really want x and y here, which I've been thinking of as being the full sequence?
+%EM Note we'll need to change a and b below.
 To do this, let $p_{n,\ell,r}(x,y)$ for $\ell + r < n$ denote
@@
 This central, smaller subsequence is called the ``base'' of the T-mer.
-We refer to such patterns as ``T-mers'', as depicted in Figure~\ref{fig:Tmers}.
-%EM Shall we define w here?
-%PR No, I think that'll muddy the waters.
-%EM My counter-argument is that we need to make it clear that the overhang of the T-mers is different than than the context-sensitivity width.
-This quantity is found by marginalizing $p_m$,
+We refer to such patterns as \emph{T-mers}, as depicted in Figure~\ref{fig:Tmers}.
+We emphasize that the overhang of the T-mers is typically bigger than the
+context-sensitivity width of the underlying model (Figure~\ref{fig:overview}).
+
+These $p_{n,\ell,r}(x,y)$ probabilities can be found by marginalizing $p_m$,
 summing over the possible inital and final substrings of $x$
@@
     \text{for}\; x \in \calS^n, \; y \in \calS^{n-\ell-r} ,
+    \label{eq:pmMarg}
 \end{align}
@@
 \begin{align*}
-    \like(y|x) &= \prod_i \P\{ Y_i = y_i \given Y_{<i} = y_{<i} \ \text{and}\; X=x \} .
+    \like(y \mid x) &= \prod_i \P\{ Y_i = y_i \given Y_{<i} = y_{<i} \ \text{and}\; X=x \} .
 \end{align*}
@@
             p_{2\ell+1,\ell,\ell+1}(x_{i-3\ell}^{(4\ell+1)}, y_{i-2\ell}^{(2\ell)})
-        }
-    \end{split} .
-\end{align}
-By rearranging the product,
-this gives us an approximation for the full likelihood as the ratio of the likelihoods of all $(2\ell+1,\ell,\ell)$ T-mers in $(x,y)$
-to that of of all $(2\ell+1,\ell,\ell+1)$ T-mers.
+        }.
+    \end{split}
+\end{align}
+
+%EM I think you said that we could just treat the state of all positions before the beginning of the sequence and after the end as having another distinct state. If that's true I can write it out.
+
+Let $A_{n, \ell, r}(x,y)$ be the product of all probability of all $(n, \ell, r)$ T-mers transitioning from $x$ to $y$, i.e.\
+\[
+A_{n, \ell, r}(x,y) = \prod_{i=1}^L p_{n, \ell, r}\left(x_{i-\ell-m}^{(n)}, y_{i-m}^{(n-\ell-r)}\right).
+\]
+
+We can then express our likelihood as
+
+\[
+    \like(y \mid x) = \frac{ A_{2\ell+1, \ell, \ell}(x,y) }{ A_{2\ell+1, \ell, \ell+1}(x,y) }.
+\]
+
 % Define $n_\ell(x,y;a,b)$ to be the number of times pattern $a$ is seen in sequence $x$
@@
 and the next section for some intuition.
-%EM Intro notation for the big product.
-%EM Also, what about the edge cases? When you say "all the t-mers", I think that includes t-mers where some of the "arms" are sticking off the sequence? Or are we making the sequence into a circle?
 
@@
 These \emph{residual} counts are usually most usefully computed for T-mers shorter than those used to fit the model.
-%EM Let's use a and b
-For each T-mer $(x,y)$, write $O_{x,y}$ for the number of this T-mer observed in the data,
-and $E_{x,y}$ for the number expected under the model.
-(The expected number is the number of occurrences of $x$ in the initial sequence
-multiplied by the probability $p_{n,\ell,r}(t;x,y)$, where $n$ is the length of $x$,
+For each T-mer $(a,b)$, write $O_{a,b}$ for the number of this T-mer observed in the data,
+and $E_{a,b}$ for the number expected under the model.
+(The expected number is the number of occurrences of $a$ in the initial sequence
+multiplied by the probability $p_{n,\ell,r}(t;a,b)$, where $n$ is the length of $a$,
 and $\ell$, $r$ are the appropriate overhangs.)
@@
 \begin{align}
-    Z_{x,y} = \frac{ O_{x,y} - E_{x,y} }{ \sqrt{n E_{x,y}} },
+    Z_{a,b} = \frac{ O_{a,b} - E_{a,b} }{ \sqrt{n E_{a,b}} },
 \end{align}
@@
 
-%EM This needs some notation updating. I think we want to commit to the n, \ell, r notation just above? I can do this.
-For inference, we need to compute the $|\calS|^{m+2\ell} \times |\calS|^m$ matrix $F$ whose $(x,y)^\text{th}$ entry is $p_{m,\ell}(t,x,y)$,
+For inference, we need to compute the $|\calS|^{n} \times |\calS|^m$ matrix $F$ whose $(a,b)^\text{th}$ entry is $p_{n,\ell,r}(t,a,b)$,
 for various $t$.
-This matrix is a projection of the $|\calS|^{m+2\ell} \times |\calS|^{m+2\ell}$ matrix whose $(x,z)^\text{th}$ entry is
+This matrix is a projection of the $|\calS|^{n} \times |\calS|^{n}$ matrix whose $(a,c)^\text{th}$ entry is
 \begin{align}
-    p_{m+2\ell}(x,z) = \left( e^{t G(m+2\ell)} \right)_{x,z} ,
-\end{align}
-where $G(m+2\ell)$ is the sparse $|\calS|^{m+2\ell} \times |\calS|^{m+2\ell}$ matrix defined in \eqref{eqn:G_defn}.
-The ``projection'' we need just marginalizes over long patterns $z$ that match the shorter pattern $y$:
-if we define the matrix $U$ in this way, so that $U_{zy}=1$ if $z_\ell^m=y$ and $U_{zy}=0$ otherwise,
+    p_{n}(a,c) = \left( e^{t G(n)} \right)_{a,c} ,
+\end{align}
+where $G(n)$ is the sparse $|\calS|^{n} \times |\calS|^{n}$ matrix defined in \eqref{eqn:G_defn}.
+The ``projection'' we need just marginalizes over long (i.e.\ length $n$) patterns $c$ that match the shorter (i.e.\ length $m$) pattern $b$
+as in \eqref{eq:pmMarg}:
+if we define the $n \times m$ matrix $U$ so that $U_{cb}=1$ if $c_\ell^m=b$ and $U_{cb}=0$ otherwise,
 then
 \begin{align} \label{eqn:Tmer_trans}
-    p_{m,\ell}(t,x,y) = \sum_{z \in \calS^{m+2\ell}} \left( e^{t G(m+2\ell)} \right)_{xz} U_{zy} ,
-\end{align}
-so in fact we don't need the entire matrix $e^{t G(m+2\ell)}$,
+    p_{m,\ell}(t,a,b) = \sum_{c \in \calS^{n}} \left( e^{t G(n)} \right)_{ac} U_{cb} ,
+\end{align}
+so in fact we don't need the entire matrix $e^{t G(n)}$,
 just the product of this matrix with each of the columns of $U$.
@@
 
-%EM Is it worth noting that (IIUC) we could have additional sparsity based on the underlying (not T-mer) model?
-E.g.
-when you are are doing stepwise addition of parts to your model, you aren't using all possible 2m+1 - mers?
-%EM E think
 Using the R package expm \citep{R_expm}, this makes computation quite feasible:
-with four possible states and $\ell=2$ and $m=1$, so that $G(m+2\ell)$ is a $1024 \times 1024$ matrix,
+with four possible states and $\ell=2$ and $m=1$, so that $G(n)$ is a $1024 \times 1024$ matrix,
 computing $e^{t G(5) }$ takes 13 seconds, while computing $e^{t G(5)} U$ takes only 0.3 seconds.
+%EM In the above, don't we only provide the recipe for the full likelihood when m = 2 ell + 1?
 Increasing to $\ell=4$ and $m=1$ or $\ell=3$ and $m=2$ is still feasible, taking $e^{t G(8)} U$ in 42 and 47 seconds, respectively
@@
 
-%EM What if we want to do derivative-based optimization of t? Then we'll have sparse times exp(sparse) times U. Still seems fast?
-%EM Add simple statement
+Although we haven't used such tools here, taking the derivative with respect to $t$ can also be performed using sparse matrix operations.
 
@@
 under many sets of parameters.
-First note that $G(m+2\ell)$ has, assuming only single-position changes,
-$(1+m(|\calS|-1)) |\calS|^{m+2\ell}$ nonzero entries, since each of the $|\calS|^{m+2\ell}$ can change in $m$ places.
-This will determine how the computation scales with $m$, $\ell$, and $|\calS|$,
+Let's assume that the T-mers only allow single-position changes.
+First note that in this case $G(n)$ has
+$(1+m(|\calS|-1)) |\calS|^{n}$ nonzero entries, since each of the $|\calS|^{n}$ can change in $m$ places.
+This will determine how the computation scales with $m$, $n$, and $|\calS|$,
 once we precompute a number of things.
@@
 sparse matrix representations of $G$ store only $g$ along with information about the rows and columns these are found in.
+%EM Here we are using lower j, rather than our "new" notation setup, but it doesn't seem like a big deal. This is a local note.
 Each $g_i$ is a linear combination of mutation rates $\mu_j$,
@@
 
-%EM I skipped for now.
 In phylogenetic applications, rather than ``before'' and ``after'' observations,
+%EM Describe the root sequence, or a prob distribution on it?
+%EM We want to be able to talk about the probability of seeing a certain n-mer... is that obvious what that random process is on the whole-sequence level?
 we get two (or more) observations evolved from a common root.
@@
 using the pattern frequencies at the root.
-Concretely: pick a random location $I$ in the sequence and let $X_{I-\ell}^{m+2\ell}(0) = \rho$ be the (long) pattern seen there at the root,
-and write $\pi(z)$ for the frequency of pattern $z$ in $X(0)$, i.e., $\P\{\rho=z\}=\pi(z)$.
+%EM Could we make it \rho_I to emphasize that it's a random quantity?
+Concretely: pick a random location $I$ in the sequence and let $X_{I-\ell}^{n}(0) = \rho$ be the (long) pattern seen there at the root.
+Write $\pi(z)$ for the frequency of a given pattern $z$ in $X(0)$, i.e., $\P\{\rho=z\}=\pi(z)$.
 % Given the corresponding pattern in $X(t_X)$, the distribution of $\rho$ is proportional to
 % \[
-%     \P\{ \rho=z \mid X_{I-\ell}^{m+2\ell}(t_X)=x\} \propto \pi(z) p_{m+2\ell}(t,z,x) .
+%     \P\{ \rho=z \mid X_{I-\ell}^{n}(t_X)=x\} \propto \pi(z) p_{n}(t,z,x) .
 % \]
@@
 \begin{align} \label{eqn:phylo_likelihood}
-    \P\{X_{I-\ell}^{m+2\ell}(t_X)=x \text{ and } Y_I^\ell(t_Y)=y \} = \sum_{z \in \calS^{m+2\ell}} \pi(z) p_{m+2\ell}(t_X,z,x) p_{m,\ell}(t_Y,z,y) .
+    \P\{X_{I-\ell}^{n}(t_X)=x \text{ and } Y_I^\ell(t_Y)=y \} = \sum_{z \in \calS^{n}} \pi(z) \, p_{n,\ell,r}(t_X,z,x) \, p_{n,\ell,r}(t_Y,z,y) .
 \end{align}
@@
 given the frequencies at the root, as described in greater generality below.
-These calculations can be done using the the same sparse matrix methods as above.
+These calculations can be done using the same sparse matrix methods as above.
 For instance, to compute
 \[
-    \sum_z \pi(z) p_{m+2\ell}(t_X,z,x) p_{m,\ell}(t_Y,z,y)
-    = \sum_z \left( e^{t_X G(m+2\ell)} \right)_{zx} \pi(z) \sum_w \left( e^{t_X G(m+2\ell)} \right)_{zw} U_{wy} ,
+    %EM This is super clever!
+    \sum_{z \in \calS^{n}} \pi(z) \, p_{n,\ell,r}(t_X,z,x) \, p_{n,\ell,r}(t_Y,z,y)
+    = \sum_{z \in \calS^{n}} \left( e^{t_X G(n)} \right)_{zx} \pi(z) \sum_{w \in \calS^{n}} \left( e^{t_Y G(n)} \right)_{zw} U_{wy} ,
 \]
-we can first compute $\sum_w \left( e^{t_X G(m+2\ell)} \right)_{zw} U_{wy}$ as above,
-multiply rows by $\pi(z)$, and then matrix multiply by the transpose of $G(m+2\ell)$.
-
+we can first compute $\sum_w \left( e^{t_Y G(n)} \right)_{zw} U_{wy}$ as above,
+multiply rows by $\pi(z)$, and then matrix multiply by the transpose of $e^{t_X G(n)}$.
+
+%EM It seems like we'd like to write out how this works for the entire set of sequences, as you did for the "full likelihood" section.
+%EM I think we have a bunch of ratios as before, where r is extended by 1 in the numerator?
+
+%EM I think this should be pruning, not peeling.
+%EM Want to check the modifications made above before continuing those to here.
 \subsection{Phylogenetic peeling}
 
-%EM Peeling vs pruning?
 The general case is a modification of Felsenstein's ``peeling'' algorithm.
@@
 with the three-taxon tree of figure \ref{fig:peeling} as an example,
-where we are counting occurrences of (long) $m+2\ell$-tuples in taxon $X$,
+where we are counting occurrences of (long) $n$-tuples in taxon $X$,
 and (shorter) $m$-tuples in taxa $Y$ and $Z$.
@@
   \begin{split}
-    & \P\{X_{I-\ell}^{m+2\ell}=x \text{ and } Y_I^\ell=y \text{ and } Z_I^\ell=z \} \\
-    &\qquad = \sum_{u \in \calS^{m+2\ell}} \pi(u) p_{m,\ell}(t_C,u,z) \sum_{v \in \calS^{m+2\ell}} p_{m+2\ell}(t_{AB},u,v) p_{m+2\ell}(t_{A},v,x) p_{m,\ell}(t_B,v,y)
+    & \P\{X_{I-\ell}^{n}=x \text{ and } Y_I^\ell=y \text{ and } Z_I^\ell=z \} \\
+    &\qquad = \sum_{u \in \calS^{n}} \pi(u) p_{m,\ell}(t_C,u,z) \sum_{v \in \calS^{n}} p_{n}(t_{AB},u,v) p_{n}(t_{A},v,x) p_{m,\ell}(t_B,v,y)
   \end{split}
@@
 
-  \item Compute $M_1(u,z) = p_{m,\ell}(t_C,u,z)$ (a $|\calS|^{m+2\ell} \times |\calS|^m$ matrix)
-
-  \item Compute $M_2(v,z) = \sum_{u \in \calS^{m+2\ell}}  p_{m+2\ell}(t_{AB},u,v) \pi(u) M_1(u,z)$ (still a $|\calS|^{m+2\ell} \times |\calS|^m$ matrix)
-
-  \item Compute $M_3(v,y) = p_{m,\ell}(t_B,v,y)$ (a $|\calS|^{m+2\ell} \times |\calS|^m$ matrix)
-
-  \item Compute $M_4(x,(y,z)) = \sum_{v \in \calS^{m+2\ell}} p_{m+2\ell}(t_{A},v,x) M_2(v,z) M_3(v,y)$ (stored as a $|\calS|^{m+2\ell} \times |\calS|^{2m}$ matrix).
+  \item Compute $M_1(u,z) = p_{m,\ell}(t_C,u,z)$ (a $|\calS|^{n} \times |\calS|^m$ matrix)
+
+  \item Compute $M_2(v,z) = \sum_{u \in \calS^{n}}  p_{n}(t_{AB},u,v) \pi(u) M_1(u,z)$ (still a $|\calS|^{n} \times |\calS|^m$ matrix)
+
+  \item Compute $M_3(v,y) = p_{m,\ell}(t_B,v,y)$ (a $|\calS|^{n} \times |\calS|^m$ matrix)
+
+  \item Compute $M_4(x,(y,z)) = \sum_{v \in \calS^{n}} p_{n}(t_{A},v,x) M_2(v,z) M_3(v,y)$ (stored as a $|\calS|^{n} \times |\calS|^{2m}$ matrix).
 
@@
 Then $M_4$ gives all the likelihoods.
-This is depicted in the figure, writing $\left(e^{tG}\right)_{ab}$ for the matrix $p_{m+2\ell}(t,a,b)$,
+This is depicted in the figure, writing $\left(e^{tG}\right)_{ab}$ for the matrix $p_{n}(t,a,b)$,
 and $P$ for the projection matrix that gives $p_{m,\ell}(t,a,b) = \left( e^{tG} P\right)_{ab}$.
@@
 
-  \item[4'.] Compute $M_4(x,w) = \sum_{v \in \calS^{m+2\ell}} p_{m+2\ell}(t_{A},v,x) M_3(v,w) M_4(v,w)$ (stored as a $|\calS|^{m+2\ell} \times |\calS|^{m}$ matrix).
+  \item[4'.] Compute $M_4(x,w) = \sum_{v \in \calS^{n}} p_{n}(t_{A},v,x) M_3(v,w) M_4(v,w)$ (stored as a $|\calS|^{n} \times |\calS|^{m}$ matrix).
 
@@
 \paragraph{Limitations}
-%EM This is true for the implementation, but it's not clear that we couldn't do some of the same tricks for calculating likelihoods if we were handed a model.
 The model presented assumes that the same process occurs at every location along the sequence,
@@
 on all tips further away from $a$ (i.e., below that node if it was rooted at $a$).
-For a node $w$ this will be stored as a $|\calS|^{m+2\ell} \times |\calS|^{cm}$ matrix, for some $c$,
+For a node $w$ this will be stored as a $|\calS|^{n} \times |\calS|^{cm}$ matrix, for some $c$,
 and denoted $M(w)$.

### 2021-10-31 writeups/context-dependence.tex
working through the approx proof

@@
 \begin{align} \label{eqn:Tmer_trans}
-    p_{m,\ell}(t,a,b) = \sum_{c \in \calS^{n}} \left( e^{t G(n)} \right)_{ac} U_{cb} ,
+    p_{n,\ell,r}(t,a,b) = \sum_{c \in \calS^{n}} \left( e^{t G(n)} \right)_{ac} U_{cb} ,
 \end{align}
@@
 %EM I suggest that we could re-introduce the mu^* notation again and remind people that these potential changes are uniform.
-Recall that in our construction of the process we first chose the times and sites
-at which potential changes might occur.
-Suppose a potential change occurs at site $i-w$, at time $t$,
+
+As in the ``Simulation'' section, we start by considering a broader collection
+of ``possible changes'' sampled from a homogeneous
+Poisson process at a rate $\mu^*$ per site.
+Suppose a possible change occurs at site $i-w$, at time $t$,
+%EM should this be "change in rate"?
 which therefore might result in a change in state at site $i$.
@@
 thus extending the sequence on which the change depends to the right.
-The number of such potential changes is Poisson with mean $\mu^* w t$,
-%EM Suggest "and their location is uniform..."
-and will extend the amount of sequence on which the change at site $i$ depends
+The number of such potential changes have a uniform location
+and their count is Poisson with mean $\mu^* w t$;
+these extend the amount of sequence on which the change at site $i$ depends
+%EM my understanding is that it's just uniform for a single mutation, but that's not obvious from the description.
 by a uniform random number of sites between 1 and $w$.
@@
 \begin{lemma}
+%EM Let's rename this X to ... V? We already have an X.
     Let $X = \sum_{j=1}^N W_j$,
@@
 For two positions $(i,s)$ and $(j,t)$ in sequence $\times$ time,
-let $r(i,s;j,t) = |j-i|/|t-s|$ be the slope of the line between them,
+let $r(i,s;j,t) = |j-i|/|t-s|$ be the absolute value of the
+slope of the line between them,
 and if $I$ and $J$ are subsets of sites, let $r(I,s;J,t) = \min\{r(i,s;j,t) : i\in I, j \in J\}$
@@
 \begin{align}
-%EM Before we had i,s,j,t, and now we have i,t,j,s. It doesn't make a difference, but should we choose one?
-%EM Do we want to specify a max allowable range for j as well on the RHS?
-    N_i = \{ (j,s) : r(i,t; j,s) \le r_*, \; 0 \le s \le t \} .
+    N_i = \{ (j,s) : r(i,s; j,t) \le r_*, \; j \leq N, \; 0 \le s \le t \} .
 \end{align}
@@
 and $i \nsim j$ otherwise.
-We can think of $N_i$ as the parts of ancestral sequences in which a mutation might have affected the state at $(i,t)$,.
-so if $i \nsim j$, then $i$ and $j$ are (approximately) independent at time $t$, given the sequence at time 0.
+We can think of $N_i$ as the parts of ancestral sequences in which a mutation might have affected the state at $(i,t)$.
+So, if $i \nsim j$, then $i$ and $j$ are (approximately) independent at time $t$, given the sequence at time 0.
 Concretely,
@@
 \end{align}
+
+%EM Would it be more direct to make a connection between
+%EM \P\{ X_i(t) = y_i \given X_{A(L(i))}(0) = x_{A(L(i))} \text{ and } X_{L(i)}(t) = y_{L(i)} \} ,
+%EM and
+%EM p_{2\ell+1,\ell,\ell}(x_{i-3\ell}^{(4\ell+1)}, y_{i-2\ell}^{(2\ell+1)})
 

### 2021-11-19 writeups/context-dependence.tex
introduction pass

@@
 
-It now common to encounter models
-consisting of a lattice of sites,
-each site taking one of a finite collection of possible states,
+Early models of DNA sequence mutation
+assumed that each nucleotide was equally likely to mutate
+into any other nucleotide \citep{jukes1969evolution}.
+These were followed by others that incorporated differences 
+in the mutation process, such as higher transition to transversion ratio
+\citep{hasegawa1985dating,felsenstein1996hidden}.
+Although the rates of mutation of a nucleotide
+are known to depend on not only the nucleotide itself
+but also nearby sequence context \citep{seplyarskiy2021population},
+it is difficult to take this context into account.
+
+This situation from biological sequence analysis
+is an example of a more commonly encountered class of models:
+a lattice of sites,
+with each site taking one of a finite collection of possible states,
 and whose stochastic, temporal evolution is Markov and governed by a set of local rules.
@@
 is similar to the ``coding'' scheme introduced by \citet{besag1974spatial},
-that conditions on a set of sites that makes the remaining observations independent (thanks to the Gibbs property);
+that conditions on a set of sites that makes the remaining observations independent;
+% (thanks to the Gibbs property);
 consistency of such methods has been shown by \citet{comets1992consistency} and reviewed by \citet{larribe2011composite}.
@@
 Indeed, molecular studies have demonstrated that the spectrum of new mutations in humans
-is highly context dependent \citep{schaibley2013influence}
+is highly context dependent \citep{schaibley2013influence,gao2019overlooked,rodriguezgalindo2020germline}
 and that multinucleotide substitutions are relatively common \citep{schrider2011pervasive,terekhanova2013prevalence,harris2013errorprone}.
 In humans, the molecular processes underlying mutations are best studied in contexts of intense somatic mutation:
-cancer and the hypervariable regions that generate immune system diversity \citep{cobey2015evolution},
+cancer and the hypervariable regions that generate immune system diversity \citep{cobey2015evolution,herediagenestar2020extreme},
 which provide quite different cellular contexts but share commonalities due to shared molecular machinery.
@@
 and (c) any functional effects of an error (e.g., lethal mutations will not be seen in a population census).
-(We are concerned with effects that are \emph{homogeneous} across the sequence,
-so do not consider further the many constraints on protein-coding sequence \citep[reviewed in][]{thorne2007protein}.)
+We are concerned with effects that are \emph{homogeneous} across the sequence,
+so do not consider further the many constraints on protein-coding sequence \citep[reviewed in][]{thorne2007protein}.
 Some of these effects -- particularly, activity levels of different enzymes in the germline --
@@
 and so comparisons between widely separated species will identify time-averaged mutation rates.
-Recent studies have used a variety of exploratory techniques to identify sets of mutational patterns
+Studies have used a variety of exploratory techniques to identify sets of mutational patterns
 co-occurring in comparisons (a) between different cancer types \citep{alexandrov2013signatures,shiraishi2015simple} and
@@
 However, phylogenetic and population genetic methods usually ignore such dependencies
-in the interest of computational efficiency and lack of understanding of the underlying process,
+in the interest of computational efficiency,
 but some progress has been made.
@@
 Others have made progress using simplifying assumptions \citep{berard2012accurate},
-or by other approximations \citep{christensen2005pseudolikelihood}.
+or by other approximations \citep{christensen2005pseudolikelihood,saunders2007insights}.
 \citet{pedersen2000dependent}, later extended by
-\citet{Hwang2004-pj,hobolth2008markov,baele2010using}, used data augmentation and an MCMC algorithm to do Bayesian inference;
+\citet{Hwang2004-pj,hobolth2008markov} and \citet{baele2010using}, used data augmentation and an MCMC algorithm to do Bayesian inference;
 \citet{lunter2004nucleotide} used an approximate matrix decomposition;
-while \citet{siepel2004phylogenetic} and later \citep{baele2010modelling} computed likelihoods by assuming a model that is Markov along the genome;
+while \citet{siepel2004phylogenetic} and later \citep{baele2010modelling} computed likelihoods by assuming a model that is Markov along the genome.
 however, all these methods are quite computationally intensive.
-
-Model with fixed flanking \citet{saunders2007insights}.
 
@@
 Parameter estimation for Ising model without temporal dynamics
-is relatively well-undersood \citep{pickard1982inference,frigessi1990parameter},
-but there does not seem to be similar work on the case when more than one observation is made of the dynamical system.
+is relatively well-understood \citep{pickard1982inference,frigessi1990parameter},
+but the problem of dynamical observations is less well-studied.
 
@@
         An overview of the method for a neighbor-dependent model.
-        The red triangle represents the ``range of influence'' of mutations on a site of interest (the site at the bottom corner of the triangle).
-        The blue T shape is called a \emph{T-mer}, which is how our approximation to the process is parameterized.
-        The bottom is called the \emph{base} of the T-mer, which is considered the outcome of a probabilistic process parameterized by additional \emph{overhang} sequence extending to the left and right of the base.%
+        The red triangle represents the ``range of influence'',
+        i.e., possible locations at which a mutations is likely
+        to affect the mutation outcome at
+        a site of interest (the site at the bottom corner of the triangle).
+        The blue T shape is called a \emph{T-mer},
+        which is how our approximation to the process is parameterized.
+        The central segment of the T is called the \emph{base} of the T-mer,
+        and the key building block computes the transition probability
+        for this base
+        conditioned on the ancestral state at the \emph{overhang}
+        of the top of the T.
         \label{fig:overview}
@@
 
-Before diving into a formal problem specification, we offer an intuitive explanation of how the method works.
-The difficulty of context-sensitive models is that, in principle, one may have a sequence of events with cascading consequences that involves a lot of the sequence.
-In the extreme case, one may have a sequence of mutations from, say, the left boundary of the sequence to the right boundary of the sequence, such that the probability of each mutation has been changed by that of the previous one.
-
-However, the starting point of the work presented here is that for typical models such a sequence of events is improbable.
-For that reason, we can perform computation using a local context which, although bigger than the context specified in the context-sensitive model, is much smaller than the entire sequence.
-We call the building blocks of this computation \emph{T-mers} (Figure~\ref{fig:overview}) which parameterize transition probabilities for a small section of sequence (the \emph{base} of the T-mer) based on additional sequence (the \emph{overhang} of the T-mer).
-We can bound the probability that the influence of a mutation outside the overhang of the T-mer will impact the sites at the base of the T-mer, which translates into an estimate of approximation error across the sequence when using a composite likelihood calculation based on T-mers.
+Before diving into a formal problem specification,
+we offer an intuitive explanation of how the method works.
+The difficulty of context-sensitive models is that, in principle,
+one may have a sequence of events with cascading consequences that involves a lot of the sequence.
+In the extreme case, one may have a sequence of mutations from,
+say, the left end of the sequence to the right end of the sequence,
+such that the probability of each mutation has been changed by that of the previous one.
+
+However, the starting point of the work presented here is that
+for typical models such a sequence of events is improbable.
+For that reason, we can compute the transition probability for each segment of the sequence
+using a local context which,
+although bigger than the context which determines the local transition rates,
+is much smaller than the entire sequence.
+We call the building blocks of this computation \emph{T-mers} (Figure~\ref{fig:overview})
+which parameterize transition probabilities for a small section of sequence
+(the \emph{base} of the T-mer) based on additional sequence (the \emph{overhang} of the T-mer).
+We can then bound the probability that the influence of a mutation
+outside the overhang of the T-mer will impact the sites at the base of the T-mer,
+which translates into an estimate of approximation error in our calculation of the likelihood.
+% note: we are not doing composite likelihood
 

### 2021-12-04 writeups/context-dependence.tex
.

@@
 \newcommand{\nR}{\mbox{R}} % purine, A or G
-\newcommand{\nY}{\mbox{Y}} % pyramidine, C or T
-\newcommand{\join}{\oplus}  % matches
+\newcommand{\nY}{\mbox{Y}} % pyrimidine, C or T
+\newcommand{\join}{\oplus}  % concatenation
 \newcommand{\st}{\colon}  % such that
@@
 but also nearby sequence context \citep{seplyarskiy2021population},
-it is difficult to take this context into account.
+it is difficult to take this context into account when performing inference.
 
 This situation from biological sequence analysis
-is an example of a more commonly encountered class of models:
+is an example of a more commonly encountered class of models in statistics:
 a lattice of sites,
@@
 cancer and the hypervariable regions that generate immune system diversity \citep{cobey2015evolution,herediagenestar2020extreme},
-which provide quite different cellular contexts but share commonalities due to shared molecular machinery.
+which provide quite different cellular contexts but can share commonalities due to shared molecular machinery.
 In general, substitution rates depend on sequence context due to the interaction of a great many factors including:
@@
 To formalize this notion of local dependence,
-first define a \textit{pattern} to mean a contiguous sequence of states;
-let $|u|$ denote the length of the pattern $u$.
-Given $x \in \calS$,
+first define a \textit{pattern} to mean a contiguous sequence of states,
+and let $|u|$ denote the length of a pattern $u$.
+Given a state $x \in \calS$,
 let $x_i^{(h)} = (x_i, x_{i+1}, \ldots, x_{i+h-1})$ denote the pattern of $x$ of length $h$ beginning at location $i$.
@@
 % We will mention this when necessary.
-The dynamics of our stochastic process are determined by the set of allowed transitions and associated rates:
+The dynamics of our stochastic process
+are determined by the set of allowed transitions and associated rates:
 given two patterns $u$ and $v$ of common length $h$,
@@
 written as ``$u \to v$ at rate $\mu$''.
-The set of all transition triples is denoted $\calT$.
+A model of context-dependent mutation
+is defined by a set of all transition triples,
+which we denote $\calT$.
+
+Although our main focus is on nucleotides,
+it is informative to consider other, simpler, models as well.
+Here is one such example.
 
 \begin{example}[TASEP]
-  The \emph{Totally Asymmetric Simple Exclusion Process} labels each site as ``empty'' or ``occupied'' (0 or 1, respectively),
-  and says that each occupied site, independently at rate $\lambda$, checks the site to the right to see if it is empty,
-  and if it is, moves there.
+  In the \emph{Totally Asymmetric Simple Exclusion Process}
+  describes a set of particles moving among some sites,
+  with at most one particle per site.
+  Each site is either ``empty'' or ``occupied''
+  (0 or 1, respectively),
+  and each particle, independently at rate $\lambda$,
+  checks whether the site to the right is empty,
+  and if so, moves there.
   This process therefore has only one transition triple:
@@
   This only has one parameter.
-  If we allow particles to exit from the right, and new particles to enter from the left
+  If we allow particles to exit from the right, and new particles to enter from the left,
   then from observing only starting and ending configurations
@@
 and that the rate at which each possible change occurs is modulated by a function of the energy difference the change would produce.
-This is natural if that transition triples describe \emph{proposed} changes,
+This is natural if transition triples describe \emph{proposed} changes,
 and that the probability a proposed change actually occurs depends on how much it affects the energy.
@@
 \begin{example}[Genomic GC content]
-    A sequence of genome can be written using A, C, G, and T;
+    A genome sequence can be written using A, C, G, and T;
     in the most general model of independent mutation across sites, each of the 12 possible transitions occurs at its own rate.
@@
     with probability approximately $(1-\exp(-2 s))/(1-\exp(-2 s N_e))$ \citep{kimura1962probability}.
-    If the mutation occurs at rate $\mu$ per individual, the total rate it occurs at in the population is $\mu N_e$.
+    If the mutation occurs at rate $\mu$ per individual,
+    the total rate it occurs at in the population is $\mu N_e$.
     GC-biased gene conversion effectively means that sequences with more G/Cs are selected against.
@@
     so that $E(y)-E(x)$ is the net change in number of G's and C's multiplied by $s$.
-    Also, assume a $\phi$ function expressed in terms of an energy difference $e$:
+    Also, define $\phi(e)$ to be the expected rate of fixations of ``selected'' mutations
+    with advantage $e$:
     \begin{align*}
-        \phi(e) = (1-\exp(-2e))/(1-\exp(-2eN)).
+        \phi(e) = N_e (1-\exp(-2e))/(1-\exp(-2eN)).
     \end{align*}
-    Thus, if a given change that occurs with rate $\mu$ would change a sequence $x$ to $y$,
-    then $\mu \, N_e \phi(E(y)-E(x))$ is the rate at which the mutation
+    With these definitions,
+    if a given change that occurs with rate $\mu$ would change a sequence $x$ to $y$,
+    then $\mu \, \phi(E(y)-E(x))$ is the rate at which the mutation
     appears and successfully takes over in the population.
@@
     \end{align*}
-
-    This process has been studied by XXX.
 \end{example}
@@
 and for those positions $x$ matches with $u$ while $y$ matches with $v$.
-Let $\calT$ be the set of these transition triples.
-
+
+As before, let $\calT$ be the set of transition triples for the model.
 As noted above, there may be more than one way to mutate a sequence $x$ to get $y$:
@@
 \]
-The rate ${G(n)}_{x,y}$ is then the sum of all matching transition rates,
+The rate ${G(n)}_{x,y}$ is then the sum of all transition rates that can take $x$ to $y$,
 multiplied by the fitness term,
@@
 and if there are no triples $(\mu,u,v)$ with $x \xrightarrow{i,u,v} y$ for some $i$, then ${G(n)}_{x,y}=0$.
-Indeed, we anticipate this matrix to be sparse because for most pairs $(x,y)$ there are no transition triples that can modify $x$ to make $y$.
+To make this the generator matrix for a Markov process,
+we also set each $G(n)_{x,x}$ so that rows sum to zero.
+Indeed, this matrix is very sparse because for most pairs $(x,y)$ there are no transition triples that can change $x$ into $y$.
 
@@
 \[
-    \mu^* = \max \left\{ \sum_{y \in \calS^L} \sum_{(\mu, u, v) \in J(i,x,y)} \mu \qquad \st x \in \calS^L, \quad 1 \le i \le L \right\} .
+    \mu^* 
+    % = \max \left\{ \sum_{y \in \calS^L} \sum_{(\mu, u, v) \in J(i,x,y)} \mu \qquad \st x \in \calS^L, \quad 1 \le i \le L \right\} .
+    = \max_x \sum_{y \neq x} G(L)_{x,y} .
 \]
+Note that we do not actually need to construct $G(L)$ to find the maximum rate.
 Now sample possible changes:
@@
 with probability proportional to the corresponding rate, i.e.,
-out of the possible transitions $\calT$, the $j^\text{th}$ one is chosen
+out of the possible transitions $\calT$, transition triple $(\mu(j), u(j), v(j))$ is chosen
 with probability
@@
 q_j = \begin{cases}
-    % TODO peter, fix notation, please.
-    \mu^{(j)}/\mu^* \qquad & \text{if } x_i^{(|u^{(j)}|)} = u^{(j)}  \\
+    \mu(j)/\mu^* \qquad & \text{if } x_i^{(|u(j)|)} = u(j)  \\
     0 \qquad & \text{otherwise.}
@@
 gives the probability that the state remains the same.
-(By construction of $\mu_*$, $q_0\ge 0$.)
-If triple $(\mu^{(j)}, u^{(j)}, v^{(j)})$ is chosen, then
-the substring of $x$ matching $u^{(j)}$ that begins at position $i$
-is replaced with $v^j$ to produce $X(s_k)$.
-
-For instance, to simulate the TASEP, one has only to let $\mu^*=\lambda$, and at each $c_{ij}$
-check if at site $i$ there is a 1 followed by a 0,
+(By construction of $\mu_*$, $q_0 \ge 0$.)
+If triple $(\mu(j), u(j), v(j))$ is chosen, then
+the substring of $x$ matching $u(j)$ that begins at position $i_k$
+is replaced with $v(j)$ to produce $X(s_k)$.
+
+For instance, to simulate the TASEP, one has only to let $\mu^*=\lambda$,
+and at each possible change
+check if at site $i_k$ there is a 1 followed by a 0,
 and if so, switch them.
@@
 How can we extract the information?
-These are Markov processes, on the state space $\calS^n$,
+These are Markov processes, on the state space $\calS^L$,
 so the full likelihood function is given by \eqref{eqn:full_likelihood},
-but doing anything with the $|\calS|^n \times |\calS|^n$ matrix $G_n$ is clearly infeasible even for moderately sized $n$.
-We can, however, compute \eqref{eqn:full_likelihood} for smaller $n$,
+but doing anything with the $|\calS|^L \times |\calS|^L$ matrix $G_L$ is clearly infeasible even for moderately sized $L$.
+We can, however, compute \eqref{eqn:full_likelihood} with smaller $G_n$,
 so the first thing that one might think to do is to break the sequence up into many blocks of length $m$,
@@
 we can compute \emph{something} correctly,
-and that something will be sufficient to compute the full likelihood
+and that will be sufficient to compute the full likelihood
 to an arbitrarily good approximation.
@@
 
-These $p_{n,\ell,r}(x,y)$ probabilities can be found by marginalizing $p_m$,
-summing over the possible inital and final substrings of $x$
+These $p_{n,\ell,r}(x,y)$ probabilities can be found by marginalizing $p_n$,
+summing over the possible initial and final substrings of $x$
 (denoted $a$ and $b$):
@@
     p_{n,\ell,r}(x,y)
-    .
-\end{align}
+    ,
+\end{align}
+with the approximation getting better for longer $\ell$ and $r$.
+
+Equation \eqref{eq:pmMarg} could then be used to compute a composite likelihood,
+which we denote $A_{n, \ell, r}(x,y)$.
+This is the product of the probabilities
+of transition for all $(n, \ell, r)$ T-mers between $x$ to $y$, i.e.,
+with $m = n - \ell - r$,
+\begin{align}\label{eqn:composite_like}
+    A_{n, \ell, r}(x,y)
+        =
+        \prod_{i=1}^{L-m+1} p_{n, \ell, r}\left(x_i^{(n)}, y_{i+\ell}^{(m)}\right).
+\end{align}
+To keep notation simple we've ignored the boundaries of the sequence;
+to deal properly with these,
+on the right-hand side of the equation $\ell$ should be replaced with $\min(\ell, \max(i-1, 0))$,
+and $r$ should be replaced with $\min(r, \max(n-i-\ell-m+1, 0))$.
 
@@
 \begin{align*}
-    \like(y \mid x) &= \prod_i \P\{ Y_i = y_i \given Y_{<i} = y_{<i} \ \text{and}\; X=x \} .
+    \like(y \mid x) &= \prod_{i=1}^L \P\{ Y_i = y_i \given Y_{<i} = y_{<i} \ \text{and}\; X=x \} .
 \end{align*}
-It turns out that
-we can approximate the terms in this sum
-by discarding approximately independent sites from the conditioning.
+Each term in this product is the probability of $Y_i$, given all sites in $X$
+and all sites in $Y$ to the left of $i$.
+We can compute an approximation to this probability as in \eqref{eqn:window_approx}
+by discarding those sites from $X$ and $Y$ that are approximately independent of $Y_i$.
+Which are these?
+
 These approximately independent sites are exactly those that are too far away from a given site to impact its evolution.
@@
         \frac{
-            p_{2\ell+1,\ell,\ell}(x_{i-3\ell}^{(4\ell+1)}, y_{i-2\ell}^{(2\ell+1)})
+            p_{2\ell+1,\ell+1,\ell}(x_{i-3\ell}^{(4\ell+1)}, y_{i-2\ell}^{(2\ell+1)})
         }{
@@
 
-%EM I think you said that we could just treat the state of all positions before the beginning of the sequence and after the end as having another distinct state. If that's true I can write it out.
-
-Let $A_{n, \ell, r}(x,y)$ be the product of all probability of all $(n, \ell, r)$ T-mers transitioning from $x$ to $y$, i.e.\
+We can then simply express our likelihood as
 \[
-A_{n, \ell, r}(x,y) = \prod_{i=1}^L p_{n, \ell, r}\left(x_{i-\ell-m}^{(n)}, y_{i-m}^{(n-\ell-r)}\right).
+    \like(y \mid x) = \frac{ A_{2\ell+1, \ell+1, \ell}(x,y) }{ A_{2\ell+1, \ell, \ell+1}(x,y) }.
 \]
-
-We can then express our likelihood as
-
-\[
-    \like(y \mid x) = \frac{ A_{2\ell+1, \ell, \ell}(x,y) }{ A_{2\ell+1, \ell, \ell+1}(x,y) }.
-\]
-
 % Define $n_\ell(x,y;a,b)$ to be the number of times pattern $a$ is seen in sequence $x$
@@
 
+%EM TODO grant acknowledgements
+
 \bibliography{context-dependence}

### 2021-12-07 writeups/context-dependence.tex
notes from chat

--- 2021-12-04 writeups/context-dependence.tex
@@
     \begin{center}
+        %EM Make width like in figure 2. 2 and 2.
+        %EM Bottom of triangle should be three bases.
         \includegraphics{overview}
@@
 % \end{align}
+
+%EM It seems reasonable that we should be able to get away using a smaller T-mer.
+%EM Question: how much context should we include; argument in appendix.
+%EM Would be nice to show convergence of full likelihood eventually.
+
 See Appendix~\ref{ss:approx_pf} for a more rigorous statement and proof,
@@
 computing $e^{t G(5) }$ takes 13 seconds, while computing $e^{t G(5)} U$ takes only 0.3 seconds.
-%EM In the above, don't we only provide the recipe for the full likelihood when m = 2 ell + 1?
 Increasing to $\ell=4$ and $m=1$ or $\ell=3$ and $m=2$ is still feasible, taking $e^{t G(8)} U$ in 42 and 47 seconds, respectively
@@
 As seen in Table~\ref{tab:ising_coverage},
-%EM I think this credible interval statement is interesting. So it's definitely this narrowness vs just having the estimate be off?
 credible intervals were somewhat too narrow for shorter T-mers,

### 2021-12-08 writeups/context-dependence.tex
plr edits until disco

@@
 \begin{abstract}
-
+    Although the rates at which positions in the genome mutate
+    are known to depend not only on the nucleotide to be mutated
+    but also on neighboring nucleotides,
+    it remains challenging to do phylogenetic inference
+    using models of context-dependent mutation.
+    In these models,
+    the effects of one mutation may in principle propagate
+    to faraway locations,
+    making it difficult to compute exactly likelihoods.
+    This paper shows how to use bounds on the propagation of dependency
+    to compute likelihoods of mutation of a given segment of genome
+    by marginalizing over sufficiently long flanking sequence.
+    This can be used for maximum likelihood or Bayesian inference.
+    Protocols examining residuals and iterative model refinement
+    are also discussed.
+    Tools for efficiently working with these models
+    are provided in an R package,
+    that could be used in other applications.
+    The method is used to examine context dependence of mutations
+    since the common ancestor of humans and chimpanzee.
 \end{abstract}
@@
 and let $|u|$ denote the length of a pattern $u$.
-Given a state $x \in \calS$,
+Given a sequence $x \in \calS^L$,
 let $x_i^{(h)} = (x_i, x_{i+1}, \ldots, x_{i+h-1})$ denote the pattern of $x$ of length $h$ beginning at location $i$.
@@
   This only has one parameter.
-  If we allow particles to exit from the right, and new particles to enter from the left,
+  If we allow particles to exit from the right and new particles to enter from the left
+  (e.g., by adding special symbols for the start and end of the sequence),
   then from observing only starting and ending configurations
@@
 but these changes only take effect with probability $\phi(E(y) - E(x))$.
-We have described this idea in terms natural for a physics model such as an Ising model in which high energy configurations are disfavored, however it is also suitable for a population genetics model in which $\phi$ describes the probability of a mutant allele given a certain fitness change from the parent.
+We have described this idea in terms natural for a physics model
+in which high energy configurations are disfavored,
+however it is also suitable for a population genetics model
+in which $\phi$ describes the probability of fixation of a mutant allele
+given a certain fitness change from the parent.
 
@@
     in the most general model of independent mutation across sites, each of the 12 possible transitions occurs at its own rate.
-    Furthermore, it is a well-known observation that in many species,
+    Furthermore, in many species
     adjacent, methylated CG dinucleotides (``CpG sites'') have a much higher mutation rate to TG and CA
@@
     Counteracting this trend towards increased G/C is GC-biased gene conversion \citep{glemin2015quantification},
-    which is effectively a selective pressure against G or C bases.
+    which acts effectively as a selective pressure against G or C bases.
     We model the evolution of a \emph{single} sequence, not a population of sequences,
@@
     GC-biased gene conversion effectively means that sequences with more G/Cs are selected against.
-    This is incorporated as follows: set the energy of pattern of pattern \nC{} or \nG{} to be $s$,
+    This is incorporated as follows: set the energy of patterns \nC{} and \nG{} to be $s$,
     so that $E(y)-E(x)$ is the net change in number of G's and C's multiplied by $s$.
     Also, define $\phi(e)$ to be the expected rate of fixations of ``selected'' mutations
-    with advantage $e$:
+    with advantage $e$ relative to the per-individual, per-site mutation rate:
     \begin{align*}
@@
 \end{example}
+
+We close with a final example from statistical physics, to be used later:
 
@@
 changes occur at rate $\mu^*$ independently at each of the $L$ sites.)
-Now, suppose we have determined the state just before time $s_k$ to be $X(s_k) = x$.
-Then, a new state is then chosen at position $i_k$
+Now, suppose we have determined the state just before time $s_k$ to be $X(s_{k-1}) = x$.
+Then, the new state $X(s_k)$ is chosen by applying a mutation at position $i_k$
 with probability proportional to the corresponding rate, i.e.,
@@
 we can compute \emph{something} correctly,
-and that will be sufficient to compute the full likelihood
-to an arbitrarily good approximation.
+and that will be sufficient to compute the full likelihood,
+at least arbitrarily.
 %EM do we really want x and y here, which I've been thinking of as being the full sequence?
@@
 and all sites in $Y$ to the left of $i$.
-We can compute an approximation to this probability as in \eqref{eqn:window_approx}
-by discarding those sites from $X$ and $Y$ that are approximately independent of $Y_i$.
-Which are these?
-
-These approximately independent sites are exactly those that are too far away from a given site to impact its evolution.
+It seems reasonable that, similar to equation \eqref{eqn:window_approx} above,
+the process at position $i$ should be approximately independent
+of distant positions, given the neighboring positions.
+Suppose that $\ell$ is long enough that approximation~\eqref{eqn:window_approx} holds
+with $\ell = r$.
+Then we know that site $i$ approximately depends only on mutations within distance $\ell$.
+Furthermore, the sites up to distance $2 \ell$ to the left of $i$
+also depend on these mutations.
+We can now use \eqref{eqn:window_approx} to restrict the dependency on $x$.
 Recall that the last base of $x_i^{(h)}$ is $x_{i+h-1}$.
 So, $x_{i-3\ell}^{(4\ell+1)}$ runs from $i-3\ell$ to $i + \ell$, inclusive, while $y_{i-2\ell}^{(2\ell+1)}$ runs from $i-2\ell$ to $i$, inclusive.
-If $\ell$ is long enough that approximation~\eqref{eqn:window_approx} holds, then:
+This suggests that the following approximation should be good:
 \begin{align} \label{eqn:full_approx}
@@
 
-We can then simply express our likelihood as
+Since the full likelihood is a product across sites,
+we can then simply express an approximate full likelihood as
+a ratio of two composite likelihoods:
 \[
-    \like(y \mid x) = \frac{ A_{2\ell+1, \ell+1, \ell}(x,y) }{ A_{2\ell+1, \ell, \ell+1}(x,y) }.
+    \like(y \mid x) \approx \frac{ A_{2\ell+1, \ell+1, \ell}(x,y) }{ A_{2\ell+1, \ell, \ell+1}(x,y) }.
 \]
@@
 
-See Appendix~\ref{ss:approx_pf} for a more rigorous statement and proof,
-and the next section for some intuition.
+See the next section for more discussion of the approximations.
 
@@
 If $\lambda t = 0.75$ (three-quarters of the sequence will have changed!),
-then taking $\ell = r = 3$ allows us to compute Tmer probabilities to within an error of 0.007,
+then taking $\ell = r = 3$ allows us to compute T-mer probabilities to within an error of 0.007,
 since this is the probability that a Poisson with mean 0.75 will be larger than 3.
@@
 
-To obtain the full likelihood, in \eqref{eqn:full_approx}, in addition to an overhang of $\ell$
-we needed the base of the T-mer to be of length $2\ell+1$.
-This is because we have also conditioned on the final sequence to the left of the focal site, $Y_{<i}$,
-and although the process at $i$ only depends on events within $\pm \ell$,
-these events may affect other sites in $Y$ going back a distance $2\ell$.
+Appendix~\ref{ss:approx_pf} makes this argument more precise;
+however, the argument for the full likelihood is slipperier,
+and we can only provide the rough argument above.
+To see one difficulty, consider the TASEP model,
+with $x_1 = y_n = 1$, and all remaining positions equal to 0.
+The full probability of this pair of sequences is very small,
+while the conditional probability of $Y_n=1$ given $X=10\cdots0$ and $Y_{<n} = 0\cdots0$
+is large since the 1 that initially began in the first position
+had to go somewhere.
+However, if we do not include the first position in the conditioning,
+this conditional probability is very different.
+% More generally, it is harder because conditioning on the final sequence $y$
+% in effect allows dependencies to propagate back in time as well as forward.
+Fortunately, such situations are rare,
+and the TASEP model provides the extreme case of dependency propagation.
+In practice, we recommend that the likelihood is computed with equation \eqref{eqn:full_approx}
+using successively larger values of $\ell$
+to check for convergence.
 
@@
 For inference, we need to compute the $|\calS|^{n} \times |\calS|^m$ matrix $F$ whose $(a,b)^\text{th}$ entry is $p_{n,\ell,r}(t,a,b)$,
-for various $t$.
+for various $t$ (and $m = n - \ell - r$).
 This matrix is a projection of the $|\calS|^{n} \times |\calS|^{n}$ matrix whose $(a,c)^\text{th}$ entry is
@@
 where $G(n)$ is the sparse $|\calS|^{n} \times |\calS|^{n}$ matrix defined in \eqref{eqn:G_defn}.
-The ``projection'' we need just marginalizes over long (i.e.\ length $n$) patterns $c$ that match the shorter (i.e.\ length $m$) pattern $b$
+The ``projection'' we need just marginalizes over long (i.e., length $n$) patterns $c$ that match the shorter (i.e., length $m$) pattern $b$
 as in \eqref{eq:pmMarg}:
-if we define the $n \times m$ matrix $U$ so that $U_{cb}=1$ if $c_\ell^m=b$ and $U_{cb}=0$ otherwise,
+if we define the $n \times m$ matrix $U$ so that $U_{cb}=1$ if $c_\ell^{(m)}=b$ and $U_{cb}=0$ otherwise,
 then
@@
 
-Although we haven't used such tools here, taking the derivative with respect to $t$ can also be performed using sparse matrix operations.
+% Although we haven't used such tools here, taking the derivative with respect to $t$ can also be performed using sparse matrix operations.
 
@@
 First note that in this case $G(n)$ has
-$(1+m(|\calS|-1)) |\calS|^{n}$ nonzero entries, since each of the $|\calS|^{n}$ can change in $m$ places.
-This will determine how the computation scales with $m$, $n$, and $|\calS|$,
+$(1+n(|\calS|-1)) |\calS|^{n}$ nonzero entries,
+since each of the $|\calS|^{n}$ sequences can change in $n$ places.
+This will determine how the computation scales with $n$ and $|\calS|$,
 once we precompute a number of things.
-Let $g = (g_1, \ldots, g_d)$ be the nonzero entries of $G$, in some fixed ordering;
-sparse matrix representations of $G$ store only $g$ along with information about the rows and columns these are found in.
-%EM Here we are using lower j, rather than our "new" notation setup, but it doesn't seem like a big deal. This is a local note.
+Let $g = (g_1, \ldots, g_d)$ be the nonzero entries of $G(n)$, in some fixed ordering;
+sparse matrix representations of $G(n)$ store only $g$ along with information about the rows and columns these are found in.
 Each $g_i$ is a linear combination of mutation rates $\mu_j$,
@@
 say $g_i = \sum_j A_{ij} \mu_j \phi(\sum_k B_{ik} e_k)$,
-so by precomputing the matrices $A$ and $B$ we can update $G$ with new parameter values
+so by precomputing the matrices $A$ and $B$ we can update $G(n)$ with new parameter values
 using only two matrix multiplications and evaluation of $\phi()$ over a vector.
@@
 Concretely: pick a random location $I$ in the sequence and let $X_{I-\ell}^{n}(0) = \rho$ be the (long) pattern seen there at the root.
-Write $\pi(z)$ for the frequency of a given pattern $z$ in $X(0)$, i.e., $\P\{\rho=z\}=\pi(z)$.
+Write $\pi(z)$ for the frequency of a given pattern $z$ seen there, i.e., $\P\{\rho=z\}=\pi(z)$.
 % Given the corresponding pattern in $X(t_X)$, the distribution of $\rho$ is proportional to
@@
     \caption{
-        A depiction of the four steps in computing the likelihood of finding particular ``short'' subsequence at $Y$ and $Z$
+        A depiction of the four steps in computing the likelihood
+        of finding particular ``short'' subsequences at $Y$ and $Z$
         given the ``long'' subsequence at $X$: see text for details.
@@
 \subsection{Hominid data}
+\label{hominid_section}
 
@@
 
-In exploratory application to real data, how does one decide when to stop adading mutational motifs?
+In exploratory application to real data, how does one decide when to stop adding mutational motifs?
 Perhaps the simplest method conceptually is to
@@
 (b) choose new motifs to add based on statistically significant residuals,
-(c) repeat until no residuals are statisically significant.
+(c) repeat until no residuals are statistically significant.
 To test this method,
@@
 \ldots which again clearly call for adding the mutation tuples $TCC \to TTC$ and $AGG \to AAG$.
+These were indeed the motifs used when simulating the data.
 After adding these in,
@@
 (judged sufficient by examination of diagnostic plots).
-We repeated this procedure separately on 60 independently simulated sequences,
+We repeated this procedure separately on 61 independently simulated sequences,
 and show the resulting credible intervals in Figure~\ref{fig:ising_coverage}.
@@
 
-We also applied the method to changes between the human and chimpanzee reference sequences.
+We also applied the method to changes between the human and chimpanzee reference sequences,
+using the model described above.
 
@@
         Estimated average mutation rates
-        for the motifs of Table \ref{tab:hominid_model}
+        for the motifs in the ``Hominid data'' section
         since the common ancestor of humans and chimpanzee.

### 2021-12-08 writeups/context-dependence.tex
Clearing a few comments and updated version of figure

@@
 and the TASEP model provides the extreme case of dependency propagation.
-In practice, we recommend that the likelihood is computed with equation \eqref{eqn:full_approx}
+In practice, we recommend computing the likelihood with equation \eqref{eqn:full_approx}
 using successively larger values of $\ell$
@@
 These can then be compared to a standard Normal distribution.
-
-%EM This is stepwise model addition.
 
@@
 In phylogenetic applications, rather than ``before'' and ``after'' observations,
-%EM Describe the root sequence, or a prob distribution on it?
-%EM We want to be able to talk about the probability of seeing a certain n-mer... is that obvious what that random process is on the whole-sequence level?
 we get two (or more) observations evolved from a common root.
@@
 \[
-    %EM This is super clever!
     \sum_{z \in \calS^{n}} \pi(z) \, p_{n,\ell,r}(t_X,z,x) \, p_{n,\ell,r}(t_Y,z,y)

### 2021-12-09 writeups/context-dependence.tex
spelling

@@
 For instance, at each site may sit
-a nucelotide whose mutation rate depends on physical properties determined by the nearby DNA sequence;
+a nucleotide whose mutation rate depends on physical properties determined by the nearby DNA sequence;
 a cell whose infection status depends on the state of its neighbors;
@@
 while \citet{siepel2004phylogenetic} and later \citep{baele2010modelling} computed likelihoods by assuming a model that is Markov along the genome.
-however, all these methods are quite computationally intensive.
+However, all these methods are quite computationally intensive.
 
@@
 Each $g_i$ is a linear combination of mutation rates $\mu_j$,
-multiplied by a function ($\phi$) of a linear combination of energy coefficents $e_j$,
+multiplied by a function ($\phi$) of a linear combination of energy coefficients $e_j$,
 say $g_i = \sum_j A_{ij} \mu_j \phi(\sum_k B_{ik} e_k)$,
@@
 aligned regions below either 3000bp (promoter flanking regions) or 1000bp (all other types).
-We then counted Tmers of appropriate length in each of the eight resulting sets of aligned regions, omitting any Tmers that included gaps or repeat-masked positions,
+We then counted T-mers of appropriate length in each of the eight resulting sets of aligned regions, omitting any T-mers that included gaps or repeat-masked positions,
 and fit models separately to each of the eight data sets.
@@
 
-It may seem tenous to look for mutational signatures associated with cancer in the germline,
+It may seem tenuous to look for mutational signatures associated with cancer in the germline,
 or absurd that UV damage could play a significant role.
@@
 It avoids approximations made by previous methods thanks to the use of T-mers
-and explict quantification of the necessary length of the ``arms'' of the T-mers.
+and explicit quantification of the necessary length of the ``arms'' of the T-mers.
 We first discuss possible uses of the method,
@@
 The model presented assumes that the same process occurs at every location along the sequence,
-and that the length of the sequnce does not change (no indels).
+and that the length of the sequence does not change (no indels).
 Neither of these assumptions are likely to be true for any data set of DNA sequence;
 however, both assumptions are also made the state of the art in phylogenetics.
-Some relaxation of the assumptions is possible -- for instnace,
+Some relaxation of the assumptions is possible -- for instance,
 we can easily allow heterogeneity between sites by scaling rates by an overall gamma-distributed random factor \citep{yang1994maximum}.
-Short deletions might be accomodated by introducing a ``gap'' letter to the alphabet,
+Short deletions might be accommodated by introducing a ``gap'' letter to the alphabet,
 but insertions may be problematic.
@@
 since all computations scale with the number of possible patterns at the long end of the T-mer.
-Although the model is efficent enough to deal with fairly long T-mers,
+Although the model is efficient enough to deal with fairly long T-mers,
 this exponential growth puts hard constraints on the application of the method.
@@
         Diagram of things in the text:
-        (top) Dependency neighboorhods.
+        (top) Dependency neighborhoods.
         (bottom) What we need to compute $\P\{X_i(t)=x \given X(0), X_{<i}(t)\}$:
@@
 resulting in a slight underestimate of the chance of change.
-A better approximation would be to use the marginal frequencies in some way to average over surroudning sequence
+A better approximation would be to use the marginal frequencies in some way to average over surrounding sequence
 -- but, this requires marginal frequencies to be known, and these are not stationary.
@@
 Given $\delta_{xy}$, this gives a linear equation for $\delta \mu_i$ that may or may not be solvable,
-but certainly is if we have all the single-base transitions as paramters.
+but certainly is if we have all the single-base transitions as parameters.
 

### 2021-12-09 writeups/context-dependence.tex
edits and tweaks

@@
 can be well-approximated by marginalizing over a finite amount of surrounding context.
+We then use this to conjecture an approximation for the full likelihood.
 Motivated by the problem of inferring context-dependent mutation rates from diverged nucleotide sequences,
@@
     \begin{center}
-        %EM Make width like in figure 2. 2 and 2.
-        %EM Bottom of triangle should be three bases.
         \includegraphics{overview}
@@
         An overview of the method for a neighbor-dependent model.
-        The red triangle represents the ``range of influence'',
-        i.e., possible locations at which a mutations is likely
+        The red trapezoidal region represents the ``range of influence'',
+        i.e., possible locations and times at which a mutations is likely
         to affect the mutation outcome at
-        a site of interest (the site at the bottom corner of the triangle).
+        the sites of interest (the three sites in the center of the final sequence).
         The blue T shape is called a \emph{T-mer},
         which is how our approximation to the process is parameterized.
-        The central segment of the T is called the \emph{base} of the T-mer,
+        The central, smaller segment of the T is called the \emph{base} of the T-mer,
         and the key building block computes the transition probability
-        for this base
+        for this subsequence
         conditioned on the ancestral state at the \emph{overhang}
@@
 
-\peter{refer to figure}
-
 The problem at hand is to infer the parameters of the model based on the final state of the model after it has evolved from some known initial state.
@@
         and the third site in the final sequence (bottom).
+        \textbf{(C)} Diagram of heuristic argument for full likelihood approximation:
+        site $i$ depends on the blue squares in $N_i$;
+        the sites to the left of this in $L(i)$ are those other sites in $X(t)$
+        whose dependency neighborhoods overlap $N_i$.
+        Red squares give the dependency neighborhood of $L(i)$,
+        so we condition on the row of red and blue sites at the top.
+        On the bottom, we have $2w+1$ sites; at the top we have $4w+1$;
+        here $w=2$.
         \label{fig:Tmers}
@@
 %EM Note we'll need to change a and b below.
+%PR I think this is okay? x and y are not exactly the same here, but are still sequences?
+%PR To me it seems less confusing than introducing another variable?
 To do this, let $p_{n,\ell,r}(x,y)$ for $\ell + r < n$ denote
 the probability that an initial subsequence $x$ of length $n$, after time $t$,
-is found to match a smaller subsequence $y$ of length $n-\ell-r$, offset by $\ell$ sites.
+is found to match a smaller subsequence $y$ of length $n-\ell-r$, offset by $\ell$ sites
+(see Figure \ref{Tmer-dependency}).
 This central, smaller subsequence is called the ``base'' of the T-mer.
@@
 the probability that a given sequence $x$ evolves into another, $y$, over a given period of time.
-This can be computed as the product of the probability of each site in $y$
+This full likelihood can be computed as the product of the probability of each site in $y$
 given $x$ and all sites in $y$ further to the left.
@@
 of distant positions, given the neighboring positions.
+
+Here is a heuristic argument, depicted in Figure~\ref{fig:Tmers}C.
 Suppose that $\ell$ is long enough that approximation~\eqref{eqn:window_approx} holds
@@
 depends on the initial sequence at that site and within distance $w$ on either side\ldots
-\emph{and} the outcomes of any mutations that have occurred within that window.
+\emph{and} the outcomes of any mutations that have occurred within that window
+(see Figure \ref{Tmer-dependency}).
 
@@
 
-Appendix~\ref{ss:approx_pf} makes this argument more precise;
-however, the argument for the full likelihood is slipperier,
-and we can only provide the rough argument above.
+Appendix~\ref{ss:approx_pf} makes this argument more precise.
+The argument for the full likelihood is more difficult,
+so at present our proposal above remains a conjecture.
 To see one difficulty, consider the TASEP model,
@@
 observed only at times $t_X$ and $t_Y$ respectively.
-
-If the approximation described above holds,
-then we can write the probability that we see (long) pattern $x$ in $X(t_X)$ juxtaposed with (short) pattern $y$ in $Y(t_Y)$
-using the pattern frequencies at the root.
-%EM Could we make it \rho_I to emphasize that it's a random quantity?
-Concretely: pick a random location $I$ in the sequence and let $X_{I-\ell}^{n}(0) = \rho$ be the (long) pattern seen there at the root.
-Write $\pi(z)$ for the frequency of a given pattern $z$ seen there, i.e., $\P\{\rho=z\}=\pi(z)$.
-% Given the corresponding pattern in $X(t_X)$, the distribution of $\rho$ is proportional to
+At first, one might think to compute probabilities
+with a ``long'' sequence at the root and ``short'' sequences at every tip.
+However, this does not work out, since the root is unobserved
+and short sequences at every tip give us no information about what the root sequence
+in the overhangs might be.
+
+Instead, we will compute the probability that
+we see (long) pattern $x$ in $X(t_X)$ juxtaposed with (short) pattern $y$ in $Y(t_Y)$,
+summing over possible (long) patterns at the root.
+Concretely: pick a random location $I$ in the sequence
+and let $X_{I-\ell}^{(n)}(0) = \rho_I$ be the (long) pattern seen there at the root.
+Write $\pi(z)$ for the frequency of a given pattern $z$ seen there, i.e., $\P\{\rho_I=z\}=\pi(z)$.
+% Given the corresponding pattern in $X(t_X)$, the distribution of $\rho_I$ is proportional to
 % \[
-%     \P\{ \rho=z \mid X_{I-\ell}^{n}(t_X)=x\} \propto \pi(z) p_{n}(t,z,x) .
+%     \P\{ \rho_I=z \mid X_{I-\ell}^{(n)}(t_X)=x\} \propto \pi(z) p_{n}(t,z,x) .
 % \]
@@
 \begin{align} \label{eqn:phylo_likelihood}
-    \P\{X_{I-\ell}^{n}(t_X)=x \text{ and } Y_I^\ell(t_Y)=y \} = \sum_{z \in \calS^{n}} \pi(z) \, p_{n,\ell,r}(t_X,z,x) \, p_{n,\ell,r}(t_Y,z,y) .
+    \P\{X_{I-\ell}^{(n)}(t_X)=x \text{ and } Y_I^{(\ell)}(t_Y)=y \}
+    =
+    \sum_{z \in \calS^{n}} \pi(z) \, p_{n,0,0}(t_X,z,x) \, p_{n,\ell,r}(t_Y,z,y) .
 \end{align}
+Denote this quantity $p_{n,\ell,r}(x, y)$.
 This can be computed without much more effort than the simpler case above,
@@
 \[
-    \sum_{z \in \calS^{n}} \pi(z) \, p_{n,\ell,r}(t_X,z,x) \, p_{n,\ell,r}(t_Y,z,y)
-    = \sum_{z \in \calS^{n}} \left( e^{t_X G(n)} \right)_{zx} \pi(z) \sum_{w \in \calS^{n}} \left( e^{t_Y G(n)} \right)_{zw} U_{wy} ,
+    \sum_{z \in \calS^{n}}
+        \pi(z) \, p_{n,0,0}(t_X,z,x) \, p_{n,\ell,r}(t_Y,z,y)
+    =
+    \sum_{z \in \calS^{n}}
+    \left( e^{t_X G(n)} \right)_{zx}
+    \pi(z)
+    \sum_{w \in \calS^{n}} \left( e^{t_Y G(n)} \right)_{zw} U_{wy} ,
 \]
@@
 
-%EM It seems like we'd like to write out how this works for the entire set of sequences, as you did for the "full likelihood" section.
-%EM I think we have a bunch of ratios as before, where r is extended by 1 in the numerator?
+We might approximate the full likelihood as follows,
+assuming for the sake of notational simplicity that we have a few positions
+of $X$ observed before the start and after then end of the sequence
+(so that, e.g., we can refer to $x_{n+1}, \ldots, x_{n+r}$):
+\begin{align*}
+    \like(x, y)
+    &=
+    \prod_{i=1}^L
+        \P\{ X_{i+r} = x_{i+r},\; Y_i = y_i
+            \given X_{<i+r} = x_{<i+r},\; Y_{<i} = y_{<i} \} \\
+    &\approx
+    \prod_{i=1}^L
+        \P\{ X_{i+r} = x_{i+r},\; Y_i = y_i
+            \given X_{i-\ell}^{(n-1)} = x_{i-\ell}^{(n-1)},\; Y_{i}^{(m-1)} = y_{i}^{(m-1)} \} \\
+    &=
+    \prod_{i=1}^L
+    \frac{
+        p_{n,\ell,r}\left( x_{i-\ell}^{(n)}, y_i^{(m)} \right)
+    }{
+        p_{n-1,\ell,r}\left( x_{i-\ell}^{(n-1)}, y_i^{(m-1)} \right)
+    }
+\end{align*}
+We expect this to be a good approximation to the true likelihood
+when the same conditions above are met, taking $t$ to the be the largest distance
+between two taxa in the tree.
 
 %EM I think this should be pruning, not peeling.
+%PR I'll take your word for it! Wikipedia suggests either would do?
 %EM Want to check the modifications made above before continuing those to here.
-\subsection{Phylogenetic peeling}
-
-The general case is a modification of Felsenstein's ``peeling'' algorithm.
+\subsection{Phylogenetic pruning}
+
+The general case is a modification of Felsenstein's ``pruning'' algorithm.
 
@@
     \begin{center}
-    \includegraphics{peeling-schematic}
+    \includegraphics{pruning-schematic}
     \end{center}
@@
         given the ``long'' subsequence at $X$: see text for details.
-        \label{fig:peeling}
+        \label{fig:pruning}
     }
@@
 To compute the likelihood on a more complicated tree, consider the following,
-with the three-taxon tree of figure \ref{fig:peeling} as an example,
+with the three-taxon tree of figure \ref{fig:pruning} as an example,
 where we are counting occurrences of (long) $n$-tuples in taxon $X$,
@@
     & \P\{X_{I-\ell}^{n}=x \text{ and } Y_I^\ell=y \text{ and } Z_I^\ell=z \} \\
-    &\qquad = \sum_{u \in \calS^{n}} \pi(u) p_{m,\ell}(t_C,u,z) \sum_{v \in \calS^{n}} p_{n}(t_{AB},u,v) p_{n}(t_{A},v,x) p_{m,\ell}(t_B,v,y)
+    &\qquad
+      =
+      \sum_{u \in \calS^{n}} \pi(u) p_{m,\ell,r}(t_C,u,z)
+      \sum_{v \in \calS^{n}} p_{n,0,0}(t_{AB},u,v) p_{n,0,0}(t_{A},v,x) p_{m,\ell,r}(t_B,v,y)
   \end{split}
@@
 
-  \item Compute $M_1(u,z) = p_{m,\ell}(t_C,u,z)$ (a $|\calS|^{n} \times |\calS|^m$ matrix)
-
-  \item Compute $M_2(v,z) = \sum_{u \in \calS^{n}}  p_{n}(t_{AB},u,v) \pi(u) M_1(u,z)$ (still a $|\calS|^{n} \times |\calS|^m$ matrix)
-
-  \item Compute $M_3(v,y) = p_{m,\ell}(t_B,v,y)$ (a $|\calS|^{n} \times |\calS|^m$ matrix)
-
-  \item Compute $M_4(x,(y,z)) = \sum_{v \in \calS^{n}} p_{n}(t_{A},v,x) M_2(v,z) M_3(v,y)$ (stored as a $|\calS|^{n} \times |\calS|^{2m}$ matrix).
+  \item Compute $M_1(u,z) = p_{m,\ell,r}(t_C,u,z)$ (a $|\calS|^{n} \times |\calS|^m$ matrix)
+
+  \item Compute $M_2(v,z) = \sum_{u \in \calS^{n}}  p_{n,0,0}(t_{AB},u,v) \pi(u) M_1(u,z)$ (still a $|\calS|^{n} \times |\calS|^m$ matrix)
+
+  \item Compute $M_3(v,y) = p_{m,\ell,r}(t_B,v,y)$ (a $|\calS|^{n} \times |\calS|^m$ matrix)
+
+  \item Compute $M_4(x,(y,z)) = \sum_{v \in \calS^{n}} p_{n,0,0}(t_{A},v,x) M_2(v,z) M_3(v,y)$ (stored as a $|\calS|^{n} \times |\calS|^{2m}$ matrix).
 
@@
 Then $M_4$ gives all the likelihoods.
-This is depicted in the figure, writing $\left(e^{tG}\right)_{ab}$ for the matrix $p_{n}(t,a,b)$,
-and $P$ for the projection matrix that gives $p_{m,\ell}(t,a,b) = \left( e^{tG} P\right)_{ab}$.
+This is depicted in the figure, writing $\left(e^{tG}\right)_{ab}$ for the matrix $p_{n,0,0}(t,a,b)$,
+and $U$ for the projection matrix that gives $p_{m,\ell,r}(t,a,b) = \left( e^{tG} U\right)_{ab}$.
 
@@
 
-  \item[4'.] Compute $M_4(x,w) = \sum_{v \in \calS^{n}} p_{n}(t_{A},v,x) M_3(v,w) M_4(v,w)$ (stored as a $|\calS|^{n} \times |\calS|^{m}$ matrix).
+  \item[4'.] Compute $M_4(x,w) = \sum_{v \in \calS^{n}} p_{n,0,0}(t_{A},v,x) M_3(v,w) M_4(v,w)$ (stored as a $|\calS|^{n} \times |\calS|^{m}$ matrix).
 
@@
 
-The algorithm for a general tree is written out in Appendix \ref{ss:peeling_algorithm}.
+The algorithm for a general tree is written out in Appendix \ref{ss:pruning_algorithm}.
 
@@
 for doing statistical inference using observations of a particle model
-in which probabilities of change are determined only by nearby sites.
+in which probabilities of change are affected by nearby sites.
 Such models appear in many fields, for instance,
@@
 
-In other applications, reconstruction of an unknown sequence may of interest.
+In other applications, reconstruction of an unknown sequence may be of interest.
 For instance, training data might be used to estimate parameters of a noisy transmission line,
@@
 Neither of these assumptions are likely to be true for any data set of DNA sequence;
-however, both assumptions are also made the state of the art in phylogenetics.
+however, both assumptions are also commonly made in phylogenetics.
 Some relaxation of the assumptions is possible -- for instance,
@@
 Short deletions might be accommodated by introducing a ``gap'' letter to the alphabet,
-but insertions may be problematic.
+but insertions are problematic.
 
@@
 
-More things to do: cancer; include gaps; more taxa; leading/lagging strand...
-
 \subsection{Acknowledgements}
@@
 Many thanks are also due Jessica Crisci for curating the human-chimp data.
+The majority of this work was done while PR was at USC.
 
@@
     \caption{
-        As in Figure \ref{fig:biochem_results}, but with reduce vertical axis range
+        As in Figure \ref{fig:biochem_results}, but with reduced vertical axis range
         to see in more detail:
         estimated average mutation rates
-        for the motifs of Table \ref{tab:hominid_model}
+        for the motifs in the ``Hominid data'' section
         since the common ancestor of humans and chimpanzee.
@@
         then rates are in units of mutations per 6,000 years.
-        Shown are posterior medians (points) and 95\% credible intervals (lines, which are quite small in all cases).
+        Shown are posterior medians (points) and 95\% credible intervals (lines).
         \label{sfig:biochem_results_zoom}}
@@
 
-\paragraph{Full likelihood}
-Now we use these ideas to quantify and prove approximation~\eqref{eqn:full_approx}.
-For two positions $(i,s)$ and $(j,t)$ in sequence $\times$ time,
-let $r(i,s;j,t) = |j-i|/|t-s|$ be the absolute value of the
-slope of the line between them,
-and if $I$ and $J$ are subsets of sites, let $r(I,s;J,t) = \min\{r(i,s;j,t) : i\in I, j \in J\}$
-be the minimum.
-Above we have shown that there is a speed of dependency propagation $r_*$
-such that if $r(I,s;J,t) > r_*$,
-then $X_I(s)$ and $X_J(t)$ are nearly independent.
-Now, fix $t$ and for each site $i$ let $N_i$ denote its $r_*$-neighborhood,
-i.e.,
-\begin{align}
-    N_i = \{ (j,s) : r(i,s; j,t) \le r_*, \; j \leq N, \; 0 \le s \le t \} .
-\end{align}
-These are depicted in Figure~\ref{fig:neighborhoods}A.
-We will write $i \sim j$ if $|i-j| \le 2 r_* t$,
-or equivalently, if $N_i \cap N_j \neq \emptyset$,
-and $i \nsim j$ otherwise.
-We can think of $N_i$ as the parts of ancestral sequences in which a mutation might have affected the state at $(i,t)$.
-So, if $i \nsim j$, then $i$ and $j$ are (approximately) independent at time $t$, given the sequence at time 0.
-Concretely,
-we have a small $\epsilon$ such that for any $i \nsim j$,
-\begin{align*}
-    &\big|
-    \log( \P\{ X_i(t) = x_i \text{ and } X_j(t) = x_j \} ) \\
-    &\qquad {} -
-    \log( \P\{ X_i(t) = x_i \} \P\{ X_j(t) = x_j \} )
-    ) \big|
-    \le \epsilon .
-\end{align*}
-
-\begin{figure}
-    \begin{center}
-    %EM This looks like N12 to me. Am I miscounting?
-    %EM I suggest we include t and s in the diagram too.
-        \includegraphics{context-neighborhoods-defns}
-
-        \vspace{2em}
-
-        \includegraphics{context-neighborhoods-widths}
-    \end{center}
-    \caption{
-        %EM TODO make this A and B so we can refer to them in the text.
-        Diagram of things in the text:
-        (top) Dependency neighborhoods.
-        (bottom) What we need to compute $\P\{X_i(t)=x \given X(0), X_{<i}(t)\}$:
-        site $i$ depends on the blue squares in $N_i$;
-        the sites to the left of this in $L(i)$ are those other sites in $X(t)$
-        whose dependency neighborhoods overlap $N_i$;
-        any further sites to the left of this are basically independent of $X_i(t)$;
-        the red squares give the dependency neighborhood of $L(i)$,
-        so we need to condition on the row of red and blue sites at the top.
-        On the bottom, we have $2w+1$ sites; at the top we have $4w+1$;
-        here $w=2$.
-        \label{fig:neighborhoods}
-    }
-\end{figure}
-
-What we'd like to know is the likelihood, i.e.,
-\begin{align}
-    \like(y|x)
-    &:=
-    \P\{ X(t) = y \given X(0) = x \} .
-\end{align}
-If we define $X_{<i}=(X_1, \ldots, X_{i-1})$, then
-\begin{align}
-    \like(y|x)
-    &=
-    \prod_{i=1}^n
-    \P\{ X_i(t) = y_i \given X(0) = x \text{ and } X_{<i}(t) = y_{<i} \} .
-\end{align}
-Now, defining $L(i) = \{ j < i \st j \sim i \}$
-to be the sites to the left of $i$ whose dependency neighborhoods overlap with $i$'s,
-and $A(I) = \{ j \st |j-i| \le r_*t \text{ for some } i \in I\}$
-to be the ancestral sites in the dependency neighborhood of $I$,
-\begin{align}
-&\big|
-    \log( \P\{ X_i(t) = y_i \given X(0) = x \text{ and } X_{<i}(t) = y_{<i} \} ) \\
-& \qquad {} -
-    \log( \P\{ X_i(t) = y_i \given X_{A(L(i))}(0) = x_{A(L(i))} \text{ and } X_{L(i)}(t) = y_{L(i)} \} )
- \big| \\
-&\qquad \qquad \le
-    \epsilon ,
-\end{align}
-so that if we define
-\begin{align}
-    \alike(y|x)
-    &=
-    \prod_{i=1}^n
-    \P\{ X_i(t) = y_i \given X_{A(L(i))}(0) = x_{A(L(i))} \text{ and } X_{L(i)}(t) = y_{L(i)} \} ,
-\end{align}
-then (with $\loglike = \log \like$),
-\begin{align}
-    \left| \aloglike(y|x) - \loglike(y|x) \right|
-    \le n \epsilon .
-\end{align}
-
-%EM Would it be more direct to make a connection between
-%EM \P\{ X_i(t) = y_i \given X_{A(L(i))}(0) = x_{A(L(i))} \text{ and } X_{L(i)}(t) = y_{L(i)} \} ,
-%EM and
-%EM p_{2\ell+1,\ell,\ell}(x_{i-3\ell}^{(4\ell+1)}, y_{i-2\ell}^{(2\ell+1)})
-
-This is good, because we can compute $\aloglike$:
-let $w=r_*t$ be radius of a dependency neighborhood,
-so that the width of $L(i)$ is $2w$,
-and for a sequence $x$ define $x^{(m)}_i = (x_{i-m+1}, \ldots, x_i)$:
-\begin{align}
-    \alike(y|x)
-    &=
-    \prod_{i=1}^n
-    \frac{
-        p_{w,2w+1}(y^{(2w+1)}_i|x^{(4w+1)}_{i+w})
-    }{
-        p_{w,2w}(y^{(2w)}_{i-1}|x^{(4w)}_{i-1+w})
-    }
-\end{align}
-This is the ratio of the likelihoods of all the $(w,2w+1)$ T-mers,
-divided by the likelihoods of all the $(w,2w)$ T-mers.
-
 \section{Peeling algorithm}
-\label{ss:peeling_algorithm}
+\label{ss:pruning_algorithm}
 
@@
 
-  \item Set $M(b_i) = P$ for each $b_i$.
+  \item Set $M(b_i) = U$ for each $b_i$.
 
@@
 
-%%%%
-\section{Other likelihood functions}
-
-\subsection{Summary statistics}
-
-Since we cannot compute reasonably the full likelihood for the data,
-we will need to choose appropriate summary statistics
-that contain as much information about the process as possible,
-and have a tractable expression for their marginal likelihood.
-For simplicity,
-we assume in this section that approximation \eqref{eqn:window_approx} holds exactly.
-If $i_1 < \ldots < i_{n(x)}$ are a set of nonoverlapping sites where the initial sequence matches $x$
--- i.e., $X_{i_k}^{(2\ell+m)}(0)=x$ and $i_{k+1} > i_k + 2\ell+m$ for all $k$ %En dash close to math could be confusing. This sentence is a doozy.
-then the counts $N(x,y)=\#\{ i_k \st X_{i_k+\ell}^{(m)}(t)=y\}$ (the number of these sites that match the smaller pattern $y$ at time $t$)
-are multinomial, with probabilities $\{p_{\ell,m}(x,y)\st y \in \calS^m\}$.
-For instance, if $\ell=3$ then $N(\text{ACTCAC}, \text{CGC})$ counts the number of occurrences of
-\begin{center}
-\begin{tabular}{c|ccccccc}
- $x$ &  C  & A & C & T & C & A & C \\
- $y$ &  $\centerdot$  & $\centerdot$  & C & G & C & $\centerdot$  & $\centerdot$
-\end{tabular}
-\end{center}
-where $\cdot$ represents any base.
-More generally, if we partition the sequence into windows of length $2\ell+m$,
-and let $N(x,y)$ be the number of times we see a window in which $x$ matches $X(0)$ and $y$ matches $X(t)$,
-then these counts are independently multinomial across each $x$.
-% so the marginal likelihood of the statistics $N(x,y)$ is
-% \begin{align}
-%     \prod_{x \in \calS^{2\ell+m}} N(x)! \prod_{y \in \calS^m} \frac{p_{\ell,m}(x,y)^{N(x,y)}}{N(x,y)!} ,
-% \end{align}
-% where $N(x) = \sum_y N(x,y)$.
-
-Of course, a drawback to this approach is that we throw away a good fraction of the data.
-In fact, we do not even look at a fraction $2\ell/(2\ell+m)$ of the second sequence, $X(t)$,
-that lies in the outside edges of each window.
-As before, let $N(x)$ be the number of occurrences of a pattern $x$ in $X(0)$,
-and $N(x,y)$ the number of times $x$ is seen juxtaposed with $y$ in $X(t)$,
-but this time, count over all positions,
-so that e.g., $N(x) = \#\{ 1 \le i \le n - 2\ell - m + 1 \st X_i^{(m+2\ell)}(0)=x\}$.
-
-Now we will now calculate the expectation and covariances of $N(x,y)$.
-\peter{The following is Not Right, but also we don't really use it at the moment.}
-If we assume that \eqref{eqn:window_approx} is exact, then
-\[
-    \E [N(x,y)] = N(x) \, p_{\ell,m}(x,y).
-\]
-To compute covariances, define $\chi_i(x,y)$ to be 1 if $X_{i-\ell}^{(|x|)}(0)=x$ and $X_i^{(|y|)}(t)=y$, and 0 otherwise,
-so that $N(x,y) = \sum_{i=1}^n \chi_i(x,y)$.
-Overloading notation, also define $\chi_i(x)$ to be 1 if $X_{i-\ell}^{(|x|)}(0)=x$,
-\begin{align}
-  \var[N(x,y)] &= \sum_{i=1}^n \var[\chi_i(x,y)] + \sum_{i\neq j} \cov[\chi_i(x,y), \chi_j(x,y)] \\
-               &\approx \sum_{i=1}^n \chi_i(x) p_{\ell,m}(x,y)(1-p_{\ell,m}(x,y))
-            + 2 \sum_{i=1}^{n-2\ell-m} \sum_{j=i+1}^{i+2\ell+m-1} \cov[ \chi_i(x,y) \chi_j(x,y) ] \\
-            &= N(x) p_{\ell,m}(x,y)(1-p_{\ell,m}(x,y)) +
-\end{align}
-The term $\E[\chi_1(x,y) \chi_{k+1}(x,y)]$ is the probability that $x$ matches at both site 1 and site $k+1$,
-and that $y$ matches both at site $\ell+1$ and at site $\ell+k+1$.
-For this to be nonzero, the joint pattern $(x,y)$ must allow self-overlaps.
-Similarly, if $(x_1,y_1) \neq (x_2,y_2)$ then
-\begin{align}
-  \begin{split}
-    \cov[N(x_1,y_1),N(x_2,y_2)]
-    & \approx - n (2m+4\ell) p_{\ell,m}(x_1,y_1)p_{\ell,m}(x_2,y_2) \\
-    & \qquad {} + n\sum_{k=1}^{m+2\ell} \E[\chi_1(x_1,y_1) \chi_{k+1}(x_2,y_2) + \chi_1(x_2,y_2) \chi_{k+1}(x_1,y_1)] .
-  \end{split}
-\end{align}
-
-%%%%
-\subsection{A likelihood function}
-
-The covariance calculations show that
-if $p_{\ell,m}(x,y)$ is small enough (i.e., the pattern is not common),
-then $N(x,y)$ is close to Poisson.
-Using this observation, we can estimate parameters by composite likelihood,
-using all $N(x,y)$ as data.
-Specifically, this composite likelihood is
-\begin{align} \label{eqn:comp_like}
-  \prod_{x,y} F_{(x,y)}^{N(x,y)}
-\end{align}
-where $F$ is the $|\calS|^{m+2\ell} \times |\calS|^m$ matrix whose $(x,y)^\text{th}$ entry is $p_{m,\ell}(t,x,y)$ as described in the "Computation" section.
-Maximum likelihood then gets an asymptotically unbiased estimate of the parameters. (cite)
-However, since they are not independent, to get confidence intervals, we need to either simulate, or do something different.
-
-If we use only a subset of patterns $\{(x_k,y_k)\}$ chosen so that $\E[M_1(x,y) M_{k+1}(x,y)]$ is small (or zero),
-\erick{M not defined, but I'm guessing that this is effectively saying that the probability of having two different patterns apply to the same sequences is small?}
-the entire collection $\{N(x_k,y_k)\}$ will be close to independent Poisson.
-(using Stein's method)
-Then we'll have the correct likelihood function for (this subset of) data.
-
-If changes are rare, then one way to get a set of patterns with this property is to only choose ones such that going from $x$ to $y$ entails a change;
-and any overlapping patterns must require additional changes,
-i.e., if $(x_1,y_1)$ matches at $i_1$,
-then for $(x_2,y_2)$ to match at $i_2$,
-there must be additional changes outside $(i_1,\ldots,i_1+\ell-1)$.
-
-Here is one way to obtain such a set.
-Take a pair of patterns $(x,y)$, that overlap at $m$ positions,
-and suppose that they differ at $1 \le i_1, \ldots, i_d \le m$.
-Let $\bar i = (1/d) \sum_{j=1}^d i_j$ be the mean position of the changes,
-and take only patterns with $(m-1)/2 < \bar i \le (m+1)/2$.
-For example, in the example above with $m=3$ and $\ell=2$,
-\begin{center}
-\begin{tabular}{c|ccccccc}
- $x$ &  C  & A & C & T & C & A & C \\
- $y$ &  $\centerdot$  & $\centerdot$  & C & G & C & $\centerdot$  & $\centerdot$
-\end{tabular}
-\end{center}
-$x$ and $y$ differ only at (relative) position $i_1=2$.
-
-Another way to achieve the same goal of having all overlapping patterns require additional changes is to require $i_1 = \bar i = 1$.
-No two such patterns can be shifted versions of each other
-without adding more changed sites,
-since shifting by $k$ without adding changed sites just adds $k$ to $\bar i$.
-For example, we might take
-\begin{center}
-\begin{tabular}{c|ccccccc}
- $x$ &  A & C & T & C & A & C & G \\
- $y$ &  $\centerdot$  & $\centerdot$  & G & C & A & $\centerdot$  & $\centerdot$
-\end{tabular}
-\end{center}
-
-\section{Removed bits that we probably don't want}
-
-\subsection{s5f likelihood}
-
-The s5f model is defined as follows:
-for each $(a,b,c,d,e,x) \subset \{\nA,\nC,\nG,\nT\}$,
-it gives a (relative) mutation rate
-\begin{align}
-    \gamma(c,x;a,b,d,e) = \text{( mutation rate of $c \to x$ in context $abcde$ )} .
-\end{align}
-Given these rates, the only unknown parameter is the total amount of time
-that the sequence has been mutated for.
-Suppose that $X(t)$ is a sequence of length $n$ evolving under this model,
-and let $G_n$ be the corresponding generator matrix,
-a $4^n \times 4^n$ matrix.
-(In the s5f model, the mutation rates are only specified in the middle of 5-mers,
-so the two bases on each end never mutate.)
-The transition probabilities are $e^{tG_n}$,
-and by marginalization,
-we can compute quantities like
-the probability for observing the middle set of basis
-given the flanking $\ell$ on each end:
-\begin{align}
-    p_{\ell,n-2\ell}(t;a,b)
-    &=
-    \P\{
-        (X_{\ell+1}(t), X_{\ell+2}(t), \ldots, X_{n-\ell}(t))=(b_1,b_2,\ldots,b_{n-2\ell}) \\
-    &\qquad \qquad
-        \given (X_{1}(0), X_{2}(0), \ldots, X_{n}(0))=(a_1,a_2,\ldots,a_{n})
-    \} .
-\end{align}
-If the size of the context, $\ell$, is big enough,
-then this probability will not depend significantly if the sequence $a$ is in fact embedded in a longer sequence
-(the error decreases exponentially with $\ell$).
-
-The sequence $X(t)$ can be constructed by beginning with $X(0)$,
-putting down a Poisson process of potential mutations of rate $\gamma = \max\{\gamma(c,x;a,b,d,e)\}$,
-and thinning appropriately.
-The process depends on the flanking 2bp,
-so any other mutation within 2bp of a site affects the process at that site,
-and the chance that this occurs depends on the flanking 4bp on either side.
-Concretely, the number of flanking positions in each direction
-that must be kept track of to decide what happens at a given site
-grows by 1 at rate $\gamma$, and by 2 at rate $\gamma$;
-so the amount of flanking sequence on each side affecting a given site at time $t$
-is the sum of a Poisson($\gamma t$) and twice an independent Poisson($\gamma t$).
-This has mean $3\gamma t$ and variance $5 \gamma t$;
-with high probability it will not exceed, say,
-$3\gamma t + 6 \sqrt{\gamma t}$;
-if $\gamma t < 1$ this is less than 9.
-
-%%%%%%%%
-\subsection{Edge effects and cyclicization}
-
-The choice to treat long sequences as cyclical is generally of little importance.
-However, we have defined in \eqref{eqn:G_defn} the basic building block of our approximation, $G(m)$,
-to be the transition matrix \emph{for a circular sequence} of length $m$, where $m$ may be short.
-\peter{no we haven't}
-This seems strange -- but, what else would we do at the edges?
-We want to use $G(\ell+m+r)$ to compute the chance that a given $(\ell+m+r)$-sequence ends up with another given $m$-sequence in the middle;
-and if $\ell$ and $r$ are big enough relative to $t$, this will be a good approximation regardless of what we do at the edges.
-One answer would be to not allow patterns hanging off of either end to match,
-resulting in a slight underestimate of the chance of change.
-A better approximation would be to use the marginal frequencies in some way to average over surrounding sequence
--- but, this requires marginal frequencies to be known, and these are not stationary.
-However, the leftmost site in a randomly chosen window is a sample from the marginal distribution,
-and if sites separated by distance $\ell+m+r$ are sufficiently uncorrelated,
-wrapping the sequence around is a good approximation to averaging over marginal frequencies in the surrounding sequence.
-It is easy to imagine situations where this would not be the best solution --
-but again, if $\ell$ and $r$ are large enough, it won't matter much.
-
-%%%%
-\subsection{The root distribution}
-
-We clearly don't want to add an extra parameter $\pi(z)$ for all $z \in \calS^{m+2\ell}$.
-If nothing else, it would make the inference depend on $m$ and $\ell$, which we'd like to avoid.
-One approach would be to put a Dirichlet distribution on $\pi(z)$.
-This has the nice property that you can choose priors that are consistent across different values of $m$ and $\ell$.
-We'd have to still MCMC around in the space of $\pi(z)$, so they'd be nuisance parameters from that point of view.
-Also, making the frequencies at the root random introduces correlations between counts, which we want to avoid.
-An alternative is to specify $\pi$ through lower-dimensional statistics,
-for instance to model the root state as a sample from a Markov random field
-with parameters for the marginal frequencies of each state, and pairwise (or higher) correlations between them.
-
-A Markov random field with interactions between sites no more than distance two away
-is determined by the energy function $\phi(a,b,c)$ which gives the ``energy'' of the configuration $abc$ for each $a,b,c \in \calS$.
-We use this to define the distribution on root patterns $\pi$ by
-\begin{align}  \label{eqn:root_mrf}
-    \pi(z) = \frac{1}{Z} \exp\left( - \sum_{i=1}^{|z|} \phi(z_{i-1},z_i,z_{i+1}) \right),
-\end{align}
-i.e., the exponential of the total energy of each three adjacent sites,
-where $Z$ is the normalizing constant.
-Note that it is often difficult to compute $Z$ in applications of Markov random fields,
-but here we are working with relatively short sequences,
-so we just compute \eqref{eqn:root_mrf} for all possible $z$, and divide by the sum.
-(However, for this same reason this is not equivalent to saying that $X(0)$ is a sample from the Markov random field
-and finding the distribution of patterns within this.)
-
-%%%%%
-\subsubsection{Mixing}
-
-There can be significant confounding of the frequencies at the root with the other parameters,
-which results in poor mixing.
-(See CpG results, below.)
-To see why this is, consider the following back-of-the-envelope calculation.
-Take $x \neq y$ to be single bases (so, set $\ell=0$ and $w=1$),
-so that $N(x,y)$ gives the number of times an $x$ and a $y$ are homologous.
-Let $q(x,y)$ be the instantaneous rate that $x \to y$ (averaged somehow over context?),
-so that $N(x,y)$ should be, roughly,
-\begin{align*}
-  \sum_z \pi(z) q(z,x) q(z,y) &\approx t \pi(x) (1+q(x,x)) q(x,y) + t \pi(y) q(y,x) (1+q(y,y)) ,
-\end{align*}
-where as usual $q(x,x) = (-1) \sum_{z \neq x} q(x,z)$ is (minus) the total jump rate out of $x$.
-
-So, when we propose a change to $\pi$ we might want to also adjust the rates accordingly to keep this more-or-less constant:
-if we chnage $\pi(x) \mapsto \pi(x) + \epsilon_x$, we want $\delta_{xy}$ to satisfy
-\begin{align*}
-  \sum_z \pi(z) q(z,x) q(z,y) = \sum_z (\pi(z) + \epsilon_z) ( q(z,x) + \delta_{zx} ) ( q(z,y) + \delta_{zy} ) ,
-\end{align*}
-or, to first order in $\delta$,
-\begin{align*}
-  0 &\approx \sum_z \epsilon_z q(z,x) q(z,y) + \sum_z \pi(z) \delta_{zx} q(z,y) + \sum_z \pi(z) q(z,x) \delta_{zy} \\
-   &\approx \sum_z \epsilon_z q(z,x) q(z,y) + \sum_z (\pi(z)+\epsilon_z) \delta_{zx} q(z,y) + \sum_z (\pi(z)+\epsilon_z) q(z,x) \delta_{zy} \\
-   &\approx \epsilon_x (1+q(x,x)) q(x,y) + \epsilon_y q(y,x) (1+q(y,y)) + (\pi(y)+\epsilon_y) \delta_{yx} (1+q(y,y)) + (\pi(x)+\epsilon_x) (1+q(x,x)) \delta_{xy} \\
-   &= \left( \epsilon_x q(x,y) + (\pi(x)+\epsilon_x) \delta_{xy} \right)(1+q(x,x)) + \left( \epsilon_y q(y,x) + (\pi(y)+\epsilon_y) \delta_{yx} \right)(1+q(y,y))
-\end{align*}
-One solution to this is to let $\delta_{xy} = - \epsilon_x q(x,y) / \pi(x)$.
-
-Suppose each change is a single-base change, so that changes $\delta \mu_i$ in $\mu$ contribute to $q$ by
-$\delta_{xy} = \sum_{i: x \to y} \delta \mu_i$.
-Given $\delta_{xy}$, this gives a linear equation for $\delta \mu_i$ that may or may not be solvable,
-but certainly is if we have all the single-base transitions as parameters.
-
-\subsection{Models without using potential}
-
-Here is a reformulation of the CpG and Ising models without using a potential.
-
-\begin{example}[CpG mutation]
-    \peter{version without selection}
-  A sequence of genome can be written using A, C, G, and T;
-  in the most general model of independent mutation across sites, each of the 12 possible transitions occurs at its own rate.
-  Furthermore, it is a well-known observation that in many species,
-  adjacent, methylated CG dinucleotides (``CpG sites'') have a much higher mutation rate to TG and CA
-  than either single nucleotide change under the independent model.
-  Embellishing the single-nucleotide model with this additional rate results in the model defined by
-
-  \begin{center}
-    \begin{tabular}{c@{\quad$\to$\quad}c@{\quad at rate\quad }cc}
-      $x$  &  $y$  &  $m_{xy}$ & when $x \neq y \in \{\nA,\nC,\nG,\nT\}$  \\
-      \nC\nG   &  \nT\nG   &  $\gamma$ & \\
-      \nC\nG   &  \nC\nA   &  $\gamma$ &
-    \end{tabular} .
-  \end{center}
-  Here $\gamma$ is the additional CpG rate above the base mutation rate,
-  and a total of 13 parameters.
-  Note that this model is very different in spirit to TASEP,
-  in which the total number of particles was conserved.
-
-  In this model, for a given pair of sequences, there may be more than one way to mutate a given string $x$ to get $y$:
-  for instance, a change $\nA\nC\nG \to \nA\nT\nG$ could have occurred by a single $\nC \to \nT$ mutation,
-  or by a $\nC\nG \to \nT\nG$ mutation;
-  and hence the total instantaneous rate at which ACG changes to ATG is $\gamma + m_{\nC\nT}$.
-  Although this means there are many possible parameterizations,
-  note that the model is identifiable, not overparameterized.
-
-\end{example}
-
-\begin{example}[1-D Ising model with Glauber dynamics]
-    \peter{version without selection}
-  In the Ising model, each site is labeled as either ``up'' or ``down'' ($+1$ or $-1$ respectively),
-  imagined as magnetic dipoles,
-  and the energy associated with a given state $x$ is $H(x) = - \frac{1}{2} \beta \sum_i x_i x_{i+1} - \frac{1}{2} \gamma \sum_i x_i$.
-  Here $\beta$ represents inverse temperature, and $\gamma$ represents the strength of the magnetic field (here scaled by temperature).
-  The associated stationary distribution on configurations is proportional to $\exp(-H(x))$.
-  The following ``Glauber dynamics'' preserve the stationary distribution:
-  each site, independently at rate $\lambda$,
-  forgets its spin,
-  and reconfigures to a state chosen with probability proportional to the stationary probability of the resulting configuration.
-  Ignoring transitions that don't change the state,
-  if patterns $u$ and $v$ differ at only one site then $u$ changes to $v$ at rate $\lambda/(1+\exp(H(v)-H(u)))$,
-  so
-
-    \peter{reformulate with a potential}
-
-  \begin{center}
-    \begin{tabular}{c@{\quad$\to$\quad}c@{\quad at rate\quad }c}
-      $+++$  &   $+-+$   &  $\lambda/\,(1+e^{2\beta + \gamma})$ \\
-      $++-$  &   $+--$   &  $\lambda/(1+e^{\gamma})$ \\
-      $+-+$  &   $+++$   &  $\lambda/(1+e^{-2\beta - \gamma})$ \\
-      $+--$  &   $++-$   &  $\lambda/(1+e^{-\gamma})$ \\
-      $-++$  &   $--+$   &  $\lambda/(1+e^{\gamma})$ \\
-      $-+-$  &   $---$   &  $\lambda/(1+e^{-2\beta + \gamma})$ \\
-      $--+$  &   $-++$   &  $\lambda/(1+e^{-\gamma})$ \\
-      $---$  &   $-+-$   &  $\lambda/(1+e^{2\beta - \gamma})$
-    \end{tabular}
-  \end{center}
-
-\end{example}
-
-\subsection{Longer models}
-      \begin{center}
-          \begin{tabular}{c@{\;\;$\to$\;\;}c@{\;\; {\small at rate}\;\; }c|c@{\;\;$\to$\;\;}c@{\;\; {\small at rate}\;\; }c|c@{\;\;$\to$\;\;}c@{\;\; {\small at rate}\;\; }c}
-            \nT\nC     &   \nT\nT      &  $\mu_{UV1}$    & \nC\nG       &   \nC\nT       &  $\mu_{g}$       & \nA\nA\nC\nT  &   \nA\nA\nT\nT   &  $\mu_{\AID}$ \\
-            \nG\nA     &   \nA\nA      &  $\mu_{UV1}$    & \nC\nG       &   \nA\nG       &  $\mu_{g}$       & \nA\nA\nC\nC  &   \nA\nA\nT\nC   &  $\mu_{\AID}$ \\
-            \nC\nC     &   \nC\nT      &  $\mu_{UV2}$    & \nT\nC\nA    &   \nT\nT\nA    &  $\mu_{\APOBEC}$ & \nA\nG\nC\nT  &   \nA\nG\nT\nT   &  $\mu_{\AID}$ \\
-            \nG\nG     &   \nA\nG      &  $\mu_{UV2}$    & \nT\nC\nT    &   \nT\nT\nT    &  $\mu_{\APOBEC}$ & \nA\nG\nC\nC  &   \nA\nG\nT\nC   &  $\mu_{\AID}$ \\
-            \nC\nC     &   \nT\nT      &  $\mu_{UV3}$    & \nT\nC\nA    &   \nT\nG\nA    &  $\mu_{\APOBEC}$ & \nA\nA\nC\nT  &   \nA\nA\nT\nT   &  $\mu_{\AID}$ \\
-            \nG\nG     &   \nA\nA      &  $\mu_{UV3}$    & \nT\nC\nT    &   \nT\nG\nT    &  $\mu_{\APOBEC}$ & \nA\nA\nC\nC  &   \nA\nA\nT\nC   &  $\mu_{\AID}$ \\
-            \nT\nC\nG  &   \nT\nT\nG   &  $\mu_{mUV}$    & \nT\nG\nA    &   \nT\nA\nA    &  $\mu_{\APOBEC}$ & \nA\nG\nC\nT  &   \nA\nG\nT\nT   &  $\mu_{\AID}$ \\
-            \nC\nC\nG  &   \nC\nT\nG   &  $\mu_{mUV}$    & \nA\nG\nA    &   \nA\nA\nA    &  $\mu_{\APOBEC}$ & \nA\nG\nC\nC  &   \nA\nG\nT\nC   &  $\mu_{\AID}$ \\
-            \nC\nG\nA  &   \nC\nA\nA   &  $\mu_{mUV}$    & \nT\nG\nA    &   \nT\nC\nA    &  $\mu_{\APOBEC}$ & \nT\nA\nC\nT  &   \nT\nA\nT\nT   &  $\mu_{\AID}$ \\
-            \nC\nG\nG  &   \nC\nA\nG   &  $\mu_{mUV}$    & \nA\nC\nA    &   \nA\nC\nA    &  $\mu_{\APOBEC}$ & \nT\nA\nC\nC  &   \nT\nA\nT\nC   &  $\mu_{\AID}$ \\
-            \nT\nA     &   \nT\nG      &  $\mu_{\eta}$   & \nA\nA       &    \nA\nG      &  $\mu_{\iota1}$  & \nT\nG\nC\nT  &   \nT\nG\nT\nT   &  $\mu_{\AID}$ \\
-            \nT\nA     &   \nC\nA      &  $\mu_{\eta}$   & \nT\nT       &    \nC\nT      &  $\mu_{\iota1}$  & \nT\nG\nC\nC  &   \nT\nG\nT\nC   &  $\mu_{\AID}$ \\
-            \nA\nA     &   \nG\nT      &  $\mu_{\iota2}$ & \nT\nG\nC\nT &  \nT\nG\nT\nT  &  $\mu_{\AID}$    & \nT\nA\nC\nT  &   \nT\nA\nT\nT   &  $\mu_{\AID}$ \\
-            \nT\nT     &   \nA\nC      &  $\mu_{\iota2}$ & \nT\nG\nC\nC &  \nT\nG\nT\nC  &  $\mu_{\AID}$    & \nT\nA\nC\nC  &   \nT\nA\nT\nC   &  $\mu_{\AID}$
-        \end{tabular}
-      \end{center}
-
 \end{document}

### 2021-12-09 writeups/notes.tex
edits and tweaks

--- 2013-06-12 context-dependence.tex
@@
 \usepackage{graphicx}
+\usepackage{natbib}
+\usepackage[hidelinks]{hyperref}
+\usepackage{bm}
+\usepackage{todonotes}
 
@@
 \newcommand{\calT}{\mathcal{T}}  % set of transition triples
+\newcommand{\calP}{\mathcal{P}}  % set of selection pairs
+\newcommand{\calM}{\mathcal{M}}  % set of potential mutations
 \newcommand{\nA}{\mbox{A}}  % nucleotides:
@@
 \newcommand{\nT}{\mbox{T}}
-\newcommand{\join}{\oplus}  % matches
+\newcommand{\nS}{\mbox{S}} % C or G
+\newcommand{\nW}{\mbox{W}} % A or T
+\newcommand{\nR}{\mbox{R}} % purine, A or G
+\newcommand{\nY}{\mbox{Y}} % pyrimidine, C or T
+\newcommand{\join}{\oplus}  % concatenation
 \newcommand{\st}{\colon}  % such that
@@
 \newcommand{\cov}{\mathop{\mbox{cov}}}
+\newcommand{\given}{\;\mid\;}
+\newcommand{\like}{\mathcal L}
+\newcommand{\loglike}{\ell}
+\newcommand{\alike}{\widetilde{\like}}
+\newcommand{\aloglike}{\widetilde{\loglike}}
 
-\theoremstyle{plain}
-\newtheorem{theorem}{Theorem}
-\theoremstyle{definition}
-\newtheorem{example}{Example}[section]
+\newcommand{\APOBEC}{\text{\tiny APOBEC}}
+\newcommand{\AID}{\text{\tiny AID}}
 
@@
 
-\section*{Introduction}
-
-Interacting particle systems with neighborhood structure: transition probabilities generally not expressable (except TASEP).
-Markov random fields with Glauber dynamics a case of this.
-
-Example: Ising model.
-
-Context-dependent substitution processes are an example of interest for phylogenetics and evolutionary inference.
-Usually dealt with either by using larger and larger windows (but, only up to four bp),
-or by doing MCMC over additional information (entire history of changes).
-Also, Markov-along-the-genome method.
-
-The difficulty comes from the fact that the rates of change at any one site depend in principle on arbitrarily far away sites,
-so transition matrix is ginormous.
-
-Goal: parameter inference, given before-after observations, or else observations descended from a common ancestor.
-
-Review stats lit, e.g. reviews by Friel, original by Besag (autologistic model), genetics by Larribe and Fearnead.
-
-Here restrict to one-dimensional neighborhood structure,
-to make it easier.
-
-\section{Notation and Examples}
-
-The common framework of the above examples is a one-dimensional grid of $n$ sites,
-and each site is labeled with one of a finite set of states.
-These change with time -- so, write $X_i(t)$ for the state of the $i^\mathrm{th}$ site at time $t$,
-where each $X_i(t) \in \calS$, the set of possible states.
-(To avoid edge effects, we sometimes arrange these sites in a circle, so the indices are interpreted $\mod n$.)
-The time evolution is Markov, and is determined by a set of transition rates and associated patterns.  
-``Patterns'' are sequences of states, and we let $|u|$ denote the length of the pattern $u$.
-Transition rates are assumed to be local and homogeneous, so for example saying ``pattern $u$ changes to $v$ at rate $\mu$''
-means that if $(X_i(t), \ldots, X_{i+\ell}(t)) = (u_1, \ldots, u_\ell)$ (where $\ell=|u|=|v|$),
-then $\P\{(X_i(t+dt), \ldots, X_{i+\ell}(t+dt)) = (v_1, \ldots, v_\ell) = \mu dt + o(dt)$,
-regardless of the location $i$ along the sequence.
-We call such a $(\mu,u,v)$ a ``transition triple''.
-The set of all transition triples $\calT = \{ (\mu_j,u_j,v_j) \}_{j=1}^{n_T}$ 
-determines the time evolution of the process.
-
-It will often be convenient, notationally and computationally, to avoid edge effects
-by treating the $n$ sites as a circle, and treating all indices $\mod n$.
-
-\begin{example}[TASEP]
-  The \emph{Totally Asymmetric Simple Exclusion Process} labels each site as ``empty'' or ``occupied'' (0 or 1, respectively), 
-  and says that each occupied site, independently at rate $\lambda$, checks the site to the right to see if it is empty,
-  and if it is, moves there.  
-  This process therefore has only one transition triple:
-
-  \begin{center}
-    \begin{tabular}{c@{\quad$\to$\quad}c@{\quad at rate\quad }c}
-      $u$  &  $v$  &  $\mu$  \\
-      \hline
-      01  &   10   &  $\lambda$
-    \end{tabular}
-  \end{center}
-
-  This only has one parameter, the speed.  
-  Note, however, that it may not be entirely trivial to estimate the speed,
-  since (on the circle) it may not be obvious which particle moved where.
-
-\end{example}
-
-\begin{example}[CpG mutation]
-  A sequence of genome can be written using A, C, G, and T;
-  in the most general model of single-site mutation, each of the 12 possible transitions occurs at its own rate.
-  Furthermore, it is a well-known observation that in many species, and certain contexts, 
-  a C followed by a G has an additional, higher rate with either the C changing to a T or the G changing to an A.
-  This simplest model of context-dependent substitution rates in a genome sequence has rates
-
-  \begin{center}
-    \begin{tabular}{c@{\quad$\to$\quad}c@{\quad at rate\quad }cc}
-      $u$  &  $v$  &  $\mu$  & \\
-      \hline
-      $x$  &  $y$  &  $m_{xy}$ & \qquad $x \neq y \in \{\nA,\nC,\nG,\nT\}$  \\
-      CG   &  TG   &  $\gamma$ & \\
-      CG   &  CA   &  $\gamma$ &  
-    \end{tabular}
-  \end{center}
-
-  This has 13 parameters.
-  Note that there are no conserved quantities here, unlike in TASEP,
-  which we could imagine as indestructible particles moving.
-
-\end{example}
-
-\begin{example}[Ising model with Glauber dynamics]
-  In the Ising model, each site is labeled as either ``up'' or ``down'' ($+1$ or $-1$ respectively),
-  imagined as magnetic dipoles,
-  and the energy associated with a given state $x$ is $H(x) = - \beta \sum_i x_i x_{i+1} - \gamma \sum_i x_i$.
-  The associated stationary distribution on configurations is proportional to $\exp(-H(x))$.
-  One way to add temporal dynamics that preserve the distribution 
-  is to say that each site, independently at rate $\lambda$,
-  forgets its spin, 
-  and reconfigures to a state chosen with probability proportional to the stationary probability of the resulting configuration.
-  This is known as ``Glauber dynamics'', and ignoring transitions that don't change the state,
-  we get that if patterns $u$ and $v$ differ at only one site, then $u$ changes to $v$ at rate $1/(1+\exp(H(u)-H(v)))$, so
-
-  \begin{center}
-    \begin{tabular}{c@{\quad$\to$\quad}c@{\quad at rate\quad }c}
-      $u$  &  $v$  &  $\mu$  \\
-      \hline
-      $+++$  &   $+-+$   &  $\lambda/1+e^{2\beta - \gamma})$ \\
-      $++-$  &   $+--$   &  $\lambda/(1+e^{-\gamma})$ \\
-      $+-+$  &   $+++$   &  $\lambda/(1+e^{-2\beta + \gamma})$ \\
-      $+--$  &   $++-$   &  $\lambda/(1+e^{\gamma})$ \\
-      $-++$  &   $--+$   &  $\lambda/(1+e^{-\gamma})$ \\
-      $-+-$  &   $---$   &  $\lambda/(1+e^{-2\beta - \gamma})$ \\
-      $--+$  &   $-++$   &  $\lambda/(1+e^{\gamma})$ \\
-      $---$  &   $-+-$   &  $\lambda/(1+e^{2\beta + \gamma})$ 
-    \end{tabular}
-  \end{center}
-
-  This has three parameters: the speed $\lambda$, the inverse temperature $\beta$, and the strength of magnetization $\gamma$
-  (here scaled by temperature).
-
-\end{example}
-
-\paragraph{The generator matrix}
-For clarity, we should define how the set of transition rates determines the transition rate matrix,
-the $|\calS|^n \times |\calS|^n$ matrix $G(n)$ whose $(x,y)^\text{th}$ entry gives the instantaneous rate 
-with which the process in state $x$ jumps to state $y$.
-Let $x_i^{(\ell)} = (x_i, x_{i+1}, \ldots, x_{i+\ell-1})$ be the subsequence of length $\ell$ beginning at location $i$,
-and for each $1\le i \le n$, and patterns $u,v \in \calS^\ell$ define the relation
-\[
-x \xrightarrow{i,u,v} y \qquad \text{iff} \qquad \begin{cases}
-  x_j = y_j \quad &\text{for } j<i \\
-  x_{i+k} = u_k \quad &\text{for } k < \ell \\
-  y_{i+k} = v_k \quad &\text{for } k < \ell, and \\
-  x_j = y_j \quad &\text{for } j\ge i+\ell ,
-\end{cases}
-\]
-i.e.\ if $x$ and $y$ match except for in $i,i+1,\ldots,i+\ell$, where $x$ matches with $u$ and $y$ matches with $v$.
-Interpet these indices $\mod n$, even for short sequences, a point we will come back to.
-We'll also need the set of transitions that can move from $x$ to $y$ at a given location: let  $J(i,x,y) = \{ j \st x \xrightarrow{i,u_j,v_j} y \}$.
-Then, the rate $G(n)_{x,y}$ is the sum of all matching transition rates,
-namely
-\begin{align} \label{eqn:G_defn}
-  G(n)_{x,y} = \sum_{i=1}^n \sum_{j \in J(i,x,y)}  \mu_j ,
-\end{align}
-and if there are no triples $(\mu,u,v)$ with $x \xrightarrow{i,u,v} y$ for some $i$, then $G(n)_{x,y}=0$.
-In principle, this gives us the transition probabilities for the process:
-\begin{align} \label{eqn:full_likelihood}
-  p_n(t;x,y) := \P\{ X(t) = y \mid X(0) = x \} = (e^{tG_n})_{xy} .
-\end{align}
-
-\section{Simulation}
-
-First, it will be useful to have a brief description of how to simulate such a process.
-Briefly, we can put down a Poisson process of times and locations of possible changes,
-at an appropriate rate $\mu_*$,
-and then resolve each possible change in temporal order.
-We situate ``changes'' at the beginning of the matching pattern,
-so $\mu_*$ can be defined as the maximum sum of rates matching  the start of a sequence:
-\[
-\mu_* = \max \left\{ \sum_{y \in \calS^n} \sum_{j\in J(1,x,y)} \mu_j \st x \in \calS^n \right\} .
-\]
-Now sample possible changes: let $s_{ij}$ denote the $j^\text{th}$ possible change at site $i$,
-so that $0 \le s_{i1} < s_{i2} < \cdots < s_{iN_i} \le t$ is a rate $\mu_*$ Poisson process.
-Then, suppose we have determined the state at time $s$ to be $X(s) = x$,
-and the next event is $s_{ij} = \min \{ s_{k\ell} : s_{k\ell}>s \}$.
-The process then jumps to a new state at times $s_{ij}$ with probability proportional to the corresponding rate.
-In other words, let
-\[
-q_k = \begin{cases}
-  \mu_k/\mu_* \qquad & \text{if } x_i^{(|u_k|)} = u_k  \\
-  0 \qquad & \text{otherwise,}
-\end{cases}
-\]
-and let $q_0 = 1-\sum_{k=1}^{n_T} q_k$.
-By construction of $\mu_*$, $q_0\ge 0$.)
-Now choose transition $k$ with probability $q_k$, and if $k>0$, replace $x_i^{|u_k|}$ with $v_k$.
-
-For instance, to simulate TASEP, one has only to let $\mu_*=\lambda$, and at each $s_{ij}$
-check if at site $i$ there is a 1 followed by a 0,
-and if so, switch them.
-
-\section{Inference}
-
-The problem at hand is to infer the parameters of the model, given the state at time 0 and later at time $t$.
-If $n$ is large enough and $t$ is not too large,
-this should be feasible.
-(If $t$ is too large, the process may ``saturate'' --
-if there are many changes at most sites, then in at least most models, 
-we will lost most of the information about the dynamics,
-retaining information only about the parameters that affect the stationary distribution.
-In TASEP, we lose all information as it approaches stationarity,
-in the Ising model we have information about $\beta$ and $\gamma$ but not $\lambda$,
-and in the CpG model it is not immediately clear.
-
-But, how can we extract the information?
-These are Markov processes, on the state space $\calS^n$, 
-so the full likelihood function is given by \eqref{eqn:full_likelihood},
-but doing anything with the $|\calS|^n \times |\calS|^n$ matrix $G_n$ is clearly infeasible.
-We can, however, compute \eqref{eqn:full_likelihood} for smaller $n$,
-so the first thing that one might think to do is to break the sequence up into many blocks of length $m$,
-and treat these as independent, taking $m$ as large as is feasible.
-Then we would define $p_m(t;x,y)$ to be the probability that a string $x$ of length $m$ evolves to string $y$ over time $t$,
-denote by $x_i^{(m)}$ the block of length $m$ beginning at index $i$,
-and obtain the approximate likelihood function
-\[
-  \prod_{k=0}^{n/m} p_m(x_{km}^{(m)},y_{km}^{(m)}) .
-\]
-Variations are possible (what do we do about the edge effects for $p_m$?),
-but in any case there is no guarentee that this is close to correct,
-as it ignore dependencies between neighboring blocks.
-
-Intuitively, we want to increase the size of the context 
--- but increasing $m$ does not do this, as there are always edge effects.
-However, we could look at the context in the initial sequence but not in the final sequence,
-including the states of an additional $\ell$ and $r$ sites on the left and right, respectively.
-Concretely, define the marginal probabilities
-\begin{align}
-  p_{\ell,m,r}(t;x,y) &= \sum_{a \in \calS^\ell} \sum_{b \in \calS^r} p_{\ell+m+r}(t;x,a \join y \join b) , \quad \text{for}\; x \in \calS^{\ell+m+r}, \; y \in \calS^m ,
-\end{align}
-where $a \join y \join b$ is the sequence composed of concatenating $a$, $y$, and $b$ together in that order.
-This is the probability that if we begin with a sequence $x$ of length $\ell+m+r$, 
-then after time $t$ we will see at the middle $m$ positions the pattern $y$,
-ignoring what is present in the leftmost $\ell$ and rightmost $r$ positions.
-In a long sequence, $p_m(x,z)$ might not be a very good approximation for $\P\{ X_i^{(m)}(t) = z | X_i(0)^{(m)} = x \}$,
-we do expect $p_{\ell,m,r}(x,y)$ to be correct, asymptotically in $\ell$ and $r$:
-\begin{align} \label{eqn:window_approx}
-  \P\{ X_i^{(m)}(t) = y | X_i(0)^{(\ell+m+r)} = x \} \approx p_{\ell,m,r}(x,y) .
-\end{align}
-We provide a concrete statement of this approximation later.
-
-%%%%
-\subsection{Summary statistics}
-
-Now we know how to compute the (marginal) probability that a particular subsequence changes into another subsequence.
-For patterns $x$ and $y$, with offset $\ell$ (so that $|x|=|y|+2\ell$),
-we can count how many times $x$ appears in $X(0)$,
-and how many times $x$ appears in $X(0)$ while $y$ appears at the same place in $X(t)$, offset by $\ell$.
-For instance, if $x$ is ACTCAC, $y$ is CGC, and $\ell=3$, this looks like:
-\begin{center}
-\begin{tabular}{c|ccccccc}
- $x$ &  C  & A & C & T & C & A & C \\
- $y$ &  $\centerdot$  & $\centerdot$  & C & G & C & $\centerdot$  & $\centerdot$  
-\end{tabular}
-\end{center}
-where $\cdot$ could be any base.
-Let $N(x)$ and $N_\ell(x,y)$ count these events, respectively.
-
-From the definitions above, 
-the expected number of times $y$ would appear juxtaposed with $x$ is approximately $N(x)p_{\ell,m,r}(x,y)$.
-Covariances are also easy to compute.
-If we define $M_i(x,y)$ to be 1 if $X_{i-\ell}^{|x|}(0)=x$ and $X_i^({|y|})(t)=y$, and 0 otherwise,
-so that $N(x,y) = \sum_{i=1}^n M_i(x,y)$, then
-\begin{align}
-  \var[N(x,y)] &= \sum_{i=1}^n \var[M_i(x,y)] + \sum_{i\neq j} \cov[M_i(x,y), M_j(x,y)] \\
-  &= n \var[M_1(x,y)] + 2n\sum_{k=1}^n \cov[M_1(x,y),M_{k+1}(x,y)] \\
-  &\approx n p_{\ell,m}(x,y)(1-p_{\ell,m}(x,y)) + 2n\sum_{k=1}^{m+2\ell} \left( \E[M_1(x,y) M_{k+1}(x,y)] - p_{\ell,m}(x,y)^2 \right) \\
-  &= n p_{\ell,m}(x,y)(1-(1+2m+4\ell)p_{\ell,m}(x,y)) + 2n\sum_{k=1}^{m+2\ell} \E[M_1(x,y) M_{k+1}(x,y)] .
-\end{align}
-The term $\E[M_1(x,y) M_{k+1}(x,y)]$ is the probability that $x$ matches at both site 1 and site $k+1$,
-and that $y$ matches both at site $\ell+1$ and at site $\ell+k+1$.
-For this to be nonzero, the joint pattern $(x,y)$ must allow self-overlaps.
-Similarly, if $(x_1,y_1) \neq (x_2,y_2)$ then
-\begin{align}
-  \begin{split}
-    \cov[N(x_1,y_1),N(x_2,y_2)] 
-    & \approx - n (2m+4\ell) p_{\ell,m}(x_1,y_1)p_{\ell,m}(x_2,y_2) \\
-    & \qquad {} + n\sum_{k=1}^{m+2\ell} \E[M_1(x_1,y_1) M_{k+1}(x_2,y_2) + M_1(x_2,y_2) M_{k+1}(x_1,y_1)] .
-  \end{split}
-\end{align}
-
-%%%%
-\subsection{A likelihood function}
-
-The covariance calculations show that
-if $p_{\ell,m}(x,y)$ is small enough (i.e.\ the pattern is not common),
-then $N(x,y)$ is close to Poisson.
-Using this observation, we can estimate parameters by composite likelihood,
-using all $N(x,y)$ as data.
-Maximum likelihood then gets an asymptotically unbiased estimate of the parameters.
-However, since they are not independent, to get confidence intervals, we need to either simulate, or do something different.
-
-If we use only a subset of patterns $\{(x_k,y_k)\}$ chosen so that $\E[M_1(x,y) M_{k+1}(x,y)]$ is small (or zero),
-the entire collection $\{N(x_k,y_k)\}$ will be close to independent Poisson.
-(using Stein's method)
-Then we'll have the correct likelihood function for (this subset of) data.
-If changes are rare, then one way to get a set of patterns with this property is to only choose ones such that going from $x$ to $y$ entails a change;
-and any overlapping patterns must require additional changes, 
-i.e.\ if $(x_1,y_1)$ matches at $i_1$, 
-then for $(x_2,y_2)$ to match at $i_2$, 
-there must be additional changes outside $(i_1,\ldots,i_1+\ell-1)$.
-
-Here is one way to obtain such a set.
-Take a pair of patterns $(x,y)$, that overlap at $m$ positions,
-and suppose that they differ at $1 \le i_1, \ldots, i_d \le m$.
-For example, in the example above with $m=3$ and $\ell=2$,
-\begin{center}
-\begin{tabular}{c|ccccccc}
- $x$ &  C  & A & C & T & C & A & C \\
- $y$ &  $\centerdot$  & $\centerdot$  & C & G & C & $\centerdot$  & $\centerdot$  
-\end{tabular}
-\end{center}
-$x$ and $y$ differ only at (relative) position $i_1=2$.
-Let $\bar i = (1/d) \sum_{j=1}^d i_j$ be the mean position of the changes,
-and take only patterns with $(m-1)/2 < \bar i \le (m+1)/2$.
-As another example,
-\begin{center}
-\begin{tabular}{c|ccccccc}
- $x$ &  A & C & T & C & A & C & G \\
- $y$ &  $\centerdot$  & $\centerdot$  & G & C & A & $\centerdot$  & $\centerdot$  
-\end{tabular}
-\end{center}
-has $i_1 = \bar i = 1$.
-No two such patterns can be shifted versions of each other
-without adding more changed sites,
-since shifting by $k$ without adding changed sites just adds $k$ to $\bar i$.
-
-%%%%
-\subsection*{Proof of the approximation}
-
-\begin{proof}
-
-In fact, the approximation improves exponentially as $\ell$ and $m$ increase --
-here is a simple argument.
-Suppose the maximum length of any transition pattern is $R = \max\{ |u| : (\mu,u,v) \in \calT \}$.
-The transition rate at site $i$ at time $s$ is determined if we know $X_{i-R}^{(2R+1)}(s)$;
-this can be determined from $X_{i-R}^{2R+1}(0)$ and the outcome of all changes happening before time $s$ at sites in $(i-R-1, \ldots, i+R)$.
-These will depend in turn on the outcome of previous possible changes.
-For each site $i$ and times $s_1 < s_2$, define the set of states $H(i,s_1,s_2)$ 
-so that if one knows the possible change points $s_{ij}$,
-then changes outside of $H(i,s_1,s_2)$ do not affect the distribution of $X_i(s_2)$.
-Concretely, define $H(i,s_1,s_2)$ by the property that $i \in H(i,s_1,s_2)$, and
-\begin{gather}
-  \text{if } s_1 < s_{jk} \le s_2 \text{ for some } i-R-1 \le j \le i+R, \text{ then } H(j,s_1,s_{jk}) \subseteq H(i,s_1,s_2) .
-\end{gather}
-In other words, $H(i,s_1,s_2)$ is the union of the $R$-neighborhoods of $i$ and all possible changes occurring in $H(i,s_1,s_2)$ between $s_1$ and $s_2$.
-For a set of states $I$, let $H(I,s_1,s_2) = \bigcup_{i\in I} H(i,s_1,s_2)$.
-Formally, the state $X_i(s_2)$ is conditionally independent of $\{X_j(s_1) \st j \notin H(i,s_1,s_2)\}$.
-Informally, the approximation \eqref{eqn:window_approx} will be good if $H$ does note expand beyond the flanking context.
-Now, MAKE THIS EASIER TO READ
-\begin{multline}
-  \P\{ X_i^{(m)}(t) = y \mid X(0) \} 
-  = \P\left\{ H(\{i,\ldots,i+m\},0,t)\subseteq\{i-\ell,\ldots,i+m+r\} \right\} \\
-   \qquad \qquad {} \times \P\left\{ X_i^{(m)}(t) = y \mid X_{i-\ell}^{(\ell+m+r)}(0), \; H(\{i,\ldots,i+m\},0,t)\subseteq\{i-\ell,\ldots,i+m+r\} \right\} \\
-     \qquad {} + \P\left\{ H(\{i,\ldots,i+m\},0,t)\nsubseteq\{i-\ell,\ldots,i+m+r\} \right\}  \\
-   \qquad\qquad {} \times \P\left\{ X_i^{(m)}(t) = y \mid X_{i-\ell}^{(\ell+m+r)}(0), \; H(\{i,\ldots,i+m\},0,t)\nsubseteq\{i-\ell,\ldots,i+m+r\} \right\} ,
-\end{multline}
-and so
-\begin{align} \label{eqn:prob_approx}
-  \left| \P\{ X_i^{(m)}(t) = y \mid X(0) \} - p_{\ell,m,r}(t;x,y) \right| \le 2 \P\left\{  H(\{I,\ldots,i+m\},0,t)\subseteq\{i-\ell,\ldots,i+m+r\} \right\}.
-\end{align}
-
-So, the error is bounded by the probability that $H(I,0,t)$ is larger than the context from $I-\ell$ to $I+r$.
-However, following $H$ backwards in time, the outer edge of $H$ expands as a continuous-time random walk:
-for instance, the left edge $L(s) = \min \{ j \in H(I,t-s,t) \}$ jumps leftwards at rate $R\mu_*$ with a random jump size
-that has mean $R/2$.
-General facts about random walks imply that \eqref{eqn:prob_approx} is bounded above by $C e^{-\alpha \ell/t}$, for some constants $C$ and $\alpha$.
-
-\end{proof}
+Here are some bits removed from the main text that might provide useful context.
 
@@
 to be the transition matrix \emph{for a circular sequence} of length $m$, where $m$ may be short.
+\peter{no we haven't}
 This seems strange -- but, what else would we do at the edges?
@@
 resulting in a slight underestimate of the chance of change.
-A better approximation would be to use the marginal frequencies in some way to average over surroudning sequence 
+A better approximation would be to use the marginal frequencies in some way to average over surrounding sequence
 -- but, this requires marginal frequencies to be known, and these are not stationary.
@@
 
-%%%%%%%%
-\section{Phylogenetic inference}
-
-In phylogenetic applications, rather than ``before'' and ``after'' observations,
-we get two (or more) observations evolved from a common root.
-In the simplest case of two taxa we have two processes $X$ and $Y$,
-with identical starting states $X(0)=Y(0)$,
-observed only at times $t_X$ and $t_Y$ respectively.
-
-If the approximation described above holds,
-then we can write the probability that we see (long) pattern $x$ in $X(t_X)$ juxtaposed with (short) pattern $y$ in $Y(t_Y)$
-using the pattern frequencies at the root.
-Concretely: pick a random location $I$ in the sequence and let $X_{I-\ell}^{m+2\ell}(0) = \rho$ be the (long) pattern seen there at the root,
-and write $\pi(z)$ for the frequency of pattern $z$ in $X(0)$, i.e.\ $\P\{\rho=z\}=\pi(z)$.
-% Given the corresponding pattern in $X(t_X)$, the distribution of $\rho$ is proportional to
-% \[
-%     \P\{ \rho=z \mid X_{I-\ell}^{m+2\ell}(t_X)=x\} \propto \pi(z) p_{m+2\ell}(t,z,x) .
-% \]
-The probability that we see $x$ and $y$ at the random location $I$ is
-\begin{align} \label{eqn:phylo_likelihood}
-    \P\{X_{I-\ell}^{m+2\ell}(t_X)=x \text{ and } Y_I^\ell(t_Y)=y \} = \sum_{z \in \calS^{m+2\ell}} \pi(z) p_{m+2\ell}(t_X,z,x) p_{m,\ell}(t_Y,z,y) .
-\end{align}
-This can be computed without much more effort than the simpler case above,
-but requires the frequencies at the root.
-
+%%%%
 \subsection{The root distribution}
@@
 \begin{align}  \label{eqn:root_mrf}
-    \pi(z) = \frac{1}{Z} \exp\left( - \sum_{i=1}^{|z|} h(z_{i-1},z_i,z_{i+1}) \right),
+    \pi(z) = \frac{1}{Z} \exp\left( - \sum_{i=1}^{|z|} \phi(z_{i-1},z_i,z_{i+1}) \right),
 \end{align}
-i.e.\ the exponential of the total energy of each three adjacent sites,
+i.e., the exponential of the total energy of each three adjacent sites,
 where $Z$ is the normalizing constant.
@@
 
-%%%%%%%%%%%
-\section{Computation}
+%%%%%
+\subsubsection{Mixing}
 
-For inference, we need to compute the $|\calS|^{m+2\ell} \times |\calS|^m$ matrix whose $(x,y)^\text{th}$ entry is $p_{m,\ell}(t,x,y)$,
-for various $t$.
-This matrix is a projection of the $|\calS|^{m+2\ell} \times |\calS|^{m+2\ell}$ matrix whose $(x,z)^\text{th}$ entry is
-\begin{align}
-    p_{m+2\ell}(x,z) = \left( e^{t G(m+2\ell)} \right)_{x,z} ,
-\end{align}
-where $G(m+2\ell)$ is the sparse $|\calS|^{m+2\ell} \times |\calS|^{m+2\ell}$ matrix defined in \eqref{eqn:G_defn}.
-The ``projection'' we need just marginalizes over long patterns $z$ that match the shorter pattern $y$:
-if we define the matrix $U$ in this way, so that $U_{zy}=1$ if $z_\ell^m=y$ and $U_{zy}=0$ otherwise,
-then
-\[
-    p_{m,\ell}(t,x,y) = \sum_{z \in \calS^{m+2\ell}} \left( e^{t G(m+2\ell)} \right)_{xz} U_{zy} ,
-\]
-so in fact we don't need the entire matrix $e^{t G(m+2\ell)}$,
-just the product of this matrix with each of the columns of $U$.
+There can be significant confounding of the frequencies at the root with the other parameters,
+which results in poor mixing.
+(See CpG results, below.)
+To see why this is, consider the following back-of-the-envelope calculation.
+Take $x \neq y$ to be single bases (so, set $\ell=0$ and $w=1$),
+so that $N(x,y)$ gives the number of times an $x$ and a $y$ are homologous.
+Let $q(x,y)$ be the instantaneous rate that $x \to y$ (averaged somehow over context?),
+so that $N(x,y)$ should be, roughly,
+\begin{align*}
+  \sum_z \pi(z) q(z,x) q(z,y) &\approx t \pi(x) (1+q(x,x)) q(x,y) + t \pi(y) q(y,x) (1+q(y,y)) ,
+\end{align*}
+where as usual $q(x,x) = (-1) \sum_{z \neq x} q(x,z)$ is (minus) the total jump rate out of $x$.
 
-Modern techniques in sparse matrix computation (Krylov methods) provide efficient ways to do this
-by only multiplying $G$ and $U$ some number of times, 
-which is much more efficient because $G$ is sparse.
+So, when we propose a change to $\pi$ we might want to also adjust the rates accordingly to keep this more-or-less constant:
+if we chnage $\pi(x) \mapsto \pi(x) + \epsilon_x$, we want $\delta_{xy}$ to satisfy
+\begin{align*}
+  \sum_z \pi(z) q(z,x) q(z,y) = \sum_z (\pi(z) + \epsilon_z) ( q(z,x) + \delta_{zx} ) ( q(z,y) + \delta_{zy} ) ,
+\end{align*}
+or, to first order in $\delta$,
+\begin{align*}
+  0 &\approx \sum_z \epsilon_z q(z,x) q(z,y) + \sum_z \pi(z) \delta_{zx} q(z,y) + \sum_z \pi(z) q(z,x) \delta_{zy} \\
+   &\approx \sum_z \epsilon_z q(z,x) q(z,y) + \sum_z (\pi(z)+\epsilon_z) \delta_{zx} q(z,y) + \sum_z (\pi(z)+\epsilon_z) q(z,x) \delta_{zy} \\
+   &\approx \epsilon_x (1+q(x,x)) q(x,y) + \epsilon_y q(y,x) (1+q(y,y)) + (\pi(y)+\epsilon_y) \delta_{yx} (1+q(y,y)) + (\pi(x)+\epsilon_x) (1+q(x,x)) \delta_{xy} \\
+   &= \left( \epsilon_x q(x,y) + (\pi(x)+\epsilon_x) \delta_{xy} \right)(1+q(x,x)) + \left( \epsilon_y q(y,x) + (\pi(y)+\epsilon_y) \delta_{yx} \right)(1+q(y,y))
+\end{align*}
+One solution to this is to let $\delta_{xy} = - \epsilon_x q(x,y) / \pi(x)$.
 
-\subsection{Updating $G$}
-
-We can also take advantage of the sparsity of $G$ to perform efficient computation of the likelihood
-under many sets of parameters.
-First note that $G(m+2\ell)$ has, assuming only single-position changes,
-$(1+m(|\calS|-1)) |\calS|^{m+2\ell}$ nonzero entries, since each of the $|\calS|^{m+2\ell}$ can change in $m$ places.
-This will determine how the computation scales with $m$, $\ell$, and $|\calS|$,
-once we precompute a number of things.
-Let $g = (g_1, \ldots, g_d)$ be the nonzero entries of $G$, in some fixed ordering;
-sparse matrix representations of $G$ store only $g$ along with information about the rows and columns these are found in.
-Each $g_i$ is a linear combination of mutation rates $\mu_j$, say $g_i = \sum_j a_{ij} \mu_j$;
-so by precomputing $a_{ij}$ we can efficiently update $G$.
-This is true if the boundary conditions are circular, mean-value, or neither.
-
-\subsection{Phylogenetic peeling}
-
-Furthermore, note that the further calculations in the phylogenetic case can be done using these methods.
-To compute
-\[
-    \sum_z \pi(z) p_{m+2\ell}(t_X,z,x) p_{m,\ell}(t_Y,z,y) 
-    = \sum_z \left( e^{t_X G(m+2\ell)} \right)_{zx} \pi(z) \sum_w \left( e^{t_X G(m+2\ell)} \right)_{zw} U_{wy} ,
-\]
-we can first compute $\sum_w \left( e^{t_X G(m+2\ell)} \right)_{zw} U_{wy}$ as above,
-multiply by $\pi(z)$, and then apply the algorithm again using the transpose of $G(m+2\ell)$.
-
-Using the R package expm, this makes computation quite feasible:
-with four possible states and $\ell=2$ and $m=1$, so that $G(m+2\ell)$ is a $1024 \times 1024$ matrix,
-computing $e^{t G(5) }$ takes 13 seconds, while computing $e^{t G(5)} U$ takes only 0.3 seconds.
-Increasing to $\ell=4$ and $m=1$ or $\ell=3$ and $m=2$ is still feasible, taking $e^{t G(8)} U$ in 42 and 47 seconds, respectively
-(and much longer for the whole matrix $e^{t G(8)}$).
-
-%%%%%%
-\section{Bayesian things}
-
-So, suppose we've got a list of patterns $(x_i,y_i)$
-and we're happy with the idea that they're independent.
-Recall that
-\[
-\E\left[ N(x,y) \mid N(x) \right] = N(x) p_{m,\ell}(x,y) ,
-\]
-Parameters $\theta$ enter implicitly through $p_{m,\ell}$.
-
-In the case that we observe the initial sequence, $N(x,y)$ are binomial with sample size $N(x)$.
-and so the likelihood is 
-\[
-\prod_x N(x)! \prod_y \frac{ p_{m,\ell}(x,y)^{N(x,y)} }{ N(x,y)! } ,
-\]
-and so the negative log likelihood is, up to a constant,
-\[
-\sum_{x,y} N(x,y) \log p_{m,\ell}(x,y) .
-\]
-
-%%%%%%%
-\section{Simulation results}
-
-The method works well to find point estimates of the parameters.
-Obtaining posterior density estimates using the approximation likelihood function on a subset of patterns
-works in some cases;
-in other cases the likelihood surface is still too peaked, underestimating the uncertainty.
-
-\subsection{TASEP}
-
-Strings of length $10^5$ were evolved under TASEP for varying amounts of time,
-and pattern occurrences counted.
-Running 10,000 MCMC iterations with window sizes of $\ell=w=3$ (total length 9) on these takes around 40 minutes.
-Apparently, the patterns used were not sufficiently independent in this case,
-as the posterior is too sharp.
-
-Here's a short example sequence with $\lambda=1$:
-\begin{center}
- \setlength{\tabcolsep}{0pt}
-\begin{tabular}{cccccccccccccccccccccccccccccccccccccccccccccccccccccccccccc}
-X&X&X&O&O&X&O&O&O&O&X&O&X&O&X&X&O&O&X&O&O&X&O&X&X&X&O&X&O&O&O&X&X&O&X&O&O&O&X&X&O&O&O&X&O&O&O&O&X&O&X&X&O&X&X&O&O&X&O&O \\
-$\centerdot$&O&X&O&X&O&O&O&X&$\centerdot$&O&X&O&O&X&X&X&o&x&$\centerdot$&o&O&X&x&x&O&X&O&X&o&o&O&X&X&$\centerdot$&$\centerdot$&o&o&O&X&X&o&$\centerdot$&O&X&$\centerdot$&o&o&$\centerdot$&$\centerdot$&O&X&X&x&O&X&o&O&X&$\centerdot$
-\end{tabular}
-\end{center}
-here particles are X's, moving to the right; and spaces are O's; 
-in the second line, 
-$\centerdot$ is a position that has not considered a change; lowercase letters are positions that considered a change but did not;
-and upper case are positions that did change.
-
-\begin{figure}
-  \begin{center}
-    \includegraphics[width=\textwidth]{tasep/all-mcmc-runs}
-  \end{center}
-  \caption{ Results from MCMC estimates for six different data sets of the parameter in TASEP (mutation rate multiplied by time).  
-  Vertical line shows the true value.
-  }
-\end{figure}
-
-\subsection{Ising model}
-
-Strings of length $10^6$ were evolved under Glauber dynamics on the Ising model with varying parameters,
-and pattern occurrences counted.
-As for TASEP, unning 10,000 MCMC iterations with window sizes of $\ell=w=3$ (total length 9) on these takes around 40 minutes.
-In this case, the true parameter values have good distribution within the estimated posterior distributions.
-
-Here is an example, with all parameters equal to 1, for $t=.5$:
-\begin{center}
- \setlength{\tabcolsep}{0pt}
-\begin{tabular}{cccccccccccccccccccccccccccccccccccccccccccccccccccccccccccc}
-X&X&O&X&X&O&O&X&O&X&O&X&X&O&O&X&O&O&O&O&X&X&X&O&X&O&X&O&X&O&O&O&X&X&O&X&X&X&X&X&O&X&O&X&O&X&X&X&X&O&O&X&X&X&X&X&X&O&O&O \\
-x&x&o&x&X&X&O&x&o&x&$\centerdot$&x&x&O&X&X&$\centerdot$&O&X&O&X&x&x&o&x&o&x&o&x&O&X&O&$\centerdot$&$\centerdot$&O&O&X&X&O&X&X&X&o&x&o&$\centerdot$&X&O&X&o&o&x&X&X&X&O&X&o&$\centerdot$&$\centerdot$ 
-\end{tabular}
-\end{center}
-The notation is as above; except that the entire matched pattern involved in a change is uppercase:
-to simulate, we look in windows of size 3; so if XOO changes to XXO, the whole thing is uppercase, not just the middle X.
-
-\begin{figure}
-  \begin{center}
-    \includegraphics{writeup-plots/selsims-2013-05-28-17-12-0275615-traces}
-  \end{center}
-  \caption{ 
-  MCMC traces of the parameters in the Ising model across 100,000 iterations.
-  The units of $\lambda$ are in mean number of changes per site.
-  Vertical line shows the true value.
-  }
-\end{figure}
-
-\begin{figure}
-  \begin{center}
-    \includegraphics[width=\textwidth,page=1]{ising/all-mcmc-runs}
-  \end{center}
-  \caption{ Results from MCMC estimates of the time-scaled mutation rate parameter $\lambda t$,
-  in ten different simulated datasets under the Ising model.
-  Vertical line shows the true value.
-  }
-\end{figure}
-
-\begin{figure}
-  \begin{center}
-    \includegraphics[width=\textwidth,page=2]{ising/all-mcmc-runs}
-  \end{center}
-  \caption{ Results from MCMC estimates of the inverse temperature $\beta$ in the Ising model. 
-  Vertical line shows the true value.
-  }
-\end{figure}
-
-\begin{figure}
-  \begin{center}
-    \includegraphics[width=\textwidth,page=3]{ising/all-mcmc-runs}
-  \end{center}
-  \caption{ Results from MCMC estimates of the magnetization parameter $\beta$ in the Ising model. 
-  Vertical line shows the true value.
-  }
-\end{figure}
-
-%%%
-\subsection{CpG mutation, on a tree}
-
-Nucleotide sequences of length $10^6$ were simulated down two (unequal) branches of a tree,
-with all single-base substitution rates $m_{xy}$ set equal (varying from a total of .01 to .1 across the tree)
-and an additional CpG rate $\gamma$ set to either 0 or three times the single-base rate.
-Analysis was done with $\ell=2$ and $w=1$ (a five-to-one window).
-
-The state at the root of the tree was independently chosen bases with equal frequencies, 
-and these frequencies at the root were added as parameters to the model (with a Dirichlet prior).
-
-Here is a short example, with all parameters equal to 1, with each branch of the tree having length $0.1$:
-\begin{center}
- \setlength{\tabcolsep}{0pt}
-\begin{tabular}{cccccccccccccccccccccccccccccccccccccccccccccccccccccccccccc}
-T&G&G&A&G&A&G&G&T&C&T&G&C&C&A&T&A&G&A&C&G&G&G&C&G&C&C&G&C&C&T&A&G&G&A&G&C&C&G&C&G&T&T&T&T&G&C&G&T&G&G&T&G&T&T&G&T&G&G&C \\
-$\centerdot$&$\centerdot$&g&C&G&a&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&T&A&C&A&$\centerdot$&$\centerdot$&$\centerdot$&T&A&T&$\centerdot$&$\centerdot$&$\centerdot$&T&G&c&C&C&A&C&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&G&G&G&t&A&T&$\centerdot$&$\centerdot$&G&A&g&C&G&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&t&T&G&A \\
-$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&A&T&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&T&T&G&T&$\centerdot$&$\centerdot$&G&A&T&$\centerdot$&$\centerdot$&$\centerdot$&G&G&c&t&A&A&G&C&$\centerdot$&$\centerdot$&$\centerdot$&G&T&G&A&t&t&t&g&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&G&A&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&A&C
-\end{tabular}
-\end{center}
-Top is ancestral sequence (``unobserved''); bottom two are at the tips.
-
-The point estimates are good, 
-as are the MCMC estimates,
-but there seems to be mixing trouble having to do with the root frequencies.
-
-\begin{figure}
-  \begin{center}
-    \includegraphics{writeup-plots/selsims-2013-06-03-13-17-0790276-initfreqs}
-  \end{center}
-  \caption{ 
-  MCMC traces across 660,000 iterations of the frequencies at the root of the (two-taxon) tree in the CpG model.
-  Vertical line shows the true value.
-  }
-\end{figure}
-
-\begin{figure}
-  \begin{center}
-    \includegraphics{writeup-plots/selsims-2013-06-03-13-17-0790276-mutrates}
-  \end{center}
-  \caption{ 
-  MCMC traces across 660,000 iterations of several mutation rate parameters in the CpG model.
-  Units are in mean number of substitutions per base across the entire tree.
-  Vertical line shows the true value.
-  }
-\end{figure}
-
-%%%%%%%%
-\section{What next}
-
-\begin{enumerate}
-
-  \item Allow more than single-base distribution at the root.
-
-  \item What's up with the mixing problem in root frequencies?
-
-  \item See if counting in nonoverlapping windows fixes the peakiness problem with TASEP.
-
-  \item See what's feasible with a larger alphabet -- 60 codons -- and more parameters.
-
-  \item Show that taking no context ($\ell=0$) can be positively misleading in some context.
-
-  \item Do a three-taxa tree.
-
-  \item How to identify motifs in the residuals?
-
-\end{enumerate}
+Suppose each change is a single-base change, so that changes $\delta \mu_i$ in $\mu$ contribute to $q$ by
+$\delta_{xy} = \sum_{i: x \to y} \delta \mu_i$.
+Given $\delta_{xy}$, this gives a linear equation for $\delta \mu_i$ that may or may not be solvable,
+but certainly is if we have all the single-base transitions as parameters.
 

### 2021-12-09 writeups/context-dependence.tex
further work

@@
 \usepackage{todonotes}
+\usepackage{authblk}
+\renewcommand\Affilfont{\itshape\small}
 
@@
 
-% Model-based inference neighborhood dependence
-% Nearby sites affecting the likelihood
-% Rigorous inference with neighborhood dependency
-% Statistical inference with local/neighborhood interdependency
-% It depends on the context:
-%
-
+% TODO include github information
+
+%EM I'm going to suggest "dependence propagation" rather than information
 \title{Inference using context-dependent models of mutation via bounds on information propagation}
-\author{}
+\author[1]{Peter L Ralph}
+\author[2,3,4,5]{Frederick A Matsen IV}
+\author[1,*]{Peter L Ralph}
+
+\affil[1]{Institute of Ecology \& Evolution, Department of Biology, University of Oregon}
+\affil[2]{Computational Biology Program, Fred Hutchinson Cancer Research Center, United States}
+\affil[3]{Department of Genome Sciences, University of Washington}
+\affil[4]{Department of Statistics, University of Washington}
+\affil[5]{Howard Hughes Medical Institute}
+\affil[*]{Corresponding author: plr@uoregon.edu}
 
@@
     Although the rates at which positions in the genome mutate
-    are known to depend not only on the nucleotide to be mutated
+    are known to depend not only on the nucleotide to be mutated,
     but also on neighboring nucleotides,
@@
     to faraway locations,
-    making it difficult to compute exactly likelihoods.
+    making it difficult to compute exact likelihoods.
     This paper shows how to use bounds on the propagation of dependency
@@
     \caption{
-        An overview of the method for a neighbor-dependent model.
-        The red trapezoidal region represents the ``range of influence'',
+        An overview of the method for a neighbor-dependent model, showing an
+        initial sequence on top, a collection of mutations progressing through
+        time, and a final sequence.
+        The pink trapezoidal region represents the ``range of influence'',
         i.e., possible locations and times at which a mutations is likely
@@
         for this subsequence
-        conditioned on the ancestral state at the \emph{overhang}
-        of the top of the T.
+        conditioned on the ancestral state at the additional \emph{overhang}
+        sites of the top of the T.
         \label{fig:overview}
@@
 at least arbitrarily.
-%EM do we really want x and y here, which I've been thinking of as being the full sequence?
-%EM Note we'll need to change a and b below.
-%PR I think this is okay? x and y are not exactly the same here, but are still sequences?
-%PR To me it seems less confusing than introducing another variable?
 To do this, let $p_{n,\ell,r}(x,y)$ for $\ell + r < n$ denote
@@
 is found to match a smaller subsequence $y$ of length $n-\ell-r$, offset by $\ell$ sites
-(see Figure \ref{Tmer-dependency}).
+(see Figure \ref{fig:Tmers}).
 This central, smaller subsequence is called the ``base'' of the T-mer.
@@
 \emph{and} the outcomes of any mutations that have occurred within that window
-(see Figure \ref{Tmer-dependency}).
+(see Figure \ref{fig:Tmers}).
 
@@
 and if the sequence has evolved for time $t$ then we expect no more than a Poisson($\lambda t$) number of changes at each site.
+%EM suggest "every site will have 75\% chance of changing", which is a little different than what you say because of multiple hits, right?
 If $\lambda t = 0.75$ (three-quarters of the sequence will have changed!),
@@
 \begin{align} \label{eqn:phylo_likelihood}
+    %EM I think the superscript of Y_I should be (m)
     \P\{X_{I-\ell}^{(n)}(t_X)=x \text{ and } Y_I^{(\ell)}(t_Y)=y \}
@@
 We might approximate the full likelihood as follows,
+%EM In the previous full likelihood section, we discuss choosing l, r, and n, and use those in the equations.
+%EM This makes the equations look a little different.
+Do we want to just clarify that we are writing the equations for general l, r, n, and that we refer the reader to the previous choice to choose these quantities?
 assuming for the sake of notational simplicity that we have a few positions
@@
 
-%EM I think this should be pruning, not peeling.
-%PR I'll take your word for it! Wikipedia suggests either would do?
-%EM Want to check the modifications made above before continuing those to here.
 \subsection{Phylogenetic pruning}
@@
     \begin{center}
+    %EM We need to put these superscripts in parens, no? Also, should we use n instead of x+2l?
     \includegraphics{pruning-schematic}
@@
 To compute the likelihood on a more complicated tree, consider the following,
-with the three-taxon tree of figure \ref{fig:pruning} as an example,
+with the three-taxon tree of Figure~\ref{fig:pruning} as an example,
 where we are counting occurrences of (long) $n$-tuples in taxon $X$,
@@
       =
+      %EM I think this should be n, ell, r, not m, ell, r
       \sum_{u \in \calS^{n}} \pi(u) p_{m,\ell,r}(t_C,u,z)
@@
 
+%EM I'm not really following this section. If we are only calculating likelihood for places where y and z are the same, why is phylogenetic inference useful?
 In total there are $|\calS|^{3m+2\ell}$ possible data combinations.
@@
             \begin{tabular}{c@{\quad$\to$\quad}c@{\quad at rate\quad }c|c@{\quad$\to$\quad}c@{\quad at rate\quad }c|c@{\quad$\to$\quad}c@{\quad at rate\quad }c}
-                \nT  &   \nA   &  $\mu_{AT}$ & \nA  &   \nG   &  $\mu_{r}$ & \nA  &   \nG   &  $\mu_{v}$ \\
-                \nA  &   \nT   &  $\mu_{AT}$ & \nG  &   \nA   &  $\mu_{r}$ & \nG  &   \nA   &  $\mu_{v}$ \\
-                \nC  &   \nG   &  $\mu_{CG}$ & \nC  &   \nT   &  $\mu_{r}$ & \nT  &   \nC   &  $\mu_{v}$ \\
-                \nG  &   \nC   &  $\mu_{CG}$ & \nT  &   \nC   &  $\mu_{r}$ & \nC  &   \nT   &  $\mu_{v}$ \\
+                %EM I think there were some typos in this table that I fixed.
+                \nT  &   \nA   &  $\mu_{AT}$ & \nA  &   \nC   &  $\mu_{r}$ & \nA  &   \nG   &  $\mu_{v}$ \\
+                \nA  &   \nT   &  $\mu_{AT}$ & \nC  &   \nA   &  $\mu_{r}$ & \nG  &   \nA   &  $\mu_{v}$ \\
+                \nC  &   \nG   &  $\mu_{CG}$ & \nG  &   \nT   &  $\mu_{r}$ & \nT  &   \nC   &  $\mu_{v}$ \\
+                \nG  &   \nC   &  $\mu_{CG}$ & \nT  &   \nG   &  $\mu_{r}$ & \nC  &   \nT   &  $\mu_{v}$ \\
             \end{tabular}
@@
 some length 2 residuals had statistically significant residuals,
+%EM In my compilation there is a float that interrupts this flow.
 but none so large as those seen in (3,1,1) T-mer residuals, shown here:
@@
   \end{center}
-  \caption{ Results from iterative fitting to a $10^6$bp sequence
+  \caption{ Results from iterative fitting to a $10^6$ bp sequence
     evolved to have 28.5\% sequence divergence
@@
         Shown are posterior medians (points) and 95\% credible intervals (lines, which are quite small in all cases).
-        Figure \ref{sfig:biochem_results_zoom} shows the smaller rates in more detail,
-        and figure \ref{sfig:biochem_results_relative} shows the rates between categories relative to each other.
+        Figure~\ref{sfig:biochem_results_zoom} shows the smaller rates in more detail,
+        and Figure~\ref{sfig:biochem_results_relative} shows the rates between categories relative to each other.
         \label{fig:biochem_results}}
@@
 fast enough to allow Bayesian inference from Monte Carlo sampling.
-The method is fast thanks to its dependence only only local neighborhoods
+The method is fast thanks to its dependence on only local neighborhoods
 (so the time does not scale with sequence length once T-mer abundances are counted)
@@
 Simon Tavar\'e, Rasmus Nielsen, Graham Coop, Yaniv Brandvain, and Sergey Nuzhdin.
-Many thanks are also due Jessica Crisci for curating the human-chimp data.
+Many thanks are also due to Jessica Crisci for curating the human-chimp data.
 The majority of this work was done while PR was at USC.
 
-%EM TODO grant acknowledgements
+FAM's work supported by NIH grants R01 AI146028 and U01 AI150747.
+Dr.\ Matsen is an Investigator of the Howard Hughes Medical Institute.
 

### 2021-12-09 writeups/context-dependence.tex
erick edits

@@
 and if the sequence has evolved for time $t$ then we expect no more than a Poisson($\lambda t$) number of changes at each site.
-%EM suggest "every site will have 75\% chance of changing", which is a little different than what you say because of multiple hits, right?
-If $\lambda t = 0.75$ (three-quarters of the sequence will have changed!),
+If $\lambda t = 0.75$ ($1-e^{-3/4}=52\%$ of the sequence will have changed!),
 then taking $\ell = r = 3$ allows us to compute T-mer probabilities to within an error of 0.007,
@@
 \begin{align} \label{eqn:phylo_likelihood}
-    %EM I think the superscript of Y_I should be (m)
-    \P\{X_{I-\ell}^{(n)}(t_X)=x \text{ and } Y_I^{(\ell)}(t_Y)=y \}
+    \P\{X_{I-\ell}^{(n)}(t_X)=x \text{ and } Y_I^{(m)}(t_Y)=y \}
     =
@@
 We might approximate the full likelihood as follows,
-%EM In the previous full likelihood section, we discuss choosing l, r, and n, and use those in the equations.
-%EM This makes the equations look a little different.
-Do we want to just clarify that we are writing the equations for general l, r, n, and that we refer the reader to the previous choice to choose these quantities?
-assuming for the sake of notational simplicity that we have a few positions
+after choosing $n$, $\ell$, and $r$ appropriately (see above).
+For the sake of notational simplicity, assume that we have a few positions
 of $X$ observed before the start and after then end of the sequence
-(so that, e.g., we can refer to $x_{n+1}, \ldots, x_{n+r}$):
+(so that, e.g., we can refer to $x_{n+1}, \ldots, x_{n+r}$).
 \begin{align*}
@@
     \begin{center}
-    %EM We need to put these superscripts in parens, no? Also, should we use n instead of x+2l?
     \includegraphics{pruning-schematic}
@@
       =
-      %EM I think this should be n, ell, r, not m, ell, r
-      \sum_{u \in \calS^{n}} \pi(u) p_{m,\ell,r}(t_C,u,z)
-      \sum_{v \in \calS^{n}} p_{n,0,0}(t_{AB},u,v) p_{n,0,0}(t_{A},v,x) p_{m,\ell,r}(t_B,v,y)
+      \sum_{u \in \calS^{n}} \pi(u) p_{n,\ell,r}(t_C,u,z)
+      \sum_{v \in \calS^{n}} p_{n,0,0}(t_{AB},u,v) p_{n,0,0}(t_{A},v,x) p_{n,\ell,r}(t_B,v,y)
   \end{split}
@@
 
-%EM I'm not really following this section. If we are only calculating likelihood for places where y and z are the same, why is phylogenetic inference useful?
 In total there are $|\calS|^{3m+2\ell}$ possible data combinations.
-However, we can reduce dimensionality by replacing the last step with
+However, we can reduce dimensionality by only computing probabilities
+for certain (more common) patterns: for instance,
+we'd compute the likelihood for situations where $y$ and $z$ agree (but may differ from $x$
+by replacing step (4) with:
 \begin{enumerate}
@@
 \end{enumerate}
-i.e., only computing the likelihood for cases with $y=z=w$.
-On a three-taxon tree it is not clear that this is desirable, since missing out on double transitions
-might be a significant loss of information,
+On a three-taxon tree it is not clear that this is desirable,
 but on larger trees it seems likely that we'd want to restrict to combinations with most sites agreeing across the tree.
@@
             \begin{tabular}{c@{\quad$\to$\quad}c@{\quad at rate\quad }c|c@{\quad$\to$\quad}c@{\quad at rate\quad }c|c@{\quad$\to$\quad}c@{\quad at rate\quad }c}
-                %EM I think there were some typos in this table that I fixed.
                 \nT  &   \nA   &  $\mu_{AT}$ & \nA  &   \nC   &  $\mu_{r}$ & \nA  &   \nG   &  $\mu_{v}$ \\
@@
 under the model given in Table \ref{tab:cpg_results},
-obtained by embellishing the basic CpG hypermutability
+obtained by embellishing the basic CpG hypermutability model
 with the longer mutational motif identified by \citet{harris2015evidence}.
@@
 using a constrained quasi-Newton method (``optim(.., method='L-BFGS-B')'' in R).
-The top and bottom of the table of (2,1,0) and (2,0,1) T-mer residuals of this model were as follows:
+The top and bottom of the table of (2,1,0) and (2,0,1) T-mer residuals of this model were
+as shown in Table~\ref{tab:resid1}.
+
+\begin{table}
+  \begin{center}
+        \begin{tabular}{c@{\quad$\to$\quad}c@{\quad at rate\quad }ll}
+          \hline
+            x & y & truth & MLE \\
+          \hline
+          \nA  &   \nT           & 0.10 & 0.0996 \\
+          \nT  &   \nA           & 0.10 & 0.0982 \\
+          \nC  &   \nG           & 0.15 & 0.1505 \\
+          \nG  &   \nC           & 0.15 & 0.1497 \\
+          \nA  &   \nC           & 0.08 & 0.0784 \\
+          \nT  &   \nG           & 0.08 & 0.0792 \\
+          \nA  &   \nG           & 0.08 & 0.0798 \\
+          \nT  &   \nC           & 0.08 & 0.0795 \\
+          \nC  &   \nA           & 0.12 & 0.1202 \\
+          \nG  &   \nT           & 0.12 & 0.1196 \\
+          \nC  &   \nT           & 0.12 & 0.1198 \\
+          \nG  &   \nA           & 0.12 & 0.1193 \\
+       \nC\nG  &  \nT\nG         & 0.40 & 0.3978 \\
+       \nC\nG  &  \nC\nA         & 0.40 & 0.3978 \\
+    \nT\nC\nC  &  \nT\nT\nC      & 0.10 & 0.1033 \\
+    \nA\nG\nG  &  \nA\nA\nG      & 0.10 & 0.1033 \\
+           \hline
+        \end{tabular}
+  \end{center}
+  \caption{
+      True parameter values, and estimates obtained
+      from iterative fitting to a $10^6$ bp sequence
+    evolved to have 28.5\% sequence divergence
+    with the above mutational motifs. The transitions $CG \to TG$ and $CG \to CA$
+    were constrained during fitting to be equal (DNA strand symmetry),
+    as were the transitions $TCC \to TTC$ and $AGG \to AAG$.
+    \label{tab:cpg_results} }
+\end{table}
+
+\begin{table}
 % cpg-plus-epsilon/32140/base-model-fit-5-3-l1-resids.2.1.l0.tsv
@@
     \end{center}
+    \caption{
+        Largest residuals of shape (2,1,0) and (2,0,1) after fitting a model with only ``basic''
+        (single-nucleotide) mutations to sequence simulated under the model
+        of Table~\ref{tab:cpg_results}, that also includes CpG hypermutability
+        as well as $TCC \to TTC$ and $AGG \to AAG$ mutations.
+        Note that in the table, the underscore means ``any nucleotide''.
+        \label{tab:resid1}
+    }
+\end{table}
+
 The situation clearly calls for addition of the CpG mutation motif.
@@
 some length 2 residuals had statistically significant residuals,
-%EM In my compilation there is a float that interrupts this flow.
-but none so large as those seen in (3,1,1) T-mer residuals, shown here:
+but none so large as those seen in (3,1,1) T-mer residuals, shown in Table~\ref{tab:resid2}.
+
+\begin{table}
     \begin{center}
@@
     \end{center}
+    \caption{
+        Largest residuals of shape (3,1,1) after fitting a model with ``basic''
+        (single-nucleotide) and CpG mutations to sequence simulated under the mdoel
+        of Table~\ref{tab:cpg_results}, that also includes $TCC \to TTC$ and $AGG \to AAG$ mutations.
+        Note that in the table, the underscore means ``any nucleotide''.
+        \label{tab:resid2}
+    }
+\end{table}
+
 \ldots which again clearly call for adding the mutation tuples $TCC \to TTC$ and $AGG \to AAG$.
@@
 %     class = "data.frame")
-
-\begin{table}
-  \begin{center}
-        \begin{tabular}{c@{\quad$\to$\quad}c@{\quad at rate\quad }ll}
-          \hline
-            x & y & truth & MLE \\
-          \hline
-          \nA  &   \nT           & 0.10 & 0.0996 \\
-          \nT  &   \nA           & 0.10 & 0.0982 \\
-          \nC  &   \nG           & 0.15 & 0.1505 \\
-          \nG  &   \nC           & 0.15 & 0.1497 \\
-          \nA  &   \nC           & 0.08 & 0.0784 \\
-          \nT  &   \nG           & 0.08 & 0.0792 \\
-          \nA  &   \nG           & 0.08 & 0.0798 \\
-          \nT  &   \nC           & 0.08 & 0.0795 \\
-          \nC  &   \nA           & 0.12 & 0.1202 \\
-          \nG  &   \nT           & 0.12 & 0.1196 \\
-          \nC  &   \nT           & 0.12 & 0.1198 \\
-          \nG  &   \nA           & 0.12 & 0.1193 \\
-       \nC\nG  &  \nT\nG         & 0.40 & 0.3978 \\
-       \nC\nG  &  \nC\nA         & 0.40 & 0.3978 \\
-    \nT\nC\nC  &  \nT\nT\nC      & 0.10 & 0.1033 \\
-    \nA\nG\nG  &  \nA\nA\nG      & 0.10 & 0.1033 \\
-           \hline
-        \end{tabular}
-  \end{center}
-  \caption{ Results from iterative fitting to a $10^6$ bp sequence
-    evolved to have 28.5\% sequence divergence
-    with the above mutational motifs. The transitions $CG \to TG$ and $CG \to CA$
-    were constrained to be equal (DNA strand symmetry),
-    as were the transitions $TCC \to TTC$ and $AGG \to AAG$.
-    \label{tab:cpg_results} }
-\end{table}
 
@@
 affect the mutation rate of the central site has $w=1$.)
-%EM I suggest that we could re-introduce the mu^* notation again and remind people that these potential changes are uniform.
-
-As in the ``Simulation'' section, we start by considering a broader collection
-of ``possible changes'' sampled from a homogeneous
-Poisson process at a rate $\mu^*$ per site.
+
+Recall that, as described in the ``Simulation'' section,
+we may represent the process by first laying down a homogeneous Poisson process
+with rate $\mu^*$ per site of ``potential mutations'',
+where $\mu^*$ is the maximum possible mutation rate,
+and then resolving these appropriately.
+(In particular, the times and locations of these potential mutations are chosen uniformly.)
 Suppose a possible change occurs at site $i-w$, at time $t$,
-%EM should this be "change in rate"?
-which therefore might result in a change in state at site $i$.
+which therefore might result in a change in mutation rate at site $i$.
 To resolve this, we need to know the outcome of any other potential changes
@@
 thus extending the sequence on which the change depends to the right.
-The number of such potential changes have a uniform location
-and their count is Poisson with mean $\mu^* w t$;
-these extend the amount of sequence on which the change at site $i$ depends
-%EM my understanding is that it's just uniform for a single mutation, but that's not obvious from the description.
-by a uniform random number of sites between 1 and $w$.
+The number of such potential changes is Poisson with mean $\mu^* w t$;
+if this number is nonzero, then
+this extends the amount of sequence on which the change at site $i$ depends
+by a uniform number between $1$ and $w$.
 If one occurs, say at time $0 < s < t$,
@@
 \begin{lemma}
-%EM Let's rename this X to ... V? We already have an X.
-    Let $X = \sum_{j=1}^N W_j$,
+    Let $V = \sum_{j=1}^N W_j$,
     where $N$ is Poisson with mean $\lambda$
@@
     \begin{align}
-        \P\{ X > k w \}
+        \P\{ V > k w \}
         \le
@@
 \begin{proof}
-    First note that $\P\{X > kw\} \le \P\{N > k\}$.
+    First note that $\P\{V > kw\} \le \P\{N > k\}$.
     Now, use the fact that if $f(n) = n (n-1) \cdots (n-k) / (k+1)!$

### 2021-12-10 writeups/context-dependence.tex
Finishing my pass through main text

@@
 
-%EM It seems reasonable that we should be able to get away using a smaller T-mer.
-%EM Question: how much context should we include; argument in appendix.
-%EM Would be nice to show convergence of full likelihood eventually.
-
 See the next section for more discussion of the approximations.
@@
 to check for convergence.
-
-%%%%%%%%
-\subsection{Model fit and model selection}
-
-Fitting a model to nucleotide data would likely require iterative modification of the model.
-The difference between observed and expected T-mer counts provide a good assessment of how and whether a model can be improved.
-These \emph{residual} counts are usually most usefully computed for T-mers shorter than those used to fit the model.
-For each T-mer $(a,b)$, write $O_{a,b}$ for the number of this T-mer observed in the data,
-and $E_{a,b}$ for the number expected under the model.
-(The expected number is the number of occurrences of $a$ in the initial sequence
-multiplied by the probability $p_{n,\ell,r}(t;a,b)$, where $n$ is the length of $a$,
-and $\ell$, $r$ are the appropriate overhangs.)
-If the T-mers were nonoverlapping, this would form a standard contingency table,
-and standard practice would be to divide the residuals by the square root of the expected counts
-and compare these to a standard Normal distribution.
-However, there are are $n$ possible such tables --
-so, we compute normalized residuals as
-\begin{align}
-    Z_{a,b} = \frac{ O_{a,b} - E_{a,b} }{ \sqrt{n E_{a,b}} },
-\end{align}
-which is equivalent to averaging the $z$-scores of the $n$ possible tables of nonoverlapping counts.
-These can then be compared to a standard Normal distribution.
 
@@
 
-The general case is a modification of Felsenstein's ``pruning'' algorithm.
+The general case is a modification of Felsenstein's ``pruning'' algorithm\citep{felsenstein1981evolutionary}.
 
@@
 for certain (more common) patterns: for instance,
-we'd compute the likelihood for situations where $y$ and $z$ agree (but may differ from $x$
+one could compute the likelihood for situations where $y$ and $z$ agree (but may differ from $x$
 by replacing step (4) with:
@@
 
-  \item[4'.] Compute $M_4(x,w) = \sum_{v \in \calS^{n}} p_{n,0,0}(t_{A},v,x) M_3(v,w) M_4(v,w)$ (stored as a $|\calS|^{n} \times |\calS|^{m}$ matrix).
+  \item[$4'$.] Compute $M_4(x,w) = \sum_{v \in \calS^{n}} p_{n,0,0}(t_{A},v,x) M_3(v,w) M_4(v,w)$ (stored as a $|\calS|^{n} \times |\calS|^{m}$ matrix).
 
@@
 
-%%%%%%%%%%%
-\subsection{Hominid data}
-\label{hominid_section}
-
-We also applied the model to human--chimpanzee divergence data,
-restricting the comparison to putative regulatory sequences,
-avoiding the additional constraints of coding sequences.
-To do this, we extracted regions in the human (hg38)--chimpanzee (panTro4) alignment % \citep[from the UCSC Genome Browser][]{pantroalign}
-according to four types of regulatory feature in the Ensembl Regulatory Build \citep[release 81][]{zerbino2015ensembl}:
-``CTCF binding site'', ``enhancer'', ``open chromatin region'', and ``promoter flanking region''.
-These were furthermore divided into regions that either did or did not overlap a transcription start or end site of a known gene
-from the hg38 Annotations Database \citep{kent2002human}
-and filtered for length:
-aligned regions below either 3000bp (promoter flanking regions) or 1000bp (all other types).
-We then counted T-mers of appropriate length in each of the eight resulting sets of aligned regions, omitting any T-mers that included gaps or repeat-masked positions,
-and fit models separately to each of the eight data sets.
-
-We constructed models from three, nested, sets of mutation motifs.
-All models are strand-symmetric.
-\begin{itemize}
-    \item[(basic)]
-        Single-base changes, with separate rates for transitions ($\mu_v$),
-        \nA$\leftrightarrow$\nT{} ($\mu_{AT}$), \nC$\leftrightarrow$\nG{} ($\mu_{CG}$),
-        and other transversions ($\mu_{r}$):
-          \begin{center}
-            \begin{tabular}{c@{\quad$\to$\quad}c@{\quad at rate\quad }c|c@{\quad$\to$\quad}c@{\quad at rate\quad }c|c@{\quad$\to$\quad}c@{\quad at rate\quad }c}
-                \nT  &   \nA   &  $\mu_{AT}$ & \nA  &   \nC   &  $\mu_{r}$ & \nA  &   \nG   &  $\mu_{v}$ \\
-                \nA  &   \nT   &  $\mu_{AT}$ & \nC  &   \nA   &  $\mu_{r}$ & \nG  &   \nA   &  $\mu_{v}$ \\
-                \nC  &   \nG   &  $\mu_{CG}$ & \nG  &   \nT   &  $\mu_{r}$ & \nT  &   \nC   &  $\mu_{v}$ \\
-                \nG  &   \nC   &  $\mu_{CG}$ & \nT  &   \nG   &  $\mu_{r}$ & \nC  &   \nT   &  $\mu_{v}$ \\
-            \end{tabular}
-          \end{center}
-
-      \item[(CpG)] In addition to the above, an additional CpG rate:
-          \begin{center}
-            \begin{tabular}{c@{\quad$\to$\quad}c@{\quad at rate\quad }c}
-                CG  &   TG   &  $\nu_{CpG}$ \\
-                CG  &   CA   &  $\nu_{CpG}$ \\
-            \end{tabular}
-          \end{center}
-
-      \item[(repair)] In addition to the above, we included a collection of mutation motifs associated with known DNA lesion/repair pathways
-          curated from the literature \citep[reviewed in][]{sale2012yfamily,goodman2013translesion,roberts2014hypermutation,cobey2015evolution,goodman2002errorprone}:
-          deamination due to repair of UV-induced pyramidine dimers \citep[$\mu_{UV}$,][]{sale2012yfamily,sinha2002uvinduced};
-          repair of guanine adducts at CpG sites \citet[$\mu_g$,][]{pfeifer2006mutagenesis};
-          deamination by AID \citep[$\mu_{AID}$,][]{teng2007immunoglobulin,kasar2015wholegenome};
-          deamination by members of the APOBEC family \citep[$\mu_\APOBEC$,][]{alexandrov2013deciphering};
-          and repair by error-prone polymerases pol $\eta$ \citep[$\mu_\eta$,][]{alexandrov2013deciphering};
-          and pol $\iota$ \citep[$\mu_{\iota}$,][]{roberts2014hypermutation,maul2016polymerase}.
-          These are denoted here using ambiguity codes
-          \nR{}: \nA{} or \nG{},
-          \nY{}: \nC{} or \nT{},
-          \nW{}: \nA{} or \nT{}, and
-          \nS{}: \nC{} or \nG{};
-          so for instance
-          ``\nT\nC\nW$\to$\nT\nT\nW''
-          means that both
-          \nT\nC\nA$\to$\nT\nT\nA{}
-          and
-          \nT\nC\nT$\to$\nT\nT\nT{}
-          share the same mutation rate.
-          \noindent
-      \begin{center}
-          \begin{tabular}{c@{\;\;$\to$\;\;}c@{\;\; {\small at rate}\;\; }c|c@{\;\;$\to$\;\;}c@{\;\; {\small at rate}\;\; }c}
-            \nC\nG     &   \nC\nT      &  $\mu_{g}$      & \nA\nA       &   \nA\nG       &  $\mu_{\iota1}$   \\
-            \nC\nG     &   \nA\nG      &  $\mu_{g}$      & \nT\nT       &   \nC\nT       &  $\mu_{\iota1}$   \\
-            \nT\nC     &   \nT\nT      &  $\mu_{UV}$     & \nA\nA       &   \nG\nT       &  $\mu_{\iota2}$   \\
-            \nG\nA     &   \nA\nA      &  $\mu_{UV}$     & \nT\nT       &   \nA\nC       &  $\mu_{\iota2}$   \\
-            \nC\nC     &   \nC\nT      &  $\mu_{UV}$     & \nT\nC\nW    &   \nT\nT\nW    &  $\mu_{\APOBEC}$  \\
-            \nG\nG     &   \nA\nG      &  $\mu_{UV}$     & \nT\nC\nW    &   \nT\nG\nW    &  $\mu_{\APOBEC}$  \\
-            \nC\nC     &   \nT\nT      &  $\mu_{UVd}$    & \nS\nG\nA    &   \nS\nA\nA    &  $\mu_{\APOBEC}$  \\
-            \nG\nG     &   \nA\nA      &  $\mu_{UVd}$    & \nS\nG\nA    &   \nS\nC\nA    &  $\mu_{\APOBEC}$  \\
-            \nT\nA     &   \nT\nG      &  $\mu_{\eta}$   & \nW\nR\nC\nY &  \nW\nR\nT\nY  &  $\mu_{\AID}$     \\
-            \nT\nA     &   \nC\nA      &  $\mu_{\eta}$   & \nR\nG\nY\nS &  \nR\nA\nY\nS  &  $\mu_{\AID}$
-        \end{tabular}
-      \end{center}
-
-\end{itemize}
-
-The mutational motifs included above are far from complete
--- context-specific affinities and error rates \textit{in vivo}
-are not known for most DNA repair pathways --
-but neither do they exhaust the hypotheses available in the literature.
-For instance, many enzymes act more than once locally when they act,
-such as AID \citep{senavirathne2015activationinduced} or error-prone polymerases involved in repair \citep{maul2016polymerase}.
+%%%%%%%%
+\subsection{Model fit and model selection}
+%EM I promoted this section here because we use residuals a lot in the results.
+Fitting a model to nucleotide data will likely require iterative modification of the model.
+The difference between observed and expected T-mer counts provide a good assessment of how and whether a model can be improved.
+%EM Why shorter? Because there are fewer of them?
+These \emph{residual} counts are usually most usefully computed for T-mers shorter than those used to fit the model.
+%EM Should we replace a and b with x and y here to agree with the rest of the doc?
+For each T-mer $(a,b)$, write $O_{a,b}$ for the number of this T-mer observed in the data,
+and $E_{a,b}$ for the number expected under the model.
+(The expected number is the number of occurrences of $a$ in the initial sequence
+multiplied by the probability $p_{n,\ell,r}(t;a,b)$, where $n$ is the length of $a$,
+and $\ell$, $r$ are the appropriate overhangs.)
+If the T-mers were nonoverlapping, this would form a standard contingency table,
+and standard practice would be to divide the residuals by the square root of the expected counts
+and compare these to a standard Normal distribution.
+However, there are are $n$ possible such tables --
+so, we compute normalized residuals as
+\begin{align}
+    Z_{a,b} = \frac{ O_{a,b} - E_{a,b} }{ \sqrt{n E_{a,b}} },
+\end{align}
+which is equivalent to averaging the $z$-scores of the $n$ possible tables of nonoverlapping counts.
+These can then be compared to a standard Normal distribution.
 
@@
 
-In exploratory application to real data, how does one decide when to stop adding mutational motifs?
-Perhaps the simplest method conceptually is to
+We benchmarked the performance of stepwise model selection:
 (a) fit a simple model,
@@
 The resulting sequences differed at 28.5\% of the sites.
-We then fit an unconstrained model of single-nucleotide substitution using (5,1,1) T-mers
+
+To begin, we fit an unconstrained model of single-nucleotide substitution using (5,1,1) T-mers
 (three sites at the base with an overhang of one on each side)
-using a constrained quasi-Newton method (``optim(.., method='L-BFGS-B')'' in R).
-The top and bottom of the table of (2,1,0) and (2,0,1) T-mer residuals of this model were
-as shown in Table~\ref{tab:resid1}.
-
-\begin{table}
+using a constrained quasi-Newton method (\verb|optim(.., method='L-BFGS-B')| in R).
+
+\begin{table}[ht]
   \begin{center}
@@
           \hline
+          %EM I'm confused why we have MLE parameters for these single NT terms because I don't see those being used in the iterative model fitting. I thought we started with a (5,1,1) model.
           \nA  &   \nT           & 0.10 & 0.0996 \\
@@
   \caption{
-      True parameter values, and estimates obtained
-      from iterative fitting to a $10^6$ bp sequence
+      True parameter values, and MLE estimates obtained
+      from iterative fitting as described in the text to a $10^6$ bp sequence
     evolved to have 28.5\% sequence divergence
-    with the above mutational motifs. The transitions $CG \to TG$ and $CG \to CA$
+    with the above mutational motifs. The transitions $\nC\nG \to \nT\nG$ and $\nC\nG \to \nC\nA$
     were constrained during fitting to be equal (DNA strand symmetry),
-    as were the transitions $TCC \to TTC$ and $AGG \to AAG$.
+    as were the transitions $\nT\nC\nC \to \nT\nT\nC$ and $\nA\nG\nG \to \nA\nA\nG$.
     \label{tab:cpg_results} }
@@
 
+We find that the (2,1,0) and (2,0,1) T-mer residuals of this model are dominated by
+terms involving CG (Table~\ref{tab:resid1}).
+%EM Stupid question: why weren't these picked up by the 5,1,1 model?
+%EM Why isn't the starting point a (1,0,0) model?
+These results clearly call for addition of the CpG mutation motif.
+
 \begin{table}
+%EM This table is BIG. And it seems like our conclusions only stem from the very biggest residuals.
+%EM What if we cut of anything that doesn't have an absolute Z-score bigger than 10?
 % cpg-plus-epsilon/32140/base-model-fit-5-3-l1-resids.2.1.l0.tsv
@@
                 \nA\nG  & \nC\_   &  10968   &   11760    &   -791.73   &	 -5.16  \\
+                %EM It looks like this is missing an entry?
                 \nG\nG  & \nG\_   & 118805   &  120975    &  -2170.25   &    \\
@@
         of Table~\ref{tab:cpg_results}, that also includes CpG hypermutability
-        as well as $TCC \to TTC$ and $AGG \to AAG$ mutations.
+        as well as $\nT\nC\nC \to \nT\nT\nC$ and $\nA\nG\nG \to \nA\nA\nG$ mutations.
         Note that in the table, the underscore means ``any nucleotide''.
@@
 
-The situation clearly calls for addition of the CpG mutation motif.
-After adding these patterns into the model and re-fitting,
+%EM "These patterns": Is it exactly the patterns in the true model?
+%EM Would we need to have prior information about CpGs to pick those additional terms?
+After adding these patterns into the model and re-fitting, we searched for evidence of additional processes happening in the data.
 % starting at the parameters inferred from the first run and a CpG rate of 0.0,
-some length 2 residuals had statistically significant residuals,
-but none so large as those seen in (3,1,1) T-mer residuals, shown in Table~\ref{tab:resid2}.
+%EM what is "statistically significant" here? I searched around and found 1.96 corresponding to 0.05.
+To do so, we calculated length 2 residuals and (3,1,1) T-mer residuals.
+Although some of the length 2 residuals were statistically significant,
+none were as large as those seen in (3,1,1) T-mer residuals, which are shown in Table~\ref{tab:resid2}.
+These residuals clearly call for adding the mutation tuples $\nT\nC\nC \to \nT\nT\nC$ and $\nA\nG\nG \to \nA\nA\nG$.
+These were indeed the motifs used when simulating the data.
+After adding these in,
+% and initializing all parameters to 0.1
+no further (3,1,1) T-mer residuals achieved
+statistical significance after a Bonferroni correction;
+the final parameter values are shown in Table \ref{tab:cpg_results};
+parameter estimates differ from the truth by less than 3\% (mostly within 1\%).
 
@@
         Largest residuals of shape (3,1,1) after fitting a model with ``basic''
-        (single-nucleotide) and CpG mutations to sequence simulated under the mdoel
-        of Table~\ref{tab:cpg_results}, that also includes $TCC \to TTC$ and $AGG \to AAG$ mutations.
+        (single-nucleotide) and CpG mutations to sequence simulated under the model
+        of Table~\ref{tab:cpg_results}, that also includes $\nT\nC\nC \to \nT\nT\nC$ and $\nA\nG\nG \to \nA\nA\nG$ mutations.
         Note that in the table, the underscore means ``any nucleotide''.
@@
 \end{table}
-
-\ldots which again clearly call for adding the mutation tuples $TCC \to TTC$ and $AGG \to AAG$.
-These were indeed the motifs used when simulating the data.
-After adding these in,
-% and initializing all parameters to 0.1
-no further (3,1,1) T-mer residuals achieved
-statistical significance after a Bonferroni correction;
-the final parameter values are shown in Table \ref{tab:cpg_results};
-parameter estimates differ from the truth by less than 3\% (mostly within 1\%).
 
@@
 The usage of fast sparse matrix methods described above make likelihood computation fast enough
-to use a Metropolis-Hasting-based Markov chain Monte Carlo method
+to use a Metropolis-Hastings-based Markov chain Monte Carlo method
 to sample from the posterior distribution on parameters.
@@
 counted $(5,3,1)$ T-mers,
-and used the mcmc package \citep{geyer2017mcmc} to run a random walk sampler for 10,000 steps
+and used the \texttt{mcmc} package \citep{geyer2017mcmc} to run a random walk sampler for 10,000 steps
 (judged sufficient by examination of diagnostic plots).
@@
     \begin{center}
+        %EM Should I be looking at the gray bars or black bars? What are they? I'm guessing 95% is gray, because that corresponds to good coverage.
         \includegraphics{writeup-plots/coverage_results}
@@
 
-We also applied the method to changes between the human and chimpanzee reference sequences,
-using the model described above.
+We also applied the method to human--chimpanzee divergence data,
+restricting the comparison to putative regulatory sequences,
+avoiding the additional constraints of coding sequences.
+To do this, we extracted regions in the human (hg38)--chimpanzee (panTro4) alignment % \citep[from the UCSC Genome Browser][]{pantroalign}
+according to four types of regulatory feature in the Ensembl Regulatory Build \citep[release 81][]{zerbino2015ensembl}:
+``CTCF binding site'', ``enhancer'', ``open chromatin region'', and ``promoter flanking region''.
+These were furthermore divided into regions that either did or did not overlap a transcription start or end site of a known gene
+from the hg38 Annotations Database \citep{kent2002human}
+and filtered for length:
+aligned regions below either 3000bp (promoter flanking regions) or 1000bp (all other types).
+We then counted T-mers of appropriate length in each of the eight resulting sets of aligned regions, omitting any T-mers that included gaps or repeat-masked positions,
+and fit models separately to each of the eight data sets.
+
+We constructed models from three, nested, sets of mutation motifs.
+All models are strand-symmetric.
+\begin{itemize}
+    \item[(basic)]
+        Single-base changes, with separate rates for transitions ($\mu_v$),
+        \nA$\leftrightarrow$\nT{} ($\mu_{AT}$), \nC$\leftrightarrow$\nG{} ($\mu_{CG}$),
+        and other transversions ($\mu_{r}$):
+          \begin{center}
+            \begin{tabular}{c@{\quad$\to$\quad}c@{\quad at rate\quad }c|c@{\quad$\to$\quad}c@{\quad at rate\quad }c|c@{\quad$\to$\quad}c@{\quad at rate\quad }c}
+                \nT  &   \nA   &  $\mu_{AT}$ & \nA  &   \nC   &  $\mu_{r}$ & \nA  &   \nG   &  $\mu_{v}$ \\
+                \nA  &   \nT   &  $\mu_{AT}$ & \nC  &   \nA   &  $\mu_{r}$ & \nG  &   \nA   &  $\mu_{v}$ \\
+                \nC  &   \nG   &  $\mu_{CG}$ & \nG  &   \nT   &  $\mu_{r}$ & \nT  &   \nC   &  $\mu_{v}$ \\
+                \nG  &   \nC   &  $\mu_{CG}$ & \nT  &   \nG   &  $\mu_{r}$ & \nC  &   \nT   &  $\mu_{v}$ \\
+            \end{tabular}
+          \end{center}
+
+      \item[(CpG)] In addition to the above, an additional CpG rate:
+          \begin{center}
+            \begin{tabular}{c@{\quad$\to$\quad}c@{\quad at rate\quad }c}
+                CG  &   TG   &  $\nu_{CpG}$ \\
+                CG  &   CA   &  $\nu_{CpG}$ \\
+            \end{tabular}
+          \end{center}
+
+      \item[(repair)] In addition to the above, we included a collection of mutation motifs associated with known DNA lesion/repair pathways
+          curated from the literature \citep[reviewed in][]{sale2012yfamily,goodman2013translesion,roberts2014hypermutation,cobey2015evolution,goodman2002errorprone}:
+          deamination due to repair of UV-induced pyramidine dimers \citep[$\mu_{UV}$,][]{sale2012yfamily,sinha2002uvinduced};
+          repair of guanine adducts at CpG sites \citet[$\mu_g$,][]{pfeifer2006mutagenesis};
+          deamination by AID \citep[$\mu_\AID$,][]{teng2007immunoglobulin,kasar2015wholegenome};
+          deamination by members of the APOBEC family \citep[$\mu_\APOBEC$,][]{alexandrov2013deciphering};
+          and repair by error-prone polymerases pol $\eta$ \citep[$\mu_\eta$,][]{alexandrov2013deciphering};
+          and pol $\iota$ \citep[$\mu_{\iota}$,][]{roberts2014hypermutation,maul2016polymerase}.
+          These are denoted here using ambiguity codes
+          \nR{}: \nA{} or \nG{},
+          \nY{}: \nC{} or \nT{},
+          \nW{}: \nA{} or \nT{}, and
+          \nS{}: \nC{} or \nG{};
+          so for instance
+          ``\nT\nC\nW$\to$\nT\nT\nW''
+          means that both
+          \nT\nC\nA$\to$\nT\nT\nA{}
+          and
+          \nT\nC\nT$\to$\nT\nT\nT{}
+          share the same mutation rate.
+          \noindent
+      \begin{center}
+          \begin{tabular}{c@{\;\;$\to$\;\;}c@{\;\; {\small at rate}\;\; }c|c@{\;\;$\to$\;\;}c@{\;\; {\small at rate}\;\; }c}
+            \nC\nG     &   \nC\nT      &  $\mu_{g}$      & \nA\nA       &   \nA\nG       &  $\mu_{\iota1}$   \\
+            \nC\nG     &   \nA\nG      &  $\mu_{g}$      & \nT\nT       &   \nC\nT       &  $\mu_{\iota1}$   \\
+            \nT\nC     &   \nT\nT      &  $\mu_{UV}$     & \nA\nA       &   \nG\nT       &  $\mu_{\iota2}$   \\
+            \nG\nA     &   \nA\nA      &  $\mu_{UV}$     & \nT\nT       &   \nA\nC       &  $\mu_{\iota2}$   \\
+            \nC\nC     &   \nC\nT      &  $\mu_{UV}$     & \nT\nC\nW    &   \nT\nT\nW    &  $\mu_{\APOBEC}$  \\
+            \nG\nG     &   \nA\nG      &  $\mu_{UV}$     & \nT\nC\nW    &   \nT\nG\nW    &  $\mu_{\APOBEC}$  \\
+            \nC\nC     &   \nT\nT      &  $\mu_{UVd}$    & \nS\nG\nA    &   \nS\nA\nA    &  $\mu_{\APOBEC}$  \\
+            \nG\nG     &   \nA\nA      &  $\mu_{UVd}$    & \nS\nG\nA    &   \nS\nC\nA    &  $\mu_{\APOBEC}$  \\
+            \nT\nA     &   \nT\nG      &  $\mu_{\eta}$   & \nW\nR\nC\nY &  \nW\nR\nT\nY  &  $\mu_{\AID}$     \\
+            \nT\nA     &   \nC\nA      &  $\mu_{\eta}$   & \nR\nG\nY\nS &  \nR\nA\nY\nS  &  $\mu_{\AID}$
+        \end{tabular}
+      \end{center}
+
+\end{itemize}
+
+The mutational motifs included above are far from complete
+-- context-specific affinities and error rates \textit{in vivo}
+are not known for most DNA repair pathways --
+but neither do they exhaust the hypotheses available in the literature.
+For instance, many enzymes act more than once locally when they act,
+such as AID \citep{senavirathne2015activationinduced} or error-prone polymerases involved in repair \citep{maul2016polymerase}.
+
+Nevertheless, we find significant effects for a number of biochemically-informed mutation features (Figure~\ref{fig:biochem_results}).
+Specifically, we find strong CpG and UV damage effects.
+%EM Kelley saw the same thing https://paperpile.com/shared/aAYSgW ...
+It may seem tenuous to look for mutational signatures associated with cancer in the germline,
+or absurd that UV damage could play a significant role.
+This is in part due to the ``streetlight effect'' --
+we can't look for signatures that are not characterized --
+but it is also reasonable that shared repair machinery might induce similar context-dependent errors.
+More likely sources of error in the germline include the effects of recombination
+\citep{arbeithuber2015crossovers,myers2010drive}.
 
@@
 
-It may seem tenuous to look for mutational signatures associated with cancer in the germline,
-or absurd that UV damage could play a significant role.
-This is in part due to the ``streetlight effect'' --
-we can't look for signatures that are not characterized --
-but it is also reasonable that shared repair machinery might induce similar context-dependent errors.
-More likely sources of error in the germline include the effects of recombination
-\citep{arbeithuber2015crossovers,myers2010drive}.
-
 %%%%%%%%%%%%%%%%%%%%
@@
 
+%EM given the full likelihood arguments, can we really say provably correct?
 This paper presents a tractable, provably correct method
@@
 Such models appear in many fields, for instance,
+%EM are semicolons really legal here?
 as the dynamic counterparts to Markov random fields;
@@
 
+%EM approximate full likelihoods?
 The methods described here provide a way to efficiently calculate full likelihoods,
@@
 The model presented assumes that the same process occurs at every location along the sequence,
-and that the length of the sequence does not change (no indels).
+and that the length of the sequence does not change (no insertions or deletions).
 Neither of these assumptions are likely to be true for any data set of DNA sequence;
@@
 we can easily allow heterogeneity between sites by scaling rates by an overall gamma-distributed random factor \citep{yang1994maximum}.
+%EM Does a gap character really make sense? It seems like this changes the interpretation of the tmer length... I'd skip.
 Short deletions might be accommodated by introducing a ``gap'' letter to the alphabet,
 but insertions are problematic.
+
+%EM Full likelihood conjecture?
 

### 2021-12-10 writeups/context-dependence.tex
A few things for appearance

@@
 \bibliographystyle{plainnat}
+
+% http://bytesizebio.net/2013/03/11/adding-supplementary-tables-and-figures-in-latex/
+\newcommand{\beginsupplement}{%
+        \setcounter{table}{0}
+        \renewcommand{\thetable}{S\arabic{table}}%
+        \setcounter{figure}{0}
+        \renewcommand{\thefigure}{S\arabic{figure}}%
+     }
 
@@
 
+\clearpage
+
+\beginsupplement
+
 \appendix
@@
 
-\begin{figure}
+\begin{figure}[h!]
     \begin{center}
@@
     \caption{
-        As in \ref{fig:ising_coverage}:
+        As in Figure~\ref{fig:ising_coverage}:
         Credible intervals for the four parameters of the dynamic Ising model,
@@
 \end{figure}
+
+\clearpage
 

### 2021-12-11 writeups/context-dependence.tex
edits for EMs

@@
 \subsection{Model fit and model selection}
-%EM I promoted this section here because we use residuals a lot in the results.
 Fitting a model to nucleotide data will likely require iterative modification of the model.
 The difference between observed and expected T-mer counts provide a good assessment of how and whether a model can be improved.
-%EM Why shorter? Because there are fewer of them?
-These \emph{residual} counts are usually most usefully computed for T-mers shorter than those used to fit the model.
-%EM Should we replace a and b with x and y here to agree with the rest of the doc?
-For each T-mer $(a,b)$, write $O_{a,b}$ for the number of this T-mer observed in the data,
-and $E_{a,b}$ for the number expected under the model.
-(The expected number is the number of occurrences of $a$ in the initial sequence
-multiplied by the probability $p_{n,\ell,r}(t;a,b)$, where $n$ is the length of $a$,
+These \emph{residual} counts are usually most usefully computed for T-mers shorter than those used to fit the model,
+because the smaller total number of possible T-mers makes it easier to identify the shared patterns.
+For each T-mer $(x,y)$, write $O_{x,y}$ for the number of this T-mer observed in the data,
+and $E_{x,y}$ for the number expected under the model.
+(The expected number is the number of occurrences of $x$ in the initial sequence
+multiplied by the probability $p_{n,\ell,r}(t;x,y)$, where $n$ is the length of $x$,
 and $\ell$, $r$ are the appropriate overhangs.)
@@
 \begin{align}
-    Z_{a,b} = \frac{ O_{a,b} - E_{a,b} }{ \sqrt{n E_{a,b}} },
+    Z_{x,y} = \frac{ O_{x,y} - E_{x,y} }{ \sqrt{n E_{x,y}} },
 \end{align}
@@
 
-To begin, we fit an unconstrained model of single-nucleotide substitution using (5,1,1) T-mers
-(three sites at the base with an overhang of one on each side)
+To begin, we fit an unconstrained model of single-nucleotide substitution
 using a constrained quasi-Newton method (\verb|optim(.., method='L-BFGS-B')| in R).
+Recall that for inference, we next need to pick a T-mer shape
+with which to compute likelihoods,
+and the shape of this T-mer is only determined by the shape of dependencies in the generative model
+in that it determines the minimum base and overhang lengths.
+We use (5,1,1) T-mers: three sites at the base with an overhang of one on each side.
+Since this initial model has no context-dependence,
+we could do the first step of model fitting using (1,0,0) T-mers (i.e., single positions),
+but for simplicity we use the same T-mer shape for all steps.
+
+After fitting the model, we can look at residuals using any T-mer shape we like.
+At this first step, we computed the (2,1,0) and (2,0,1) T-mer residuals,
+shown in Table~\ref{tab:resid1}.
+The largest residuals are an excess of $\nC \to \nT$ mutations in $\nC\nG$ dinucleotides,
+and the reverse-complement pattern ($z$ scores of 88).
+Complementing this are deficits of $\nC \to \nT$ mutations in other contexts,
+since the $\nC \to \nT$ mutation rate has been fit to be an average across contexts.
+This clearly calls for addition of the $\nC\nG \to \nT\nG$ mutation motif (and reverse complement),
+which was used in simulating the data.
 
@@
           \hline
-          %EM I'm confused why we have MLE parameters for these single NT terms because I don't see those being used in the iterative model fitting. I thought we started with a (5,1,1) model.
           \nA  &   \nT           & 0.10 & 0.0996 \\
@@
 
-We find that the (2,1,0) and (2,0,1) T-mer residuals of this model are dominated by
-terms involving CG (Table~\ref{tab:resid1}).
-%EM Stupid question: why weren't these picked up by the 5,1,1 model?
-%EM Why isn't the starting point a (1,0,0) model?
-These results clearly call for addition of the CpG mutation motif.
-
 \begin{table}
@@
 %EM What if we cut of anything that doesn't have an absolute Z-score bigger than 10?
+%PR True, but I'd like to show that the $z$-scores do down to something smaller?
+%PR I've cut out four lines...
 % cpg-plus-epsilon/32140/base-model-fit-5-3-l1-resids.2.1.l0.tsv
@@
                 \nA\nG  & \nC\_   &  10968   &   11760    &   -791.73   &	 -5.16  \\
-                %EM It looks like this is missing an entry?
-                \nG\nG  & \nG\_   & 118805   &  120975    &  -2170.25   &    \\
                 \hline
-                \nA\nG  & \nT\_   &  17280   &   16515    &    765.20   &  4.21   \\
                 \nG\nT  & \nC\_   &  20490   &   19507    &    983.44   &  4.97   \\
@@
                 \nC\nA  &  \_\nG  &   10839  &   11882  &   -1043.39  &   -6.76 \\
-                \nC\nG  &  \_\nC  &   18354  &   19348  &    -994.19  &   -5.05 \\
                 \hline
-                \nC\nT  &  \_\nA  &   17007  &   16078  &     929.33  &    5.18 \\
                 \nT\nC  &  \_\nG  &   20691  &   19509  &    1182.40  &    5.98 \\
@@
 %EM Would we need to have prior information about CpGs to pick those additional terms?
+%PR No - I've added explanation - see above.
 After adding these patterns into the model and re-fitting, we searched for evidence of additional processes happening in the data.
-% starting at the parameters inferred from the first run and a CpG rate of 0.0,
-%EM what is "statistically significant" here? I searched around and found 1.96 corresponding to 0.05.
 To do so, we calculated length 2 residuals and (3,1,1) T-mer residuals.
-Although some of the length 2 residuals were statistically significant,
-none were as large as those seen in (3,1,1) T-mer residuals, which are shown in Table~\ref{tab:resid2}.
-These residuals clearly call for adding the mutation tuples $\nT\nC\nC \to \nT\nT\nC$ and $\nA\nG\nG \to \nA\nA\nG$.
-These were indeed the motifs used when simulating the data.
+Although some of the length 2 residuals had large $z$-scores,
+none were as large as those seen in (3,1,1) T-mer residuals,
+which are shown in Table~\ref{tab:resid2}.
+These show an excess of $\nT\nC\nC \to \nT\nT\nC$ and $\nA\nG\nG \to \nA\nA\nG$ changes,
+which were indeed the motifs used when simulating the data.
 After adding these in,
@@
 no further (3,1,1) T-mer residuals achieved
-statistical significance after a Bonferroni correction;
-the final parameter values are shown in Table \ref{tab:cpg_results};
+statistical significance after computing $p$-values with the Gaussian CDF
+and applying a Bonferroni correction.
+The final parameter values are shown in Table \ref{tab:cpg_results};
 parameter estimates differ from the truth by less than 3\% (mostly within 1\%).
@@
     \begin{center}
-        %EM Should I be looking at the gray bars or black bars? What are they? I'm guessing 95% is gray, because that corresponds to good coverage.
         \includegraphics{writeup-plots/coverage_results}
@@
         differing at around 18.6\% of the sites.
-        Shown are relative intervals, obtained by dividing by the true value.
+        Dark bars show the 50\% credible intervals, and lighter grey bars the 95\% credible intervals.
+        Values shown are relative parameters, obtained by dividing by the true value.
         \label{fig:ising_coverage}}
@@
 Nevertheless, we find significant effects for a number of biochemically-informed mutation features (Figure~\ref{fig:biochem_results}).
-Specifically, we find strong CpG and UV damage effects.
-%EM Kelley saw the same thing https://paperpile.com/shared/aAYSgW ...
+Specifically, we find strong CpG and UV damage effects \citep[][also seen by]{harris2015evidence}.
 It may seem tenuous to look for mutational signatures associated with cancer in the germline,
-or absurd that UV damage could play a significant role.
-This is in part due to the ``streetlight effect'' --
-we can't look for signatures that are not characterized --
-but it is also reasonable that shared repair machinery might induce similar context-dependent errors.
-More likely sources of error in the germline include the effects of recombination
+or surprising that UV damage could play a significant role,
+but it is reasonable that shared repair machinery might induce similar context-dependent errors.
+Other likely sources of error in the germline include the effects of recombination
 \citep{arbeithuber2015crossovers,myers2010drive}.
@@
 
-%EM given the full likelihood arguments, can we really say provably correct?
-This paper presents a tractable, provably correct method
+This paper presents a tractable method
 for doing statistical inference using observations of a particle model
@@
 Such models appear in many fields, for instance,
-%EM are semicolons really legal here?
-as the dynamic counterparts to Markov random fields;
-many spin glass models;
+as the dynamic counterparts to Markov random fields or spin glass models,
 and models of DNA sequence evolution.
 
-%EM approximate full likelihoods?
-The methods described here provide a way to efficiently calculate full likelihoods,
+The methods described here provide a way to efficiently approximate full likelihoods,
 which in many circumstances provide the optimal basis for inference \citep{neyman1933problem},
@@
 we can easily allow heterogeneity between sites by scaling rates by an overall gamma-distributed random factor \citep{yang1994maximum}.
-%EM Does a gap character really make sense? It seems like this changes the interpretation of the tmer length... I'd skip.
-Short deletions might be accommodated by introducing a ``gap'' letter to the alphabet,
-but insertions are problematic.
-
-%EM Full likelihood conjecture?
+Insertions and deletions are problematic, however.
+
+Furthermore, we have not been able to put
+quantitative bounds on how well we can approximate the full likelihood
+with a given T-mer shape.
+One reason this is difficult is that there are special cases of initial and final sequence
+for which the likelihood is \emph{not} approximable by looking only at T-mer counts;
+however, we believe such situations to be rare,
+so that the full likelihood can be well approximated on a set of outcomes of high probability.
+The TASEP model represents in some ways the worst-case of dependency propagation,
+and has an analytic solution \citep{schutz1997exact}.
+Futher theoretical work in this direction would be welcome.
+In practice,
+the approximation may be checked by computing the likelihood
+using successively longer T-mers,
+and checking for convergence as the length (of both base and overhang) increase.
 

### 2021-12-11 writeups/context-dependence.tex
github

@@
 % TODO after submission: think about the BCR parameter regime.
-
-% TODO include github information
 
@@
 These can then be compared to a standard Normal distribution.
+
+%%%%%%%
+\section{Software}
+
+The methods described here
+are implemented in R
+and freely available at \url{https://github.com/petrelharp/context}.
+The core implementation of T-mer counting and likelihood evaluation
+is provided in an R package, \texttt{contextual},
+and is supplemented by (documented) code to simulate from context-dependent models
+and perform maximum likelihood or Bayesian inference.
 

### 2021-12-11 writeups/context-dependence.tex
title tweak
a few cleanups and a question

@@
 
-%EM I'm going to suggest "dependence propagation" rather than information
-\title{Inference using context-dependent models of mutation via bounds on information propagation}
+\title{Inference using context-dependent models of mutation via bounds on propagation of dependency}
 \author[1]{Peter L Ralph}
@@
 
-To begin, we fit an unconstrained model of single-nucleotide substitution
+To begin, we fit an unconstrained non-context-dependent
+model of single-nucleotide substitution
 using a constrained quasi-Newton method (\verb|optim(.., method='L-BFGS-B')| in R).
@@
 \begin{table}
-%EM This table is BIG. And it seems like our conclusions only stem from the very biggest residuals.
-%EM What if we cut of anything that doesn't have an absolute Z-score bigger than 10?
-%PR True, but I'd like to show that the $z$-scores do down to something smaller?
-%PR I've cut out four lines...
 % cpg-plus-epsilon/32140/base-model-fit-5-3-l1-resids.2.1.l0.tsv
@@
 
-%EM "These patterns": Is it exactly the patterns in the true model?
-%EM Would we need to have prior information about CpGs to pick those additional terms?
-%PR No - I've added explanation - see above.
 After adding these patterns into the model and re-fitting, we searched for evidence of additional processes happening in the data.
@@
 and has an analytic solution \citep{schutz1997exact}.
-Futher theoretical work in this direction would be welcome.
+Further theoretical work in this direction would be welcome.
 In practice,
@@
 
-\section{Peeling algorithm}
+\section{Pruning algorithm}
 \label{ss:pruning_algorithm}
@@
 on all tips further away from $a$ (i.e., below that node if it was rooted at $a$).
+%EM I think I understand that the c is the number of tips on that side of the tree?
+%EM So we should think about y and z below of concatenation of short sequences on the tips of the tree, and (y,z) as the concatenation operator?
 For a node $w$ this will be stored as a $|\calS|^{n} \times |\calS|^{cm}$ matrix, for some $c$,
@@
   \item While there are $u_j$ whose children $w_1$ and $w_2$ both have matrices computed,
+  %EM I'm not sure why there is a single-item enumeration block here
     \begin{enumerate}
@@
         \begin{align}
+          %EM Is there a typo here? (x,z) appears 2x
           M(u_j)_{x,(y,z)} = \left( e^{t(w_1) G} M(w_1) \right)_{x,y} \left( e^{t(w_2) G} M(w_2)_{x,z} \right)_{x,z} .
@@
     \begin{align}
-      M(\rho)_{x,y} = \pi(x) \left( e^{t(u(\rho)) G} M(w(\rho)) \right)_{x,y}
+      M(\rho)_{x,y} = \pi(x) \left( e^{t(u(\rho)) G} M(w(\rho)) \right)_{x,y}.
     \end{align}

### 2021-12-11 writeups/context-dependence.tex
EM edits

@@
 
-The general case is a modification of Felsenstein's ``pruning'' algorithm\citep{felsenstein1981evolutionary}.
+The general case is a modification of Felsenstein's ``pruning'' algorithm \citep{felsenstein1981evolutionary}.
 
@@
 on all tips further away from $a$ (i.e., below that node if it was rooted at $a$).
-%EM I think I understand that the c is the number of tips on that side of the tree?
-%EM So we should think about y and z below of concatenation of short sequences on the tips of the tree, and (y,z) as the concatenation operator?
-For a node $w$ this will be stored as a $|\calS|^{n} \times |\calS|^{cm}$ matrix, for some $c$,
-and denoted $M(w)$.
+For a node $w$ this will be stored as a $|\calS|^{n} \times |\calS|^{cm}$ matrix, denoted $M(w)$,
+where $c$ is the number of tips for which $w$ lies on the path from $a$ to the tip.
 Rows of $M(u_j)$ sum to 1,
@@
 and $M(u_j)$ takes the value at $u_j$ as given.
+Below, we index columns of $M(u)$ by lists of short patterns,
+and leave the specific ordering to implementation.
+(For instance, if $c=2$ then a generic entry is $M(u)_{x,(y,z)}$,
+where the long pattern $x$ indexes the rows and the pair of short patterns $(y,z)$ indexes the columns;
+in the software all matrices retain column names making this unambiguous.)
 
@@
 
-  \item While there are $u_j$ whose children $w_1$ and $w_2$ both have matrices computed,
-  %EM I'm not sure why there is a single-item enumeration block here
-    \begin{enumerate}
-      \item let
+  \item While there are $u_j$ whose children $w_1$ and $w_2$ both have matrices computed, let
         \begin{align}
-          %EM Is there a typo here? (x,z) appears 2x
-          M(u_j)_{x,(y,z)} = \left( e^{t(w_1) G} M(w_1) \right)_{x,y} \left( e^{t(w_2) G} M(w_2)_{x,z} \right)_{x,z} .
+          M(u_j)_{x,(y,z)} = \left( e^{t(w_1) G} M(w_1) \right)_{x,y} \left( e^{t(w_2) G} M(w_2) \right)_{x,z} .
         \end{align}
-    \end{enumerate}
 
@@
 
-  \item For each $1 \le k < \ell$,
-    \begin{enumerate}
-      \item let
+  \item For each $1 \le k < \ell$, let
         \begin{align}
@@
         \end{align}
-    \end{enumerate}
 

### 2022-01-18 writeups/context-dependence.tex
Changing the title
authors and funding
Adding reviews and suggestions

@@
 
-\title{Inference using context-dependent models of mutation via bounds on propagation of dependency}
-\author[1]{Peter L Ralph}
+\title{Enabling inference for context-dependent models of mutation by bounding the propagation of dependency}
 \author[2,3,4,5]{Frederick A Matsen IV}
@@
 Dr.\ Matsen is an Investigator of the Howard Hughes Medical Institute.
+PLR's work was supported by NIH award R01HG010774
 
@@
 for nodes not on the path from $a$ to the root,
-we compute the chance that that each ``long'' sequence observed at that node
+we compute the chance that each ``long'' sequence observed at that node
 is matched by each possible combination of ``short'' sequences on the tips of the tree below it.

### 2022-03-07 writeups/context-dependence.tex
start at responses

@@
      }
+
+% Responses to reviews
+\input{reviews/review-response-commands}
+% set this to show line numbers and include responses to reviews or not
+\newif\ifreviewresponses
+\reviewresponsestrue  % include them
+% \reviewresponsesfalse  % don't include them
+\newcommand{\responsefile}{reviews/reviews.v1.tex}  % name of the review responses file
+
+\newif\ifsubmission
+\submissiontrue
+% \submissionfalse
+\ifsubmission
+    % % stuff for submission
+    \usepackage{lineno}
+    % allow line numbers around math environments
+    % from https://tex.stackexchange.com/questions/43648/why-doesnt-lineno-number-a-paragraph-when-it-is-followed-by-an-align-equation
+        \newcommand*\patchAmsMathEnvironmentForLineno[1]{%
+          \expandafter\let\csname old#1\expandafter\endcsname\csname #1\endcsname
+          \expandafter\let\csname oldend#1\expandafter\endcsname\csname end#1\endcsname
+          \renewenvironment{#1}%
+             {\linenomath\csname old#1\endcsname}%
+             {\csname oldend#1\endcsname\endlinenomath}}% 
+        \newcommand*\patchBothAmsMathEnvironmentsForLineno[1]{%
+          \patchAmsMathEnvironmentForLineno{#1}%
+          \patchAmsMathEnvironmentForLineno{#1*}}%
+        \AtBeginDocument{%
+        \patchBothAmsMathEnvironmentsForLineno{equation}%
+        \patchBothAmsMathEnvironmentsForLineno{align}%
+        \patchBothAmsMathEnvironmentsForLineno{flalign}%
+        \patchBothAmsMathEnvironmentsForLineno{alignat}%
+        \patchBothAmsMathEnvironmentsForLineno{gather}%
+        \patchBothAmsMathEnvironmentsForLineno{multline}%
+        }
+\else
+    % for the nice version
+\fi
 
@@
     In the Ising model, each site is labeled as either ``up'' or ``down'' ($+1$ or $-1$ respectively),
-    imagined as a string of $n$ magnetic dipoles,
-    and the energy associated with a given state $x$ is $E(x) = - \frac{1}{2} \beta \sum_{i=1}^{n-1} x_i x_{i+1} - \frac{1}{2} \gamma \sum_{i=1}^n x_i$.
+    imagined as a string of $L$ magnetic dipoles,
+    and the energy associated with a given state $x$ is $E(x) = - \frac{1}{2} \beta \sum_{i=1}^{L-1} x_i x_{i+1} - \frac{1}{2} \gamma \sum_{i=1}^L x_i$.
     Here $\beta$ represents inverse temperature, and $\gamma$ represents the strength of the magnetic field (here scaled by temperature).
@@
 We now describe concretely how the set of transition rates on patterns determines the transition rate matrix for complete sequences, that is,
-the $|\calS|^n \times |\calS|^n$ matrix $G(n)$ whose $(x,y)^\text{th}$ entry gives the instantaneous rate
-with which the process in state $x \in \calS^n$ jumps to state $y \in \calS^n$.
+the $|\calS|^L \times |\calS|^L$ matrix $G(L)$ whose $(x,y)^\text{th}$ entry gives the instantaneous rate
+with which the process in state $x \in \calS^L$ jumps to state $y \in \calS^L$.
 Recall that $x_i^{(h)} = (x_i, x_{i+1}, \ldots, x_{i+h-1})$ denotes the subsequence of length $h$ beginning at location $i$.
-For each $1\le i \le n$, pattern length $h > 0$, and patterns $u,v \in \calS^h$ define the relation
+For each $1\le i \le L$, pattern length $h > 0$, and patterns $u,v \in \calS^h$ define the relation
 \[
@@
 and for those positions $x$ matches with $u$ while $y$ matches with $v$.
+\revpoint{2}{3}
 
@@
 \]
-The rate ${G(n)}_{x,y}$ is then the sum of all transition rates that can take $x$ to $y$,
+The rate ${G(L)}_{x,y}$ is then the sum of all transition rates that can take $x$ to $y$,
 multiplied by the fitness term,
@@
 \begin{align} \label{eqn:G_defn}
-    {G(n)}_{x,y} = \phi\left(E(y)-E(x)\right) \, \sum_{i=1}^n \sum_{(\mu, u, v) \in J(i,x,y)}  \mu ,
+    {G(L)}_{x,y} = \phi\left(E(y)-E(x)\right) \, \sum_{i=1}^L \sum_{(\mu, u, v) \in J(i,x,y)}  \mu ,
 \end{align}
-and if there are no triples $(\mu,u,v)$ with $x \xrightarrow{i,u,v} y$ for some $i$, then ${G(n)}_{x,y}=0$.
+and if there are no triples $(\mu,u,v)$ with $x \xrightarrow{i,u,v} y$ for some $i$, then ${G(L)}_{x,y}=0$.
 To make this the generator matrix for a Markov process,
-we also set each $G(n)_{x,x}$ so that rows sum to zero.
+we also set each $G(L)_{x,x}$ so that rows sum to zero.
 Indeed, this matrix is very sparse because for most pairs $(x,y)$ there are no transition triples that can change $x$ into $y$.
@@
 \begin{align} \label{eqn:full_likelihood}
-    p_n(t;x,y) := \P\{ X(t) = y \mid X(0) = x \} = {\left(e^{tG(n)}\right)}_{xy} ,
+    p_n(t;x,y) := \P\{ X(t) = y \mid X(0) = x \} = {\left(e^{tG(L)}\right)}_{xy} ,
 \end{align}
 which would then provide a route to parameter estimation.
-In practice, the size of these matrices makes direct application obviously impractical for anything but very small $n$.
+In practice, the size of these matrices makes direct application obviously impractical for anything but very small $L$.
 This paper proposes an alternate strategy.
 
+%%%%%%%%%%%%%%%%%%%%%
 \section{Simulation}
@@
 or ``uniformization'' \citep{hobolth2009simulation}.
+\revpoint{2}{5}
 Briefly, this  works by first sampling a homogeneous Poisson process of times and locations of possible changes,
@@
 so the full likelihood function is given by \eqref{eqn:full_likelihood},
-but doing anything with the $|\calS|^L \times |\calS|^L$ matrix $G_L$ is clearly infeasible even for moderately sized $L$.
-We can, however, compute \eqref{eqn:full_likelihood} with smaller $G_n$,
+but doing anything with the $|\calS|^L \times |\calS|^L$ matrix $G(L)$
+is clearly infeasible even for moderately sized $L$.
+We can, however, compute \eqref{eqn:full_likelihood} with smaller $G(n)$,
 so the first thing that one might think to do is to break the sequence up into many blocks of length $m$,
@@
 to check for convergence.
+
+\paragraph{Boundary conditions} \revpoint{2}{1}
+Above, we have chosen to disallow changes whose patterns extend beyond the end of the sequence,
+i.e., the pattern $u$ (and hence, also the replacement $v$) of each transition triple
+must fit entirely within the existing sequence.
+Alternatively, we might assume periodic boundary conditions,
+which are natural for the Ising or TASEP models.
+Or, if the sequence we consider is embedded within a larger, unknown sequence,
+then we might assume a statistical model for the flanking sequence,
+averaging over the rates implied by states chosen from a stationary distribution.
+Our choice here of ``empty'' boundary conditions
+has the nice property that the total rate of changes on a given sub-segment of sequence
+is monotonically increasing in the length of the overhang,
+which gives us a simple check that the overhang is long enough
+(by seeing whether the total rate of change has converged).
+However, the software also allows periodic boundary conditions.
 
@@
         calculated using $(6,2,2)$ T-mers,
-        from 61 MCMC runs on independently simulated sequences of $10^6$ bases
+        from 61 MCMC runs on independently simulated sequences of $10^6$ spins
         differing at around 18.6\% of the sites.
@@
 (or sample from its posterior distribution).
+There is no conceptual barrier to applying these methods to lattices in higher dimensions,
+such as noisy images or particle systems used to model other physical processes.
+However, computational costs will grow with the size of the local neighborhood used
+even faster than in one dimension. \revpoint{2}{2}
 
@@
 
+%%%%%%%%%%%%%%%%%
+\includereviews
+
 \end{document}

### 2022-03-07 writeups/reviews/reviews.v1.tex
start at responses

%%%%%%
%%
%%  Don't reorder the reviewer points; that'll mess up the automatic referencing!
%%
%%%%%

\begin{minipage}[b]{2.5in}
  Resubmission Cover Letter \\
  {\it Evolution}
\end{minipage}
\hfill
\begin{minipage}[b]{2.5in}
    Joshua Schiffman \\
    \emph{and} Peter Ralph \\
  \today
\end{minipage}

\vskip 2em

\noindent
{\bf To the Editor(s) -- }

\vskip 1em

Our revised manuscript.

\noindent \hspace{4em}
\begin{minipage}{3in}
\noindent
{\bf Sincerely,}

\vskip 2em

{\bf 
Erick Matsen and
Peter Ralph
}\\
\end{minipage}

\vskip 4em

\pagebreak

%%%%%%%%%%%%%%
\reviewersection{1}

\begin{quote}
In their article ``Enabling inference for context-dependent models of
mutation by bounding the propagation of dependency'' Matsen and Ralph
discuss methodology to infer mutation rates in context dependent
models of DNA evolution. With the advent of more sequence data it
became clear that the first very simple models of DNA evolution are
only a crude approximation to the ``real'' process. The extension of
such models to also include neighbor dependencies is very timely.
However, the inclusion of such processes is easier said than done -
without ``bounding the propagation of dependency'' such models are
computational not tractable.

The paper is very well written and gives sufficient detail.

The source code of the mentioned R package is available at github.com.
\end{quote}

Thanks very much for the kind words!

%%%%%%%%%%%%%%
\reviewersection{2}

\begin{quote}
In this manuscript the authors address the problem how to infer model
parameters and model ``ingredients'' (allowed transitions) in models of
sequence evolution, in which the rate of mutations depends on the
local sequence context of a site. This is a significant problem and
surprisingly difficult. The authors present a solution where they
decompose the full transition matrix into elements that map
subsequences into even shorter subsequences. This is an approximation,
but they give a bound on the approximation and even more importantly
show that the method works on two sets of simulated data. They then
apply the method to real sequence data from humans and chimpanzee and
find interesting rates for different processes relevant for sequence
evolution.
\end{quote}

% > This is overall a very nice work, which in addition is presented very
% > lucidly, but there are a few points the authors should address before
% > publication:

\begin{point}{}
Could the authors provide some thoughts about sequences with
periodic boundary conditions. That would be a natural setup both for
the TASEP that they use as one of their examples and for bacterial
genomes. Would they expect the method to still work and/or does
anything need to be changed?
\end{point}

\reply{
This is something that we thought about a good deal early on in this project,
but removed to avoid introducing additional complexities.
We agree it's a natural question to ask,
and so have inserted some dicussion \revref.
}

\begin{point}{}
Since the authors in the introduction bring up contexts like wild
fires, land use patterns, and cell-cell interactions, could the
authors in the discussion at least speculate about the possibility of
extending this approach to 2D or 3D systems?
\end{point}

\reply{
Good idea; we've inserted a brief note. \revref
}

\begin{quote}
3) There are a few places where the presentation could be improved:
\end{quote}

\begin{point}{}
It is somewhat confusing that in the definition of the
% x --i,u,v--> y
$ x \xrightarrow{i,u,v} y$
relation toward the end of section 1 \revref{}
the indices $i$ of the sequences
$x$ and $y$ are one-based while the indices $k$ of the patterns $u$ and $v$ are
zero-based. Can this be made more consistent?
\end{point}

\reply{
Thanks for pointing this out,
but after trying a few solutions out
we'd like to keep it as-is.
We think that the indexing on transition triples (for $u$ and $v$)
should certainly be zero-based, because they are offsets.
On the other hand, there would be no problem doing zero-based indexing for sequences ($x$ and $y$),
but since these indexes are absolute positions it seems less important,
and we think that $\sum_{i=1}^n$ is slightly more clear than $\sum_{i=0}^{n-1}$.
(We tried inserting a note about this, but felt that even the note made things more confusing.)
}

\begin{point}{}
It is quite confusing that the length of the sequence is $n$ in
section 1 and then switches to $L$ in section 2. Later $n$ becomes the
length of a subsequence block. The notation should ideally be made
consistent between the sections, or, at a minimum, the reader should
be warned of these notation changes.
\end{point}

\reply{
This is a good point; we've made this change.
(We had been trying to arrange things to introduce the generator matrix
using the notation $G(n)$ rather than $G(L)$,
as that's how it would mostly appear later,
but this was not making this more clear.)
}

\begin{point}{}
This referee would take some objection to calling the simulation
algorithm in section 2 a Gillespie algorithm. The whole point of the
Gillespie algorithm is to advance time in such a way that every step
of the algorithm results in a transition. This is not the case in the
algorithm the authors present since they always step forward with the
maximal possible rate but then allow non-transitions.
\end{point}

\reply{
We don't entirely agree here:
one could also say that
the point of the Gillespie algorithm is to obtain an exact trajectory
of a continuous-time system
by only considering those discrete times when there is a transition.
This view doesn't disallow ``do-nothing'' transitions.
We wouldn't mind removing the phrase entirely,
but on the other hand the phrase is widely used
(see for instance a nice Wikipedia page and 8,500 google scholar hits),
so we'd like to leave it in to give readers new to the subject
some more pointers into the background literature.
}

\begin{point}{3d}
There is an inconsistency between equations (5), (6), and the
equation right below (6). In the numerator the identification between
the symbols must be that (n,l,r) in (5) is (2l+1,l+1,l) in (6).
However, the upper index on x is n in (5) but 4l+1 != 2l+1 in (6).
That is true in the denominator as well. Also,
m=n-l-r=(2l+1)-(l+1)-l=0 != 2l+1 (if the authors meant that n was
actually 4l+1, 4l+1-(l+1)-l=2l != 2l+1 does not work much better
either) and in the denominator it is not clear why m should change
given that l and r just get swapped and m only depends on their sum.
Maybe this referee totally misunderstands the notation here, but this
fact alone means that something needs to be improved in the
presentation in this area.
\end{point}

\reply{
TODO: Sorry about this. I suggest the solution is to change the
first subscript on p to 4l+1.
}

\begin{point}{3e}
It is somewhat counterintuitive that the caption of figure 4 talks
of ``bases''. This seems like taking the analogy a bit too far.
Shouldn't it be ``spins''?
\end{point}

\reply{
Good point; done.
}

\begin{point}{3f}
Remove a ``that'' in ``that that'' in the fifth line of appendix C.
\end{point}

\reply{
Fixed, thanks.
}

### 2022-03-07 writeups/context-dependence.tex
Will comments

@@
 
-    Counteracting this trend towards increased G/C is GC-biased gene conversion \citep{glemin2015quantification},
-    which acts effectively as a selective pressure against G or C bases.
+    Counteracting this trend towards decreased G/C
+    is GC-biased gene conversion \citep{glemin2015quantification},
+    which acts effectively as a selective pressure in favor of G or C bases.
     We model the evolution of a \emph{single} sequence, not a population of sequences,
@@
     the total rate it occurs at in the population is $\mu N_e$.
-    GC-biased gene conversion effectively means that sequences with more G/Cs are selected against.
+    GC-biased gene conversion effectively means that sequences with more G/Cs are selected for.
     This is incorporated as follows: set the energy of patterns \nC{} and \nG{} to be $s$,
     so that $E(y)-E(x)$ is the net change in number of G's and C's multiplied by $s$.
+    We then expect $s$ to be positive.
     Also, define $\phi(e)$ to be the expected rate of fixations of ``selected'' mutations
@@
     and the energy associated with a given state $x$ is $E(x) = - \frac{1}{2} \beta \sum_{i=1}^{L-1} x_i x_{i+1} - \frac{1}{2} \gamma \sum_{i=1}^L x_i$.
-    Here $\beta$ represents inverse temperature, and $\gamma$ represents the strength of the magnetic field (here scaled by temperature).
+    Here $\beta$ controls the strength of coupling between neighboring dipoles,
+    and $\gamma$ controls the strength of an external magnetic field
+    (and both are scaled by temperature).
     The associated stationary distribution on configurations is proportional to $\exp(-E(x))$.
@@
   x_k = y_k \quad &\text{for } k<i \\
-  x_{i+k} = u_k \quad &\text{for } 0 \le k < h \\
-  y_{i+k} = v_k \quad &\text{for } 0 \le k < h,\ \text{and} \\
+  x_{i+j} = u_j \quad &\text{for } 0 \le j < h \\
+  y_{i+j} = v_j \quad &\text{for } 0 \le j < h,\ \text{and} \\
   x_k = y_k \quad &\text{for } k\ge i+h ,
@@
 \[
-  \prod_{k=0}^{n/m} p_m(x_{km+1}^{(m)},y_{km+1}^{(m)}) .
+  \prod_{k=0}^{n/m} p_m(t;x_{km+1}^{(m)},y_{km+1}^{(m)}) .
 \]
@@
 and that will be sufficient to compute the full likelihood,
-at least arbitrarily.
-To do this, let $p_{n,\ell,r}(x,y)$ for $\ell + r < n$ denote
+at least approximately.
+To do this, let $p_{n,\ell,r}(t;x,y)$ for $\ell + r < n$ denote
 the probability that an initial subsequence $x$ of length $n$, after time $t$,
@@
 
-These $p_{n,\ell,r}(x,y)$ probabilities can be found by marginalizing $p_n$,
+These $p_{n,\ell,r}(t;x,y)$ probabilities can be found by marginalizing $p_n$,
 summing over the possible initial and final substrings of $x$
@@
 where $a \join y \join b$ is the sequence composed of concatenating $a$, $y$, and $b$ together in that order.
-We expect $p_{n,\ell,r}(x,y)$ to be asymptotically correct when $\ell$ and $r$ are large (and $t$ is fixed).
+We expect $p_{n,\ell,r}(t;x,y)$ to be asymptotically correct when $\ell$ and $r$ are large (and $t$ is fixed).
 In Appendix~\ref{ss:approx_pf}, we show that this is so, namely that
@@
 is later observed to have its center matching $y$
-is close to $p_{n,\ell,r}(x,y)$, regardless of wider context.
+is close to $p_{n,\ell,r}(t;x,y)$, regardless of wider context.
 Formally, for each $i$,
@@
     \approx
-    p_{n,\ell,r}(x,y)
+    p_{n,\ell,r}(t;x,y)
     ,
@@
         =
-        \prod_{i=1}^{L-m+1} p_{n, \ell, r}\left(x_i^{(n)}, y_{i+\ell}^{(m)}\right).
+        \prod_{i=1}^{L-m+1} p_{n, \ell, r}\left(t;x_i^{(n)}, y_{i+\ell}^{(m)}\right).
 \end{align}
@@
         \frac{
-            p_{2\ell+1,\ell+1,\ell}(x_{i-3\ell}^{(4\ell+1)}, y_{i-2\ell}^{(2\ell+1)})
+            p_{4\ell+1,\ell,\ell}(t; x_{i-3\ell}^{(4\ell+1)}, y_{i-2\ell}^{(2\ell+1)})
         }{
-            p_{2\ell+1,\ell,\ell+1}(x_{i-3\ell}^{(4\ell+1)}, y_{i-2\ell}^{(2\ell)})
+            p_{4\ell+1,\ell,\ell+1}(t; x_{i-3\ell}^{(4\ell+1)}, y_{i-2\ell}^{(2\ell)})
         }.
@@
 \end{align}
+\revpoint{2}{6}
 
@@
 \[
-    \like(y \mid x) \approx \frac{ A_{2\ell+1, \ell+1, \ell}(x,y) }{ A_{2\ell+1, \ell, \ell+1}(x,y) }.
+    \like(y \mid x) \approx \frac{ A_{4\ell+1, \ell, \ell}(x,y) }{ A_{4\ell+1, \ell, \ell+1}(x,y) }.
 \]
@@
 %     &\approx
-%         \sum_{a \in \calS^n} \sum_{b \in \calS^{n-2\ell}} n_\ell(x,y;a,b) \log p_{n,\ell,\ell}(a,b)
+%         \sum_{a \in \calS^n} \sum_{b \in \calS^{n-2\ell}} n_\ell(x,y;a,b) \log p_{n,\ell,\ell}(t;a,b)
 %         -
-%         \sum_{c \in \calS^n} \sum_{d \in \calS^{n-2\ell-1}} n_\ell(x,y;c,d) \log p_{n,\ell,\ell+1}(c,d)
+%         \sum_{c \in \calS^n} \sum_{d \in \calS^{n-2\ell-1}} n_\ell(x,y;c,d) \log p_{n,\ell,\ell+1}(t;c,d)
 % \end{align}
@@
 \begin{align}
-    p_{n}(a,c) = \left( e^{t G(n)} \right)_{a,c} ,
+    p_{n}(t;a,c) = \left( e^{t G(n)} \right)_{a,c} ,
 \end{align}
@@
 \begin{align} \label{eqn:Tmer_trans}
-    p_{n,\ell,r}(t,a,b) = \sum_{c \in \calS^{n}} \left( e^{t G(n)} \right)_{ac} U_{cb} ,
+    p_{n,\ell,r}(t;a,b) = \sum_{c \in \calS^{n}} \left( e^{t G(n)} \right)_{ac} U_{cb} ,
 \end{align}
@@
     =
-    \sum_{z \in \calS^{n}} \pi(z) \, p_{n,0,0}(t_X,z,x) \, p_{n,\ell,r}(t_Y,z,y) .
+    \sum_{z \in \calS^{n}} \pi(z) \, p_{n,0,0}(t_X;z,x) \, p_{n,\ell,r}(t_Y;z,y) .
 \end{align}
@@
     \sum_{z \in \calS^{n}}
-        \pi(z) \, p_{n,0,0}(t_X,z,x) \, p_{n,\ell,r}(t_Y,z,y)
+        \pi(z) \, p_{n,0,0}(t_X;z,x) \, p_{n,\ell,r}(t_Y;z,y)
     =
@@
 along the way substantial encouragement and useful suggestions were provided by
-Simon Tavar\'e, Rasmus Nielsen, Graham Coop, Yaniv Brandvain, and Sergey Nuzhdin.
+Will DeWitt, Simon Tavar\'e, Rasmus Nielsen, Graham Coop, Yaniv Brandvain, and Sergey Nuzhdin.
 Many thanks are also due to Jessica Crisci for curating the human-chimp data.

### 2022-03-07 writeups/reviews/reviews.v1.tex
Will comments

@@
 
-\begin{point}{3d}
-There is an inconsistency between equations (5), (6), and the
-equation right below (6). In the numerator the identification between
-the symbols must be that (n,l,r) in (5) is (2l+1,l+1,l) in (6).
-However, the upper index on x is n in (5) but 4l+1 != 2l+1 in (6).
+\begin{point}{}
+There is an inconsistency between equations \eqref{eqn:composite_like}, \eqref{eqn:full_approx},
+and the equation right below \eqref{eqn:full_approx}. In the numerator the identification between
+the symbols must be that $(n,\ell,r)$
+in \eqref{eqn:composite_like} is $(2\ell+1,\ell+1, \ell)$ in \eqref{eqn:full_approx}.
+However, the upper index on $x$ is $n$ in \eqref{eqn:composite_like}
+but $4\ell+1 \neq 2\ell+1$ in \eqref{eqn:full_approx}.
 That is true in the denominator as well. Also,
-m=n-l-r=(2l+1)-(l+1)-l=0 != 2l+1 (if the authors meant that n was
-actually 4l+1, 4l+1-(l+1)-l=2l != 2l+1 does not work much better
-either) and in the denominator it is not clear why m should change
-given that l and r just get swapped and m only depends on their sum.
+$m=n-\ell-r=(2\ell+1)-(\ell+1)-\ell=0 \neq 2\ell+1$
+(if the authors meant that $n$ was
+actually $4\ell+1$, then $4\ell+1-(\ell+1)-\ell=2\ell != 2\ell+1$ does not work much better
+either) and in the denominator it is not clear why $m$ should change
+given that $\ell$ and $r$ just get swapped and m only depends on their sum.
 Maybe this referee totally misunderstands the notation here, but this
@@
 \reply{
-TODO: Sorry about this. I suggest the solution is to change the
-first subscript on p to 4l+1.
+Whoops - sorry about that.
+The inconsistency stemmed from our switching around notation in a previous draft
+but forgetting to update it in equation \eqref{eqn:full_approx}.
+(This is confusing, which is why we have Figure~\ref{fig:Tmers}C to refer to.)
+The notation is now fixed. \revref
 }

### 2022-03-07 writeups/reviews/review-responses.tex
actually commit the reviews now

%%%%%%
%%
%%  Don't reorder the reviewer points; that'll mess up the automatic referencing!
%%
%%%%%

\begin{minipage}[b]{2.5in}
  Resubmission Cover Letter \\
  {\it Evolution}
\end{minipage}
\hfill
\begin{minipage}[b]{2.5in}
    Joshua Schiffman \\
    \emph{and} Peter Ralph \\
  \today
\end{minipage}

\vskip 2em

\noindent
{\bf To the Editor(s) -- }

\vskip 1em

We are writing to submit a revision of our manuscript, ``{System drift and speciation}'' (ID 18-0316), for your review. 
We sincerely apologize for the delayed resubmission but hope that you will find the
revised manuscript significantly improved.

We thank you and the two reviewers for providing constructive suggestions 
and feedback that has helped to improve the manuscript. We have followed 
the suggestions closely and hope to have fully addressed all the concerns and 
suggestions as detailed in our point-by-point response below.

In our original submission, we studied the effect of gene regulatory network evolution on speciation,
first by applying linear system theory to characterize the large space of phenotypically equivalent network organizations,
and second by using quantitative genetics to show that neutral genetic drift on this network space can lead to speciation over plausible timescales.

We agree with the reviewers that the combination of these two approaches, while interesting,
was not seamless and that the quantitative genetics section required further development, 
and thus have followed the recommendation of both the Editor and Associate Editor,
by breaking the manuscript into two papers.
One paper -- the present revision -- covers the application of system theory to regulatory networks,
and the other (which we plan to resubmit soon) will cover the quantitative genetics.
In addition to detailed responses to the reviews,
we are submitting a color-coded diff to make it easy to see what has been removed.
We are happy with the result -- as predicted by the reviewers, the resulting paper stands alone
without the quantitative genetics results.

We hope that by focusing this paper on system theory, we have clarified our results, which include a mathematical description of phenotypically equivalent network space 
and its implications for speciation under neutral genetic drift. These results suggest that the space of phenotypically equivalent network organizations is substantial,
and often not reproductively compatible, leaving many opportunities for independently evolving populations to become reproductively incompatible.
The linear system framework applied in this manuscript explicitly describes possible molecular pathways and naturally predicts Haldane's rule.

We would like to note that the previous version of this paper as a preprint has since received 10 citations as well as attracted
substantial positive attention as a preprint (e.g., see tweets at
\url{https://www.biorxiv.org/content/10.1101/231209v2}). Furthermore, this paper has been downloaded more than 2,326 times (\url{https://rxivist.org/papers/24911}),
making it the 155$^{\text{th}}$ most downloaded evolutionary biology paper available on the \emph{bioRxiv}, as of this writing.

\noindent \hspace{4em}
\begin{minipage}{3in}
\noindent
{\bf Sincerely,}

\vskip 2em

{\bf 
Joshua Schiffman and
Peter Ralph
}\\
\end{minipage}

\vskip 4em

\pagebreak

%%%%%%%%%%%%%%
\reviewersection{AE}

\begin{quote}
I have received the evaluations of two reviewers, and I have read the paper
myself. First, sorry for the long time it took to review your manuscript; it
was reviewed on the timescale of mathematicians (for Reviewer2), but both
reviewers provided thorough evaluations of your work (which in my opinion is
better than a quick but superficial review).
\end{quote}

Thanks for the apology,
but given the delay in our resubmission, we can hardly complain, at this point.

\begin{quote}
Your manuscript is quite long already, and the reviewers' suggestions of
modifications and clarification, which need to be implemented, will make the
manuscript even longer. I however share the reviewers' opinion that the
manuscript's two parts are only loosely related. My suggestion is therefore to
publish them separately. Although connecting the systems biology and the
quantitative genetics parts is indeed an exciting endeavor, this is not really
achieved in the current version of the manuscript, and it would be more
profitable to first better describe each part separately. Regarding the
connexion between the two parts, please also pay particular attention to R2's
first specific comment about dimensionality.

This may be because I am more familiar with Fisher's geometric model than
systems biology models, but it seems to me that the first part is more novel,
and should therefore deserve your attention first, should you follow my
suggestion of publishing the two parts separately.
\end{quote}

Thanks for the careful attention to the paper.
We have decided to go with this suggestion,
and have split the paper in two;
this paper is the ``systems biology'' part of it.
We've left in reviewer comments regarding the ``Fisher's geometric model'' portion below,
commenting that the relevant bit ``has been removed''.

\begin{point}{Format:}
    Please add line numbers to your manuscript; it is straightforward
with the lineno package in LaTeX.  Please also do not increase text width too
much, as this decreases legibility (LaTeX's default settings already optimize
the number of characters per line).  Finally, be careful about not confusing
citep and citet (a lot of missing parentheses around citations in the second part).
\end{point}

\reply{
    Apologies!
  We've added line numbers and corrected improperly formatted citations. 
}

\begin{point}{all figures:}
    Please ensure that all axes legends and labels are big enough
and not distorted (same of arrow labels on the diagrams). Please also make sure
that the lines are thick enough to be visible.
\end{point}

\reply{
  We have regenerated figures to address this.
}

\begin{point}{Above eq (2):}
    ``Under quite general assumptions'': quickly mention them?
\end{point}

\reply{
  Good idea.
  The formula is quite general,
  and a more precise statement of the assumptions would veer too far afield
  (perhaps needing definitions of function spaces),
  but we've added a hopefully clarifying comment.
  \revref.
}

\begin{point}{eq (2):} 
    Do you need the first term on the rhs given that you assume that $\kappa(0)=0$?
\end{point}

\reply{
    Ah, good point. We've removed it. \revref
}

\begin{point}{Example 1:}
    ``and so its phenotype'' whose phenotype? (the subject of the sentence is ``the input'')
\end{point}

\reply{
    Fixed. \revref
}

\begin{point}{p5, Note on implementation}
    ``the closure of span(B)'' may not be
    understandable by most of Evolution's readers; please explain what this means.
\end{point}

\reply{
    We've added a clarifying note \revref{}
    but we're hoping that readers not familiar with linear algebra will realize
    that they can skip this ``note on implementation'' without loss of continuity.
}

\begin{point}{Figures 1-3:}
    In figures 1 and 3, there are two types of arrow heads (+ and -
effects), but not in Figure 2, probably because the sign of the effect depends
on the value of $\tau$. Maybe use a  different type of arrow heads for
interactions that may change signs, to avoid the confusion with positive
effects?
\end{point}

\reply{
  We think that adding a third type of arrowhead may only add to the confusion, so now we are using only the standard arrowhead.
}

\begin{point}{p7, Hybrid incompatibility:}
    Please provide a rationale for choosing a particular weight $\rho$.
\end{point}

\reply{
    Good suggestion; added. \revref
}

\begin{point}{p11, 2nd paragraph}
    Try to better relate these different studies to yours
(e.g., what you add, what they do and you do not). In addition to R1's
suggestions, consider \citet{weinreich2013fishers} and \citet{blanquart2016epistasis}.
% Weinreich, D. M. and Knies, J. L. (2013), FISHER'S GEOMETRIC MODEL OF ADAPTATION MEETS THE FUNCTIONAL SYNTHESIS: DATA ON PAIRWISE EPISTASIS FOR FITNESS YIELDS INSIGHTS INTO THE SHAPE AND SIZE OF PHENOTYPE SPACE. Evolution, 67: 2957-2972. doi:10.1111/evo.12156
% François Blanquart and Thomas Bataillon Epistasis and the Structure of Fitness Landscapes: Are Experimental Fitness Landscapes Compatible with Fisher’s Geometric Model?  GENETICS Early online April 6, 2016
\end{point}

\reply{
    This part has been removed.
}

\begin{point}{p12 Hybridization:}
    What is $\mathcal{X}$?
\end{point}

\reply{
    This part has been removed.
}

\begin{point}{p15}
    ``The importance of including neutral directions in these models, which
is not usually done'' -- Do not line of isofitness correspond to neutral
directions, and aren't these neutral directions already included in those
models?
\end{point}

\reply{
    This part has been removed.
}

%%%%%%%%%%%%%%
\reviewersection{1}

\begin{quote}
The ms combines two parts that are only loosely linked: one ``systems biology''
part and a ``popgen'' part. The combination of both parts is appealing, primarily
because we lack such combined approaches and it is not easy to come up with an
adequate framework. Below, I will discuss both parts in turn.
\end{quote}

This manuscript is mostly the ``systems biology'' part,
but we think that the potential applications to ``popgen'' are still reasonably visible.
(And, we plan to resubmit the ``popgen'' part separately.)

% \begin{quote}
%     \textbf{(I)} The first (systems biology) part makes use of the concept of
% ``nonidentifiability'' of chemical reaction networks and develops these concepts
% in the context of a genotype to phenotype map. In particular, this map is
% analyzed for the case where it can be modeled as a linear system.
% \end{quote}

\begin{point}{}
    First note that the concept of ``nonidentifiability'' is closely related to
themes that have been discussed in evolutionary biology for a long time, under
names like ``mutational robustness'', ``network neutrality'', ``canalization'',
``redundancy'' etc. In particular, if you characterize your model by ``many
distinct (and mutationally connected) molecular pathways can realize identical
phenotypes'', note that this essentially describes what has been called a
``neutral network'' and studied under this name in many articles. Also concrete
developmental networks have been studied in this context (eg, von Dassow etal
2000, Nature 406:188–192). Discussion of this literature - and more importantly
discussion of terms and notions used here relative to terms that have been used
elsewhere - is largely missing. I won't be able to summarize all this here, but
an older review is the Evolution Perspective piece ``Evolution and detection of
genetic robustness'' (de Visser et al 2003). You can work work backward and
forward from there.  
\end{point}

\reply{
  Thanks for the additional suggestions!
  We've reworked the introduction, adding these (and many other) additional references
  into the (large and growing) literature. \revref
  We hope it's more representative now.
}

\begin{point}{}
    Related to this: I find it non-intuitive to start the introduction with the
notion of ``nonidentifiability'', which is a consequence rather than a cause, the
cause being what you later call ``phenotypic equivalence'' and what had been
discussed under other names in other articles.
\end{point}

\reply{
  Good point. In the introdcution we now introduce nonidentifiability after discussing general cases of phenotypic equivalence. \revref
}

\begin{point}{}
    Still related: You give long lists of references in the introduction, but do
not provide the reader with any information about specific contributions. A bit
more would be helpful (starting with True and Haag coined the term
``developmental system drift'', etc).
\end{point}

\reply{
    We've tried to explain things better in the revised Introduction. \revref
    However, we find assigning specific contributions to be difficult
    because similar ideas appear in many disparate parts of the literature.
}

\begin{point}{4.}
    Your ms does not simply assume that there is a large ``neutral network''
underlying a phenotype, but suggests a mechanistic underpinning to create this
neutral space. Since much of the appeal of the first part of the paper is
connected to this fact, I'd like to see some more in-depth discussion.
\textit{(suggestions to follow)}
\end{point}

\reply{
    Yes, that's right! We've followed the suggestions (see below),
    and have added additional discussion to the Discussion.
}

\begin{point}{(mechanistic models)}
    You should probably mention somewhere that mechanistic models of neutral
  spaces exist in the (quite different) context of RNA folding (papers by Hofacker, Fontana, etal)
\end{point}

\reply{
  Added \revref.
}

\begin{point}{(mechanistic models)}
    For development, Andreas Wagner once suggested a discrete version of a linear
  model (iterated matrix multiplication). Variants of this have later been used
by others (eg Draghi and GP Wagner, I think). How is your model related to these approaches?
\end{point}

\reply{
    This is a natural quesiton. We've addressed it after introducting the model.
    \revref
}

\begin{point}{(mechanistic models)}
    You model development as a linear system and this assumption is essential
   for all further steps (the explicit solution of the dynamics and the Kalman
decomposition). For me, the only justification of a linear model (other than
mathematical convenience) is local approximation. Do you agree? This will be
important for the second part of the ms, see below.
\end{point}

\reply{
    We do agree, and have clarified this in the Discussion. \revref
}

\begin{point}{(mechanistic models)}
    You gain a lot of ``neutrality'' by the assumption that the dimension of the
  ``krypotype'' is larger than the dimension of the trait. If the trait is the
expression of a gene at the top of some pathway, this is certainly true.
However, if we include further traits that are affected by the same genes
(i.e., pleiotropy) this is no longer clear. The space of ``all traits under
selection'' of an organism is awfully high-dimensional and pleiotropy is
wide-spread (even if we do not believe in ``omnigenic'' models). It is not clear
to me from what kind of data we could learn more about these dimensions and I
do not expect an answer in your ms, but the issue deserves more discussion.
Currently, ``pleiotropy'' is not even mentioned in the ms.   
\end{point}

\reply{
    Counterintuitively, it is not necessary to have the kryptotype have a higher dimension
    than the phenotype to have degrees of freedom.
    The precise details depend on what you want to treat as fixed or evolvable --
    for instance, should the output matrix $C$ be subject to drift? --
    but we don't think it's controversial to suppose that there's a bunch of genes
    whose expression levels aren't the direct target of selection.
    We have added discussion of this in the Discussion. \revref
}

\begin{point}{Introduction:}
    ``Genotypes encoding identical phenotypes can even persist stably within a species'' - if there is population structure, I suppose.
\end{point}

\reply{
    Yes, that's right; we've clarified this. \revref
}

\begin{point}{Introduction:}
    ``It is not a new observation that there is often more than one way to do the same thing, and that this may lead to speciation'' - Isn't this what people would call a neutral Dobzhansky-Muller incompatibility?
\end{point}

\reply{
    That'd be one way to do it --
    and, it turns out, Bateson thought BDMIs would probably be neutral;
    we've added this. \revref{}
    (The quote from \citet{bateson1909heredity} is
    \textit{``Now if the sterility
of the cross-bred be really the consequence of the meeting of two
complementary factors, we see that the phenomenon could only be
produced among the divergent offspring of one species by the acquisition of at least two new factors; for if the acquisition of a single
factor caused sterility the line would then end. Moreover each factor
must be separately acquired by distinct individuals, for if both were
present together, the possessors would by hypothesis be sterile. And
in order to imitate the case of species each of these factors must be
acquired by distinct breeds. The factors need not, and probably would
not, produce any other perceptible effects; they might, like the colour-
factors present in white flowers, make no difference in the form or
other characters.``})
}

\begin{point}{Results:}
    (Eq. 1) development as a linear system: has this been done before (probably yes?) - If so, references?
\end{point}

\reply{
    We wouldn't be surprised if it had, but we're not aware of it --
    other papers all seem to take the discrete-time approach as in \citet{wagner1994evolution}.
}

\begin{point}{Results:}
    ``Of course, neither of these are necessarily true for real systems'' - delete ``necessarily''?
\end{point}

\reply{
    This sentence has been removed.
}

\begin{point}{Eq. 3:}
    What does a coordinate change mean biologically? In our JC, we discussed for a while whether this is just a different parametrization of the exact same biological thing (like transformation to principle components). I think I now understand that it is more, but this should be explained better.
\end{point}

\reply{
    Biologically, a coordinate change may produce an unrecognizably different system
    (repression changed to activation or the like);
    we've added a hopefully clarifying sentence. \revref
}

\begin{point}{}
    ``Since gene networks can grow or shrink following gene duplications and deletions,
these additional degrees of freedom can apply in principle to any system.'' - are there examples of equivalent gene networks with a different size due to gene duplication or loss?
\end{point}

\reply{
    We're not aware of any biological examples of ``equivalent'' networks
    in more than one species at all,
    but we think it's not clear whether we understand the dynamics of \emph{any} gene regulatory network
    well enough to determine equivalence,
    so this is an absence of evidence, not evidence of absence.
    (There are plenty of examples of homologous networks with missing genes,
    but there's so many uncertainties we don't think bringing this up will clarify things.)
    % PLR: I don't think we need this, as the reviewers are aware of these examples.
    % Over larger evolutionary distances there is some evidence that structurally distinct
    % isofunctional networks of different size do indeed exist. \emph{E.g.}, the networks underlying circadian
    % rhythm in cyanobacteria and mammals, as discussed by \citet{sancar2008intelligent}.
    % It seems possible, in principle, that, on a smaller scale, such differences could exist between more closely related populations.

}

\begin{point}{Kalman decomposition:}
    Define the sub-matrices directly when first used.
\end{point}

\reply{
    Good point; we now say ``in block matrix form'' and add more explanation. \revref
}

\begin{point}{Figure 2:}
    Colors are hardly visible
\end{point}

\reply{
    We've regenerated figures to improve clarity. \revref
}

\begin{quote}
    \textbf{(II)} The popgen part. This part models the ``system drift'', which is nothing else
but neutral drift of a population along a high-fitness ridge in an epistatic
landscape (conceptual figure 4), right? If two populations in allopatry drift
in different directions, this can lead to hybrid incompatibilities, which are
uncovered upon secondary contact. Fitness is modeled by the weighted distance
from the optimal impulse response function (eq 5). This is a natural
assumption.
\end{quote}

\reply{
    We've removed the ``popgen'' part, but some of the comments below still apply.
}

\begin{point}{}
    For system drift, you assume that ``Selection will tend to restrain this
motion, but movement along the optimal set N is unconstrained, and so we expect
the population mean to drift along the optimal set like a particle diffusing.''
I see two major problems with this view. Both lead to slower divergence and
therefore run against your conclusions.
    \textit{(further points to follow)}
\end{point}

\reply{
    We have removed this part.
}

\begin{point}{(drift along a ridge)}
    In the presence of epistasis, evolution on a neutral network (a high-fitness
    ridge) is *not* due to drift alone, but also affected by (weak second order)
    selection in favor of mutational robustness / genetic canalization. In contrast
    to what you write, diffusion on the set of network coefficients corresponding
    to the optimal phenotype is not unbiased - even if the optimal phenotypes do
    indeed all have the same fitness. Instead, selection drives the population to
    ``thicker'' parts of the network where the mean fitness of the population
    (including a cloud of mutants) is higher that on a narrow ridge. This is the
    basis of the evolution of robustness/canalization. It is possible to account
    for this effect, see \citet{hermisson2003epistasis,alvarez2009effects}, 
    or also \citet{rice1998evolution,neutral_network_physics}.
    Note that epistasis is necessary for the neutral evolution of incompatibilities,
    which is what you are aiming for.
\end{point}

\reply{
    Thanks for pointing out this important omission.
    We have added mention of this (and these citations),
    including a paragraph to the discussion. \revref
}

\begin{point}{(drift along a ridge)}
    a second problem results from the fact that evolution on a high-fitness ridge
  often requires coordinated changes at many loci. Take the oscillator system
that you use as an example: simultaneous changes at two genes are required to
maintain the phenotype. This leads to a phenomenon called ``adaptive inertia''
(see \citet{baatz1997adaptive} and \citet{alvarez2009effects}),
which effectively slows down the movement along the ridge considerably. This
problem applies, in particular, in a ``house-of-cards'' mutation regime when
where rarely two mutations occur together on the same haplotype. In small
populations, it typically requires that populations drift through shallow
fitness valleys. While this is possible, it slows down the process.  
\end{point}

\reply{
    Again, this is an important point covered in additional discussion.
    \revref
}

\begin{point}{(drift along a ridge)}
    The relevance of both effects could be studied by simulations in a simple
example (eg the oscillator that is used as an illustration in the ms anyway).
\end{point}

\reply{
    We agree, and hope to do this in our follow-up for this part of the paper.
}

\begin{point}{}
    If speciation due to accumulation of incompatibilities does not occur in
allopatry, but under (even weak) gene flow, some degree of positive selection
is always needed (\citet{bank2012limits} for a proof). This also
means that even weak gene flow will counteract the process described in the ms.
\end{point}

\reply{
    Also a good point - we've added mention of this as well. \revref
}

\begin{point}{}
    Population isolates and genetic load: Isn't this exactly ``founder effect
speciation''?
\end{point}

\reply{
    We have removed this part.
}

\begin{point}{}
    Your analysis of the fitness loss in F1 and F2 hybrids rests on a local
Taylor expansion of the fitness landscape. This is adequate given that the
underlying linear network is also only locally valid (see above). However, you
then apply this to a discussion of hybrid incompatibility (Haldane's rule etc).
We are thus interested in \emph{very large} fitness costs for hybrids. This does not
seems to be compatible: to get hybrid incompatibility or sterility, you need to
have epsilon sufficiently large (below equation 5). But then $\epsilon^2$ is no
longer smaller than $\epsilon$. In other words: it seems to me that for the
discussion of hybrid incompatibilities you apply your model of the first part
beyond the local range where it is valid.
\end{point}

\reply{
    This is a good point, but tricky since it's not clear what ``small'' is here.
    A fitness difference of 0.2 is quite strong selection, for instance,
    although it is a fairly small number in some ways.
    Furthermore, the same picture holds in the opposing cases of a single locus
    or many loci, illustrated in Figure~\ref{fig:conceptual_fig}.
    In any case, we now rely only on the conceptual idea rather than the numerics of this estimate.
}

%%%%%%%%%%%%%%
\reviewersection{2}

\begin{quote}
The present paper is divided into two parts. In the first part, the authors propose a framework
in which distinct genetic architectures can produce the same phenotype. (In the fitness landscape
terminology, this would correspond to the existence of an evolutionary ridge in a fitness landscape.)
The starting point is a nice analogy with linear systems theory. The authors highlight the fact
that in general, two distinct linear (differential) systems can respond identically for any input; i.e.,
different systems (genotypes) can always produce the same output (phenotype) given the same input
(environment). Further, the set of equivalent systems can be nicely characterized through Kalman
decomposition, thus providing a nice characterization of level sets in the underlying fitness landscape.
All along the first part of the paper, this analogy is well illustrated using a simple (yet quite nice)
example of oscillating gene transcription. I especially like Fig. 5 where it is shown that even if $F_1$
hybrids have a phenotypic response close to their parents, $F_2$ hybrids can behave in a drastically
different manner.

In the second part of the paper, motivated by the previous results, the authors explore a general
quantitative genetics model in which populations can drift stochastically near a set of equivalent
and optimal systems. Since the optimal set (or evolutionary ridge) is not closed under averaging or
recombination, two isolated populations can drift apart and accumulate enough genetic differences
so that they do not produce any viable offspring. Using some heuristics, several expressions are
derived to quantify the accumulation of genetic incompatibilities.

Overall, I think the paper is well written. The framework developed in the first part of the paper
is very interesting and that the analogy with system theory is quite enlightening. I am somehow
less convinced by the second part. First, I find the arguments a bit sketchy (see below) and not so
easy to follow. Secondly, it is not entirely clear to me what is the main contribution of this part
compared to previous works. It seems to me that the main result is somehow contains in the fact
that the variance (or ``segregation variance'') of an $F_2$ population is given by
$$\sigma^2_S + 4 \omega \sigma^2_N T / N_e$$
which was already derived Slatkin and Lande according to the authors (except for the explicit
expression of $\omega$, but again I am a little bit confused by the arguments derived in the appendix).

In summary, I think this paper could be a nice contribution to Evolution. However, I am also
convinced that the second part of the paper would require more work, or more explicit reference
to previous works (equation, section etc.) if the authors do not want to re-derive already existing
formula.
\end{quote}

Thanks for the encouraging words.
As discssed above, we've separated out the second (less convincing) part,
and the manuscript now only covers the first part.

\begin{point}{end of p6:} 
    ``we assume ... $((A + A' )/2, (B + B' )/2, (C + C' )/2)$'' . It is claimed \revreffull{2}{1}
    that two kryptotypes need not have the same dimension. In this
    case, the previous sum does not make sense, right?
\end{point}

\reply{
    Good point; added that caveat. \revref
}

\begin{point}{p7: fig 4.}
    ``The distribution of $F_2$ ... homozygotes''. I do not really grasp the meaning of this
sentence. More importantly, I find the purple cloud of dots ($F_2$ population) quite confusing.
It seems to me that the purple populations should be a cloud of points concentrated around
a point sitting on the red dotted line, i.e., a Gaussian distribution with the mean at the
average of the two parental populations. According to the authors, the distribution is
bimodal with a peak below the optimal set (as displayed on the figure) and another distinct
peak sitting on the other side of the optimal. In fact, if I understand the computations of
the second part correctly, the red and purple means should coincide, but the red variance
should simply be greater.
\end{point}

\reply{
    We have improved Figure~\ref{fig:conceptual_fig} and made the accompanying discussion
    more concrete to hopefully better demonstrate the point.
}

\begin{point}{p7: end of the page.}
    Why do you need $\rho$ to be square integrable?
\end{point}

\reply{
    Good point -- clearly some assumptions jointly on the long-time behavior of $h(t)$ and $\rho$ 
    are required for $D$ to be finite, but just saying that $\rho$ is square integrable doesn't suffice;
    we've removed the caveat. 
    \revref
    (It is always well-defined, anyhow, since the integrand is nonnegative.)
}

\begin{point}{p8:}
    ``A Taylor expansion of $D(h_\epsilon)$ .. ''. It is not clear to me at all. Could you provide some
extra explanation (e.g., in the appendix)?
\end{point}

\reply{
    We have put back in the somewhat lengthy Appendix on this subject 
    that we'd removed before submission. \revref
}

\begin{point}{p9. Fig 5.} 
    In the left panel, it seems to me that the two parents are homozygotes since there
is a single $F_1$ possible offspring (dashed blue curve). It would be worth being more explicit.
For the right panels, could you be more explicit on the number of curves. If I understand
correctly, there are 16 possibilities due to recombination (2 per entries of the matrix) since
$F_1$ individuals are heterozygotes. Is that correct?
Finally, the labels on the y axis are not easy to read.
\end{point}

\reply{
    We've added ``homozygous'' to the caption,
    and mentioned that there are $3^4 = 81$ possible $F_2$s -- 
    this is because it only matters if, at each of the four matrix entries, 
    the offspring is heterozygous, homozygous for parent 1, or homozygous for parent 2.
}

\begin{point}{p11. Paragraph system drift.}
    ``move a random distance $\sigma$''. What is $\sigma$? I think it should
be $\sigma_N$ (the std deviation in the direction of the evolutionary ridge) to be consistent with
the assumption that the population drifts along the optimal set. I believe that this is what
is assumed thereafter. Also, the sentence ``It therefore seems .... as cloud of points of width
$\sigma$'' is not very accurate, since the covariance matrix is not the identity.
\end{point}

\reply{
    This part has been removed.
}

\begin{point}{p 11.}
    Approximating the optimal set $\mathcal{N}$ by a quadratic surface should only be accurate if
we look at the genetic divergence at small time scales. This should be at least mentioned.
\end{point}

\reply{
    This part has been removed.
}

\begin{point}{end of p11.}
    $1/( \frac{d}{du} D(x + uz))$ should be evaluated at $u = 0$.
\end{point}

\reply{
    This part has been removed.
}

\begin{point}{p12. Third paragraph}
    $\sqrt{4 \omega T/N_e} \sim \gamma/\sigma_N$.
    I guess the underlying assumption here is that
    $\sigma_S \ll \sqrt{4 \omega \sigma^2_N T/N_e}$?
\end{point}

\reply{
    This part has been removed.
}

\begin{point}{p12. before eq. 7.}
    $\mu = c_\mu \gamma T /N_e$. Why is $\mu$ proportional to $\gamma$?
\end{point}

\reply{
    This part has been removed.
}

\begin{point}{Fig 7.}
    I have one important issue with this figure (and the assumptions of the underly-
ing quantitative genetic model). If one wants to be consistent with the assumption that
parental populations drift along the evolutionary ridge, I think one would need to assume
that selection is strong enough to constraint the mean of the population on the surface. This
would presumably require that $\sigma S /\gamma \ll 1$. First, I think this assumption (or something
alternative to that) should be made explicit in the text. Secondly, it seems to me that this
assumption is not satisfied for panel A and C: under the range of parameters proposed by
the authors, the parental populations could easily drift away from the optimal set, and in
particular, the heuristics derived in the main text would not be satisfied.
\end{point}

\reply{
    This part has been removed.
}

\begin{point}{}
    Finally, it would be worth mentioning several old and recent works relating genetic drift
    to speciation: \citet{yamaguchi2013first}, several articles by Gavrilets et al. (I think several
    citations are missing here), and \citet{pina2019does}.
\end{point}

\reply{
  Thank you for the references. We've added citations to the Introduction. \revref
}

### 2022-03-08 writeups/context-dependence.tex
fixing the cover letter and making a suggestion

@@
         which is how our approximation to the process is parameterized.
-        The central, smaller segment of the T is called the \emph{base} of the T-mer,
-        and the key building block computes the transition probability
+        The central, smaller segment of the T is called the \emph{base} of the T-mer.
+        The key building block for our method computes the transition probability
         for this subsequence
@@
 %%%%%
-\subsection{Choice of overhang}
+\ubsection{Choice of overhang}
 

### 2022-03-08 writeups/reviews/reviews.v1.tex
fixing the cover letter and making a suggestion

@@
   Resubmission Cover Letter \\
-  {\it Evolution}
+  {\it Journal of Computational Biology}
 \end{minipage}
@@
 \begin{minipage}[b]{2.5in}
-    Joshua Schiffman \\
+    Erick Matsen \\
     \emph{and} Peter Ralph \\
@@
 
-Our revised manuscript.
+Thank you for your evaluation of our manuscript
+``Enabling inference for context-dependent models of mutation by bounding the propagation of dependency,''
+and thank you to the reviewers for their constructive feedback.
+
+Please find attached our revised manuscript, which includes a ``diff'' of the manuscript with the previous version.
+We have fully addressed the reviews, which have resulted in a number of important improvements and fixes.
+
+\vskip 1em
 
@@
 
+%EM This one looks a little out of place. I think it'd be clearer to just comment it out.
 \begin{quote}

### 2022-03-08 writeups/context-dependence.tex
A bit of space and some suggestions

@@
 \begin{align} \label{eqn:full_likelihood}
-    p_n(t;x,y) := \P\{ X(t) = y \mid X(0) = x \} = {\left(e^{tG(L)}\right)}_{xy} ,
+    p_n(t;\,x,y) := \P\{ X(t) = y \mid X(0) = x \} = {\left(e^{tG(L)}\right)}_{xy} ,
 \end{align}
@@
 is clearly infeasible even for moderately sized $L$.
-We can, however, compute \eqref{eqn:full_likelihood} with smaller $G(n)$,
+We can, however, compute \eqref{eqn:full_likelihood} using $G(n)$ for some smaller $n$,
 so the first thing that one might think to do is to break the sequence up into many blocks of length $m$,
 and treat these as independent.
-Then, defining $p_m(t;x,y)$ to be the probability that a string $x$ of length $m$ evolves to string $y$ over time $t$,
+Then, defining $p_m(t;\,x,y)$ to be the probability that a string $x$ of length $m$ evolves to string $y$ over time $t$,
 we would obtain the approximate likelihood function
 \[
-  \prod_{k=0}^{n/m} p_m(t;x_{km+1}^{(m)},y_{km+1}^{(m)}) .
+  \prod_{k=0}^{n/m} p_m(t;\,x_{km+1}^{(m)},y_{km+1}^{(m)}) .
 \]
@@
 at least approximately.
-To do this, let $p_{n,\ell,r}(t;x,y)$ for $\ell + r < n$ denote
+To do this, let $p_{n,\ell,r}(t;\,x,y)$ for $\ell + r < n$ denote
 the probability that an initial subsequence $x$ of length $n$, after time $t$,
@@
 
-These $p_{n,\ell,r}(t;x,y)$ probabilities can be found by marginalizing $p_n$,
+These $p_{n,\ell,r}(t;\,x,y)$ probabilities can be found by marginalizing $p_n$,
 summing over the possible initial and final substrings of $x$
@@
 \begin{align}
-    p_{n,\ell,r}(t;x,y)
+    p_{n,\ell,r}(t;\,x,y)
     &=
-    \sum_{a \in \calS^\ell} \sum_{b \in \calS^r} p_n(t;x,a \join y \join b) ,
+    \sum_{a \in \calS^\ell} \sum_{b \in \calS^r} p_n(t;\,x,a \join y \join b) ,
     \quad
@@
 where $a \join y \join b$ is the sequence composed of concatenating $a$, $y$, and $b$ together in that order.
-We expect $p_{n,\ell,r}(t;x,y)$ to be asymptotically correct when $\ell$ and $r$ are large (and $t$ is fixed).
+We expect $p_{n,\ell,r}(t;\,x,y)$ to be asymptotically correct when $\ell$ and $r$ are large (and $t$ is fixed).
 In Appendix~\ref{ss:approx_pf}, we show that this is so, namely that
@@
 is later observed to have its center matching $y$
-is close to $p_{n,\ell,r}(t;x,y)$, regardless of wider context.
+is close to $p_{n,\ell,r}(t;\,x,y)$, regardless of wider context.
 Formally, for each $i$,
@@
     \approx
-    p_{n,\ell,r}(t;x,y)
+    p_{n,\ell,r}(t;\,x,y)
     ,
@@
         =
-        \prod_{i=1}^{L-m+1} p_{n, \ell, r}\left(t;x_i^{(n)}, y_{i+\ell}^{(m)}\right).
+        \prod_{i=1}^{L-m+1} p_{n, \ell, r}\left(t;\,x_i^{(n)}, y_{i+\ell}^{(m)}\right).
 \end{align}
@@
         \frac{
-            p_{4\ell+1,\ell,\ell}(t; x_{i-3\ell}^{(4\ell+1)}, y_{i-2\ell}^{(2\ell+1)})
+            p_{4\ell+1,\ell,\ell}(t;\, x_{i-3\ell}^{(4\ell+1)}, y_{i-2\ell}^{(2\ell+1)})
         }{
-            p_{4\ell+1,\ell,\ell+1}(t; x_{i-3\ell}^{(4\ell+1)}, y_{i-2\ell}^{(2\ell)})
+            p_{4\ell+1,\ell,\ell+1}(t;\, x_{i-3\ell}^{(4\ell+1)}, y_{i-2\ell}^{(2\ell)})
         }.
@@
 %     &\approx
-%         \sum_{a \in \calS^n} \sum_{b \in \calS^{n-2\ell}} n_\ell(x,y;a,b) \log p_{n,\ell,\ell}(t;a,b)
+%         \sum_{a \in \calS^n} \sum_{b \in \calS^{n-2\ell}} n_\ell(x,y;a,b) \log p_{n,\ell,\ell}(t;\,a,b)
 %         -
-%         \sum_{c \in \calS^n} \sum_{d \in \calS^{n-2\ell-1}} n_\ell(x,y;c,d) \log p_{n,\ell,\ell+1}(t;c,d)
+%         \sum_{c \in \calS^n} \sum_{d \in \calS^{n-2\ell-1}} n_\ell(x,y;c,d) \log p_{n,\ell,\ell+1}(t;\,c,d)
 % \end{align}
@@
 Above, we have chosen to disallow changes whose patterns extend beyond the end of the sequence,
+%EM This i.e. seems out of place
 i.e., the pattern $u$ (and hence, also the replacement $v$) of each transition triple
@@
 \begin{align}
-    p_{n}(t;a,c) = \left( e^{t G(n)} \right)_{a,c} ,
+    p_{n}(t;\,a,c) = \left( e^{t G(n)} \right)_{a,c} ,
 \end{align}
@@
 \begin{align} \label{eqn:Tmer_trans}
-    p_{n,\ell,r}(t;a,b) = \sum_{c \in \calS^{n}} \left( e^{t G(n)} \right)_{ac} U_{cb} ,
+    p_{n,\ell,r}(t;\,a,b) = \sum_{c \in \calS^{n}} \left( e^{t G(n)} \right)_{ac} U_{cb} ,
 \end{align}
@@
 (The expected number is the number of occurrences of $x$ in the initial sequence
-multiplied by the probability $p_{n,\ell,r}(t;x,y)$, where $n$ is the length of $x$,
+multiplied by the probability $p_{n,\ell,r}(t;\,x,y)$, where $n$ is the length of $x$,
 and $\ell$, $r$ are the appropriate overhangs.)
@@
 such as noisy images or particle systems used to model other physical processes.
+%EM I suggest taking out "used"
 However, computational costs will grow with the size of the local neighborhood used
@@
 Dr.\ Matsen is an Investigator of the Howard Hughes Medical Institute.
-PLR's work was supported by NIH award R01HG010774
+PLR's work was supported by NIH award R01HG010774.
 

### 2022-03-09 writeups/context-dependence.tex
final fix-ups

@@
 and that multinucleotide substitutions are relatively common \citep{schrider2011pervasive,terekhanova2013prevalence,harris2013errorprone}.
+\citet{dewitt2021nonparametric} has even identified how these mutational signatures in humans
+have varied over recent time.
 In humans, the molecular processes underlying mutations are best studied in contexts of intense somatic mutation:
@@
 so do not consider further the many constraints on protein-coding sequence \citep[reviewed in][]{thorne2007protein}.
-Some of these effects -- particularly, activity levels of different enzymes in the germline --
-will likely vary over evolutionary time,
-and so comparisons between widely separated species will identify time-averaged mutation rates.
+% Some of these effects -- particularly, activity levels of different enzymes in the germline --
+% will likely vary over evolutionary time,
+% and so comparisons between widely separated species will identify time-averaged mutation rates.
 Studies have used a variety of exploratory techniques to identify sets of mutational patterns
-co-occurring in comparisons (a) between different cancer types \citep{alexandrov2013signatures,shiraishi2015simple} and
-(b) between different human populations \citep{harris2015evidence,harris2016rapid,mathieson2017differences},
+co-occurring in comparisons between different cancer types \citep{alexandrov2013signatures,shiraishi2015simple} and
+between different human populations \citep{harris2015evidence,harris2016rapid,mathieson2017differences}
 to disentangle distinct signals presumably coming from the action of distinct sources of mutation.
@@
 \citet{lunter2004nucleotide} used an approximate matrix decomposition;
-while \citet{siepel2004phylogenetic} and later \citep{baele2010modelling} computed likelihoods by assuming a model that is Markov along the genome.
+while \citet{siepel2004phylogenetic} and later \citet{baele2010modelling} computed likelihoods by assuming a model that is Markov along the genome.
 However, all these methods are quite computationally intensive.
@@
 relaxing into states dependent on the energy of the resulting configuration.
-Parameter estimation for Ising model without temporal dynamics
+Parameter estimation for the Ising model without temporal dynamics
 is relatively well-understood \citep{pickard1982inference,frigessi1990parameter},
@@
 
-This general class of models are known in the probability literature as
+This general class of models is known in the probability literature as
 interacting particle systems \citep{liggett2005ips} with neighborhood structure
@@
 \begin{example}[TASEP]
-  In the \emph{Totally Asymmetric Simple Exclusion Process}
+  The \emph{Totally Asymmetric Simple Exclusion Process}
   describes a set of particles moving among some sites,
@@
     in the most general model of independent mutation across sites, each of the 12 possible transitions occurs at its own rate.
-    Furthermore, in many species
+    Furthermore, in many species,
     adjacent, methylated CG dinucleotides (``CpG sites'') have a much higher mutation rate to TG and CA
@@
     in a population of effective size $N_e$ becomes ubiquitous, rather than dying out,
-    with probability approximately $(1-\exp(-2 s))/(1-\exp(-2 s N_e))$ \citep{kimura1962probability}.
+    with probability approximately $(1-\exp(-2 s))/(1-\exp(-4 s N_e))$ \citep{kimura1962probability}.
     If the mutation occurs at rate $\mu$ per individual,
-    the total rate it occurs at in the population is $\mu N_e$.
-    GC-biased gene conversion effectively means that sequences with more G/Cs are selected for.
+    the total rate it occurs at in the population is $2 \mu N_e$.
+    The effect of GC-biased gene conversion can be modeled as a selective pressure in favor of G/Cs.
     This is incorporated as follows: set the energy of patterns \nC{} and \nG{} to be $s$,
@@
     \begin{align*}
-        \phi(e) = N_e (1-\exp(-2e))/(1-\exp(-2eN)).
+        \phi(e) = 2 N_e (1-\exp(-2 e))/(1-\exp(-4 e N_e)).
     \end{align*}
@@
 \]
-Note that we do not actually need to construct $G(L)$ to find the maximum rate.
+Note that we do not actually need to construct $G(L)$ to find $\mu^*$:
+since $\mu^*$ must only be greater than this maximum value (not necessarily equal to it),
+we can take $\mu^*$ to be $L$ multiplied by the maximum ``per-site'' rate.
 Now sample possible changes:
@@
 \]
-(Note that this is equivalent to saying that possible
-changes occur at rate $\mu^*$ independently at each of the $L$ sites.)
+(Note that this is equivalent to placing a Poisson process of possible
+changes at rate $\mu^*$ independently at each of the $L$ sites.)
 Now, suppose we have determined the state just before time $s_k$ to be $X(s_{k-1}) = x$.
@@
 the probability that an initial subsequence $x$ of length $n$, after time $t$,
-is found to match a smaller subsequence $y$ of length $n-\ell-r$, offset by $\ell$ sites
-(see Figure \ref{fig:Tmers}).
-This central, smaller subsequence is called the ``base'' of the T-mer.
-We refer to such patterns as \emph{T-mers}, as depicted in Figure~\ref{fig:Tmers}.
-We emphasize that the overhang of the T-mers is typically bigger than the
+is found to match a smaller subsequence $y$ of length $n-\ell-r$, offset by $\ell$ sites.
+We refer to such patterns as \emph{T-mers}, as depicted in Figure~\ref{fig:Tmers},
+and this central, smaller subsequence is called the ``base'' of the T-mer.
+The overhang of the T-mers is typically bigger than the
 context-sensitivity width of the underlying model (Figure~\ref{fig:overview}).
@@
 to deal properly with these,
-on the right-hand side of the equation $\ell$ should be replaced with $\min(\ell, \max(i-1, 0))$,
-and $r$ should be replaced with $\min(r, \max(n-i-\ell-m+1, 0))$.
+$\ell$ and $r$ should be made smaller near the left and right ends of the sequence, respectively
+% on the right-hand side of the equation $\ell$ should be replaced with $\min(\ell, i-1)$,
+% and $r$ should be replaced with $\min(r, L-i-\ell-m+1)$.
 
@@
 %%%%%
-\ubsection{Choice of overhang}
+\subsection{Choice of overhang}
 
@@
 and if the sequence has evolved for time $t$ then we expect no more than a Poisson($\lambda t$) number of changes at each site.
-If $\lambda t = 0.75$ ($1-e^{-3/4}=52\%$ of the sequence will have changed!),
+If $\lambda t = 0.75$ (in which case $1-e^{-3/4}=52\%$ of the sequence will have changed!),
 then taking $\ell = r = 3$ allows us to compute T-mer probabilities to within an error of 0.007,
 since this is the probability that a Poisson with mean 0.75 will be larger than 3.
-If $\lambda t = 0.25$ we would do even better with $\ell = r = 2$.
+If $\lambda t = 0.25$ we would do even better with only $\ell = r = 2$.
 
@@
 
-For inference, we need to compute the $|\calS|^{n} \times |\calS|^m$ matrix $F$ whose $(a,b)^\text{th}$ entry is $p_{n,\ell,r}(t,a,b)$,
-for various $t$ (and $m = n - \ell - r$).
+For inference, we need to compute the $|\calS|^{n} \times |\calS|^m$ matrix $F$ whose $(a,b)^\text{th}$ entry is $p_{n,\ell,r}(t;a,b)$,
+for various $t$.
 This matrix is a projection of the $|\calS|^{n} \times |\calS|^{n}$ matrix whose $(a,c)^\text{th}$ entry is
@@
 where $G(n)$ is the sparse $|\calS|^{n} \times |\calS|^{n}$ matrix defined in \eqref{eqn:G_defn}.
-The ``projection'' we need just marginalizes over long (i.e., length $n$) patterns $c$ that match the shorter (i.e., length $m$) pattern $b$
+The ``projection'' we need just marginalizes over long (i.e., length $n$) patterns $c$ that match the shorter
+(i.e., length $m=n - \ell - r$) pattern $b$
 as in \eqref{eq:pmMarg}:
@@
 We might approximate the full likelihood as follows,
-after choosing $n$, $\ell$, and $r$ appropriately (see above).
+after choosing $n$, $\ell$, and $r$ as above.
 For the sake of notational simplicity, assume that we have a few positions
-of $X$ observed before the start and after then end of the sequence
+of $X$ observed before the start and after the end of the sequence
 (so that, e.g., we can refer to $x_{n+1}, \ldots, x_{n+r}$).
@@
   \begin{split}
-    & \P\{X_{I-\ell}^{n}=x \text{ and } Y_I^\ell=y \text{ and } Z_I^\ell=z \} \\
+      & \P\{X_{I-\ell}^{(n)}=x \text{ and } Y_I^{(m)}=y \text{ and } Z_I^{(m)}=z \} \\
     &\qquad
@@
 for certain (more common) patterns: for instance,
-one could compute the likelihood for situations where $y$ and $z$ agree (but may differ from $x$
+one could compute the likelihood for situations where $y$ and $z$ agree (but may differ from $x$)
 by replacing step (4) with:
@@
 such as AID \citep{senavirathne2015activationinduced} or error-prone polymerases involved in repair \citep{maul2016polymerase}.
-
-Nevertheless, we find significant effects for a number of biochemically-informed mutation features (Figure~\ref{fig:biochem_results}).
-Specifically, we find strong CpG and UV damage effects \citep[][also seen by]{harris2015evidence}.
 It may seem tenuous to look for mutational signatures associated with cancer in the germline,
@@
 \citep{arbeithuber2015crossovers,myers2010drive}.
+
+The strongest context-specific signals we find are a high CpG rate (Figure~\ref{fig:biochem_results}),
+and a strong \emph{negative} rate for the UV damage motif.
+A negative estimate implies that the corresponding motifs mutate \emph{less}
+than expected based on other rates -- so, for instance,
+a lower \nG\nA$\to$\nA\nA\ rate than expected based on the \nG$\to$\nA\ rate,
+perhaps because the latter rate was estimated to be particularly high.
+Notably, we find a small but significant rate of UV-associated dimer formation,
+also seen by \citet{harris2015evidence}.
+All rates were statistically significantly nonzero except those associated with pol $\iota$.
 
@@
 \paragraph{Modeling}
-Careful building of complex models in this framework may take care,
-but no more than in any other class of models (e.g., multivariate linear regression).
+Careful building of complex models in this framework can be difficult,
+as in variable selection more generally  (e.g., multivariate linear modeling).
 It is easy to specify nonidentifiable models --
@@
         -
-        p_{n,\ell,r}(x,y)
+        p_{n,\ell,r}(t;x,y)
     \right|
@@
     Now, $\E[f(N)] = \lambda^{k+1} / (k+1)!$ via direct calculation or the
-    formula for the $k+1$st factorial moment for the Poisson distribution.
+    formula for the $k+1^\text{st}$ factorial moment for the Poisson distribution.
 \end{proof}

### 2022-03-09 writeups/context-dependence.tex
.

@@
 There is no conceptual barrier to applying these methods to lattices in higher dimensions,
-such as noisy images or particle systems used to model other physical processes.
-%EM I suggest taking out "used"
+such as noisy images or particle systems modeling other physical processes.
 However, computational costs will grow with the size of the local neighborhood used

### 2022-04-19 writeups/context-dependence.tex
biblatex version

@@
-\documentclass{article}
+\documentclass[english]{article}
 \usepackage{fullpage}
@@
 \usepackage{graphicx}
-\usepackage{natbib}
+\usepackage{babel}
+\usepackage[natbib=true,style=authoryear,backend=biber]{biblatex}
 \usepackage[hidelinks]{hyperref}
@@
 \renewcommand\Affilfont{\itshape\small}
+
+\addbibresource{context-dependence.bib}
 
@@
 
-\bibliographystyle{plainnat}
-
 % http://bytesizebio.net/2013/03/11/adding-supplementary-tables-and-figures-in-latex/
@@
 
-% Responses to reviews
-\input{reviews/review-response-commands}
-% set this to show line numbers and include responses to reviews or not
-\newif\ifreviewresponses
-\reviewresponsestrue  % include them
-% \reviewresponsesfalse  % don't include them
-\newcommand{\responsefile}{reviews/reviews.v1.tex}  % name of the review responses file
-
-\newif\ifsubmission
-\submissiontrue
-% \submissionfalse
-\ifsubmission
-    % % stuff for submission
-    \usepackage{lineno}
-    % allow line numbers around math environments
-    % from https://tex.stackexchange.com/questions/43648/why-doesnt-lineno-number-a-paragraph-when-it-is-followed-by-an-align-equation
-        \newcommand*\patchAmsMathEnvironmentForLineno[1]{%
-          \expandafter\let\csname old#1\expandafter\endcsname\csname #1\endcsname
-          \expandafter\let\csname oldend#1\expandafter\endcsname\csname end#1\endcsname
-          \renewenvironment{#1}%
-             {\linenomath\csname old#1\endcsname}%
-             {\csname oldend#1\endcsname\endlinenomath}}%
-        \newcommand*\patchBothAmsMathEnvironmentsForLineno[1]{%
-          \patchAmsMathEnvironmentForLineno{#1}%
-          \patchAmsMathEnvironmentForLineno{#1*}}%
-        \AtBeginDocument{%
-        \patchBothAmsMathEnvironmentsForLineno{equation}%
-        \patchBothAmsMathEnvironmentsForLineno{align}%
-        \patchBothAmsMathEnvironmentsForLineno{flalign}%
-        \patchBothAmsMathEnvironmentsForLineno{alignat}%
-        \patchBothAmsMathEnvironmentsForLineno{gather}%
-        \patchBothAmsMathEnvironmentsForLineno{multline}%
-        }
-\else
-    % for the nice version
-\fi
-
 \begin{document}
-
-% TODO after submission: think about the BCR parameter regime.
 
@@
 
-\subsection{Acknowledgements}
+\section*{Acknowledgements}
 The initial impetus for this project came from Matt Dean;
@@
 
+\section*{Author contributions}
+PLR designed the method and wrote the first drafts of the code.
+PLR and FAM tested the code and wrote the paper.
+
+\section*{Authors' disclosure}
+The authors declare no conflict of interest.
+
+\section*{Funding statement}
 FAM's work supported by NIH grants R01 AI146028 and U01 AI150747.
@@
 
-\bibliography{context-dependence}
+\printbibliography
 

### 2022-04-25 writeups/context-dependence.tex
submission template

@@
-\documentclass[english]{article}
-\usepackage{fullpage}
+\documentclass[english]{Liebert_Author_revised}
 \usepackage{amsmath,amssymb}
@@
 \usepackage{babel}
-\usepackage[natbib=true,style=authoryear,backend=biber]{biblatex}
-\usepackage[hidelinks]{hyperref}
+\usepackage{natbib}
 \usepackage{bm}
-\usepackage{todonotes}
-\usepackage{authblk}
-\renewcommand\Affilfont{\itshape\small}
-
-\addbibresource{context-dependence.bib}
-
-\newcommand{\peter}[1]{\todo[color=green!40]{Peter: #1}}
-\newcommand{\erick}[1]{\todo[color=purple!40]{Erick: #1}}
 
@@
 
-\theoremstyle{plain}
-\newtheorem{theorem}{Theorem}
-\newtheorem{lemma}{Lemma}
 \theoremstyle{definition}
@@
 \title{Enabling inference for context-dependent models of mutation by bounding the propagation of dependency}
-\author[2,3,4,5]{Frederick A Matsen IV}
-\author[1,*]{Peter L Ralph}
-
-\affil[1]{Institute of Ecology \& Evolution, Department of Biology, University of Oregon}
-\affil[2]{Computational Biology Program, Fred Hutchinson Cancer Research Center, United States}
-\affil[3]{Department of Genome Sciences, University of Washington}
-\affil[4]{Department of Statistics, University of Washington}
-\affil[5]{Howard Hughes Medical Institute}
-\affil[*]{Corresponding author: plr@uoregon.edu}
+\author{
+    Frederick A Matsen IV$^{2,3,4,5}$,
+    Peter L Ralph$^{1,*}$\\
+$^1$Institute of Ecology \& Evolution, Department of Biology, University of Oregon \\
+$^2$Computational Biology Program, Fred Hutchinson Cancer Research Center, United States \\
+$^3$Department of Genome Sciences, University of Washington \\
+$^4$Department of Statistics, University of Washington \\
+$^5$Howard Hughes Medical Institute \\
+$^*$To whom correspondence should be addressed: plr@uoregon.edu
+}
 
@@
 and for those positions $x$ matches with $u$ while $y$ matches with $v$.
-\revpoint{2}{3}
 
@@
 or ``uniformization'' \citep{hobolth2009simulation}.
-\revpoint{2}{5}
 Briefly, this  works by first sampling a homogeneous Poisson process of times and locations of possible changes,
@@
 \end{align}
-\revpoint{2}{6}
 
@@
 
-\paragraph{Boundary conditions} \revpoint{2}{1}
+\paragraph{Boundary conditions}
 Above, we have chosen to disallow changes whose patterns extend beyond the end of the sequence,
@@
 However, computational costs will grow with the size of the local neighborhood used
-even faster than in one dimension. \revpoint{2}{2}
+even faster than in one dimension.
 
@@
 
-\printbibliography
+\bibliographystyle{plainnat-revised}
+\bibliography{context-dependence}
 
@@
 
-%%%%%%%%%%%%%%%%%
-\includereviews
-
 \end{document}
